### 2024-11-20 main.tex
add permutation tree

\documentclass{article}

\usepackage[utf8]{inputenc}
\usepackage[margin=0.8in]{geometry}

\usepackage{color}
\usepackage{amsmath}
%\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage[all]{xy}
\usepackage{multirow}
\usepackage{paralist}
\usepackage{hhline}
\usepackage{bm}

\usepackage{wasysym}
\usepackage{extarrows}
\usepackage{tikz}

\input{notations}

% add new commands for comments here
\newcommand{\yx}[1]{\textit{\color{blue}[YX] : #1}}

\newcommand{\modify}[1]{{\color{red}#1}}

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{stmaryrd}
\usepackage{hyperref}

\usepackage{braket}

\title{\textbf{DiracDec C++/Coq Implementation\cite{ExampleCitation}}}
\author{Yingte Xu}

\begin{document}

\maketitle

%%%%%%%%%%%%%%%%
\section{Scalar: Minimal Example}

We first consider a minimal example of the scalars.

\subsection{Modeling AC by Sequence}
The problem is presented in the normal AC term-rewriting setting, with a minor modification of flattening the AC symbols. This is the model implemented in Coq, and we consider the 
C++ optimization only as a efficient searching algorithm for the rewriting trace.

\newcommand{\ZERO}{\mathsf{0}}
\newcommand{\ONE}{\mathsf{1}}
\newcommand{\ADDS}{\mathsf{ADDS}}
\newcommand{\MLTS}{\mathsf{MLTS}}
\newcommand{\CONJ}{\mathsf{CONJ}}

\begin{definition}[scalar syntax]
    The syntax for scalars is defined as follows.
    \[ 
    S ::= \ZERO\ |\ \ONE\ |\ \ADDS ( S^+ )\ |\ \MLTS ( S^+ )\ |\ \CONJ(S).
    \]
    Here $S^+$ means a sequence of one or more $S$ terms, i.e., AC symbols are flattened and have indefinite arities.
\end{definition}

The rewriting rules are presented as follows.

\begin{figure}[h]
    \begin{align*}
        & \text{(R-FLATTEN)} &&
        \textcolor{gray}{\text{A special rule to flatten all AC symbols within one call.}} \\
        & \text{(R-ADDSID)} &&
        \ADDS(a) \to a \\
        & \text{(R-ADDS0)} &&
        \ADDS((seq:\_ \_)\ \ZERO) \to \ADDS(seq) \\
        & && \textcolor{gray}{\text{This rule removes all $\ZERO$ occurances and keeps the order of remaining subterms.}} \\
        & \text{(R-MLTSID)} &&
        \MLTS(a) \to a \\
        & \text{(R-MLTS0)} &&
        \MLTS((seq: \_ \_)\ \ZERO) \to \ZERO \\
        & \text{(R-MLTS1)} &&
        \MLTS((seq: \_ \_)\ \ONE) \to \MLTS(seq) \\
        & && \textcolor{gray}{\text{Similar to (R-ADDS0).}} \\
        & \text{(R-MLTS2)} &&
        \MLTS((seq1: \_ \_ \_)\ \ADDS(a_1\ a_2 \cdots a_n)\ (seq2: \_ \_ \_)) \\
        & && \to \ADDS(\MLTS(seq1\ a_1\ seq2)\ \MLTS(seq1\ a_2\ seq2) \cdots \MLTS(seq1\ a_n\ seq2)) \\
        & && \textcolor{gray}{\text{This rule matches the first $\ADDS$ subterm in the list. $seq1$ or $seq2$ can be empty, but not both.}} \\
        & \text{(R-CONJ0)} &&
        \CONJ(\ZERO) \to \ZERO \\
        & \text{(R-CONJ1)} &&
        \CONJ(\ONE) \to \ONE \\
        & \text{(R-CONJ2)} &&
        \CONJ(\ADDS(a_1\ a_2\ \cdots a_n)) \to \ADDS(\CONJ(a_1)\ \CONJ(a_2) \cdots \CONJ(a_n)) \\
        & \text{(R-CONJ3)} &&
        \CONJ(\MLTS(a_1\ a_2\ \cdots a_n)) \to \MLTS(\CONJ(a_1)\ \CONJ(a_2) \cdots \CONJ(a_n)) \\
        & \text{(R-CONJ4)} &&
        \CONJ(\CONJ(a)) \to a
    \end{align*}
\end{figure}

Notice that all the rules behaves the same as in the term count mapping implementation in C++. That is, the transformation from sequence representation to mapping representaiton commutes with all rewritings.

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Rewriting Control and Intermediate Language}
The associativity is already handled by the (R-FLATTEN) rule. In order to decide two terms $A$ and $B$ are equivalent under commutativity, we need to proof that $A$ can be transformed into $B$ with a structured permutation, which is described by the \textit{permutation tree}.

\begin{definition}[permutation tree]
    The syntax for permutation trees are inductively defined below:
    \[
    P ::= \mathsf{E}\ |\ [(i:P)^+].
    \]
    Here $i$ represents positive numbers.
\end{definition}
We always only consider \textit{well-formed} permutation trees. That is, if $P \equiv [i_1:P_1\ i_2:P_2\ \cdots\ i_n:P_n]$, then $\{i_1, ... i_n\}$ forms the set of integers from $0$ to $n-1$.

We can transform a term A with a suitable permutation tree. The transformation is defined as

\begin{align*}
    \texttt{apply}(A, P) :=\ & \texttt{match}\ P\ \texttt{with} \\
    & \quad |\ \textsf{E} \Rightarrow A \\
    & \quad |\ [i_1:P_1\ \cdots\ i_n:P_n] \Rightarrow \texttt{A.head}(\texttt{apply(A.args[$i_1$], $P_1$)}\ \cdots\ \texttt{apply(A.args[$i_n$], $P_n$)}) \\
    & \texttt{end}
\end{align*}

% \subsection{Coq Implementation}
% The following abilities should be implemented on the Coq side:
% \begin{itemize}
%     \item \textbf{Deciding $E$}. For the C++ implementation, the data structure itself already reflects $E$. Therefore, on the Coq side, this should also be decided automatically, implemneted as a lemma 
%     \[
%         \texttt{EQ\_E : forall (A\ B: Term), Optional(A = B)}
%     \]
%     \item \textbf{The computational function to output the equivalence lemma for all rules.} For example, for the (R-ADDS0) rule, there should be such a lemma:
%     \[
%         \texttt{E\_R\_ADDS0 : forall (A\ B: Term), Optional(A = B)}
%     \]
%     which outputs the proof for the equivalence proof when $A$ matches the left side of (R-ADDS0) and is rewriten into B.
% \end{itemize}

% With these Coq implementations, the C++ can output the script for equivalence proof in the following format:

% \begin{align*}
%     & \texttt{rewrite (EQ\_E S1 T1).}\\
%     & \texttt{rewrite (EQ\_RULE1 A1 B1).}\\
%     & \texttt{rewrite (EQ\_E S2 T2).}\\
%     & \texttt{rewrite (EQ\_RULE2 A2 B2).}\\
%     & \texttt{rewrite (EQ\_E S3 T3).}\\
%     & \texttt{rewrite (EQ\_RULE3 A3 B3).}\\
%     & \cdots
% \end{align*}

% where at each step, the C++ solver rewrites the subterm \texttt{Ai} using rule \texttt{RULEi}. The Coq prover will synchronize with C++ solver, because both systems replace all occurances of $Ai$ into $Bi$.

% \yx{Actually the \texttt{B} inputs can be inferred and is not necessary.}

% \yx{Question: How to decide the subterms \texttt{Si} and \texttt{Ti}.}

% \yx{Question: The syntax for terms that Coq accepts.}

% \section{About Typing}
% I plan to implement a typing system in C++, replicating the one in Coq.

\bibliographystyle{plain}
\bibliography{ref}

\end{document}

### 2024-11-25 main.tex
CoC typings

@@
     \end{align*}
+    \caption{The rewriting rules for scalars.}
 \end{figure}
@@
 
-% \section{About Typing}
-% I plan to implement a typing system in C++, replicating the one in Coq.
+\clearpage
+\section{Typing System}
+The typing system modifies the one for Calculus of Construction, adding special symbols for Dirac notations.
+According to the CoC typing rules, the environment and context can be modeled by as two stacks.
+It is obvious that popping assumptions/definitions from any context or from a environment with empty context will not destroy the well-formedness.
+
+One important reason for using dependent type is that, the types of Dirac notatinos (i.e., ket, bra and operators) depends on the basis.
+
+\newcommand{\reduce}{\triangleright}
+
+\newcommand{\SORT}{\mathsf{Sort}}
+\newcommand{\WF}{\mathcal{WF}}
+\newcommand{\Type}{\mathsf{Type}}
+\newcommand{\BaseS}{\mathsf{Base}}
+
+\newcommand{\SType}{\mathcal{S}}
+\newcommand{\KType}{\mathcal{K}}
+\newcommand{\BType}{\mathcal{B}}
+\newcommand{\OType}{\mathcal{O}}
+
+\newcommand{\ZEROK}{\mathbf{0}_\mathcal{K}}
+\newcommand{\ZEROB}{\mathbf{0}_\mathcal{B}}
+\newcommand{\ZEROO}{\mathbf{0}_\mathcal{O}}
+\newcommand{\ONEO}{\mathbf{1}_\mathcal{O}}
+
+\begin{figure}[h]
+    \begin{gather*}
+        \textbf{W-Empty} \qquad
+        \frac{}{\WF([])[]}\\
+        \\
+        \textbf{W-Local-Assum} \qquad
+        \frac{E[\Gamma]\vdash T:s \qquad s \in \mathcal{S} \qquad x \notin \Gamma}{\WF(E)[\Gamma::(x:T)]}
+        \qquad \qquad
+        \textbf{W-Local-Def} \qquad
+        \frac{E[\Gamma]\vdash t : T \qquad x \notin \Gamma}{\WF(E)[\Gamma ::(x:=t:T)]} \\
+        \\
+        \textbf{W-Global-Assum} \qquad
+        \frac{E[]\vdash T:s \qquad s \in \mathcal{S} \qquad c \notin E}{\WF(E; c:T)[]}
+        \qquad \qquad
+        \textbf{W-Global-Def} \qquad
+        \frac{E[]\vdash t:T \qquad c \notin E}{\WF(E; c:=t:T)[]} \\
+        \\
+        \textbf{Ax-Type} \qquad
+        \frac{\WF(E)[\Gamma]}{E[\Gamma]\vdash \Type(i):\Type(i+1)} \\
+        \\
+        \textbf{Var} \qquad
+        \frac{\WF(E)[\Gamma] \qquad (x:T) \in \Gamma \text{ or } (x := t : T) \in \Gamma \text{ for some $t$}}{E[\Gamma] \vdash x : T} \\
+        \\
+        \textbf{Const} \qquad
+        \frac{\WF(E)[\Gamma] \qquad (c:T) \in E \text{ or } (c := t : T) \in E \text{ for some $t$}}{E[\Gamma] \vdash c : T} \\
+        \\
+        \textbf{Prod-Type} \qquad
+        \frac{E[\Gamma] \vdash T:\Type(i)\qquad E[\Gamma::(x:T)]\vdash U:\Type(i)}{E[\Gamma] \vdash \forall x: T, U : \Type(i)} \\
+        \\
+        \textbf{Lam} \qquad
+        \frac{E[\Gamma] \vdash \forall x:T, U : s \qquad E[\Gamma :: (x:T)]\vdash t : U}{E[\Gamma] \vdash \lambda x:T . t : \forall x : T, U} \\
+        \\
+        \textbf{App} \qquad
+        \frac{E[\Gamma] \vdash t:\forall x:U, T \qquad E[\Gamma] \vdash u:U}{E[\Gamma] \vdash (t\ u):T\{x/u\}} \\
+        \\
+        \textbf{Subtyping} \qquad
+        \frac{E[\Gamma] \vdash U : s \qquad E[\Gamma] \vdash t : T \qquad E[\Gamma] \vdash T \leq_{\beta\delta\eta} U}{E[\Gamma]\vdash t:U}
+    \end{gather*}
+    \caption{
+    The typing rules for CoC. 
+    We only consider computational types, removing the sorts of $\mathsf{Set}$ and $\mathsf{Prop}$.}
+\end{figure}
+
+\begin{figure}[h]
+    \begin{gather*}
+        \frac{E[\Gamma]\vdash t =_{\beta\delta\eta} u}{E[\Gamma]\vdash t \leq_{\beta\delta\eta} u} 
+        \qquad
+        \frac{i \leq j}{E[\Gamma] \vdash \Type(i) \leq_{\beta\delta\eta} \Type(j)}
+        \qquad
+        \frac{E[\Gamma] \vdash T =_{\beta\delta\eta} U \qquad E[\Gamma :: (x:T)]\vdash T' \leq_{\beta\delta\eta} U'}{E[\Gamma] \vdash \forall x : T, T' \leq_{\beta\delta\eta} \forall x : U, U'}
+    \end{gather*}
+    \caption{The rules to decide subtyping.}
+\end{figure}
+
+\begin{figure}[h]
+    \begin{gather*}
+        \textbf{Type-Base} \qquad
+        \frac{\WF(E)[\Gamma]}{E[\Gamma]\vdash \BaseS : \Type(1)} 
+        \qquad \qquad
+        \textbf{Type-Scalar} \qquad
+        \frac{\WF(E)[\Gamma]}{E[\Gamma]\vdash \SType : \Type(1)} \\
+        \\
+        \textbf{Type-Ket} \qquad
+        \frac{E[\Gamma]\vdash \sigma : \BaseS}{E[\Gamma]\vdash \KType(\sigma) : \Type(1)}
+        \qquad \qquad
+        \textbf{Type-Bra} \qquad
+        \frac{E[\Gamma]\vdash \sigma : \BaseS}{E[\Gamma]\vdash \BType(\sigma) : \Type(1)} \\
+        \\
+        \textbf{Type-Opt} \qquad
+        \frac{E[\Gamma]\vdash \sigma : \BaseS \qquad E[\Gamma]\vdash \tau : \BaseS}{E[\Gamma]\vdash \OType(\sigma, \tau) : \Type(1)}
+    \end{gather*}
+    \caption{Typing rules for Dirac notation types.}
+\end{figure}
+
+\begin{figure}[h]
+    \begin{gather*}
+        \textbf{Sca-0} \qquad
+        \frac{\WF(E)[\Gamma]}{E[\Gamma] \vdash 0 : \SType}
+        \qquad \qquad
+        \textbf{Sca-1} \qquad
+        \frac{\WF(E)[\Gamma]}{E[\Gamma] \vdash 1 : \SType}
+        \qquad \qquad
+        \textbf{Sca-Delta} \qquad
+        \frac{            
+            \begin{aligned}
+                & E[\Gamma]\vdash \sigma : \BaseS \\
+                & E[\Gamma]\vdash s : \sigma \qquad E[\Gamma]\vdash t : \sigma
+            \end{aligned}} {E[\Gamma] \vdash \delta_{s, t} : \SType} \\
+        \\
+        \textbf{Sca-Add} \qquad
+        \frac{E[\Gamma]\vdash a : \SType \qquad E[\Gamma]\vdash b : \SType}{E[\Gamma]\vdash a + b : \SType}
+        \qquad \qquad
+        \textbf{Sca-Mul} \qquad
+        \frac{E[\Gamma]\vdash a : \SType \qquad E[\Gamma]\vdash b : \SType}{E[\Gamma]\vdash a \times b : \SType} \\
+        \\
+        \textbf{Sca-Conj} \qquad
+        \frac{E[\Gamma] \vdash a : \SType}{E[\Gamma] \vdash a^*:\SType}
+        \qquad \qquad
+        \textbf{Sca-Dot} \qquad
+        \frac{E[\Gamma]\vdash B : \BType(\sigma) \qquad E[\Gamma]\vdash K : \KType(\sigma)}{E[\Gamma] \vdash B \cdot K : \SType}
+    \end{gather*}
+    \caption{Scalar typing rules.}
+\end{figure}
+
+\begin{figure}[h]
+    \begin{gather*}
+        \textbf{Base-Def} \qquad
+        \frac{\WF(E)[\Gamma] \qquad \text{for all $i$, } a_i \notin \Gamma \text{ and } a_i \notin E}
+        {E[\Gamma]\vdash \{a_1, a_2, \cdots, a_n\} : \BaseS} \\
+        \\
+        \textbf{Base-Element} \qquad
+        \frac{\WF(E)[\Gamma] \qquad c \in \{a_1, a_2, \cdots a_n\}}{E[\Gamma]\vdash c : \{a_1, a_2, \cdots, a_n\}} \\
+        \\
+        \textbf{Base-Prod} \qquad
+        \frac{E[\Gamma]\vdash \sigma : \BaseS \qquad E[\Gamma]\vdash \tau : \BaseS}{E[\Gamma]\vdash \sigma \times \tau : \BaseS} \\
+        \\
+        \textbf{Pair-Base} \qquad
+        \frac{
+            \begin{aligned}
+                & E[\Gamma]\vdash \sigma : \BaseS \qquad E[\Gamma]\vdash s : \sigma \\
+                & E[\Gamma]\vdash \tau : \BaseS \qquad E[\Gamma]\vdash t : \tau
+            \end{aligned}} {E[\Gamma]\vdash (s, t) : \sigma \times \tau} 
+        \textbf{}
+    \end{gather*}
+    \caption{
+    Typing rules for bases. 
+    Here the definitions $\{a_1, a_2, \cdots a_n\}$ of different order elements are equivalent.}
+\end{figure}
+
+\begin{figure}[h]
+    \begin{gather*}
+        \textbf{Ket-0} \qquad
+        \frac{E[\Gamma] \vdash \sigma : \BaseS}{E[\Gamma] \vdash \ZEROK(\sigma) : \KType(\sigma)} 
+        \qquad \qquad
+        \textbf{Ket-Base} \qquad
+        \frac{E[\Gamma] \vdash \sigma : \BaseS \qquad E[\Gamma]\vdash t : \sigma}{E[\Gamma] \vdash \ket{t} : \KType(\sigma)} \\
+        \\
+        \textbf{Ket-Adj} \qquad
+        \frac{E[\Gamma] \vdash B : \BType(\sigma)}{E[\Gamma] \vdash B^\dagger : \KType(\sigma)} 
+        \qquad \qquad
+        \textbf{Ket-Sca} \qquad
+        \frac{E[\Gamma] \vdash a : \SType \qquad E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash a.K : \KType(\sigma)} \\
+        \\
+        \textbf{Ket-Add} \qquad
+        \frac{E[\Gamma] \vdash K_1 : \KType(\sigma) \qquad E[\Gamma] \vdash K_2 : \KType(\sigma)}{E[\Gamma] \vdash K_1 + K_2 : \KType(\sigma)} \\
+        \\
+        \textbf{Ket-Dot} \qquad
+        \frac{E[\Gamma] \vdash O : \OType(\sigma, \tau) \qquad E[\Gamma] \vdash K : \KType(\tau)}{E[\Gamma] \vdash O \cdot K : \KType(\sigma)} \\
+        \\
+        \textbf{Ket-Tsr} \qquad
+        \frac{E[\Gamma] \vdash K_1 : \KType(\sigma) \qquad E[\Gamma] \vdash K_2 : \KType(\tau)} {E[\Gamma] \vdash K_1 \otimes K_2 : \KType(\sigma \times \tau)}
+    \end{gather*}
+    \caption{Ket typing rules.}
+\end{figure}
+
+\begin{figure}[h]
+    \begin{gather*}
+        \textbf{Bra-0} \qquad
+        \frac{E[\Gamma] \vdash \sigma : \BaseS}{E[\Gamma] \vdash \ZEROB(\sigma) : \BType(\sigma)} 
+        \qquad \qquad
+        \textbf{Bra-Base} \qquad
+        \frac{E[\Gamma] \vdash \sigma : \BaseS \qquad E[\Gamma]\vdash t : \sigma}{E[\Gamma] \vdash \bra{t} : \BType(\sigma)} \\
+        \\
+        \textbf{Bra-Adj} \qquad
+        \frac{E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash K^\dagger : \BType(\sigma)} 
+        \qquad \qquad
+        \textbf{Bra-Sca} \qquad
+        \frac{E[\Gamma] \vdash a : \SType \qquad E[\Gamma] \vdash B : \BType(\sigma)}{E[\Gamma] \vdash a.B : \BType(\sigma)} \\
+        \\
+        \textbf{Bra-Add} \qquad
+        \frac{E[\Gamma] \vdash B_1 : \BType(\sigma) \qquad E[\Gamma] \vdash B_2 : \BType(\sigma)}{E[\Gamma] \vdash B_1 + B_2 : \BType(\sigma)} \\
+        \\
+        \textbf{Bra-Dot} \qquad
+        \frac{E[\Gamma] \vdash B : \KType(\sigma) \qquad E[\Gamma] \vdash O : \OType(\sigma, \tau)}{E[\Gamma] \vdash B \cdot O : \BType(\tau)} \\
+        \\
+        \textbf{Bra-Tsr} \qquad
+        \frac{E[\Gamma] \vdash B_1 : \BType(\sigma) \qquad E[\Gamma] \vdash B_2 : \BType(\tau)} {E[\Gamma] \vdash B_1 \otimes B_2 : \BType(\sigma \times \tau)}
+    \end{gather*}
+    \caption{Bra typing rules.}
+\end{figure}
+
+\begin{figure}[h]
+    \begin{gather*}
+        \textbf{Opt-0} \qquad
+        \frac{E[\Gamma] \vdash \sigma : \BaseS \qquad E[\Gamma] \vdash \tau : \BaseS}{E[\Gamma] \vdash \ZEROO(\sigma, \tau) : \OType(\sigma, \tau)} 
+        \qquad \qquad
+        \textbf{Opt-1} \qquad
+        \frac{E[\Gamma] \vdash \sigma : \BaseS}{E[\Gamma] \vdash \ONEO(\sigma) : \OType(\sigma, \sigma)} \\
+        \\
+        \textbf{Opt-Adj} \qquad
+        \frac{E[\Gamma] \vdash O : \OType(\sigma, \tau)}{E[\Gamma] \vdash O^\dagger : \OType(\tau, \sigma)} 
+        \qquad \qquad
+        \textbf{Opt-Sca} \qquad
+        \frac{E[\Gamma] \vdash a : \SType \qquad E[\Gamma] \vdash O : \OType(\sigma, \tau)}{E[\Gamma] \vdash a.O : \OType(\sigma, \tau)} \\
+        \\
+        \textbf{Opt-Add} \qquad
+        \frac{E[\Gamma] \vdash O_1 : \OType(\sigma, \tau) \qquad E[\Gamma] \vdash O_2 : \OType(\sigma, \tau)}{E[\Gamma] \vdash O_1 + O_2 : \OType(\sigma, \tau)} \\
+        \\
+        \textbf{Opt-Outer} \qquad
+        \frac{E[\Gamma]\vdash K : \KType(\sigma) \qquad E[\Gamma]\vdash B : \BType(\tau)}{E[\Gamma]\vdash K \cdot B : \OType(\sigma, \tau)} \\
+        \\
+        \textbf{Opt-Dot} \qquad
+        \frac{E[\Gamma] \vdash O_1 : \OType(\sigma, \tau) \qquad E[\Gamma] \vdash O_2 : \OType(\tau, \rho)}{E[\Gamma] \vdash O_1 \cdot O_2 : \OType(\sigma, \rho)} \\
+        \\
+        \textbf{Opt-Tsr} \qquad
+        \frac{E[\Gamma] \vdash O_1 : \OType(\sigma_1, \tau_1) \qquad E[\Gamma] \vdash O_2 : \OType(\sigma_2, \tau_2)} {E[\Gamma] \vdash O_1 \otimes O_2 : \OType(\sigma_1 \times \sigma_2, \tau_1 \times \tau_2)}
+    \end{gather*}
+    \caption{Operator typing rules.}
+\end{figure}
+
+\clearpage
+\section{Conversions and Reductions}
+\begin{figure}[h]
+    \begin{gather*}
+        \textbf{Alpha} \qquad \text{(bound variable renaming)}
+        \qquad \qquad
+        \textbf{Beta} \qquad
+        \frac{}{E[\Gamma] \vdash ((\lambda x:T.t)\ u) \reduce_\beta t\{x/u\}} \\
+        \\
+        \textbf{Delta-Local} \qquad
+        \frac{\WF(E)[\Gamma] \qquad (x:=t:T) \in \Gamma}{E[\Gamma] \vdash x \reduce_\Delta t} 
+        \qquad
+        \textbf{Delta-Global} \qquad
+        \frac{\WF(E)[\Gamma] \qquad (c:=t:T) \in E}{E[\Gamma] \vdash c \reduce_\delta t} \\
+        \\
+        \textbf{Eta} \qquad
+        \frac{E[\Gamma]\vdash t : \forall x : T , U}{E[\Gamma] \vdash t =_\eta \lambda x : T . (t\ x)}
+    \end{gather*}
+    \caption{Conversions and reductions in CoC.}
+\end{figure}
+
+\begin{figure}[h]
+    \begin{gather*}
+        \cdots
+    \end{gather*}
+    \caption{Dirac notation reduction rules}
+\end{figure}
+
+About big operators: maybe we should use the "sum operators"?
+
+\clearpage
 

### 2024-11-25 main.tex
change to typed lambda calculus

--- 2024-11-20 main.tex
@@
 
-% \section{About Typing}
-% I plan to implement a typing system in C++, replicating the one in Coq.
+\clearpage
+\section{Typing System}
+The typing system modifies the one for Calculus of Construction, adding special symbols for Dirac notations.
+According to the CoC typing rules, the environment and context can be modeled by as two stacks.
+It is obvious that popping assumptions/definitions from any context or from a environment with empty context will not destroy the well-formedness.
+
+One important reason for using dependent type is that, the types of Dirac notatinos (i.e., ket, bra and operators) depends on the basis.
+
+\newcommand{\reduce}{\triangleright}
+
+\newcommand{\Sort}{\mathsf{Sort}}
+\newcommand{\WF}{\mathcal{WF}}
+\newcommand{\Type}{\mathsf{Type}}
+\newcommand{\BaseS}{\mathsf{Base}}
+
+\newcommand{\SType}{\mathcal{S}}
+\newcommand{\KType}{\mathcal{K}}
+\newcommand{\BType}{\mathcal{B}}
+\newcommand{\OType}{\mathcal{O}}
+
+\newcommand{\ZEROK}{\mathbf{0}_\mathcal{K}}
+\newcommand{\ZEROB}{\mathbf{0}_\mathcal{B}}
+\newcommand{\ZEROO}{\mathbf{0}_\mathcal{O}}
+\newcommand{\ONEO}{\mathbf{1}_\mathcal{O}}
+
+\begin{figure}[h]
+    \begin{gather*}
+        \textbf{W-Empty} \qquad
+        \frac{}{\WF([])}\\
+        \\
+        \textbf{W-Assum} \qquad
+        \frac{E\vdash T:\Type \qquad c \notin E}{\WF(E; c:T)}
+        \qquad \qquad
+        \textbf{W-Def} \qquad
+        \frac{E\vdash t:T \qquad c \notin E}{\WF(E; c:=t:T)} \\
+        \\
+        \textbf{Const} \qquad
+        \frac{\WF(E) \qquad (c:T) \in E \text{ or } (c := t : T) \in E \text{ for some $t$}}{E \vdash c : T} \\
+        \\
+        \textbf{Arrow} \qquad
+        \frac{E \vdash T:\Type \qquad E \vdash U:\Type}{E \vdash T \to U : \Type} \\
+        \\
+        \textbf{Lam} \qquad
+        \frac{E;x:T \vdash t : U}{E \vdash (\lambda x:T . t) : T \to U}
+        \qquad \qquad
+        \textbf{App} \qquad
+        \frac{E \vdash t:U \to T \qquad E \vdash u:U}{E \vdash (t\ u):T}
+    \end{gather*}
+    \caption{The typing rules for the typed lambda calculus with environment.}
+\end{figure}
+
+\begin{figure}[h]
+    \begin{gather*}
+        \textbf{Type-Base} \qquad
+        \frac{\WF(E)}{E \vdash \BaseS : \Type} 
+        \qquad \qquad
+        \textbf{Base-Subtype} \qquad
+        \frac{E \vdash T : \BaseS}{E \vdash T : \Type} \\
+        \\
+        \textbf{Type-Ket} \qquad
+        \frac{E \vdash \sigma : \BaseS}{E \vdash \KType(\sigma) : \Type}
+        \qquad \qquad
+        \textbf{Type-Bra} \qquad
+        \frac{E \vdash \sigma : \BaseS}{E \vdash \BType(\sigma) : \Type} \\
+        \\
+        \textbf{Type-Opt} \qquad
+        \frac{E \vdash \sigma : \BaseS \qquad E \vdash \tau : \BaseS}{E \vdash \OType(\sigma, \tau) : \Type}
+        \qquad \qquad
+        \textbf{Type-Scalar} \qquad
+        \frac{\WF(E)}{E \vdash \SType : \Type}
+    \end{gather*}
+    \caption{Typing rules for Dirac notation types.}
+\end{figure}
+
+\begin{figure}[h]
+    \begin{gather*}
+        \textbf{Sca-0} \qquad
+        \frac{\WF(E)}{E \vdash 0 : \SType}
+        \qquad \qquad
+        \textbf{Sca-1} \qquad
+        \frac{\WF(E)}{E \vdash 1 : \SType}
+        \qquad \qquad
+        \textbf{Sca-Delta} \qquad
+        \frac{            
+            \begin{aligned}
+                & E\vdash \sigma : \BaseS \\
+                & E\vdash s : \sigma \qquad E\vdash t : \sigma
+            \end{aligned}} {E \vdash \delta_{s, t} : \SType} \\
+        \\
+        \textbf{Sca-Add} \qquad
+        \frac{E\vdash a : \SType \qquad E\vdash b : \SType}{E\vdash a + b : \SType}
+        \qquad \qquad
+        \textbf{Sca-Mul} \qquad
+        \frac{E\vdash a : \SType \qquad E\vdash b : \SType}{E\vdash a \times b : \SType} \\
+        \\
+        \textbf{Sca-Conj} \qquad
+        \frac{E \vdash a : \SType}{E \vdash a^*:\SType}
+        \qquad \qquad
+        \textbf{Sca-Dot} \qquad
+        \frac{E\vdash B : \BType(\sigma) \qquad E\vdash K : \KType(\sigma)}{E \vdash B \cdot K : \SType}
+    \end{gather*}
+    \caption{Scalar typing rules.}
+\end{figure}
+
+\begin{figure}[h]
+    \begin{gather*}
+        \textbf{Base-Def} \qquad
+        \frac{\WF(E) \qquad a_i \notin E \text{ for all $i$}}
+        {E\vdash \{a_1, a_2, \cdots, a_n\} : \BaseS}
+        \qquad \qquad
+        \textbf{Base-Element} \qquad
+        \frac{\WF(E) \qquad c \in \{a_1, a_2, \cdots a_n\}}{E\vdash c : \{a_1, a_2, \cdots, a_n\}} \\
+        \\
+        \textbf{Base-Prod} \qquad
+        \frac{E\vdash \sigma : \BaseS \qquad E\vdash \tau : \BaseS}{E\vdash \sigma \times \tau : \BaseS}
+        \qquad \qquad
+        \textbf{Pair-Base} \qquad
+        \frac{
+            \begin{aligned}
+                & E\vdash \sigma : \BaseS \qquad E\vdash s : \sigma \\
+                & E\vdash \tau : \BaseS \qquad E\vdash t : \tau
+            \end{aligned}} {E\vdash (s, t) : \sigma \times \tau} 
+        \textbf{}
+    \end{gather*}
+    \caption{Typing rules for bases. }
+\end{figure}
+
+\begin{figure}[h]
+    \begin{gather*}
+        \textbf{Ket-0} \qquad
+        \frac{E \vdash \sigma : \BaseS}{E \vdash \ZEROK(\sigma) : \KType(\sigma)} 
+        \qquad \qquad
+        \textbf{Ket-Base} \qquad
+        \frac{E \vdash \sigma : \BaseS \qquad E\vdash t : \sigma}{E \vdash \ket{t} : \KType(\sigma)} \\
+        \\
+        \textbf{Ket-Adj} \qquad
+        \frac{E \vdash B : \BType(\sigma)}{E \vdash B^\dagger : \KType(\sigma)} 
+        \qquad \qquad
+        \textbf{Ket-Sca} \qquad
+        \frac{E \vdash a : \SType \qquad E \vdash K : \KType(\sigma)}{E \vdash a.K : \KType(\sigma)} \\
+        \\
+        \textbf{Ket-Add} \qquad
+        \frac{E \vdash K_1 : \KType(\sigma) \qquad E \vdash K_2 : \KType(\sigma)}{E \vdash K_1 + K_2 : \KType(\sigma)}
+        \qquad \qquad
+        \textbf{Ket-Dot} \qquad
+        \frac{E \vdash O : \OType(\sigma, \tau) \qquad E \vdash K : \KType(\tau)}{E \vdash O \cdot K : \KType(\sigma)} \\
+        \\
+        \textbf{Ket-Tsr} \qquad
+        \frac{E \vdash K_1 : \KType(\sigma) \qquad E \vdash K_2 : \KType(\tau)} {E \vdash K_1 \otimes K_2 : \KType(\sigma \times \tau)}
+    \end{gather*}
+    \caption{Ket typing rules.}
+\end{figure}
+
+\begin{figure}[h]
+    \begin{gather*}
+        \textbf{Bra-0} \qquad
+        \frac{E \vdash \sigma : \BaseS}{E \vdash \ZEROB(\sigma) : \BType(\sigma)} 
+        \qquad \qquad
+        \textbf{Bra-Base} \qquad
+        \frac{E \vdash \sigma : \BaseS \qquad E\vdash t : \sigma}{E \vdash \bra{t} : \BType(\sigma)} \\
+        \\
+        \textbf{Bra-Adj} \qquad
+        \frac{E \vdash K : \KType(\sigma)}{E \vdash K^\dagger : \BType(\sigma)} 
+        \qquad \qquad
+        \textbf{Bra-Sca} \qquad
+        \frac{E \vdash a : \SType \qquad E \vdash B : \BType(\sigma)}{E \vdash a.B : \BType(\sigma)} \\
+        \\
+        \textbf{Bra-Add} \qquad
+        \frac{E \vdash B_1 : \BType(\sigma) \qquad E \vdash B_2 : \BType(\sigma)}{E \vdash B_1 + B_2 : \BType(\sigma)}
+        \qquad \qquad
+        \textbf{Bra-Dot} \qquad
+        \frac{E \vdash B : \KType(\sigma) \qquad E \vdash O : \OType(\sigma, \tau)}{E \vdash B \cdot O : \BType(\tau)} \\
+        \\
+        \textbf{Bra-Tsr} \qquad
+        \frac{E \vdash B_1 : \BType(\sigma) \qquad E \vdash B_2 : \BType(\tau)} {E \vdash B_1 \otimes B_2 : \BType(\sigma \times \tau)}
+    \end{gather*}
+    \caption{Bra typing rules.}
+\end{figure}
+
+\begin{figure}[h]
+    \begin{gather*}
+        \textbf{Opt-0} \qquad
+        \frac{E \vdash \sigma : \BaseS \qquad E \vdash \tau : \BaseS}{E \vdash \ZEROO(\sigma, \tau) : \OType(\sigma, \tau)} 
+        \qquad \qquad
+        \textbf{Opt-1} \qquad
+        \frac{E \vdash \sigma : \BaseS}{E \vdash \ONEO(\sigma) : \OType(\sigma, \sigma)} \\
+        \\
+        \textbf{Opt-Adj} \qquad
+        \frac{E \vdash O : \OType(\sigma, \tau)}{E \vdash O^\dagger : \OType(\tau, \sigma)} 
+        \qquad \qquad
+        \textbf{Opt-Sca} \qquad
+        \frac{E \vdash a : \SType \qquad E \vdash O : \OType(\sigma, \tau)}{E \vdash a.O : \OType(\sigma, \tau)} \\
+        \\
+        \textbf{Opt-Add} \qquad
+        \frac{E \vdash O_1 : \OType(\sigma, \tau) \qquad E \vdash O_2 : \OType(\sigma, \tau)}{E \vdash O_1 + O_2 : \OType(\sigma, \tau)}
+        \qquad \qquad
+        \textbf{Opt-Outer} \qquad
+        \frac{E\vdash K : \KType(\sigma) \qquad E\vdash B : \BType(\tau)}{E\vdash K \cdot B : \OType(\sigma, \tau)} \\
+        \\
+        \textbf{Opt-Dot} \qquad
+        \frac{E \vdash O_1 : \OType(\sigma, \tau) \qquad E \vdash O_2 : \OType(\tau, \rho)}{E \vdash O_1 \cdot O_2 : \OType(\sigma, \rho)} \\
+        \\
+        \textbf{Opt-Tsr} \qquad
+        \frac{E \vdash O_1 : \OType(\sigma_1, \tau_1) \qquad E \vdash O_2 : \OType(\sigma_2, \tau_2)} {E \vdash O_1 \otimes O_2 : \OType(\sigma_1 \times \sigma_2, \tau_1 \times \tau_2)}
+    \end{gather*}
+    \caption{Operator typing rules.}
+\end{figure}
+
+\clearpage
+\section{Conversions and Reductions}
+\begin{figure}[h]
+    \begin{gather*}
+        \textbf{Alpha} \qquad \text{(bound variable renaming)}
+        \qquad \qquad
+        \textbf{Beta} \qquad
+        \frac{}{E \vdash ((\lambda x:T.t)\ u) \reduce_\beta t\{x/u\}} \\
+        \\
+        \textbf{Delta} \qquad
+        \frac{\WF(E) \qquad (c:=t:T) \in E}{E \vdash c \reduce_\delta t}
+        \qquad \qquad
+        \textbf{Eta} \qquad
+        \frac{E\vdash t : T \to U}{E \vdash \lambda x : T . (t\ x) \reduce_\eta t}
+    \end{gather*}
+    \caption{Conversions and reductions for the typed lambda calculus.}
+\end{figure}
+
+\begin{figure}[h]
+    \begin{gather*}
+        \cdots
+    \end{gather*}
+    \caption{Dirac notation reduction rules}
+\end{figure}
+
+About big operators: maybe we should use the "sum operators"?
+
+\clearpage
 

### 2024-11-29 main.tex
adding rules

--- 2024-11-20 main.tex
@@
 \usepackage{bm}
+\usepackage{longtable}
+\usepackage{makecell}
 
@@
 
+\newcommand{\reduce}{\triangleright}
+
+\newcommand{\Sort}{\mathsf{Sort}}
+\newcommand{\WF}{\mathcal{WF}}
+
+\newcommand{\Type}{\mathsf{Type}}
+\newcommand{\BaseS}{\mathsf{Base}}
+
+\newcommand{\SType}{\mathcal{S}}
+\newcommand{\KType}{\mathcal{K}}
+\newcommand{\BType}{\mathcal{B}}
+\newcommand{\OType}{\mathcal{O}}
+
+\newcommand{\ZEROK}{\mathbf{0}_\mathcal{K}}
+\newcommand{\ZEROB}{\mathbf{0}_\mathcal{B}}
+\newcommand{\ZEROO}{\mathbf{0}_\mathcal{O}}
+
+\newcommand{\PAIR}{\mathsf{PAIR}}
+
 \newcommand{\ZERO}{\mathsf{0}}
@@
 \newcommand{\ADDS}{\mathsf{ADDS}}
-\newcommand{\MLTS}{\mathsf{MLTS}}
+\newcommand{\ADD}{\mathsf{ADD}}
+\newcommand{\MULS}{\mathsf{MULS}}
+\newcommand{\MUL}{\mathsf{MUL}}
 \newcommand{\CONJ}{\mathsf{CONJ}}
+\newcommand{\CJG}{\mathsf{CJG}}
+\newcommand{\ADJ}{\mathsf{ADJ}}
+\newcommand{\DELTA}{\mathsf{DELTA}}
+\newcommand{\DOT}{\mathsf{DOT}}
+\newcommand{\SCR}{\mathsf{SCR}}
+\newcommand{\TSR}{\mathsf{TSR}}
+\newcommand{\KET}{\mathsf{KET}}
+\newcommand{\BRA}{\mathsf{BRA}}
+\newcommand{\ONEO}{\mathbf{1}_\mathcal{O}}
 
@@
     \[ 
-    S ::= \ZERO\ |\ \ONE\ |\ \ADDS ( S^+ )\ |\ \MLTS ( S^+ )\ |\ \CONJ(S).
+    S ::= \ZERO\ |\ \ONE\ |\ \ADDS ( S^+ )\ |\ \MULS ( S^+ )\ |\ \CONJ(S).
     \]
@@
         & && \textcolor{gray}{\text{This rule removes all $\ZERO$ occurances and keeps the order of remaining subterms.}} \\
-        & \text{(R-MLTSID)} &&
-        \MLTS(a) \to a \\
-        & \text{(R-MLTS0)} &&
-        \MLTS((seq: \_ \_)\ \ZERO) \to \ZERO \\
-        & \text{(R-MLTS1)} &&
-        \MLTS((seq: \_ \_)\ \ONE) \to \MLTS(seq) \\
+        & \text{(R-MULSID)} &&
+        \MULS(a) \to a \\
+        & \text{(R-MULS0)} &&
+        \MULS((seq: \_ \_)\ \ZERO) \to \ZERO \\
+        & \text{(R-MULS1)} &&
+        \MULS((seq: \_ \_)\ \ONE) \to \MULS(seq) \\
         & && \textcolor{gray}{\text{Similar to (R-ADDS0).}} \\
-        & \text{(R-MLTS2)} &&
-        \MLTS((seq1: \_ \_ \_)\ \ADDS(a_1\ a_2 \cdots a_n)\ (seq2: \_ \_ \_)) \\
-        & && \to \ADDS(\MLTS(seq1\ a_1\ seq2)\ \MLTS(seq1\ a_2\ seq2) \cdots \MLTS(seq1\ a_n\ seq2)) \\
+        & \text{(R-MULS2)} &&
+        \MULS((seq1: \_ \_ \_)\ \ADDS(a_1\ a_2 \cdots a_n)\ (seq2: \_ \_ \_)) \\
+        & && \to \ADDS(\MULS(seq1\ a_1\ seq2)\ \MULS(seq1\ a_2\ seq2) \cdots \MULS(seq1\ a_n\ seq2)) \\
         & && \textcolor{gray}{\text{This rule matches the first $\ADDS$ subterm in the list. $seq1$ or $seq2$ can be empty, but not both.}} \\
@@
         & \text{(R-CONJ3)} &&
-        \CONJ(\MLTS(a_1\ a_2\ \cdots a_n)) \to \MLTS(\CONJ(a_1)\ \CONJ(a_2) \cdots \CONJ(a_n)) \\
+        \CONJ(\MULS(a_1\ a_2\ \cdots a_n)) \to \MULS(\CONJ(a_1)\ \CONJ(a_2) \cdots \CONJ(a_n)) \\
         & \text{(R-CONJ4)} &&
@@
     & \quad |\ \textsf{E} \Rightarrow A \\
-    & \quad |\ [i_1:P_1\ \cdots\ i_n:P_n] \Rightarrow \texttt{A.head}(\texttt{apply(A.args[$i_1$], $P_1$)}\ \cdots\ \texttt{apply(A.args[$i_n$], $P_n$)}) \\
+    & \quad |\ [i_1:P_1\ \cdots\ i_n:P_n] \Rightarrow \texttt{A.head}(\texttt{apply(A.args[$i_1$], $P_{i_1}$)}\ \cdots\ \texttt{apply(A.args[$i_n$], $P_{i_n}$)}) \\
     & \texttt{end}
@@
 
-% \section{About Typing}
-% I plan to implement a typing system in C++, replicating the one in Coq.
+\clearpage
+
+\section{Language Syntax}
+\begin{definition}[syntax]
+    The syntax for Dirac notation types is defined as 
+    \begin{align*}
+        T ::=\ & x\ |\ \SType\ |\ \KType(\sigma)\ |\ \BType(\sigma)\ |\ \OType(\sigma_1, \sigma_2)\ |\ T_1 \to T_2, \\
+        \sigma ::=\ & x\ |\ \sigma_1 \times \sigma_2.
+    \end{align*}
+    The syntax for Dirac notation terms is defined as
+    \begin{align*}
+        e ::=\ & x\ |\ (e_1, e_2) \\
+        & |\ 0\ |\ 1\ |\ e_1 + \cdots + e_n \ |\ e_1 \times \cdots \times e_n\ |\ e^* \ |\ \delta_{e_1, e_2}\ \\
+        & |\ \ZEROK(\sigma)\ |\ \ZEROB(\sigma)\ |\ \ZEROO(\sigma_1, \sigma_2)\ |\ \ONEO(\sigma) \\
+        & |\ \ket{e}\ |\ \bra{t}\ |\ e^\dagger\ |\ e_1.e_2\ |\ e_1 \cdot e_2\ |\ e_1 \otimes e_2.
+    \end{align*}
+\end{definition}
+ % after consideration I still think keep DOT as binary is the best choice. It reduces the unstability of complicated matching in sequences.
+
+\section{Typing System}
+The typing system modifies the one for Calculus of Construction, adding special symbols for Dirac notations.
+According to the CoC typing rules, the environment and context can be modeled by as two stacks.
+It is obvious that popping assumptions/definitions from any context or from a environment with empty context will not destroy the well-formedness.
+
+One important reason for using dependent type is that, the types of Dirac notatinos (i.e., ket, bra and operators) depends on the basis.
+
+\begin{figure}[h]
+    \begin{gather*}
+        \textbf{W-Empty} \qquad
+        \frac{}{\WF([])}
+        \qquad \qquad
+        \textbf{W-Assum-Type} \qquad
+        \frac{\WF(\Gamma) \qquad T \notin \Gamma}{\WF(\Gamma; T:\Type)}\\
+        \\
+        \textbf{W-Assum-Term} \qquad
+        \frac{\Gamma\vdash T:\Type \qquad x \notin \Gamma}{\WF(\Gamma; x:T)}
+        \qquad \qquad
+        \textbf{W-Def} \qquad
+        \frac{\Gamma\vdash t:T \qquad x \notin \Gamma}{\WF(\Gamma; x:=t:T)} \\
+        \\
+        \textbf{Var} \qquad
+        \frac{\WF(\Gamma) \qquad (x:T) \in \Gamma \text{ or } (x := t : T) \in \Gamma \text{ for some $t$}}{\Gamma \vdash x : T} \\
+        \\
+        \textbf{Arrow} \qquad
+        \frac{\Gamma \vdash T:\Type \qquad \Gamma \vdash U:\Type}{\Gamma \vdash T \to U : \Type} \\
+        \\
+        \textbf{Lam} \qquad
+        \frac{\Gamma;x:T \vdash t : U}{\Gamma \vdash (\lambda x:T . t) : T \to U}
+        \qquad \qquad
+        \textbf{App} \qquad
+        \frac{\Gamma \vdash t:U \to T \qquad \Gamma \vdash u:U}{\Gamma \vdash (t\ u):T}
+    \end{gather*}
+    \caption{The typing rules for the typed lambda calculus with environment.}
+\end{figure}
+
+\begin{figure}[h]
+    \begin{gather*}
+        \textbf{Type-Base} \qquad
+        \frac{\WF(\Gamma)}{\Gamma \vdash \BaseS : \Type} 
+        \qquad \qquad
+        \textbf{Base-Subtype} \qquad
+        \frac{\Gamma \vdash T : \BaseS}{\Gamma \vdash T : \Type} \\
+        \\
+        \textbf{Type-Ket} \qquad
+        \frac{\Gamma \vdash \sigma : \BaseS}{\Gamma \vdash \KType(\sigma) : \Type}
+        \qquad \qquad
+        \textbf{Type-Bra} \qquad
+        \frac{\Gamma \vdash \sigma : \BaseS}{\Gamma \vdash \BType(\sigma) : \Type} \\
+        \\
+        \textbf{Type-Opt} \qquad
+        \frac{\Gamma \vdash \sigma : \BaseS \qquad \Gamma \vdash \tau : \BaseS}{\Gamma \vdash \OType(\sigma, \tau) : \Type}
+        \qquad \qquad
+        \textbf{Type-Scalar} \qquad
+        \frac{\WF(\Gamma)}{\Gamma \vdash \SType : \Type}
+    \end{gather*}
+    \caption{Typing rules for Dirac notation types.}
+\end{figure}
+
+\begin{figure}[h]
+    \begin{gather*}
+        \textbf{Sca-0} \qquad
+        \frac{\WF(\Gamma)}{\Gamma \vdash 0 : \SType}
+        \qquad \qquad
+        \textbf{Sca-1} \qquad
+        \frac{\WF(\Gamma)}{\Gamma \vdash 1 : \SType}
+        \qquad \qquad
+        \textbf{Sca-Delta} \qquad
+        \frac{            
+            \begin{aligned}
+                & \Gamma\vdash \sigma : \BaseS \\
+                & \Gamma\vdash s : \sigma \qquad \Gamma\vdash t : \sigma
+            \end{aligned}} {\Gamma \vdash \delta_{s, t} : \SType} \\
+        \\
+        \textbf{Sca-Add} \qquad
+        \frac{\Gamma\vdash a_i : \SType \text{ for all $i$}}{\Gamma\vdash a_1 + \cdots + a_n : \SType}
+        \qquad \qquad
+        \textbf{Sca-Mul} \qquad
+        \frac{\Gamma\vdash a_i : \SType \text{ for all $i$}}{\Gamma\vdash a_1 \times \cdots \times a_n : \SType} \\
+        \\
+        \textbf{Sca-Conj} \qquad
+        \frac{\Gamma \vdash a : \SType}{\Gamma \vdash a^*:\SType}
+        \qquad \qquad
+        \textbf{Sca-Dot} \qquad
+        \frac{\Gamma\vdash B : \BType(\sigma) \qquad \Gamma\vdash K : \KType(\sigma)}{\Gamma \vdash B \cdot K : \SType}
+    \end{gather*}
+    \caption{Scalar typing rules.}
+\end{figure}
+
+\begin{figure}[h]
+    \begin{gather*}
+        % this two rules can be considered later
+        % \textbf{Base-Def} \qquad
+        % \frac{\WF(\Gamma) \qquad a_i \notin \Gamma \text{ for all $i$}}
+        % {\Gamma\vdash \{a_1, a_2, \cdots, a_n\} : \BaseS}
+        % \qquad \qquad
+        % \textbf{Base-Element} \qquad
+        % \frac{\WF(\Gamma) \qquad c \in \{a_1, a_2, \cdots a_n\}}{\Gamma\vdash c : \{a_1, a_2, \cdots, a_n\}} \\
+        % \\
+        \textbf{Base-Prod} \qquad
+        \frac{\Gamma\vdash \sigma : \BaseS \qquad \Gamma\vdash \tau : \BaseS}{\Gamma\vdash \sigma \times \tau : \BaseS}
+        \qquad \qquad
+        \textbf{Pair-Base} \qquad
+        \frac{
+            \begin{aligned}
+                & \Gamma\vdash \sigma : \BaseS \qquad \Gamma\vdash s : \sigma \\
+                & \Gamma\vdash \tau : \BaseS \qquad \Gamma\vdash t : \tau
+            \end{aligned}} {\Gamma\vdash (s, t) : \sigma \times \tau} 
+        \textbf{}
+    \end{gather*}
+    \caption{Typing rules for bases. }
+\end{figure}
+
+\begin{figure}[h]
+    \begin{gather*}
+        \textbf{Ket-0} \qquad
+        \frac{\Gamma \vdash \sigma : \BaseS}{\Gamma \vdash \ZEROK(\sigma) : \KType(\sigma)} 
+        \qquad \qquad
+        \textbf{Ket-Base} \qquad
+        \frac{\Gamma \vdash \sigma : \BaseS \qquad \Gamma\vdash t : \sigma}{\Gamma \vdash \ket{t} : \KType(\sigma)} \\
+        \\
+        \textbf{Ket-Adj} \qquad
+        \frac{\Gamma \vdash B : \BType(\sigma)}{\Gamma \vdash B^\dagger : \KType(\sigma)} 
+        \qquad \qquad
+        \textbf{Ket-Sca} \qquad
+        \frac{\Gamma \vdash a : \SType \qquad \Gamma \vdash K : \KType(\sigma)}{\Gamma \vdash a.K : \KType(\sigma)} \\
+        \\
+        \textbf{Ket-Add} \qquad
+        \frac{\Gamma \vdash K_i : \KType(\sigma) \text{ for all $i$}}{\Gamma \vdash K_1 + \cdots + K_n : \KType(\sigma)}
+        \qquad \qquad
+        \textbf{Ket-Dot} \qquad
+        \frac{\Gamma \vdash O : \OType(\sigma, \tau) \qquad \Gamma \vdash K : \KType(\tau)}{\Gamma \vdash O \cdot K : \KType(\sigma)} \\
+        \\
+        \textbf{Ket-Tsr} \qquad
+        \frac{\Gamma \vdash K_1 : \KType(\sigma) \qquad \Gamma \vdash K_2 : \KType(\tau)} {\Gamma \vdash K_1 \otimes K_2 : \KType(\sigma \times \tau)}
+    \end{gather*}
+    \caption{Ket typing rules.}
+\end{figure}
+
+\begin{figure}[h]
+    \begin{gather*}
+        \textbf{Bra-0} \qquad
+        \frac{\Gamma \vdash \sigma : \BaseS}{\Gamma \vdash \ZEROB(\sigma) : \BType(\sigma)} 
+        \qquad \qquad
+        \textbf{Bra-Base} \qquad
+        \frac{\Gamma \vdash \sigma : \BaseS \qquad \Gamma\vdash t : \sigma}{\Gamma \vdash \bra{t} : \BType(\sigma)} \\
+        \\
+        \textbf{Bra-Adj} \qquad
+        \frac{\Gamma \vdash K : \KType(\sigma)}{\Gamma \vdash K^\dagger : \BType(\sigma)} 
+        \qquad \qquad
+        \textbf{Bra-Sca} \qquad
+        \frac{\Gamma \vdash a : \SType \qquad \Gamma \vdash B : \BType(\sigma)}{\Gamma \vdash a.B : \BType(\sigma)} \\
+        \\
+        \textbf{Bra-Add} \qquad
+        \frac{\Gamma \vdash B_i : \BType(\sigma) \text{ for all $i$}}{\Gamma \vdash B_1 + \cdots + B_n : \BType(\sigma)}
+        \qquad \qquad
+        \textbf{Bra-Dot} \qquad
+        \frac{\Gamma \vdash B : \KType(\sigma) \qquad \Gamma \vdash O : \OType(\sigma, \tau)}{\Gamma \vdash B \cdot O : \BType(\tau)} \\
+        \\
+        \textbf{Bra-Tsr} \qquad
+        \frac{\Gamma \vdash B_1 : \BType(\sigma) \qquad \Gamma \vdash B_2 : \BType(\tau)} {\Gamma \vdash B_1 \otimes B_2 : \BType(\sigma \times \tau)}
+    \end{gather*}
+    \caption{Bra typing rules.}
+\end{figure}
+
+\begin{figure}[h]
+    \begin{gather*}
+        \textbf{Opt-0} \qquad
+        \frac{\Gamma \vdash \sigma : \BaseS \qquad \Gamma \vdash \tau : \BaseS}{\Gamma \vdash \ZEROO(\sigma, \tau) : \OType(\sigma, \tau)} 
+        \qquad \qquad
+        \textbf{Opt-1} \qquad
+        \frac{\Gamma \vdash \sigma : \BaseS}{\Gamma \vdash \ONEO(\sigma) : \OType(\sigma, \sigma)} \\
+        \\
+        \textbf{Opt-Adj} \qquad
+        \frac{\Gamma \vdash O : \OType(\sigma, \tau)}{\Gamma \vdash O^\dagger : \OType(\tau, \sigma)} 
+        \qquad \qquad
+        \textbf{Opt-Sca} \qquad
+        \frac{\Gamma \vdash a : \SType \qquad \Gamma \vdash O : \OType(\sigma, \tau)}{\Gamma \vdash a.O : \OType(\sigma, \tau)} \\
+        \\
+        \textbf{Opt-Add} \qquad
+        \frac{\Gamma \vdash O_i : \OType(\sigma, \tau) \text{ for all $i$}}{\Gamma \vdash O_1 + \cdots + O_n : \OType(\sigma, \tau)}
+        \qquad \qquad
+        \textbf{Opt-Outer} \qquad
+        \frac{\Gamma\vdash K : \KType(\sigma) \qquad \Gamma\vdash B : \BType(\tau)}{\Gamma\vdash K \cdot B : \OType(\sigma, \tau)} \\
+        \\
+        \textbf{Opt-Dot} \qquad
+        \frac{\Gamma \vdash O_1 : \OType(\sigma, \tau) \qquad \Gamma \vdash O_2 : \OType(\tau, \rho)}{\Gamma \vdash O_1 \cdot O_2 : \OType(\sigma, \rho)} \\
+        \\
+        \textbf{Opt-Tsr} \qquad
+        \frac{\Gamma \vdash O_1 : \OType(\sigma_1, \tau_1) \qquad \Gamma \vdash O_2 : \OType(\sigma_2, \tau_2)} {\Gamma \vdash O_1 \otimes O_2 : \OType(\sigma_1 \times \sigma_2, \tau_1 \times \tau_2)}
+    \end{gather*}
+    \caption{Operator typing rules.}
+\end{figure}
+
+\clearpage
+\section{Conversions and Reductions}
+\begin{figure}[h]
+    \begin{gather*}
+        \textbf{Alpha} \qquad \text{(bound variable renaming)}
+        \qquad \qquad
+        \textbf{Beta} \qquad
+        \frac{}{\Gamma \vdash ((\lambda x:T.t)\ u)\ \reduce_\beta\ t\{x/u\}} \\
+        \\
+        \textbf{Delta} \qquad
+        \frac{\WF(\Gamma) \qquad (c:=t:T) \in \Gamma}{\Gamma \vdash c\ \reduce_\delta\ t}
+        \qquad \qquad
+        \textbf{Eta} \qquad
+        \frac{\Gamma\vdash t : T \to U}{\Gamma \vdash \lambda x : T . (t\ x)\ \reduce_\eta\ t}
+    \end{gather*}
+    \caption{Conversions and reductions for the typed lambda calculus.}
+\end{figure}
+
+\renewcommand{\arraystretch}{1.2} % Increases row height by 50%
+\begin{longtable}{cl}
+    \caption{sefsefsef}\\
+    \hline
+    \textbf{Rule} & \textbf{Description} \\
+    \hline
+    \endfirsthead
+
+    \hline
+    \textbf{Rule} & \textbf{Description} \\
+    \hline
+    \endhead
+
+    \hline
+    \multicolumn{2}{r}{\textit{Continued on the next page}} \\
+    \hline
+    \endfoot
+
+    \hline
+    \endlastfoot
+
+    %
+    R-FLAT
+    & $\ADD(a_1 \cdots \ADD(b_1 \cdots b_m) \cdots a_n)\ \reduce \ADD(a_1 \cdots b_1 \cdots b_m \cdots a_n)$ \\
+    & $\MUL(a_1 \cdots \MUL(b_1 \cdots b_m) \cdots a_n)\ \reduce \MUL(a_1 \cdots b_1 \cdots b_m \cdots a_n)$ \\
+    & \textcolor{gray}{A special rule to flatten all AC symbols within one call.} \\
+    %
+    R-ADDID 
+    & $\ADD(a) \ \reduce\ a$ \\
+    %
+    R-ADDS0 
+    & $\ADD(a_1 \cdots \ZERO \cdots a_n) \ \reduce\ \ADD(a_1 \cdots a_n)$ \\
+    & \textcolor{gray}{This rule removes all $\ZERO$ occurances and keeps the order of remaining subterms.} \\
+    R-MULID 
+    & $\MUL(a) \ \reduce\ a$ \\
+    %
+    R-MUL0
+    & $\MUL(a_1 \cdots \ZERO \cdots a_n) \ \reduce\ \ZERO$ \\
+    %
+    R-MUL1
+    & $\MUL(a_1 \cdots \ONE \cdots a_n) \ \reduce\ \MUL(a_1 \cdots a_n)$ \\
+    & \textcolor{gray}{Similar to (R-ADDS0).} \\
+    %
+    R-MUL2
+    & $\MUL(b_1 \cdots \ADD(a_1\ a_2 \cdots a_n) \cdots b_m)$ \\
+    & $\reduce\ \ADD(\MUL(b_1 \cdots a_1\cdots b_m)\ \MUL(b_1 \cdots a_2\cdots b_m) \cdots \MUL(b_1\cdots a_n \cdots b_m))$ \\
+    & \textcolor{gray}{This rule matches the first $\ADD$ subterm in the list.} \\
+    %
+    R-CJG0
+    & $\CJG(\ZERO) \ \reduce\ \ZERO$ \\
+    %
+    R-CJG1
+    & $\CJG(\ONE) \ \reduce\ \ONE$ \\
+    %
+    R-CJG2
+    & $ \CJG(\ADD(a_1\ \cdots a_n)) \ \reduce\ \ADD(\CJG(a_1) \cdots \CJG(a_n))$ \\
+    %
+    R-CJG3
+    & $ \CJG(\MUL(a_1\ \cdots a_n)) \ \reduce\ \MUL(\CJG(a_1) \cdots \CJG(a_n))$ \\
+    %
+    R-CJG4
+    & $ \CJG(\CJG(a)) \ \reduce\ a$ \\
+    % 
+    R-DELTA0
+    & $ \DELTA(a\ a) \ \reduce\ \ONE$ \\
+    %
+    R-DELTA1
+    & $ \DELTA(\PAIR(a\ b)\ \PAIR(c\ d)) \ \reduce\ \MUL(\DELTA(a\ c)\ \DELTA(b\ d))$ \\
+    %
+    R-KET0
+    & $[\Gamma \vdash K : \KType(\sigma)]\vdash[\Gamma \vdash \SCR(\ZERO\ K)\ \reduce\ \ZEROK(\sigma)]$ \\
+\end{longtable}
+
+About big operators: maybe we should use the "sum operators"?
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{\texttt{Diracoq} language}
+
+\begin{align*}
+    cmd ::=&\ \texttt{Def}(\texttt{ID}\ term) \\
+        &|\ \texttt{Def}(\texttt{ID}\ term\ type) \\
+        &|\ \texttt{Var}(\texttt{ID}\ term) \\
+        &|\ \texttt{Check}(term) \\
+        &|\ \texttt{Show}(\texttt{ID}) \\
+        &|\ \texttt{ShowAll} \\
+        &|\ \texttt{Normalize}(term) \\
+        &|\ \texttt{CheckEq}(term\ term) \\
+    type ::=&\ \texttt{Type}\ |\ \texttt{Arrow}(type\ type) \\
+            &|\ \texttt{Base} \\
+    term ::=&\ \texttt{Type}\ |\ \texttt{fun}(\texttt{ID}\ type\ term)\ |\ \texttt{apply}(term\ term)\ |\ \texttt{ID} \\
+\end{align*}
+
+\clearpage
 

### 2024-12-01 main.tex
syntax/typing/rules proposed

@@
 \usepackage{makecell}
+\usepackage{mdframed}
 
@@
 
+% framed environment
+\newmdenv[
+  skipabove=5pt,
+  skipbelow=5pt,
+  leftmargin=0pt,
+  rightmargin=0pt,
+  innerleftmargin=5pt,
+  innerrightmargin=5pt,
+  innertopmargin=5pt,
+  innerbottommargin=5pt,
+  linecolor=black,
+]{framedeq}
+
 \title{\textbf{DiracDec C++/Coq Implementation\cite{ExampleCitation}}}
-\author{Yingte Xu}
 
@@
 %%%%%%%%%%%%%%%%
-\section{Scalar: Minimal Example}
-
-We first consider a minimal example of the scalars.
-
-\subsection{Modeling AC by Sequence}
-The problem is presented in the normal AC term-rewriting setting, with a minor modification of flattening the AC symbols. This is the model implemented in Coq, and we consider the 
-C++ optimization only as a efficient searching algorithm for the rewriting trace.
 
@@
 \newcommand{\ONEO}{\mathbf{1}_\mathcal{O}}
-
-\begin{definition}[scalar syntax]
-    The syntax for scalars is defined as follows.
-    \[ 
-    S ::= \ZERO\ |\ \ONE\ |\ \ADDS ( S^+ )\ |\ \MULS ( S^+ )\ |\ \CONJ(S).
-    \]
-    Here $S^+$ means a sequence of one or more $S$ terms, i.e., AC symbols are flattened and have indefinite arities.
-\end{definition}
-
-The rewriting rules are presented as follows.
-
-\begin{figure}[h]
-    \begin{align*}
-        & \text{(R-FLATTEN)} &&
-        \textcolor{gray}{\text{A special rule to flatten all AC symbols within one call.}} \\
-        & \text{(R-ADDSID)} &&
-        \ADDS(a) \to a \\
-        & \text{(R-ADDS0)} &&
-        \ADDS((seq:\_ \_)\ \ZERO) \to \ADDS(seq) \\
-        & && \textcolor{gray}{\text{This rule removes all $\ZERO$ occurances and keeps the order of remaining subterms.}} \\
-        & \text{(R-MULSID)} &&
-        \MULS(a) \to a \\
-        & \text{(R-MULS0)} &&
-        \MULS((seq: \_ \_)\ \ZERO) \to \ZERO \\
-        & \text{(R-MULS1)} &&
-        \MULS((seq: \_ \_)\ \ONE) \to \MULS(seq) \\
-        & && \textcolor{gray}{\text{Similar to (R-ADDS0).}} \\
-        & \text{(R-MULS2)} &&
-        \MULS((seq1: \_ \_ \_)\ \ADDS(a_1\ a_2 \cdots a_n)\ (seq2: \_ \_ \_)) \\
-        & && \to \ADDS(\MULS(seq1\ a_1\ seq2)\ \MULS(seq1\ a_2\ seq2) \cdots \MULS(seq1\ a_n\ seq2)) \\
-        & && \textcolor{gray}{\text{This rule matches the first $\ADDS$ subterm in the list. $seq1$ or $seq2$ can be empty, but not both.}} \\
-        & \text{(R-CONJ0)} &&
-        \CONJ(\ZERO) \to \ZERO \\
-        & \text{(R-CONJ1)} &&
-        \CONJ(\ONE) \to \ONE \\
-        & \text{(R-CONJ2)} &&
-        \CONJ(\ADDS(a_1\ a_2\ \cdots a_n)) \to \ADDS(\CONJ(a_1)\ \CONJ(a_2) \cdots \CONJ(a_n)) \\
-        & \text{(R-CONJ3)} &&
-        \CONJ(\MULS(a_1\ a_2\ \cdots a_n)) \to \MULS(\CONJ(a_1)\ \CONJ(a_2) \cdots \CONJ(a_n)) \\
-        & \text{(R-CONJ4)} &&
-        \CONJ(\CONJ(a)) \to a
-    \end{align*}
-    \caption{The rewriting rules for scalars.}
-\end{figure}
-
-Notice that all the rules behaves the same as in the term count mapping implementation in C++.
-That is, the transformation from sequence representation to mapping representaiton commutes with all rewritings.
+\newcommand{\OUTER}{\mathsf{OUTER}}
+\newcommand{\MULK}{\mathsf{MULK}}
+\newcommand{\MULB}{\mathsf{MULB}}
+\newcommand{\MULO}{\mathsf{MULO}}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Rewriting Control and Intermediate Language}
-The associativity is already handled by the (R-FLATTEN) rule.
-In order to decide two terms $A$ and $B$ are equivalent under commutativity, we need to proof that $A$ can be transformed into $B$ with a structured permutation, which is described by the \textit{permutation tree}.
-
-\begin{definition}[permutation tree]
-    The syntax for permutation trees are inductively defined below:
-    \[
-    P ::= \mathsf{E}\ |\ [(i:P)^+].
-    \]
-    Here $i$ represents positive numbers.
-\end{definition}
-We always only consider \textit{well-formed} permutation trees. That is, if $P \equiv [i_1:P_1\ i_2:P_2\ \cdots\ i_n:P_n]$, then $\{i_1, ... i_n\}$ forms the set of integers from $0$ to $n-1$.
-
-We can transform a term A with a suitable permutation tree. The transformation is defined as
-
-\begin{align*}
-    \texttt{apply}(A, P) :=\ & \texttt{match}\ P\ \texttt{with} \\
-    & \quad |\ \textsf{E} \Rightarrow A \\
-    & \quad |\ [i_1:P_1\ \cdots\ i_n:P_n] \Rightarrow \texttt{A.head}(\texttt{apply(A.args[$i_1$], $P_{i_1}$)}\ \cdots\ \texttt{apply(A.args[$i_n$], $P_{i_n}$)}) \\
-    & \texttt{end}
-\end{align*}
-
-% \subsection{Coq Implementation}
-% The following abilities should be implemented on the Coq side:
-% \begin{itemize}
-%     \item \textbf{Deciding $E$}.
-For the C++ implementation, the data structure itself already reflects $E$.
-Therefore, on the Coq side, this should also be decided automatically, implemneted as a lemma
-%     \[
-%         \texttt{EQ\_E : forall (A\ B: Term), Optional(A = B)}
-%     \]
-%     \item \textbf{The computational function to output the equivalence lemma for all rules.} For example, for the (R-ADDS0) rule, there should be such a lemma:
-%     \[
-%         \texttt{E\_R\_ADDS0 : forall (A\ B: Term), Optional(A = B)}
-%     \]
-%     which outputs the proof for the equivalence proof when $A$ matches the left side of (R-ADDS0) and is rewriten into B.
-% \end{itemize}
-
-% With these Coq implementations, the C++ can output the script for equivalence proof in the following format:
-
-% \begin{align*}
-%     & \texttt{rewrite (EQ\_E S1 T1).}\\
-%     & \texttt{rewrite (EQ\_RULE1 A1 B1).}\\
-%     & \texttt{rewrite (EQ\_E S2 T2).}\\
-%     & \texttt{rewrite (EQ\_RULE2 A2 B2).}\\
-%     & \texttt{rewrite (EQ\_E S3 T3).}\\
-%     & \texttt{rewrite (EQ\_RULE3 A3 B3).}\\
-%     & \cdots
-% \end{align*}
-
-% where at each step, the C++ solver rewrites the subterm \texttt{Ai} using rule \texttt{RULEi}.
-The Coq prover will synchronize with C++ solver, because both systems replace all occurances of $Ai$ into $Bi$.
-
-% \yx{Actually the \texttt{B} inputs can be inferred and is not necessary.}
-
-% \yx{Question: How to decide the subterms \texttt{Si} and \texttt{Ti}.}
-
-% \yx{Question: The syntax for terms that Coq accepts.}
-
-\clearpage
 
@@
         e ::=\ & x\ |\ (e_1, e_2) \\
-        & |\ 0\ |\ 1\ |\ e_1 + \cdots + e_n \ |\ e_1 \times \cdots \times e_n\ |\ e^* \ |\ \delta_{e_1, e_2}\ \\
+        & |\ 0\ |\ 1\ |\ \ADDS(e_1 \cdots e_n) \ |\ e_1 \times \cdots \times e_n\ |\ e^* \ |\ \delta_{e_1, e_2}\ |\ \DOT(e_1\ e_2) \\
         & |\ \ZEROK(\sigma)\ |\ \ZEROB(\sigma)\ |\ \ZEROO(\sigma_1, \sigma_2)\ |\ \ONEO(\sigma) \\
-        & |\ \ket{e}\ |\ \bra{t}\ |\ e^\dagger\ |\ e_1.e_2\ |\ e_1 \cdot e_2\ |\ e_1 \otimes e_2.
+        & |\ \ket{e}\ |\ \bra{t}\ |\ e^\dagger\ |\ e_1.e_2\ |\ \ADD(e_1 \cdots e_n)\ |\ e_1 \otimes e_2 \\
+        & |\ \MULK(e_1\ e_2)\ |\ \MULB(e_1\ e_2)\ |\ \OUTER(e_1\ e_2)\ |\ \MULO(e_1\ e_2).
     \end{align*}
@@
 
+Compared to [?], this syntax for Dirac notations merges the symbols with overlapped properties, such as the addition and scaling symbols for ket, bra and operator.
+Here $\ADDS$ and $\ADD$ are two different AC symbols representing the scalar addition and the linear algebra addition respectively.
+They will be denoted as $a_1 + \cdots + a_n$ and $X_1 + \cdots + X_n$.
+There are five kinds of linear algebra multiplications among ket, bra and operator, whose properties are similar but still diverge to some extent.
+For example, the rules $(O_1 \cdot O_2) \cdot K\ \reduce\ O_1 \cdot (O_2 \cdot K)$ and $B \cdot (O_1 \cdot O_2) \ \reduce\ (B \cdot O_1) \cdot O_2$ indicate that the sorting of multiplication sequences depends on the subterm types.
+To avoid frequent but unnecessary type checkings, we encode the typing information by using five different symbols, namely $\DOT$, $\MULK$, $\MULB$, $\OUTER$ and $\MULO$.
+They are denoted as $B\cdot K$, $K_1 \cdot K_2$, $B_1 \cdot B_2$, $K \cdot B$ and $O_1 \cdot O_2$, respectively.
+
 \section{Typing System}
-The typing system modifies the one for Calculus of Construction, adding special symbols for Dirac notations.
-According to the CoC typing rules, the environment and context can be modeled by as two stacks.
-It is obvious that popping assumptions/definitions from any context or from a environment with empty context will not destroy the well-formedness.
-
-One important reason for using dependent type is that, the types of Dirac notatinos (i.e., ket, bra and operators) depends on the basis.
-
+The type checking of Dirac notations involves maintaining a well-formed context $\Gamma$, which specifies the definitions and typing assumtpions for variables.
+Notice that the type checking $\Gamma \vdash t : T$ is decidable.
 \begin{figure}[h]
@@
         \qquad \qquad
-        \textbf{Ket-Sca} \qquad
+        \textbf{Ket-Scr} \qquad
         \frac{\Gamma \vdash a : \SType \qquad \Gamma \vdash K : \KType(\sigma)}{\Gamma \vdash a.K : \KType(\sigma)} \\
@@
         \qquad \qquad
-        \textbf{Ket-Dot} \qquad
+        \textbf{Ket-MulK} \qquad
         \frac{\Gamma \vdash O : \OType(\sigma, \tau) \qquad \Gamma \vdash K : \KType(\tau)}{\Gamma \vdash O \cdot K : \KType(\sigma)} \\
@@
         \qquad \qquad
-        \textbf{Bra-Sca} \qquad
+        \textbf{Bra-Scr} \qquad
         \frac{\Gamma \vdash a : \SType \qquad \Gamma \vdash B : \BType(\sigma)}{\Gamma \vdash a.B : \BType(\sigma)} \\
@@
         \qquad \qquad
-        \textbf{Bra-Dot} \qquad
+        \textbf{Bra-MulB} \qquad
         \frac{\Gamma \vdash B : \KType(\sigma) \qquad \Gamma \vdash O : \OType(\sigma, \tau)}{\Gamma \vdash B \cdot O : \BType(\tau)} \\
@@
         \qquad \qquad
-        \textbf{Opt-Sca} \qquad
+        \textbf{Opt-Scr} \qquad
         \frac{\Gamma \vdash a : \SType \qquad \Gamma \vdash O : \OType(\sigma, \tau)}{\Gamma \vdash a.O : \OType(\sigma, \tau)} \\
@@
         \\
-        \textbf{Opt-Dot} \qquad
+        \textbf{Opt-MulO} \qquad
         \frac{\Gamma \vdash O_1 : \OType(\sigma, \tau) \qquad \Gamma \vdash O_2 : \OType(\tau, \rho)}{\Gamma \vdash O_1 \cdot O_2 : \OType(\sigma, \rho)} \\
@@
 
-\renewcommand{\arraystretch}{1.2} % Increases row height by 50%
+\yx{Alpha rule will be implemented by renaming, and we can leave it for the future.}
+
+\newenvironment{ruletable}[1]
+{
     \begin{longtable}{cl}
-    \caption{sefsefsef}\\
+    \caption{#1}\\
     \hline
@@
     \endlastfoot
+}
+{
+    \end{longtable}
+}
 
+\renewcommand{\arraystretch}{1.2} % Increases row height by 50%
+
+\begin{ruletable}{The special flattening rule.}
     %
-    R-FLAT
-    & $\ADD(a_1 \cdots \ADD(b_1 \cdots b_m) \cdots a_n)\ \reduce \ADD(a_1 \cdots b_1 \cdots b_m \cdots a_n)$ \\
-    & $\MUL(a_1 \cdots \MUL(b_1 \cdots b_m) \cdots a_n)\ \reduce \MUL(a_1 \cdots b_1 \cdots b_m \cdots a_n)$ \\
-    & \textcolor{gray}{A special rule to flatten all AC symbols within one call.} \\
+    R-FLATTEN
+    & $a_1 + \cdots + (b_1 + \cdots + b_m) + \cdots + a_n\ \reduce\ a_1 + \cdots + b_1 + \cdots + b_m + \cdots + a_n$ \\
+    & $a_1 \times \cdots \times (b_1 \times \cdots \times b_m) \times \cdots \times a_n\ \reduce\ a_1 \times \cdots \times b_1 \times \cdots \times b_m \times \cdots \times a_n$ \\
+    & $X_1 + \cdots + (X_1' + \cdots + X_m') + \cdots + X_n\ \reduce\ X_1 + \cdots + X_1' + \cdots + X_m' + \cdots + X_n$ \\
+    & \textcolor{gray}{A special rule to flatten all AC symbols within one call.}    
+\end{ruletable}
+
+\begin{ruletable}{Rules for scalar addition and multiplication.}
+    %
+    R-ADDSID 
+    & $+(a) \ \reduce\ a$ \\
+    & \textcolor{gray}{Reduce the term if the AC symbol $+$ only has one argument.} \\
+    %
+    R-MULSID 
+    & $\times(a) \ \reduce\ a$ \\
+    & \textcolor{gray}{Similar to (R-ADDSID).} \\
+    %
+    R-ADDS0 
+    & $a_1 + \cdots + 0 + \cdots + a_n \ \reduce\ a_1 + \cdots + a_n$ \\
+    & \textcolor{gray}{This rule removes all $\ZERO$ occurances and keeps the order of remaining subterms.} \\
+    %
+    R-MULS0
+    & $a_1 \times \cdots \times 0 \times \cdots \times a_n \ \reduce\ 0$ \\
+    %
+    R-MULS1
+    & $a_1 \times \cdots \times \ONE \times \cdots \times a_n \ \reduce\ a_1 \times \cdots \times a_n$ \\
+    & \textcolor{gray}{Similar to (R-ADDS0).} \\
+    %
+    R-MULS2
+    & $b_1 \times \cdots \times (a_1 + \cdots + a_n) \times \cdots \times b_m$ \\
+    & $\reduce\ (b_1 \times \cdots \times a_1 \times \cdots \times b_m) + \cdots + (b_1 \times \cdots \times a_n \times \cdots \times b_m)$ \\
+    & \textcolor{gray}{This rule matches the first scalar addition subterm in the list.} \\
+\end{ruletable}
+
+\begin{ruletable}{Rules for other scalar symbols.}
+    %
+    R-CONJ0
+    & $0^* \ \reduce\ 0$ \\
+    %
+    R-CONJ1
+    & $1^* \ \reduce\ 1$ \\
+    %
+    R-CONJ2
+    & $(a_1 + \cdots + a_n)^* \ \reduce\ a_1^* + \cdots + a_n^*$ \\
+    %
+    R-CONJ3
+    & $(a_1 \times \cdots \times a_n)^* \ \reduce\ a_1^* \times \cdots \times a_n^*$ \\
+    %
+    R-CONJ4
+    & $ (a^*)^* \ \reduce\ a$ \\
+    % 
+    R-CONJ5
+    & $ \delta_{s, t}^* \ \reduce\ \delta_{s, t}$ \\
+    %
+    R-CONJ6
+    & $ (B \cdot K)^* \ \reduce\ K^\dagger \cdot B^\dagger $ \\
+    %
+    R-DOT0
+    & $ \ZEROB(\sigma) \cdot K \ \reduce\ 0 $ \\
+    %
+    R-DOT1
+    & $ B \cdot \ZEROK(\sigma) \ \reduce\ 0 $ \\
+    %
+    R-DOT2
+    & $ (a.B) \cdot K \ \reduce\ a \times (B \cdot K) $ \\
+    %
+    R-DOT3
+    & $ B \cdot (a.K) \ \reduce\ a \times (B \cdot K) $ \\
+    %
+    R-DOT4
+    & $ (B_1 + \cdots + B_n) \cdot K \ \reduce\ B_1 \cdot K + \cdots + B_n \cdot K $ \\
+    %
+    R-DOT5
+    & $ B \cdot (K_1 + \cdots + K_n) \ \reduce\ B \cdot K_1 + \cdots + B \cdot K_n $ \\
+    %
+    R-DOT6
+    & $ \bra{s} \cdot \ket{t} \ \reduce\ \delta_{s, t} $ \\
+    %
+    R-DOT7
+    & $ (B_1 \otimes B_2) \cdot \ket{(s, t)} \ \reduce\ (B_1 \cdot \ket{s}) \times (B_2 \cdot \ket{t}) $ \\
+    %
+    R-DOT8
+    & $ \bra{(s, t)} \cdot (K_1 \otimes K_2) \ \reduce\ (\bra{s} \cdot K_1) \times (\bra{t} \cdot K_2) $ \\
+    %
+    R-DOT9
+    & $ (B_1 \otimes B_2) \cdot (K_1 \otimes K_2) \ \reduce\ (B_1 \cdot K_1) \times (B_2 \cdot K_2) $ \\
+    %
+    R-DOT10
+    & $ (B \cdot O) \cdot K \ \reduce\ B \cdot (O \cdot K) $ \\
+    %
+    R-DOT11
+    & $ \bra{(s, t)} \cdot ((O_1 \otimes O_2) \cdot K) \ \reduce\ ((\bra{s} \cdot O_1) \otimes (\bra{t} \cdot O_2)) \cdot K $ \\
+    %
+    R-DOT12
+    & $ (B_1 \otimes B_2) \cdot ((O_1 \otimes O_2) \cdot K) \ \reduce\ ((B_1 \cdot O_1) \otimes (B_2 \cdot O_2)) \cdot K $ \\
+    %
+    R-DELTA0
+    & $ \delta_{a, a} \ \reduce\ 1$ \\
+    %
+    R-DELTA1
+    & $ \delta_{(a, b), (c, d)} \ \reduce\ \delta_{a, c} \times \delta_{b, d}$ \\
+\end{ruletable}
+
+\begin{ruletable}{Rules for linear algebra scaling.}
+    %
+    R-SCR0
+    & $ 1.X \ \reduce\ X $ \\
+    %
+    R-SCR1
+    & $ a.(b.X) \ \reduce\ (a \times b).X $ \\
+    %
+    R-SCR2
+    & $ a.(X_1 + \cdots + X_n) \ \reduce\ a.X_1 + \cdots + a.X_n $ \\
+    %
+    R-SCRK0
+    & $ K : \KType(\sigma) \Rightarrow 0.K \ \reduce\ \ZEROK(\sigma)$ \\
+    %
+    R-SCRK1
+    & $ a.\ZEROK(\sigma) \ \reduce\ \ZEROK(\sigma) $ \\
+    % 
+    R-SCRB0
+    & $ B : \BType(\sigma) \Rightarrow 0.B \ \reduce\ \ZEROB(\sigma)$ \\
+    % 
+    R-SCRB1
+    & $ a.\ZEROB(\sigma) \ \reduce\ \ZEROB(\sigma) $ \\
+    % 
+    R-SCRO0
+    & $ O : \OType(\sigma, \tau) \Rightarrow 0.O \ \reduce\ \ZEROO(\sigma, \tau)$ \\
+    % 
+    R-SCRO1
+    & $ a.\ZEROO(\sigma, \tau) \ \reduce\ \ZEROO(\sigma, \tau) $ \\
+\end{ruletable}
+
+\begin{ruletable}{Rules for linear algebra addition. }
     %
     R-ADDID
-    & $\ADD(a) \ \reduce\ a$ \\
+    & $+(X) \ \reduce\ X$ \\
+    & \textcolor{gray}{Reduce the term if the AC symbol $+$ only has one argument.} \\
     %
-    R-ADDS0 
-    & $\ADD(a_1 \cdots \ZERO \cdots a_n) \ \reduce\ \ADD(a_1 \cdots a_n)$ \\
-    & \textcolor{gray}{This rule removes all $\ZERO$ occurances and keeps the order of remaining subterms.} \\
-    R-MULID 
-    & $\MUL(a) \ \reduce\ a$ \\
+    R-ADD0
+    & $a_1 + \cdots + X + \cdots + X + \cdots + a_n \ \reduce\ a_1 + \cdots + a_n + \cdots + (1+1).X$ \\
+    & \textcolor{gray}{\makecell[l]{This rule matches the first $X$ in the list satisfying the pattern. \\ The result $(1+1).X$ will be placed at the end of the list.}} \\
     %
-    R-MUL0
-    & $\MUL(a_1 \cdots \ZERO \cdots a_n) \ \reduce\ \ZERO$ \\
+    R-ADD1
+    & $a_1 + \cdots + X + \cdots + a.X + \cdots + a_n \ \reduce\ a_1 + \cdots + a_n + (1+a).X$ \\
+    & \textcolor{gray}{Similar to (R-ADD0).} \\
     %
-    R-MUL1
-    & $\MUL(a_1 \cdots \ONE \cdots a_n) \ \reduce\ \MUL(a_1 \cdots a_n)$ \\
+    R-ADD2
+    & $a_1 + \cdots + a.X + \cdots + X + \cdots + a_n \ \reduce\ a_1 + \cdots + a_n + (a+1).X$ \\
+    & \textcolor{gray}{Similar to (R-ADD0).} \\
+    %
+    R-ADD3
+    & $a_1 + \cdots + a.X + \cdots + b.X + \cdots + a_n \ \reduce\ a_1 + \cdots + a_n + (a+b).X$ \\
+    & \textcolor{gray}{Similar to (R-ADD0).} \\
+    %
+    R-ADDK0
+    & $K_1 + \cdots + \ZEROK(\sigma) + \cdots + K_n\ \reduce K_1 + \cdots + K_n$ \\
     & \textcolor{gray}{Similar to (R-ADDS0).} \\
     %
-    R-MUL2
-    & $\MUL(b_1 \cdots \ADD(a_1\ a_2 \cdots a_n) \cdots b_m)$ \\
-    & $\reduce\ \ADD(\MUL(b_1 \cdots a_1\cdots b_m)\ \MUL(b_1 \cdots a_2\cdots b_m) \cdots \MUL(b_1\cdots a_n \cdots b_m))$ \\
-    & \textcolor{gray}{This rule matches the first $\ADD$ subterm in the list.} \\
+    R-ADDB0
+    & $B_1 + \cdots + \ZEROB(\sigma) + \cdots + B_n \ \reduce\ B_1 + \cdots + B_n$ \\
+    & \textcolor{gray}{Similar to (R-ADDS0).} \\
     %
-    R-CJG0
-    & $\CJG(\ZERO) \ \reduce\ \ZERO$ \\
+    R-ADDO0
+    & $O_1 + \cdots + \ZEROO(\sigma, \tau) + \cdots + O_n \ \reduce\ O_1 + \cdots + O_n$ 
+    \\
+    & \textcolor{gray}{Similar to (R-ADDS0).} \\
+\end{ruletable}
+
+\begin{ruletable}{Rules for adjoint.}
     %
-    R-CJG1
-    & $\CJG(\ONE) \ \reduce\ \ONE$ \\
+    R-ADJ0
+    & $ (X^\dagger)^\dagger \ \reduce\ X $ \\
     %
-    R-CJG2
-    & $ \CJG(\ADD(a_1\ \cdots a_n)) \ \reduce\ \ADD(\CJG(a_1) \cdots \CJG(a_n))$ \\
+    R-ADJ1
+    & $ (a.X)^\dagger \ \reduce\ (a^*).(X^\dagger) $ \\
     %
-    R-CJG3
-    & $ \CJG(\MUL(a_1\ \cdots a_n)) \ \reduce\ \MUL(\CJG(a_1) \cdots \CJG(a_n))$ \\
+    R-ADJ2
+    & $ (X_1 + \cdots + X_n)^\dagger \ \reduce\ X_1^\dagger + \cdots + X_n^\dagger $ \\
     %
-    R-CJG4
-    & $ \CJG(\CJG(a)) \ \reduce\ a$ \\
+    R-ADJK0
+    & $ \ZEROB(\sigma)^\dagger \ \reduce\ \ZEROK(\sigma) $ \\
     %
-    R-DELTA0
-    & $ \DELTA(a\ a) \ \reduce\ \ONE$ \\
+    R-ADJK1
+    & $ \bra{t}^\dagger \ \reduce\ \ket{t} $ \\
     %
-    R-DELTA1
-    & $ \DELTA(\PAIR(a\ b)\ \PAIR(c\ d)) \ \reduce\ \MUL(\DELTA(a\ c)\ \DELTA(b\ d))$ \\
+    R-ADJK2
+    & $ (B \cdot O)^\dagger\ \reduce\ O^\dagger \cdot B^\dagger $ \\
     %
-    R-KET0
-    & $[\Gamma \vdash K : \KType(\sigma)]\vdash[\Gamma \vdash \SCR(\ZERO\ K)\ \reduce\ \ZEROK(\sigma)]$ \\
-\end{longtable}
+    R-ADJK3
+    & $ (B_1 \otimes B_2)^\dagger \ \reduce\ B_1^\dagger \otimes B_2^\dagger $ \\
+    %
+    R-ADJB0
+    & $ \ZEROK(\sigma)^\dagger \ \reduce\ \ZEROB(\sigma) $ \\
+    %
+    R-ADJB1
+    & $ \ket{t}^\dagger \ \reduce\ \bra{t} $ \\
+    %
+    R-ADJB2
+    & $ (O \cdot K)^\dagger\ \reduce\ K^\dagger \cdot O^\dagger $ \\
+    %
+    R-ADJB3
+    & $ (K_1 \otimes K_2)^\dagger \ \reduce\ K_1^\dagger \otimes K_2^\dagger $ \\
+    %
+    R-ADJO0
+    & $ \ZEROO(\sigma, \tau)^\dagger \ \reduce\ \ZEROO(\tau, \sigma) $ \\
+    %
+    R-ADJO1
+    & $ \ONEO(\sigma)^\dagger \ \reduce\ \ONEO(\sigma)$ \\
+    %
+    R-ADJO2
+    & $ (K \cdot B)^\dagger \ \reduce\ B^\dagger \cdot K^\dagger$ \\
+    %
+    R-ADJO3
+    & $ (O_1 \cdot O_2)^\dagger \ \reduce\ O_2^\dagger \cdot O_1^\dagger $ \\
+    %
+    R-ADJO4
+    & $ (O_1 \otimes O_2)^\dagger \ \reduce\ O_1^\dagger \otimes O_2^\dagger $ 
+\end{ruletable}
 
-About big operators: maybe we should use the "sum operators"?
+\begin{ruletable}{Rules for tensor product.}
+    R-TSR0
+    & $ (a.X_1) \otimes X_2 \ \reduce\ a.(X_1 \otimes X_2) $ \\
+    %
+    R-TSR1
+    & $ X_1 \otimes (a.X_2) \ \reduce\ a.(X_1 \otimes X_2) $ \\
+    %
+    R-TSR2
+    & $ (X_1 + \cdots + X_n) \otimes X' \reduce X_1 \otimes X' + \cdots + X_n \otimes X' $ \\
+    %
+    R-TSR3
+    & $ X' \otimes (X_1 + \cdots + X_n) \reduce X' \otimes X_1 + \cdots + X' \otimes X_n $ \\
+    %
+    R-TSRK0
+    & $ K : \KType(\tau) \Rightarrow \ZEROK(\sigma) \otimes K\ \reduce\ \ZEROK(\sigma \times \tau) $ \\
+    %
+    R-TSRK1
+    & $ K : \KType(\tau) \Rightarrow K \otimes \ZEROK(\sigma)\ \reduce\ \ZEROK(\tau \times \sigma) $ \\
+    %
+    R-TSRK2
+    & $\ket{s} \otimes \ket{t} \ \reduce\ \ket{(s, t)}$ \\
+    %
+    R-TSRB0
+    & $ B : \BType(\tau) \Rightarrow \ZEROB(\sigma) \otimes B\ \reduce\ \ZEROB(\sigma \times \tau) $ \\
+    %
+    R-TSRB1
+    & $ B : \BType(\tau) \Rightarrow B \otimes \ZEROB(\sigma)\ \reduce\ \ZEROB(\tau \times \sigma) $ \\
+    %
+    R-TSRB2
+    & $\bra{s} \otimes \bra{t} \ \reduce\ \bra{(s, t)}$ \\
+    %
+    R-TSRO0
+    & $ O : \OType(\sigma, \tau) \Rightarrow O \otimes \ZEROO(\sigma', \tau') \ \reduce\ \ZEROO(\sigma \times \sigma', \tau \times \tau') $ \\
+    % 
+    R-TSRO1
+    & $ O : \OType(\sigma, \tau) \Rightarrow \ZEROO(\sigma', \tau') \otimes O \ \reduce\ \ZEROO(\sigma' \times \sigma, \tau' \times \tau) $ \\
+    %
+    R-TSRO2
+    & $\ONEO(\sigma) \otimes \ONEO(\tau)\ \reduce\ \ONEO(\sigma \times \tau)$ \\
+    %
+    R-TSRO3
+    & $ (K_1 \cdot B_1) \otimes (K_2 \cdot B_2)\ \reduce\ (K_1 \otimes K_2) \cdot (B_1 \otimes B_2) $ \\
+    %
+\end{ruletable}
+
+\begin{ruletable}{Rule for $O\cdot K$.}
+    %
+    R-MULK0
+    & $ \ZEROO(\sigma, \tau) \cdot K \ \reduce\ \ZEROK(\sigma) $ \\
+    %
+    R-MULK1
+    & $ O : \OType(\sigma, \tau) \Rightarrow O \cdot \ZEROK(\tau) \ \reduce\ \ZEROK(\sigma) $ \\
+    %
+    R-MULK2
+    & $ \ONEO(\sigma) \cdot K \ \reduce K $ \\
+    %
+    R-MULK3
+    & $ (a.O) \cdot K \ \reduce\ a.(O \cdot K) $ \\
+    %
+    R-MULK4
+    & $ O \cdot (a.K) \ \reduce\ a.(O \cdot K) $ \\
+    %
+    R-MULK5
+    & $ (O_1 + \cdots + O_n) \cdot K \ \reduce\ O_1 \cdot K + \cdots + O_n \cdot K $ \\
+    %
+    R-MULK6
+    & $ O \cdot (K_1 + \cdots + K_n) \ \reduce\ O \cdot K_1 + \cdots + O \cdot K_n $ \\
+    %
+    R-MULK7
+    & $ (K_1 \cdot B) \cdot K_2 \ \reduce\ (B \cdot K_2).K_1 $ \\
+    %
+    R-MULK8
+    & $ (O_1 \cdot O_2) \cdot K \ \reduce\ O_1 \cdot (O_2 \cdot K) $ \\
+    %
+    R-MULK9
+    & $ (O_1 \otimes O_2) \cdot ((O_1' \otimes O_2') \cdot K)\ \reduce\ ((O_1 \cdot O_1') \otimes (O_2 \cdot O_2')) \cdot K $ \\
+    %
+    R-MULK10
+    & $ (O_1 \otimes O_2) \cdot \ket{(s, t)}\ \reduce\ (O_1 \cdot \ket{s}) \otimes (O_2 \cdot \ket{t}) $ \\
+    %
+    R-MULK11
+    & $ (O_1 \otimes O_2) \cdot (K_1 \otimes K_2)\ \reduce\ (O_1 \cdot K_1) \otimes (O_2 \cdot K_2) $
+\end{ruletable}
+
+\begin{ruletable}{Rule for $B\cdot O$.}
+    %
+    R-MULB0
+    & $ B \cdot \ZEROO(\sigma, \tau) \ \reduce\ \ZEROB(\tau) $ \\
+    %
+    R-MULB1
+    & $ O : \OType(\sigma, \tau) \Rightarrow \ZEROB(\sigma) \cdot O\ \reduce\ \ZEROB(\tau) $ \\
+    %
+    R-MULB2
+    & $ B \cdot \ONEO(\sigma) \ \reduce B $ \\
+    %
+    R-MULB3
+    & $ (a.B) \cdot O \ \reduce\ a.(B \cdot O) $ \\
+    %
+    R-MULB4
+    & $ B \cdot (a.O) \ \reduce\ a.(B \cdot O) $ \\
+    %
+    R-MULB5
+    & $ (B_1 + \cdots + B_n) \cdot O \ \reduce\ B_1 \cdot O + \cdots + B_n \cdot O $ \\
+    %
+    R-MULB6
+    & $ B \cdot (O_1 + \cdots + O_n) \ \reduce\ B \cdot O_1 + \cdots + B \cdot O_n $ \\
+    %
+    R-MULB7
+    & $ B_1 \cdot (K \cdot B_2) \ \reduce\ (B_1 \cdot K).B_2 $ \\
+    %
+    R-MULB8
+    & $ B \cdot (O_1 \cdot O_2) \ \reduce\ (B \cdot O_1) \cdot O_2 $ \\
+    %
+    R-MULB9
+    & $ (B \cdot (O_1' \otimes O_2')) \cdot (O_1 \otimes O_2) \ \reduce\ B \cdot ((O_1' \otimes O_2') \cdot (O_1 \otimes O_2)) $ \\
+    %
+    R-MULB10
+    & $ \bra{(s, t)} \cdot (O_1 \otimes O_2)\ \reduce\ (\bra{s} \cdot O_1) \otimes (\bra{t} \cdot O_2) $ \\
+    %
+    R-MULB11
+    & $ (B_1 \otimes B_2) \cdot (O_1 \otimes O_2)\ \reduce\ (B_1 \cdot O_1) \otimes (B_2 \cdot O_2) $
+\end{ruletable}
+
+\begin{ruletable}{Rules for $K \cdot B$.}
+    %
+    R-OUTER0
+    & $ B : \BType(\tau) \Rightarrow \mathbf{0}_\mathcal{K}(\sigma) \cdot B\ \reduce\ \mathbf{0}_\mathcal{O}(\sigma, \tau) $ \\
+    %
+    R-OUTER1
+    & $ K : \KType(\sigma) \Rightarrow K \cdot \mathbf{0}_\mathcal{B}(\tau)\ \reduce\ \mathbf{0}_\mathcal{O}(\sigma, \tau) $ \\
+    %
+    R-OUTER2
+    & $ (a.K) \cdot B\ \reduce\ a.(K \cdot B) $ \\
+    %
+    R-OUTER3
+    & $ K \cdot (a.B)\ \reduce\ a.(K \cdot B) $ \\
+    %
+    R-OUTER4
+    & $ (K_1 + \cdots + K_n) \cdot B\ \reduce\ K_1 \cdot B + \cdots + K_n \cdot B $ \\
+    %
+    R-OUTER5
+    & $ K \cdot (B_1 + \cdots + B_n)\ \reduce\ K \cdot B_1 + \cdots + K \cdot B_n $ \\
+\end{ruletable}
+
+\begin{ruletable}{Rules for $O_1 \cdot O_2$.}
+    %
+    R-MULO0
+    & $ O : \OType(\tau, \rho) \Rightarrow \ZEROO(\sigma, \tau) \cdot O\ \reduce\ \ZEROO(\sigma, \rho) $ \\
+    %
+    R-MULO1
+    & $ O : \OType(\sigma, \tau) \Rightarrow O \cdot \ZEROO(\tau, \rho)\ \reduce\ \mathbf{0}_\mathcal{O}(\sigma, \rho) $ \\
+    %
+    R-MULO2
+    & $ \ONEO(\sigma) \cdot O \ \reduce\ O $ \\
+    %
+    R-MULO3
+    & $ O \cdot \ONEO(\sigma) \ \reduce\ O $ \\
+    %
+    R-MULO4
+    & $ (K \cdot B) \cdot O \ \reduce\ K \cdot (B \cdot O) $ \\
+    %
+    R-MULO5
+    & $ O \cdot (K \cdot B) \ \reduce\ (O \cdot K) \cdot B $ \\
+    %
+    R-MULO6
+    & $ (a.O_1) \cdot O_2 \ \reduce\ a.(O_1 \cdot O_2) $ \\
+    %
+    R-MULO7
+    & $ O_1 \cdot (a.O_2) \ \reduce\ a.(O_1 \cdot O_2) $ \\
+    %
+    R-MULO8
+    & $ (O_1 + \cdots + O_n) \cdot O'\ \reduce\ O_1 \cdot O' + \cdots + O_n \cdot O' $ \\
+    %
+    R-MULO9
+    & $ O' \cdot (O_1 + \cdots + O_n)\ \reduce\ O' \cdot O_1 + \cdots + O' \cdot O_n $ \\
+    %
+    R-MULO10
+    & $ (O_1 \cdot O_2) \cdot O_3\ \reduce\ O_1 \cdot (O_2 \cdot O_3) $ \\
+    %
+    R-MULO11
+    & $ (O_1 \otimes O_2) \cdot (O_1' \otimes O_2')\ \reduce\ (O_1 \cdot O_1') \otimes (O_2 \cdot O_2') $ \\
+    %
+    R-MULO12
+    & $ (O_1 \otimes O_2) \cdot ((O_1' \otimes O_2') \cdot O_3)\ \reduce\ ((O_1 \cdot O_1') \otimes (O_2 \cdot O_2')) \cdot O_3 $ \\  
+\end{ruletable}
+
+% About big operators: maybe we should use the "sum operators"?
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\clearpage
 
@@
 
+\section{Rewriting Control and Intermediate Language}
+The associativity is already handled by the (R-FLATTEN) rule.
+In order to decide two terms $A$ and $B$ are equivalent under commutativity, we need to proof that $A$ can be transformed into $B$ with a structured permutation, which is described by the \textit{permutation tree}.
+
+\begin{definition}[permutation tree]
+    The syntax for permutation trees are inductively defined below:
+    \[
+    P ::= \mathsf{E}\ |\ [(i:P)^+].
+    \]
+    Here $i$ represents positive numbers.
+\end{definition}
+We always only consider \textit{well-formed} permutation trees. That is, if $P \equiv [i_1:P_1\ i_2:P_2\ \cdots\ i_n:P_n]$, then $\{i_1, ... i_n\}$ forms the set of integers from $0$ to $n-1$.
+
+We can transform a term A with a suitable permutation tree. The transformation is defined as
+
+\begin{align*}
+    \texttt{apply}(A, P) :=\ & \texttt{match}\ P\ \texttt{with} \\
+    & \quad |\ \textsf{E} \Rightarrow A \\
+    & \quad |\ [i_1:P_1\ \cdots\ i_n:P_n] \Rightarrow \texttt{A.head}(\texttt{apply(A.args[$i_1$], $P_{i_1}$)}\ \cdots\ \texttt{apply(A.args[$i_n$], $P_{i_n}$)}) \\
+    & \texttt{end}
+\end{align*}
+
 \clearpage

### 2024-12-03 main.tex
updated ADJ rules

@@
     R-ADD0
-    & $a_1 + \cdots + X + \cdots + X + \cdots + a_n \ \reduce\ a_1 + \cdots + a_n + \cdots + (1+1).X$ \\
+    & $Y_1 + \cdots + X + \cdots + X + \cdots + Y_n \ \reduce\ Y_1 + \cdots + Y_n + \cdots + (1+1).X$ \\
     & \textcolor{gray}{\makecell[l]{This rule matches the first $X$ in the list satisfying the pattern. \\ The result $(1+1).X$ will be placed at the end of the list.}} \\
@@
     R-ADD1
-    & $a_1 + \cdots + X + \cdots + a.X + \cdots + a_n \ \reduce\ a_1 + \cdots + a_n + (1+a).X$ \\
+    & $Y_1 + \cdots + X + \cdots + a.X + \cdots + Y_n \ \reduce\ Y_1 + \cdots + Y_n + (1+a).X$ \\
     & \textcolor{gray}{Similar to (R-ADD0).} \\
@@
     R-ADD2
-    & $a_1 + \cdots + a.X + \cdots + X + \cdots + a_n \ \reduce\ a_1 + \cdots + a_n + (a+1).X$ \\
+    & $Y_1 + \cdots + a.X + \cdots + X + \cdots + Y_n \ \reduce\ Y_1 + \cdots + Y_n + (a+1).X$ \\
     & \textcolor{gray}{Similar to (R-ADD0).} \\
@@
     R-ADD3
-    & $a_1 + \cdots + a.X + \cdots + b.X + \cdots + a_n \ \reduce\ a_1 + \cdots + a_n + (a+b).X$ \\
+    & $Y_1 + \cdots + a.X + \cdots + b.X + \cdots + Y_n \ \reduce\ Y_1 + \cdots + Y_n + (a+b).X$ \\
     & \textcolor{gray}{Similar to (R-ADD0).} \\
@@
     %
+    R-ADJ3
+    & $ (X \otimes Y)^\dagger \ \reduce\ X^\dagger \otimes Y^\dagger$ \\
+    %
     R-ADJK0
@@
     %
-    R-ADJK3
-    & $ (B_1 \otimes B_2)^\dagger \ \reduce\ B_1^\dagger \otimes B_2^\dagger $ \\
-    %
     R-ADJB0
@@
     & $ (O \cdot K)^\dagger\ \reduce\ K^\dagger \cdot O^\dagger $ \\
-    %
-    R-ADJB3
-    & $ (K_1 \otimes K_2)^\dagger \ \reduce\ K_1^\dagger \otimes K_2^\dagger $ \\
     %
@@
     & $ (O_1 \cdot O_2)^\dagger \ \reduce\ O_2^\dagger \cdot O_1^\dagger $ \\
-    %
-    R-ADJO4
-    & $ (O_1 \otimes O_2)^\dagger \ \reduce\ O_1^\dagger \otimes O_2^\dagger $ 
 \end{ruletable}

### 2024-12-05 main.tex
sum rules implemented.

@@
 \newcommand{\OType}{\mathcal{O}}
+\newcommand{\SET}{\mathsf{Set}}
 
@@
 
+\section{TODO}
+\begin{itemize}
+    \item The Coq side should be able to decide the alpha-equivalence of terms.
+\end{itemize}
+
 \section{Language Syntax}
@@
     \begin{align*}
-        T ::=\ & x\ |\ \SType\ |\ \KType(\sigma)\ |\ \BType(\sigma)\ |\ \OType(\sigma_1, \sigma_2)\ |\ T_1 \to T_2, \\
+        T ::=\ & x\ |\ \SType\ |\ \KType(\sigma)\ |\ \BType(\sigma)\ |\ \OType(\sigma_1, \sigma_2)\ |\ T_1 \to T_2\ |\ \SET(\sigma), \\
         \sigma ::=\ & x\ |\ \sigma_1 \times \sigma_2.
@@
         & |\ \ket{e}\ |\ \bra{t}\ |\ e^\dagger\ |\ e_1.e_2\ |\ \ADD(e_1 \cdots e_n)\ |\ e_1 \otimes e_2 \\
-        & |\ \MULK(e_1\ e_2)\ |\ \MULB(e_1\ e_2)\ |\ \OUTER(e_1\ e_2)\ |\ \MULO(e_1\ e_2).
+        & |\ \MULK(e_1\ e_2)\ |\ \MULB(e_1\ e_2)\ |\ \OUTER(e_1\ e_2)\ |\ \MULO(e_1\ e_2) \\
+        & |\ \mathbf{U}(e)\ |\ e_1 \star e_2\ |\ \sum_{e_1} e_2.
     \end{align*}
@@
 They are denoted as $B\cdot K$, $K_1 \cdot K_2$, $B_1 \cdot B_2$, $K \cdot B$ and $O_1 \cdot O_2$, respectively.
+
+Usually, the sum body is specified by an abstraction. Therefore we use notation $\sum_{i \in s} X$ to denote $\sum_{s} \lambda i . X$.
 
@@
         \textbf{Type-Scalar} \qquad
-        \frac{\WF(\Gamma)}{\Gamma \vdash \SType : \Type}
+        \frac{\WF(\Gamma)}{\Gamma \vdash \SType : \Type} \\
+        \\
+        \textbf{Type-Set} \qquad
+        \frac{\Gamma \vdash \sigma : \BaseS}{\Gamma \vdash \SET(\sigma) : \Type}
     \end{gather*}
@@
     \caption{Operator typing rules.}
+\end{figure}
+
+\begin{figure}[h]
+    \begin{gather*}
+        \textbf{Set-U} \qquad
+        \frac{\Gamma \vdash \sigma : \BaseS}{\Gamma \vdash \mathbf{U}(\sigma) : \SET(\sigma)} 
+        \qquad \qquad
+        \textbf{Set-Prod} \qquad
+        \frac{\Gamma \vdash A : \SET(\sigma) \qquad \Gamma \vdash B : \SET(\tau)}{\Gamma \vdash A \star B : \SET(\sigma \times \tau)}
+    \end{gather*}
+    \caption{Set typing rules.}
+\end{figure}
+
+\begin{figure}[h]
+    \begin{gather*}
+        \textbf{Sum-Scalar} \qquad
+        \frac{\Gamma \vdash s : \SET(\sigma) \qquad \Gamma \vdash f : \sigma \to \SType}{\Gamma \vdash \sum_{s} f : \SType} 
+        \qquad \qquad
+        \textbf{Sum-Ket} \qquad
+        \frac{\Gamma \vdash s : \SET(\sigma) \qquad \Gamma \vdash f : \sigma \to \KType(\tau)}{\Gamma \vdash \sum_{s} f : \KType(\tau)} \\
+        \\
+        \textbf{Sum-Bra} \qquad
+        \frac{\Gamma \vdash s : \SET(\sigma) \qquad \Gamma \vdash f : \sigma \to \BType(\tau)}{\Gamma \vdash \sum_{s} f : \BType(\tau)} 
+        \qquad \qquad
+        \textbf{Sum-Opt} \qquad
+        \frac{\Gamma \vdash s : \SET(\sigma) \qquad \Gamma \vdash f : \sigma \to \OType(\tau, \rho)}{\Gamma \vdash \sum_{s} f : \OType(\tau, \rho)}
+    \end{gather*}
+    \caption{Sum typing rules.}
 \end{figure}
@@
 
-% About big operators: maybe we should use the "sum operators"?
+\begin{ruletable}{Rules for sets.}
+    %
+    R-SET0
+    & $ \mathbf{U}(\sigma) \star \mathbf{U}(\tau) \ \reduce\ \mathbf{U}(\sigma \times \tau) $
+\end{ruletable}
+
+\begin{ruletable}{Rules for sum operators.}
+    %
+    R-SUM-CONST0
+    & $ \sum_{x \in s} 0 \ \reduce\ 0 $ \\
+    %
+    R-SUM-CONST1
+    & $ \sum_{x \in s} \ZEROK(\sigma)\ \reduce\ \ZEROK(\sigma) $ \\
+    %
+    R-SUM-CONST2
+    & $ \sum_{x \in s} \ZEROB(\sigma)\ \reduce\ \ZEROB(\sigma) $ \\
+    %
+    R-SUM-CONST3
+    & $ \sum_{x \in s} \ZEROO(\sigma, \tau)\ \reduce\ \ZEROO(\sigma, \tau) $ \\
+    %
+    R-SUM-CONST4
+    & $ \ONEO(\sigma) \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)} \ket{i} \cdot \bra{i} $
+\end{ruletable}
+
+\begin{ruletable}{Rules for eliminating $\delta_{s, t}$.}
+    %
+    R-SUM-ELIM0
+    & $ i \text{ free in } t \vdash \sum_{i \in \mathbf{U}(\sigma)} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} \delta_{i, t} \ \reduce\ \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n}  1$ \\
+    %
+    R-SUM-ELIM1
+    & $ i \text{ free in } t \vdash \sum_{i \in \mathbf{U}(\sigma)} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (\delta_{i, t} \times a) \ \reduce\ \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} a\{i/t\} $ \\
+    %
+    R-SUM-ELIM2
+    & $ i \text{ free in } t \vdash \sum_{i \in \mathbf{U}(\sigma)} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (\delta_{i, t}.A) \ \reduce\ \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} A\{i/t\} $ \\
+    %
+    R-SUM-ELIM3
+    & $ i \text{ free in } t \vdash \sum_{i \in \mathbf{U}(\sigma)} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (\delta_{i, t} \times a).A \ \reduce\ \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n}  a\{i/t\}.A\{i/t\} $ \\
+    R-SUM-ELIM4
+    & $ \sum_{i \in M} \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n}  \delta_{i, j} \ \reduce\ \sum_{i \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} 1 $ \\
+    %
+    R-SUM-ELIM5
+    & $ \sum_{i \in M} \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (\delta_{i, j} \times a) \ \reduce\ \sum_{i \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} a\{i/j\} $ \\
+    %
+    R-SUM-ELIM6
+    & $ \sum_{i \in M} \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (\delta_{i, j}.A) \ \reduce\ \sum_{i \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} A\{i/j\} $ \\
+    %
+    R-SUM-ELIM7
+    & $ \sum_{i \in M} \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (\delta_{i, j} \times a).A \ \reduce\ \sum_{i \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n}  a\{i/j\}.A\{i/j\} $ \\
+\end{ruletable}
+
+\begin{ruletable}{Rules for pushing terms into sum operators}
+    %
+    R-SUM-PUSH0
+    & $ (\sum_{i \in M} a) \times b \ \reduce\ \sum_{i \in M} (a \times b) $ \\
+    %
+    R-SUM-PUSH1
+    & $ (\sum_{i \in M} a)^* \ \reduce\ \sum_{i \in M} a^* $ \\
+    %
+    R-SUM-PUSH2
+    & $ (\sum_{i \in M} X)^\dagger \ \reduce\ \sum_{i \in M} X^\dagger $ \\
+    %
+    R-SUM-PUSH3
+    & $ a.(\sum_{i \in M} X) \ \reduce\ \sum_{i \in M} (a.X) $ \\
+    %
+    R-SUM-PUSH4
+    & $ (\sum_{i \in M} a).X \ \reduce\ \sum_{i \in M} (a.X) $ \\
+    %
+    R-SUM-PUSH5
+    & $ (\sum_{i \in M} B)\cdot K \ \reduce\ \sum_{i \in M}(B \cdot K) $ \\
+    %
+    R-SUM-PUSH6
+    & $ (\sum_{i \in M} O)\cdot K \ \reduce\ \sum_{i \in M}(O \cdot K) $ \\
+    %
+    R-SUM-PUSH7
+    & $ (\sum_{i \in M} B)\cdot O \ \reduce\ \sum_{i \in M}(B \cdot O) $ \\
+    %
+    R-SUM-PUSH8
+    & $ (\sum_{i \in M} K)\cdot B \ \reduce\ \sum_{i \in M}(K \cdot B) $ \\
+    %
+    R-SUM-PUSH9
+    & $ (\sum_{i \in M} O_1)\cdot O_2 \ \reduce\ \sum_{i \in M}(O_1 \cdot O_2) $ \\
+    %
+    R-SUM-PUSH10
+    & $ B \cdot (\sum_{i \in M} K) \ \reduce\ \sum_{i \in M}(B \cdot K) $ \\
+    %
+    R-SUM-PUSH11
+    & $ O \cdot (\sum_{i \in M} K) \ \reduce\ \sum_{i \in M}(O \cdot K) $ \\
+    %
+    R-SUM-PUSH12
+    & $ B \cdot (\sum_{i \in M} O) \ \reduce\ \sum_{i \in M}(B \cdot O) $ \\
+    %
+    R-SUM-PUSH13
+    & $ K \cdot (\sum_{i \in M} B) \ \reduce\ \sum_{i \in M}(K \cdot B) $ \\
+    %
+    R-SUM-PUSH14
+    & $ O_1 \cdot (\sum_{i \in M} O_2) \ \reduce\ \sum_{i \in M}(O_1 \cdot O_2) $ \\
+    %
+    R-SUM-PUSH15
+    & $ (\sum_{i \in M} K_1) \otimes K_2 \ \reduce\ \sum_{i \in M} (K_1 \otimes K_2) $ \\
+    %
+    R-SUM-PUSH16
+    & $ (\sum_{i \in M} B_1) \otimes B_2 \ \reduce\ \sum_{i \in M} (B_1 \otimes B_2) $ \\
+    %
+    R-SUM-PUSH17
+    & $ (\sum_{i \in M} O_1) \otimes O_2 \ \reduce\ \sum_{i \in M} (O_1 \otimes O_2) $ \\
+    %
+    R-SUM-PUSH18
+    & $ K_1 \otimes (\sum_{i \in M} K_2) \ \reduce\ \sum_{i \in M} (K_1 \otimes K_2) $ \\
+    %
+    R-SUM-PUSH19
+    & $ B_1 \otimes (\sum_{i \in M} B_2) \ \reduce\ \sum_{i \in M} (B_1 \otimes B_2) $ \\
+    %
+    R-SUM-PUSH20
+    & $ O_1 \otimes (\sum_{i \in M} O_2) \ \reduce\ \sum_{i \in M} (O_1 \otimes O_2) $ \\
+\end{ruletable}
+\yx{Note: because we apply type checking on variables, and stick to unique bound variables, these pushing in operations are always sound.}
+
+\begin{ruletable}{Rules for addition and index in sum}
+    %
+    R-SUM-ADD0
+    & $\sum_{i \in M}(X_1 + \cdots + X_n) \ \reduce\ (\sum_{i \in M} X_1) + \cdots + (\sum_{i \in M} X_n) $ \\
+    %
+    R-SUM-ADD1
+    & $Y_1 + \cdots + \sum_{i \in M}(a.X) + \cdots + \sum_{i \in M}X + Y_n \ \reduce\ Y_1 + \cdots + Y_n + \sum_{i \in M}(a+1).X $ \\
+    %
+    R-SUM-ADD2
+    & $Y_1 + \cdots + \sum_{i \in M}X + \cdots + \sum_{i \in M}(a.X) + Y_n \ \reduce\ Y_1 + \cdots + Y_n + \sum_{i \in M}(1+a).X $ \\
+    %
+    R-SUM-ADD3
+    & $Y_1 + \cdots + \sum_{i \in M}(a.X) + \cdots + \sum_{i \in M}(b.X) + Y_n \ \reduce\ Y_1 + \cdots + Y_n + \sum_{i \in M}(a+b).X $ \\
+    %
+    R-SUM-INDEX0
+    & $ \sum_{i \in \mathbf{U}(\sigma \times \tau)} A \ \reduce\ \sum_{j \in \mathbf{U}(\sigma)} \sum_{k \in \mathbf{U}(\tau)} A\{i/(j, k)\} $ \\
+    %
+    R-SUM-INDEX1
+    & $ \sum_{i \in M_1 \star M_2} A \ \reduce\ \sum_{j \in M_1} \sum_{k \in M_2} A\{i/(j, k)\} $
+\end{ruletable}
 
@@
 
+\input{labelled_dirac.tex}
+
+\clearpage
+
 \bibliographystyle{plain}

### 2024-12-05 main.tex
sum rules fixed

@@
 \end{figure}
+\yx{The Lam rule is problematic. It still allows a ill-formed dependent type.}
 
@@
 
-\begin{ruletable}{Rules for eliminating $\delta_{s, t}$.}
+\begin{ruletable}{Rules for eliminating $\delta_{s, t}$. Note that these rules will match the $\delta$ operator modulo the commutativity of its arguments.}
     %
@@
     R-SUM-ELIM1
-    & $ i \text{ free in } t \vdash \sum_{i \in \mathbf{U}(\sigma)} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (\delta_{i, t} \times a) \ \reduce\ \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} a\{i/t\} $ \\
+    & $ i \text{ free in } t \vdash \sum_{i \in \mathbf{U}(\sigma)} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (a_1 \times \cdots \times \delta_{i, t} \times \cdots \times a_n) $ \\
+    & $ \reduce\ \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} a_1\{i/t\} \times \cdots \times a_n\{i/t\} $ \\
     %
@@
     R-SUM-ELIM3
-    & $ i \text{ free in } t \vdash \sum_{i \in \mathbf{U}(\sigma)} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (\delta_{i, t} \times a).A \ \reduce\ \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n}  a\{i/t\}.A\{i/t\} $ \\
+    & $ i \text{ free in } t \vdash \sum_{i \in \mathbf{U}(\sigma)} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (a_1 \times \cdots \times \delta_{i, t} \times \cdots \times a_n).A $ \\
+    & $ \reduce\ \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n}  (a_1\{i/t\} \times \cdots \times a_n\{i/t\}).A\{i/t\} $ \\
     R-SUM-ELIM4
-    & $ \sum_{i \in M} \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n}  \delta_{i, j} \ \reduce\ \sum_{i \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} 1 $ \\
+    & $ \sum_{i \in M} \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n}  \delta_{i, j} \ \reduce\ \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} 1 $ \\
     %
     R-SUM-ELIM5
-    & $ \sum_{i \in M} \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (\delta_{i, j} \times a) \ \reduce\ \sum_{i \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} a\{i/j\} $ \\
+    & $ \sum_{i \in M} \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (a_1 \times \dots \times \delta_{i, j} \times \cdots \times a_n) $ \\
+    & $ \reduce\ \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (a_1\{j/i\} \times \cdots \times a_n\{j/i\}) $ \\
     %
     R-SUM-ELIM6
-    & $ \sum_{i \in M} \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (\delta_{i, j}.A) \ \reduce\ \sum_{i \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} A\{i/j\} $ \\
+    & $ \sum_{i \in M} \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (\delta_{i, j}.A) \ \reduce\ \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} A\{j/i\} $ \\
     %
     R-SUM-ELIM7
-    & $ \sum_{i \in M} \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (\delta_{i, j} \times a).A \ \reduce\ \sum_{i \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n}  a\{i/j\}.A\{i/j\} $ \\
+    & $ \sum_{i \in M} \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (a_1 \times \cdots \times \delta_{i, j} \times \cdots \times a_n).A $ \\
+    & $ \reduce\ \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (a_1\{j/i\} \times \cdots \times a_n\{j/i\}).A\{j/i\} $ \\
 \end{ruletable}
@@
     R-SUM-PUSH0
-    & $ (\sum_{i \in M} a) \times b \ \reduce\ \sum_{i \in M} (a \times b) $ \\
+    & $ b_1 \times \cdots \times (\sum_{i \in M} a) \times \cdots \times b_n \ \reduce\ \sum_{i \in M} (b_1 \times \cdots \times a \times \cdots \times b_n) $ \\
     %
@@
     R-SUM-PUSH15
-    & $ (\sum_{i \in M} K_1) \otimes K_2 \ \reduce\ \sum_{i \in M} (K_1 \otimes K_2) $ \\
+    & $ (\sum_{i \in M} X_1) \otimes X_2 \ \reduce\ \sum_{i \in M} (X_1 \otimes X_2) $ \\
     %
     R-SUM-PUSH16
-    & $ (\sum_{i \in M} B_1) \otimes B_2 \ \reduce\ \sum_{i \in M} (B_1 \otimes B_2) $ \\
-    %
-    R-SUM-PUSH17
-    & $ (\sum_{i \in M} O_1) \otimes O_2 \ \reduce\ \sum_{i \in M} (O_1 \otimes O_2) $ \\
-    %
-    R-SUM-PUSH18
-    & $ K_1 \otimes (\sum_{i \in M} K_2) \ \reduce\ \sum_{i \in M} (K_1 \otimes K_2) $ \\
-    %
-    R-SUM-PUSH19
-    & $ B_1 \otimes (\sum_{i \in M} B_2) \ \reduce\ \sum_{i \in M} (B_1 \otimes B_2) $ \\
-    %
-    R-SUM-PUSH20
-    & $ O_1 \otimes (\sum_{i \in M} O_2) \ \reduce\ \sum_{i \in M} (O_1 \otimes O_2) $ \\
+    & $ X_1 \otimes (\sum_{i \in M} X_2) \ \reduce\ \sum_{i \in M} (X_1 \otimes X_2) $
 \end{ruletable}
@@
 \begin{ruletable}{Rules for addition and index in sum}
+    %
+    R-SUM-ADDS0
+    & $\sum_{i \in M}(a_1 + \cdots + a_n) \ \reduce\ (\sum_{i \in M} a_1) + \cdots + (\sum_{i \in M} a_n) $ \\
     %
@@
     R-SUM-INDEX1
-    & $ \sum_{i \in M_1 \star M_2} A \ \reduce\ \sum_{j \in M_1} \sum_{k \in M_2} A\{i/(j, k)\} $
+    & $ \sum_{i \in M_1 \star M_2} A \ \reduce\ \sum_{j \in M_1} \sum_{k \in M_2} A\{i/(j, k)\} $ \\
+    %
+    R-SUM-SWAP
+    & $ M_1 < M_2 \vdash \sum_{i \in M_2} \sum_{j \in M_1} X \ \reduce\ \sum_{j \in M_1} \sum_{i \in M_2} X $
 \end{ruletable}
+
+\yx{The rules R-SUM-ADD1 to R-SUM-ADD3 needs identical sum terms, which requires the rewriting to be after the alpha normalization. We don't implement them for now.}
 

### 2024-12-08 main.tex
typing system updated. now using type index

@@
 
+\newcommand{\Index}{\mathsf{Index}}
 \newcommand{\Type}{\mathsf{Type}}
-\newcommand{\BaseS}{\mathsf{Base}}
+\newcommand{\Basis}{\mathsf{Basis}}
 
@@
 \begin{itemize}
-    \item The Coq side should be able to decide the alpha-equivalence of terms.
+    \item Alpha equivalence is checked by a bound variable renaming on the whole term.
+    \item arrow type of $\Type \to \Type$ are forbidden. Check whether typing rules comply with this restriction.
 \end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \begin{definition}[syntax]
+    The syntax for type indices are defined as
+    \begin{align*}
+        \sigma ::=\ & x\ |\ \sigma_1 \times \sigma_2.
+    \end{align*}
     The syntax for Dirac notation types is defined as 
     \begin{align*}
-        T ::=\ & x\ |\ \SType\ |\ \KType(\sigma)\ |\ \BType(\sigma)\ |\ \OType(\sigma_1, \sigma_2)\ |\ T_1 \to T_2\ |\ \SET(\sigma), \\
-        \sigma ::=\ & x\ |\ \sigma_1 \times \sigma_2.
+        T ::=\ & x\ |\ \Basis(\sigma)\ |\ \SType\ |\ \KType(\sigma)\ |\ \BType(\sigma)\ |\ \OType(\sigma_1, \sigma_2)\ |\ T_1 \to T_2\ |\ \forall x.T\ |\ \SET(\sigma). \\
     \end{align*}
@@
 Notice that the type checking $\Gamma \vdash t : T$ is decidable.
+
+\begin{definition}[context]
+    \begin{align*}
+        \Gamma ::= []\ |\ \Gamma; x : \Index\ |\ \Gamma; T : \Type\ |\ \Gamma; x : T\ |\ \Gamma; x := t : T.
+    \end{align*}
+\end{definition}
+
 \begin{figure}[h]
@@
         \textbf{W-Empty} \qquad
-        \frac{}{\WF([])}
-        \qquad \qquad
+        \frac{}{\WF([])} \\
+        \\
+        \textbf{W-Assum-Index} \qquad
+        \frac{\WF(\Gamma) \qquad \sigma \notin \Gamma}{\WF(\Gamma; \sigma : \Index)}\qquad \qquad
         \textbf{W-Assum-Type} \qquad
-        \frac{\WF(\Gamma) \qquad T \notin \Gamma}{\WF(\Gamma; T:\Type)}\\
+        \frac{\WF(\Gamma) \qquad T \notin \Gamma}{\WF(\Gamma; T : \Type)} \\
         \\
         \textbf{W-Assum-Term} \qquad
-        \frac{\Gamma\vdash T:\Type \qquad x \notin \Gamma}{\WF(\Gamma; x:T)}
-        \qquad \qquad
-        \textbf{W-Def} \qquad
-        \frac{\Gamma\vdash t:T \qquad x \notin \Gamma}{\WF(\Gamma; x:=t:T)} \\
-        \\
-        \textbf{Var} \qquad
+        \frac{\Gamma \vdash T : \Type \qquad x \notin \Gamma}{\WF(\Gamma; x:T)}
+        \qquad \qquad
+        \textbf{W-Def-Term} \qquad
+        \frac{\Gamma\vdash t:T \qquad x \notin \Gamma}{\WF(\Gamma; x:=t:T)}
+    \end{gather*}
+    \caption{Rules for a well-formed context.}
+\end{figure}
+
+\begin{figure}
+    \begin{gather*}
+        \textbf{Index-Var} \qquad
+        \frac{\WF(\Gamma) \qquad \sigma : \Index \in \Gamma}{\Gamma \vdash \sigma : \Index}
+        \qquad \qquad
+        \textbf{Index-Prod} \qquad
+        \frac{\Gamma \vdash \sigma : \Index \qquad \Gamma \vdash \tau : \Index}{\Gamma \vdash \sigma \times \tau : \Index}
+    \end{gather*}
+    \caption{Rules for type index.}
+\end{figure}
+
+\begin{figure}
+    \begin{gather*}
+        \textbf{Type-Var} \qquad
+        \frac{\WF(\Gamma) \qquad T : \Type \in \Gamma}{\Gamma \vdash T : \Type}\\
+        \\
+        \textbf{Type-Arrow} \qquad
+        \frac{\Gamma \vdash T : \Type \qquad \Gamma \vdash U : \Type}{\Gamma \vdash T \to U : \Type}
+        \qquad \qquad
+        \textbf{Type-Index} \qquad
+        \frac{\Gamma ; \sigma : \Index \vdash U : \Type}{\Gamma \vdash \forall \sigma.U : \Type} \\
+        \\        
+        \textbf{Type-Basis} \qquad
+        \frac{\Gamma \vdash \sigma : \Index}{\Gamma \vdash \Basis(\sigma) : \Type}  \\
+        \\
+        \textbf{Type-Ket} \qquad
+        \frac{\Gamma \vdash \sigma : \Index}{\Gamma \vdash \KType(\sigma) : \Type}
+        \qquad \qquad
+        \textbf{Type-Bra} \qquad
+        \frac{\Gamma \vdash \sigma : \Index}{\Gamma \vdash \BType(\sigma) : \Type} \\
+        \\
+        \textbf{Type-Opt} \qquad
+        \frac{\Gamma \vdash \sigma : \Index \qquad \Gamma \vdash \tau : \Index}{\Gamma \vdash \OType(\sigma, \tau) : \Type}
+        \qquad \qquad
+        \textbf{Type-Scalar} \qquad
+        \frac{\WF(\Gamma)}{\Gamma \vdash \SType : \Type} \\
+        \\
+        \textbf{Type-Set} \qquad
+        \frac{\Gamma \vdash \sigma : \Index}{\Gamma \vdash \SET(\sigma) : \Type}
+    \end{gather*}
+    \caption{Rules for types.}
+\end{figure}
+
+\begin{figure}[h]
+    \begin{gather*}
+        \textbf{Term-Var} \qquad
         \frac{\WF(\Gamma) \qquad (x:T) \in \Gamma \text{ or } (x := t : T) \in \Gamma \text{ for some $t$}}{\Gamma \vdash x : T} \\
-        \\
-        \textbf{Arrow} \qquad
-        \frac{\Gamma \vdash T:\Type \qquad \Gamma \vdash U:\Type}{\Gamma \vdash T \to U : \Type} \\
         \\
@@
         \qquad \qquad
-        \textbf{App} \qquad
+        \textbf{Index} \qquad
+        \frac{\Gamma; \sigma : \Index \vdash t : U \qquad x \notin \Gamma}{\Gamma \vdash (\lambda x.t) : \forall \sigma.U} \\
+        \\
+        \textbf{App-Arrow} \qquad
         \frac{\Gamma \vdash t:U \to T \qquad \Gamma \vdash u:U}{\Gamma \vdash (t\ u):T}
-    \end{gather*}
-    \caption{The typing rules for the typed lambda calculus with environment.}
-\end{figure}
-\yx{The Lam rule is problematic. It still allows a ill-formed dependent type.}
-
-\begin{figure}[h]
-    \begin{gather*}
-        \textbf{Type-Base} \qquad
-        \frac{\WF(\Gamma)}{\Gamma \vdash \BaseS : \Type} 
-        \qquad \qquad
-        \textbf{Base-Subtype} \qquad
-        \frac{\Gamma \vdash T : \BaseS}{\Gamma \vdash T : \Type} \\
-        \\
-        \textbf{Type-Ket} \qquad
-        \frac{\Gamma \vdash \sigma : \BaseS}{\Gamma \vdash \KType(\sigma) : \Type}
-        \qquad \qquad
-        \textbf{Type-Bra} \qquad
-        \frac{\Gamma \vdash \sigma : \BaseS}{\Gamma \vdash \BType(\sigma) : \Type} \\
-        \\
-        \textbf{Type-Opt} \qquad
-        \frac{\Gamma \vdash \sigma : \BaseS \qquad \Gamma \vdash \tau : \BaseS}{\Gamma \vdash \OType(\sigma, \tau) : \Type}
-        \qquad \qquad
-        \textbf{Type-Scalar} \qquad
-        \frac{\WF(\Gamma)}{\Gamma \vdash \SType : \Type} \\
-        \\
-        \textbf{Type-Set} \qquad
-        \frac{\Gamma \vdash \sigma : \BaseS}{\Gamma \vdash \SET(\sigma) : \Type}
-    \end{gather*}
-    \caption{Typing rules for Dirac notation types.}
+        \qquad \qquad
+        \textbf{App-Index} \qquad
+        \frac{\Gamma \vdash t:\forall x.U \qquad \Gamma \vdash u : \Index}{\Gamma \vdash (t\ u):U\{x/u\}}
+    \end{gather*}
+    \caption{Rules for variable and function typings.}
 \end{figure}
@@
         \textbf{Sca-Delta} \qquad
-        \frac{            
-            \begin{aligned}
-                & \Gamma\vdash \sigma : \BaseS \\
-                & \Gamma\vdash s : \sigma \qquad \Gamma\vdash t : \sigma
-            \end{aligned}} {\Gamma \vdash \delta_{s, t} : \SType} \\
+        \frac{ \Gamma\vdash s : \Basis(\sigma) \qquad \Gamma\vdash t : \Basis(\sigma) } {\Gamma \vdash \delta_{s, t} : \SType} \\
         \\
@@
         % \frac{\WF(\Gamma) \qquad a_i \notin \Gamma \text{ for all $i$}}
-        % {\Gamma\vdash \{a_1, a_2, \cdots, a_n\} : \BaseS}
+        % {\Gamma\vdash \{a_1, a_2, \cdots, a_n\} : \Basis}
         % \qquad \qquad
@@
         % \\
-        \textbf{Base-Prod} \qquad
-        \frac{\Gamma\vdash \sigma : \BaseS \qquad \Gamma\vdash \tau : \BaseS}{\Gamma\vdash \sigma \times \tau : \BaseS}
-        \qquad \qquad
         \textbf{Pair-Base} \qquad
         \frac{
-            \begin{aligned}
-                & \Gamma\vdash \sigma : \BaseS \qquad \Gamma\vdash s : \sigma \\
-                & \Gamma\vdash \tau : \BaseS \qquad \Gamma\vdash t : \tau
-            \end{aligned}} {\Gamma\vdash (s, t) : \sigma \times \tau} 
+            \Gamma\vdash s : \Basis(\sigma) \qquad \Gamma\vdash t : \Basis(\tau)} {\Gamma\vdash (s, t) : \Basis(\sigma \times \tau)} 
         \textbf{}
     \end{gather*}
-    \caption{Typing rules for bases. }
+    \caption{Typing rules for Basis. }
 \end{figure}
@@
         \textbf{Ket-0} \qquad
-        \frac{\Gamma \vdash \sigma : \BaseS}{\Gamma \vdash \ZEROK(\sigma) : \KType(\sigma)} 
+        \frac{\Gamma \vdash \sigma : \Index}{\Gamma \vdash \ZEROK(\sigma) : \KType(\sigma)} 
         \qquad \qquad
         \textbf{Ket-Base} \qquad
-        \frac{\Gamma \vdash \sigma : \BaseS \qquad \Gamma\vdash t : \sigma}{\Gamma \vdash \ket{t} : \KType(\sigma)} \\
+        \frac{\Gamma\vdash t : \Basis(\sigma)}{\Gamma \vdash \ket{t} : \KType(\sigma)} \\
         \\
@@
         \textbf{Bra-0} \qquad
-        \frac{\Gamma \vdash \sigma : \BaseS}{\Gamma \vdash \ZEROB(\sigma) : \BType(\sigma)} 
+        \frac{\Gamma \vdash \sigma : \Index}{\Gamma \vdash \ZEROB(\sigma) : \BType(\sigma)} 
         \qquad \qquad
         \textbf{Bra-Base} \qquad
-        \frac{\Gamma \vdash \sigma : \BaseS \qquad \Gamma\vdash t : \sigma}{\Gamma \vdash \bra{t} : \BType(\sigma)} \\
+        \frac{\Gamma\vdash t : \Basis(\sigma)}{\Gamma \vdash \bra{t} : \BType(\sigma)} \\
         \\
@@
         \textbf{Opt-0} \qquad
-        \frac{\Gamma \vdash \sigma : \BaseS \qquad \Gamma \vdash \tau : \BaseS}{\Gamma \vdash \ZEROO(\sigma, \tau) : \OType(\sigma, \tau)} 
+        \frac{\Gamma \vdash \sigma : \Index \qquad \Gamma \vdash \tau : \Index}{\Gamma \vdash \ZEROO(\sigma, \tau) : \OType(\sigma, \tau)} 
         \qquad \qquad
         \textbf{Opt-1} \qquad
-        \frac{\Gamma \vdash \sigma : \BaseS}{\Gamma \vdash \ONEO(\sigma) : \OType(\sigma, \sigma)} \\
+        \frac{\Gamma \vdash \sigma : \Index}{\Gamma \vdash \ONEO(\sigma) : \OType(\sigma, \sigma)} \\
         \\
@@
         \textbf{Set-U} \qquad
-        \frac{\Gamma \vdash \sigma : \BaseS}{\Gamma \vdash \mathbf{U}(\sigma) : \SET(\sigma)} 
+        \frac{\Gamma \vdash \sigma : \Index}{\Gamma \vdash \mathbf{U}(\sigma) : \SET(\sigma)} 
         \qquad \qquad
@@
         \textbf{Sum-Scalar} \qquad
-        \frac{\Gamma \vdash s : \SET(\sigma) \qquad \Gamma \vdash f : \sigma \to \SType}{\Gamma \vdash \sum_{s} f : \SType} 
-        \qquad \qquad
+        \frac{\Gamma \vdash s : \SET(\sigma) \qquad \Gamma \vdash f : \Basis(\sigma) \to \SType}{\Gamma \vdash \sum_{s} f : \SType} \\
+        \\
         \textbf{Sum-Ket} \qquad
-        \frac{\Gamma \vdash s : \SET(\sigma) \qquad \Gamma \vdash f : \sigma \to \KType(\tau)}{\Gamma \vdash \sum_{s} f : \KType(\tau)} \\
+        \frac{\Gamma \vdash s : \SET(\sigma) \qquad \Gamma \vdash f : \Basis(\sigma) \to \KType(\tau)}{\Gamma \vdash \sum_{s} f : \KType(\tau)} \\
         \\
         \textbf{Sum-Bra} \qquad
-        \frac{\Gamma \vdash s : \SET(\sigma) \qquad \Gamma \vdash f : \sigma \to \BType(\tau)}{\Gamma \vdash \sum_{s} f : \BType(\tau)} 
-        \qquad \qquad
+        \frac{\Gamma \vdash s : \SET(\sigma) \qquad \Gamma \vdash f : \Basis(\sigma) \to \BType(\tau)}{\Gamma \vdash \sum_{s} f : \BType(\tau)} \\
+        \\
         \textbf{Sum-Opt} \qquad
-        \frac{\Gamma \vdash s : \SET(\sigma) \qquad \Gamma \vdash f : \sigma \to \OType(\tau, \rho)}{\Gamma \vdash \sum_{s} f : \OType(\tau, \rho)}
+        \frac{\Gamma \vdash s : \SET(\sigma) \qquad \Gamma \vdash f : \Basis(\sigma) \to \OType(\tau, \rho)}{\Gamma \vdash \sum_{s} f : \OType(\tau, \rho)}
     \end{gather*}
@@
 \end{figure}
-
-\yx{Alpha rule will be implemented by renaming, and we can leave it for the future.}
 

### 2024-12-10 main.tex
minor adjustments
changed to de Bruijn index design

@@
     \begin{align*}
-        T ::=\ & x\ |\ \Basis(\sigma)\ |\ \SType\ |\ \KType(\sigma)\ |\ \BType(\sigma)\ |\ \OType(\sigma_1, \sigma_2)\ |\ T_1 \to T_2\ |\ \forall x.T\ |\ \SET(\sigma). \\
+        T ::=\ & x\ |\ \Basis(\sigma)\ |\ \SType\ |\ \KType(\sigma)\ |\ \BType(\sigma)\ |\ \OType(\sigma_1, \sigma_2)\ |\ T_1 \to T_2\ |\ \forall .T\ |\ \SET(\sigma). \\
     \end{align*}
@@
     \begin{align*}
-        e ::=\ & x\ |\ (e_1, e_2) \\
+        e ::=\ & x\ |\ \$i\ |\ (e_1, e_2)\ |\ \lambda : T.e\ |\ \lambda.e\ |\ e_1\ e_2 \\
         & |\ 0\ |\ 1\ |\ \ADDS(e_1 \cdots e_n) \ |\ e_1 \times \cdots \times e_n\ |\ e^* \ |\ \delta_{e_1, e_2}\ |\ \DOT(e_1\ e_2) \\
@@
  % after consideration I still think keep DOT as binary is the best choice. It reduces the unstability of complicated matching in sequences.
-
+Here $i$ is a natural number and $\$i$ represents the $i$-th bound variable in de Bruijn notation. 
 Compared to [?], this syntax for Dirac notations merges the symbols with overlapped properties, such as the addition and scaling symbols for ket, bra and operator.
@@
 
-Usually, the sum body is specified by an abstraction. Therefore we use notation $\sum_{i \in s} X$ to denote $\sum_{s} \lambda i . X$.
+Usually, the sum body is specified by an abstraction. Therefore we use notation $\sum_{i \in s} X$ to denote $\sum_{s} \lambda i : T . X$.
 
 \section{Typing System}
-The type checking of Dirac notations involves maintaining a well-formed context $\Gamma$, which specifies the definitions and typing assumtpions for variables.
-Notice that the type checking $\Gamma \vdash t : T$ is decidable.
-
-\begin{definition}[context]
+The type checking of Dirac notations involves maintaining a well-formed environment and context $E[\Gamma]$, which specifies the definitions and typing assumtpions for variables.
+The environment and the conetxt are defined as follows.
+
+\begin{definition}[environment and context]
     \begin{align*}
-        \Gamma ::= []\ |\ \Gamma; x : \Index\ |\ \Gamma; T : \Type\ |\ \Gamma; x : T\ |\ \Gamma; x := t : T.
+        E ::=\ & []\ |\ E; x : \Index\ |\ E; T : \Type\ |\ E; x : T\ |\ E; x := t : T. \\
+        \Gamma ::=\ & []\ |\ \Gamma; \Index\ |\ \Gamma; T.
     \end{align*}
@@
 
+Note that in the following rules, $x : \Index$ and $x : \Type$ are not considered as $x : T$ judgements.
+The context $\Gamma$ acts as the stack to indicate the typing of de Bruijn notations.
+The notation $\Gamma_i$ is defined as the $i$-th element in $\Gamma$ from the end.
+That is, if $\Gamma \equiv \Gamma'; T$, then $\Gamma_0 = T$ and $\Gamma_i = \Gamma'_{i-1}$.
+
 \begin{figure}[h]
@@
         \textbf{W-Empty} \qquad
-        \frac{}{\WF([])} \\
-        \\
-        \textbf{W-Assum-Index} \qquad
-        \frac{\WF(\Gamma) \qquad \sigma \notin \Gamma}{\WF(\Gamma; \sigma : \Index)}\qquad \qquad
-        \textbf{W-Assum-Type} \qquad
-        \frac{\WF(\Gamma) \qquad T \notin \Gamma}{\WF(\Gamma; T : \Type)} \\
-        \\
-        \textbf{W-Assum-Term} \qquad
-        \frac{\Gamma \vdash T : \Type \qquad x \notin \Gamma}{\WF(\Gamma; x:T)}
+        \frac{}{\WF([])[]} \\
+        \\
+        \textbf{W-AssumE-Index} \qquad
+        \frac{\WF(E)[] \qquad x \notin E}{\WF(E; x : \Index)[]}\qquad \qquad
+        \textbf{W-AssumE-Type} \qquad
+        \frac{\WF(E)[] \qquad x \notin E}{\WF(E; x : \Type)} \\
+        \\
+        \textbf{W-AssumE-Term} \qquad
+        \frac{E[] \vdash T : \Type \qquad x \notin E}{\WF(E; x:T)[]}
         \qquad \qquad
         \textbf{W-Def-Term} \qquad
-        \frac{\Gamma\vdash t:T \qquad x \notin \Gamma}{\WF(\Gamma; x:=t:T)}
-    \end{gather*}
-    \caption{Rules for a well-formed context.}
+        \frac{E[]\vdash t:T \qquad x \notin E}{\WF(E; x:=t:T)[]} \\
+        \\
+        \textbf{W-AssumC-Index} \qquad
+        \frac{\WF(E)[\Gamma]}{\WF(E)[\Gamma; \Index]}
+        \qquad \qquad
+        \textbf{W-AssumC-Term} \qquad
+        \frac{E[\Gamma] \vdash T : \Type}{\WF(E)[\Gamma; T]}
+    \end{gather*}
+    \caption{Rules for a well-formed environment and context.}
 \end{figure}
@@
         \textbf{Index-Var} \qquad
-        \frac{\WF(\Gamma) \qquad \sigma : \Index \in \Gamma}{\Gamma \vdash \sigma : \Index}
+        \frac{\WF(E)[\Gamma] \qquad x : \Index \in E}{E[\Gamma] \vdash x : \Index}
         \qquad \qquad
         \textbf{Index-Prod} \qquad
-        \frac{\Gamma \vdash \sigma : \Index \qquad \Gamma \vdash \tau : \Index}{\Gamma \vdash \sigma \times \tau : \Index}
+        \frac{E[\Gamma] \vdash \sigma : \Index \qquad E[\Gamma] \vdash \tau : \Index}{E[\Gamma] \vdash \sigma \times \tau : \Index}
     \end{gather*}
@@
     \begin{gather*}
-        \textbf{Type-Var} \qquad
-        \frac{\WF(\Gamma) \qquad T : \Type \in \Gamma}{\Gamma \vdash T : \Type}\\
+        \textbf{Type-Context-Var} \qquad
+        \frac{\WF(E)[\Gamma]}{E[\Gamma]\vdash \$i : \Gamma_i}
+        \\
         \\
         \textbf{Type-Arrow} \qquad
-        \frac{\Gamma \vdash T : \Type \qquad \Gamma \vdash U : \Type}{\Gamma \vdash T \to U : \Type}
+        \frac{E[\Gamma] \vdash T : \Type \qquad E[\Gamma] \vdash U : \Type}{E[\Gamma] \vdash T \to U : \Type}
         \qquad \qquad
         \textbf{Type-Index} \qquad
-        \frac{\Gamma ; \sigma : \Index \vdash U : \Type}{\Gamma \vdash \forall \sigma.U : \Type} \\
+        \frac{E[\Gamma ; \Index] \vdash U : \Type}{E[\Gamma] \vdash \forall.U : \Type} \\
         \\        
         \textbf{Type-Basis} \qquad
-        \frac{\Gamma \vdash \sigma : \Index}{\Gamma \vdash \Basis(\sigma) : \Type}  \\
+        \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \Basis(\sigma) : \Type}  \\
         \\
         \textbf{Type-Ket} \qquad
-        \frac{\Gamma \vdash \sigma : \Index}{\Gamma \vdash \KType(\sigma) : \Type}
+        \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \KType(\sigma) : \Type}
         \qquad \qquad
         \textbf{Type-Bra} \qquad
-        \frac{\Gamma \vdash \sigma : \Index}{\Gamma \vdash \BType(\sigma) : \Type} \\
+        \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \BType(\sigma) : \Type} \\
         \\
         \textbf{Type-Opt} \qquad
-        \frac{\Gamma \vdash \sigma : \Index \qquad \Gamma \vdash \tau : \Index}{\Gamma \vdash \OType(\sigma, \tau) : \Type}
+        \frac{E[\Gamma] \vdash \sigma : \Index \qquad E[\Gamma] \vdash \tau : \Index}{E[\Gamma] \vdash \OType(\sigma, \tau) : \Type}
         \qquad \qquad
         \textbf{Type-Scalar} \qquad
-        \frac{\WF(\Gamma)}{\Gamma \vdash \SType : \Type} \\
+        \frac{\WF(E)[\Gamma]}{E[\Gamma] \vdash \SType : \Type} \\
         \\
         \textbf{Type-Set} \qquad
-        \frac{\Gamma \vdash \sigma : \Index}{\Gamma \vdash \SET(\sigma) : \Type}
+        \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \SET(\sigma) : \Type}
     \end{gather*}
@@
         \textbf{Term-Var} \qquad
-        \frac{\WF(\Gamma) \qquad (x:T) \in \Gamma \text{ or } (x := t : T) \in \Gamma \text{ for some $t$}}{\Gamma \vdash x : T} \\
+        \frac{\WF(E)[\Gamma] \qquad (x:T) \in E \text{ or } (x := t : T) \in E \text{ for some $t$}}{E[\Gamma] \vdash x : T} \\
         \\
         \textbf{Lam} \qquad
-        \frac{\Gamma;x:T \vdash t : U}{\Gamma \vdash (\lambda x:T . t) : T \to U}
+        \frac{E[\Gamma; T] \vdash t : U}{E[\Gamma] \vdash (\lambda : T . t) : T \to U}
         \qquad \qquad
         \textbf{Index} \qquad
-        \frac{\Gamma; \sigma : \Index \vdash t : U \qquad x \notin \Gamma}{\Gamma \vdash (\lambda x.t) : \forall \sigma.U} \\
+        \frac{E[\Gamma; \Index] \vdash t : U}{E[\Gamma] \vdash (\lambda.t) : \forall .U} \\
         \\
         \textbf{App-Arrow} \qquad
-        \frac{\Gamma \vdash t:U \to T \qquad \Gamma \vdash u:U}{\Gamma \vdash (t\ u):T}
+        \frac{E[\Gamma] \vdash t:U \to T \qquad E[\Gamma] \vdash u:U}{E[\Gamma] \vdash (t\ u):T}
         \qquad \qquad
         \textbf{App-Index} \qquad
-        \frac{\Gamma \vdash t:\forall x.U \qquad \Gamma \vdash u : \Index}{\Gamma \vdash (t\ u):U\{x/u\}}
-    \end{gather*}
-    \caption{Rules for variable and function typings.}
-\end{figure}
-
-\begin{figure}[h]
-    \begin{gather*}
-        \textbf{Sca-0} \qquad
-        \frac{\WF(\Gamma)}{\Gamma \vdash 0 : \SType}
-        \qquad \qquad
-        \textbf{Sca-1} \qquad
-        \frac{\WF(\Gamma)}{\Gamma \vdash 1 : \SType}
-        \qquad \qquad
-        \textbf{Sca-Delta} \qquad
-        \frac{ \Gamma\vdash s : \Basis(\sigma) \qquad \Gamma\vdash t : \Basis(\sigma) } {\Gamma \vdash \delta_{s, t} : \SType} \\
-        \\
-        \textbf{Sca-Add} \qquad
-        \frac{\Gamma\vdash a_i : \SType \text{ for all $i$}}{\Gamma\vdash a_1 + \cdots + a_n : \SType}
-        \qquad \qquad
-        \textbf{Sca-Mul} \qquad
-        \frac{\Gamma\vdash a_i : \SType \text{ for all $i$}}{\Gamma\vdash a_1 \times \cdots \times a_n : \SType} \\
-        \\
-        \textbf{Sca-Conj} \qquad
-        \frac{\Gamma \vdash a : \SType}{\Gamma \vdash a^*:\SType}
-        \qquad \qquad
-        \textbf{Sca-Dot} \qquad
-        \frac{\Gamma\vdash B : \BType(\sigma) \qquad \Gamma\vdash K : \KType(\sigma)}{\Gamma \vdash B \cdot K : \SType}
-    \end{gather*}
-    \caption{Scalar typing rules.}
+        \frac{E[\Gamma] \vdash t:\forall .U \qquad E[\Gamma] \vdash u : \Index}{E[\Gamma] \vdash (t\ u):U[u]}
+    \end{gather*}
+    \caption{Rules for variable and function typings. Here $U[u]$ means instantiate the first bound variable with $u$ in $U$.}
 \end{figure}
@@
         \frac{
-            \Gamma\vdash s : \Basis(\sigma) \qquad \Gamma\vdash t : \Basis(\tau)} {\Gamma\vdash (s, t) : \Basis(\sigma \times \tau)} 
+            E[\Gamma]\vdash s : \Basis(\sigma) \qquad E[\Gamma]\vdash t : \Basis(\tau)} {E[\Gamma]\vdash (s, t) : \Basis(\sigma \times \tau)} 
         \textbf{}
@@
     \begin{gather*}
+        \textbf{Sca-0} \qquad
+        \frac{\WF(E)[\Gamma]}{E[\Gamma] \vdash 0 : \SType}
+        \qquad \qquad
+        \textbf{Sca-1} \qquad
+        \frac{\WF(E)[\Gamma]}{E[\Gamma] \vdash 1 : \SType} \\
+        \\
+        \textbf{Sca-Delta} \qquad
+        \frac{ E[\Gamma]\vdash s : \Basis(\sigma) \qquad E[\Gamma]\vdash t : \Basis(\sigma) } {E[\Gamma] \vdash \delta_{s, t} : \SType} \\
+        \\
+        \textbf{Sca-Add} \qquad
+        \frac{E[\Gamma]\vdash a_i : \SType \text{ for all $i$}}{E[\Gamma]\vdash a_1 + \cdots + a_n : \SType}
+        \qquad \qquad
+        \textbf{Sca-Mul} \qquad
+        \frac{E[\Gamma]\vdash a_i : \SType \text{ for all $i$}}{E[\Gamma]\vdash a_1 \times \cdots \times a_n : \SType} \\
+        \\
+        \textbf{Sca-Conj} \qquad
+        \frac{E[\Gamma] \vdash a : \SType}{E[\Gamma] \vdash a^*:\SType}
+        \qquad \qquad
+        \textbf{Sca-Dot} \qquad
+        \frac{E[\Gamma]\vdash B : \BType(\sigma) \qquad E[\Gamma]\vdash K : \KType(\sigma)}{E[\Gamma] \vdash B \cdot K : \SType}
+    \end{gather*}
+    \caption{Scalar typing rules.}
+\end{figure}
+
+\begin{figure}[h]
+    \begin{gather*}
         \textbf{Ket-0} \qquad
-        \frac{\Gamma \vdash \sigma : \Index}{\Gamma \vdash \ZEROK(\sigma) : \KType(\sigma)} 
+        \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \ZEROK(\sigma) : \KType(\sigma)} 
         \qquad \qquad
         \textbf{Ket-Base} \qquad
-        \frac{\Gamma\vdash t : \Basis(\sigma)}{\Gamma \vdash \ket{t} : \KType(\sigma)} \\
+        \frac{E[\Gamma]\vdash t : \Basis(\sigma)}{E[\Gamma] \vdash \ket{t} : \KType(\sigma)} \\
         \\
         \textbf{Ket-Adj} \qquad
-        \frac{\Gamma \vdash B : \BType(\sigma)}{\Gamma \vdash B^\dagger : \KType(\sigma)} 
+        \frac{E[\Gamma] \vdash B : \BType(\sigma)}{E[\Gamma] \vdash B^\dagger : \KType(\sigma)} 
         \qquad \qquad
         \textbf{Ket-Scr} \qquad
-        \frac{\Gamma \vdash a : \SType \qquad \Gamma \vdash K : \KType(\sigma)}{\Gamma \vdash a.K : \KType(\sigma)} \\
+        \frac{E[\Gamma] \vdash a : \SType \qquad E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash a.K : \KType(\sigma)} \\
         \\
         \textbf{Ket-Add} \qquad
-        \frac{\Gamma \vdash K_i : \KType(\sigma) \text{ for all $i$}}{\Gamma \vdash K_1 + \cdots + K_n : \KType(\sigma)}
+        \frac{E[\Gamma] \vdash K_i : \KType(\sigma) \text{ for all $i$}}{E[\Gamma] \vdash K_1 + \cdots + K_n : \KType(\sigma)}
         \qquad \qquad
         \textbf{Ket-MulK} \qquad
-        \frac{\Gamma \vdash O : \OType(\sigma, \tau) \qquad \Gamma \vdash K : \KType(\tau)}{\Gamma \vdash O \cdot K : \KType(\sigma)} \\
+        \frac{E[\Gamma] \vdash O : \OType(\sigma, \tau) \qquad E[\Gamma] \vdash K : \KType(\tau)}{E[\Gamma] \vdash O \cdot K : \KType(\sigma)} \\
         \\
         \textbf{Ket-Tsr} \qquad
-        \frac{\Gamma \vdash K_1 : \KType(\sigma) \qquad \Gamma \vdash K_2 : \KType(\tau)} {\Gamma \vdash K_1 \otimes K_2 : \KType(\sigma \times \tau)}
+        \frac{E[\Gamma] \vdash K_1 : \KType(\sigma) \qquad E[\Gamma] \vdash K_2 : \KType(\tau)} {E[\Gamma] \vdash K_1 \otimes K_2 : \KType(\sigma \times \tau)}
     \end{gather*}
@@
         \textbf{Bra-0} \qquad
-        \frac{\Gamma \vdash \sigma : \Index}{\Gamma \vdash \ZEROB(\sigma) : \BType(\sigma)} 
+        \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \ZEROB(\sigma) : \BType(\sigma)} 
         \qquad \qquad
         \textbf{Bra-Base} \qquad
-        \frac{\Gamma\vdash t : \Basis(\sigma)}{\Gamma \vdash \bra{t} : \BType(\sigma)} \\
+        \frac{E[\Gamma]\vdash t : \Basis(\sigma)}{E[\Gamma] \vdash \bra{t} : \BType(\sigma)} \\
         \\
         \textbf{Bra-Adj} \qquad
-        \frac{\Gamma \vdash K : \KType(\sigma)}{\Gamma \vdash K^\dagger : \BType(\sigma)} 
+        \frac{E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash K^\dagger : \BType(\sigma)} 
         \qquad \qquad
         \textbf{Bra-Scr} \qquad
-        \frac{\Gamma \vdash a : \SType \qquad \Gamma \vdash B : \BType(\sigma)}{\Gamma \vdash a.B : \BType(\sigma)} \\
+        \frac{E[\Gamma] \vdash a : \SType \qquad E[\Gamma] \vdash B : \BType(\sigma)}{E[\Gamma] \vdash a.B : \BType(\sigma)} \\
         \\
         \textbf{Bra-Add} \qquad
-        \frac{\Gamma \vdash B_i : \BType(\sigma) \text{ for all $i$}}{\Gamma \vdash B_1 + \cdots + B_n : \BType(\sigma)}
+        \frac{E[\Gamma] \vdash B_i : \BType(\sigma) \text{ for all $i$}}{E[\Gamma] \vdash B_1 + \cdots + B_n : \BType(\sigma)}
         \qquad \qquad
         \textbf{Bra-MulB} \qquad
-        \frac{\Gamma \vdash B : \KType(\sigma) \qquad \Gamma \vdash O : \OType(\sigma, \tau)}{\Gamma \vdash B \cdot O : \BType(\tau)} \\
+        \frac{E[\Gamma] \vdash B : \KType(\sigma) \qquad E[\Gamma] \vdash O : \OType(\sigma, \tau)}{E[\Gamma] \vdash B \cdot O : \BType(\tau)} \\
         \\
         \textbf{Bra-Tsr} \qquad
-        \frac{\Gamma \vdash B_1 : \BType(\sigma) \qquad \Gamma \vdash B_2 : \BType(\tau)} {\Gamma \vdash B_1 \otimes B_2 : \BType(\sigma \times \tau)}
+        \frac{E[\Gamma] \vdash B_1 : \BType(\sigma) \qquad E[\Gamma] \vdash B_2 : \BType(\tau)} {E[\Gamma] \vdash B_1 \otimes B_2 : \BType(\sigma \times \tau)}
     \end{gather*}
@@
         \textbf{Opt-0} \qquad
-        \frac{\Gamma \vdash \sigma : \Index \qquad \Gamma \vdash \tau : \Index}{\Gamma \vdash \ZEROO(\sigma, \tau) : \OType(\sigma, \tau)} 
+        \frac{E[\Gamma] \vdash \sigma : \Index \qquad E[\Gamma] \vdash \tau : \Index}{E[\Gamma] \vdash \ZEROO(\sigma, \tau) : \OType(\sigma, \tau)} 
         \qquad \qquad
         \textbf{Opt-1} \qquad
-        \frac{\Gamma \vdash \sigma : \Index}{\Gamma \vdash \ONEO(\sigma) : \OType(\sigma, \sigma)} \\
+        \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \ONEO(\sigma) : \OType(\sigma, \sigma)} \\
         \\
         \textbf{Opt-Adj} \qquad
-        \frac{\Gamma \vdash O : \OType(\sigma, \tau)}{\Gamma \vdash O^\dagger : \OType(\tau, \sigma)} 
+        \frac{E[\Gamma] \vdash O : \OType(\sigma, \tau)}{E[\Gamma] \vdash O^\dagger : \OType(\tau, \sigma)} 
         \qquad \qquad
         \textbf{Opt-Scr} \qquad
-        \frac{\Gamma \vdash a : \SType \qquad \Gamma \vdash O : \OType(\sigma, \tau)}{\Gamma \vdash a.O : \OType(\sigma, \tau)} \\
+        \frac{E[\Gamma] \vdash a : \SType \qquad E[\Gamma] \vdash O : \OType(\sigma, \tau)}{E[\Gamma] \vdash a.O : \OType(\sigma, \tau)} \\
         \\
         \textbf{Opt-Add} \qquad
-        \frac{\Gamma \vdash O_i : \OType(\sigma, \tau) \text{ for all $i$}}{\Gamma \vdash O_1 + \cdots + O_n : \OType(\sigma, \tau)}
+        \frac{E[\Gamma] \vdash O_i : \OType(\sigma, \tau) \text{ for all $i$}}{E[\Gamma] \vdash O_1 + \cdots + O_n : \OType(\sigma, \tau)}
         \qquad \qquad
         \textbf{Opt-Outer} \qquad
-        \frac{\Gamma\vdash K : \KType(\sigma) \qquad \Gamma\vdash B : \BType(\tau)}{\Gamma\vdash K \cdot B : \OType(\sigma, \tau)} \\
+        \frac{E[\Gamma]\vdash K : \KType(\sigma) \qquad E[\Gamma]\vdash B : \BType(\tau)}{E[\Gamma]\vdash K \cdot B : \OType(\sigma, \tau)} \\
         \\
         \textbf{Opt-MulO} \qquad
-        \frac{\Gamma \vdash O_1 : \OType(\sigma, \tau) \qquad \Gamma \vdash O_2 : \OType(\tau, \rho)}{\Gamma \vdash O_1 \cdot O_2 : \OType(\sigma, \rho)} \\
+        \frac{E[\Gamma] \vdash O_1 : \OType(\sigma, \tau) \qquad E[\Gamma] \vdash O_2 : \OType(\tau, \rho)}{E[\Gamma] \vdash O_1 \cdot O_2 : \OType(\sigma, \rho)} \\
         \\
         \textbf{Opt-Tsr} \qquad
-        \frac{\Gamma \vdash O_1 : \OType(\sigma_1, \tau_1) \qquad \Gamma \vdash O_2 : \OType(\sigma_2, \tau_2)} {\Gamma \vdash O_1 \otimes O_2 : \OType(\sigma_1 \times \sigma_2, \tau_1 \times \tau_2)}
+        \frac{E[\Gamma] \vdash O_1 : \OType(\sigma_1, \tau_1) \qquad E[\Gamma] \vdash O_2 : \OType(\sigma_2, \tau_2)} {E[\Gamma] \vdash O_1 \otimes O_2 : \OType(\sigma_1 \times \sigma_2, \tau_1 \times \tau_2)}
     \end{gather*}
@@
         \textbf{Set-U} \qquad
-        \frac{\Gamma \vdash \sigma : \Index}{\Gamma \vdash \mathbf{U}(\sigma) : \SET(\sigma)} 
+        \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \mathbf{U}(\sigma) : \SET(\sigma)} 
         \qquad \qquad
         \textbf{Set-Prod} \qquad
-        \frac{\Gamma \vdash A : \SET(\sigma) \qquad \Gamma \vdash B : \SET(\tau)}{\Gamma \vdash A \star B : \SET(\sigma \times \tau)}
+        \frac{E[\Gamma] \vdash A : \SET(\sigma) \qquad E[\Gamma] \vdash B : \SET(\tau)}{E[\Gamma] \vdash A \star B : \SET(\sigma \times \tau)}
     \end{gather*}
@@
         \textbf{Sum-Scalar} \qquad
-        \frac{\Gamma \vdash s : \SET(\sigma) \qquad \Gamma \vdash f : \Basis(\sigma) \to \SType}{\Gamma \vdash \sum_{s} f : \SType} \\
+        \frac{E[\Gamma] \vdash s : \SET(\sigma) \qquad E[\Gamma] \vdash f : \Basis(\sigma) \to \SType}{E[\Gamma] \vdash \sum_{s} f : \SType} \\
         \\
         \textbf{Sum-Ket} \qquad
-        \frac{\Gamma \vdash s : \SET(\sigma) \qquad \Gamma \vdash f : \Basis(\sigma) \to \KType(\tau)}{\Gamma \vdash \sum_{s} f : \KType(\tau)} \\
+        \frac{E[\Gamma] \vdash s : \SET(\sigma) \qquad E[\Gamma] \vdash f : \Basis(\sigma) \to \KType(\tau)}{E[\Gamma] \vdash \sum_{s} f : \KType(\tau)} \\
         \\
         \textbf{Sum-Bra} \qquad
-        \frac{\Gamma \vdash s : \SET(\sigma) \qquad \Gamma \vdash f : \Basis(\sigma) \to \BType(\tau)}{\Gamma \vdash \sum_{s} f : \BType(\tau)} \\
+        \frac{E[\Gamma] \vdash s : \SET(\sigma) \qquad E[\Gamma] \vdash f : \Basis(\sigma) \to \BType(\tau)}{E[\Gamma] \vdash \sum_{s} f : \BType(\tau)} \\
         \\
         \textbf{Sum-Opt} \qquad
-        \frac{\Gamma \vdash s : \SET(\sigma) \qquad \Gamma \vdash f : \Basis(\sigma) \to \OType(\tau, \rho)}{\Gamma \vdash \sum_{s} f : \OType(\tau, \rho)}
+        \frac{E[\Gamma] \vdash s : \SET(\sigma) \qquad E[\Gamma] \vdash f : \Basis(\sigma) \to \OType(\tau, \rho)}{E[\Gamma] \vdash \sum_{s} f : \OType(\tau, \rho)}
     \end{gather*}
@@
     \begin{gather*}
-        \textbf{Alpha} \qquad \text{(bound variable renaming)}
-        \qquad \qquad
-        \textbf{Beta} \qquad
-        \frac{}{\Gamma \vdash ((\lambda x:T.t)\ u)\ \reduce_\beta\ t\{x/u\}} \\
+        \textbf{Beta-Arrow} \qquad
+        \frac{}{E[\Gamma] \vdash ((\lambda : T.t)\ u)\ \reduce\ t[u]} 
+        \qquad \qquad
+        \textbf{Beta-Index} \qquad
+        \frac{}{E[\Gamma] \vdash ((\lambda.t)\ u)\ \reduce\ t[u]} 
+        \\
         \\
         \textbf{Delta} \qquad
-        \frac{\WF(\Gamma) \qquad (c:=t:T) \in \Gamma}{\Gamma \vdash c\ \reduce_\delta\ t}
-        \qquad \qquad
-        \textbf{Eta} \qquad
-        \frac{\Gamma\vdash t : T \to U}{\Gamma \vdash \lambda x : T . (t\ x)\ \reduce_\eta\ t}
+        \frac{\WF(E)[\Gamma] \qquad (c:=t:T) \in E}{E[\Gamma] \vdash c\ \reduce\ t} \\
+        \\
+        \textbf{Eta-Arrow} \qquad
+        \frac{}{E[\Gamma] \vdash \lambda : T . (t\ x)\ \reduce\ t}
+        \qquad \qquad
+        \textbf{Eta-Index} \qquad
+        \frac{}{E[\Gamma] \vdash \lambda . (t\ x)\ \reduce\ t}
     \end{gather*}

### 2024-12-14 main.tex
de Bruijn expression rules updated

@@
 \begin{itemize}
-    \item Alpha equivalence is checked by a bound variable renaming on the whole term.
-    \item arrow type of $\Type \to \Type$ are forbidden. Check whether typing rules comply with this restriction.
+    \item \ 
+\end{itemize}
+
+\section{Some practices of de Bruijn substitution and application}
+\begin{itemize}
+    \item $\lambda x .((\lambda y. \lambda z. \lambda w. y)\ x) \reduce \lambda x. \lambda z. \lambda w. x $ 
+    \\
+    $ \lambda . ((\lambda . \lambda . \lambda. 2)\ 0) \reduce \lambda . \lambda . \lambda.2$
+
+    \item $\lambda x . ((\lambda y. \lambda z. y)\ (X\ x)) \reduce \lambda x . \lambda z . (X\ x)$
+    \\
+    $\lambda . ((\lambda. \lambda. 1)\ (X\ 0)) \reduce \lambda. \lambda. (X\ 1)$
+
 \end{itemize}
@@
 
-Usually, the sum body is specified by an abstraction. Therefore we use notation $\sum_{i \in s} X$ to denote $\sum_{s} \lambda i : T . X$.
+Usually, the sum body is specified by an abstraction. Therefore we use notation $\sum_s X$ to denote $\sum_{s} \lambda : T . X$ as well.
 
@@
         \textbf{Eta-Arrow} \qquad
-        \frac{}{E[\Gamma] \vdash \lambda : T . (t\ x)\ \reduce\ t}
+        \frac{}{E[\Gamma] \vdash \lambda : T . (t\ \$0)\ \reduce\ t}
         \qquad \qquad
         \textbf{Eta-Index} \qquad
-        \frac{}{E[\Gamma] \vdash \lambda . (t\ x)\ \reduce\ t}
+        \frac{}{E[\Gamma] \vdash \lambda . (t\ \$0)\ \reduce\ t}
     \end{gather*}
@@
     R-SUM-CONST0
-    & $ \sum_{x \in s} 0 \ \reduce\ 0 $ \\
+    & $ \sum_{s} 0 \ \reduce\ 0 $ \\
     %
     R-SUM-CONST1
-    & $ \sum_{x \in s} \ZEROK(\sigma)\ \reduce\ \ZEROK(\sigma) $ \\
+    & $ \sum_{s} \ZEROK(\sigma)\ \reduce\ \ZEROK(\sigma) $ \\
     %
     R-SUM-CONST2
-    & $ \sum_{x \in s} \ZEROB(\sigma)\ \reduce\ \ZEROB(\sigma) $ \\
+    & $ \sum_{s} \ZEROB(\sigma)\ \reduce\ \ZEROB(\sigma) $ \\
     %
     R-SUM-CONST3
-    & $ \sum_{x \in s} \ZEROO(\sigma, \tau)\ \reduce\ \ZEROO(\sigma, \tau) $ \\
+    & $ \sum_{s} \ZEROO(\sigma, \tau)\ \reduce\ \ZEROO(\sigma, \tau) $ \\
     %
     R-SUM-CONST4
-    & $ \ONEO(\sigma) \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)} \ket{i} \cdot \bra{i} $
+    & $ \ONEO(\sigma) \ \reduce\ \sum_{\mathbf{U}(\sigma)} \ket{i} \cdot \bra{i} $
 \end{ruletable}
@@
     R-SUM-ELIM0
-    & $ i \text{ free in } t \vdash \sum_{i \in \mathbf{U}(\sigma)} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} \delta_{i, t} \ \reduce\ \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n}  1$ \\
+    & $ \$i \text{ free in } t \vdash \sum_{\mathbf{U}(\sigma)} \sum_{s_1} \cdots \sum_{s_i} \delta_{\$i, t} \ \reduce\ \sum_{s_1} \cdots \sum_{s_i}  1$ \\
     %
     R-SUM-ELIM1
-    & $ i \text{ free in } t \vdash \sum_{i \in \mathbf{U}(\sigma)} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (a_1 \times \cdots \times \delta_{i, t} \times \cdots \times a_n) $ \\
-    & $ \reduce\ \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} a_1\{i/t\} \times \cdots \times a_n\{i/t\} $ \\
+    & $ \$i \text{ free in } t \vdash \sum_{\mathbf{U}(\sigma)} \sum_{s_1} \cdots \sum_{s_i} (a_1 \times \cdots \times \delta_{\$i, t} \times \cdots \times a_n) $ \\
+    & $ \reduce\ \sum_{s_1} \cdots \sum_{s_i} a_1\{\$i/t\} \times \cdots \times a_n\{\$i/t\} $ \\
     %
     R-SUM-ELIM2
-    & $ i \text{ free in } t \vdash \sum_{i \in \mathbf{U}(\sigma)} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (\delta_{i, t}.A) \ \reduce\ \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} A\{i/t\} $ \\
+    & $ \$i \text{ free in } t \vdash \sum_{\mathbf{U}(\sigma)} \sum_{s_1} \cdots \sum_{s_i} (\delta_{\$i, t}.A) \ \reduce\ \sum_{s_1} \cdots \sum_{s_i} A\{\$i/t\} $ \\
     %
     R-SUM-ELIM3
-    & $ i \text{ free in } t \vdash \sum_{i \in \mathbf{U}(\sigma)} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (a_1 \times \cdots \times \delta_{i, t} \times \cdots \times a_n).A $ \\
-    & $ \reduce\ \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n}  (a_1\{i/t\} \times \cdots \times a_n\{i/t\}).A\{i/t\} $ \\
+    & $ \$i \text{ free in } t \vdash \sum_{\mathbf{U}(\sigma)} \sum_{s_1} \cdots \sum_{s_i} (a_1 \times \cdots \times \delta_{\$i, t} \times \cdots \times a_i).A $ \\
+    & $ \reduce\ \sum_{s_1} \cdots \sum_{s_i}  (a_1\{\$i/t\} \times \cdots \times a_n\{\$i/t\}).A\{\$i/t\} $ \\
     R-SUM-ELIM4
-    & $ \sum_{i \in M} \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n}  \delta_{i, j} \ \reduce\ \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} 1 $ \\
+    & $ \sum_{M} \sum_{M} \sum_{s_1} \cdots \sum_{s_i}  \delta_{\$(i + 1), \$i} \ \reduce\ \sum_{M} \sum_{s_1} \cdots \sum_{k_n \in s_i} 1 $ \\
     %
     R-SUM-ELIM5
-    & $ \sum_{i \in M} \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (a_1 \times \dots \times \delta_{i, j} \times \cdots \times a_n) $ \\
-    & $ \reduce\ \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (a_1\{j/i\} \times \cdots \times a_n\{j/i\}) $ \\
+    & $ \sum_{M} \sum_{M} \sum_{s_1} \cdots \sum_{s_i} (a_1 \times \dots \times \delta_{\$(i+1), \$i} \times \cdots \times a_n) $ \\
+    & $ \reduce\ \sum_{M} \sum_{s_1} \cdots \sum_{s_n} (a_1\{\$i/\$(i+1)\} \times \cdots \times a_n\{\$i/\$(i+1)\}) $ \\
     %
     R-SUM-ELIM6
-    & $ \sum_{i \in M} \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (\delta_{i, j}.A) \ \reduce\ \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} A\{j/i\} $ \\
+    & $ \sum_{M} \sum_{M} \sum_{s_1} \cdots \sum_{s_i} (\delta_{\$(i+1), \$i}.A) \ \reduce\ \sum_{M} \sum_{s_1} \cdots \sum_{s_i} A\{\$i/\$(i+1)\} $ \\
     %
     R-SUM-ELIM7
-    & $ \sum_{i \in M} \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (a_1 \times \cdots \times \delta_{i, j} \times \cdots \times a_n).A $ \\
-    & $ \reduce\ \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (a_1\{j/i\} \times \cdots \times a_n\{j/i\}).A\{j/i\} $ \\
+    & $ \sum_{M} \sum_{M} \sum_{s_1} \cdots \sum_{s_i} (a_1 \times \cdots \times \delta_{\$(i+1), \$i} \times \cdots \times a_n).A $ \\
+    & $ \reduce\ \sum_{M} \sum_{s_1} \cdots \sum_{s_i} (a_1\{\$i/\$(i+1)\} \times \cdots \times a_n\{\$i/\$(i+1)\}).A\{\$i/\$(i+1)\} $ \\
 \end{ruletable}
@@
     R-SUM-PUSH0
-    & $ b_1 \times \cdots \times (\sum_{i \in M} a) \times \cdots \times b_n \ \reduce\ \sum_{i \in M} (b_1 \times \cdots \times a \times \cdots \times b_n) $ \\
+    & $ b_1 \times \cdots \times (\sum_{M} a) \times \cdots \times b_n \ \reduce\ \sum_{M} (b_1 \times \cdots \times a \times \cdots \times b_n) $ \\
     %
     R-SUM-PUSH1
-    & $ (\sum_{i \in M} a)^* \ \reduce\ \sum_{i \in M} a^* $ \\
+    & $ (\sum_{M} a)^* \ \reduce\ \sum_{M} a^* $ \\
     %
     R-SUM-PUSH2
-    & $ (\sum_{i \in M} X)^\dagger \ \reduce\ \sum_{i \in M} X^\dagger $ \\
+    & $ (\sum_{M} X)^\dagger \ \reduce\ \sum_{M} X^\dagger $ \\
     %
     R-SUM-PUSH3
-    & $ a.(\sum_{i \in M} X) \ \reduce\ \sum_{i \in M} (a.X) $ \\
+    & $ a.(\sum_{M} X) \ \reduce\ \sum_{M} (a.X) $ \\
     %
     R-SUM-PUSH4
-    & $ (\sum_{i \in M} a).X \ \reduce\ \sum_{i \in M} (a.X) $ \\
+    & $ (\sum_{M} a).X \ \reduce\ \sum_{M} (a.X) $ \\
     %
     R-SUM-PUSH5
-    & $ (\sum_{i \in M} B)\cdot K \ \reduce\ \sum_{i \in M}(B \cdot K) $ \\
+    & $ (\sum_{M} B)\cdot K \ \reduce\ \sum_{M}(B \cdot K) $ \\
     %
     R-SUM-PUSH6
-    & $ (\sum_{i \in M} O)\cdot K \ \reduce\ \sum_{i \in M}(O \cdot K) $ \\
+    & $ (\sum_{M} O)\cdot K \ \reduce\ \sum_{M}(O \cdot K) $ \\
     %
     R-SUM-PUSH7
-    & $ (\sum_{i \in M} B)\cdot O \ \reduce\ \sum_{i \in M}(B \cdot O) $ \\
+    & $ (\sum_{M} B)\cdot O \ \reduce\ \sum_{M}(B \cdot O) $ \\
     %
     R-SUM-PUSH8
-    & $ (\sum_{i \in M} K)\cdot B \ \reduce\ \sum_{i \in M}(K \cdot B) $ \\
+    & $ (\sum_{M} K)\cdot B \ \reduce\ \sum_{M}(K \cdot B) $ \\
     %
     R-SUM-PUSH9
-    & $ (\sum_{i \in M} O_1)\cdot O_2 \ \reduce\ \sum_{i \in M}(O_1 \cdot O_2) $ \\
+    & $ (\sum_{M} O_1)\cdot O_2 \ \reduce\ \sum_{M}(O_1 \cdot O_2) $ \\
     %
     R-SUM-PUSH10
-    & $ B \cdot (\sum_{i \in M} K) \ \reduce\ \sum_{i \in M}(B \cdot K) $ \\
+    & $ B \cdot (\sum_{M} K) \ \reduce\ \sum_{M}(B \cdot K) $ \\
     %
     R-SUM-PUSH11
-    & $ O \cdot (\sum_{i \in M} K) \ \reduce\ \sum_{i \in M}(O \cdot K) $ \\
+    & $ O \cdot (\sum_{M} K) \ \reduce\ \sum_{M}(O \cdot K) $ \\
     %
     R-SUM-PUSH12
-    & $ B \cdot (\sum_{i \in M} O) \ \reduce\ \sum_{i \in M}(B \cdot O) $ \\
+    & $ B \cdot (\sum_{M} O) \ \reduce\ \sum_{M}(B \cdot O) $ \\
     %
     R-SUM-PUSH13
-    & $ K \cdot (\sum_{i \in M} B) \ \reduce\ \sum_{i \in M}(K \cdot B) $ \\
+    & $ K \cdot (\sum_{M} B) \ \reduce\ \sum_{M}(K \cdot B) $ \\
     %
     R-SUM-PUSH14
-    & $ O_1 \cdot (\sum_{i \in M} O_2) \ \reduce\ \sum_{i \in M}(O_1 \cdot O_2) $ \\
+    & $ O_1 \cdot (\sum_{M} O_2) \ \reduce\ \sum_{M}(O_1 \cdot O_2) $ \\
     %
     R-SUM-PUSH15
-    & $ (\sum_{i \in M} X_1) \otimes X_2 \ \reduce\ \sum_{i \in M} (X_1 \otimes X_2) $ \\
+    & $ (\sum_{M} X_1) \otimes X_2 \ \reduce\ \sum_{M} (X_1 \otimes X_2) $ \\
     %
     R-SUM-PUSH16
-    & $ X_1 \otimes (\sum_{i \in M} X_2) \ \reduce\ \sum_{i \in M} (X_1 \otimes X_2) $
+    & $ X_1 \otimes (\sum_{M} X_2) \ \reduce\ \sum_{M} (X_1 \otimes X_2) $
 \end{ruletable}
@@
     R-SUM-ADDS0
-    & $\sum_{i \in M}(a_1 + \cdots + a_n) \ \reduce\ (\sum_{i \in M} a_1) + \cdots + (\sum_{i \in M} a_n) $ \\
+    & $\sum_{M}(a_1 + \cdots + a_n) \ \reduce\ (\sum_{M} a_1) + \cdots + (\sum_{M} a_n) $ \\
     %
     R-SUM-ADD0
-    & $\sum_{i \in M}(X_1 + \cdots + X_n) \ \reduce\ (\sum_{i \in M} X_1) + \cdots + (\sum_{i \in M} X_n) $ \\
+    & $\sum_{M}(X_1 + \cdots + X_n) \ \reduce\ (\sum_{M} X_1) + \cdots + (\sum_{M} X_n) $ \\
     %
@@
     R-SUM-INDEX0
-    & $ \sum_{i \in \mathbf{U}(\sigma \times \tau)} A \ \reduce\ \sum_{j \in \mathbf{U}(\sigma)} \sum_{k \in \mathbf{U}(\tau)} A\{i/(j, k)\} $ \\
+    & $ \sum_{\mathbf{U}(\sigma \times \tau)} A \ \reduce\ \sum_{\mathbf{U}(\sigma)} \sum_{\mathbf{U}(\tau)} \textsf{adjust}(A, 1, 2)\{\$0/(\$2, \$1)\} $ \\
     %
     R-SUM-INDEX1
-    & $ \sum_{i \in M_1 \star M_2} A \ \reduce\ \sum_{j \in M_1} \sum_{k \in M_2} A\{i/(j, k)\} $ \\
+    & $ \sum_{M_1 \star M_2} A \ \reduce\ \sum_{M_1} \sum_{M_2} \textsf{adjust}(A, 1, 2)\{\$0/(\$2, \$1)\} $ \\
     %
     R-SUM-SWAP
-    & $ M_1 < M_2 \vdash \sum_{i \in M_2} \sum_{j \in M_1} X \ \reduce\ \sum_{j \in M_1} \sum_{i \in M_2} X $
+    & $ M_1 < M_2 \vdash \sum_{M_2} \sum_{M_1} X \ \reduce\ \sum_{M_1} \sum_{M_2} \textsf{swap01}(X) $
 \end{ruletable}

### 2024-12-17 main.tex
tool basics implementated

--- 2024-12-10 main.tex
@@
 
+\section{Consideration}
+\begin{itemize}
+    \item AC equivalence and alpha equivalence.
+    \item Efficient AC equivalence checking by sorting on syntax.
+    \item Sorting requires that alpha equivalent terms have the same syntax.
+    \item de Bruijn expression satisfies this requirement, but the typing and substitution becomes very complicated in the type index scenario.
+    \item only compute de Bruijn form in the equivalence checking and sorting phase.
+\end{itemize}
+
 \section{TODO}
 \begin{itemize}
-    \item Alpha equivalence is checked by a bound variable renaming on the whole term.
-    \item arrow type of $\Type \to \Type$ are forbidden. Check whether typing rules comply with this restriction.
+    \item implemented trace output for the whole pipeline.
+    \item Note that the context should also be maintained during rewriting matching.
+    \item better parser and output.
 \end{itemize}
@@
     \begin{align*}
-        T ::=\ & x\ |\ \Basis(\sigma)\ |\ \SType\ |\ \KType(\sigma)\ |\ \BType(\sigma)\ |\ \OType(\sigma_1, \sigma_2)\ |\ T_1 \to T_2\ |\ \forall .T\ |\ \SET(\sigma). \\
+        T ::=\ & x\ |\ \Basis(\sigma)\ |\ \SType\ |\ \KType(\sigma)\ |\ \BType(\sigma)\ |\ \OType(\sigma_1, \sigma_2)\ |\ T_1 \to T_2\ |\ \forall x.T\ |\ \SET(\sigma). \\
     \end{align*}
@@
     \begin{align*}
-        e ::=\ & x\ |\ \$i\ |\ (e_1, e_2)\ |\ \lambda : T.e\ |\ \lambda.e\ |\ e_1\ e_2 \\
+        e ::=\ & x\ |\ (e_1, e_2)\ |\ \lambda x : T.e\ |\ \mu x.e\ |\ e_1\ e_2 \\
         & |\ 0\ |\ 1\ |\ \ADDS(e_1 \cdots e_n) \ |\ e_1 \times \cdots \times e_n\ |\ e^* \ |\ \delta_{e_1, e_2}\ |\ \DOT(e_1\ e_2) \\
@@
 
-Usually, the sum body is specified by an abstraction. Therefore we use notation $\sum_{i \in s} X$ to denote $\sum_{s} \lambda i : T . X$.
+Usually, the sum body is specified by an abstraction. Therefore we use notation $\sum_s X$ to denote $\sum_{x \in s} \lambda x : T . X$ as well.
 
@@
         E ::=\ & []\ |\ E; x : \Index\ |\ E; T : \Type\ |\ E; x : T\ |\ E; x := t : T. \\
-        \Gamma ::=\ & []\ |\ \Gamma; \Index\ |\ \Gamma; T.
+        \Gamma ::=\ & []\ |\ \Gamma; x : \Index\ |\ \Gamma; x : T.
     \end{align*}
@@
 Note that in the following rules, $x : \Index$ and $x : \Type$ are not considered as $x : T$ judgements.
-The context $\Gamma$ acts as the stack to indicate the typing of de Bruijn notations.
-The notation $\Gamma_i$ is defined as the $i$-th element in $\Gamma$ from the end.
-That is, if $\Gamma \equiv \Gamma'; T$, then $\Gamma_0 = T$ and $\Gamma_i = \Gamma'_{i-1}$.
 
@@
         \textbf{W-AssumE-Type} \qquad
-        \frac{\WF(E)[] \qquad x \notin E}{\WF(E; x : \Type)} \\
+        \frac{\WF(E)[] \qquad x \notin E}{\WF(E; x : \Type)[]} \\
         \\
@@
         \textbf{W-AssumC-Index} \qquad
-        \frac{\WF(E)[\Gamma]}{\WF(E)[\Gamma; \Index]}
+        \frac{\WF(E)[\Gamma]}{\WF(E)[\Gamma; x : \Index]}
         \qquad \qquad
         \textbf{W-AssumC-Term} \qquad
-        \frac{E[\Gamma] \vdash T : \Type}{\WF(E)[\Gamma; T]}
+        \frac{E[\Gamma] \vdash T : \Type}{\WF(E)[\Gamma; x : T]}
     \end{gather*}
@@
         \textbf{Index-Var} \qquad
-        \frac{\WF(E)[\Gamma] \qquad x : \Index \in E}{E[\Gamma] \vdash x : \Index}
+        \frac{\WF(E)[\Gamma] \qquad x : \Index \in E[\Gamma]}{E[\Gamma] \vdash x : \Index}
         \qquad \qquad
@@
     \begin{gather*}
-        \textbf{Type-Context-Var} \qquad
-        \frac{\WF(E)[\Gamma]}{E[\Gamma]\vdash \$i : \Gamma_i}
-        \\
-        \\
         \textbf{Type-Arrow} \qquad
@@
         \textbf{Type-Index} \qquad
-        \frac{E[\Gamma ; \Index] \vdash U : \Type}{E[\Gamma] \vdash \forall.U : \Type} \\
+        \frac{E[\Gamma ; x : \Index] \vdash U : \Type}{E[\Gamma] \vdash \forall x.U : \Type} \\
         \\        
@@
         \textbf{Term-Var} \qquad
-        \frac{\WF(E)[\Gamma] \qquad (x:T) \in E \text{ or } (x := t : T) \in E \text{ for some $t$}}{E[\Gamma] \vdash x : T} \\
+        \frac{\WF(E)[\Gamma] \qquad (x:T) \in E[\Gamma] \text{ or } (x := t : T) \in E \text{ for some $t$}}{E[\Gamma] \vdash x : T} \\
         \\
         \textbf{Lam} \qquad
-        \frac{E[\Gamma; T] \vdash t : U}{E[\Gamma] \vdash (\lambda : T . t) : T \to U}
+        \frac{E[\Gamma; x : T] \vdash t : U}{E[\Gamma] \vdash (\lambda x : T . t) : T \to U}
         \qquad \qquad
         \textbf{Index} \qquad
-        \frac{E[\Gamma; \Index] \vdash t : U}{E[\Gamma] \vdash (\lambda.t) : \forall .U} \\
+        \frac{E[\Gamma; x : \Index] \vdash t : U}{E[\Gamma] \vdash (\mu x.t) : \forall x.U} \\
         \\
@@
         \textbf{App-Index} \qquad
-        \frac{E[\Gamma] \vdash t:\forall .U \qquad E[\Gamma] \vdash u : \Index}{E[\Gamma] \vdash (t\ u):U[u]}
+        \frac{E[\Gamma] \vdash t:\forall x.U \qquad E[\Gamma] \vdash u : \Index}{E[\Gamma] \vdash (t\ u):U\{x/u\}}
     \end{gather*}
@@
 \section{Conversions and Reductions}
+
 \begin{figure}[h]
     \begin{gather*}
+        \frac{E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash K \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)}(\bra{i}\cdot K).\ket{i}} \\
+        \\
+        \frac{E[\Gamma] \vdash B : \BType(\sigma)}{E[\Gamma] \vdash B \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)}(B \cdot \ket{i}).\bra{i}} \\
+        \\
+        \frac{E[\Gamma] \vdash O : \OType(\sigma, \tau)}{E[\Gamma] \vdash O \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)} \sum_{j \in \mathbf{U}(\tau)}(\bra{i} \cdot O \cdot \ket{j}).(\ket{i} \cdot \bra{j})}
+    \end{gather*}
+\end{figure}
+
+\begin{figure}[h]
+    \begin{gather*}
         \textbf{Beta-Arrow} \qquad
-        \frac{}{E[\Gamma] \vdash ((\lambda : T.t)\ u)\ \reduce\ t[u]} 
+        \frac{}{E[\Gamma] \vdash ((\lambda x : T.t)\ u)\ \reduce\ t\{x/u\}} 
         \qquad \qquad
         \textbf{Beta-Index} \qquad
-        \frac{}{E[\Gamma] \vdash ((\lambda.t)\ u)\ \reduce\ t[u]} 
+        \frac{}{E[\Gamma] \vdash ((\mu x.t)\ u)\ \reduce\ t\{x/u\}} 
         \\
@@
         \textbf{Eta-Arrow} \qquad
-        \frac{}{E[\Gamma] \vdash \lambda : T . (t\ x)\ \reduce\ t}
+        \frac{}{E[\Gamma] \vdash \lambda x : T . (t\ x)\ \reduce\ t}
         \qquad \qquad
         \textbf{Eta-Index} \qquad
-        \frac{}{E[\Gamma] \vdash \lambda . (t\ x)\ \reduce\ t}
+        \frac{}{E[\Gamma] \vdash \mu x . (t\ x)\ \reduce\ t}
     \end{gather*}
@@
         &|\ \texttt{ShowAll} \\
-        &|\ \texttt{Normalize}(term) \\
+        &|\ \texttt{Normalize}(term)\ |\ \texttt{Normalize}(term\ \textsf{Trace}) \\
         &|\ \texttt{CheckEq}(term\ term) \\

### 2024-12-30 main.tex
update 2024 end

@@
     \item only compute de Bruijn form in the equivalence checking and sorting phase.
+    \item pipeline: 
+        \begin{enumerate}
+            \item preprocessing
+            \item 1st rewritings
+            \item variable expansion
+            \item 2nd rewritings 
+            \item computing de Bruijn and sorting
+        \end{enumerate}
+
+    \item I found that typing of inner bound variables also requires modifying the context.
+    Need to modify the tree visiting algorithm
+    (Typed term rewriting with bound variables is not that easy)    
 \end{itemize}
@@
     \item implemented trace output for the whole pipeline.
-    \item Note that the context should also be maintained during rewriting matching.
     \item better parser and output.
+    \item add Mathematica support for scalars
+    \item output trace to Coq
+    \item consider labelled Dirac notations
+    \item better trace output
+    \item better error logic
+    \item better format output
+\end{itemize}
+
+\section{Things to Note}
+\begin{itemize}
+    \item The context should also be maintained during rewriting matching.
+    \item We don't allow eta reduction. It will intertwine with SUM-SWAP and break the confluence.
 \end{itemize}
@@
     \begin{align*}
-        \sigma ::=\ & x\ |\ \sigma_1 \times \sigma_2.
+        \sigma ::=\ & x\ |\ \sigma_1 \times \sigma_2\ |\ \{c_0, c_1, \cdots, c_n\}.
     \end{align*}
@@
         \textbf{Index-Prod} \qquad
-        \frac{E[\Gamma] \vdash \sigma : \Index \qquad E[\Gamma] \vdash \tau : \Index}{E[\Gamma] \vdash \sigma \times \tau : \Index}
+        \frac{E[\Gamma] \vdash \sigma : \Index \qquad E[\Gamma] \vdash \tau : \Index}{E[\Gamma] \vdash \sigma \times \tau : \Index} \\
+        \\
+        \textbf{Index-Concrete} \qquad
+        \frac{\WF(E)[\Gamma] \qquad c_i \notin E[\Gamma] \text{ for all } i \qquad c_i \text{ are unique}}{E[\Gamma] \vdash \{c_1, c_2, \cdots, c_n\} : \Index}
     \end{gather*}
@@
         \textbf{Delta} \qquad
-        \frac{\WF(E)[\Gamma] \qquad (c:=t:T) \in E}{E[\Gamma] \vdash c\ \reduce\ t} \\
-        \\
-        \textbf{Eta-Arrow} \qquad
-        \frac{}{E[\Gamma] \vdash \lambda x : T . (t\ x)\ \reduce\ t}
-        \qquad \qquad
-        \textbf{Eta-Index} \qquad
-        \frac{}{E[\Gamma] \vdash \mu x . (t\ x)\ \reduce\ t}
+        \frac{\WF(E)[\Gamma] \qquad (c:=t:T) \in E}{E[\Gamma] \vdash c\ \reduce\ t}
     \end{gather*}
@@
 \yx{The rules R-SUM-ADD1 to R-SUM-ADD3 needs identical sum terms, which requires the rewriting to be after the alpha normalization. We don't implement them for now.}
+
+\section{Preprocessing}
+\begin{figure}
+    \begin{gather*}
+        \frac{E[\Gamma] \vdash B : \BType(\sigma) \qquad E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash B \circ K : \SType}
+    \end{gather*}
+    \caption{The typing rules for preprocessing}
+\end{figure}
 
@@
 
+Comment \texttt{(* ... *)} can be inserted between commands.
+
+These are tye parsing rules for different expressions:
+\begin{itemize}
+    \item \texttt{Def ID := term.} --- \texttt{Def(ID term)}
+    \item \texttt{Def ID := term : type.} --- \texttt{Def(ID term type)}
+    \item \texttt{Var ID : type.} --- \texttt{Var(ID type)}
+    \item \texttt{Check ID.} --- \texttt{Check(term)}
+    \item \texttt{Show ID.} --- \texttt{Show(ID)}
+    \item \texttt{ShowAll.} --- \texttt{ShowAll}
+    \item \texttt{Normalize term.} --- \texttt{Normalize(term)}
+    \item \texttt{Normalize term with trace.} --- \texttt{Normalize(term Trace)}
+    \item \texttt{Check term = term.} --- \texttt{CheckEq(term term)}
+    \item \texttt{T1 -> T2} --- \texttt{Arrow(T1 T2)}.
+    \item \texttt{forall x. T} --- \texttt{Forall(x T)}
+    \item \texttt{(e1, e2)} --- \texttt{PAIR(e1 e2)}
+    \item \texttt{fun x : T => e} --- \texttt{fun(x T e)}
+    \item \texttt{idx x => e} --- \texttt{idx(x e)}
+    \item \texttt{e1 @ e2} --- \texttt{COMPO(e1 e2)}
+    \item \texttt{e1 + ... + en} --- \texttt{ADDG(e1 ... en)}
+    \item \texttt{e1 * ... * en} --- \texttt{STAR(e1 ... en)}
+    \item \texttt{e1\^{}*} --- \texttt{CONJ(e1)}
+    \item \texttt{delta(e1, e2)} --- \texttt{DELTA(e1 e2)}
+    \item \texttt{|e>} --- \texttt{KET(e)}
+    \item \texttt{<e|} --- \texttt{BRA(e)}
+    \item \texttt{e1\^{}D} --- \texttt{ADJ(e1)}
+    \item \texttt{e1.e2} --- \texttt{SCR(e1 e2)}
+    \item \texttt{Sum(i in s, e)} --- \texttt{SSUM(i s e)}
+\end{itemize}
+
+These symbols will be transformed into internal language:
+\begin{itemize}
+    \item $A \circ B$ : $S \circ S$, $S \circ K$, $S \circ B$, $S \circ O$, $K \circ S$, $K \circ K$, $K \circ B$, $B \circ S$, $B \circ K$, $B \circ B$, $B \circ O$, $O \circ S$, $O \circ K$, $O \circ O$, $f\circ a$ (arrow), $f\circ a$ (index).
+    \item \texttt{STAR(a ... b)} :  $\sigma_1 \times \sigma_2$, $\MULS(a \cdots b)$, $O_1 \otimes O_2$, $ M_1 \star M_2$
+    \item \texttt{ADDG(e1 ... en)} : $\ADDS(e1 \cdots en)$, $\ADD(e1 \cdots en)$
+    \item \texttt{SSUM(i S e)} : \texttt{SUM(s FUN(i T e))}
+\end{itemize}
+
 \section{Rewriting Control and Intermediate Language}

### 2025-01-03 main.tex
update 2025/1/3

@@
 \begin{itemize}
+    \item We can emphasize that we are moving from pure term rewriting to a hybrid algorithm.
     \item AC equivalence and alpha equivalence.
@@
             \item preprocessing
+            \item rename unique bound variable names
             \item 1st rewritings
@@
             \item 2nd rewritings 
-            \item computing de Bruijn and sorting
+            \item sorting modulo bound variables
+            \item sorting successive sum.
+The order depends on there occurances in the last sorting result.
+If no occurance, then the order will depend on the set (for sum) and the type (for lambda abstraction only).
+            \item computing de Bruijn
         \end{enumerate}
@@
     \item implemented trace output for the whole pipeline.
-    \item better parser and output.
+    \item better output.
     \item add Mathematica support for scalars
@@
     \item better format output
+    \item remove term bank and hash consing
 \end{itemize}
@@
     \item We don't allow eta reduction. It will intertwine with SUM-SWAP and break the confluence.
+    \item In each Delta reduction, the bound variables are replaced with unique variables. This should help solve the problem of conflicting variable names during substitution.
+    \item I guess it may still be necessary to try all different bound variable assignments.
 \end{itemize}
@@
     \begin{align*}
-        \sigma ::=\ & x\ |\ \sigma_1 \times \sigma_2\ |\ \{c_0, c_1, \cdots, c_n\}.
+        \sigma ::=\ & x\ |\ \sigma_1 \times \sigma_2\ |\ \mathsf{Qdit}_n.
     \end{align*}
@@
         \\
-        \textbf{Index-Concrete} \qquad
-        \frac{\WF(E)[\Gamma] \qquad c_i \notin E[\Gamma] \text{ for all } i \qquad c_i \text{ are unique}}{E[\Gamma] \vdash \{c_1, c_2, \cdots, c_n\} : \Index}
+        \textbf{Index-Qudit} \qquad
+        \frac{\WF(E)[\Gamma]}{E[\Gamma] \vdash \mathsf{Qudit}_n : \Index}
     \end{gather*}
@@
     R-SUM-INDEX1
-    & $ \sum_{i \in M_1 \star M_2} A \ \reduce\ \sum_{j \in M_1} \sum_{k \in M_2} A\{i/(j, k)\} $ \\
-    %
-    R-SUM-SWAP
-    & $ M_1 < M_2 \vdash \sum_{i \in M_2} \sum_{j \in M_1} X \ \reduce\ \sum_{j \in M_1} \sum_{i \in M_2} X $
+    & $ \sum_{i \in M_1 \star M_2} A \ \reduce\ \sum_{j \in M_1} \sum_{k \in M_2} A\{i/(j, k)\} $
 \end{ruletable}

### 2025-01-13 labelled.tex
integrate Li's notes on labelled, change to LNCS format

\newcommand{\fst}{\mathsf{fst}\ }
\newcommand{\snd}{\mathsf{snd}\ }
\newcommand{\var}{\mathsf{var}\ }
\newcommand{\Var}{\mathsf{Var}\ }
\newcommand{\reg}{\mathsf{Reg}\ }
\newcommand{\DType}{\mathcal{D}}
\newcommand{\cR}{\mathcal{R}}
\newcommand{\cN}{\mathcal{N}}
\newcommand{\tD}{\tilde{D}}
\newcommand{\te}{\tilde{e}}
\newcommand{\tT}{\tilde{T}}
\newcommand{\tADD}{\widetilde{ADD}}
\newcommand{\bU}{\mathbf{U}}
\newcommand{\<}{\langle}
\newcommand{\simp}{\mathsf{Simp}}
\newcommand{\List}{\mathsf{list}}
\renewcommand{\>}{\rangle}

\section{Labelled Dirac Notation}

\lz{I suggest not adding first and second. Of course, we can, decide this later after discussion.}

\lz{We further only consider constant registers.}

Assume the name set of registers $\cR$.
\begin{definition}[quantum registers]
  \begin{align*}
    R ::= r\in\cR \mid (R, R)
  \end{align*}
\end{definition}

\begin{definition}[register variable set]
The variable set of a register is defined inductively by:
\begin{itemize}
    \item $R\equiv r$: $\var R = \{r\}$;
    \item $R\equiv (R_1,R_2)$: $\var R = \var R_1 \cup\var R_2$.
\end{itemize}  
We define the following relations for quantum registers:
\begin{itemize}
  \item $R$ \textbf{belongs to} $Q$, written as $R\ \text{in}\ Q$, if $\var R\subseteq \var Q$;
  \item $R$ \textbf{is disjoint with} $Q$, written as $R \| Q$, if $\var R \cap \var Q = \emptyset$.
\end{itemize}
\end{definition}

\lz{Since the construction of the quantum register is known, the variable set can be directly computed. Similarly, we use the (computable) set as dependent parameter of the type of Labelled Dirac Term.}

\textbf{Remark: } Set operations: $\cup$ for union; $\cap$ for intersection; $\setminus$ for difference. So,
% \begin{itemize}
$ S_1 \cap S_2 \equiv S_1 \cup S_2 \setminus (S_1 \setminus S_2) \setminus (S_2 \setminus S_1) $.
% \end{itemize}

\begin{definition}[labelled core language]
  The \textbf{labelled core langauge} includes all symbols in the core language of Dirac notation, as well as symbols defined below.
  Here, $s\subseteq \cR$ is the set of register's name, it is computable and thus we do not introduce syntax for it.
  \begin{align*}
    T & ::= \DType(s,s) \mid \mathsf{Reg}(\sigma) \\
    e & ::= R \mid |i\>_r \mid {}_r\<i| \mid e_R \mid e_{R;R} \mid  
    e \otimes e \mid e \cdot e \\
  \end{align*}
  In other words, we don't allow variables for labelled core language for now. In particular, $|i\>_r$ and ${}_r\<i|$ are used as reserved syntax for eliminating registers.
\end{definition}

\subsubsection*{Typing rules}
\lz{context? environment? Since quantum registers are not bound variables, we might just add it to the environment.}

The extra syntax for the environment.
\begin{gather}
    E ::= \dots \mid E; r : \reg(\sigma)
\end{gather}

\begin{gather}
  \textbf{W-AssumE-Reg}\qquad 
  \frac{E[]\vdash \sigma : \Index\qquad r\notin E}{\WF[E; r : \reg(\sigma)[]]}
\end{gather}

\begin{gather*}
  \textbf{Type-Labelled} \qquad
  \frac{E[\Gamma] \vdash r : \reg(\sigma_r) \text{ for all $r$ in $s_1$ and $s_2$} }{E[\Gamma] \vdash \DType(s_1, s_2) : \textsf{Type}} \\
  \\
  \textbf{Reg-Var} \qquad
  \frac{\WF(E[\Gamma]) \qquad r : \reg(\sigma) \in E}{E[\Gamma] \vdash r : \reg(\sigma)}  \\ 
  \\
  \textbf{Reg-Pair}\qquad
  \frac{E[\Gamma] \vdash R : \reg \sigma\qquad E[\Gamma] \vdash Q : \reg \tau\qquad R\| Q}{E[\Gamma] \vdash (R,Q) : \reg \sigma \times \tau}
\end{gather*}

\begin{gather*}
    \textbf{Type-L-Base-Ket}\qquad 
    \frac{r : \reg \sigma \in E\qquad E[\Gamma] \vdash i : \Basis(\sigma)}{E[\Gamma] \vdash |i\>_r : \DType(\{r\}; \emptyset)} \\
    \textbf{Type-L-Base-Bra}\qquad 
    \frac{r : \reg \sigma \in E \qquad E[\Gamma] \vdash i : \Basis(\sigma)}{E[\Gamma] \vdash {}_r\<i| : \DType(\emptyset; \{r\})} \\
    \textbf{Type-L-Ket}\qquad 
    \frac{E[\Gamma] \vdash R : \reg \sigma\qquad E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash K_R : \DType(\var R; \emptyset)} \\
    \textbf{Type-L-Bra}\qquad 
    \frac{E[\Gamma] \vdash R : \reg \sigma\qquad E[\Gamma] \vdash B : \BType(\sigma)}{E[\Gamma] \vdash B_R : \DType(\emptyset; \var R)} \\
    \textbf{Type-L-Opt}\qquad 
    \frac{E[\Gamma] \vdash R_1 : \reg \sigma_1\qquad 
    E[\Gamma] \vdash R_2 : \reg \sigma_2\qquad 
    E[\Gamma] \vdash O : \OType(\sigma_1; \sigma_2)}
    {E[\Gamma] \vdash O_{R_1;R_2} : \DType(\var R_1; \var R_2)} \\
    \textbf{Type-L-Conj}\qquad 
    \frac{E[\Gamma] \vdash \tD : \DType(s_1;s_2)}{E[\Gamma] \vdash \tD^\dagger : \DType(s_2;s_1)} \qquad
    \textbf{Type-L-SCL}\qquad 
    \frac{E[\Gamma] \vdash S : \SType\qquad E[\Gamma] \vdash \tD : \DType(s_1;s_2)}{E[\Gamma] \vdash S.\tD : \DType(s_1;s_2)} \\
    \textbf{Type-L-ADD}\qquad
    \frac{E[\Gamma] \vdash \tD_i : \DType(s_1;s_2)\quad \text{forall } i}{E[\Gamma] \vdash \tD_1+\cdots+\tD_n : \DType(s_1;s_2)} \\
    \textbf{Type-L-TSR}\qquad
    \frac{E[\Gamma] \vdash \tD_1 : \DType(s_1;s_1')\qquad E[\Gamma] \vdash \tD_2 : \DType(s_2;s_2')\qquad s_1\cap s_2 = \emptyset\qquad s_1'\cap s_2' = \emptyset}{E[\Gamma] \vdash \tD_1\otimes\tD_2 : \DType(s_1\cup s_2;s_1'\cup s_2')} \\
    \textbf{Type-L-DOT}\qquad
    \frac{E[\Gamma] \vdash \tD_1 : \DType(s_1;s_1')\qquad E[\Gamma] \vdash \tD_2 : \DType(s_2;s_2')\qquad 
    s_1 \cap s_2 \backslash s_1' = \emptyset \qquad
    s_2' \cap s_1' \backslash s_2 = \emptyset}
    {E[\Gamma] \vdash \tD_1\cdot\tD_2 : \DType(s_1 \cup (s_2\backslash s_1'); s_2' \cup (s_1'\backslash s_2))}
\end{gather*}

\subsubsection*{Eliminating the registers}

\begin{enumerate}
  \item For any register $R$ s.t. $E[\Gamma]\vdash R : \reg \sigma$, suppose $\var R = \{r_1,r_2,\cdots,r_n\}$, 
  we introduce variables $i_{r_k} : \Basis(\sigma_{r_k})$ with $r_k : \reg \sigma_{r_k} \in E$ for $k = 1,\cdots, n$. We reconstruct the basis $|i_R\>$ (which is of type $\KType(\sigma)$) and $\<i_R|$ (which is of type $\BType(\sigma)$) of $R$ by:
  \begin{itemize}
    \item $R = r_k$, $|i_R\> \triangleq |i_{r_k}\>$ and $\<i_R| \triangleq \<i_{r_k}|$;
    \item $R = (R_1,R_2)$: $|i_R\> \triangleq |i_{R_1}\> \otimes |i_{R_2}\>$ and $\<i_R| \triangleq \<i_{R_1}| \otimes \<i_{R_2}|$.
  \end{itemize}
    \item Expansion of all $K_R,B_R,O_{R,R'}$:
    \begin{align*}
      &K_R = \sum_{i_{r_1}\in\bU(\sigma_{r_1})}\cdots \sum_{i_{r_n}\in\bU(\sigma_{r_n})} (\<i_R|\cdot K). (|i_{r_1}\>_{r_i}\otimes\cdots\otimes|i_{r_n}\>_{r_n}).\\
      &B_R = \sum_{i_{r_1}\in\bU(\sigma_{r_1})}\cdots \sum_{i_{r_n}\in\bU(\sigma_{r_n})} (B\cdot |i_R\>). ({}_{r_1}\<i_{r_1}|\otimes\cdots\otimes{}_{r_n}\<i_{r_n}|).\\
      &O_{R,R'} = \sum_{i_{r_1}\in\bU(\sigma_{r_1})}\cdots \sum_{i_{r_n}\in\bU(\sigma_{r_n})}
      \sum_{i_{r'_1}\in\bU(\sigma_{r'_1})}\cdots \sum_{i_{r'_{n'}}\in\bU(\sigma_{r'_{n'}})} \\
      &\qquad\qquad\qquad (\<i_R|\cdot O\cdot |i_{R'}\>).\Big((|i_{r_1}\>_{r_i}\otimes\cdots\otimes|i_{r_n}\>_{r_n})\otimes({}_{r'_1}\<i_{r'_1}|\otimes\cdots\otimes{}_{r'_{n'}}\<i_{r'_{n'}}|)\Big).\\
    \end{align*}
    \item Rewriting to push big operators and split: \textbf{Beta-Arrow}, \textbf{Beta-Index}, \textbf{Delta}

\begin{ruletable}{The special flattening rule.}
    %
    R-ADJDK
    & $ ({}_r\<i|)^\dagger \ \reduce\ |i\>_r$ \\
    %
    R-ADJDB
    & $ (|i\>_r)^\dagger \ \reduce\ {}_r\<i|$ \\
    & \textcolor{gray}{Dealing with adjoint; now we don't have any $\tD^\dagger$ by combining push rules.} \\
    R-ADJD0
    & $ (\tD_1 \otimes \tD_2)^\dagger \ \reduce\ \tD_1^\dagger \otimes \tD_2^\dagger$ \\
    %
    R-ADJD1
    & $ (\tD_1\cdot \tD_2)^\dagger \ \reduce\ \tD_2^\dagger \cdot \tD_1^\dagger$ \\
    %
    R-SCRD0
    & $ D_1 \otimes \cdots \otimes (a.\tD_n) \otimes \cdots \otimes \tD_m \ \reduce\ a.(\tD_1 \otimes \cdots \otimes \tD_m) $ \\
    %
    R-SCRD1
    & $ (a.\tD_1) \cdot \tD_2 \ \reduce\ a.(\tD_1 \cdot \tD_2) $ \\
    %
    R-SCRD2
    & $ \tD_1 \cdot (a.\tD_2) \ \reduce\ a.(\tD_1 \cdot \tD_2) $ \\
    & \textcolor{gray}{Dealing with scaling} \\
    %
    R-TSRD0
    & $ (\tD_1 + \cdots + \tD_n) \otimes \tD\ \reduce\ \tD_1 \otimes \tD + \cdots + \tD_n \otimes \tD $ \\
    %
    R-DOTD0
    & $ (\tD_1 + \cdots + \tD_n) \cdot \tD\ \reduce\ \tD_1 \cdot \tD + \cdots + \tD_n \cdot \tD $ \\
    %
    R-DOTD1
    & $ \tD \cdot (\tD_1 + \cdots + \tD_n)\ \reduce\ \tD \cdot \tD_1 + \cdots + \tD \cdot \tD_n $ \\
    & \textcolor{gray}{Distributivity of $\cdot$ and $\otimes$. } \\
    %
    R-SUM-PUSHD0
    & $ (\sum_{i \in M} \tD_1) \otimes \tD_2 \ \reduce\ \sum_{i \in M} (\tD_1 \otimes \tD_2) $ \\
    %
    R-SUM-PUSHD2
    & $ (\sum_{i \in M} \tD_1) \cdot \tD_2 \ \reduce\ \sum_{i \in M} (\tD_1 \cdot \tD_2) $ \\
    %
    R-SUM-PUSHD3
    & $ \tD_1 \cdot (\sum_{i \in M} \tD_2) \ \reduce\ \sum_{i \in M} (\tD_1 \cdot \tD_2) $ \\
    %
    R-L-SORT0
    & $ A : \DType(s_1, s_2), B : \DType(s_1', s_2'), s_2 \cap s_1'=\emptyset \Rightarrow A \cdot B \ \reduce\ A \otimes B $ \\
    %
    R-L-SORT1
    & ${}_r\bra{i}\cdot\ket{j}_r \ \reduce\ \delta_{i, j}$ \\
    %
    R-L-SORT2
    & ${}_r\bra{i}\cdot(Y_1 \otimes \cdots \otimes \ket{j}_r \otimes \cdots \otimes Y_m) \ \reduce\ \delta_{i, j}.(Y_1  \otimes \cdots \otimes Y_m)$ \\
    %
    R-L-SORT3
    & $(X_1 \otimes \cdots \otimes {}_r\bra{i} \otimes \cdots \otimes X_n) \cdot \ket{j}_r \ \reduce\ \delta_{i,j}.(X_1 \otimes \cdots \otimes X_n)$ \\
    %
    R-L-SORT1
    & $ (X_1 \otimes \cdots \otimes {}_r\bra{i} \otimes \cdots \otimes X_n) \cdot (Y_1 \otimes \cdots \otimes \ket{j}_r \otimes \cdots \otimes Y_m) $ \\
    & $\reduce\ \delta_{i,j}.(X_1 \otimes \cdots \otimes X_n) \cdot (Y_1 \otimes \cdots \otimes Y_m)$
\end{ruletable}
    After the rewriting of this step, the expression we have is always in the following form:
    $$
    \Big(\sum_{i_1}\cdots\sum_{i_n} a.\tD_f\Big) + \cdots + \Big(\sum_{i_1}\cdots\sum_{i_{n'}} a.\tD_f\Big)
    $$
    where $\tD_f$ is generated by :
    $$\tD_f ::= |i\>_r \mid {}_r\<i| \mid \te\otimes\te \mid \te\cdot\te. $$
    \item Simplify $\tD_f$. Let $\cN$ be the variable name set. For every $\tD_f$, we define three functions $\mathsf{Can}(\tD_f) : \List\ (\cN\times\cN)$, $\mathsf{CDom}(\tD_f), \mathsf{Dom}(\tD_f) : \List\ (\cN\times \cR)$ that compute the cancellation pairs of variables, codomain and domain.
    They are inductively defined by the structure of $\tD_f$:
    \begin{itemize}
        \item $\tD_f \equiv |i\>_r$: $\mathsf{Can}(\tD_f) = []$, $\mathsf{CDom}(\tD_f) = [(i,r)]$, $\mathsf{Dom}(\tD_f) = []$.
        \item $\tD_f \equiv {}_r\<i|$: $\mathsf{Can}(\tD_f) = []$, $\mathsf{CDom}(\tD_f) = []$, $\mathsf{Dom}(\tD_f) = [(i,r)]$.
        \item $\tD_f \equiv \tD_{1f}\otimes\tD_{2f}$: $\mathsf{Can}(\tD_f) = \mathsf{Can}(\tD_{1f}) ; \mathsf{Can}(\tD_{2f})$, $\mathsf{CDom}(\tD_f) = \mathsf{CDom}(\tD_{1f}) ; \mathsf{CDom}(\tD_{2f})$, $\mathsf{Dom}(\tD_f) = \mathsf{Dom}(\tD_{1f}) ; \mathsf{Dom}(\tD_{2f})$. (Here, $;$ is the concatenation of lists)
        \item $\tD_f \equiv \tD_{1f}\cdot\tD_{2f}$: recursively execute:
        \begin{enumerate}
            \item Set $L_{\rm can} = []$, $L_{\rm dom} = \mathsf{Dom}(\tD_{1f})$, $L_{\rm cdom} = \mathsf{CDom}(\tD_{2f})$.
            \item If there exists $(i_1,r)\in L_{\rm dom}$ and $(i_2,r)\in L_{\rm cdom}$, then $L_{\rm can} = L_{\rm can}; (i_1,i_2)$, $L_{\rm dom} = L_{\rm dom}\backslash(i_1,r)$ and $L_{\rm cdom} = L_{\rm cdom}\backslash(i_2,r)$. (Here, $\backslash$ means to remove the element from the list.)
            \item Repeat step (b) until it fails.
        \end{enumerate}
        Then, 
        $\mathsf{Can}(\tD_f) = \mathsf{Can}(\tD_{1f}) ; \mathsf{Can}(\tD_{2f}); L_{\rm can}$, 
        $\mathsf{CDom}(\tD_f) = \mathsf{CDom}(\tD_{1f}) ; L_{\rm cdom}$, $\mathsf{Dom}(\tD_f) = \mathsf{Dom}(\tD_{f}) ; L_{\rm dom}$. 
    \end{itemize}
    Suppose $\mathsf{CDom}(\tD_{f}) = \{(i_1,r_1),\cdots,(i_n,r_n)\}$ and $\mathsf{Dom}(\tD_{f}) = \{(i'_1,r'_1),\cdots,(i'_{n'},r'_{n'})\}$ where the list is sorted by a default order of variables in $\cR$. We define:
    $$D_f = \bigg(\prod_{(i,j)\in \mathsf{Can}(\tD_f)}\delta_{i,j}\bigg).\Big((|i_1\>\otimes\cdots\otimes|i_n\>) \cdot (\<i'_1|\otimes\cdots\otimes\<i'_{n'}|)\Big).$$
    If $\mathsf{Dom}$ or $\mathsf{CDom}$ is an empty list, the $D_f$ is just a Ket or Bra.
    Then we claim:
    \begin{itemize}
        \item If $E[\Gamma]\vdash \tD_f : \DType(s_1,s_2)$, then $\{r_1,\cdots,r_n\} = s_1$ and
        $\{r'_1,\cdots,r'_{n'}\} = s_2$.
        \item Set $R = ((\cdots(r_1,r_2),\cdots),r_n)$ and $R' = ((\cdots(r'_1,r'_2),\cdots),r'_{n'})$, then
        $$\tD_{f} = (D_f)_{R,R'}$$
    \end{itemize}
    \item For the obtained formula in (3), we replace each $\tD_f$ by $D_f$ which is computed by step (4). Note that $D_f$ is a Dirac term without labels, so the equivalence of two labelled Dirac terms is reduced to decide the equivalence of two Dirac terms. This is guaranteed if two labelled Dirac terms are of the same type (by typing rules).
\end{enumerate}

### 2025-01-13 main.tex
integrate Li's notes on labelled, change to LNCS format

@@
-\documentclass{article}
-
-\usepackage[utf8]{inputenc}
-\usepackage[margin=0.8in]{geometry}
+% This is samplepaper.tex, a sample chapter demonstrating the
+% LLNCS macro package for Springer Computer Science proceedings;
+% Version 2.21 of 2022/01/12
+%
+\documentclass[runningheads]{llncs}
+
+% \usepackage[utf8]{inputenc}
+% \usepackage[margin=0.8in]{geometry}
 
@@
 \usepackage{graphicx}
-\usepackage{amsthm}
+% \usepackage{amsthm}
 \usepackage{stmaryrd}
@@
 \newcommand{\yx}[1]{\textit{\color{blue}[YX] : #1}}
+\newcommand{\lz}[1]{\textit{\color{red}[LZ] : #1}}
 
@@
 
-\title{\textbf{DiracDec C++/Coq Implementation\cite{ExampleCitation}}}
-
+%
+\usepackage[T1]{fontenc}
+% T1 fonts will be used to generate the final print and online PDFs,
+% so please use T1 fonts in your manuscript whenever possible.
+% Other font encondings may result in incorrect characters.
+%
+\usepackage{graphicx}
+% Used for displaying a sample figure. If possible, figure files should
+% be included in EPS format.
+%
+% If you use the hyperref package, please uncomment the following two lines
+% to display URLs in blue roman font according to Springer's eBook style:
+\usepackage{color}
+\renewcommand\UrlFont{\color{blue}\rmfamily}
+\urlstyle{rm}
+%
 \begin{document}
-
-\maketitle
+%
+\title{DiracDec C++/Coq Implementation\cite{ExampleCitation}}
+
+%
+%\titlerunning{Abbreviated paper title}
+% If the paper title is too long for the running head, you can set
+% an abbreviated paper title here
+%
+\author{First Author\inst{1}\orcidID{0000-1111-2222-3333} \and
+Second Author\inst{2,3}\orcidID{1111-2222-3333-4444} \and
+Third Author\inst{3}\orcidID{2222--3333-4444-5555}}
+%
+\authorrunning{F. Author et al.}
+% First names are abbreviated in the running head.
+% If there are more than two authors, 'et al.' is used.
+%
+\institute{Princeton University, Princeton NJ 08544, USA \and
+Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
+\email{lncs@springer.com}\\
+\url{http://www.springer.com/gp/computer-science/lncs} \and
+ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
+\email{\{abc,lncs\}@uni-heidelberg.de}}
+%
+\maketitle              % typeset the header of the contribution
+%
+\begin{abstract}
+The abstract should briefly summarize the contents of the paper in
+150--250 words.
+
+\keywords{First keyword  \and Second keyword \and Another keyword.}
+\end{abstract}
+%
+%
+%
 
@@
     \item I guess it may still be necessary to try all different bound variable assignments.
+    \item the $\bar{U_{AC}}$, $\bar{U_{BC}}$ terms can be modelled by quantum registers.
 \end{itemize}
@@
     \begin{align*}
-        \sigma ::=\ & x\ |\ \sigma_1 \times \sigma_2\ |\ \mathsf{Qdit}_n.
+        \sigma ::=\ & x \mid \sigma_1 \times \sigma_2 \mid \mathsf{Qdit}_n.
     \end{align*}
@@
     \begin{align*}
-        T ::=\ & x\ |\ \Basis(\sigma)\ |\ \SType\ |\ \KType(\sigma)\ |\ \BType(\sigma)\ |\ \OType(\sigma_1, \sigma_2)\ |\ T_1 \to T_2\ |\ \forall x.T\ |\ \SET(\sigma). \\
+        T ::=\ & x \mid \Basis(\sigma) \mid \SType \mid \KType(\sigma) \mid \BType(\sigma) \mid \OType(\sigma_1, \sigma_2) \mid T_1 \to T_2 \mid \forall x.T \mid \SET(\sigma). \\
     \end{align*}
@@
     \begin{align*}
-        e ::=\ & x\ |\ (e_1, e_2)\ |\ \lambda x : T.e\ |\ \mu x.e\ |\ e_1\ e_2 \\
-        & |\ 0\ |\ 1\ |\ \ADDS(e_1 \cdots e_n) \ |\ e_1 \times \cdots \times e_n\ |\ e^* \ |\ \delta_{e_1, e_2}\ |\ \DOT(e_1\ e_2) \\
-        & |\ \ZEROK(\sigma)\ |\ \ZEROB(\sigma)\ |\ \ZEROO(\sigma_1, \sigma_2)\ |\ \ONEO(\sigma) \\
-        & |\ \ket{e}\ |\ \bra{t}\ |\ e^\dagger\ |\ e_1.e_2\ |\ \ADD(e_1 \cdots e_n)\ |\ e_1 \otimes e_2 \\
-        & |\ \MULK(e_1\ e_2)\ |\ \MULB(e_1\ e_2)\ |\ \OUTER(e_1\ e_2)\ |\ \MULO(e_1\ e_2) \\
-        & |\ \mathbf{U}(e)\ |\ e_1 \star e_2\ |\ \sum_{e_1} e_2.
+        e ::=\ & x \mid (e_1, e_2) \mid \lambda x : T.e \mid \mu x.e \mid e_1\ e_2 \\
+        & |\ 0 \mid 1 \mid \ADDS(e_1 \cdots e_n)  \mid e_1 \times \cdots \times e_n \mid e^*  \mid \delta_{e_1, e_2} \mid \DOT(e_1\ e_2) \\
+        & |\ \ZEROK(\sigma) \mid \ZEROB(\sigma) \mid \ZEROO(\sigma_1, \sigma_2) \mid \ONEO(\sigma) \\
+        & |\ \ket{e} \mid \bra{t} \mid e^\dagger \mid e_1.e_2 \mid \ADD(e_1 \cdots e_n) \mid e_1 \otimes e_2 \\
+        & |\ \MULK(e_1\ e_2) \mid \MULB(e_1\ e_2) \mid \OUTER(e_1\ e_2) \mid \MULO(e_1\ e_2) \\
+        & |\ \mathbf{U}(e) \mid e_1 \star e_2 \mid \sum_{e_1} e_2.
     \end{align*}
@@
     \begin{align*}
-        E ::=\ & []\ |\ E; x : \Index\ |\ E; T : \Type\ |\ E; x : T\ |\ E; x := t : T. \\
-        \Gamma ::=\ & []\ |\ \Gamma; x : \Index\ |\ \Gamma; x : T.
+        E ::=\ & [] \mid E; x : \Index \mid E; T : \Type \mid E; x : T \mid E; x := t : T. \\
+        \Gamma ::=\ & [] \mid \Gamma; x : \Index \mid \Gamma; x : T.
     \end{align*}
@@
     & $ \reduce\ \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (a_1\{j/i\} \times \cdots \times a_n\{j/i\}).A\{j/i\} $ \\
+    %
+    R-SUM-ELIM8
+    & $ \sum_{i \in M} \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} ((a_1 \times \cdots \times \delta_{i, j} \times \cdots \times a_n) + \cdots + (b_1 \times \cdots \times \delta_{i, j} \times \cdots \times b_n)).A $ \\
+    & $ \reduce\ \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} ((a_1\{j/i\} \times \cdots \times a_n\{j/i\}) + \cdots + (b_1\{j/i\} \times \cdots \times b_n\{j/i\})).A\{j/i\} $ \\
 \end{ruletable}
@@
     & $\sum_{i \in M}(a_1 + \cdots + a_n) \ \reduce\ (\sum_{i \in M} a_1) + \cdots + (\sum_{i \in M} a_n) $ \\
+    % %
+    % R-SUM-ADDS1
+    % & $\sum_{i \in M}(b_1 \times \cdots \times (a_1 + \cdots + a_n) \times \cdots \times b_m) $ \\
+    % & $ \reduce\ (\sum_{i \in M} (b_1 \times \cdots \times a_1 \times \cdots \times b_m) + \cdots + (\sum_{i \in M} (b_1 \times \cdots \times a_n \times \cdots \times b_m)) $ \\
     %
@@
     & $\sum_{i \in M}(X_1 + \cdots + X_n) \ \reduce\ (\sum_{i \in M} X_1) + \cdots + (\sum_{i \in M} X_n) $ \\
-    %
     R-SUM-ADD1
-    & $Y_1 + \cdots + \sum_{i \in M}(a.X) + \cdots + \sum_{i \in M}X + Y_n \ \reduce\ Y_1 + \cdots + Y_n + \sum_{i \in M}(a+1).X $ \\
-    %
-    R-SUM-ADD2
-    & $Y_1 + \cdots + \sum_{i \in M}X + \cdots + \sum_{i \in M}(a.X) + Y_n \ \reduce\ Y_1 + \cdots + Y_n + \sum_{i \in M}(1+a).X $ \\
-    %
-    R-SUM-ADD3
-    & $Y_1 + \cdots + \sum_{i \in M}(a.X) + \cdots + \sum_{i \in M}(b.X) + Y_n \ \reduce\ Y_1 + \cdots + Y_n + \sum_{i \in M}(a+b).X $ \\
+    & $ Y_1 + \cdots + Y_n + \sum_{i \in M}(a+b).X \ \reduce\ Y_1 + \cdots + \sum_{i \in M}(a.X) + \cdots + \sum_{i \in M}(b.X) + Y_n$ \\
     %
@@
 \yx{The rules R-SUM-ADD1 to R-SUM-ADD3 needs identical sum terms, which requires the rewriting to be after the alpha normalization. We don't implement them for now.}
+
+\begin{ruletable}{Rules for Qubit}
+    %
+    R-QBIT-DELTA
+    & $\delta_{0, 1} \ \reduce\ 0$ \\
+    %
+    R-QBIT-ONEO
+    & $\ONEO(QBIT)\ \reduce\ \ket{0}\bra{0} + \ket{1}\bra{1} $ \\
+    %
+    R-QBIT-SUM
+    & $\sum_{i \in \mathbf{U}(QBIT)} A\ \reduce\ A\{i/0\} + A\{i/1\}$
+\end{ruletable}
 
@@
         &|\ \texttt{ShowAll} \\
-        &|\ \texttt{Normalize}(term)\ |\ \texttt{Normalize}(term\ \textsf{Trace}) \\
+        &|\ \texttt{Normalize}(term) \mid \texttt{Normalize}(term\ \textsf{Trace}) \\
         &|\ \texttt{CheckEq}(term\ term) \\
-    type ::=&\ \texttt{Type}\ |\ \texttt{Arrow}(type\ type) \\
+    type ::=&\ \texttt{Type} \mid \texttt{Arrow}(type\ type) \\
             &|\ \texttt{Base} \\
-    term ::=&\ \texttt{Type}\ |\ \texttt{fun}(\texttt{ID}\ type\ term)\ |\ \texttt{apply}(term\ term)\ |\ \texttt{ID} \\
+    term ::=&\ \texttt{Type} \mid \texttt{fun}(\texttt{ID}\ type\ term) \mid \texttt{apply}(term\ term) \mid \texttt{ID} \\
 \end{align*}
@@
     \[
-    P ::= \mathsf{E}\ |\ [(i:P)^+].
+    P ::= \mathsf{E} \mid [(i:P)^+].
     \]
@@
 
-\input{labelled_dirac.tex}
-
-\clearpage
-
-\bibliographystyle{plain}
+\input{labelled}
+%\input{labelled_dirac.tex}
+
+\begin{credits}
+    \subsubsection{\ackname} A bold run-in heading in small font size at the end of the paper is
+    used for general acknowledgments, for example: This study was funded
+    by X (grant number Y).
+
+    \subsubsection{\discintname}
+    It is now necessary to declare any competing interests or to specifically
+    state that the authors have no competing interests. Please place the
+    statement with a bold run-in heading in small font size beneath the
+    (optional) acknowledgments\footnote{If EquinOCS, our proceedings submission
+    system, is used, then the disclaimer can be provided directly in the system.},
+    for example: The authors have no competing interests to declare that are
+    relevant to the content of this article. Or: Author A has received research
+    grants from Company W. Author B has received a speaker honorarium from
+    Company X and owns stock in Company Y. Author C is a member of committee Z.
+\end{credits}
+
+%
+% ---- Bibliography ----
+%
+% BibTeX users should specify bibliography style 'splncs04'.
+% References will then be sorted and formatted in the correct style.
+%
+\bibliographystyle{splncs04}
 \bibliography{ref}
-
+%
 \end{document}

### 2025-01-14 flowchart.tex
2025/1/14

\tikzstyle{rewriting} = [rectangle, rounded corners, minimum width=3cm, minimum height=0.8cm,text centered, draw=black, fill=red!30]
\tikzstyle{process} = [rectangle, minimum width=3cm, minimum height=0.8cm, text centered, draw=black, fill=blue!30]
\tikzstyle{decision} = [diamond, minimum width=3cm, minimum height=0.8cm, text centered, draw=black, fill=green!30]
\tikzstyle{arrow} = [thick,->,>=stealth]

\begin{figure}
    \center
    \begin{tikzpicture}[node distance=1.2cm]

    % \node (rename) [process] {rename unique bound variable names};
    \node (rewrite1) [rewriting] {first rewritings};
    \node (expansion) [process, below of=rewrite1] {variable expansion};
    \node (rewrite2) [rewriting, below of=expansion] {second rewritings};
    \node (sorting) [process, below of=rewrite2] {sorting modulo bound variables};
    \node (successive) [process, below of=sorting] {sorting successive summation};
    \node (deBruijn) [process, below of=successive] {de Bruijn normalization};

    % \draw [arrow] (rename) -- (rewrite1);
    \draw [arrow] (rewrite1) -- (expansion);
    \draw [arrow] (expansion) -- (rewrite2);
    \draw [arrow] (rewrite2) -- (sorting);
    \draw [arrow] (sorting) -- (successive);
    \draw [arrow] (successive) -- (deBruijn);

    \end{tikzpicture}
    \caption{A flowchart for normalization of Dirac notations.}
    \label{fig: normalization}
\end{figure}

### 2025-01-14 labelled.tex
2025/1/14

@@
 
-\begin{gather}
-  \textbf{W-AssumE-Reg}\qquad 
-  \frac{E[]\vdash \sigma : \Index\qquad r\notin E}{\WF[E; r : \reg(\sigma)[]]}
-\end{gather}
-
-\begin{gather*}
-  \textbf{Type-Labelled} \qquad
-  \frac{E[\Gamma] \vdash r : \reg(\sigma_r) \text{ for all $r$ in $s_1$ and $s_2$} }{E[\Gamma] \vdash \DType(s_1, s_2) : \textsf{Type}} \\
-  \\
-  \textbf{Reg-Var} \qquad
-  \frac{\WF(E[\Gamma]) \qquad r : \reg(\sigma) \in E}{E[\Gamma] \vdash r : \reg(\sigma)}  \\ 
-  \\
-  \textbf{Reg-Pair}\qquad
-  \frac{E[\Gamma] \vdash R : \reg \sigma\qquad E[\Gamma] \vdash Q : \reg \tau\qquad R\| Q}{E[\Gamma] \vdash (R,Q) : \reg \sigma \times \tau}
-\end{gather*}
-
-\begin{gather*}
-    \textbf{Type-L-Base-Ket}\qquad 
-    \frac{r : \reg \sigma \in E\qquad E[\Gamma] \vdash i : \Basis(\sigma)}{E[\Gamma] \vdash |i\>_r : \DType(\{r\}; \emptyset)} \\
-    \textbf{Type-L-Base-Bra}\qquad 
-    \frac{r : \reg \sigma \in E \qquad E[\Gamma] \vdash i : \Basis(\sigma)}{E[\Gamma] \vdash {}_r\<i| : \DType(\emptyset; \{r\})} \\
-    \textbf{Type-L-Ket}\qquad 
-    \frac{E[\Gamma] \vdash R : \reg \sigma\qquad E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash K_R : \DType(\var R; \emptyset)} \\
-    \textbf{Type-L-Bra}\qquad 
-    \frac{E[\Gamma] \vdash R : \reg \sigma\qquad E[\Gamma] \vdash B : \BType(\sigma)}{E[\Gamma] \vdash B_R : \DType(\emptyset; \var R)} \\
-    \textbf{Type-L-Opt}\qquad 
-    \frac{E[\Gamma] \vdash R_1 : \reg \sigma_1\qquad 
-    E[\Gamma] \vdash R_2 : \reg \sigma_2\qquad 
-    E[\Gamma] \vdash O : \OType(\sigma_1; \sigma_2)}
-    {E[\Gamma] \vdash O_{R_1;R_2} : \DType(\var R_1; \var R_2)} \\
-    \textbf{Type-L-Conj}\qquad 
-    \frac{E[\Gamma] \vdash \tD : \DType(s_1;s_2)}{E[\Gamma] \vdash \tD^\dagger : \DType(s_2;s_1)} \qquad
-    \textbf{Type-L-SCL}\qquad 
-    \frac{E[\Gamma] \vdash S : \SType\qquad E[\Gamma] \vdash \tD : \DType(s_1;s_2)}{E[\Gamma] \vdash S.\tD : \DType(s_1;s_2)} \\
-    \textbf{Type-L-ADD}\qquad
-    \frac{E[\Gamma] \vdash \tD_i : \DType(s_1;s_2)\quad \text{forall } i}{E[\Gamma] \vdash \tD_1+\cdots+\tD_n : \DType(s_1;s_2)} \\
-    \textbf{Type-L-TSR}\qquad
-    \frac{E[\Gamma] \vdash \tD_1 : \DType(s_1;s_1')\qquad E[\Gamma] \vdash \tD_2 : \DType(s_2;s_2')\qquad s_1\cap s_2 = \emptyset\qquad s_1'\cap s_2' = \emptyset}{E[\Gamma] \vdash \tD_1\otimes\tD_2 : \DType(s_1\cup s_2;s_1'\cup s_2')} \\
-    \textbf{Type-L-DOT}\qquad
-    \frac{E[\Gamma] \vdash \tD_1 : \DType(s_1;s_1')\qquad E[\Gamma] \vdash \tD_2 : \DType(s_2;s_2')\qquad 
-    s_1 \cap s_2 \backslash s_1' = \emptyset \qquad
-    s_2' \cap s_1' \backslash s_2 = \emptyset}
-    {E[\Gamma] \vdash \tD_1\cdot\tD_2 : \DType(s_1 \cup (s_2\backslash s_1'); s_2' \cup (s_1'\backslash s_2))}
-\end{gather*}
-
 \subsubsection*{Eliminating the registers}
@@
   \end{itemize}
-    \item Expansion of all $K_R,B_R,O_{R,R'}$:
-    \begin{align*}
-      &K_R = \sum_{i_{r_1}\in\bU(\sigma_{r_1})}\cdots \sum_{i_{r_n}\in\bU(\sigma_{r_n})} (\<i_R|\cdot K). (|i_{r_1}\>_{r_i}\otimes\cdots\otimes|i_{r_n}\>_{r_n}).\\
-      &B_R = \sum_{i_{r_1}\in\bU(\sigma_{r_1})}\cdots \sum_{i_{r_n}\in\bU(\sigma_{r_n})} (B\cdot |i_R\>). ({}_{r_1}\<i_{r_1}|\otimes\cdots\otimes{}_{r_n}\<i_{r_n}|).\\
-      &O_{R,R'} = \sum_{i_{r_1}\in\bU(\sigma_{r_1})}\cdots \sum_{i_{r_n}\in\bU(\sigma_{r_n})}
-      \sum_{i_{r'_1}\in\bU(\sigma_{r'_1})}\cdots \sum_{i_{r'_{n'}}\in\bU(\sigma_{r'_{n'}})} \\
-      &\qquad\qquad\qquad (\<i_R|\cdot O\cdot |i_{R'}\>).\Big((|i_{r_1}\>_{r_i}\otimes\cdots\otimes|i_{r_n}\>_{r_n})\otimes({}_{r'_1}\<i_{r'_1}|\otimes\cdots\otimes{}_{r'_{n'}}\<i_{r'_{n'}}|)\Big).\\
-    \end{align*}
-    \item Rewriting to push big operators and split: \textbf{Beta-Arrow}, \textbf{Beta-Index}, \textbf{Delta}
+    \item Expansion of all $K_R,B_R,O_{R,R'}$
 
-\begin{ruletable}{The special flattening rule.}
-    %
-    R-ADJDK
-    & $ ({}_r\<i|)^\dagger \ \reduce\ |i\>_r$ \\
-    %
-    R-ADJDB
-    & $ (|i\>_r)^\dagger \ \reduce\ {}_r\<i|$ \\
-    & \textcolor{gray}{Dealing with adjoint; now we don't have any $\tD^\dagger$ by combining push rules.} \\
-    R-ADJD0
-    & $ (\tD_1 \otimes \tD_2)^\dagger \ \reduce\ \tD_1^\dagger \otimes \tD_2^\dagger$ \\
-    %
-    R-ADJD1
-    & $ (\tD_1\cdot \tD_2)^\dagger \ \reduce\ \tD_2^\dagger \cdot \tD_1^\dagger$ \\
-    %
-    R-SCRD0
-    & $ D_1 \otimes \cdots \otimes (a.\tD_n) \otimes \cdots \otimes \tD_m \ \reduce\ a.(\tD_1 \otimes \cdots \otimes \tD_m) $ \\
-    %
-    R-SCRD1
-    & $ (a.\tD_1) \cdot \tD_2 \ \reduce\ a.(\tD_1 \cdot \tD_2) $ \\
-    %
-    R-SCRD2
-    & $ \tD_1 \cdot (a.\tD_2) \ \reduce\ a.(\tD_1 \cdot \tD_2) $ \\
-    & \textcolor{gray}{Dealing with scaling} \\
-    %
-    R-TSRD0
-    & $ (\tD_1 + \cdots + \tD_n) \otimes \tD\ \reduce\ \tD_1 \otimes \tD + \cdots + \tD_n \otimes \tD $ \\
-    %
-    R-DOTD0
-    & $ (\tD_1 + \cdots + \tD_n) \cdot \tD\ \reduce\ \tD_1 \cdot \tD + \cdots + \tD_n \cdot \tD $ \\
-    %
-    R-DOTD1
-    & $ \tD \cdot (\tD_1 + \cdots + \tD_n)\ \reduce\ \tD \cdot \tD_1 + \cdots + \tD \cdot \tD_n $ \\
-    & \textcolor{gray}{Distributivity of $\cdot$ and $\otimes$. } \\
-    %
-    R-SUM-PUSHD0
-    & $ (\sum_{i \in M} \tD_1) \otimes \tD_2 \ \reduce\ \sum_{i \in M} (\tD_1 \otimes \tD_2) $ \\
-    %
-    R-SUM-PUSHD2
-    & $ (\sum_{i \in M} \tD_1) \cdot \tD_2 \ \reduce\ \sum_{i \in M} (\tD_1 \cdot \tD_2) $ \\
-    %
-    R-SUM-PUSHD3
-    & $ \tD_1 \cdot (\sum_{i \in M} \tD_2) \ \reduce\ \sum_{i \in M} (\tD_1 \cdot \tD_2) $ \\
-    %
-    R-L-SORT0
-    & $ A : \DType(s_1, s_2), B : \DType(s_1', s_2'), s_2 \cap s_1'=\emptyset \Rightarrow A \cdot B \ \reduce\ A \otimes B $ \\
-    %
-    R-L-SORT1
-    & ${}_r\bra{i}\cdot\ket{j}_r \ \reduce\ \delta_{i, j}$ \\
-    %
-    R-L-SORT2
-    & ${}_r\bra{i}\cdot(Y_1 \otimes \cdots \otimes \ket{j}_r \otimes \cdots \otimes Y_m) \ \reduce\ \delta_{i, j}.(Y_1  \otimes \cdots \otimes Y_m)$ \\
-    %
-    R-L-SORT3
-    & $(X_1 \otimes \cdots \otimes {}_r\bra{i} \otimes \cdots \otimes X_n) \cdot \ket{j}_r \ \reduce\ \delta_{i,j}.(X_1 \otimes \cdots \otimes X_n)$ \\
-    %
-    R-L-SORT1
-    & $ (X_1 \otimes \cdots \otimes {}_r\bra{i} \otimes \cdots \otimes X_n) \cdot (Y_1 \otimes \cdots \otimes \ket{j}_r \otimes \cdots \otimes Y_m) $ \\
-    & $\reduce\ \delta_{i,j}.(X_1 \otimes \cdots \otimes X_n) \cdot (Y_1 \otimes \cdots \otimes Y_m)$
-\end{ruletable}
     After the rewriting of this step, the expression we have is always in the following form:
@@
     $$\tD_f ::= |i\>_r \mid {}_r\<i| \mid \te\otimes\te \mid \te\cdot\te. $$
-    \item Simplify $\tD_f$.
-Let $\cN$ be the variable name set.
-For every $\tD_f$, we define three functions $\mathsf{Can}(\tD_f) : \List\ (\cN\times\cN)$, $\mathsf{CDom}(\tD_f), \mathsf{Dom}(\tD_f) : \List\ (\cN\times \cR)$ that compute the cancellation pairs of variables, codomain and domain.
-    They are inductively defined by the structure of $\tD_f$:
-    \begin{itemize}
-        \item $\tD_f \equiv |i\>_r$: $\mathsf{Can}(\tD_f) = []$, $\mathsf{CDom}(\tD_f) = [(i,r)]$, $\mathsf{Dom}(\tD_f) = []$.
-        \item $\tD_f \equiv {}_r\<i|$: $\mathsf{Can}(\tD_f) = []$, $\mathsf{CDom}(\tD_f) = []$, $\mathsf{Dom}(\tD_f) = [(i,r)]$.
-        \item $\tD_f \equiv \tD_{1f}\otimes\tD_{2f}$: $\mathsf{Can}(\tD_f) = \mathsf{Can}(\tD_{1f}) ; \mathsf{Can}(\tD_{2f})$, $\mathsf{CDom}(\tD_f) = \mathsf{CDom}(\tD_{1f}) ; \mathsf{CDom}(\tD_{2f})$, $\mathsf{Dom}(\tD_f) = \mathsf{Dom}(\tD_{1f}) ; \mathsf{Dom}(\tD_{2f})$.
-(Here, $;$ is the concatenation of lists)
-        \item $\tD_f \equiv \tD_{1f}\cdot\tD_{2f}$: recursively execute:
-        \begin{enumerate}
-            \item Set $L_{\rm can} = []$, $L_{\rm dom} = \mathsf{Dom}(\tD_{1f})$, $L_{\rm cdom} = \mathsf{CDom}(\tD_{2f})$.
-            \item If there exists $(i_1,r)\in L_{\rm dom}$ and $(i_2,r)\in L_{\rm cdom}$, then $L_{\rm can} = L_{\rm can}; (i_1,i_2)$, $L_{\rm dom} = L_{\rm dom}\backslash(i_1,r)$ and $L_{\rm cdom} = L_{\rm cdom}\backslash(i_2,r)$.
-(Here, $\backslash$ means to remove the element from the list.)
-            \item Repeat step (b) until it fails.
-        \end{enumerate}
-        Then, 
-        $\mathsf{Can}(\tD_f) = \mathsf{Can}(\tD_{1f}) ; \mathsf{Can}(\tD_{2f}); L_{\rm can}$, 
-        $\mathsf{CDom}(\tD_f) = \mathsf{CDom}(\tD_{1f}) ; L_{\rm cdom}$, $\mathsf{Dom}(\tD_f) = \mathsf{Dom}(\tD_{f}) ; L_{\rm dom}$. 
-    \end{itemize}
+    \item
     Suppose $\mathsf{CDom}(\tD_{f}) = \{(i_1,r_1),\cdots,(i_n,r_n)\}$ and $\mathsf{Dom}(\tD_{f}) = \{(i'_1,r'_1),\cdots,(i'_{n'},r'_{n'})\}$ where the list is sorted by a default order of variables in $\cR$.

### 2025-01-14 main.tex
2025/1/14

@@
 
-\input{notations}
-
 % add new commands for comments here
@@
 \usepackage{braket}
+\usepackage{cleveref}
 
-% framed environment
-\newmdenv[
-  skipabove=5pt,
-  skipbelow=5pt,
-  leftmargin=0pt,
-  rightmargin=0pt,
-  innerleftmargin=5pt,
-  innerrightmargin=5pt,
-  innertopmargin=5pt,
-  innerbottommargin=5pt,
-  linecolor=black,
-]{framedeq}
+\usepackage{tikz}
+\usetikzlibrary{shapes.geometric, arrows}
+
+\newenvironment{ruletable}[1]
+{
+    \begin{longtable}{cl}
+    \caption{#1}\\
+    \hline
+    \textbf{Rule} & \textbf{Description} \\
+    \hline
+    \endfirsthead
+
+    \hline
+    \textbf{Rule} & \textbf{Description} \\
+    \hline
+    \endhead
+
+    % \hline
+    % \multicolumn{2}{r}{\textit{Continued on the next page}} \\
+    \hline
+    \endfoot
+
+    \hline
+    \endlastfoot
+}
+{
+    \end{longtable}
+}
 
@@
 %
-\title{DiracDec C++/Coq Implementation\cite{ExampleCitation}}
+\title{Dirace: Practical Proof Automation of Dirac Notation Equations}
 
-%
-%\titlerunning{Abbreviated paper title}
-% If the paper title is too long for the running head, you can set
-% an abbreviated paper title here
-%
-\author{First Author\inst{1}\orcidID{0000-1111-2222-3333} \and
-Second Author\inst{2,3}\orcidID{1111-2222-3333-4444} \and
-Third Author\inst{3}\orcidID{2222--3333-4444-5555}}
-%
-\authorrunning{F. Author et al.}
-% First names are abbreviated in the running head.
-% If there are more than two authors, 'et al.' is used.
-%
-\institute{Princeton University, Princeton NJ 08544, USA \and
-Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
-\email{lncs@springer.com}\\
-\url{http://www.springer.com/gp/computer-science/lncs} \and
-ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
-\email{\{abc,lncs\}@uni-heidelberg.de}}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% %
+% %\titlerunning{Abbreviated paper title}
+% % If the paper title is too long for the running head, you can set
+% % an abbreviated paper title here
+% %
+% \author{First Author\inst{1}\orcidID{0000-1111-2222-3333} \and
+% Second Author\inst{2,3}\orcidID{1111-2222-3333-4444} \and
+% Third Author\inst{3}\orcidID{2222--3333-4444-5555}}
+% %
+% \authorrunning{F. Author et al.}
+% % First names are abbreviated in the running head.
+% % If there are more than two authors, 'et al.' is used.
+% %
+% \institute{Princeton University, Princeton NJ 08544, USA \and
+% Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
+% \email{lncs@springer.com}\\
+% \url{http://www.springer.com/gp/computer-science/lncs} \and
+% ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
+% \email{\{abc,lncs\}@uni-heidelberg.de}}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 %
@@
 \begin{abstract}
-The abstract should briefly summarize the contents of the paper in
-150--250 words.
 
-\keywords{First keyword  \and Second keyword \and Another keyword.}
+Dirac notations are widely used in theoretical reasonings of quantum computation and quantum information.
+The previous work proved that the equivalence of basic Dirac notations is decidable, and provided a
+corresponding term rewriting system.
+This work focuses on developing the theory and tool that practically solves this problem. Compared to the
+previous result, this work includes a typing system and a simplified term rewriting system.
+A much more efficient algorithm.
+And a language and algorithm for labelled Dirac notation is proposed as the final contribution.
+We implement the whole algorithm in C++ with a Mathematica backend for theories of scalars, which 
+decides the 158 examples in seconds.
+
+% \keywords{First keyword  \and Second keyword \and Another keyword.}
 \end{abstract}
@@
 %%%%%%%%%%%%%%%%
+\newcommand*{\sem}[1]{{\llbracket #1 \rrbracket}}
+\newcommand{\DiracDec}{\textsf{DiracDec}}
 
@@
 
+\section{Introduction}
+
+In 1939, Dirac proposed his notation~\cite{dirac1939new} for quantum mechanics, which is designed to represent linear algebra formulae in a compact and convenient way.
+For instance, $a\ket{\psi} + b\ket{\phi}$ indicates the addition of two vectors, i.e., the superposition of two states $\ket{\psi}$ and $\ket{\phi}$.
+Dirac notation is also widely accepted as the working language in quantum computation and quantum information.
+The reasonings of Dirac notations play a fundamental role, but there was few works on automating this procedure.
+
+The absence of a working Dirac notation solver is also a main obstacle for proof automation of quantum programming languages.
+In these works, Dirac notation are used to define the program states, operations and assertions. In order to automate the verification procedure,
+we need to simplify and check the equivalence of pre-conditions.
+
+Recently, the work by Yingte et al.[] provides a theory to decide the equivalence of Dirac notations, as well as a prototype implementation in Mathematica.
+They proved that the equivalence of basic Dirac notations are decidable.
+Their system sticks with a pure term rewriting system, which allows to prove important properties such as confluence and termination.
+Even though, there is still a gap to a practical solver for Dirac notation equivalence.
+The Mathematica implementation is hard to be integrated into other tools. 
+The second concern is the efficiency. To decide the theory $E$, \DiracDec\ iterates all possible permutations, which is simple and direct but has factorial complexity.
+And lastly, the previous work does not consider labelled Dirac notation.
+The original work cannot intergrate with well with the Scalar system in mathematica.
+And there were a few examples that failded to decide.
+
+This work refines the pure term-rewriting based system into a hybrid decision procedure,
+overcomes the problems mentioned above and focuses on practical
+automated equational reasonings. Our main contributions are:
+\begin{itemize}
+    \item We propose a rigorous typing system for Dirac notation. Defined symbols (e.g., transpose and trace) are modelled by functions.
+        Also, the typing system removes the difficulty of disambiguation, therefore the language and the term rewriting system are much simplified.
+    \item An efficient algorithm to decide the extra equational theories $E$ is provided.
+The basic idea is normalization by sorting, which considers the interplay of AC symbols, SUM-SWAP and $\alpha$-equivalence.
+This algorithm reduces the complexity from the factorial level to the polynomial level.
+    \item The language to support variable labels is proposed.
+\end{itemize}
+
+\paragraph{Related Work} Other previous works explore the language and decision procedure to express quantum computation differently.
+
+\section{Preliminary and Motivation}
+The preliminary for the problem are two folds: Dirac notation and the equational logic in universal algebra.
+
+Quantum states live in complex Hilbert spaces. We use vectors in the space to descirbe pure quantum states, and operations corresponds to linear transformations. 
+In 1939, Dirac proposed his notation~\cite{dirac1939new} for quantum mechanics, which is essentially a wrapping of the linear algebra language. 
+Dirac notation uses the ket $\ket{i}$ and the bra $\bra{i}$ to indicate bases of the space and the dual space.
+Together with other variable symbols, they are composed with each other in sequence, and the composition will be interpreted into different operations, depending on the type of operands.
+For example, $\braket{i|j}$ represents the inner product of $\bra{i}$ and $\ket{j}$, which is a scalar, while $\ket{i}\bra{j}$ represents the outer product, resulting in an operator.
+\[
+    \braket{i|\phi}\braket{\psi|j} = \bra{i} (\ket{\phi}\bra{\psi}) \ket{j},
+\]
+and they are equivalent for all variables.
+Dirac notation also use $\otimes$ to indicate the vectors and operators in the tensor product space.
+
+In this manner, we can express long formulae in succint Dirac notations.
+The notation further enjoys the property that the interpretation is independent on the order of composition, thus parentheses can be omitted.
+For example, the formula \(\bra{i}\ket{\phi}\bra{\psi}\ket{j}\) can be understood as
+
+With the concrete basis, Dirac notations can be interpreted as matrices. For example ...
+
+We use universal algebra and equational logic to formally represent Dirac notations and the reasoning procedure in computers.
+
+\yx{mention the dilemma of simplicity and efficiency}
+
+\section{Language, Typing and Semantics}
+The syntax of Dirac notations involves three layers: the index, the type and the term. 
+Index represents classical datatypes, and they appear in type expression to denote the type of Hilbert spaces and sets.
+
+\begin{definition}[language syntax]
+    The syntax for type indices are defined as
+    \begin{align*}
+        \sigma ::=\ & x \mid \sigma_1 \times \sigma_2 \mid \mathsf{Bit}.
+    \end{align*}
+    The syntax for Dirac notation types is defined as 
+    \begin{align*}
+        T ::=\ & \Basis(\sigma) \mid \SType \mid \KType(\sigma) \mid \BType(\sigma) \mid \OType(\sigma_1, \sigma_2) \mid T_1 \to T_2 \mid \forall x.T \mid \SET(\sigma). \\
+    \end{align*}
+    The syntax for Dirac notation terms is defined as
+    \begin{align*}
+        e ::=\ & x \mid \lambda x : T.e \mid \mu x.e \mid e_1\ e_2 \mid e_1 \circ e_2 \\
+        & |\ \hat{0} \mid \hat{1} \mid (e_1, e_2) \\
+        & |\ 0 \mid 1 \mid \ADDS(e_1 \cdots e_n)  \mid e_1 \times \cdots \times e_n \mid e^*  \mid \delta_{e_1, e_2} \mid \DOT(e_1\ e_2) \\
+        & |\ \ZEROK(\sigma) \mid \ZEROB(\sigma) \mid \ZEROO(\sigma_1, \sigma_2) \mid \ONEO(\sigma) \\
+        & |\ \ket{e} \mid \bra{t} \mid e^\dagger \mid e_1.e_2 \mid \ADD(e_1 \cdots e_n) \mid e_1 \otimes e_2 \\
+        & |\ \MULK(e_1\ e_2) \mid \MULB(e_1\ e_2) \mid \OUTER(e_1\ e_2) \mid \MULO(e_1\ e_2) \\
+        & |\ \mathbf{U}(e) \mid e_1 \star e_2 \mid \sum_{e_1} e_2.
+    \end{align*}
+\end{definition}
+ % after consideration I still think keep DOT as binary is the best choice. It reduces the unstability of complicated matching in sequences.
+Here $i$ is a natural number and $\$i$ represents the $i$-th bound variable in de Bruijn notation. 
+Compared to [?], this syntax for Dirac notations merges the symbols with overlapped properties, such as the addition and scaling symbols for ket, bra and operator.
+Here $\ADDS$ and $\ADD$ are two different AC symbols representing the scalar addition and the linear algebra addition respectively.
+They will be denoted as $a_1 + \cdots + a_n$ and $X_1 + \cdots + X_n$.
+There are five kinds of linear algebra multiplications among ket, bra and operator, whose properties are similar but still diverge to some extent.
+For example, the rules $(O_1 \cdot O_2) \cdot K\ \reduce\ O_1 \cdot (O_2 \cdot K)$ and $B \cdot (O_1 \cdot O_2) \ \reduce\ (B \cdot O_1) \cdot O_2$ indicate that the sorting of multiplication sequences depends on the subterm types.
+To avoid frequent but unnecessary type checkings, we encode the typing information by using five different symbols, namely $\DOT$, $\MULK$, $\MULB$, $\OUTER$ and $\MULO$.
+They are denoted as $B\cdot K$, $K_1 \cdot K_2$, $B_1 \cdot B_2$, $K \cdot B$ and $O_1 \cdot O_2$, respectively.
+
+Usually, the sum body is specified by an abstraction. Therefore we use notation $\sum_s X$ to denote $\sum_{x \in s} \lambda x : T . X$ as well.
+
+\subsection{Typing System}
+
+The type checking of our language involves maintaining a well-formed environment and context $E[\Gamma]$, which is defined as follows.
+
+\begin{definition}[environment and context]
+    \begin{align*}
+        E ::=\ & [] \mid E; x : \Index \mid E; x : T \mid E; x := t : T. \\
+        \Gamma ::=\ & [] \mid \Gamma; x : \Index \mid \Gamma; x : T.
+    \end{align*}
+\end{definition}
+
+The environment and the context are sequences of assumptions $x : T$ or definitions $x := t : T$.
+With the environment, we can declare the type of variable symbols, and encode more operations on Dirac notations as definitions, such as the trace operator.
+Also, the existence of lambda abstractions requires a context of bound variables.
+
+We say an expression $t$ has type $X$ in context $E[\Gamma]$, if the typing judgement $E[\Gamma] \vdash t : X$ can be proved through the typing rules in~\Cref{sec: full typing rules}.
+Here we present and explain the rules selectively.
+Firstly, well-formed contexts $\WF(E)[\Gamma]$ are built in the incremental way, e.g.:
+\[
+    \frac{}{\WF([])[]}
+    \qquad
+    \frac{\WF(E)[] \qquad x \notin E}{\WF(E; x : \Index)[]}
+    \qquad
+    \frac{E[]\vdash t:T \qquad x \notin E}{\WF(E; x:=t:T)[]}.
+\]
+Starting from an empty context, we can assume new index symbols, and assume or define symbols with checked types.
+Based on the well-formed context, typing judgements can be proved by information from $E[\Gamma]$, or built inductively.
+In the following rules, for example, the condition $x : \Index \in E[\Gamma]$ is true if $E$ or $\Gamma$ has the assumption in their sequences, and $\sigma \times \tau$ is a index if both $\sigma$ and $\tau$ are typed as the index.
+$\KType(\sigma)$ and $\OType(\sigma, \tau)$ will be valid types for kets and operators, if their arguments are typed as the index.
+\begin{gather*}
+    \frac{\WF(E)[\Gamma] \qquad x : \Index \in E[\Gamma]}{E[\Gamma] \vdash x : \Index}
+    \qquad
+    \frac{E[\Gamma] \vdash \sigma : \Index \qquad E[\Gamma] \vdash \tau : \Index}{E[\Gamma] \vdash \sigma \times \tau : \Index} \\
+    \\
+    \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \KType(\sigma) : \Type}
+    \qquad
+    \frac{E[\Gamma] \vdash \sigma : \Index \qquad E[\Gamma] \vdash \tau : \Index}{E[\Gamma] \vdash \OType(\sigma, \tau) : \Type}
+\end{gather*}
+
+The Dirac notations will then be typed accordingly.
+For example, the ket syntax $\ket{t}$ has type $\KType(\sigma)$, if $t$ is typed as a basis term of index $\sigma$.
+Also, the inner product of a bra and a ket with the same type index $\sigma$ is typed as the scalar.
+This corresponds to the constraint of inner products that vectors should be in the same Hilbert space.
+\begin{gather*}
+    \frac{E[\Gamma]\vdash t : \Basis(\sigma)}{E[\Gamma] \vdash \ket{t} : \KType(\sigma)}
+    \qquad
+    \frac{E[\Gamma]\vdash B : \BType(\sigma) \qquad E[\Gamma]\vdash K : \KType(\sigma)}{E[\Gamma] \vdash B \cdot K : \SType}
+\end{gather*}
+
+The typing for functions and applications follow the common practice.
+We specify the syntax and typing rule for functions of indices $\mu x.
+t$: here $x$ is a bound variable of typed as $\Index$, and the type $U\{x/u\}$ of application $(t u)$ is obtained by replacing $x$ with the index instance $u$.
+
+\begin{gather*}
+    \frac{E[\Gamma; x : T] \vdash t : U}{E[\Gamma] \vdash (\lambda x : T . t) : T \to U}
+    \qquad
+    \frac{E[\Gamma] \vdash t:U \to T \qquad E[\Gamma] \vdash u:U}{E[\Gamma] \vdash (t\ u):T} \\
+    \\
+    \frac{E[\Gamma; x : \Index] \vdash t : U}{E[\Gamma] \vdash (\mu x.t) : \forall x.U}
+    \qquad
+    \frac{E[\Gamma] \vdash t:\forall x.U \qquad E[\Gamma] \vdash u : \Index}{E[\Gamma] \vdash (t\ u):U\{x/u\}}
+\end{gather*}
+
+The big operator sum is modelled by folding a function on a set, therefore the typing rule is as follows:
+\[
+    \frac{E[\Gamma] \vdash s : \SET(\sigma) \qquad E[\Gamma] \vdash f : \Basis(\sigma) \to \KType(\tau)}{E[\Gamma] \vdash \sum_{s} f : \KType(\tau)}.
+\]
+
+And lastly we have the typing rules for composition $x \circ y$. As is the case for casual Dirac notation, the typing of composition depends on the types of operands.
+\begin{gather*}
+    \frac{E[\Gamma] \vdash x : \SType \qquad E[\Gamma] \vdash y : \KType(\sigma)}{E[\Gamma] \vdash x \circ y : \KType(\sigma)}
+    \qquad
+    \frac{E[\Gamma] \vdash x : \OType(\sigma, \tau) \qquad E[\Gamma] \vdash y : \KType(\tau)}{E[\Gamma] \vdash x \circ y : \KType(\sigma)}
+\end{gather*}
+
+\begin{lemma}
+    The typing of expressions are decidable and unique.
+\end{lemma}
+
+\subsection{Semantics}
+
+Semantics assign the meanings to the expressions, and the goal of the decision procedure is to decide whether
+two expressions have the equivalent semantics. We consider two ways of definition: the axiomatic equations and the denotational semantics.
+
+The denotational way interprets every expression as a linear algebra concept, and equivalence is considered in the common mathematical sense.
+This explanation formalizes the original definition of Dirac notations, and best describes the target of the decision procedure.
+
+The semantics by equations, on the other hand, is an abstraction and axiomization. From the operational view, each equation declares a 
+valid rewriting operation, and two expressions are equiavlent if and only if they can be rewritten into the same form using the axioms.
+
+\section{Decision procedure for Dirac notations}
+
+The following two sections talk about how to decide the equivalence of Dirac notations.
+In the previous work, the equational axioms are separated into two parts: a set $E$ of equations that can not be decided by rewriting, and the remaining equations to be decided by term rewriting.
+Here, the equational theories $E$ inlude:
+\begin{itemize}
+    \item Commutativity of symbols $a + b$, $a\times b$ and $\delta_{s, t}$,
+    \item $\alpha$-equivalence of bound variables, i.e., $\lambda x.t = \lambda y . t\{x/y\}$,
+    \item swapping successive summations, i.e., $\sum_{i \in s_1} \sum_{j \in s_2} A = \sum_{j \in s_2} \sum_{i \in s_1}A$, and
+    \item equational theories for scalars.
+\end{itemize}
+
+In this work, the general idea is to carry through the normalization procedure, so that semantical equivalence can be directly checked by the syntax of normal forms.
+The procedure of the normalization is displayed in~\Cref{fig: normalization}.
+
+\input{flowchart.tex}
+
+The last three steps deals with the equational theory $E$, while the first three steps use term rewriting to work on the structure of Dirac notations.
+
+\subsection{Normalization modulo $E$ by Term Rewriting}
+Term rewriting rules, written as $l -> r$, are used to normalize terms by recursively matching the subterms of the term with the left hand side $l$, and replace them with the corresponding right hand side $r$.
+The procedure terminates when no more rewritings can be made, and the order of rewritings will be irrelevant if the term rewriting system is \textit{confluent}, which is a desirable property.
+
+To express the term rewriting system, the previous work adheres to use the naive universal algebra, where each function symbol has a fixed arity, and the left hand side pattern only allows constant symbols and variables.
+This constraint enables them to check the confluence and termination of their term rewriting system with other tools.
+
+Here to enable more efficient algorithms, our langauge uses AC symbols with a variable arity.
+
+The full list of rewriting rules are in~\Cref{sec: rewriting rules}. Here we present some of them to illustrate the design idea.
+
+When using variable arity AC symbols, the associativity can be normalized by the following flattening rule:
+
+\[
+a_1 + \cdots + (b_1 + \cdots + b_m) + \cdots + a_n
+\ \reduce\ a_1 + \cdots + b_1 + \cdots + b_m + \cdots + a_n,
+\]
+
+while commutativity is left for a sorting algorithm later on.
+
+The term rewriting rule follows the system in the previous work.
+
+The general idea is to reduce all the possible calculations, and transform multiplication into tensor product as much as possible.
+
+And some rules exists for completeness. As an example, given the following rule snippet
+\begin{align*}
+    %
+    & \text{(R-DOT10)}
+    && (B \cdot O) \cdot K \ \reduce\ B \cdot (O \cdot K), \\
+    %
+    & \text{(R-DOT11)}
+    && \bra{(s, t)} \cdot ((O_1 \otimes O_2) \cdot K) \ \reduce\ ((\bra{s} \cdot O_1) \otimes (\bra{t} \cdot O_2)) \cdot K, \\
+    %
+    & \text{(R-MULB10)}
+    && \bra{(s, t)} \cdot (O_1 \otimes O_2)\ \reduce\ (\bra{s} \cdot O_1) \otimes (\bra{t} \cdot O_2),
+\end{align*}
+the normalization of term $(\bra{(s, t)} \cdot (O_1 \otimes O_2)) \cdot K$ have two rewriting paths: (a). apply (R-MULB10) and get $((\bra{s} \cdot O_1) \otimes (\bra{t} \cdot O_2)) \cdot K$, or
+(b).
+first apply (R-DOT10) and sort the term into $\bra{(s, t)} \cdot ((O_1 \otimes O_2) \cdot K)$, and then apply (R-DOT11) to get the same result.
+Here (R-DOT11) is for completeness of cases similar to this example.
+
+One important technique revealed in the previous work is the expansion of variables:
+
+\begin{gather*}
+    \frac{E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash K \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)}(\bra{i}\cdot K).\ket{i}}
+    \qquad \qquad
+    \frac{E[\Gamma] \vdash B : \BType(\sigma)}{E[\Gamma] \vdash B \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)}(B \cdot \ket{i}).\bra{i}} \\
+    \\
+    \frac{E[\Gamma] \vdash O : \OType(\sigma, \tau)}{E[\Gamma] \vdash O \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)} \sum_{j \in \mathbf{U}(\tau)}(\bra{i} \cdot O \cdot \ket{j}).(\ket{i} \cdot \bra{j})}
+\end{gather*}
+The above three rules are obviously not terminating, and this is why we have the rewriting-expansion-rewriting steps in the decision procedure.
+On the other hand, we discovered that doing the expansion on all variables only once is already sufficient.
+
+\begin{lemma}
+    Let $\text{expand}(e)$ indicate the result of expanding all variables in $e$ once.
+    For all well-typed term $e$ in $E[\Gamma]$, we have $\text{expand}(\text{expand}(e)) \downarrow = \text{expand}(e)\downarrow$.
+\end{lemma}
+\begin{proof}
+\end{proof}
+
+\section{Deciding Equational Theory $E$}
+
+The previous work did not consider the algorithm to decide equational theory $E$.
+In their Mathematica implementation, it is implemented by a unification, which tries to find a substitution of summation bound variables that makes the two expressions syntactically equivalent.
+To decide AC-equivalence and (SUM-SWAP), they iterate through all the permutations, and the complexity is factorial to the number of AC symbol arguments and successive summations.
+
+A standard approach to decide this permutation equivalence is to normalize by sorting with a given order.
+For example, given the dictionary order $a < b < c$, the term $b + c + a$ (and any other AC equivalent ones) will be normalized into $a + b + c$.
+For our setting, there are two related difficulties: how to assign such an order to all terms in our language, and how to normalize with respect to AC-equivalence and (SUM-SWAP) at the same time.
+
+At last, we will prove that equivalence by this normalization procedure is sound in the semantics.
+
+\begin{theorem}[soundness]
+    For any well-formed context $E[\Gamma]$ and well-typed expressions $e_1$ and $e_2$, if $e_1\downarrow = e_2\downarrow$, then $\sem{e_1} = \sem{e_2}$.
+\end{theorem}
+\begin{proof}
+
+\end{proof}
+
+% Labelled Dirac Notation
+
+\input{labelled}
+
+\section{Implementation and Case Study}
+
+\section{Conclusion}
+
+\clearpage
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Consideration}
 \begin{itemize}
-    \item We can emphasize that we are moving from pure term rewriting to a hybrid algorithm.
-    \item AC equivalence and alpha equivalence.
-    \item Efficient AC equivalence checking by sorting on syntax.
     \item Sorting requires that alpha equivalent terms have the same syntax.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Language Syntax}
-\begin{definition}[syntax]
-    The syntax for type indices are defined as
-    \begin{align*}
-        \sigma ::=\ & x \mid \sigma_1 \times \sigma_2 \mid \mathsf{Qdit}_n.
-    \end{align*}
-    The syntax for Dirac notation types is defined as 
-    \begin{align*}
-        T ::=\ & x \mid \Basis(\sigma) \mid \SType \mid \KType(\sigma) \mid \BType(\sigma) \mid \OType(\sigma_1, \sigma_2) \mid T_1 \to T_2 \mid \forall x.T \mid \SET(\sigma). \\
-    \end{align*}
-    The syntax for Dirac notation terms is defined as
-    \begin{align*}
-        e ::=\ & x \mid (e_1, e_2) \mid \lambda x : T.e \mid \mu x.e \mid e_1\ e_2 \\
-        & |\ 0 \mid 1 \mid \ADDS(e_1 \cdots e_n)  \mid e_1 \times \cdots \times e_n \mid e^*  \mid \delta_{e_1, e_2} \mid \DOT(e_1\ e_2) \\
-        & |\ \ZEROK(\sigma) \mid \ZEROB(\sigma) \mid \ZEROO(\sigma_1, \sigma_2) \mid \ONEO(\sigma) \\
-        & |\ \ket{e} \mid \bra{t} \mid e^\dagger \mid e_1.e_2 \mid \ADD(e_1 \cdots e_n) \mid e_1 \otimes e_2 \\
-        & |\ \MULK(e_1\ e_2) \mid \MULB(e_1\ e_2) \mid \OUTER(e_1\ e_2) \mid \MULO(e_1\ e_2) \\
-        & |\ \mathbf{U}(e) \mid e_1 \star e_2 \mid \sum_{e_1} e_2.
-    \end{align*}
-\end{definition}
- % after consideration I still think keep DOT as binary is the best choice. It reduces the unstability of complicated matching in sequences.
-Here $i$ is a natural number and $\$i$ represents the $i$-th bound variable in de Bruijn notation. 
-Compared to [?], this syntax for Dirac notations merges the symbols with overlapped properties, such as the addition and scaling symbols for ket, bra and operator.
-Here $\ADDS$ and $\ADD$ are two different AC symbols representing the scalar addition and the linear algebra addition respectively.
-They will be denoted as $a_1 + \cdots + a_n$ and $X_1 + \cdots + X_n$.
-There are five kinds of linear algebra multiplications among ket, bra and operator, whose properties are similar but still diverge to some extent.
-For example, the rules $(O_1 \cdot O_2) \cdot K\ \reduce\ O_1 \cdot (O_2 \cdot K)$ and $B \cdot (O_1 \cdot O_2) \ \reduce\ (B \cdot O_1) \cdot O_2$ indicate that the sorting of multiplication sequences depends on the subterm types.
-To avoid frequent but unnecessary type checkings, we encode the typing information by using five different symbols, namely $\DOT$, $\MULK$, $\MULB$, $\OUTER$ and $\MULO$.
-They are denoted as $B\cdot K$, $K_1 \cdot K_2$, $B_1 \cdot B_2$, $K \cdot B$ and $O_1 \cdot O_2$, respectively.
-
-Usually, the sum body is specified by an abstraction. Therefore we use notation $\sum_s X$ to denote $\sum_{x \in s} \lambda x : T . X$ as well.
-
-\section{Typing System}
-The type checking of Dirac notations involves maintaining a well-formed environment and context $E[\Gamma]$, which specifies the definitions and typing assumtpions for variables.
-The environment and the conetxt are defined as follows.
-
-\begin{definition}[environment and context]
-    \begin{align*}
-        E ::=\ & [] \mid E; x : \Index \mid E; T : \Type \mid E; x : T \mid E; x := t : T. \\
-        \Gamma ::=\ & [] \mid \Gamma; x : \Index \mid \Gamma; x : T.
-    \end{align*}
-\end{definition}
-
-Note that in the following rules, $x : \Index$ and $x : \Type$ are not considered as $x : T$ judgements.
-
-\begin{figure}[h]
-    \begin{gather*}
-        \textbf{W-Empty} \qquad
-        \frac{}{\WF([])[]} \\
-        \\
-        \textbf{W-AssumE-Index} \qquad
-        \frac{\WF(E)[] \qquad x \notin E}{\WF(E; x : \Index)[]}\qquad \qquad
-        \textbf{W-AssumE-Type} \qquad
-        \frac{\WF(E)[] \qquad x \notin E}{\WF(E; x : \Type)[]} \\
-        \\
-        \textbf{W-AssumE-Term} \qquad
-        \frac{E[] \vdash T : \Type \qquad x \notin E}{\WF(E; x:T)[]}
-        \qquad \qquad
-        \textbf{W-Def-Term} \qquad
-        \frac{E[]\vdash t:T \qquad x \notin E}{\WF(E; x:=t:T)[]} \\
-        \\
-        \textbf{W-AssumC-Index} \qquad
-        \frac{\WF(E)[\Gamma]}{\WF(E)[\Gamma; x : \Index]}
-        \qquad \qquad
-        \textbf{W-AssumC-Term} \qquad
-        \frac{E[\Gamma] \vdash T : \Type}{\WF(E)[\Gamma; x : T]}
-    \end{gather*}
-    \caption{Rules for a well-formed environment and context.}
-\end{figure}
-
-\begin{figure}
-    \begin{gather*}
-        \textbf{Index-Var} \qquad
-        \frac{\WF(E)[\Gamma] \qquad x : \Index \in E[\Gamma]}{E[\Gamma] \vdash x : \Index}
-        \qquad \qquad
-        \textbf{Index-Prod} \qquad
-        \frac{E[\Gamma] \vdash \sigma : \Index \qquad E[\Gamma] \vdash \tau : \Index}{E[\Gamma] \vdash \sigma \times \tau : \Index} \\
-        \\
-        \textbf{Index-Qudit} \qquad
-        \frac{\WF(E)[\Gamma]}{E[\Gamma] \vdash \mathsf{Qudit}_n : \Index}
-    \end{gather*}
-    \caption{Rules for type index.}
-\end{figure}
-
-\begin{figure}
-    \begin{gather*}
-        \textbf{Type-Arrow} \qquad
-        \frac{E[\Gamma] \vdash T : \Type \qquad E[\Gamma] \vdash U : \Type}{E[\Gamma] \vdash T \to U : \Type}
-        \qquad \qquad
-        \textbf{Type-Index} \qquad
-        \frac{E[\Gamma ; x : \Index] \vdash U : \Type}{E[\Gamma] \vdash \forall x.U : \Type} \\
-        \\        
-        \textbf{Type-Basis} \qquad
-        \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \Basis(\sigma) : \Type}  \\
-        \\
-        \textbf{Type-Ket} \qquad
-        \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \KType(\sigma) : \Type}
-        \qquad \qquad
-        \textbf{Type-Bra} \qquad
-        \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \BType(\sigma) : \Type} \\
-        \\
-        \textbf{Type-Opt} \qquad
-        \frac{E[\Gamma] \vdash \sigma : \Index \qquad E[\Gamma] \vdash \tau : \Index}{E[\Gamma] \vdash \OType(\sigma, \tau) : \Type}
-        \qquad \qquad
-        \textbf{Type-Scalar} \qquad
-        \frac{\WF(E)[\Gamma]}{E[\Gamma] \vdash \SType : \Type} \\
-        \\
-        \textbf{Type-Set} \qquad
-        \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \SET(\sigma) : \Type}
-    \end{gather*}
-    \caption{Rules for types.}
-\end{figure}
-
-\begin{figure}[h]
-    \begin{gather*}
-        \textbf{Term-Var} \qquad
-        \frac{\WF(E)[\Gamma] \qquad (x:T) \in E[\Gamma] \text{ or } (x := t : T) \in E \text{ for some $t$}}{E[\Gamma] \vdash x : T} \\
-        \\
-        \textbf{Lam} \qquad
-        \frac{E[\Gamma; x : T] \vdash t : U}{E[\Gamma] \vdash (\lambda x : T . t) : T \to U}
-        \qquad \qquad
-        \textbf{Index} \qquad
-        \frac{E[\Gamma; x : \Index] \vdash t : U}{E[\Gamma] \vdash (\mu x.t) : \forall x.U} \\
-        \\
-        \textbf{App-Arrow} \qquad
-        \frac{E[\Gamma] \vdash t:U \to T \qquad E[\Gamma] \vdash u:U}{E[\Gamma] \vdash (t\ u):T}
-        \qquad \qquad
-        \textbf{App-Index} \qquad
-        \frac{E[\Gamma] \vdash t:\forall x.U \qquad E[\Gamma] \vdash u : \Index}{E[\Gamma] \vdash (t\ u):U\{x/u\}}
-    \end{gather*}
-    \caption{Rules for variable and function typings. Here $U[u]$ means instantiate the first bound variable with $u$ in $U$.}
-\end{figure}
-
-\begin{figure}[h]
-    \begin{gather*}
-        % this two rules can be considered later
-        % \textbf{Base-Def} \qquad
-        % \frac{\WF(\Gamma) \qquad a_i \notin \Gamma \text{ for all $i$}}
-        % {\Gamma\vdash \{a_1, a_2, \cdots, a_n\} : \Basis}
-        % \qquad \qquad
-        % \textbf{Base-Element} \qquad
-        % \frac{\WF(\Gamma) \qquad c \in \{a_1, a_2, \cdots a_n\}}{\Gamma\vdash c : \{a_1, a_2, \cdots, a_n\}} \\
-        % \\
-        \textbf{Pair-Base} \qquad
-        \frac{
-            E[\Gamma]\vdash s : \Basis(\sigma) \qquad E[\Gamma]\vdash t : \Basis(\tau)} {E[\Gamma]\vdash (s, t) : \Basis(\sigma \times \tau)} 
-        \textbf{}
-    \end{gather*}
-    \caption{Typing rules for Basis. }
-\end{figure}
-
-\begin{figure}[h]
-    \begin{gather*}
-        \textbf{Sca-0} \qquad
-        \frac{\WF(E)[\Gamma]}{E[\Gamma] \vdash 0 : \SType}
-        \qquad \qquad
-        \textbf{Sca-1} \qquad
-        \frac{\WF(E)[\Gamma]}{E[\Gamma] \vdash 1 : \SType} \\
-        \\
-        \textbf{Sca-Delta} \qquad
-        \frac{ E[\Gamma]\vdash s : \Basis(\sigma) \qquad E[\Gamma]\vdash t : \Basis(\sigma) } {E[\Gamma] \vdash \delta_{s, t} : \SType} \\
-        \\
-        \textbf{Sca-Add} \qquad
-        \frac{E[\Gamma]\vdash a_i : \SType \text{ for all $i$}}{E[\Gamma]\vdash a_1 + \cdots + a_n : \SType}
-        \qquad \qquad
-        \textbf{Sca-Mul} \qquad
-        \frac{E[\Gamma]\vdash a_i : \SType \text{ for all $i$}}{E[\Gamma]\vdash a_1 \times \cdots \times a_n : \SType} \\
-        \\
-        \textbf{Sca-Conj} \qquad
-        \frac{E[\Gamma] \vdash a : \SType}{E[\Gamma] \vdash a^*:\SType}
-        \qquad \qquad
-        \textbf{Sca-Dot} \qquad
-        \frac{E[\Gamma]\vdash B : \BType(\sigma) \qquad E[\Gamma]\vdash K : \KType(\sigma)}{E[\Gamma] \vdash B \cdot K : \SType}
-    \end{gather*}
-    \caption{Scalar typing rules.}
-\end{figure}
-
-\begin{figure}[h]
-    \begin{gather*}
-        \textbf{Ket-0} \qquad
-        \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \ZEROK(\sigma) : \KType(\sigma)} 
-        \qquad \qquad
-        \textbf{Ket-Base} \qquad
-        \frac{E[\Gamma]\vdash t : \Basis(\sigma)}{E[\Gamma] \vdash \ket{t} : \KType(\sigma)} \\
-        \\
-        \textbf{Ket-Adj} \qquad
-        \frac{E[\Gamma] \vdash B : \BType(\sigma)}{E[\Gamma] \vdash B^\dagger : \KType(\sigma)} 
-        \qquad \qquad
-        \textbf{Ket-Scr} \qquad
-        \frac{E[\Gamma] \vdash a : \SType \qquad E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash a.K : \KType(\sigma)} \\
-        \\
-        \textbf{Ket-Add} \qquad
-        \frac{E[\Gamma] \vdash K_i : \KType(\sigma) \text{ for all $i$}}{E[\Gamma] \vdash K_1 + \cdots + K_n : \KType(\sigma)}
-        \qquad \qquad
-        \textbf{Ket-MulK} \qquad
-        \frac{E[\Gamma] \vdash O : \OType(\sigma, \tau) \qquad E[\Gamma] \vdash K : \KType(\tau)}{E[\Gamma] \vdash O \cdot K : \KType(\sigma)} \\
-        \\
-        \textbf{Ket-Tsr} \qquad
-        \frac{E[\Gamma] \vdash K_1 : \KType(\sigma) \qquad E[\Gamma] \vdash K_2 : \KType(\tau)} {E[\Gamma] \vdash K_1 \otimes K_2 : \KType(\sigma \times \tau)}
-    \end{gather*}
-    \caption{Ket typing rules.}
-\end{figure}
-
-\begin{figure}[h]
-    \begin{gather*}
-        \textbf{Bra-0} \qquad
-        \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \ZEROB(\sigma) : \BType(\sigma)} 
-        \qquad \qquad
-        \textbf{Bra-Base} \qquad
-        \frac{E[\Gamma]\vdash t : \Basis(\sigma)}{E[\Gamma] \vdash \bra{t} : \BType(\sigma)} \\
-        \\
-        \textbf{Bra-Adj} \qquad
-        \frac{E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash K^\dagger : \BType(\sigma)} 
-        \qquad \qquad
-        \textbf{Bra-Scr} \qquad
-        \frac{E[\Gamma] \vdash a : \SType \qquad E[\Gamma] \vdash B : \BType(\sigma)}{E[\Gamma] \vdash a.B : \BType(\sigma)} \\
-        \\
-        \textbf{Bra-Add} \qquad
-        \frac{E[\Gamma] \vdash B_i : \BType(\sigma) \text{ for all $i$}}{E[\Gamma] \vdash B_1 + \cdots + B_n : \BType(\sigma)}
-        \qquad \qquad
-        \textbf{Bra-MulB} \qquad
-        \frac{E[\Gamma] \vdash B : \KType(\sigma) \qquad E[\Gamma] \vdash O : \OType(\sigma, \tau)}{E[\Gamma] \vdash B \cdot O : \BType(\tau)} \\
-        \\
-        \textbf{Bra-Tsr} \qquad
-        \frac{E[\Gamma] \vdash B_1 : \BType(\sigma) \qquad E[\Gamma] \vdash B_2 : \BType(\tau)} {E[\Gamma] \vdash B_1 \otimes B_2 : \BType(\sigma \times \tau)}
-    \end{gather*}
-    \caption{Bra typing rules.}
-\end{figure}
-
-\begin{figure}[h]
-    \begin{gather*}
-        \textbf{Opt-0} \qquad
-        \frac{E[\Gamma] \vdash \sigma : \Index \qquad E[\Gamma] \vdash \tau : \Index}{E[\Gamma] \vdash \ZEROO(\sigma, \tau) : \OType(\sigma, \tau)} 
-        \qquad \qquad
-        \textbf{Opt-1} \qquad
-        \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \ONEO(\sigma) : \OType(\sigma, \sigma)} \\
-        \\
-        \textbf{Opt-Adj} \qquad
-        \frac{E[\Gamma] \vdash O : \OType(\sigma, \tau)}{E[\Gamma] \vdash O^\dagger : \OType(\tau, \sigma)} 
-        \qquad \qquad
-        \textbf{Opt-Scr} \qquad
-        \frac{E[\Gamma] \vdash a : \SType \qquad E[\Gamma] \vdash O : \OType(\sigma, \tau)}{E[\Gamma] \vdash a.O : \OType(\sigma, \tau)} \\
-        \\
-        \textbf{Opt-Add} \qquad
-        \frac{E[\Gamma] \vdash O_i : \OType(\sigma, \tau) \text{ for all $i$}}{E[\Gamma] \vdash O_1 + \cdots + O_n : \OType(\sigma, \tau)}
-        \qquad \qquad
-        \textbf{Opt-Outer} \qquad
-        \frac{E[\Gamma]\vdash K : \KType(\sigma) \qquad E[\Gamma]\vdash B : \BType(\tau)}{E[\Gamma]\vdash K \cdot B : \OType(\sigma, \tau)} \\
-        \\
-        \textbf{Opt-MulO} \qquad
-        \frac{E[\Gamma] \vdash O_1 : \OType(\sigma, \tau) \qquad E[\Gamma] \vdash O_2 : \OType(\tau, \rho)}{E[\Gamma] \vdash O_1 \cdot O_2 : \OType(\sigma, \rho)} \\
-        \\
-        \textbf{Opt-Tsr} \qquad
-        \frac{E[\Gamma] \vdash O_1 : \OType(\sigma_1, \tau_1) \qquad E[\Gamma] \vdash O_2 : \OType(\sigma_2, \tau_2)} {E[\Gamma] \vdash O_1 \otimes O_2 : \OType(\sigma_1 \times \sigma_2, \tau_1 \times \tau_2)}
-    \end{gather*}
-    \caption{Operator typing rules.}
-\end{figure}
-
-\begin{figure}[h]
-    \begin{gather*}
-        \textbf{Set-U} \qquad
-        \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \mathbf{U}(\sigma) : \SET(\sigma)} 
-        \qquad \qquad
-        \textbf{Set-Prod} \qquad
-        \frac{E[\Gamma] \vdash A : \SET(\sigma) \qquad E[\Gamma] \vdash B : \SET(\tau)}{E[\Gamma] \vdash A \star B : \SET(\sigma \times \tau)}
-    \end{gather*}
-    \caption{Set typing rules.}
-\end{figure}
-
-\begin{figure}[h]
-    \begin{gather*}
-        \textbf{Sum-Scalar} \qquad
-        \frac{E[\Gamma] \vdash s : \SET(\sigma) \qquad E[\Gamma] \vdash f : \Basis(\sigma) \to \SType}{E[\Gamma] \vdash \sum_{s} f : \SType} \\
-        \\
-        \textbf{Sum-Ket} \qquad
-        \frac{E[\Gamma] \vdash s : \SET(\sigma) \qquad E[\Gamma] \vdash f : \Basis(\sigma) \to \KType(\tau)}{E[\Gamma] \vdash \sum_{s} f : \KType(\tau)} \\
-        \\
-        \textbf{Sum-Bra} \qquad
-        \frac{E[\Gamma] \vdash s : \SET(\sigma) \qquad E[\Gamma] \vdash f : \Basis(\sigma) \to \BType(\tau)}{E[\Gamma] \vdash \sum_{s} f : \BType(\tau)} \\
-        \\
-        \textbf{Sum-Opt} \qquad
-        \frac{E[\Gamma] \vdash s : \SET(\sigma) \qquad E[\Gamma] \vdash f : \Basis(\sigma) \to \OType(\tau, \rho)}{E[\Gamma] \vdash \sum_{s} f : \OType(\tau, \rho)}
-    \end{gather*}
-    \caption{Sum typing rules.}
-\end{figure}
-
-\clearpage
-\section{Conversions and Reductions}
-
-\begin{figure}[h]
-    \begin{gather*}
-        \frac{E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash K \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)}(\bra{i}\cdot K).\ket{i}} \\
-        \\
-        \frac{E[\Gamma] \vdash B : \BType(\sigma)}{E[\Gamma] \vdash B \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)}(B \cdot \ket{i}).\bra{i}} \\
-        \\
-        \frac{E[\Gamma] \vdash O : \OType(\sigma, \tau)}{E[\Gamma] \vdash O \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)} \sum_{j \in \mathbf{U}(\tau)}(\bra{i} \cdot O \cdot \ket{j}).(\ket{i} \cdot \bra{j})}
-    \end{gather*}
-\end{figure}
-
-\begin{figure}[h]
-    \begin{gather*}
-        \textbf{Beta-Arrow} \qquad
-        \frac{}{E[\Gamma] \vdash ((\lambda x : T.t)\ u)\ \reduce\ t\{x/u\}} 
-        \qquad \qquad
-        \textbf{Beta-Index} \qquad
-        \frac{}{E[\Gamma] \vdash ((\mu x.t)\ u)\ \reduce\ t\{x/u\}} 
-        \\
-        \\
-        \textbf{Delta} \qquad
-        \frac{\WF(E)[\Gamma] \qquad (c:=t:T) \in E}{E[\Gamma] \vdash c\ \reduce\ t}
-    \end{gather*}
-    \caption{Conversions and reductions for the typed lambda calculus.}
-\end{figure}
-
-\newenvironment{ruletable}[1]
-{
-    \begin{longtable}{cl}
-    \caption{#1}\\
-    \hline
-    \textbf{Rule} & \textbf{Description} \\
-    \hline
-    \endfirsthead
-
-    \hline
-    \textbf{Rule} & \textbf{Description} \\
-    \hline
-    \endhead
-
-    \hline
-    \multicolumn{2}{r}{\textit{Continued on the next page}} \\
-    \hline
-    \endfoot
-
-    \hline
-    \endlastfoot
-}
-{
-    \end{longtable}
-}
-
-\renewcommand{\arraystretch}{1.2} % Increases row height by 50%
-
-\begin{ruletable}{The special flattening rule.}
-    %
-    R-FLATTEN
-    & $a_1 + \cdots + (b_1 + \cdots + b_m) + \cdots + a_n\ \reduce\ a_1 + \cdots + b_1 + \cdots + b_m + \cdots + a_n$ \\
-    & $a_1 \times \cdots \times (b_1 \times \cdots \times b_m) \times \cdots \times a_n\ \reduce\ a_1 \times \cdots \times b_1 \times \cdots \times b_m \times \cdots \times a_n$ \\
-    & $X_1 + \cdots + (X_1' + \cdots + X_m') + \cdots + X_n\ \reduce\ X_1 + \cdots + X_1' + \cdots + X_m' + \cdots + X_n$ \\
-    & \textcolor{gray}{A special rule to flatten all AC symbols within one call.}    
-\end{ruletable}
-
-\begin{ruletable}{Rules for scalar addition and multiplication.}
-    %
-    R-ADDSID 
-    & $+(a) \ \reduce\ a$ \\
-    & \textcolor{gray}{Reduce the term if the AC symbol $+$ only has one argument.} \\
-    %
-    R-MULSID 
-    & $\times(a) \ \reduce\ a$ \\
-    & \textcolor{gray}{Similar to (R-ADDSID).} \\
-    %
-    R-ADDS0 
-    & $a_1 + \cdots + 0 + \cdots + a_n \ \reduce\ a_1 + \cdots + a_n$ \\
-    & \textcolor{gray}{This rule removes all $\ZERO$ occurances and keeps the order of remaining subterms.} \\
-    %
-    R-MULS0
-    & $a_1 \times \cdots \times 0 \times \cdots \times a_n \ \reduce\ 0$ \\
-    %
-    R-MULS1
-    & $a_1 \times \cdots \times \ONE \times \cdots \times a_n \ \reduce\ a_1 \times \cdots \times a_n$ \\
-    & \textcolor{gray}{Similar to (R-ADDS0).} \\
-    %
-    R-MULS2
-    & $b_1 \times \cdots \times (a_1 + \cdots + a_n) \times \cdots \times b_m$ \\
-    & $\reduce\ (b_1 \times \cdots \times a_1 \times \cdots \times b_m) + \cdots + (b_1 \times \cdots \times a_n \times \cdots \times b_m)$ \\
-    & \textcolor{gray}{This rule matches the first scalar addition subterm in the list.} \\
-\end{ruletable}
-
-\begin{ruletable}{Rules for other scalar symbols.}
-    %
-    R-CONJ0
-    & $0^* \ \reduce\ 0$ \\
-    %
-    R-CONJ1
-    & $1^* \ \reduce\ 1$ \\
-    %
-    R-CONJ2
-    & $(a_1 + \cdots + a_n)^* \ \reduce\ a_1^* + \cdots + a_n^*$ \\
-    %
-    R-CONJ3
-    & $(a_1 \times \cdots \times a_n)^* \ \reduce\ a_1^* \times \cdots \times a_n^*$ \\
-    %
-    R-CONJ4
-    & $ (a^*)^* \ \reduce\ a$ \\
-    % 
-    R-CONJ5
-    & $ \delta_{s, t}^* \ \reduce\ \delta_{s, t}$ \\
-    %
-    R-CONJ6
-    & $ (B \cdot K)^* \ \reduce\ K^\dagger \cdot B^\dagger $ \\
-    %
-    R-DOT0
-    & $ \ZEROB(\sigma) \cdot K \ \reduce\ 0 $ \\
-    %
-    R-DOT1
-    & $ B \cdot \ZEROK(\sigma) \ \reduce\ 0 $ \\
-    %
-    R-DOT2
-    & $ (a.B) \cdot K \ \reduce\ a \times (B \cdot K) $ \\
-    %
-    R-DOT3
-    & $ B \cdot (a.K) \ \reduce\ a \times (B \cdot K) $ \\
-    %
-    R-DOT4
-    & $ (B_1 + \cdots + B_n) \cdot K \ \reduce\ B_1 \cdot K + \cdots + B_n \cdot K $ \\
-    %
-    R-DOT5
-    & $ B \cdot (K_1 + \cdots + K_n) \ \reduce\ B \cdot K_1 + \cdots + B \cdot K_n $ \\
-    %
-    R-DOT6
-    & $ \bra{s} \cdot \ket{t} \ \reduce\ \delta_{s, t} $ \\
-    %
-    R-DOT7
-    & $ (B_1 \otimes B_2) \cdot \ket{(s, t)} \ \reduce\ (B_1 \cdot \ket{s}) \times (B_2 \cdot \ket{t}) $ \\
-    %
-    R-DOT8
-    & $ \bra{(s, t)} \cdot (K_1 \otimes K_2) \ \reduce\ (\bra{s} \cdot K_1) \times (\bra{t} \cdot K_2) $ \\
-    %
-    R-DOT9
-    & $ (B_1 \otimes B_2) \cdot (K_1 \otimes K_2) \ \reduce\ (B_1 \cdot K_1) \times (B_2 \cdot K_2) $ \\
-    %
-    R-DOT10
-    & $ (B \cdot O) \cdot K \ \reduce\ B \cdot (O \cdot K) $ \\
-    %
-    R-DOT11
-    & $ \bra{(s, t)} \cdot ((O_1 \otimes O_2) \cdot K) \ \reduce\ ((\bra{s} \cdot O_1) \otimes (\bra{t} \cdot O_2)) \cdot K $ \\
-    %
-    R-DOT12
-    & $ (B_1 \otimes B_2) \cdot ((O_1 \otimes O_2) \cdot K) \ \reduce\ ((B_1 \cdot O_1) \otimes (B_2 \cdot O_2)) \cdot K $ \\
-    %
-    R-DELTA0
-    & $ \delta_{a, a} \ \reduce\ 1$ \\
-    %
-    R-DELTA1
-    & $ \delta_{(a, b), (c, d)} \ \reduce\ \delta_{a, c} \times \delta_{b, d}$ \\
-\end{ruletable}
-
-\begin{ruletable}{Rules for linear algebra scaling.}
-    %
-    R-SCR0
-    & $ 1.X \ \reduce\ X $ \\
-    %
-    R-SCR1
-    & $ a.(b.X) \ \reduce\ (a \times b).X $ \\
-    %
-    R-SCR2
-    & $ a.(X_1 + \cdots + X_n) \ \reduce\ a.X_1 + \cdots + a.X_n $ \\
-    %
-    R-SCRK0
-    & $ K : \KType(\sigma) \Rightarrow 0.K \ \reduce\ \ZEROK(\sigma)$ \\
-    %
-    R-SCRK1
-    & $ a.\ZEROK(\sigma) \ \reduce\ \ZEROK(\sigma) $ \\
-    % 
-    R-SCRB0
-    & $ B : \BType(\sigma) \Rightarrow 0.B \ \reduce\ \ZEROB(\sigma)$ \\
-    % 
-    R-SCRB1
-    & $ a.\ZEROB(\sigma) \ \reduce\ \ZEROB(\sigma) $ \\
-    % 
-    R-SCRO0
-    & $ O : \OType(\sigma, \tau) \Rightarrow 0.O \ \reduce\ \ZEROO(\sigma, \tau)$ \\
-    % 
-    R-SCRO1
-    & $ a.\ZEROO(\sigma, \tau) \ \reduce\ \ZEROO(\sigma, \tau) $ \\
-\end{ruletable}
-
-\begin{ruletable}{Rules for linear algebra addition. }
-    %
-    R-ADDID
-    & $+(X) \ \reduce\ X$ \\
-    & \textcolor{gray}{Reduce the term if the AC symbol $+$ only has one argument.} \\
-    %
-    R-ADD0
-    & $Y_1 + \cdots + X + \cdots + X + \cdots + Y_n \ \reduce\ Y_1 + \cdots + Y_n + \cdots + (1+1).X$ \\
-    & \textcolor{gray}{\makecell[l]{This rule matches the first $X$ in the list satisfying the pattern. \\ The result $(1+1).X$ will be placed at the end of the list.}} \\
-    %
-    R-ADD1
-    & $Y_1 + \cdots + X + \cdots + a.X + \cdots + Y_n \ \reduce\ Y_1 + \cdots + Y_n + (1+a).X$ \\
-    & \textcolor{gray}{Similar to (R-ADD0).} \\
-    %
-    R-ADD2
-    & $Y_1 + \cdots + a.X + \cdots + X + \cdots + Y_n \ \reduce\ Y_1 + \cdots + Y_n + (a+1).X$ \\
-    & \textcolor{gray}{Similar to (R-ADD0).} \\
-    %
-    R-ADD3
-    & $Y_1 + \cdots + a.X + \cdots + b.X + \cdots + Y_n \ \reduce\ Y_1 + \cdots + Y_n + (a+b).X$ \\
-    & \textcolor{gray}{Similar to (R-ADD0).} \\
-    %
-    R-ADDK0
-    & $K_1 + \cdots + \ZEROK(\sigma) + \cdots + K_n\ \reduce K_1 + \cdots + K_n$ \\
-    & \textcolor{gray}{Similar to (R-ADDS0).} \\
-    %
-    R-ADDB0
-    & $B_1 + \cdots + \ZEROB(\sigma) + \cdots + B_n \ \reduce\ B_1 + \cdots + B_n$ \\
-    & \textcolor{gray}{Similar to (R-ADDS0).} \\
-    %
-    R-ADDO0
-    & $O_1 + \cdots + \ZEROO(\sigma, \tau) + \cdots + O_n \ \reduce\ O_1 + \cdots + O_n$ 
-    \\
-    & \textcolor{gray}{Similar to (R-ADDS0).} \\
-\end{ruletable}
-
-\begin{ruletable}{Rules for adjoint.}
-    %
-    R-ADJ0
-    & $ (X^\dagger)^\dagger \ \reduce\ X $ \\
-    %
-    R-ADJ1
-    & $ (a.X)^\dagger \ \reduce\ (a^*).(X^\dagger) $ \\
-    %
-    R-ADJ2
-    & $ (X_1 + \cdots + X_n)^\dagger \ \reduce\ X_1^\dagger + \cdots + X_n^\dagger $ \\
-    %
-    R-ADJ3
-    & $ (X \otimes Y)^\dagger \ \reduce\ X^\dagger \otimes Y^\dagger$ \\
-    %
-    R-ADJK0
-    & $ \ZEROB(\sigma)^\dagger \ \reduce\ \ZEROK(\sigma) $ \\
-    %
-    R-ADJK1
-    & $ \bra{t}^\dagger \ \reduce\ \ket{t} $ \\
-    %
-    R-ADJK2
-    & $ (B \cdot O)^\dagger\ \reduce\ O^\dagger \cdot B^\dagger $ \\
-    %
-    R-ADJB0
-    & $ \ZEROK(\sigma)^\dagger \ \reduce\ \ZEROB(\sigma) $ \\
-    %
-    R-ADJB1
-    & $ \ket{t}^\dagger \ \reduce\ \bra{t} $ \\
-    %
-    R-ADJB2
-    & $ (O \cdot K)^\dagger\ \reduce\ K^\dagger \cdot O^\dagger $ \\
-    %
-    R-ADJO0
-    & $ \ZEROO(\sigma, \tau)^\dagger \ \reduce\ \ZEROO(\tau, \sigma) $ \\
-    %
-    R-ADJO1
-    & $ \ONEO(\sigma)^\dagger \ \reduce\ \ONEO(\sigma)$ \\
-    %
-    R-ADJO2
-    & $ (K \cdot B)^\dagger \ \reduce\ B^\dagger \cdot K^\dagger$ \\
-    %
-    R-ADJO3
-    & $ (O_1 \cdot O_2)^\dagger \ \reduce\ O_2^\dagger \cdot O_1^\dagger $ \\
-\end{ruletable}
-
-\begin{ruletable}{Rules for tensor product.}
-    R-TSR0
-    & $ (a.X_1) \otimes X_2 \ \reduce\ a.(X_1 \otimes X_2) $ \\
-    %
-    R-TSR1
-    & $ X_1 \otimes (a.X_2) \ \reduce\ a.(X_1 \otimes X_2) $ \\
-    %
-    R-TSR2
-    & $ (X_1 + \cdots + X_n) \otimes X' \reduce X_1 \otimes X' + \cdots + X_n \otimes X' $ \\
-    %
-    R-TSR3
-    & $ X' \otimes (X_1 + \cdots + X_n) \reduce X' \otimes X_1 + \cdots + X' \otimes X_n $ \\
-    %
-    R-TSRK0
-    & $ K : \KType(\tau) \Rightarrow \ZEROK(\sigma) \otimes K\ \reduce\ \ZEROK(\sigma \times \tau) $ \\
-    %
-    R-TSRK1
-    & $ K : \KType(\tau) \Rightarrow K \otimes \ZEROK(\sigma)\ \reduce\ \ZEROK(\tau \times \sigma) $ \\
-    %
-    R-TSRK2
-    & $\ket{s} \otimes \ket{t} \ \reduce\ \ket{(s, t)}$ \\
-    %
-    R-TSRB0
-    & $ B : \BType(\tau) \Rightarrow \ZEROB(\sigma) \otimes B\ \reduce\ \ZEROB(\sigma \times \tau) $ \\
-    %
-    R-TSRB1
-    & $ B : \BType(\tau) \Rightarrow B \otimes \ZEROB(\sigma)\ \reduce\ \ZEROB(\tau \times \sigma) $ \\
-    %
-    R-TSRB2
-    & $\bra{s} \otimes \bra{t} \ \reduce\ \bra{(s, t)}$ \\
-    %
-    R-TSRO0
-    & $ O : \OType(\sigma, \tau) \Rightarrow O \otimes \ZEROO(\sigma', \tau') \ \reduce\ \ZEROO(\sigma \times \sigma', \tau \times \tau') $ \\
-    % 
-    R-TSRO1
-    & $ O : \OType(\sigma, \tau) \Rightarrow \ZEROO(\sigma', \tau') \otimes O \ \reduce\ \ZEROO(\sigma' \times \sigma, \tau' \times \tau) $ \\
-    %
-    R-TSRO2
-    & $\ONEO(\sigma) \otimes \ONEO(\tau)\ \reduce\ \ONEO(\sigma \times \tau)$ \\
-    %
-    R-TSRO3
-    & $ (K_1 \cdot B_1) \otimes (K_2 \cdot B_2)\ \reduce\ (K_1 \otimes K_2) \cdot (B_1 \otimes B_2) $ \\
-    %
-\end{ruletable}
-
-\begin{ruletable}{Rule for $O\cdot K$.}
-    %
-    R-MULK0
-    & $ \ZEROO(\sigma, \tau) \cdot K \ \reduce\ \ZEROK(\sigma) $ \\
-    %
-    R-MULK1
-    & $ O : \OType(\sigma, \tau) \Rightarrow O \cdot \ZEROK(\tau) \ \reduce\ \ZEROK(\sigma) $ \\
-    %
-    R-MULK2
-    & $ \ONEO(\sigma) \cdot K \ \reduce K $ \\
-    %
-    R-MULK3
-    & $ (a.O) \cdot K \ \reduce\ a.(O \cdot K) $ \\
-    %
-    R-MULK4
-    & $ O \cdot (a.K) \ \reduce\ a.(O \cdot K) $ \\
-    %
-    R-MULK5
-    & $ (O_1 + \cdots + O_n) \cdot K \ \reduce\ O_1 \cdot K + \cdots + O_n \cdot K $ \\
-    %
-    R-MULK6
-    & $ O \cdot (K_1 + \cdots + K_n) \ \reduce\ O \cdot K_1 + \cdots + O \cdot K_n $ \\
-    %
-    R-MULK7
-    & $ (K_1 \cdot B) \cdot K_2 \ \reduce\ (B \cdot K_2).K_1 $ \\
-    %
-    R-MULK8
-    & $ (O_1 \cdot O_2) \cdot K \ \reduce\ O_1 \cdot (O_2 \cdot K) $ \\
-    %
-    R-MULK9
-    & $ (O_1 \otimes O_2) \cdot ((O_1' \otimes O_2') \cdot K)\ \reduce\ ((O_1 \cdot O_1') \otimes (O_2 \cdot O_2')) \cdot K $ \\
-    %
-    R-MULK10
-    & $ (O_1 \otimes O_2) \cdot \ket{(s, t)}\ \reduce\ (O_1 \cdot \ket{s}) \otimes (O_2 \cdot \ket{t}) $ \\
-    %
-    R-MULK11
-    & $ (O_1 \otimes O_2) \cdot (K_1 \otimes K_2)\ \reduce\ (O_1 \cdot K_1) \otimes (O_2 \cdot K_2) $
-\end{ruletable}
-
-\begin{ruletable}{Rule for $B\cdot O$.}
-    %
-    R-MULB0
-    & $ B \cdot \ZEROO(\sigma, \tau) \ \reduce\ \ZEROB(\tau) $ \\
-    %
-    R-MULB1
-    & $ O : \OType(\sigma, \tau) \Rightarrow \ZEROB(\sigma) \cdot O\ \reduce\ \ZEROB(\tau) $ \\
-    %
-    R-MULB2
-    & $ B \cdot \ONEO(\sigma) \ \reduce B $ \\
-    %
-    R-MULB3
-    & $ (a.B) \cdot O \ \reduce\ a.(B \cdot O) $ \\
-    %
-    R-MULB4
-    & $ B \cdot (a.O) \ \reduce\ a.(B \cdot O) $ \\
-    %
-    R-MULB5
-    & $ (B_1 + \cdots + B_n) \cdot O \ \reduce\ B_1 \cdot O + \cdots + B_n \cdot O $ \\
-    %
-    R-MULB6
-    & $ B \cdot (O_1 + \cdots + O_n) \ \reduce\ B \cdot O_1 + \cdots + B \cdot O_n $ \\
-    %
-    R-MULB7
-    & $ B_1 \cdot (K \cdot B_2) \ \reduce\ (B_1 \cdot K).B_2 $ \\
-    %
-    R-MULB8
-    & $ B \cdot (O_1 \cdot O_2) \ \reduce\ (B \cdot O_1) \cdot O_2 $ \\
-    %
-    R-MULB9
-    & $ (B \cdot (O_1' \otimes O_2')) \cdot (O_1 \otimes O_2) \ \reduce\ B \cdot ((O_1' \otimes O_2') \cdot (O_1 \otimes O_2)) $ \\
-    %
-    R-MULB10
-    & $ \bra{(s, t)} \cdot (O_1 \otimes O_2)\ \reduce\ (\bra{s} \cdot O_1) \otimes (\bra{t} \cdot O_2) $ \\
-    %
-    R-MULB11
-    & $ (B_1 \otimes B_2) \cdot (O_1 \otimes O_2)\ \reduce\ (B_1 \cdot O_1) \otimes (B_2 \cdot O_2) $
-\end{ruletable}
-
-\begin{ruletable}{Rules for $K \cdot B$.}
-    %
-    R-OUTER0
-    & $ B : \BType(\tau) \Rightarrow \mathbf{0}_\mathcal{K}(\sigma) \cdot B\ \reduce\ \mathbf{0}_\mathcal{O}(\sigma, \tau) $ \\
-    %
-    R-OUTER1
-    & $ K : \KType(\sigma) \Rightarrow K \cdot \mathbf{0}_\mathcal{B}(\tau)\ \reduce\ \mathbf{0}_\mathcal{O}(\sigma, \tau) $ \\
-    %
-    R-OUTER2
-    & $ (a.K) \cdot B\ \reduce\ a.(K \cdot B) $ \\
-    %
-    R-OUTER3
-    & $ K \cdot (a.B)\ \reduce\ a.(K \cdot B) $ \\
-    %
-    R-OUTER4
-    & $ (K_1 + \cdots + K_n) \cdot B\ \reduce\ K_1 \cdot B + \cdots + K_n \cdot B $ \\
-    %
-    R-OUTER5
-    & $ K \cdot (B_1 + \cdots + B_n)\ \reduce\ K \cdot B_1 + \cdots + K \cdot B_n $ \\
-\end{ruletable}
-
-\begin{ruletable}{Rules for $O_1 \cdot O_2$.}
-    %
-    R-MULO0
-    & $ O : \OType(\tau, \rho) \Rightarrow \ZEROO(\sigma, \tau) \cdot O\ \reduce\ \ZEROO(\sigma, \rho) $ \\
-    %
-    R-MULO1
-    & $ O : \OType(\sigma, \tau) \Rightarrow O \cdot \ZEROO(\tau, \rho)\ \reduce\ \mathbf{0}_\mathcal{O}(\sigma, \rho) $ \\
-    %
-    R-MULO2
-    & $ \ONEO(\sigma) \cdot O \ \reduce\ O $ \\
-    %
-    R-MULO3
-    & $ O \cdot \ONEO(\sigma) \ \reduce\ O $ \\
-    %
-    R-MULO4
-    & $ (K \cdot B) \cdot O \ \reduce\ K \cdot (B \cdot O) $ \\
-    %
-    R-MULO5
-    & $ O \cdot (K \cdot B) \ \reduce\ (O \cdot K) \cdot B $ \\
-    %
-    R-MULO6
-    & $ (a.O_1) \cdot O_2 \ \reduce\ a.(O_1 \cdot O_2) $ \\
-    %
-    R-MULO7
-    & $ O_1 \cdot (a.O_2) \ \reduce\ a.(O_1 \cdot O_2) $ \\
-    %
-    R-MULO8
-    & $ (O_1 + \cdots + O_n) \cdot O'\ \reduce\ O_1 \cdot O' + \cdots + O_n \cdot O' $ \\
-    %
-    R-MULO9
-    & $ O' \cdot (O_1 + \cdots + O_n)\ \reduce\ O' \cdot O_1 + \cdots + O' \cdot O_n $ \\
-    %
-    R-MULO10
-    & $ (O_1 \cdot O_2) \cdot O_3\ \reduce\ O_1 \cdot (O_2 \cdot O_3) $ \\
-    %
-    R-MULO11
-    & $ (O_1 \otimes O_2) \cdot (O_1' \otimes O_2')\ \reduce\ (O_1 \cdot O_1') \otimes (O_2 \cdot O_2') $ \\
-    %
-    R-MULO12
-    & $ (O_1 \otimes O_2) \cdot ((O_1' \otimes O_2') \cdot O_3)\ \reduce\ ((O_1 \cdot O_1') \otimes (O_2 \cdot O_2')) \cdot O_3 $ \\  
-\end{ruletable}
-
-\begin{ruletable}{Rules for sets.}
-    %
-    R-SET0
-    & $ \mathbf{U}(\sigma) \star \mathbf{U}(\tau) \ \reduce\ \mathbf{U}(\sigma \times \tau) $
-\end{ruletable}
-
-\begin{ruletable}{Rules for sum operators.}
-    %
-    R-SUM-CONST0
-    & $ \sum_{x \in s} 0 \ \reduce\ 0 $ \\
-    %
-    R-SUM-CONST1
-    & $ \sum_{x \in s} \ZEROK(\sigma)\ \reduce\ \ZEROK(\sigma) $ \\
-    %
-    R-SUM-CONST2
-    & $ \sum_{x \in s} \ZEROB(\sigma)\ \reduce\ \ZEROB(\sigma) $ \\
-    %
-    R-SUM-CONST3
-    & $ \sum_{x \in s} \ZEROO(\sigma, \tau)\ \reduce\ \ZEROO(\sigma, \tau) $ \\
-    %
-    R-SUM-CONST4
-    & $ \ONEO(\sigma) \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)} \ket{i} \cdot \bra{i} $
-\end{ruletable}
-
-\begin{ruletable}{Rules for eliminating $\delta_{s, t}$. Note that these rules will match the $\delta$ operator modulo the commutativity of its arguments.}
-    %
-    R-SUM-ELIM0
-    & $ i \text{ free in } t \vdash \sum_{i \in \mathbf{U}(\sigma)} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} \delta_{i, t} \ \reduce\ \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n}  1$ \\
-    %
-    R-SUM-ELIM1
-    & $ i \text{ free in } t \vdash \sum_{i \in \mathbf{U}(\sigma)} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (a_1 \times \cdots \times \delta_{i, t} \times \cdots \times a_n) $ \\
-    & $ \reduce\ \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} a_1\{i/t\} \times \cdots \times a_n\{i/t\} $ \\
-    %
-    R-SUM-ELIM2
-    & $ i \text{ free in } t \vdash \sum_{i \in \mathbf{U}(\sigma)} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (\delta_{i, t}.A) \ \reduce\ \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} A\{i/t\} $ \\
-    %
-    R-SUM-ELIM3
-    & $ i \text{ free in } t \vdash \sum_{i \in \mathbf{U}(\sigma)} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (a_1 \times \cdots \times \delta_{i, t} \times \cdots \times a_n).A $ \\
-    & $ \reduce\ \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n}  (a_1\{i/t\} \times \cdots \times a_n\{i/t\}).A\{i/t\} $ \\
-    R-SUM-ELIM4
-    & $ \sum_{i \in M} \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n}  \delta_{i, j} \ \reduce\ \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} 1 $ \\
-    %
-    R-SUM-ELIM5
-    & $ \sum_{i \in M} \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (a_1 \times \dots \times \delta_{i, j} \times \cdots \times a_n) $ \\
-    & $ \reduce\ \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (a_1\{j/i\} \times \cdots \times a_n\{j/i\}) $ \\
-    %
-    R-SUM-ELIM6
-    & $ \sum_{i \in M} \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (\delta_{i, j}.A) \ \reduce\ \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} A\{j/i\} $ \\
-    %
-    R-SUM-ELIM7
-    & $ \sum_{i \in M} \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (a_1 \times \cdots \times \delta_{i, j} \times \cdots \times a_n).A $ \\
-    & $ \reduce\ \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (a_1\{j/i\} \times \cdots \times a_n\{j/i\}).A\{j/i\} $ \\
-    %
-    R-SUM-ELIM8
-    & $ \sum_{i \in M} \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} ((a_1 \times \cdots \times \delta_{i, j} \times \cdots \times a_n) + \cdots + (b_1 \times \cdots \times \delta_{i, j} \times \cdots \times b_n)).A $ \\
-    & $ \reduce\ \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} ((a_1\{j/i\} \times \cdots \times a_n\{j/i\}) + \cdots + (b_1\{j/i\} \times \cdots \times b_n\{j/i\})).A\{j/i\} $ \\
-\end{ruletable}
-
-\begin{ruletable}{Rules for pushing terms into sum operators}
-    %
-    R-SUM-PUSH0
-    & $ b_1 \times \cdots \times (\sum_{i \in M} a) \times \cdots \times b_n \ \reduce\ \sum_{i \in M} (b_1 \times \cdots \times a \times \cdots \times b_n) $ \\
-    %
-    R-SUM-PUSH1
-    & $ (\sum_{i \in M} a)^* \ \reduce\ \sum_{i \in M} a^* $ \\
-    %
-    R-SUM-PUSH2
-    & $ (\sum_{i \in M} X)^\dagger \ \reduce\ \sum_{i \in M} X^\dagger $ \\
-    %
-    R-SUM-PUSH3
-    & $ a.(\sum_{i \in M} X) \ \reduce\ \sum_{i \in M} (a.X) $ \\
-    %
-    R-SUM-PUSH4
-    & $ (\sum_{i \in M} a).X \ \reduce\ \sum_{i \in M} (a.X) $ \\
-    %
-    R-SUM-PUSH5
-    & $ (\sum_{i \in M} B)\cdot K \ \reduce\ \sum_{i \in M}(B \cdot K) $ \\
-    %
-    R-SUM-PUSH6
-    & $ (\sum_{i \in M} O)\cdot K \ \reduce\ \sum_{i \in M}(O \cdot K) $ \\
-    %
-    R-SUM-PUSH7
-    & $ (\sum_{i \in M} B)\cdot O \ \reduce\ \sum_{i \in M}(B \cdot O) $ \\
-    %
-    R-SUM-PUSH8
-    & $ (\sum_{i \in M} K)\cdot B \ \reduce\ \sum_{i \in M}(K \cdot B) $ \\
-    %
-    R-SUM-PUSH9
-    & $ (\sum_{i \in M} O_1)\cdot O_2 \ \reduce\ \sum_{i \in M}(O_1 \cdot O_2) $ \\
-    %
-    R-SUM-PUSH10
-    & $ B \cdot (\sum_{i \in M} K) \ \reduce\ \sum_{i \in M}(B \cdot K) $ \\
-    %
-    R-SUM-PUSH11
-    & $ O \cdot (\sum_{i \in M} K) \ \reduce\ \sum_{i \in M}(O \cdot K) $ \\
-    %
-    R-SUM-PUSH12
-    & $ B \cdot (\sum_{i \in M} O) \ \reduce\ \sum_{i \in M}(B \cdot O) $ \\
-    %
-    R-SUM-PUSH13
-    & $ K \cdot (\sum_{i \in M} B) \ \reduce\ \sum_{i \in M}(K \cdot B) $ \\
-    %
-    R-SUM-PUSH14
-    & $ O_1 \cdot (\sum_{i \in M} O_2) \ \reduce\ \sum_{i \in M}(O_1 \cdot O_2) $ \\
-    %
-    R-SUM-PUSH15
-    & $ (\sum_{i \in M} X_1) \otimes X_2 \ \reduce\ \sum_{i \in M} (X_1 \otimes X_2) $ \\
-    %
-    R-SUM-PUSH16
-    & $ X_1 \otimes (\sum_{i \in M} X_2) \ \reduce\ \sum_{i \in M} (X_1 \otimes X_2) $
-\end{ruletable}
-\yx{Note: because we apply type checking on variables, and stick to unique bound variables, these pushing in operations are always sound.}
-
-\begin{ruletable}{Rules for addition and index in sum}
-    %
-    R-SUM-ADDS0
-    & $\sum_{i \in M}(a_1 + \cdots + a_n) \ \reduce\ (\sum_{i \in M} a_1) + \cdots + (\sum_{i \in M} a_n) $ \\
-    % %
-    % R-SUM-ADDS1
-    % & $\sum_{i \in M}(b_1 \times \cdots \times (a_1 + \cdots + a_n) \times \cdots \times b_m) $ \\
-    % & $ \reduce\ (\sum_{i \in M} (b_1 \times \cdots \times a_1 \times \cdots \times b_m) + \cdots + (\sum_{i \in M} (b_1 \times \cdots \times a_n \times \cdots \times b_m)) $ \\
-    %
-    R-SUM-ADD0
-    & $\sum_{i \in M}(X_1 + \cdots + X_n) \ \reduce\ (\sum_{i \in M} X_1) + \cdots + (\sum_{i \in M} X_n) $ \\
-    R-SUM-ADD1
-    & $ Y_1 + \cdots + Y_n + \sum_{i \in M}(a+b).X \ \reduce\ Y_1 + \cdots + \sum_{i \in M}(a.X) + \cdots + \sum_{i \in M}(b.X) + Y_n$ \\
-    %
-    R-SUM-INDEX0
-    & $ \sum_{i \in \mathbf{U}(\sigma \times \tau)} A \ \reduce\ \sum_{j \in \mathbf{U}(\sigma)} \sum_{k \in \mathbf{U}(\tau)} A\{i/(j, k)\} $ \\
-    %
-    R-SUM-INDEX1
-    & $ \sum_{i \in M_1 \star M_2} A \ \reduce\ \sum_{j \in M_1} \sum_{k \in M_2} A\{i/(j, k)\} $
-\end{ruletable}
-
-\yx{The rules R-SUM-ADD1 to R-SUM-ADD3 needs identical sum terms, which requires the rewriting to be after the alpha normalization. We don't implement them for now.}
-
-\begin{ruletable}{Rules for Qubit}
-    %
-    R-QBIT-DELTA
-    & $\delta_{0, 1} \ \reduce\ 0$ \\
-    %
-    R-QBIT-ONEO
-    & $\ONEO(QBIT)\ \reduce\ \ket{0}\bra{0} + \ket{1}\bra{1} $ \\
-    %
-    R-QBIT-SUM
-    & $\sum_{i \in \mathbf{U}(QBIT)} A\ \reduce\ A\{i/0\} + A\{i/1\}$
-\end{ruletable}
-
-\section{Preprocessing}
-\begin{figure}
-    \begin{gather*}
-        \frac{E[\Gamma] \vdash B : \BType(\sigma) \qquad E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash B \circ K : \SType}
-    \end{gather*}
-    \caption{The typing rules for preprocessing}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\clearpage
 
@@
 
-\section{Rewriting Control and Intermediate Language}
-The associativity is already handled by the (R-FLATTEN) rule.
-In order to decide two terms $A$ and $B$ are equivalent under commutativity, we need to proof that $A$ can be transformed into $B$ with a structured permutation, which is described by the \textit{permutation tree}.
-
-\begin{definition}[permutation tree]
-    The syntax for permutation trees are inductively defined below:
-    \[
-    P ::= \mathsf{E} \mid [(i:P)^+].
-    \]
-    Here $i$ represents positive numbers.
-\end{definition}
-We always only consider \textit{well-formed} permutation trees. That is, if $P \equiv [i_1:P_1\ i_2:P_2\ \cdots\ i_n:P_n]$, then $\{i_1, ... i_n\}$ forms the set of integers from $0$ to $n-1$.
-
-We can transform a term A with a suitable permutation tree. The transformation is defined as
-
-\begin{align*}
-    \texttt{apply}(A, P) :=\ & \texttt{match}\ P\ \texttt{with} \\
-    & \quad |\ \textsf{E} \Rightarrow A \\
-    & \quad |\ [i_1:P_1\ \cdots\ i_n:P_n] \Rightarrow \texttt{A.head}(\texttt{apply(A.args[$i_1$], $P_{i_1}$)}\ \cdots\ \texttt{apply(A.args[$i_n$], $P_{i_n}$)}) \\
-    & \texttt{end}
-\end{align*}
-
-\clearpage
-
-\input{labelled}
-%\input{labelled_dirac.tex}
-
-\begin{credits}
-    \subsubsection{\ackname} A bold run-in heading in small font size at the end of the paper is
-    used for general acknowledgments, for example: This study was funded
-    by X (grant number Y).
-
-    \subsubsection{\discintname}
-    It is now necessary to declare any competing interests or to specifically
-    state that the authors have no competing interests. Please place the
-    statement with a bold run-in heading in small font size beneath the
-    (optional) acknowledgments\footnote{If EquinOCS, our proceedings submission
-    system, is used, then the disclaimer can be provided directly in the system.},
-    for example: The authors have no competing interests to declare that are
-    relevant to the content of this article. Or: Author A has received research
-    grants from Company W. Author B has received a speaker honorarium from
-    Company X and owns stock in Company Y. Author C is a member of committee Z.
-\end{credits}
+% \begin{credits}
+%     \subsubsection{\ackname} A bold run-in heading in small font size at the end of the paper is
+%     used for general acknowledgments, for example: This study was funded
+%     by X (grant number Y).
+% \end{credits}
 
@@
 %
+
+\appendix
+\include{appendix}
+
 \end{document}

### 2025-01-17 labelled.tex
2025/1/17

@@
-\newcommand{\fst}{\mathsf{fst}\ }
-\newcommand{\snd}{\mathsf{snd}\ }
-\newcommand{\var}{\mathsf{var}\ }
-\newcommand{\Var}{\mathsf{Var}\ }
-\newcommand{\reg}{\mathsf{Reg}\ }
+\newcommand{\var}{\mathsf{var}}
+\newcommand{\reg}{\mathsf{Reg}}
 \newcommand{\DType}{\mathcal{D}}
@@
 
-\lz{I suggest not adding first and second. Of course, we can, decide this later after discussion.}
+Labelled Dirac notation uses register names to indicate the quantum system of vectors and operators.
+This allows us to talk aboud the states and operations locally, without referring to the whole system.
+For instance, assume $Q$ and $R$ are two registers, we have
+\[
+    M_Q \cdot \ket{\Phi}_{(Q, R)} = ((M \otimes I) \ket{\Phi})_{(Q, R)}.
+\]
+In the left hand side, the state of two subsystems is $\ket{\Phi}$, and we apply quantum operation $M$ on the system $Q$.
 
-\lz{We further only consider constant registers.}
+In this section, we introduce the language and typing of registers and labelled Dirac notations, and demonstrate how to transform the equivalence problem into the Dirac notations studied above.
 
-Assume the name set of registers $\cR$.
+We assume the set of register symbols is $\cR$. The syntax of quantum registers is defined as follows.
 \begin{definition}[quantum registers]
@@
 
+Registers can be composed by pairs of $(R, R)$, and this structure corresponds to the structure of tensor product spaces in Dirac notations.
+
+To reason about the registers, we define their variable set as the enumeration of all register symbols involved.
+
 \begin{definition}[register variable set]
@@
 \begin{itemize}
-    \item $R\equiv r$: $\var R = \{r\}$;
-    \item $R\equiv (R_1,R_2)$: $\var R = \var R_1 \cup\var R_2$.
-\end{itemize}  
-We define the following relations for quantum registers:
-\begin{itemize}
-  \item $R$ \textbf{belongs to} $Q$, written as $R\ \text{in}\ Q$, if $\var R\subseteq \var Q$;
-  \item $R$ \textbf{is disjoint with} $Q$, written as $R \| Q$, if $\var R \cap \var Q = \emptyset$.
+    \item $R\equiv r$: $\var(R) = \{r\}$;
+    \item $R\equiv (R_1,R_2)$: $\var(R) = \var(R_1) \cup\var(R_2)$.
 \end{itemize}
 \end{definition}
-
-\lz{Since the construction of the quantum register is known, the variable set can be directly computed.
-Similarly, we use the (computable) set as dependent parameter of the type of Labelled Dirac Term.}
 
@@
 
-\begin{definition}[labelled core language]
-  The \textbf{labelled core langauge} includes all symbols in the core language of Dirac notation, as well as symbols defined below.
-  Here, $s\subseteq \cR$ is the set of register's name, it is computable and thus we do not introduce syntax for it.
+\begin{definition}[labelled Dirac notation]
+  The \textbf{labelled Dirac notation} includes all Dirac notation symbols and the generators defined below.
+  Here, $s\subseteq \cR$ is a register variable set.
   \begin{align*}
@@
     e & ::= R \mid |i\>_r \mid {}_r\<i| \mid e_R \mid e_{R;R} \mid
-    e \otimes e \mid e \cdot e \\
+    e \otimes e \otimes \cdots \otimes e \mid e \cdot e \\
   \end{align*}
-  In other words, we don't allow variables for labelled core language for now. In particular, $|i\>_r$ and ${}_r\<i|$ are used as reserved syntax for eliminating registers.
 \end{definition}
 
+$\DType(s_1, s_2)$ is the unified type for all labelled Dirac notations, where $s_1$ indicates the codomain systems and $s_2$ indicates the domain systems.
+For instance, labelled ket has type $\DType(s_1, \emptyset)$, and labelled bra has type $\DType(\emptyset, s_2)$.
+$\reg(\sigma)$ are types for registers $R$, and the index $\sigma$ indicates the type of Hilbert space represented by the register.
+Terms also include the labelled notation $e_R$ for bra, ket and $e_{R;R}$ for operators.
+We introduce new dot and tensor product symbols for labelled Dirac notations.
+In labelled Dirac notation, the structure of tensor product does not matter.
+Therefore $\otimes$ is an AC symbol.
+Following the unified type $\DType(s,s)$, all kinds of multiplications are represented by the same dot product $e \cdot e$.
+Finally, $\ket{i}_r$ and ${}_r\bra{i}$ are labelled basis for the normal form of labelled Dirac notations, where $r$ are registers symbols in $\mathcal{R}$.
+
 \subsubsection*{Typing rules}
-\lz{context? environment? Since quantum registers are not bound variables, we might just add it to the environment.}
 
-The extra syntax for the environment.
-\begin{gather}
-    E ::= \dots \mid E; r : \reg(\sigma)
-\end{gather}
+\[
+    \frac{E[\Gamma] \vdash  \sigma : \Index}{E[\Gamma] \vdash \reg(\sigma) : \Type}
+    \qquad
+    \frac{E[\Gamma] \vdash r : \reg(\sigma_r) \text{ for all $r$ in $s_1$ and $s_2$} }{E[\Gamma] \vdash \DType(s_1, s_2) : \textsf{Type}} 
+\]
+
+\[
+    \frac{E[\Gamma] \vdash R : \reg(\sigma)\qquad E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash K_R : \DType(\var R, \emptyset)}
+    \qquad
+    \frac{E[\Gamma] \vdash D : \DType(s_1,s_2)}{E[\Gamma] \vdash D^\dagger : \DType(s_2,s_1)}
+\]
+
+The typing rule for tensor product is
+\[
+    \frac{
+        E[\Gamma] \vdash D_i : \DType(s_i,s_i') \qquad
+        \bigcap_i s_i = \emptyset \qquad
+        \bigcap_i s_i' = \emptyset
+    }
+    {E[\Gamma] \vdash D_1 \otimes \cdots \otimes D_i : \DType(\bigcup_i s_i, \bigcup_i s_i')}.
+\]
+
+The typing rule for dot product is
+\[
+    \qquad
+    \frac{
+        \begin{aligned}
+            E[\Gamma] \vdash D_1 : \DType(s_1,s_1') \\
+            E[\Gamma] \vdash D_2 : \DType(s_2,s_2')
+        \end{aligned}
+        \qquad 
+        \begin{aligned}
+            s_1 \cap s_2 \backslash s_1' = \emptyset \\
+            s_2' \cap s_1' \backslash s_2 = \emptyset
+        \end{aligned}
+    }
+    {E[\Gamma] \vdash D_1\cdot D_2 : \DType(s_1 \cup (s_2\backslash s_1'), s_2' \cup (s_1'\backslash s_2))}.
+\]
+
+\subsection{Normalization}
+The main idea of normalizing labelled Dirac notations is \textit{label elimination}. It expands $e_R$ and $e_{R;R}$ by tensor products of labelled basis, which are formalized as rules below.
+
+\begin{align*}
+    & K_R \ \reduce\ \sum_{i_{r_1}\in\bU(\sigma_{r_1})}\cdots \sum_{i_{r_n}\in\bU(\sigma_{r_n})} (\<i_R|\cdot K). (|i_{r_1}\>_{r_i}\otimes\cdots\otimes|i_{r_n}\>_{r_n}) \\
+    %
+    & B_R \ \reduce\ \sum_{i_{r_1}\in\bU(\sigma_{r_1})}\cdots \sum_{i_{r_n}\in\bU(\sigma_{r_n})} (B\cdot |i_R\>). ({}_{r_1}\<i_{r_1}|\otimes\cdots\otimes{}_{r_n}\<i_{r_n}|) \\
+    %
+    & O_{R,R'} \ \reduce\ \sum_{i_{r_1}\in\bU(\sigma_{r_1})}\cdots \sum_{i_{r_n}\in\bU(\sigma_{r_n})}
+    \sum_{i_{r'_1}\in\bU(\sigma_{r'_1})}\cdots \sum_{i_{r'_{n'}}\in\bU(\sigma_{r'_{n'}})} \\
+    & \qquad (\<i_R|\cdot O\cdot |i_{R'}\>).(|i_{r_1}\>_{r_i}\otimes\cdots\otimes|i_{r_n}\>_{r_n} \otimes {}_{r'_1}\<i_{r'_1}|\otimes\cdots\otimes{}_{r'_{n'}}\<i_{r'_{n'}}|)
+\end{align*}
+
+\begin{align*}
+    & \textrm{(R-SUM-PUSHD0)}
+    && X_1 \otimes \cdots (\sum_{i \in M} D) \cdots \otimes X_2\ \reduce\ \sum_{i \in M} (X_1 \otimes \cdots D \cdots \otimes X_n) \\
+    %
+    & \textrm{(R-SUM-PUSHD1)}
+    && (\sum_{i \in M} D_1) \cdot D_2 \ \reduce\ \sum_{i \in M} (D_1 \cdot D_2) \\
+\end{align*}
+
+\begin{align*}
+    %
+    & \textrm{(R-L-SORT0)}
+    && A : \DType(s_1, s_2), B : \DType(s_1', s_2'), s_2 \cap s_1'=\emptyset \Rightarrow A \cdot B \ \reduce\ A \otimes B \\
+    %
+    & \textrm{(R-L-SORT1)}
+    && {}_r\bra{i}\cdot\ket{j}_r \ \reduce\ \delta_{i, j} \\
+    %
+    & \textrm{(R-L-SORT2)}
+    && {}_r\bra{i}\cdot(Y_1 \otimes \cdots \otimes \ket{j}_r \otimes \cdots \otimes Y_m) \ \reduce\ \delta_{i, j}.(Y_1  \otimes \cdots \otimes Y_m) \\
+    %
+\end{align*}
 

### 2025-01-17 main.tex
2025/1/17

@@
 
+\input{lstlisting}
+
 \usepackage{tikz}
@@
 }
+
+% define C++ logo
+\newcommand{\CC}{C\nolinebreak\hspace{-.05em}\raisebox{.4ex}{\tiny\bf +}\nolinebreak\hspace{-.10em}\raisebox{.4ex}{\tiny\bf +}}
+\def\CC{{C\nolinebreak[4]\hspace{-.05em}\raisebox{.4ex}{\tiny\bf ++}}}
 
@@
 
-Dirac notations are widely used in theoretical reasonings of quantum computation and quantum information.
-The previous work proved that the equivalence of basic Dirac notations is decidable, and provided a
-corresponding term rewriting system.
-This work focuses on developing the theory and tool that practically solves this problem. Compared to the
-previous result, this work includes a typing system and a simplified term rewriting system.
-A much more efficient algorithm.
-And a language and algorithm for labelled Dirac notation is proposed as the final contribution.
-We implement the whole algorithm in C++ with a Mathematica backend for theories of scalars, which 
-decides the 158 examples in seconds.
+    Dirac notations is the fundamental language in quantum computation and quantum information reasoning.
+Recently, the term rewriting system DiracDec~\cite{diracdec} was introduced to automate equational reasoning with Dirac notations, a critical yet time-intensive component of formalization tasks.
+This work bases upon and extends DiracDec, aiming to develop a solver optimized for practical applications.
+Enhancements include an improved typing system, a simplified language and rewriting system, more efficient algorithms, and added support for labelled Dirac notations.
+The solver Dirace, implemented in \CC\ with a Mathematica backend, demonstrates superior decision-making power and computational efficiency, successfully addressing all examples in its benchmark suite.
 
@@
 For instance, $a\ket{\psi} + b\ket{\phi}$ indicates the addition of two vectors, i.e., the superposition of two states $\ket{\psi}$ and $\ket{\phi}$.
-Dirac notation is also widely accepted as the working language in quantum computation and quantum information.
-The reasonings of Dirac notations play a fundamental role, but there was few works on automating this procedure.
+Dirac notation is now widely accepted as the working language in quantum computation and quantum information.
+The reasonings of Dirac notation play a fundamental role in research and application, just like boolean and integer logic to classical computer science.
+For example, formalization works of quantum algorithms and quantum programming languages involves plenty of equational proofs of Dirac notation, which are critical, repeating but time-intensive jobs.
+Dirac notation are also used in quantum programming languages to define the program states, operations and assertions.
+In order to automate the verification procedure, we need to simplify and check the equivalence of pre-conditions.
+However, unlike the existence of SAT and SMT solver, we are still in need of a working Dirac notation solver, and this absence has become an obstacle of many areas.
 
-The absence of a working Dirac notation solver is also a main obstacle for proof automation of quantum programming languages.
-In these works, Dirac notation are used to define the program states, operations and assertions. In order to automate the verification procedure,
-we need to simplify and check the equivalence of pre-conditions.
+Recently, Xu et al.~\cite{diracdec} proposed a theory to decide the equivalence of Dirac notations, as well as a prototype implementation in Mathematica called DiracDec.
+They proved that the equivalence of basic Dirac notations are decidable.
+Their decision procedure sticks with a pure term rewriting system, which allows to prove important properties such as confluence and termination of the system.
+Even though, there is still a gap between DiracDec and a practical solver for Dirac notation equivalence.
 
-Recently, the work by Yingte et al.[] provides a theory to decide the equivalence of Dirac notations, as well as a prototype implementation in Mathematica.
-They proved that the equivalence of basic Dirac notations are decidable.
-Their system sticks with a pure term rewriting system, which allows to prove important properties such as confluence and termination.
-Even though, there is still a gap to a practical solver for Dirac notation equivalence.
-The Mathematica implementation is hard to be integrated into other tools. 
-The second concern is the efficiency. To decide the theory $E$, \DiracDec\ iterates all possible permutations, which is simple and direct but has factorial complexity.
-And lastly, the previous work does not consider labelled Dirac notation.
-The original work cannot intergrate with well with the Scalar system in mathematica.
-And there were a few examples that failded to decide.
+One problem is the algorithm efficiency for equivalence out of the scope of rewritings.
+DiracDec decides the whole equational theory by rewriting modulo $E$, where $E$ is a set of equational axioms that cannot be decided by normalization in rewritings, e.g.
+the associativity and commutativity (AC) of some function symbols.
+DiracDec uses a direct by inefficient algorithm to decide $E$, which search through all possible permutations and has factorial complexity.
+This is witnessed in their evaluation: DiracDec is not good at dealing with "computational" examples, which has many AC symbols and are therefore time consuming.
 
-This work refines the pure term-rewriting based system into a hybrid decision procedure,
-overcomes the problems mentioned above and focuses on practical
-automated equational reasonings. Our main contributions are:
+Another problem is the lack of labelled Dirac notations.
+which makes the reasoning more convenient and scaleable.
+
+One last problem is about usability.
+The typing of DiracDec only deals with Dirac notations, which is not sufficient for a working scenario, where defined symbols are desirable, and variables have their typing assumptions in the context.
+Moreover, to avoid type checking during term rewriting, DiracDec uses different symbols for the same casual notation to disambiguate, which 
+is not close to the natural way of using the langauge, and makes the system bloated.
+It is hard to integrate the Mathematica implementation as solvers into into other tools.  
+Although DiracDec solve most examples in the benchmark, including some very complicated equations, there are still a few examples that DiracDec failds to decide.
+
+Aimming at building a practical solver, this work bases on and extends the DiracDec result, transform the term rewriting system into a hybrid decision procedure, and overcomes the problems mentioned above.
+Our main contributions and techniques are:
 \begin{itemize}
-    \item We propose a rigorous typing system for Dirac notation. Defined symbols (e.g., transpose and trace) are modelled by functions.
-        Also, the typing system removes the difficulty of disambiguation, therefore the language and the term rewriting system are much simplified.
-    \item An efficient algorithm to decide the extra equational theories $E$ is provided.
+    \item An efficient algorithm to decide the extra equational theories $E$.
 The basic idea is normalization by sorting, which considers the interplay of AC symbols, SUM-SWAP and $\alpha$-equivalence.
 This algorithm reduces the complexity from the factorial level to the polynomial level.
-    \item The language to support variable labels is proposed.
+    \item The support for constant register labels. The equivalence decision of labelled Dirac notations is then reduced into the unlabelled situation.
+    \item A more usable \CC\ solver Dirace with usable language and typing. Defined symbols (e.g., transpose and trace) are modelled by functions.
 \end{itemize}
 
-\paragraph{Related Work} Other previous works explore the language and decision procedure to express quantum computation differently.
+As an evaluation, we tested Dirace on the benchmark of DiracDec and new example for labelled Dirac notations.
+Dirace successfully decides all the examples that are expressable in its language, including those failed by DiracDec because of complexity or insufficient decision power.
 
@@
 Quantum states live in complex Hilbert spaces. We use vectors in the space to descirbe pure quantum states, and operations corresponds to linear transformations. 
-In 1939, Dirac proposed his notation~\cite{dirac1939new} for quantum mechanics, which is essentially a wrapping of the linear algebra language. 
-Dirac notation uses the ket $\ket{i}$ and the bra $\bra{i}$ to indicate bases of the space and the dual space.
+Dirac notation are used to express the linear algebra formula in a convenient and compat way.
+It uses the ket $\ket{i}$ and the bra $\bra{i}$ to indicate bases of the space and the dual space.
 Together with other variable symbols, they are composed with each other in sequence, and the composition will be interpreted into different operations, depending on the type of operands.
 For example, $\braket{i|j}$ represents the inner product of $\bra{i}$ and $\ket{j}$, which is a scalar, while $\ket{i}\bra{j}$ represents the outer product, resulting in an operator.
+Dirac notation also use $\otimes$ to indicate the vectors and operators in the tensor product space.
+
+Dirac notation further enjoys the property that the interpretation is independent on the order of composition, thus parentheses can be omitted.
+For example, the formula \(\bra{i}\ket{\phi}\bra{\psi}\ket{j}\) can be understood as
 \[
@@
 and they are equivalent for all variables.
-Dirac notation also use $\otimes$ to indicate the vectors and operators in the tensor product space.
 
-In this manner, we can express long formulae in succint Dirac notations.
-The notation further enjoys the property that the interpretation is independent on the order of composition, thus parentheses can be omitted.
-For example, the formula \(\bra{i}\ket{\phi}\bra{\psi}\ket{j}\) can be understood as
+With the concrete basis, Dirac notations can be interpreted as matrices.
+For example, the ket of two basis of qubit states are column vectors $\ket{0} = \begin{bmatrix} 1 \\ 0 \end{bmatrix}$ and $\ket{1} = \begin{bmatrix} 0 \\ 1 \end{bmatrix}$, and the corresponding bras are row vectors in the dual space $\bra{0} = \begin{bmatrix} 1 & 0 \end{bmatrix}$ and $\bra{1} = \begin{bmatrix} 0 & 1 \end{bmatrix}$.
+The calculation between Dirac notations then corresponds to matrix operatoins.
+For example, inner product $\braket{0| 1} = \begin{bmatrix} 0 & 1 \end{bmatrix} \begin{bmatrix} 1 \\ 0 \end{bmatrix} = 0$, and tensor product $\ket{0} \otimes \ket{1} = \ket{(0, 1)} = \begin{bmatrix} 1 \\ 0 \end{bmatrix} \otimes \begin{bmatrix} 0 \\ 1 \end{bmatrix} = \begin{bmatrix} 0 \\ 1 \\ 0 \\ 0 \end{bmatrix}$.
 
-With the concrete basis, Dirac notations can be interpreted as matrices. For example ...
+In practice, Dirac notations are combined with other syntax for better expressiveness.
+For example, the big operator sum $\sum_{i \in S} A$ are used in theoretical reasonings.
+We also use labelled Dirac notation, where Dirac notations are subscripted with registers to denote the quantum system in consideration, e.g.
+$\ket{i}_q \otimes \ket{i}_r$.
 
-We use universal algebra and equational logic to formally represent Dirac notations and the reasoning procedure in computers.
+We use universal algebra and equational logic to formally represent Dirac notations and the reasoning procedure.
+A universal algebra defines a signature of function symbols and the axioms.
+Terms in the algebra are constructed by constants, variables or function application on terms.
+In our case of Dirac notation, the signature consists of constructors and operations like $\ket{i}$ or $A \otimes B$, and the axioms Dirac notation equivalenc rules with equational logic.
 
-\yx{mention the dilemma of simplicity and efficiency}
+Here is an example proof of a equivalence lemma.
+\begin{example}
+    Let $q$ and $r$ be two quantum systems on $\mathcal{H}_T$ space. Let $M$ be a quantum operation on $\mathcal{H}_T$,
+    and $\ket{\Phi} = \sum_{i \in T} \ket{i}\otimes \ket{i}$ be the maximally entangled state, then we have
+    \[
+    M_q \ket{\Phi}_{q; r} = M_r^T \ket{\Phi}_{q; r}.
+    \]
+\end{example}
+
+\yx{I guess it's wise to avoid mentioning that we don't impose fixed arity.}
+
+% \yx{mention the dilemma of simplicity and efficiency}
 
@@
 The syntax of Dirac notations involves three layers: the index, the type and the term.
-Index represents classical datatypes, and they appear in type expression to denote the type of Hilbert spaces and sets.
+Terms describe the concrete instances like ket, bra and operator, which will be checked and typed.
+Index represents classical datatypes, and they appear in type expression to distinguish between different Hilbert spaces and sets.
 
-\begin{definition}[language syntax]
-    The syntax for type indices are defined as
+\begin{definition}[index syntax]
+    The syntax for type indices is
     \begin{align*}
-        \sigma ::=\ & x \mid \sigma_1 \times \sigma_2 \mid \mathsf{Bit}.
+        \sigma ::=\ & x \mid \sigma_1 \times \sigma_2 \mid \mathsf{bool}.
     \end{align*}
-    The syntax for Dirac notation types is defined as 
+\end{definition}
+Here $x$ is a variable. $\sigma_1 \times \sigma_2$ is the product type, which is used to represent the type of tensor product space, or Cartesian product sets.
+$\mathsf{bool}$ is a constant index with two basis, and is used in Dirac notations of qubit systems.
+
+\begin{definition}[type syntax]
+    The syntax for Dirac notation types is
     \begin{align*}
@@
     \end{align*}
-    The syntax for Dirac notation terms is defined as
+\end{definition}
+$\Basis(\sigma)$ is the type for basis in index $\sigma$.
+$\SType$ indicates scalars, $\KType(\sigma)$ and $\BType(\sigma)$ indicates ket and bra terms in Hilbert space $\sigma$, and $\OType(\sigma_1, \sigma_2)$ indicates linear operators with Hilbert space $\sigma_2$ as domain and $\sigma_1$ as codomain.
+$\mathsf{Set}(\sigma)$ is the type of subsets in $\sigma$ index, and is used to denote the values of bound variables in summation.
+The remaining two generators are types for functions.
+$T_1 \to T_2$ denotes normal functions, which accepts a $T_1$ type argument and results in a $T_2$ type term.
+$\forall x.
+T$ denotes index functions, which accepts an index argument $x$ and results in a $T$ type term, where $T$ can depend on $x$.
+
+\begin{definition}[term syntax]
+    The syntax for Dirac notation terms is
     \begin{align*}
@@
         & |\ \hat{0} \mid \hat{1} \mid (e_1, e_2) \\
-        & |\ 0 \mid 1 \mid \ADDS(e_1 \cdots e_n)  \mid e_1 \times \cdots \times e_n \mid e^*  \mid \delta_{e_1, e_2} \mid \DOT(e_1\ e_2) \\
+        & |\ 0 \mid 1 \mid \ADDS(e_1, \cdots, e_n)  \mid e_1 \times \cdots \times e_n \mid e^*  \mid \delta_{e_1, e_2} \mid \DOT(e_1, e_2) \\
         & |\ \ZEROK(\sigma) \mid \ZEROB(\sigma) \mid \ZEROO(\sigma_1, \sigma_2) \mid \ONEO(\sigma) \\
-        & |\ \ket{e} \mid \bra{t} \mid e^\dagger \mid e_1.e_2 \mid \ADD(e_1 \cdots e_n) \mid e_1 \otimes e_2 \\
-        & |\ \MULK(e_1\ e_2) \mid \MULB(e_1\ e_2) \mid \OUTER(e_1\ e_2) \mid \MULO(e_1\ e_2) \\
+        & |\ \ket{e} \mid \bra{t} \mid e^\dagger \mid e_1.e_2 \mid \ADD(e_1, \cdots, e_n) \mid e_1 \otimes e_2 \\
+        & |\ \MULK(e_1, e_2) \mid \MULB(e_1, e_2) \mid \OUTER(e_1, e_2) \mid \MULO(e_1, e_2) \\
         & |\ \mathbf{U}(e) \mid e_1 \star e_2 \mid \sum_{e_1} e_2.
@@
 \end{definition}
- % after consideration I still think keep DOT as binary is the best choice. It reduces the unstability of complicated matching in sequences.
-Here $i$ is a natural number and $\$i$ represents the $i$-th bound variable in de Bruijn notation. 
-Compared to [?], this syntax for Dirac notations merges the symbols with overlapped properties, such as the addition and scaling symbols for ket, bra and operator.
+
+The generators are explained in order.
+$\lambda x : T.e$ is the abstraction for normal functions, and $\mu x.e$ is the abstraction for index functions.
+$e_1\ e_2$ is function application. 
+$e_1 \circ e_2$ is the composition operator of Dirac notations.
+$\hat{0}$ and $\hat{1}$ are constant basis of $\Basis(\mathsf{bool})$ type.
+$(e_1, e_2)$ is the basis pair for product types. $0$, $1$, $\ADDS$, $e_1 \times \cdots \times e_n$ and $e^*$ are symbols for scalars. 
+The next line includes constant symbols for ket, bra and operator.
+$e^\dagger$ denotes the conjugate transpose of $e$. $e_1.e_2$ denotes scaling the term $e_2$ with scalar $e_1$.
+$\mathbf{U}(e)$ is the universal set with index $e$.
+$e_1 \star e_2$ is the Cartesian product.
+$\sum_{e_1} e_2$ is the big operator sum, which is modelled by folding the function $e_2$ over value sets $e_1$.
+Usually, the sum body is specified by an abstraction.
+Therefore we use the notation $\sum_{x \in s} X$ to denote $\sum_{s} \lambda x : T .
+X$ as well.
+
 Here $\ADDS$ and $\ADD$ are two different AC symbols representing the scalar addition and the linear algebra addition respectively. 
-They will be denoted as $a_1 + \cdots + a_n$ and $X_1 + \cdots + X_n$.
-There are five kinds of linear algebra multiplications among ket, bra and operator, whose properties are similar but still diverge to some extent.
-For example, the rules $(O_1 \cdot O_2) \cdot K\ \reduce\ O_1 \cdot (O_2 \cdot K)$ and $B \cdot (O_1 \cdot O_2) \ \reduce\ (B \cdot O_1) \cdot O_2$ indicate that the sorting of multiplication sequences depends on the subterm types.
-To avoid frequent but unnecessary type checkings, we encode the typing information by using five different symbols, namely $\DOT$, $\MULK$, $\MULB$, $\OUTER$ and $\MULO$.
+% They will be denoted as $a_1 + \cdots + a_n$ and $X_1 + \cdots + X_n$.
+There are five kinds of linear algebra multiplications among ket, bra and operator.
+We encode the typing information by using five different symbols, namely $\DOT$, $\MULK$, $\MULB$, $\OUTER$ and $\MULO$.
 They are denoted as $B\cdot K$, $K_1 \cdot K_2$, $B_1 \cdot B_2$, $K \cdot B$ and $O_1 \cdot O_2$, respectively.
 
-Usually, the sum body is specified by an abstraction. Therefore we use notation $\sum_s X$ to denote $\sum_{x \in s} \lambda x : T . X$ as well.
+Compared to DiracDec, our language for Dirac notations merges the symbols with identital properties, such as the addition and scaling symbols for ket, bra and operator.
+One important difference is that AC symbols are treated as functions with indefinite arities.
 
@@
 
-And lastly we have the typing rules for composition $x \circ y$. As is the case for casual Dirac notation, the typing of composition depends on the types of operands.
+And lastly we have the typing rules for composition $x \circ y$. Similar to casual Dirac notations, the typing of composition depends on the types of operands.
 \begin{gather*}
@@
 \end{lemma}
+
+\begin{proof}
+    The type can be calculated in the recursive manner. There is at most one typing rule for the same function symbol and argument types, which leads to the uniqueness of typing.
+\end{proof}
 
@@
 valid rewriting operation, and two expressions are equiavlent if and only if they can be rewritten into the same form using the axioms.
+
+\yx{where should we put the example?}
+To conclude the design introduction, we demonstrate part of the formalization of the preliminary example.
+\begin{example}[formalizing the preliminary example]
+    Definitions and assumptions in the environment:
+    \begin{align*}
+        & \textrm{TPO} && := \mu T_1. \mu T_2. \lambda O : \OType(T_1, T_2). \sum_{i \in \mathbf{U}(T_1)} \sum_{j \in \mathbf{U}(T_2)} \bra{i} O \ket{j} . \ket{j}\bra{i} \\
+        & && : \forall T_1. \forall T_2. \OType(T1, T2) \to \OType(T2, T1); \\
+        & \textrm{phi} && := \mu T. \sum_{i \in \mathbf{U}(T)} \sum_{j \in \mathbf{U}(T)} \ket{(i, j)} : \forall T.\KType(T \times T); \\
+        & \textrm{T} && : \Index; \\
+        & \textrm{M} && : \OType(T, T).
+    \end{align*}
+\end{example}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+As a typical example, the two terms 
+\[
+    \sum_{i \in s_1} \sum_{j \in s_2} \bra{i}A\ket{j} \times \bra{j}B\ket{i}
+    = 
+    \sum_{i \in s_2} \sum_{j \in s_1} \bra{i}B\ket{j} \times \bra{j}A\ket{i}
+\]
+are equivalent due to the swap of summations and commutativity of scalar multiplication.
+
+We propose an algorithm to ...
+
+To describe the algorithm, we introduce two notations: for a term $e = f(a_1, a_2, \cdots, a_n)$, $\textrm{head}(e)$ indicates the function symbol $f$,
+and $\textrm{arg}(e, i)$ indicates the $i$-th argument $a_i$. Variables and constants are considered as functions with zero arguments.
+\begin{definition}[order modulo bound variables]
+    Assume we have a total order of all symbols.
+    Let $\mathcal{B}$ be the set of bound variables.
+    For simplicity, we assume all bound variables are unique. 
+    The relation $e_1 < e_2$ between the two terms holds when:
+    \begin{itemize}
+        \item $e_1 \notin \mathcal{B}$ and $e_2 \in \mathcal{B}$, or
+        \item $head(e_1) < head(e_2)$, or
+        \item $head(e_1) = head(e_2)$, and there exists $n$ with $arg(e_1, n) < arg(e_2, n)$, and for all $i < n$, either $arg(e_1, i) = arg(e_2, i)$, or $arg(e_1, i) \in \mathcal{B}$ and $arg(e_2, i) \in \mathcal{B}$.
+    \end{itemize}
+\end{definition}
+
+The order depends on there occurances in the last sorting result. If no occurance, then the order will depend on the set (for sum) and the type (for lambda abstraction only).
+
+%%%%%%%%%%%%%%%%%%%%%%%
+
+The idea is to assign an order to terms, which is independent on the bound variables. Because we can have terms with nested AC symbols.
+
+And in the end, the $\alpha$-equivalence are checked by de Bruijn index~\cite{deBruijn1972lambda}, where references to bound variable names are replaced by the distance of the lambda abstraction to the variable.
+For example, the nominal lambda abstraction $\lambda x.
+x$ is transformed into $\lambda .
+1$, while $\lambda x.
+\lambda y.
+(x\ (y\ x))$ will is transformed into $\lambda.\lambda.
+(2\ (1\ 2))$.
+Therefore, transformation into de Bruijn index is at the last step of the normalization to check the equivalence of terms with different bound variable names.
+
 At last, we will prove that equivalence by this normalization procedure is sound in the semantics.
@@
 
+The main purpose of this work is to build a practical tool that works well in checking Dirac notation equations.
+The refinements and extensions above concludes in our implementation called Dirace.
+Dirace is a solver written in \CC\ working as a command line tool.
+It has a parser built by ANTLR4, and scalar reasonings are powered by a Mathematica kernel.
+The user can use commands to make definitions and assumptions in the maintained context, conduct the normalization and equivalence checking, and obtain the rewriting trace output.
+This implementation
+is tested on the benchmark of the DiracDec work, and succeeds in proving most of them efficiently.
+
+Here is the code for the preliminary example.
+
+    \begin{lstlisting}[style=dirace]
+Var T : INDEX. Var M : OTYPE[T, T].
+Def phi := idx T => Sum nv in USET[T], |(nv, nv)>.
+Var r1 : REG[T]. Var r2 : REG[T].
+CheckEq M_r1;r1 (phi T)_(r1, r2) with (TPO T T M)_r2;r2 (phi T)_(r1, r2).
+    \end{lstlisting}
+
+The first three lines use the \texttt{Var} and \texttt{Def} commands to set up the context for the Dirac notation.
+Here \texttt{T} is a type index, representing arbitrary Hilbert space type.
+\texttt{M} is assume to be an operator in Hilbert space with type \texttt{T}.
+\texttt{phi} is defined as the maximally entangled state.
+
+Besides, the symbol \texttt{TPO} represents transpose of operators, and the definition is provided as
+\begin{lstlisting}[style=dirace]
+Def TPO := idx sigma => idx tau => fun O : OTYPE[sigma, tau] => Sum i in USET[sigma], Sum j in USET[tau], (<i| O |j>).(|j> <i|).
+\end{lstlisting}
+
+And the output result is:
+    \begin{lstlisting}[style=dirace]
+The two terms are equal.
+[Normalized Term] SUM[USET[T], FUN[BASIS[T], SUM[USET[T], FUN[BASIS[T], SCR[DOT[BRA[$\texttt{\$1}$], MULK[M, KET[$\texttt{\$0}$]]], LTSR[LKET[$\texttt{\$1}$, r1], LKET[$\texttt{\$0}$, r2]]]]]]] : DTYPE[RSET[r1, r2], RSET]
+    \end{lstlisting}
+
+The normal form can be transformed into the formatted version:
+
+\[
+\sum_{\mathbf{U}(T)} \sum_{\mathbf{U}(T)} \bra{\$1}M\ket{\$ 0} . \ket{\$1}_{r1} \otimes \ket{\$0}_{r2} : \DType(\{r1, r2\}, \emptyset)
+\]
+Here $\$0$ and $\$1$ are de Bruijn indices. The result is a ket on the \{r1, r2\} system as expected.
+
+\subsection{Benchmark performance}
+It does not support fst and snd projectors.
+It does not support interactive proof, as is used in the reasonings of quantum programming language theory.
+
+\section{Related Work}
+
+Other previous works explore the language and decision procedure to express quantum computation differently.
+
 \section{Conclusion}
+Based on the first Dirac notation equational reasoning tool DiracDec, this work refines and extends the theory for practical application, and provides the solver Dirace.
+Experimental results shows that the tool can correctly identify all encoded equivalent expressions efficiently.
+We expect Dirace will play a role in areas like quantum program verification or proofs of post-quantum cryptography protocols.
 
@@
 
-\section{Consideration}
-\begin{itemize}
-    \item Sorting requires that alpha equivalent terms have the same syntax.
-    \item de Bruijn expression satisfies this requirement, but the typing and substitution becomes very complicated in the type index scenario.
-    \item only compute de Bruijn form in the equivalence checking and sorting phase.
-    \item pipeline: 
-        \begin{enumerate}
-            \item preprocessing
-            \item rename unique bound variable names
-            \item 1st rewritings
-            \item variable expansion
-            \item 2nd rewritings 
-            \item sorting modulo bound variables
-            \item sorting successive sum.
-The order depends on there occurances in the last sorting result.
-If no occurance, then the order will depend on the set (for sum) and the type (for lambda abstraction only).
-            \item computing de Bruijn
-        \end{enumerate}
-
-    \item I found that typing of inner bound variables also requires modifying the context.
-    Need to modify the tree visiting algorithm
-    (Typed term rewriting with bound variables is not that easy)    
-\end{itemize}
-
 \section{TODO}
 \begin{itemize}
-    \item implemented trace output for the whole pipeline.
     \item better output.
-    \item add Mathematica support for scalars
-    \item output trace to Coq
-    \item consider labelled Dirac notations
-    \item better trace output
     \item better error logic
-    \item better format output
-    \item remove term bank and hash consing
+    \item the $\bar{U_{AC}}$, $\bar{U_{BC}}$ terms can be modelled by quantum registers.
 \end{itemize}
@@
     \item In each Delta reduction, the bound variables are replaced with unique variables. This should help solve the problem of conflicting variable names during substitution.
-    \item I guess it may still be necessary to try all different bound variable assignments.
-    \item the $\bar{U_{AC}}$, $\bar{U_{BC}}$ terms can be modelled by quantum registers.
 \end{itemize}

### 2025-01-18 main.tex
2025/1/17

@@
 They proved that the equivalence of basic Dirac notations are decidable.
-Their decision procedure sticks with a pure term rewriting system, which allows to prove important properties such as confluence and termination of the system.
+Their algorithm sticks with a pure term rewriting system, which allows to prove important properties such as confluence and termination of the system.
 Even though, there is still a gap between DiracDec and a practical solver for Dirac notation equivalence.
@@
 
-Aimming at building a practical solver, this work bases on and extends the DiracDec result, transform the term rewriting system into a hybrid decision procedure, and overcomes the problems mentioned above.
+Aimming at building a practical solver, this work bases on and extends the DiracDec result, transform the term rewriting system into a hybrid algorithm, and overcomes the problems mentioned above.
 Our main contributions and techniques are:
@@
 
-Semantics assign the meanings to the expressions, and the goal of the decision procedure is to decide whether
-two expressions have the equivalent semantics. We consider two ways of definition: the axiomatic equations and the denotational semantics.
+Semantics define the meanings of expressions, and the goal of the desirable algorithm is to decide whether
+two expressions have the equivalent semantics. We can give the semantics by axiomatic equations and the denotational interpretations.
 
-The denotational way interprets every expression as a linear algebra concept, and equivalence is considered in the common mathematical sense.
-This explanation formalizes the original definition of Dirac notations, and best describes the target of the decision procedure.
+\subsubsection{Denotational semantics} The denotational way interprets every expression as a set theory object in linear algebra, according to a valuation mapping $v$ from variables to their values.
+This explanation formalizes the common explanation of Dirac notations, and best describes the target of the algorithm.
+The semantics of $e$ with valuation $v$ is written as $\sem{e}_v$, and two expressions $e_1$ and $e_2$ are equivalence if they have equal semantics in the mathematical sense for all valuations, written as $\sem{e_1} = \sem{e_2}$.
 
-The semantics by equations, on the other hand, is an abstraction and axiomization. From the operational view, each equation declares a 
-valid rewriting operation, and two expressions are equiavlent if and only if they can be rewritten into the same form using the axioms.
+Variables typed with $\Index$ are interpreted as finite sets, and index product $\sem{\sigma_1 \times \sigma_2} = \sem{\sigma_1} \times \sem{\sigma_2}$ are defined as Cartesian product of sets.
+Following on, every type is interpreted as a set.
+For example, the scalar type $\sem{\SType} = \mathbb{C}$ is interpreted as complex number set, the the ket and bra type $\sem{\KType(\sigma)} = \mathcal{H}_{\sem{\sigma}}$, $\sem{\KType(\sigma)} = \mathcal{H}_{\sem{\sigma}}^*$ are interpreted as the Hilbert space and dual space depending on $\sigma$.
+One special case is the delta symbol
+$\sem{\delta_{s, t}} = 
+\left \{  
+    \begin{array}{ll}
+        1, & \text{where } \sem{s} = \sem{t}, \\
+        0, & \text{where } \sem{s} \neq \sem{t}.
+    \end{array} 
+\right .
+$.
+
+The idea behind the interpretation of types and terms is to characterize the typing relation with set theory inclusion. Therefore we have the following lemma.
+\begin{lemma}
+    For all well-formed context $E[\Gamma]$, term $t$ and type $T$, if $E[\Gamma]\vdash t : T$, then for all valuations $v$, $\sem{t}_v \in \sem{T}_v$.
+\end{lemma}
+
+\subsubsection{Axiomatic semantics} 
+An other way is to express the equivalence by a set of axioms. 
+% From the operational view, each equation declares a valid rewriting operation, and two expressions are equiavlent if and only if they can be rewritten into the same form using the axioms.
+For example, we have the absorption law of zero symbols $X \cdot \mathbf{0} = \mathbf{0}$, and the bilinearity of dot product
+\begin{align*}
+    (a.X) \cdot Y = a.(X \cdot Y) \qquad X \cdot (Y_1 + Y_2) = X \cdot Y_1 + X \cdot Y_2 \\
+    X \cdot (a.Y) = a.(X \cdot Y) \qquad (X_1 + X_2) \cdot Y = X_1 \cdot Y + X_2 \cdot Y
+\end{align*}
+
+Note that some equational theories cannot be decided by term rewriting, and are collected in the set $E$:
+
+\begin{align*}
+    & \textrm{AC-equivalence} 
+    && \textrm{e.g.} \qquad a + b = b + a \qquad (a + b) + c = a + (b + c) \\
+    & \textrm{$\alpha$-equivalence}
+    && \lambda x . A = \lambda y . A\{x/y\} \\
+    & \textrm{(SUM-SWAP)}
+    && \sum_{i \in s_1} \sum_{j \in s_2} A = \sum_{j \in s_2} \sum_{i \in s_1} A
+\end{align*}
+
+The equations should also include axioms for scalars. In the implementation, we use the Mathematica kernel to decide scalar equivalences, so the axioms are not given here.
+
+The equational axioms provide an operable theory for the proof automation algorithm,
+and the denotational semantics can be considered as one model for the theory.
+In this sense, the equivalence derived by axioms is weaker than the denotational semantics.
+
+\begin{lemma}
+    For all well-formed context $E[\Gamma]$ and terms $e_1$, $e_2$, if $AX \vdash e_1 = e_2$, then $\sem{e_1} = \sem{e_2}$.
+\end{lemma}
+
+Unfortunately, the inverse does not hold: there are expressions that are equivalent in denotational interpretations but not provable through the axioms.
+One notorious example is $\delta_{i, j} \times \delta_{j, k} = \delta_{i, k} \times \delta_{j, k}$.
+Intuitively, the two terms will be evaluated to $1$ if and only if $i = j = k$, but it is difficult to incorporate such properties as axioms.
+This incapability further limits the algorithmm, which is designed to automate reasonings using the axioms, and blocks us from proving complete results with respect to the semantics.
+But our work focuses on solving the practical examples, which are covered by our axioms, as is shown by experiments.
 
@@
         & \textrm{TPO} && := \mu T_1. \mu T_2. \lambda O : \OType(T_1, T_2). \sum_{i \in \mathbf{U}(T_1)} \sum_{j \in \mathbf{U}(T_2)} \bra{i} O \ket{j} . \ket{j}\bra{i} \\
-        & && : \forall T_1. \forall T_2. \OType(T1, T2) \to \OType(T2, T1); \\
+        & && : \forall T_1. \forall T_2. \OType(T_1, T_2) \to \OType(T_2, T_1); \\
         & \textrm{phi} && := \mu T. \sum_{i \in \mathbf{U}(T)} \sum_{j \in \mathbf{U}(T)} \ket{(i, j)} : \forall T.\KType(T \times T); \\
@@
     \end{align*}
+    The symbol TPO is the encoding of operator transpose. It takes two index $T_1$ and $T_2$, and transforms the operators.
+    phi takes the index $T$ and defines the maximally entangled states. It sums over all basis in $T$, indicated by the universal set $\mathbf{U}(T)$.
+    With the assumption of index T and operator M, we can write the equivalence as 
+    \[
+        (\textrm{M} \otimes \mathbf{1}_\mathcal{O}(\textrm{T})) \cdot (\textrm{phi T}) = (\mathbf{1}_\mathcal{O}(\textrm{T}) \otimes (\textrm{TPO T T M})) \cdot (\textrm{phi T}).
+    \]
 \end{example}
@@
 
-\section{Decision procedure for Dirac notations}
+\section{Algorithm for Deciding Dirac notations equations}
 
-The following two sections talk about how to decide the equivalence of Dirac notations.
+The following two sections discuss how to decide the equivalence of Dirac notations implied by the axioms.
 In the previous work, the equational axioms are separated into two parts: a set $E$ of equations that can not be decided by rewriting, and the remaining equations to be decided by term rewriting.
@@
 \begin{itemize}
-    \item Commutativity of symbols $a + b$, $a\times b$ and $\delta_{s, t}$,
+    \item commutativity of symbols $a + b$, $a\times b$ and $\delta_{s, t}$,
     \item $\alpha$-equivalence of bound variables, i.e., $\lambda x.t = \lambda y . t\{x/y\}$,
@@
 
-Here to enable more efficient algorithms, our langauge uses AC symbols with a variable arity.
+% Here to enable more efficient algorithms, our langauge uses AC symbols with a variable arity.
 
 The full list of rewriting rules are in~\Cref{sec: rewriting rules}. Here we present some of them to illustrate the design idea.
-
-When using variable arity AC symbols, the associativity can be normalized by the following flattening rule:
+The whole rule list consists of several types.
+Firstly, we use the following flattening rule to decide the associative with AC symbols.
 
@@
 
-while commutativity is left for a sorting algorithm later on.
+and commutativity is left for a sorting algorithm later on.
 
-The term rewriting rule follows the system in the previous work.
+Some equational axioms imply obvious rewriting diraction, for example:
 
-The general idea is to reduce all the possible calculations, and transform multiplication into tensor product as much as possible.
+\begin{align*}
+    &\textrm{(R-DOT6)} && \bra{s} \cdot \ket{t} \ \reduce \delta_{s, t} \\
+    &\textrm{(R-DELTA0)} && \delta_{s, s} \ \reduce\ 1 \\
+    &\textrm{(R-MULK1)} && O : \OType(\sigma, \tau) \Rightarrow O \cdot \mathbf{0}_{\KType(\tau)} \ \reduce \mathbf{0}_{\KType(\sigma)}
+\end{align*}
 
-And some rules exists for completeness. As an example, given the following rule snippet
+The (R-DOT6) says that inner product of two basis will be evaluated to a delta expression,
+and (R-DELTA0) further transforms delta on identical basis to scalar $1$.
+(R-MULK1) corresponds to the axiom that multiplication on zero vectors results in zero. It has a condition on typing, which is calculated during rewriting.
+They share the common feature that right hand side is simpler than the left hand side.
+
+Other part of rules assign the rewriting diraction to equations according to the property of symbols and intuitive observations of their relations.
+When different rules rewrites the same term to different forms, rules for confluence purpose are added.
+
 \begin{align*}
+    %
+    & \text{(R-SCR1)} 
+    && a.(b.X) \ \reduce\ (a \times b).X \\
     %
@@
 \end{align*}
-the normalization of term $(\bra{(s, t)} \cdot (O_1 \otimes O_2)) \cdot K$ have two rewriting paths: (a). apply (R-MULB10) and get $((\bra{s} \cdot O_1) \otimes (\bra{t} \cdot O_2)) \cdot K$, or
+The (R-SCR1) rule follows the idea to reduce linear algebra scaling into scalar multiplications,
+and (R-MULB10) transform inner product into tensor product. 
+As an example of the completetion rule, here the normalization of term $(\bra{(s, t)} \cdot (O_1 \otimes O_2)) \cdot K$ have two rewriting paths: (a).
+apply (R-MULB10) and get $((\bra{s} \cdot O_1) \otimes (\bra{t} \cdot O_2)) \cdot K$, or
 (b).
@@
 
-One important technique revealed in the previous work is the expansion of variables:
+There are also rules dealing with sum expressions. For example,
+
+\begin{align*}
+    & \textrm{(R-SUM-ELIM2)}
+    && i \text{ free in } t \Rightarrow \sum_{i \in \mathbf{U}(\sigma)} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (\delta_{i, t}.A) \\
+    & && \reduce\ \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} A\{i/t\} \\
+    & \textrm{(R-SUM-PUSH5)}
+    && (\sum_{i \in M} B)\cdot K \ \reduce\ \sum_{i \in M}(B \cdot K)
+\end{align*}
+
+The (R-SUM-ELIM2) rule will try to eliminate the delta expression in summations.
+And the (R-SUM-PUSH5) rule will lift sum to the outside of inner product.
+
+One important technique revealed in the DiracDec work is the expansion of variables:
 
@@
 \end{gather*}
-The above three rules are obviously not terminating, and this is why we have the rewriting-expansion-rewriting steps in the decision procedure.
+The above three rules are not terminating, and this is why we have the rewriting-expansion-rewriting steps in the algorithm.
 On the other hand, we discovered that doing the expansion on all variables only once is already sufficient.
@@
 
-The previous work did not consider the algorithm to decide equational theory $E$.
-In their Mathematica implementation, it is implemented by a unification, which tries to find a substitution of summation bound variables that makes the two expressions syntactically equivalent.
-To decide AC-equivalence and (SUM-SWAP), they iterate through all the permutations, and the complexity is factorial to the number of AC symbol arguments and successive summations.
+In the Mathematica implementation of DiracDec, the equational theory $E$ is decided by a unification, which tries to find a substitution of summation bound variables and AC symbol arguments that makes the two expressions syntactically equivalent.
+The unification iterates through all permutations, and the complexity is factorial in general.
 
-A standard approach to decide this permutation equivalence is to normalize by sorting with a given order.
-For example, given the dictionary order $a < b < c$, the term $b + c + a$ (and any other AC equivalent ones) will be normalized into $a + b + c$.
-For our setting, there are two related difficulties: how to assign such an order to all terms in our language, and how to normalize with respect to AC-equivalence and (SUM-SWAP) at the same time.
-
-As a typical example, the two terms 
-\[
-    \sum_{i \in s_1} \sum_{j \in s_2} \bra{i}A\ket{j} \times \bra{j}B\ket{i}
-    = 
-    \sum_{i \in s_2} \sum_{j \in s_1} \bra{i}B\ket{j} \times \bra{j}A\ket{i}
-\]
-are equivalent due to the swap of summations and commutativity of scalar multiplication.
-
-We propose an algorithm to ...
-
-To describe the algorithm, we introduce two notations: for a term $e = f(a_1, a_2, \cdots, a_n)$, $\textrm{head}(e)$ indicates the function symbol $f$,
-and $\textrm{arg}(e, i)$ indicates the $i$-th argument $a_i$. Variables and constants are considered as functions with zero arguments.
-\begin{definition}[order modulo bound variables]
-    Assume we have a total order of all symbols.
-    Let $\mathcal{B}$ be the set of bound variables.
-    For simplicity, we assume all bound variables are unique. 
-    The relation $e_1 < e_2$ between the two terms holds when:
-    \begin{itemize}
-        \item $e_1 \notin \mathcal{B}$ and $e_2 \in \mathcal{B}$, or
-        \item $head(e_1) < head(e_2)$, or
-        \item $head(e_1) = head(e_2)$, and there exists $n$ with $arg(e_1, n) < arg(e_2, n)$, and for all $i < n$, either $arg(e_1, i) = arg(e_2, i)$, or $arg(e_1, i) \in \mathcal{B}$ and $arg(e_2, i) \in \mathcal{B}$.
-    \end{itemize}
-\end{definition}
-
-The order depends on there occurances in the last sorting result. If no occurance, then the order will depend on the set (for sum) and the type (for lambda abstraction only).
-
-%%%%%%%%%%%%%%%%%%%%%%%
-
-The idea is to assign an order to terms, which is independent on the bound variables. Because we can have terms with nested AC symbols.
-
-And in the end, the $\alpha$-equivalence are checked by de Bruijn index~\cite{deBruijn1972lambda}, where references to bound variable names are replaced by the distance of the lambda abstraction to the variable.
+As the first improvement, the $\alpha$-equivalence are checked by de Bruijn index~\cite{deBruijn1972lambda}, where references to bound variable names are replaced by the distance of the lambda abstraction to the variable.
 For example, the nominal lambda abstraction $\lambda x.
@@
 Therefore, transformation into de Bruijn index is at the last step of the normalization to check the equivalence of terms with different bound variable names.
+
+A standard approach to decide this permutation equivalence is to normalize by sorting with a given order.
+For example, given the dictionary order $a < b < c$, the term $b + c + a$ (and any other AC equivalent ones) will be normalized into $a + b + c$.
+For our setting, there are two difficulties intertwined together: how to assign such an order to all terms in our language, and how to normalize with respect to $E$.
+As a typical example, the two terms 
+\[
+    \sum_{i \in s_1} \sum_{j \in s_2} \bra{i}A\ket{j} \times \bra{j}B\ket{i}
+    = 
+    \sum_{i \in s_2} \sum_{j \in s_1} \bra{i}B\ket{j} \times \bra{j}A\ket{i}
+\]
+are equivalent due to the swap of summations and commutativity of scalar multiplication. However,
+directly sorting elemnts of scalar multiplication will not lead to the same form.
+
+We propose an algorithm to solve the problem by sorting without bound variables.
+The observation is that in a successive sum expression $\sum_{i \in s_1}\cdots\sum_{j \in s_n}A$, the name and order of bound variables $i, \cdots, j$ can be permuted freely.
+Therefore, all bound variables should be treated uniformly during sorting, and the order of sum can then be decided according to the position of their bound variables.
+
+In the above example, we ignore the bound variables first, and sort the sum body into $\bra{\bullet}A\ket{\bullet} \times \bra{\bullet}B\ket{\bullet}$.
+Then we swap the sum so that the bound variable at first $\bullet$ position appears at the outside, and the two terms will have the same de Bruijn normal form as $\sum_{s_1} \sum_{s_2} \bra{\$1} A \ket{\$0} \times \bra{\$0} B \ket{\$1}$.
+
+To describe the algorithm, we introduce two notations: for a term $e = f(a_1, a_2, \cdots, a_n)$, $\textrm{head}(e)$ indicates the function symbol $f$,
+and $\textrm{arg}(e, i)$ indicates the $i$-th argument $a_i$. 
+% Variables and constants are considered as functions with zero arguments.
+\begin{definition}[order without bound variables]
+    Assume we have a total order of all symbols.
+    Let $\mathcal{B}$ be the set of bound variables.
+    For simplicity, we assume all bound variables are unique.
+    The relation $e_1 =_\mathcal{B} e_2$ holds when
+    \begin{itemize}
+        \item $head(e_1) = head(e_2)$ and $arg(e_1, i) =_\mathcal{B} arg(e_2, i)$ for all $i$, or
+        \item $e_1 \in \mathcal{B}$ and $e_2 \in \mathcal{B}$.
+    \end{itemize}
+
+    The relation $e_1 <_\mathcal{B} e_2$ between the two terms holds when:
+    \begin{itemize}
+        \item $e_1 \notin \mathcal{B}$ and $e_2 \in \mathcal{B}$, or
+        \item $head(e_1) < head(e_2)$, or
+        \item $head(e_1) = head(e_2)$, and there exists $n$ with $arg(e_1, n) <_\mathcal{B} arg(e_2, n)$, where $arg(e_1, i) =_\mathcal{B} arg(e_2, i)$ for all $i < n$.
+    \end{itemize}
+\end{definition}
+It can be checked that $e_1 =_\mathcal{B} e_2$ if and only if neither $e_1 <_\mathcal{B} e_2$ or $e_2 <_\mathcal{B} e_1$ holds.
+The idea is to compare the function symbols in the top down order, and omit bound variable occurance.
+
+% The order depends on there occurances in the last sorting result. If no occurance, then the order will depend on the set (for sum) and the type (for lambda abstraction only).
+
+%%%%%%%%%%%%%%%%%%%%%%%
+
+The idea is to assign an order to terms, which is independent on the bound variables. Because we can have terms with nested AC symbols.
 
@@
 
-Other previous works explore the language and decision procedure to express quantum computation differently.
+Other previous works exploring the language and algorithm to reasong about quantum computation differently.
 

### 2025-01-20 flowchart.tex
2025/1/17
2025/1/20

@@
     \node (rewrite2) [rewriting, below of=expansion] {second rewritings};
-    \node (sorting) [process, below of=rewrite2] {sorting modulo bound variables};
-    \node (successive) [process, below of=sorting] {sorting successive summation};
-    \node (deBruijn) [process, below of=successive] {de Bruijn normalization};
+    \node (sort) [process, below of=rewrite2] {sorting without bound variables};
+    \node (swap) [process, below of=sort] {swap successive summation};
+    \node (deBruijn) [process, below of=swap] {de Bruijn normalization};
 
@@
     \draw [arrow] (expansion) -- (rewrite2);
-    \draw [arrow] (rewrite2) -- (sorting);
-    \draw [arrow] (sorting) -- (successive);
-    \draw [arrow] (successive) -- (deBruijn);
+    \draw [arrow] (rewrite2) -- (sort);
+    \draw [arrow] (sort) -- (swap);
+    \draw [arrow] (swap) -- (deBruijn);
 

### 2025-01-20 labelled.tex
2025/1/20

@@
 \]
-In the left hand side, the state of two subsystems is $\ket{\Phi}$, and we apply quantum operation $M$ on the system $Q$.
+On the left hand side, the state of two subsystems is $\ket{\Phi}$, and we apply quantum operation $M$ on the system $Q$.
+It is equivalent to extend the operation using identity operators on other subsystems (i.e.
+the cylinder extension), and consider the application in the whole system.
 
@@
   \begin{align*}
-    R ::= r\in\cR \mid (R, R)
+    R ::= r\in\cR \mid (R, R).
   \end{align*}
@@
 Registers can be composed by pairs of $(R, R)$, and this structure corresponds to the structure of tensor product spaces in Dirac notations.
-
 To reason about the registers, we define their variable set as the enumeration of all register symbols involved.
@@
 \begin{itemize}
-    \item $R\equiv r$: $\var(R) = \{r\}$;
-    \item $R\equiv (R_1,R_2)$: $\var(R) = \var(R_1) \cup\var(R_2)$.
+    \item $\var(R) = \{r\}$, if $R\equiv r$; or
+    \item $\var(R) = \var(R_1) \cup\var(R_2)$, if $R\equiv (R_1,R_2)$.
 \end{itemize}
@@
 
-\textbf{Remark: } Set operations: $\cup$ for union; $\cap$ for intersection; $\setminus$ for difference. So,
-% \begin{itemize}
-$ S_1 \cap S_2 \equiv S_1 \cup S_2 \setminus (S_1 \setminus S_2) \setminus (S_2 \setminus S_1) $.
+% \textbf{Remark: } Set operations: $\cup$ for union; $\cap$ for intersection; $\setminus$ for difference. So,
+% % \begin{itemize}
+% $ S_1 \cap S_2 \equiv S_1 \cup S_2 \setminus (S_1 \setminus S_2) \setminus (S_2 \setminus S_1) $.
 % \end{itemize}
@@
 \subsubsection*{Typing rules}
+Some typing rules are introduced here.
+The rule for $\DType(s_1, s_2)$ requires that all registers in variable set $s_1$ and $s_2$ are well-typed.
+The rule for $K_R$ demonstrates how a register label is added to the Dirac notation, and the rule for $D^\dagger$ shows that labelled Dirac notation also have symbols for calculation, such as the adjoint.
 
@@
 
-The typing rule for tensor product is
+The dot and the tensor product symbols are different from those in unlabelled Dirac notation.
+Since the goal of labels is to replace the order and structure of tensor products by the reference to registers, the tensor product becomes an AC symbol.
+The typing still checks whether the component subsystems are disjoint with each other.
 \[
@@
 \]
-
-The typing rule for dot product is
+As for the dot product, the disjointness is considered except registers contracted by multiplication.
 \[
@@
 
+This step reduces all terms in the form of $K_R$ or $O_{R;R}$.
+Other symbols on labelled Dirac notations are also reduced by rules like $(D_1 \cdot D_2)^\dagger \ \reduce\ D_2^\dagger \cdot D_1^\dagger$.
+
+The final part of rules operates on sum and dot product. They will lift summation to the outside, and eliminate the bra-ket pairs whenever possible.
+
 \begin{align*}
@@
     && (\sum_{i \in M} D_1) \cdot D_2 \ \reduce\ \sum_{i \in M} (D_1 \cdot D_2) \\
-\end{align*}
-
-\begin{align*}
     %
@@
 
-\subsubsection*{Eliminating the registers}
+These rules are added to the rewriting system in ~\Cref{sec: decide} and executed together.
+In the end, if there are no variables of $\DType(s_1, s_2)$, the expression will be reduced to the addition of big operator sum, and each sum body is Dirac notation scalar with labelled basis.
 
-\begin{enumerate}
-  \item For any register $R$ s.t. $E[\Gamma]\vdash R : \reg \sigma$, suppose $\var R = \{r_1,r_2,\cdots,r_n\}$, 
-  we introduce variables $i_{r_k} : \Basis(\sigma_{r_k})$ with $r_k : \reg \sigma_{r_k} \in E$ for $k = 1,\cdots, n$.
-We reconstruct the basis $|i_R\>$ (which is of type $\KType(\sigma)$) and $\<i_R|$ (which is of type $\BType(\sigma)$) of $R$ by:
-  \begin{itemize}
-    \item $R = r_k$, $|i_R\> \triangleq |i_{r_k}\>$ and $\<i_R| \triangleq \<i_{r_k}|$;
-    \item $R = (R_1,R_2)$: $|i_R\> \triangleq |i_{R_1}\> \otimes |i_{R_2}\>$ and $\<i_R| \triangleq \<i_{R_1}| \otimes \<i_{R_2}|$.
-  \end{itemize}
-    \item Expansion of all $K_R,B_R,O_{R,R'}$
-
-    After the rewriting of this step, the expression we have is always in the following form:
-    $$
-    \Big(\sum_{i_1}\cdots\sum_{i_n} a.\tD_f\Big) + \cdots + \Big(\sum_{i_1}\cdots\sum_{i_{n'}} a.\tD_f\Big)
-    $$
-    where $\tD_f$ is generated by :
-    $$\tD_f ::= |i\>_r \mid {}_r\<i| \mid \te\otimes\te \mid \te\cdot\te. $$
-    \item
-    Suppose $\mathsf{CDom}(\tD_{f}) = \{(i_1,r_1),\cdots,(i_n,r_n)\}$ and $\mathsf{Dom}(\tD_{f}) = \{(i'_1,r'_1),\cdots,(i'_{n'},r'_{n'})\}$ where the list is sorted by a default order of variables in $\cR$.
-We define:
-    $$D_f = \bigg(\prod_{(i,j)\in \mathsf{Can}(\tD_f)}\delta_{i,j}\bigg).\Big((|i_1\>\otimes\cdots\otimes|i_n\>) \cdot (\<i'_1|\otimes\cdots\otimes\<i'_{n'}|)\Big).$$
-    If $\mathsf{Dom}$ or $\mathsf{CDom}$ is an empty list, the $D_f$ is just a Ket or Bra.
-    Then we claim:
-    \begin{itemize}
-        \item If $E[\Gamma]\vdash \tD_f : \DType(s_1,s_2)$, then $\{r_1,\cdots,r_n\} = s_1$ and
-        $\{r'_1,\cdots,r'_{n'}\} = s_2$.
-        \item Set $R = ((\cdots(r_1,r_2),\cdots),r_n)$ and $R' = ((\cdots(r'_1,r'_2),\cdots),r'_{n'})$, then
-        $$\tD_{f} = (D_f)_{R,R'}$$
-    \end{itemize}
-    \item For the obtained formula in (3), we replace each $\tD_f$ by $D_f$ which is computed by step (4).
-Note that $D_f$ is a Dirac term without labels, so the equivalence of two labelled Dirac terms is reduced to decide the equivalence of two Dirac terms.
-This is guaranteed if two labelled Dirac terms are of the same type (by typing rules).
-\end{enumerate}
+\begin{lemma}
+    For a well-typed term $e$ in $E[\Gamma]$, if $e$ does not contain variables of $\DType(s_1, s_2)$ type, the normal form of $e$ will be
+    \[
+    \sum_{i}\cdots\sum_{j} a_1 . (\ket{i}_{p} \otimes \cdots \otimes \bra{j}_{q})
+    + \cdots +
+    \sum_{k}\cdots\sum_{l} a_m . (\ket{k}_{r} \otimes \cdots \otimes \bra{l}_{s})
+    \]
+    Where $a_i$ are scalar typed Dirac notations.
+\end{lemma}

### 2025-01-20 main.tex
2025/1/20

@@
 
-In practice, Dirac notations are combined with other syntax for better expressiveness.
-For example, the big operator sum $\sum_{i \in S} A$ are used in theoretical reasonings.
-We also use labelled Dirac notation, where Dirac notations are subscripted with registers to denote the quantum system in consideration, e.g.
-$\ket{i}_q \otimes \ket{i}_r$.
+In practice, Dirac notations are combined with other syntax, such as the big operator sum $\sum_{i \in S} A$, for better expressiveness.
+We also use labelled Dirac notations like $\ket{i}_q \otimes \ket{i}_r$ to denote the quantum system in consideration, where Dirac notations are subscripted with registers.
 
 We use universal algebra and equational logic to formally represent Dirac notations and the reasoning procedure.
-A universal algebra defines a signature of function symbols and the axioms.
+A universal algebra defines a signature of function.
 Terms in the algebra are constructed by constants, variables or function application on terms.
-In our case of Dirac notation, the signature consists of constructors and operations like $\ket{i}$ or $A \otimes B$, and the axioms Dirac notation equivalenc rules with equational logic.
+In our case of Dirac notation, the signature consists of constructors and operations like $\ket{i}$ or $A \otimes B$.
+Reasonings of Dirac notation terms follow the equational logic, where an equaitonal relation $e_1 \doteq e_2$ is defined, consistent with substitution and term construction:
+\[
+    \frac{E \vdash s \doteq t}{E \vdash v(s) \doteq v(t)}
+    \qquad
+    \frac{E \vdash s_1 \doteq t_1 \qquad \cdots \qquad E \vdash s_n \doteq t_n}{E \vdash f(s_1, \dots, s_n) \doteq f(t_1, \dots, t_n)}
+\]
+Here $v(s)$ represent a substitution of variables, and $E$ is set of Dirac notation equational axioms.
 
@@
 \end{example}
-
-\yx{I guess it's wise to avoid mentioning that we don't impose fixed arity.}
 
@@
 \yx{where should we put the example?}
-To conclude the design introduction, we demonstrate part of the formalization of the preliminary example.
-\begin{example}[formalizing the preliminary example]
+To conclude the design introduction, we demonstrate part of the formalization of the motivating example.
+\begin{example}[formalizing the motivating example]
     Definitions and assumptions in the environment:
@@
 \section{Algorithm for Deciding Dirac notations equations}
+\label{sec: decide}
 
@@
 \subsection{Normalization modulo $E$ by Term Rewriting}
-Term rewriting rules, written as $l -> r$, are used to normalize terms by recursively matching the subterms of the term with the left hand side $l$, and replace them with the corresponding right hand side $r$.
+Term rewriting rules, written as $l \to r$, are used to normalize terms by recursively matching the subterms of the term with the left hand side $l$, and replace them with the corresponding right hand side $r$.
 The procedure terminates when no more rewritings can be made, and the order of rewritings will be irrelevant if the term rewriting system is \textit{confluent}, which is a desirable property.
@@
 The idea is to compare the function symbols in the top down order, and omit bound variable occurance. 
+With the order, the sort function normalize the term with respect to AC equivalence.
+
+\begin{definition}[sort transformation]
+    For a term $e$ with bound variable set $\mathcal{B}$,
+    The sort transformation sort(e) is defined as
+    \begin{itemize}
+        \item $e$, if $e$ is a variable or constant;
+        \item $\lambda x:T. \textrm{sort}(e')$, if $e \equiv \lambda x : T. e'$;
+        \item $\mu x. \textrm{sort}(e')$, if $e \equiv \mu x. \textrm{sort}(e')$; or
+        \item $f(\textrm{sort}(a_1), \cdots, \textrm{sort}(a_n))$, if $e \equiv f(a_1, \cdots, a_n)$.
+If $f$ is an AC symbol, then the order of sort($a_i$) is sorted ascendingly according to $<_\mathcal{B}$.
+    \end{itemize}
+\end{definition}
+
+The order of bound variables can be decided according to the result of the sort transformation, and then successive summations can be normalized by swapping.
+\begin{definition}[swap transformation]
+    For a term $e$ with sorting result sort$(e)$, we can obtain a total order of all bound variables appearing in the term $e$.
+Specifically, they are ordered by their first appearances, except in function definitions $\lambda x$ and $\mu x$.
+    The swap transformation then arrange successive summation according to this order.
+\end{definition}
 
@@
 
-The idea is to assign an order to terms, which is independent on the bound variables. Because we can have terms with nested AC symbols.
+% The idea is to assign an order to terms, which is independent on the bound variables. Because we can have terms with nested AC symbols.
 
@@
 
-Here is the code for the preliminary example.
+Here is the code for the motivating example.
 

### 2025-01-23 dirace.tex
2025/1/23

\tikzstyle{module} = [draw, rounded corners, fill=blue!20, text centered]
\tikzstyle{file} = [draw, rounded corners, fill=green!20, text centered]
\tikzstyle{thirdparty} = [draw, rounded corners, fill=red!20, text centered]

\begin{figure}
    \centering
    \ttfamily
    \begin{minipage}{0.4\textwidth}
        \centering
        \begin{tikzpicture}

            % \node (rename) [process] {rename unique bound variable names};
            \node (examples) [module] {examples};
            \node (toplevel) [file, right of=examples, xshift=1cm] {toplevel.cpp};
            \node (dirace) [module, below right of=examples, xshift=0.3cm] {dirace};
            \node (ualg) [module, below left of=dirace] {ualg};
            \node (ANTLR4) [thirdparty, below of =ualg] {ANTLR4};
            \node (WSTP) [thirdparty, below right of =dirace, xshift=1cm, yshift=-0.3cm, align=center] {WSTP interface\\(Mathematica)};

            \draw [arrow] (examples) -- (dirace);
            \draw [arrow] (toplevel) -- (dirace);
            \draw [arrow] (dirace) -- (ualg);
            \draw [arrow] (ualg) -- (ANTLR4);
            \draw [arrow] (dirace) -- (WSTP);
        \end{tikzpicture}
    \end{minipage}
    \hfill
    \begin{minipage}{0.55\textwidth}
        \centering
        \begin{tikzpicture}
            \node (prover) [file] {prover.cpp};
            \node (diracparser) [file, below left of=prover, xshift = -1cm] {dirac\_parser.cpp};
            \node (reduction) [file, below right of=prover, xshift = 1cm] {reduction.cpp};
            \node (calculus) [file, below of=prover, yshift=-0.5cm]{calculus.cpp};
            \node (syntaxtheory) [file, below right of=calculus, xshift = 1cm]{syntax\_theory.cpp};
            \node (symbols) [file, below left of=calculus, yshift=-0.8cm]{symbols.cpp};

            \draw [arrow] (prover) -- (diracparser);
            \draw [arrow] (prover) -- (reduction);
            \draw [arrow] (prover) -- (calculus);
            \draw [arrow] (reduction) -- (calculus);
            \draw [arrow] (reduction) -- (syntaxtheory);
            \draw [arrow] (syntaxtheory) -- (symbols);
            \draw [arrow] (calculus) -- (symbols);
            \draw [arrow] (diracparser) -- (symbols);
        \end{tikzpicture}
    \end{minipage}
    \rmfamily
    \caption{The structure of Dirace. The left figure illustrates the whole system, and the right figure illustrates the \texttt{dirace} module. Blue nodes denote modules, red nodes denote third-party libraries, and green nodes denote files. Arrows represent dependency.}
    \label{fig: dirace structure}
\end{figure}

### 2025-01-23 labelled.tex
2025/1/23

@@
 
-These rules are added to the rewriting system in ~\Cref{sec: decide} and executed together.
+These rules are added to the rewriting system in~\Cref{sec: decide} and executed together.
 In the end, if there are no variables of $\DType(s_1, s_2)$, the expression will be reduced to the addition of big operator sum, and each sum body is Dirac notation scalar with labelled basis.
 
-\begin{lemma}
+\begin{lemma}[normal form]
+    \label{lem: labelled normal form}
     For a well-typed term $e$ in $E[\Gamma]$, if $e$ does not contain variables of $\DType(s_1, s_2)$ type, the normal form of $e$ will be

### 2025-01-23 main.tex
2025/1/23

@@
 \usetikzlibrary{shapes.geometric, arrows}
+\usetikzlibrary{trees}
 
@@
 
-    Dirac notations is the fundamental language in quantum computation and quantum information reasoning.
-Recently, the term rewriting system DiracDec~\cite{diracdec} was introduced to automate equational reasoning with Dirac notations, a critical yet time-intensive component of formalization tasks.
+    Dirac notations is the fundamental language in quantum computation and quantum information.
+Recently, the term rewriting system DiracDec~\cite{diracdec} was introduced to automate the equational reasoning with Dirac notations, a critical yet time-intensive task.
 This work bases upon and extends DiracDec, aiming to develop a solver optimized for practical applications.
 Enhancements include an improved typing system, a simplified language and rewriting system, more efficient algorithms, and added support for labelled Dirac notations.
-The solver Dirace, implemented in \CC\ with a Mathematica backend, demonstrates superior decision-making power and computational efficiency, successfully addressing all examples in its benchmark suite.
+Our solver Dirace is implemented in \CC\ with a Mathematica backend, and demonstrates advantages in decision-making power and time efficiency.
 
@@
 They proved that the equivalence of basic Dirac notations are decidable.
-Their algorithm sticks with a pure term rewriting system, which allows to prove important properties such as confluence and termination of the system.
+Their algorithm sticks with a pure term rewriting system, and is proved to be confluent and terminating.
 Even though, there is still a gap between DiracDec and a practical solver for Dirac notation equivalence.
@@
 DiracDec uses a direct by inefficient algorithm to decide $E$, which search through all possible permutations and has factorial complexity.
-This is witnessed in their evaluation: DiracDec is not good at dealing with "computational" examples, which has many AC symbols and are therefore time consuming.
+This is witnessed in their evaluation: DiracDec is not good at dealing with ``computational examples'', which has many AC symbols and are therefore time consuming.
 
-Another problem is the lack of labelled Dirac notations.
-which makes the reasoning more convenient and scaleable.
+Usability is another problem.
+DiracDec does not support labelled Dirac notation, which use registers to denote the subsystem, and express the state locally.
+The typing of DiracDec only deals with Dirac notations, which is not sufficient for a working scenario, where we want to have defined symbols and a context for the typing assumptions of variables.
+Moreover, to avoid type checking during term rewriting, DiracDec separate the same symbol (e.g. multiplication) for different types to disambiguate, which makes the system bloated.
+Also, it is hard to integrate the Mathematica implementation as a solvers into other projects.
 
-One last problem is about usability.
-The typing of DiracDec only deals with Dirac notations, which is not sufficient for a working scenario, where defined symbols are desirable, and variables have their typing assumptions in the context.
-Moreover, to avoid type checking during term rewriting, DiracDec uses different symbols for the same casual notation to disambiguate, which 
-is not close to the natural way of using the langauge, and makes the system bloated.
-It is hard to integrate the Mathematica implementation as solvers into into other tools.  
-Although DiracDec solve most examples in the benchmark, including some very complicated equations, there are still a few examples that DiracDec failds to decide.
-
-Aimming at building a practical solver, this work bases on and extends the DiracDec result, transform the term rewriting system into a hybrid algorithm, and overcomes the problems mentioned above.
+There is a trade-off between simplicity and efficiency for the system design.
+The simplicity of term rewriting in DiracDec allows them to reach satisfying theoretical results, but it is also a constraint for optimization.
+Based on DiracDec, this work aims at building a practical solver.
+It transforms the term rewriting system into a hybrid algorithm, and overcomes the problems mentioned above.
 Our main contributions and techniques are:
 \begin{itemize}
-    \item An efficient algorithm to decide the extra equational theories $E$.
-The basic idea is normalization by sorting, which considers the interplay of AC symbols, SUM-SWAP and $\alpha$-equivalence.
-This algorithm reduces the complexity from the factorial level to the polynomial level.
-    \item The support for constant register labels. The equivalence decision of labelled Dirac notations is then reduced into the unlabelled situation.
-    \item A more usable \CC\ solver Dirace with usable language and typing. Defined symbols (e.g., transpose and trace) are modelled by functions.
+    \item An efficient algorithm to decide the equational theories in $E$.
+The basic idea is to carry through equivalence checking by normalization, and the normal form for $E$ are obtained by sorting.
+% This algorithm reduces the complexity from the factorial level to the polynomial level.
+    \item The support for constant register labels. The equivalence decision of labelled Dirac notations is reduced to the unlabelled situation.
+    \item A more usable \CC\ solver Dirace with abstract language and typing. We also support to define symbols (e.g., transpose and trace) using the functions syntax.
 \end{itemize}
 
-As an evaluation, we tested Dirace on the benchmark of DiracDec and new example for labelled Dirac notations.
-Dirace successfully decides all the examples that are expressable in its language, including those failed by DiracDec because of complexity or insufficient decision power.
+As an evaluation, we tested Dirace on the benchmark of DiracDec and new examples for labelled Dirac notations. 
+The result shows significant improvement in decidability and efficiency over DiracDec.
+% Dirace successfully decides all the examples that are expressable in its language, including those failed by DiracDec because of complexity or insufficient decision power.
 
-\section{Preliminary and Motivation}
-The preliminary for the problem are two folds: Dirac notation and the equational logic in universal algebra.
+% \yx{mention the dilemma of simplicity and efficiency}
 
-Quantum states live in complex Hilbert spaces. We use vectors in the space to descirbe pure quantum states, and operations corresponds to linear transformations. 
-Dirac notation are used to express the linear algebra formula in a convenient and compat way.
+\section{Motivation and Preliminary}
+One interesting fact about the quantum world is that for two maximally entangled states, applying one operator $M$ on one subsystem is equivalent to applying $M^T$, the transpose of $M$, on the other.
+This property holds no matter how far the two systems are separated, and is expressed as an equation of Dirac notations.
+\begin{example}
+    \label{ex: motivating}
+    Let $q$ and $r$ be two quantum systems on $\mathcal{H}_T$ space. Let $M$ be a quantum operation on $\mathcal{H}_T$,
+    and $\ket{\Phi} = \sum_{i \in T} \ket{i}\otimes \ket{i}$ be the maximally entangled state, then we have
+    \[
+    M_q \ket{\Phi}_{q; r} = M_r^T \ket{\Phi}_{q; r}.
+    \]
+\end{example}
+Here $q$ and $r$ are called labels, denoting the system for Dirac notations.
+To reason about such equations using computers, we need to formalize the language and work in a proof system.
+The following paragraphs briefly introduce the Dirac notation and the universal algebra framework we use.
+
+\subsubsection{Dirac notation}
+Quantum states live in complex Hilbert spaces. We use vectors in the space to descirbe pure quantum states, and operations are described by linear transformations. 
 It uses the ket $\ket{i}$ and the bra $\bra{i}$ to indicate bases of the space and the dual space.
@@
 
+\subsubsection{Universal algebra}
 We use universal algebra and equational logic to formally represent Dirac notations and the reasoning procedure.
-A universal algebra defines a signature of function.
+A universal algebra defines a signature of function symbols.
 Terms in the algebra are constructed by constants, variables or function application on terms. 
+Other basic concepts like substitution of variables or pattern matching are also defined.
 In our case of Dirac notation, the signature consists of constructors and operations like $\ket{i}$ or $A \otimes B$.
-Reasonings of Dirac notation terms follow the equational logic, where an equaitonal relation $e_1 \doteq e_2$ is defined, consistent with substitution and term construction:
+Reasonings of Dirac notation terms follow the equational logic, which defines an equaitonal relation $e_1 \doteq e_2$ compatible with substitution and term construction:
 \[
@@
 \]
-Here $v(s)$ represent a substitution of variables, and $E$ is set of Dirac notation equational axioms.
-
-Here is an example proof of a equivalence lemma.
-\begin{example}
-    Let $q$ and $r$ be two quantum systems on $\mathcal{H}_T$ space. Let $M$ be a quantum operation on $\mathcal{H}_T$,
-    and $\ket{\Phi} = \sum_{i \in T} \ket{i}\otimes \ket{i}$ be the maximally entangled state, then we have
-    \[
-    M_q \ket{\Phi}_{q; r} = M_r^T \ket{\Phi}_{q; r}.
-    \]
-\end{example}
-
-% \yx{mention the dilemma of simplicity and efficiency}
+This logic essentially formalizes the common sense of equivalence in algebra.
 
 \section{Language, Typing and Semantics}
-The syntax of Dirac notations involves three layers: the index, the type and the term.
+The first step is to formally pin down the language for Dirac notation.
+As a reference, the language in DiracDec has a concrete design, meaning that the same syntax for different types corresponds to different symbols.
+Our intention is to move from the concrete design to a more abstract one. This is closer to the casual Dirac notation we use, and simplifies the term rewriting system.
+
+For this purpose, our language involves three layers: the index, the type and the term.
 Terms describe the concrete instances like ket, bra and operator, which will be checked and typed.
-Index represents classical datatypes, and they appear in type expression to distinguish between different Hilbert spaces and sets.
+Index represents classical datatypes, and they appear in type expressions to distinguish between different Hilbert spaces and sets.
 
@@
     \begin{align*}
-        \sigma ::=\ & x \mid \sigma_1 \times \sigma_2 \mid \mathsf{bool}.
+        \sigma ::=\ & x \mid \sigma_1 \times \sigma_2.
     \end{align*}
 \end{definition}
-Here $x$ is a variable. $\sigma_1 \times \sigma_2$ is the product type, which is used to represent the type of tensor product space, or Cartesian product sets.
-$\mathsf{bool}$ is a constant index with two basis, and is used in Dirac notations of qubit systems.
+Here $x$ is a variable. $\sigma_1 \times \sigma_2$ is the product type for tensor product space or Cartesian product sets.
+% $\mathsf{bool}$ is a constant index with two basis, and is used in Dirac notations of qubit systems.
 
@@
 T$ denotes index functions, which accepts an index argument $x$ and results in a $T$ type term, where $T$ can depend on $x$.
+Index functions are necessary for defining transformations that are polymorphic on Hilbert spaces.
 
@@
     \begin{align*}
-        e ::=\ & x \mid \lambda x : T.e \mid \mu x.e \mid e_1\ e_2 \mid e_1 \circ e_2 \\
-        & |\ \hat{0} \mid \hat{1} \mid (e_1, e_2) \\
+        e ::=\ & x \mid \lambda x : T.e \mid \mu x.e \mid e_1\ e_2 \mid (e_1, e_2) \\
         & |\ 0 \mid 1 \mid \ADDS(e_1, \cdots, e_n)  \mid e_1 \times \cdots \times e_n \mid e^*  \mid \delta_{e_1, e_2} \mid \DOT(e_1, e_2) \\
@@
 $e_1\ e_2$ is function application. 
-$e_1 \circ e_2$ is the composition operator of Dirac notations.
-$\hat{0}$ and $\hat{1}$ are constant basis of $\Basis(\mathsf{bool})$ type.
+% $e_1 \circ e_2$ is the composition operator of Dirac notations.
+% $\hat{0}$ and $\hat{1}$ are constant basis of $\Basis(\mathsf{bool})$ type.
 $(e_1, e_2)$ is the basis pair for product types. $0$, $1$, $\ADDS$, $e_1 \times \cdots \times e_n$ and $e^*$ are symbols for scalars. 
@@
 X$ as well.
-
 Here $\ADDS$ and $\ADD$ are two different AC symbols representing the scalar addition and the linear algebra addition respectively. 
@@
 We encode the typing information by using five different symbols, namely $\DOT$, $\MULK$, $\MULB$, $\OUTER$ and $\MULO$.
+
+Compared to DiracDec, the addition, adjoint, scaling and tensor product symbols are merged together.
 They are denoted as $B\cdot K$, $K_1 \cdot K_2$, $B_1 \cdot B_2$, $K \cdot B$ and $O_1 \cdot O_2$, respectively.
-
-Compared to DiracDec, our language for Dirac notations merges the symbols with identital properties, such as the addition and scaling symbols for ket, bra and operator.
-One important difference is that AC symbols are treated as functions with indefinite arities.
+The multiplications are still separated because their properties do not have to much overlap.
 
@@
 
-The type checking of our language involves maintaining a well-formed environment and context $E[\Gamma]$, which is defined as follows.
+Typing is the procedure to classify the term through a proof system in a typing context. 
+Our context consists of the environment $E$ for definitions and assumptions, and the context $\Gamma$ for bound variables. 
+$E$ and $\Gamma$ are sequences of assumptions $x : T$ or definitions $x := t : T$.
 
@@
 \end{definition}
+Definitions refer to the symbols that can be unfolded.
+Definition symbols usually have abstract meanings, and here we use them to encode more operations in Dirac notations, such as transpose and trace.
+Assumptions are the declarations of the type for variables. In the scenario of equational proofs, variable assumptions implicitly have the universal quantifier.
 
-The environment and the context are sequences of assumptions $x : T$ or definitions $x := t : T$.
-With the environment, we can declare the type of variable symbols, and encode more operations on Dirac notations as definitions, such as the trace operator.
-Also, the existence of lambda abstractions requires a context of bound variables.
-
+The type checking of our language involves maintaining a well-formed environment and context $E[\Gamma]$.
 We say an expression $t$ has type $X$ in context $E[\Gamma]$, if the typing judgement $E[\Gamma] \vdash t : X$ can be proved through the typing rules in~\Cref{sec: full typing rules}.
@@
 Based on the well-formed context, typing judgements can be proved by information from $E[\Gamma]$, or built inductively.
-In the following rules, for example, the condition $x : \Index \in E[\Gamma]$ is true if $E$ or $\Gamma$ has the assumption in their sequences, and $\sigma \times \tau$ is a index if both $\sigma$ and $\tau$ are typed as the index.
-$\KType(\sigma)$ and $\OType(\sigma, \tau)$ will be valid types for kets and operators, if their arguments are typed as the index.
+In the following rules, for example, the condition $x : \Index \in E[\Gamma]$ is true if $E$ or $\Gamma$ has the assumption in their sequences.
+$\KType(\sigma)$ will be a valid type for kets, if the argument $\sigma$ is typed as the index.
 \begin{gather*}
@@
     \qquad
-    \frac{E[\Gamma] \vdash \sigma : \Index \qquad E[\Gamma] \vdash \tau : \Index}{E[\Gamma] \vdash \sigma \times \tau : \Index} \\
-    \\
     \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \KType(\sigma) : \Type}
-    \qquad
-    \frac{E[\Gamma] \vdash \sigma : \Index \qquad E[\Gamma] \vdash \tau : \Index}{E[\Gamma] \vdash \OType(\sigma, \tau) : \Type}
 \end{gather*}
@@
 The typing for functions and applications follow the common practice. 
-We specify the syntax and typing rule for functions of indices $\mu x.
-t$: here $x$ is a bound variable of typed as $\Index$, and the type $U\{x/u\}$ of application $(t u)$ is obtained by replacing $x$ with the index instance $u$.
+Take the index function $\mu x.
+t$ as an example: here $x$ is a bound variable typed as $\Index$, and the type $U\{x/u\}$ of application $(t\ u)$ is obtained by replacing $x$ with the index instance $u$.
 
@@
 
-The big operator sum is modelled by folding a function on a set, therefore the typing rule is as follows:
+Lastly, the big operator sum is modelled by folding a function on a set, therefore the typing rule is as follows:
 \[
@@
 
-And lastly we have the typing rules for composition $x \circ y$. Similar to casual Dirac notations, the typing of composition depends on the types of operands.
-\begin{gather*}
-    \frac{E[\Gamma] \vdash x : \SType \qquad E[\Gamma] \vdash y : \KType(\sigma)}{E[\Gamma] \vdash x \circ y : \KType(\sigma)}
-    \qquad
-    \frac{E[\Gamma] \vdash x : \OType(\sigma, \tau) \qquad E[\Gamma] \vdash y : \KType(\tau)}{E[\Gamma] \vdash x \circ y : \KType(\sigma)}
-\end{gather*}
+% And lastly we have the typing rules for composition $x \circ y$. Similar to casual Dirac notations, the typing of composition depends on the types of operands.
+% \begin{gather*}
+%     \frac{E[\Gamma] \vdash x : \SType \qquad E[\Gamma] \vdash y : \KType(\sigma)}{E[\Gamma] \vdash x \circ y : \KType(\sigma)}
+%     \qquad
+%     \frac{E[\Gamma] \vdash x : \OType(\sigma, \tau) \qquad E[\Gamma] \vdash y : \KType(\tau)}{E[\Gamma] \vdash x \circ y : \KType(\sigma)}
+% \end{gather*}
 
@@
 
-Semantics define the meanings of expressions, and the goal of the desirable algorithm is to decide whether
-two expressions have the equivalent semantics. We can give the semantics by axiomatic equations and the denotational interpretations.
+% Semantics define the meanings of expressions, and the goal of the desirable algorithm is to decide whether
+% two expressions have the equivalent semantics. We can give the semantics by axiomatic equations and the denotational interpretations.
 
-\subsubsection{Denotational semantics} The denotational way interprets every expression as a set theory object in linear algebra, according to a valuation mapping $v$ from variables to their values.
-This explanation formalizes the common explanation of Dirac notations, and best describes the target of the algorithm.
+To give meanings to the syntax, we assign the semantics in the denotational way, which is a mapping from syntax to set-theoretical objects.
+
+\subsubsection{Denotational semantics} 
+The denotational semantics interpret every expression as an object in linear algebra, according to a valuation mapping $v$ from variables to their values. 
 The semantics of $e$ with valuation $v$ is written as $\sem{e}_v$, and two expressions $e_1$ and $e_2$ are equivalence if they have equal semantics in the mathematical sense for all valuations, written as $\sem{e_1} = \sem{e_2}$.
 
+The complete interpretation is listed in \yx{We need to write this}.
 Variables typed with $\Index$ are interpreted as finite sets, and index product $\sem{\sigma_1 \times \sigma_2} = \sem{\sigma_1} \times \sem{\sigma_2}$ are defined as Cartesian product of sets.
@@
 \end{lemma}
+\begin{proof}
+    Directly check all the cases.
+\end{proof}
+
+This explanation formalizes the common explanation of Dirac notations, and best describes the target of the algorithm.
+However, computers cannot reason about equivalence through mathematical interpretations. We move on define the proof system as an abstraction.
 
 \subsubsection{Axiomatic semantics} 
-An other way is to express the equivalence by a set of axioms. 
+The proof system for equivalence consists of equational logic as the framework and axioms describing the properties of Dirac notations.
+The full list is in \yx{TODO}.
+There are axioms for linear space, and other structures like tensor product and inner product.
 % From the operational view, each equation declares a valid rewriting operation, and two expressions are equiavlent if and only if they can be rewritten into the same form using the axioms.
@@
 
-Note that some equational theories cannot be decided by term rewriting, and are collected in the set $E$:
+As is mentioned in the introduction, a subset $E$ of the axioms is not decided by term rewriting. The set consists of:
 
@@
     & \textrm{AC-equivalence} 
-    && \textrm{e.g.} \qquad a + b = b + a \qquad (a + b) + c = a + (b + c) \\
+    && \textrm{e.g.} \qquad X + Y = Y + X \qquad (X + Y) + Z = X + (Y + Z) \\
     & \textrm{$\alpha$-equivalence}
     && \lambda x . A = \lambda y . A\{x/y\} \\
-    & \textrm{(SUM-SWAP)}
-    && \sum_{i \in s_1} \sum_{j \in s_2} A = \sum_{j \in s_2} \sum_{i \in s_1} A
+    & \textrm{SUM-SWAP}
+    && \sum_{i \in s_1} \sum_{j \in s_2} A = \sum_{j \in s_2} \sum_{i \in s_1} A \\
+    & \textrm{scalar theories} 
+    && \textrm{e.g.} \qquad a + 0 = a \qquad a \times (b + c) = a \times b + a \times c
 \end{align*}
 
-The equations should also include axioms for scalars. In the implementation, we use the Mathematica kernel to decide scalar equivalences, so the axioms are not given here.
+The equational theories for scalars are treated as a moduloe, and are not considered here. In the implementation, we use the Mathematica kernel to decide scalar equivalences.
 
@@
 and the denotational semantics can be considered as one model for the theory.
-In this sense, the equivalence derived by axioms is weaker than the denotational semantics.
+In this sense, the equivalence derived by axioms always imply the equivalence in interpretations.
 
@@
 \end{lemma}
+\begin{proof}
+    Direct check all cases.
+\end{proof}
 
-Unfortunately, the inverse does not hold: there are expressions that are equivalent in denotational interpretations but not provable through the axioms.
-One notorious example is $\delta_{i, j} \times \delta_{j, k} = \delta_{i, k} \times \delta_{j, k}$.
-Intuitively, the two terms will be evaluated to $1$ if and only if $i = j = k$, but it is difficult to incorporate such properties as axioms.
-This incapability further limits the algorithmm, which is designed to automate reasonings using the axioms, and blocks us from proving complete results with respect to the semantics.
-But our work focuses on solving the practical examples, which are covered by our axioms, as is shown by experiments.
+Therefore, the axioms are sound with respect to denotational semantics. 
+Unfortunately, the inverse does not hold: there are equivalences in denotational semantics that cannot be captured by these axioms. 
+However, our work focuses on solving the practical examples, which are covered by our axioms in the experiments.
 
-\yx{where should we put the example?}
-To conclude the design introduction, we demonstrate part of the formalization of the motivating example.
+To conclude the design introduction, we demonstrate the formalization of symbols used in the motivating example.
 \begin{example}[formalizing the motivating example]
-    Definitions and assumptions in the environment:
+    \label{ex: formalizing motivating}
+    Definitions and assumptions in the environment $E$:
     \begin{align*}
@@
     \end{align*}
-    The symbol TPO is the encoding of operator transpose. It takes two index $T_1$ and $T_2$, and transforms the operators.
+    The symbol TPO is the encoding of operator transpose, which is polymorphic on the Hilbert spaces by $T_1$ and $T_2$.
     phi takes the index $T$ and defines the maximally entangled states. It sums over all basis in $T$, indicated by the universal set $\mathbf{U}(T)$.
-    With the assumption of index T and operator M, we can write the equivalence as 
+    With the assumption of index T and operator M, we can write the equivalence in the non-label version, formally as 
     \[
@@
 
-\section{Algorithm for Deciding Dirac notations equations}
+\section{Algorithm for Deciding Dirac notation equations}
 \label{sec: decide}
+
+We use the term rewriting technique to decide the equivalence of Dirac notation implied by its axiomatic semantics.
+Term rewriting decides equivalence by normalizing the two terms, checking whether they have the same syntax.
+The normalization is powered by matching and substitution in universal algebra, using a set of rewriting rules.
+As
+
+% CONTINUE HERE
 
@@
 
+\begin{lemma}[completeness of transformation]
+    For any two terms $e_1$ and $e_2$ that is equivalent under AC equivalence, SUM-SWAP and $\alpha$-equivalence, they will have the same form after sort, swap and de Bruijn transformations.
+\end{lemma}
+
 % The order depends on there occurances in the last sorting result. If no occurance, then the order will depend on the set (for sum) and the type (for lambda abstraction only).
@@
 The main purpose of this work is to build a practical tool that works well in checking Dirac notation equations.
-The refinements and extensions above concludes in our implementation called Dirace.
-Dirace is a solver written in \CC\ working as a command line tool.
+The refinements and extensions above concludes in our implementation called Dirace, a solver written in \CC.
 It has a parser built by ANTLR4, and scalar reasonings are powered by a Mathematica kernel.
@@
 is tested on the benchmark of the DiracDec work, and succeeds in proving most of them efficiently. 
+It can be used from the command line interactively, or can be integrated into other \CC\ projects as a library.
 
-Here is the code for the motivating example.
+\subsubsection{Project Structure}
+The project structure is illustrated in~\Cref{fig: dirace structure}.
+\texttt{ualg} is the module for universal algebra, defining basic concepts like terms and substitutions.
+It serves as the library for \texttt{dirace}, which are then utilized in the example benchmarks and the toplevel command line application.
+The components of \texttt{dirace} are as follows:
+
+\begin{itemize}
+    \item \texttt{symbols.cpp}: the reserved symbols and AC symbols;
+    \item \texttt{syntax\_theory.cpp}: syntax related algorithms, such as de Bruijn normalization and freeness of variables;
+    \item \texttt{calculus.cpp}: type checker and intergration with Mathematica;
+    \item \texttt{reduction.cpp}: all the rewriting rules and transformations;
+    \item \texttt{dirac\_parser.cpp}: parser for Dirac notations and Dirace commands;
+    \item \texttt{prover.cpp}: the prover that maintains the context and process commands like definition or equivalence checking.
+\end{itemize}
+
+\input{dirace.tex}
+
+The internal data structure for terms is a pointer-based syntax tree following the function application style:
+\[
+    \texttt{
+        term ::= ID | ID [term (, term)*].
+    }
+\]
+The syntax tree can be an itentifier, or an application with an identifier as the function head, and several syntax trees as arguments.
+There are several Dirac notations terms and their corresponding syntax trees.
+\footnotesize{
+\begin{align*}
+    & X_1 + X_2 + X_3 && \texttt{ADD[X1, X2, X3]} 
+    \\
+    & \lambda x: \OType(T_1,  T_2). x^\dagger && \texttt{FUN[x, OTYPE[T1, T2], ADJ[x]]}
+    \\
+    & \sum_{i \in \mathbf{U}(T)} \ket{i} \bra{i} && \texttt{SUM[USET[T], FUN[i, BASIS[T], OUTER[KET[i], BRA[i]]]]}
+\end{align*}
+}
+The syntax tree structure is also compatible with the datatype of Mathematica. This improves the interoperability between Dirace and the Mathematica system, enable them to work interleavingly.
+To improve usability, Dirace also supports many special syntacies for terms,  and most Dirac notation terms will be encoded in the natural way.
+Here are some examples for the parsing syntax.
+
+\begin{figure}
+    \center
+\begin{tabular}{c >{\centering\arraybackslash}p{4cm} l}
+    \hline
+    syntax & parsing result & explanation \\
+    \hline
+    \texttt{|e>} & \texttt{KET[e]} & the ket basis\\
+    \texttt{e1 + ... + en} & \texttt{ADD[e1, ..., en]} & the addition\\
+    \texttt{e1\ e2} & \texttt{COMPO[e1, e2]} & composition in Dirac notations \\
+    \texttt{e1\^{}*} & \texttt{CONJ[e1]} & scalar conjugation \\
+    \texttt{fun i : T => X} & \texttt{FUN[i, T, X]} & lambda abstraction \\
+    \hline
+\end{tabular}
+\end{figure}
+
+Finally, Dirace uses a prover to host the computation.
+The prover maintains a well-formed context $E[\Gamma]$, and processes commands to modify the context and conduct calculations.
+The commands are listed below.
+\begin{itemize}
+    \item \texttt{\textcolor{NavyBlue}{Def} ID := term.} It defines the \texttt{ID} as the \texttt{term}, using the \textbf{W-Def} typing rule.
+    \item \texttt{\textcolor{NavyBlue}{Var} ID := term.} It make an assumption of \texttt{ID} with the \texttt{term} as type, using the \textbf{W-AssumeE} typing rules.
+    \item \texttt{\textcolor{NavyBlue}{Check} term.} Type checking the \texttt{term} and output the result.
+    \item \texttt{\textcolor{NavyBlue}{Normalize} term.} Normalize the \texttt{term} using the algorithm introduced in~\Cref{sec: decide}.
+    \item \texttt{\textcolor{NavyBlue}{CheckEq} term \textcolor{NavyBlue}{with} term.} Check the equivalence of the two terms calculating and comparing their normal forms.
+\end{itemize}
+The prover will type check the terms for each command.
+We can also use \texttt{\textcolor{NavyBlue}{Normalize} term \textcolor{NavyBlue}{with trace}.} to output the proof trace during normalization.
+The proof trace is a sequence of records, including the rule or transformation appied, the position of application, and the pre- and post-transformation terms.
+The record helps understand the normalization procedure better, and can be turned into verified proofs in theorem provers in the future.
+
+\subsubsection{Use Case}
+Here we encode the motivating~\Cref{ex: motivating}, examine and explain how it is checked in Dirace. The encoding is shown below.
 
@@
 
-The first three lines use the \texttt{Var} and \texttt{Def} commands to set up the context for the Dirac notation.
-Here \texttt{T} is a type index, representing arbitrary Hilbert space type.
-\texttt{M} is assume to be an operator in Hilbert space with type \texttt{T}.
-\texttt{phi} is defined as the maximally entangled state.
+The first three lines use the \texttt{\textcolor{NavyBlue}{Var}} and \texttt{\textcolor{NavyBlue}{Def}} commands to set up the context for the Dirac notation.
+\texttt{T} is a type index, representing arbitrary Hilbert space types.
+\texttt{M} is assume to be an operator in the Hilbert space with type \texttt{T}.
+\texttt{phi} is defined as the maximally entangled state, depending on the bound variable \texttt{T} as index.
+\texttt{r1} and \texttt{r2} are register names for the two subsystems.
 
-Besides, the symbol \texttt{TPO} represents transpose of operators, and the definition is provided as
+In the left hand side of \texttt{\textcolor{NavyBlue}{CheckEq}} command, \texttt{M\_r1;r1} denotes the labelled notation $M_{r_1; r_1}$, and \texttt{(phi T)\_(r1, r2)} denotes the entangled state $\ket{\Phi}_{(r_1, r_2)}$.
+They are connected by a white space, which is parsed into the composition of Dirac notation, and will be reduced into the operator-ket multiplication after typing.
+The right hand side is interpreted similarly, except the defined symbol \texttt{TPO} in the environment:
+
 \begin{lstlisting}[style=dirace]
@@
 
-And the output result is:
+The \texttt{TPO} symbol represents the transpose of operators, and encodes the formalization in~\Cref{ex: formalizing motivating}.
+Thanks to the design of environment and functions, many other commonly used symbols in Dirac notations are encoded and provided as defined symbols in Dirace.
+
+Within one second, the prover reports the result of equivalence with their common normal form:
     \begin{lstlisting}[style=dirace]
@@
 
-The normal form can be transformed into the formatted version:
+The normal form is in the internal syntax tree format mentioned above. A more readable interpretation is:
+\[
+\sum_{\mathbf{U}(T)} \sum_{\mathbf{U}(T)} \bra{\$1}M\ket{\$ 0} . \ket{\$1}_{r_1} \otimes \ket{\$0}_{r_2} : \DType(\{r_1, r_2\}, \emptyset).
+\]
+Here $\$0$ and $\$1$ are de Bruijn indices. The result is a ket on the $\{r_1, r_2\}$ system as expected, and follows pattern proposed in~\Cref{lem: labelled normal form}.
 
-\[
-\sum_{\mathbf{U}(T)} \sum_{\mathbf{U}(T)} \bra{\$1}M\ket{\$ 0} . \ket{\$1}_{r1} \otimes \ket{\$0}_{r2} : \DType(\{r1, r2\}, \emptyset)
-\]
-Here $\$0$ and $\$1$ are de Bruijn indices. The result is a ket on the \{r1, r2\} system as expected.
+\section{Benchmark performance}
 
-\subsection{Benchmark performance}
-It does not support fst and snd projectors.
-It does not support interactive proof, as is used in the reasonings of quantum programming language theory.
+To evaluate Dirace, we first test the examples from DiracDec benchmark and make a comparison.
+The experiments are carried out using a MacBook Pro with M3 Max chip.
+
+\begin{figure}
+    \center
+    \begin{tabular}{c|c c c|c c c}
+        \hline
+        \multirow{2}{*}{type} & \multicolumn{3}{c|}{DiracDec} & \multicolumn{3}{c}{Dirace} \\
+        \cline{2-7}
+                                 & expressable & success & time(s)           & expressable & success & time(s)                 \\
+        \hline
+        textbook(QCQI)          & 18          & 18        &    1.02        &    18      & 18          &   0.82      \\
+        CoqQ                    & 162          & 156       &    48.69       &   158     &  158   &     9.74     \\
+        Circuits                 & 2          & 2       &    17.67       &   3     &  2   &     1.4     \\
+        Jens2024                & 4          & 4         &  59.53       &   4    & 4       &  0.73     \\
+        \hline
+    \end{tabular}        
+    \caption{For DiracDec, examples that cannot be decided within 60 seconds are not included.}
+\end{figure}
+
+The timing of Dirace does not include the initialization of Mathematica kernel link, which takes about 3 seconds in the beginning.
+As to expressibility, the language for DiracDec has the support for projectors $\texttt{fst}$ and $\texttt{snd}$ on basis pairs, satisfying $\texttt{fst} (s, t) = s$ and $\texttt{snd} (s, t) = t$.
+We found this feature is almost not used, so we removed the support.
+As a result, Dirace encodes 158 examples for the CoqQ part, 4 less than DiracDec.
+For decidability, we improved the rewriting rules about sum for Dirace, and proved several more examples failed by DiracDec.
+The main difference is about their time efficiency.
+Because of our algorithm to decide AC-equivalence and SUM-SWAP, Dirace has a significant efficiency improvement, especially on those ``computational examples'' mentioned in the DiracDec paper.
+One typical example comes from the paper by Jens about the equivalence of operators for qubits.
+The system has to decompose the term on the concrete $\ket{0}$ and $\ket{1}$ basis, resulting a lot of addition elements.
+It takes DiracDec about one minute, but Dirace solves it within one second.
+
+We also built an example benchmark for labelled Dirac notation.
+\yx{We need examples.}
 
@@
 
-Other previous works exploring the language and algorithm to reasong about quantum computation differently.
+\yx{Other previous works exploring the language and algorithm to reasong about quantum computation differently.}
 
@@
 
-\section{\texttt{Diracoq} language}
-
-\begin{align*}
-    cmd ::=&\ \texttt{Def}(\texttt{ID}\ term) \\
-        &|\ \texttt{Def}(\texttt{ID}\ term\ type) \\
-        &|\ \texttt{Var}(\texttt{ID}\ term) \\
-        &|\ \texttt{Check}(term) \\
-        &|\ \texttt{Show}(\texttt{ID}) \\
-        &|\ \texttt{ShowAll} \\
-        &|\ \texttt{Normalize}(term) \mid \texttt{Normalize}(term\ \textsf{Trace}) \\
-        &|\ \texttt{CheckEq}(term\ term) \\
-    type ::=&\ \texttt{Type} \mid \texttt{Arrow}(type\ type) \\
-            &|\ \texttt{Base} \\
-    term ::=&\ \texttt{Type} \mid \texttt{fun}(\texttt{ID}\ type\ term) \mid \texttt{apply}(term\ term) \mid \texttt{ID} \\
-\end{align*}
-
-Comment \texttt{(* ... *)} can be inserted between commands.
-
-These are tye parsing rules for different expressions:
-\begin{itemize}
-    \item \texttt{Def ID := term.} --- \texttt{Def(ID term)}
-    \item \texttt{Def ID := term : type.} --- \texttt{Def(ID term type)}
-    \item \texttt{Var ID : type.} --- \texttt{Var(ID type)}
-    \item \texttt{Check ID.} --- \texttt{Check(term)}
-    \item \texttt{Show ID.} --- \texttt{Show(ID)}
-    \item \texttt{ShowAll.} --- \texttt{ShowAll}
-    \item \texttt{Normalize term.} --- \texttt{Normalize(term)}
-    \item \texttt{Normalize term with trace.} --- \texttt{Normalize(term Trace)}
-    \item \texttt{Check term = term.} --- \texttt{CheckEq(term term)}
-    \item \texttt{T1 -> T2} --- \texttt{Arrow(T1 T2)}.
-    \item \texttt{forall x. T} --- \texttt{Forall(x T)}
-    \item \texttt{(e1, e2)} --- \texttt{PAIR(e1 e2)}
-    \item \texttt{fun x : T => e} --- \texttt{fun(x T e)}
-    \item \texttt{idx x => e} --- \texttt{idx(x e)}
-    \item \texttt{e1 @ e2} --- \texttt{COMPO(e1 e2)}
-    \item \texttt{e1 + ... + en} --- \texttt{ADDG(e1 ... en)}
-    \item \texttt{e1 * ... * en} --- \texttt{STAR(e1 ... en)}
-    \item \texttt{e1\^{}*} --- \texttt{CONJ(e1)}
-    \item \texttt{delta(e1, e2)} --- \texttt{DELTA(e1 e2)}
-    \item \texttt{|e>} --- \texttt{KET(e)}
-    \item \texttt{<e|} --- \texttt{BRA(e)}
-    \item \texttt{e1\^{}D} --- \texttt{ADJ(e1)}
-    \item \texttt{e1.e2} --- \texttt{SCR(e1 e2)}
-    \item \texttt{Sum(i in s, e)} --- \texttt{SSUM(i s e)}
-\end{itemize}
-
-These symbols will be transformed into internal language:
-\begin{itemize}
-    \item $A \circ B$ : $S \circ S$, $S \circ K$, $S \circ B$, $S \circ O$, $K \circ S$, $K \circ K$, $K \circ B$, $B \circ S$, $B \circ K$, $B \circ B$, $B \circ O$, $O \circ S$, $O \circ K$, $O \circ O$, $f\circ a$ (arrow), $f\circ a$ (index).
-    \item \texttt{STAR(a ... b)} :  $\sigma_1 \times \sigma_2$, $\MULS(a \cdots b)$, $O_1 \otimes O_2$, $ M_1 \star M_2$
-    \item \texttt{ADDG(e1 ... en)} : $\ADDS(e1 \cdots en)$, $\ADD(e1 \cdots en)$
-    \item \texttt{SSUM(i S e)} : \texttt{SUM(s FUN(i T e))}
-\end{itemize}
-
 % \begin{credits}

### 2025-01-24 labelled.tex
2025/1/24

--- 2025-01-20 labelled.tex
@@
 \section{Labelled Dirac Notation}
-
+\label{sec: labelled}
 Labelled Dirac notation uses register names to indicate the quantum system of vectors and operators.
-This allows us to talk aboud the states and operations locally, without referring to the whole system.
+This allows us to express and reason aboud the states and operations locally, without referring to the whole system.
 For instance, assume $Q$ and $R$ are two registers, we have
 \[
-    M_Q \cdot \ket{\Phi}_{(Q, R)} = ((M \otimes I) \ket{\Phi})_{(Q, R)}.
+    M_Q \cdot \ket{\Phi}_{(Q, R)} = ((M \otimes I) \cdot \ket{\Phi})_{(Q, R)}.
 \]
@@
 
-In this section, we introduce the language and typing of registers and labelled Dirac notations, and demonstrate how to transform the equivalence problem into the Dirac notations studied above.
-
-We assume the set of register symbols is $\cR$. The syntax of quantum registers is defined as follows.
+In this section, we introduce the language and typing of registers and labelled Dirac notations, and demonstrate how to transform the equivalence problem into that for the Dirac notations studied above.
+The first new notion is quantum registers, and we assume they are constructed from a set $\cR$.
 \begin{definition}[quantum registers]
@@
 
+% The labelled Dirac notation is an extension on the existing type and syntax.
 \begin{definition}[labelled Dirac notation]
@@
     e & ::= R \mid |i\>_r \mid {}_r\<i| \mid e_R \mid e_{R;R} \mid
-    e \otimes e \otimes \cdots \otimes e \mid e \cdot e \\
+    e \otimes e \otimes \cdots \otimes e \mid e \cdot e
   \end{align*}
 \end{definition}
-
 $\DType(s_1, s_2)$ is the unified type for all labelled Dirac notations, where $s_1$ indicates the codomain systems and $s_2$ indicates the domain systems.
@@
 We introduce new dot and tensor product symbols for labelled Dirac notations.
-In labelled Dirac notation, the structure of tensor product does not matter.
-Therefore $\otimes$ is an AC symbol.
+In labelled Dirac notation, the structure of tensor product does not matter, therefore $\otimes$ is an AC symbol.
 Following the unified type $\DType(s,s)$, all kinds of multiplications are represented by the same dot product $e \cdot e$.
@@
 
-\subsubsection*{Typing rules}
+\subsubsection{Typing rules}
 Some typing rules are introduced here.
@@
 The rule for $K_R$ demonstrates how a register label is added to the Dirac notation, and the rule for $D^\dagger$ shows that labelled Dirac notation also have symbols for calculation, such as the adjoint.
-
 \[
@@
 \]
-
 \[
@@
 \]
-
 The dot and the tensor product symbols are different from those in unlabelled Dirac notation.
@@
 
-\subsection{Normalization}
-The main idea of normalizing labelled Dirac notations is \textit{label elimination}. It expands $e_R$ and $e_{R;R}$ by tensor products of labelled basis, which are formalized as rules below.
-
+\subsubsection{Normalization}
+The normalization procedure of Dirac notation is extended to check equivalence with labels.
+We add rules to the term rewriting system, which in general try to represent labelled Dirac notations with labelled basis and scalar coefficients.
+The first step is the label elimination.
+Take operator as an example:
 \begin{align*}
-    & K_R \ \reduce\ \sum_{i_{r_1}\in\bU(\sigma_{r_1})}\cdots \sum_{i_{r_n}\in\bU(\sigma_{r_n})} (\<i_R|\cdot K). (|i_{r_1}\>_{r_i}\otimes\cdots\otimes|i_{r_n}\>_{r_n}) \\
-    %
-    & B_R \ \reduce\ \sum_{i_{r_1}\in\bU(\sigma_{r_1})}\cdots \sum_{i_{r_n}\in\bU(\sigma_{r_n})} (B\cdot |i_R\>). ({}_{r_1}\<i_{r_1}|\otimes\cdots\otimes{}_{r_n}\<i_{r_n}|) \\
-    %
     & O_{R,R'} \ \reduce\ \sum_{i_{r_1}\in\bU(\sigma_{r_1})}\cdots \sum_{i_{r_n}\in\bU(\sigma_{r_n})}
     \sum_{i_{r'_1}\in\bU(\sigma_{r'_1})}\cdots \sum_{i_{r'_{n'}}\in\bU(\sigma_{r'_{n'}})} \\
-    & \qquad (\<i_R|\cdot O\cdot |i_{R'}\>).(|i_{r_1}\>_{r_i}\otimes\cdots\otimes|i_{r_n}\>_{r_n} \otimes {}_{r'_1}\<i_{r'_1}|\otimes\cdots\otimes{}_{r'_{n'}}\<i_{r'_{n'}}|)
+    & \qquad (\<i_R|\cdot O\cdot |i_{R'}\>).(|i_{r_1}\>_{r_i}\otimes\cdots\otimes|i_{r_n}\>_{r_n} \otimes {}_{r'_1}\<i_{r'_1}|\otimes\cdots\otimes{}_{r'_{n'}}\<i_{r'_{n'}}|).
 \end{align*}
-
-This step reduces all terms in the form of $K_R$ or $O_{R;R}$.
+The rules for $e_R$ (ket and bra) are similar. This step reduces all labelled terms $e_R$ or $e_{R;R}$.
 Other symbols on labelled Dirac notations are also reduced by rules like $(D_1 \cdot D_2)^\dagger \ \reduce\ D_2^\dagger \cdot D_1^\dagger$.
-
-The final part of rules operates on sum and dot product. They will lift summation to the outside, and eliminate the bra-ket pairs whenever possible.
-
+The final step operates on sum and dot product. They will lift summation to the outside, and eliminate the bra-ket pairs whenever possible.
 \begin{align*}
@@
     && X_1 \otimes \cdots (\sum_{i \in M} D) \cdots \otimes X_2\ \reduce\ \sum_{i \in M} (X_1 \otimes \cdots D \cdots \otimes X_n) \\
-    %
-    & \textrm{(R-SUM-PUSHD1)}
-    && (\sum_{i \in M} D_1) \cdot D_2 \ \reduce\ \sum_{i \in M} (D_1 \cdot D_2) \\
+    % %
+    % & \textrm{(R-SUM-PUSHD1)}
+    % && (\sum_{i \in M} D_1) \cdot D_2 \ \reduce\ \sum_{i \in M} (D_1 \cdot D_2) \\
     %
@@
     & \textrm{(R-L-SORT2)}
-    && {}_r\bra{i}\cdot(Y_1 \otimes \cdots \otimes \ket{j}_r \otimes \cdots \otimes Y_m) \ \reduce\ \delta_{i, j}.(Y_1  \otimes \cdots \otimes Y_m) \\
+    && {}_r\bra{i}\cdot(Y_1 \otimes \cdots \otimes \ket{j}_r \otimes \cdots \otimes Y_m) \ \reduce\ \delta_{i, j}.(Y_1  \otimes \cdots \otimes Y_m)
     %
 \end{align*}
-
-These rules are added to the rewriting system in ~\Cref{sec: decide} and executed together.
-In the end, if there are no variables of $\DType(s_1, s_2)$, the expression will be reduced to the addition of big operator sum, and each sum body is Dirac notation scalar with labelled basis.
-
-\begin{lemma}
-    For a well-typed term $e$ in $E[\Gamma]$, if $e$ does not contain variables of $\DType(s_1, s_2)$ type, the normal form of $e$ will be
+% These rules are added to the rewriting system in~\Cref{sec: decide} and executed together.
+In the end, if there are no variables of $\DType(s_1, s_2)$, the expression will be reduced to the addition of big operator sum, and each sum body is labelled basis with Dirac notation scalar coefficients:
 \[
@@
 \]
-    Where $a_i$ are scalar typed Dirac notations.
-\end{lemma}
+In this way, we only need to check the scalars to decide the equivalence of labelled Dirac notations.
+
+% \begin{lemma}[normal form]
+%     \label{lem: labelled normal form}
+%     For a well-typed term $e$ in $E[\Gamma]$, if $e$ does not contain variables of $\DType(s_1, s_2)$ type, the normal form of $e$ will be
+%     \[
+%     \sum_{i}\cdots\sum_{j} a_1 . (\ket{i}_{p} \otimes \cdots \otimes \bra{j}_{q})
+%     + \cdots +
+%     \sum_{k}\cdots\sum_{l} a_m . (\ket{k}_{r} \otimes \cdots \otimes \bra{l}_{s})
+%     \]
+%     Where $a_i$ are scalar typed Dirac notations.
+% \end{lemma}
+% \begin{proof}
+%     Labelled Dirac notations of $e_R$, $e_R;R$, 
+% \end{proof}

### 2025-01-24 main.tex
2025/1/24

@@
 \usetikzlibrary{trees}
+\tikzstyle{arrow} = [thick,->,>=stealth]
 
@@
 To reason about such equations using computers, we need to formalize the language and work in a proof system.
-The following paragraphs briefly introduce the Dirac notation and the universal algebra framework we use.
+% The following paragraphs briefly introduce the Dirac notation and the universal algebra framework we use.
 
@@
 
-With the concrete basis, Dirac notations can be interpreted as matrices.
+% With the concrete basis, Dirac notations can be interpreted as matrices.
 For example, the ket of two basis of qubit states are column vectors $\ket{0} = \begin{bmatrix} 1 \\ 0 \end{bmatrix}$ and $\ket{1} = \begin{bmatrix} 0 \\ 1 \end{bmatrix}$, and the corresponding bras are row vectors in the dual space $\bra{0} = \begin{bmatrix} 1 & 0 \end{bmatrix}$ and $\bra{1} = \begin{bmatrix} 0 & 1 \end{bmatrix}$.
-The calculation between Dirac notations then corresponds to matrix operatoins.
+% The calculation between Dirac notations then corresponds to matrix operatoins.
 For example, inner product $\braket{0| 1} = \begin{bmatrix} 0 & 1 \end{bmatrix} \begin{bmatrix} 1 \\ 0 \end{bmatrix} = 0$, and tensor product $\ket{0} \otimes \ket{1} = \ket{(0, 1)} = \begin{bmatrix} 1 \\ 0 \end{bmatrix} \otimes \begin{bmatrix} 0 \\ 1 \end{bmatrix} = \begin{bmatrix} 0 \\ 1 \\ 0 \\ 0 \end{bmatrix}$.
@@
 In our case of Dirac notation, the signature consists of constructors and operations like $\ket{i}$ or $A \otimes B$.
-Reasonings of Dirac notation terms follow the equational logic, which defines an equaitonal relation $e_1 \doteq e_2$ compatible with substitution and term construction:
-\[
-    \frac{E \vdash s \doteq t}{E \vdash v(s) \doteq v(t)}
-    \qquad
-    \frac{E \vdash s_1 \doteq t_1 \qquad \cdots \qquad E \vdash s_n \doteq t_n}{E \vdash f(s_1, \dots, s_n) \doteq f(t_1, \dots, t_n)}
-\]
-This logic essentially formalizes the common sense of equivalence in algebra.
+Reasonings of Dirac notation terms follow the equational logic.
+It defines an equaitonal relation compatible with substitution and term construction, which essentially formalizes the common sense of equivalence in algebra.
 
@@
 Index represents classical datatypes, and they appear in type expressions to distinguish between different Hilbert spaces and sets.
-
 \begin{definition}[index syntax]
@@
 Here $x$ is a variable. $\sigma_1 \times \sigma_2$ is the product type for tensor product space or Cartesian product sets.
-% $\mathsf{bool}$ is a constant index with two basis, and is used in Dirac notations of qubit systems.
 
@@
 \end{definition}
-
 The generators are explained in order.
@@
 $E$ and $\Gamma$ are sequences of assumptions $x : T$ or definitions $x := t : T$.
-
 \begin{definition}[environment and context]
@@
 t$ as an example: here $x$ is a bound variable typed as $\Index$, and the type $U\{x/u\}$ of application $(t\ u)$ is obtained by replacing $x$ with the index instance $u$.
-
 \begin{gather*}
@@
 \end{lemma}
-
 \begin{proof}
@@
 As is mentioned in the introduction, a subset $E$ of the axioms is not decided by term rewriting. The set consists of:
-
 \begin{align*}
@@
 \end{align*}
-
 The equational theories for scalars are treated as a moduloe, and are not considered here. In the implementation, we use the Mathematica kernel to decide scalar equivalences.
@@
 In this sense, the equivalence derived by axioms always imply the equivalence in interpretations.
-
 \begin{lemma}
@@
 We use the term rewriting technique to decide the equivalence of Dirac notation implied by its axiomatic semantics.
-Term rewriting decides equivalence by normalizing the two terms, checking whether they have the same syntax.
+Term rewriting decides equivalence by normalizing the two terms, and checking whether they have the same syntax.
 The normalization is powered by matching and substitution in universal algebra, using a set of rewriting rules.
-As
+As is mentioned in the sematnics, the axioms in $E$ cannot be decided by term rewriting.
+DiracDec decides $E$ by examining all possible permutations in the rewriting result.
+In this work, the general idea is to carry through the normalization procedure using sorting algorithms. 
+The procedure of the normalization is displayed below. 
+The first three steps use term rewriting to work on the structure of Dirac notations, while the last three steps deals with the equational theory $E$.
+\begin{enumerate}
+    \item \textbf{first rewritings}: expand definitions and simplify,
+    \item \textbf{variable expansion}: consider scalars expressions for completeness,
+    \item \textbf{second rewritings}: normalize modulo $E$,
+    \item \textbf{sorting without bound variables}: normalize AC-equivalence,
+    \item \textbf{swap successive summation}: normalize SUM-SWAP,
+    \item \textbf{de Bruijn normalization}: normalize $\alpha$-equivalence.
+\end{enumerate}
 
-% CONTINUE HERE
-
-The following two sections discuss how to decide the equivalence of Dirac notations implied by the axioms.
-In the previous work, the equational axioms are separated into two parts: a set $E$ of equations that can not be decided by rewriting, and the remaining equations to be decided by term rewriting.
-Here, the equational theories $E$ inlude:
-\begin{itemize}
-    \item commutativity of symbols $a + b$, $a\times b$ and $\delta_{s, t}$,
-    \item $\alpha$-equivalence of bound variables, i.e., $\lambda x.t = \lambda y . t\{x/y\}$,
-    \item swapping successive summations, i.e., $\sum_{i \in s_1} \sum_{j \in s_2} A = \sum_{j \in s_2} \sum_{i \in s_1}A$, and
-    \item equational theories for scalars.
-\end{itemize}
-
-In this work, the general idea is to carry through the normalization procedure, so that semantical equivalence can be directly checked by the syntax of normal forms.
-The procedure of the normalization is displayed in~\Cref{fig: normalization}.
-
-\input{flowchart.tex}
-
-The last three steps deals with the equational theory $E$, while the first three steps use term rewriting to work on the structure of Dirac notations.
+% \input{flowchart.tex}
 
@@
 Term rewriting rules, written as $l \to r$, are used to normalize terms by recursively matching the subterms of the term with the left hand side $l$, and replace them with the corresponding right hand side $r$.
-The procedure terminates when no more rewritings can be made, and the order of rewritings will be irrelevant if the term rewriting system is \textit{confluent}, which is a desirable property.
+The procedure terminates when no more rewritings can be made.
+The full list of rewriting rules are in~\Cref{sec: rewriting rules}, and we present some of them to illustrate the design idea.
 
-To express the term rewriting system, the previous work adheres to use the naive universal algebra, where each function symbol has a fixed arity, and the left hand side pattern only allows constant symbols and variables.
-This constraint enables them to check the confluence and termination of their term rewriting system with other tools.
-
-% Here to enable more efficient algorithms, our langauge uses AC symbols with a variable arity.
-
-The full list of rewriting rules are in~\Cref{sec: rewriting rules}. Here we present some of them to illustrate the design idea.
-The whole rule list consists of several types.
-Firstly, we use the following flattening rule to decide the associative with AC symbols.
-
+Firstly, we use the following flattening rule to decide the associative with AC symbols, because we allow functions with indefinite arities.
 \[
 a_1 + \cdots + (b_1 + \cdots + b_m) + \cdots + a_n
-\ \reduce\ a_1 + \cdots + b_1 + \cdots + b_m + \cdots + a_n,
+\ \reduce\ a_1 + \cdots + b_1 + \cdots + b_m + \cdots + a_n
 \]
-
-and commutativity is left for a sorting algorithm later on.
-
-Some equational axioms imply obvious rewriting diraction, for example:
-
+Commutativity is left for the sorting algorithm later on. 
+Many rewriting rules are obtained by assigning directions to the equational axioms, for example:
 \begin{align*}
@@
     &\textrm{(R-DELTA0)} && \delta_{s, s} \ \reduce\ 1 \\
-    &\textrm{(R-MULK1)} && O : \OType(\sigma, \tau) \Rightarrow O \cdot \mathbf{0}_{\KType(\tau)} \ \reduce \mathbf{0}_{\KType(\sigma)}
+    &\textrm{(R-MULK1)} && O : \OType(\sigma, \tau) \Rightarrow O \cdot \mathbf{0}_{\KType(\tau)} \ \reduce \mathbf{0}_{\KType(\sigma)} \\
+    &\textrm{(R-MULK11)} && (O_1 \otimes O_2) \cdot (K_1 \otimes K_2)\ \reduce\ (O_1 \cdot K_1) \otimes (O_2 \cdot K_2)
 \end{align*}
-
-The (R-DOT6) says that inner product of two basis will be evaluated to a delta expression,
+Some of them are obvious. The (R-DOT6) says that inner product of two basis will be evaluated to a delta expression,
 and (R-DELTA0) further transforms delta on identical basis to scalar $1$.
 (R-MULK1) corresponds to the axiom that multiplication on zero vectors results in zero. It has a condition on typing, which is calculated during rewriting.
-They share the common feature that right hand side is simpler than the left hand side.
+Some rules require the intuition. For example, (R-MULK11) shows that we prefer tensor product than multiplication.
 
-Other part of rules assign the rewriting diraction to equations according to the property of symbols and intuitive observations of their relations.
-When different rules rewrites the same term to different forms, rules for confluence purpose are added.
+As a reference, the term rewriting system for DiracDec is proved complete for the axioms, except the sum symbol.
+The result is obtained by checking confluence and termination of the system.
+Our rewriting rules are translations from DiracDec into the typed and abstract language, therefore completeness for corresponding symbols is also expected to hold.
 
-\begin{align*}
-    %
-    & \text{(R-SCR1)} 
-    && a.(b.X) \ \reduce\ (a \times b).X \\
-    %
-    & \text{(R-DOT10)}
-    && (B \cdot O) \cdot K \ \reduce\ B \cdot (O \cdot K), \\
-    %
-    & \text{(R-DOT11)}
-    && \bra{(s, t)} \cdot ((O_1 \otimes O_2) \cdot K) \ \reduce\ ((\bra{s} \cdot O_1) \otimes (\bra{t} \cdot O_2)) \cdot K, \\
-    %
-    & \text{(R-MULB10)}
-    && \bra{(s, t)} \cdot (O_1 \otimes O_2)\ \reduce\ (\bra{s} \cdot O_1) \otimes (\bra{t} \cdot O_2),
-\end{align*}
-The (R-SCR1) rule follows the idea to reduce linear algebra scaling into scalar multiplications,
-and (R-MULB10) transform inner product into tensor product. 
-As an example of the completetion rule, here the normalization of term $(\bra{(s, t)} \cdot (O_1 \otimes O_2)) \cdot K$ have two rewriting paths: (a).
+%%%%%%%%%%% It's better to avoid mentioning completion rules.
+% Other part of rules assign the rewriting diraction to equations according to the property of symbols and intuitive observations of their relations.
+% When different rules rewrites the same term to different forms, rules for confluence purpose are added.
+% \begin{align*}
+%     %
+%     & \text{(R-SCR1)} 
+%     && a.(b.X) \ \reduce\ (a \times b).X \\
+%     %
+%     & \text{(R-DOT10)}
+%     && (B \cdot O) \cdot K \ \reduce\ B \cdot (O \cdot K), \\
+%     %
+%     & \text{(R-DOT11)}
+%     && \bra{(s, t)} \cdot ((O_1 \otimes O_2) \cdot K) \ \reduce\ ((\bra{s} \cdot O_1) \otimes (\bra{t} \cdot O_2)) \cdot K, \\
+%     %
+%     & \text{(R-MULB10)}
+%     && \bra{(s, t)} \cdot (O_1 \otimes O_2)\ \reduce\ (\bra{s} \cdot O_1) \otimes (\bra{t} \cdot O_2),
+% \end{align*}
+% The (R-SCR1) rule follows the idea to reduce linear algebra scaling into scalar multiplications,
+% and (R-MULB10) transform inner product into tensor product. 
+% As an example of the completetion rule, here the normalization of term $(\bra{(s, t)} \cdot (O_1 \otimes O_2)) \cdot K$ have two rewriting paths: (a).
 apply (R-MULB10) and get $((\bra{s} \cdot O_1) \otimes (\bra{t} \cdot O_2)) \cdot K$, or
-(b).
+% (b).
 first apply (R-DOT10) and sort the term into $\bra{(s, t)} \cdot ((O_1 \otimes O_2) \cdot K)$, and then apply (R-DOT11) to get the same result.
@@
 
-There are also rules dealing with sum expressions. For example,
-
+We have other rules dealing with big operator sum. For example, 
 \begin{align*}
@@
 \end{align*}
-
 The (R-SUM-ELIM2) rule will try to eliminate the delta expression in summations.
 And the (R-SUM-PUSH5) rule will lift sum to the outside of inner product.
+There are no completeness guarantee, but the rules work well in practice.
 
-One important technique revealed in the DiracDec work is the expansion of variables:
-
+One technique revealed in the DiracDec work is the expansion of variables, critical for proofs with sum:
 \begin{gather*}
@@
 \end{gather*}
-The above three rules are not terminating, and this is why we have the rewriting-expansion-rewriting steps in the algorithm.
+These rules transform variables to the symbolic summation of their decomposition on the basis.
+The rules are not terminating, therefore is applied recrusively once in the second step called \textit{variable expansion}.
 On the other hand, we discovered that doing the expansion on all variables only once is already sufficient.
-
 \begin{lemma}
-    Let $\text{expand}(e)$ indicate the result of expanding all variables in $e$ once.
-    For all well-typed term $e$ in $E[\Gamma]$, we have $\text{expand}(\text{expand}(e)) \downarrow = \text{expand}(e)\downarrow$.
+    Let $\textrm{expand}(e)$ indicate the result of expanding all variables in $e$ once.
+    For all well-typed term $e$ in $E[\Gamma]$, $\textrm{expand}(\textrm{expand}(e))$ and $\textrm{expand}(e)$ have the same normal form.
 \end{lemma}
 \begin{proof}
+    If we expand a ket variable twice, the structure will be reduced to
+    \[
+        \sum_{i \in \mathbf{U}(\sigma)}(\bra{i}\cdot \sum_{j \in \mathbf{U}(\sigma)}(\bra{j}\cdot K).\ket{j}).\ket{i}
+        \ \reduce\ 
+        \sum_{i \in \mathbf{U}(\sigma)}\sum_{j \in \mathbf{U}(\sigma)}(\bra{j}\cdot K \times \braket{i|j}).\ket{i},
+    \]
+    the elimination rule of delta symbol returns the term to the one-expansion result. Similar for bra and operator.
 \end{proof}
@@
 
-In the Mathematica implementation of DiracDec, the equational theory $E$ is decided by a unification, which tries to find a substitution of summation bound variables and AC symbol arguments that makes the two expressions syntactically equivalent.
-The unification iterates through all permutations, and the complexity is factorial in general.
+In the Mathematica implementation of DiracDec, the equational theory $E$ is decided by a unification, which tries to find a substitution of summation bound variables that makes the two expressions syntactically equivalent.
+The unification iterates through all permutations of AC symbol arguments, and the complexity is factorial in the maximum number of AC symbol arguments.
 
-As the first improvement, the $\alpha$-equivalence are checked by de Bruijn index~\cite{deBruijn1972lambda}, where references to bound variable names are replaced by the distance of the lambda abstraction to the variable.
+As the first improvement, we check the $\alpha$-equivalence by de Bruijn index~\cite{deBruijn1972lambda}, where references to bound variable names are replaced by the distance of the lambda abstraction to the variable.
 For example, the nominal lambda abstraction $\lambda x.
 x$ is transformed into $\lambda .
-1$, while $\lambda x.
+0$, while $\lambda x.
 \lambda y.
 (x\ (y\ x))$ will is transformed into $\lambda.\lambda.
-(2\ (1\ 2))$.
-Therefore, transformation into de Bruijn index is at the last step of the normalization to check the equivalence of terms with different bound variable names.
+(1\ (0\ 1))$.
+Transformation into de Bruijn index is at the last step of the normalization to check the equivalence of terms with different bound variable names.
 
-A standard approach to decide this permutation equivalence is to normalize by sorting with a given order.
+The remaining axioms of AC-equivalence and SUM-SWAP claims equivalence under permutations.
+A standard approach to decide is normalization by sorting in a given order.
 For example, given the dictionary order $a < b < c$, the term $b + c + a$ (and any other AC equivalent ones) will be normalized into $a + b + c$.
-For our setting, there are two difficulties intertwined together: how to assign such an order to all terms in our language, and how to normalize with respect to $E$.
+For our setting, there are two difficulties intertwined together: how to assign such an order to all terms in our language, and how to sort for the two axioms simultaneously.
 As a typical example, the two terms 
@@
 \]
-are equivalent due to the swap of summations and commutativity of scalar multiplication. However,
-directly sorting elemnts of scalar multiplication will not lead to the same form.
+are equivalent. However, directly sorting elemnts of scalar multiplication in the lexical order will not lead to the same form.
 
@@
 In the above example, we ignore the bound variables first, and sort the sum body into $\bra{\bullet}A\ket{\bullet} \times \bra{\bullet}B\ket{\bullet}$.
-Then we swap the sum so that the bound variable at first $\bullet$ position appears at the outside, and the two terms will have the same de Bruijn normal form as $\sum_{s_1} \sum_{s_2} \bra{\$1} A \ket{\$0} \times \bra{\$0} B \ket{\$1}$.
+Then we swap the sum, so that the bound variable at first $\bullet$ position appears at the outmost position.
+The results will have the same de Bruijn normal form, namely $\sum_{s_1} \sum_{s_2} \bra{\$1} A \ket{\$0} \times \bra{\$0} B \ket{\$1}$.
 
-To describe the algorithm, we introduce two notations: for a term $e = f(a_1, a_2, \cdots, a_n)$, $\textrm{head}(e)$ indicates the function symbol $f$,
+To describe the algorithm in the following, we introduce two notations: for a term $e = f(a_1, a_2, \cdots, a_n)$, $\textrm{head}(e)$ indicates the function symbol $f$,
 and $\textrm{arg}(e, i)$ indicates the $i$-th argument $a_i$. 
@@
 With the order, the sort function normalize the term with respect to AC equivalence.
-
 \begin{definition}[sort transformation]
@@
 
-The order of bound variables can be decided according to the result of the sort transformation, and then successive summations can be normalized by swapping.
+The order of bound variables can be decided according to the result of the sort transformation, and then successive summations can be normalized using the order.
 \begin{definition}[swap transformation]
@@
 
-\begin{lemma}[completeness of transformation]
+The algorithm succeeds in all equivalences of our benchmark, and we formalize its completeness as a conjecture.
+\begin{conjecture}[completeness of transformation]
     For any two terms $e_1$ and $e_2$ that is equivalent under AC equivalence, SUM-SWAP and $\alpha$-equivalence, they will have the same form after sort, swap and de Bruijn transformations.
-\end{lemma}
+\end{conjecture}
 
@@
 
-At last, we will prove that equivalence by this normalization procedure is sound in the semantics.
-
+Lastly, we can prove that equivalence by this normalization procedure is sound with respect to the semantics. 
 \begin{theorem}[soundness]
@@
 \begin{proof}
-
+    The soundness of term rewriting holds because each rewriting rule preserves the equivalence.
+The operations in sort and swap transformation follows AC-equivalence and SUM-SWAP axioms.
+The de Bruijn normal form is sound for $\alpha$-equivalence.
 \end{proof}
@@
 The components of \texttt{dirace} are as follows:
-
 \begin{itemize}
@@
 \]
-The syntax tree can be an itentifier, or an application with an identifier as the function head, and several syntax trees as arguments.
+The syntax tree can be an identifier, or an application with an identifier as the function head, and several syntax trees as arguments.
 There are several Dirac notations terms and their corresponding syntax trees.
@@
 }
-The syntax tree structure is also compatible with the datatype of Mathematica. This improves the interoperability between Dirace and the Mathematica system, enable them to work interleavingly.
+The syntax tree structure is also compatible with the datatype of Mathematica. This improves the interoperability between Dirace and the Mathematica system, enabling them to work interleavingly.
 To improve usability, Dirace also supports many special syntacies for terms,  and most Dirac notation terms will be encoded in the natural way.
@@
 \]
-Here $\$0$ and $\$1$ are de Bruijn indices. The result is a ket on the $\{r_1, r_2\}$ system as expected, and follows pattern proposed in~\Cref{lem: labelled normal form}.
+Here $\$0$ and $\$1$ are de Bruijn indices. The result is a ket on the $\{r_1, r_2\}$ system as expected, and follows pattern proposed in~\Cref{sec: labelled}.
 
-\section{Benchmark performance}
+\subsubsection{Benchmark performance}
 
@@
 
-\yx{Other previous works exploring the language and algorithm to reasong about quantum computation differently.}
+Automated theorem proving has seen significant advancements in recent years, particularly through the development of satisfiability modulo theories (SMT) solvers,
+% SMT solvers extend propositional satisfiability by integrating theories such as arithmetic, arrays, and bit-vectors, making them suitable for solving a wide range of verification problems in both hardware and software domains.
+including prominent tools like Z3. These solvers have become essential in various fields such as formal verification, synthesis, and model checking. 
+Equational reasoning is another crucial area of research within automated theorem proving, which focuses on solving problems that involve equations between terms in an algebraic structure. 
+Equational provers, such as Vampire and E, have played a pivotal role in addressing the challenge of proving equations in first-order logic, employing sophisticated algorithms like superposition and term rewriting.
+
+Term rewriting.
+
+Other theories and tools are proposed for the formal verification of quantum computations.
+From the theoretical part, we have ... to verify zx calclus. Other tools for circuits, compilation, program.
 
@@
 Experimental results shows that the tool can correctly identify all encoded equivalent expressions efficiently.
-We expect Dirace will play a role in areas like quantum program verification or proofs of post-quantum cryptography protocols.
-
-\clearpage
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{TODO}
-\begin{itemize}
-    \item better output.
-    \item better error logic
-    \item the $\bar{U_{AC}}$, $\bar{U_{BC}}$ terms can be modelled by quantum registers.
-\end{itemize}
-
-\section{Things to Note}
-\begin{itemize}
-    \item The context should also be maintained during rewriting matching.
-    \item We don't allow eta reduction. It will intertwine with SUM-SWAP and break the confluence.
-    \item In each Delta reduction, the bound variables are replaced with unique variables. This should help solve the problem of conflicting variable names during substitution.
-\end{itemize}
+We expect Dirace will have applications in areas like quantum program verification or proofs of post-quantum cryptography protocols.
 

### 2025-01-25 labelled.tex
2025/1/25

@@
 
-In this section, we introduce the language and typing of registers and labelled Dirac notations, and demonstrate how to transform the equivalence problem into that for the Dirac notations studied above.
+In this section, we introduce the language and typing of registers and labelled Dirac notation, and demonstrate how to transform the equivalence problem into that for the Dirac notation studied above.
 The first new notion is quantum registers, and we assume they are constructed from a set $\cR$.
@@
 
-Registers can be composed by pairs of $(R, R)$, and this structure corresponds to the structure of tensor product spaces in Dirac notations.
+Registers can be composed by pairs of $(R, R)$, and this structure corresponds to the structure of tensor product spaces in Dirac notation.
 To reason about the registers, we define their variable set as the enumeration of all register symbols involved.
@@
 \end{definition}
-$\DType(s_1, s_2)$ is the unified type for all labelled Dirac notations, where $s_1$ indicates the codomain systems and $s_2$ indicates the domain systems.
+$\DType(s_1, s_2)$ is the unified type for all labelled Dirac notation, where $s_1$ indicates the codomain systems and $s_2$ indicates the domain systems.
 For instance, labelled ket has type $\DType(s_1, \emptyset)$, and labelled bra has type $\DType(\emptyset, s_2)$.
@@
 Terms also include the labelled notation $e_R$ for bra, ket and $e_{R;R}$ for operators.
-We introduce new dot and tensor product symbols for labelled Dirac notations.
+We introduce new dot and tensor product symbols for labelled Dirac notation.
 In labelled Dirac notation, the structure of tensor product does not matter, therefore $\otimes$ is an AC symbol.
 Following the unified type $\DType(s,s)$, all kinds of multiplications are represented by the same dot product $e \cdot e$.
-Finally, $\ket{i}_r$ and ${}_r\bra{i}$ are labelled basis for the normal form of labelled Dirac notations, where $r$ are registers symbols in $\mathcal{R}$.
+Finally, $\ket{i}_r$ and ${}_r\bra{i}$ are labelled basis for the normal form of labelled Dirac notation, where $r$ are registers symbols in $\mathcal{R}$.
 
@@
 The normalization procedure of Dirac notation is extended to check equivalence with labels.
-We add rules to the term rewriting system, which in general try to represent labelled Dirac notations with labelled basis and scalar coefficients.
+We add rules to the term rewriting system, which in general try to represent labelled Dirac notation with labelled basis and scalar coefficients.
 The first step is the label elimination.
@@
 The rules for $e_R$ (ket and bra) are similar. This step reduces all labelled terms $e_R$ or $e_{R;R}$.
-Other symbols on labelled Dirac notations are also reduced by rules like $(D_1 \cdot D_2)^\dagger \ \reduce\ D_2^\dagger \cdot D_1^\dagger$.
+Other symbols on labelled Dirac notation are also reduced by rules like $(D_1 \cdot D_2)^\dagger \ \reduce\ D_2^\dagger \cdot D_1^\dagger$.
 The final step operates on sum and dot product. They will lift summation to the outside, and eliminate the bra-ket pairs whenever possible.
@@
 \]
-In this way, we only need to check the scalars to decide the equivalence of labelled Dirac notations.
+In this stage, we only need to check the scalars to decide the equivalence of labelled Dirac notation.
 
@@
 %     \]
-%     Where $a_i$ are scalar typed Dirac notations.
+%     Where $a_i$ are scalar typed Dirac notation.
 % \end{lemma}
 % \begin{proof}
-%     Labelled Dirac notations of $e_R$, $e_R;R$, 
+%     Labelled Dirac notation of $e_R$, $e_R;R$, 
 % \end{proof}

### 2025-01-25 main.tex
2025/1/25

@@
 
-    Dirac notations is the fundamental language in quantum computation and quantum information.
-Recently, the term rewriting system DiracDec~\cite{diracdec} was introduced to automate the equational reasoning with Dirac notations, a critical yet time-intensive task.
+    Dirac notation is the fundamental language in quantum computation and quantum information.
+Recently, the term rewriting system DiracDec~\cite{diracdec} was introduced to automate the equational reasoning with Dirac notation, a critical yet time-intensive task.
 This work bases upon and extends DiracDec, aiming to develop a solver optimized for practical applications.
-Enhancements include an improved typing system, a simplified language and rewriting system, more efficient algorithms, and added support for labelled Dirac notations.
+Enhancements include an improved typing system, a simplified language and rewriting system, more efficient algorithms, and added support for labelled Dirac notation.
 Our solver Dirace is implemented in \CC\ with a Mathematica backend, and demonstrates advantages in decision-making power and time efficiency.
@@
 
-Recently, Xu et al.~\cite{diracdec} proposed a theory to decide the equivalence of Dirac notations, as well as a prototype implementation in Mathematica called DiracDec.
-They proved that the equivalence of basic Dirac notations are decidable.
+Recently, Xu et al.~\cite{diracdec} proposed a theory to decide the equivalence of Dirac notation, as well as a prototype implementation in Mathematica called DiracDec.
+They proved that the equivalence of basic Dirac notation are decidable.
 Their algorithm sticks with a pure term rewriting system, and is proved to be confluent and terminating.
@@
 DiracDec does not support labelled Dirac notation, which use registers to denote the subsystem, and express the state locally.
-The typing of DiracDec only deals with Dirac notations, which is not sufficient for a working scenario, where we want to have defined symbols and a context for the typing assumptions of variables.
+The typing of DiracDec only deals with Dirac notation, which is not sufficient for a working scenario, where we want to have defined symbols and a context for the typing assumptions of variables.
 Moreover, to avoid type checking during term rewriting, DiracDec separate the same symbol (e.g. multiplication) for different types to disambiguate, which makes the system bloated.
@@
 
-There is a trade-off between simplicity and efficiency for the system design.
+The system design involves a trade-off between simplicity and efficiency.
 The simplicity of term rewriting in DiracDec allows them to reach satisfying theoretical results, but it is also a constraint for optimization.
@@
 It transforms the term rewriting system into a hybrid algorithm, and overcomes the problems mentioned above.
-Our main contributions and techniques are:
+Our main technical contributions are:
 \begin{itemize}
     \item An efficient algorithm to decide the equational theories in $E$.
-The basic idea is to carry through equivalence checking by normalization, and the normal form for $E$ are obtained by sorting.
-% This algorithm reduces the complexity from the factorial level to the polynomial level.
-    \item The support for constant register labels. The equivalence decision of labelled Dirac notations is reduced to the unlabelled situation.
+The core idea is to perform equivalence checking through normalization, with the normal form for $E$ being obtained by sorting.
+    \item The support for constant register labels. The equivalence decision of labelled Dirac notation is reduced to the unlabelled situation.
     \item A more usable \CC\ solver Dirace with abstract language and typing. We also support to define symbols (e.g., transpose and trace) using the functions syntax.
@@
 
-As an evaluation, we tested Dirace on the benchmark of DiracDec and new examples for labelled Dirac notations. 
+As an evaluation, we tested Dirace on the benchmark of DiracDec and new examples for labelled Dirac notation. 
 The result shows significant improvement in decidability and efficiency over DiracDec.
@@
 One interesting fact about the quantum world is that for two maximally entangled states, applying one operator $M$ on one subsystem is equivalent to applying $M^T$, the transpose of $M$, on the other.
-This property holds no matter how far the two systems are separated, and is expressed as an equation of Dirac notations.
+This property holds no matter how far the two systems are separated, and is expressed as an equation of Dirac notation.
 \begin{example}
@@
 \end{example}
-Here $q$ and $r$ are called labels, denoting the system for Dirac notations.
-To reason about such equations using computers, we need to formalize the language and work in a proof system.
+Here $q$ and $r$ are called labels, denoting the system for Dirac notation.
+To automate reasoning about such equations, we need to formalize the language and develop a proof system.
 % The following paragraphs briefly introduce the Dirac notation and the universal algebra framework we use.
@@
 
-% With the concrete basis, Dirac notations can be interpreted as matrices.
+% With the concrete basis, Dirac notation can be interpreted as matrices.
 For example, the ket of two basis of qubit states are column vectors $\ket{0} = \begin{bmatrix} 1 \\ 0 \end{bmatrix}$ and $\ket{1} = \begin{bmatrix} 0 \\ 1 \end{bmatrix}$, and the corresponding bras are row vectors in the dual space $\bra{0} = \begin{bmatrix} 1 & 0 \end{bmatrix}$ and $\bra{1} = \begin{bmatrix} 0 & 1 \end{bmatrix}$.
-% The calculation between Dirac notations then corresponds to matrix operatoins.
+% The calculation between Dirac notation then corresponds to matrix operatoins.
 For example, inner product $\braket{0| 1} = \begin{bmatrix} 0 & 1 \end{bmatrix} \begin{bmatrix} 1 \\ 0 \end{bmatrix} = 0$, and tensor product $\ket{0} \otimes \ket{1} = \ket{(0, 1)} = \begin{bmatrix} 1 \\ 0 \end{bmatrix} \otimes \begin{bmatrix} 0 \\ 1 \end{bmatrix} = \begin{bmatrix} 0 \\ 1 \\ 0 \\ 0 \end{bmatrix}$.
 
-In practice, Dirac notations are combined with other syntax, such as the big operator sum $\sum_{i \in S} A$, for better expressiveness.
-We also use labelled Dirac notations like $\ket{i}_q \otimes \ket{i}_r$ to denote the quantum system in consideration, where Dirac notations are subscripted with registers.
+In practice, Dirac notation are combined with other syntax, such as the big operator sum $\sum_{i \in S} A$, for better expressiveness.
+We also use labelled Dirac notation like $\ket{i}_q \otimes \ket{i}_r$ to denote the quantum system in consideration, where Dirac notation are subscripted with registers.
 
 \subsubsection{Universal algebra}
-We use universal algebra and equational logic to formally represent Dirac notations and the reasoning procedure.
+We use universal algebra and equational logic to formally represent Dirac notation and the reasoning procedure.
 A universal algebra defines a signature of function symbols.
@@
 $e_1\ e_2$ is function application. 
-% $e_1 \circ e_2$ is the composition operator of Dirac notations.
+% $e_1 \circ e_2$ is the composition operator of Dirac notation.
 % $\hat{0}$ and $\hat{1}$ are constant basis of $\Basis(\mathsf{bool})$ type.
@@
 \end{definition}
-Definitions refer to the symbols that can be unfolded.
-Definition symbols usually have abstract meanings, and here we use them to encode more operations in Dirac notations, such as transpose and trace.
-Assumptions are the declarations of the type for variables. In the scenario of equational proofs, variable assumptions implicitly have the universal quantifier.
+Definitions refer to the symbols that can be unfolded. Definition symbols usually have abstract meanings, and here we use them to encode more operations in Dirac notation, such as transpose and trace.
+Assumptions are the declarations of the type for variables. In the scenario of equational proofs, variable assumptions are universally quantified implicitly.
 
@@
 
-The Dirac notations will then be typed accordingly.
+The Dirac notation will then be typed accordingly.
 For example, the ket syntax $\ket{t}$ has type $\KType(\sigma)$, if $t$ is typed as a basis term of index $\sigma$.
@@
 
-% And lastly we have the typing rules for composition $x \circ y$. Similar to casual Dirac notations, the typing of composition depends on the types of operands.
+% And lastly we have the typing rules for composition $x \circ y$. Similar to casual Dirac notation, the typing of composition depends on the types of operands.
 % \begin{gather*}
@@
 
-The complete interpretation is listed in \yx{We need to write this}.
+The complete interpretation is listed in~\Cref{sec: full denotational sem}.
 Variables typed with $\Index$ are interpreted as finite sets, and index product $\sem{\sigma_1 \times \sigma_2} = \sem{\sigma_1} \times \sem{\sigma_2}$ are defined as Cartesian product of sets.
@@
 
-This explanation formalizes the common explanation of Dirac notations, and best describes the target of the algorithm.
+This explanation formalizes the common explanation of Dirac notation, and best describes the target of the algorithm.
 However, computers cannot reason about equivalence through mathematical interpretations. We move on define the proof system as an abstraction.
@@
 \subsubsection{Axiomatic semantics} 
-The proof system for equivalence consists of equational logic as the framework and axioms describing the properties of Dirac notations.
-The full list is in \yx{TODO}.
+The proof system for equivalence consists of equational logic as the framework and axioms describing the properties of Dirac notation.
+The full list is in~\Cref{sec: full axioms}.
 There are axioms for linear space, and other structures like tensor product and inner product.
@@
 The procedure of the normalization is displayed below. 
-The first three steps use term rewriting to work on the structure of Dirac notations, while the last three steps deals with the equational theory $E$.
+The first three steps use term rewriting to work on the structure of Dirac notation, while the last three steps deals with the equational theory $E$.
 \begin{enumerate}
@@
     \item \texttt{reduction.cpp}: all the rewriting rules and transformations;
-    \item \texttt{dirac\_parser.cpp}: parser for Dirac notations and Dirace commands;
+    \item \texttt{dirac\_parser.cpp}: parser for Dirac notation and Dirace commands;
     \item \texttt{prover.cpp}: the prover that maintains the context and process commands like definition or equivalence checking.
@@
 The syntax tree can be an identifier, or an application with an identifier as the function head, and several syntax trees as arguments.
-There are several Dirac notations terms and their corresponding syntax trees.
+There are several Dirac notation terms and their corresponding syntax trees.
 \footnotesize{
@@
     \texttt{e1 + ... + en} & \texttt{ADD[e1, ..., en]} & the addition\\
-    \texttt{e1\ e2} & \texttt{COMPO[e1, e2]} & composition in Dirac notations \\
+    \texttt{e1\ e2} & \texttt{COMPO[e1, e2]} & composition in Dirac notation \\
     \texttt{e1\^{}*} & \texttt{CONJ[e1]} & scalar conjugation \\
@@
 The \texttt{TPO} symbol represents the transpose of operators, and encodes the formalization in~\Cref{ex: formalizing motivating}.
-Thanks to the design of environment and functions, many other commonly used symbols in Dirac notations are encoded and provided as defined symbols in Dirace.
+Thanks to the design of environment and functions, many other commonly used symbols in Dirac notation are encoded and provided as defined symbols in Dirace.
 
@@
         \hline
-        \multirow{2}{*}{type} & \multicolumn{3}{c|}{DiracDec} & \multicolumn{3}{c}{Dirace} \\
+        \multirow{2}{*}{source} & \multicolumn{3}{c|}{DiracDec} & \multicolumn{3}{c}{Dirace} \\
         \cline{2-7}
@@
         CoqQ                    & 162          & 156       &    48.69       &   158     &  158   &     9.74     \\
-        Circuits                 & 2          & 2       &    17.67       &   3     &  2   &     1.4     \\
-        Jens2024                & 4          & 4         &  59.53       &   4    & 4       &  0.73     \\
+        circuits                 & 2          & 2       &    17.67       &   3     &  2   &     1.4     \\
+        research paper                & 4          & 4         &  59.53       &   4    & 4       &  0.73     \\
         \hline
@@
 
-Term rewriting.
+Formal verification of quantum computation is receiving increasing attentions during these years. See~\cite{Lewis2023} for a comprehensive review. 
+Verification frameworks in Coq include the foundational formalization CoqQ and quantum circuit language QWIRE. 
+Verifications of quantum programs are also considered, such as the Hoare logic based methods [] and model checking based methods [].
+The equational reasoning of Dirac notation is crucial for establishing property proofs in these works.
+Verification theoreis and tools based on other languages are also proposed, such as PyZX for reasoning and simplification of ZX-calculus.
 
-Other theories and tools are proposed for the formal verification of quantum computations.
-From the theoretical part, we have ... to verify zx calclus. Other tools for circuits, compilation, program.
+\section{Conclusion and Future Work}
+Based on the first equational reasoning tool for Dirac notation called DiracDec, this work improves and extends the theory for practical applications, and provides the solver Dirace.
+Experiments show that the tool demonstrates advantages in decidability, efficiency and usability.
 
-\section{Conclusion}
-Based on the first Dirac notation equational reasoning tool DiracDec, this work refines and extends the theory for practical application, and provides the solver Dirace.
-Experimental results shows that the tool can correctly identify all encoded equivalent expressions efficiently.
-We expect Dirace will have applications in areas like quantum program verification or proofs of post-quantum cryptography protocols.
+We expect Dirace to have applications in areas like quantum program verification or proofs of post-quantum cryptography protocols in the future.
+One promising following up is to connect Dirace with theorem provers like Coq.
+It involves transforming theorem prover expressions into Dirace, and verify the proof trace of Dirace in theorem provers.
+Besids, most quantum program verifiers nowadays depend on matrix calculations.
+Dirace can serve as the replacement for matrix methods to enable symbolic deductions.
 

### 2025-01-25 main.tex
2025/1/25 GPT4 improved

@@
 \begin{abstract}
-
-    Dirac notation is the fundamental language in quantum computation and quantum information.
-Recently, the term rewriting system DiracDec~\cite{diracdec} was introduced to automate the equational reasoning with Dirac notation, a critical yet time-intensive task.
-This work bases upon and extends DiracDec, aiming to develop a solver optimized for practical applications.
-Enhancements include an improved typing system, a simplified language and rewriting system, more efficient algorithms, and added support for labelled Dirac notation.
-Our solver Dirace is implemented in \CC\ with a Mathematica backend, and demonstrates advantages in decision-making power and time efficiency.
+Dirac notation is a fundamental language in quantum computation and quantum information.
+Recently, the term rewriting system DiracDec~\cite{diracdec} was introduced to automate equational reasoning with Dirac notation, a critical yet time-intensive task.
+Building upon DiracDec, this work aims to develop a solver optimized for practical applications.
+Enhancements include an improved typing system, a simplified language and rewriting system, more efficient algorithms, and added support for labeled Dirac notation.
+Our solver, Dirace, is implemented in \CC\ with a Mathematica backend, demonstrating significant improvements in decision-making power and time efficiency.
 
@@
 
-In 1939, Dirac proposed his notation~\cite{dirac1939new} for quantum mechanics, which is designed to represent linear algebra formulae in a compact and convenient way.
-For instance, $a\ket{\psi} + b\ket{\phi}$ indicates the addition of two vectors, i.e., the superposition of two states $\ket{\psi}$ and $\ket{\phi}$.
-Dirac notation is now widely accepted as the working language in quantum computation and quantum information.
-The reasonings of Dirac notation play a fundamental role in research and application, just like boolean and integer logic to classical computer science.
-For example, formalization works of quantum algorithms and quantum programming languages involves plenty of equational proofs of Dirac notation, which are critical, repeating but time-intensive jobs.
-Dirac notation are also used in quantum programming languages to define the program states, operations and assertions.
-In order to automate the verification procedure, we need to simplify and check the equivalence of pre-conditions.
-However, unlike the existence of SAT and SMT solver, we are still in need of a working Dirac notation solver, and this absence has become an obstacle of many areas.
+In 1939, Dirac introduced his notation for quantum mechanics~\cite{dirac1939new}, designed to represent linear algebraic formulas in a compact and convenient form.
+For example, the expression \( a\ket{\psi} + b\ket{\phi} \) represents the superposition of two quantum states, \( \ket{\psi} \) and \( \ket{\phi} \).
+Today, Dirac notation is widely accepted as the standard language in quantum computation and quantum information.
+Its reasoning forms the foundation of research and applications, much like boolean and integer logic do in classical computer science.
 
-Recently, Xu et al.~\cite{diracdec} proposed a theory to decide the equivalence of Dirac notation, as well as a prototype implementation in Mathematica called DiracDec.
-They proved that the equivalence of basic Dirac notation are decidable.
-Their algorithm sticks with a pure term rewriting system, and is proved to be confluent and terminating.
-Even though, there is still a gap between DiracDec and a practical solver for Dirac notation equivalence.
+In quantum algorithm formalizations and quantum programming languages, Dirac notation is frequently used in equational proofs, which are critical, repetitive, and often time-intensive.
+These notations also play a key role in defining program states, operations, and assertions in quantum programming languages.
+To automate the verification of these programs, we need tools that can simplify and check the equivalence of preconditions.
+However, unlike the well-established SAT and SMT solvers for classical logic, a practical solver for Dirac notation equivalence remains an unmet need, creating a barrier for progress in several fields.
 
-One problem is the algorithm efficiency for equivalence out of the scope of rewritings.
-DiracDec decides the whole equational theory by rewriting modulo $E$, where $E$ is a set of equational axioms that cannot be decided by normalization in rewritings, e.g.
-the associativity and commutativity (AC) of some function symbols.
-DiracDec uses a direct by inefficient algorithm to decide $E$, which search through all possible permutations and has factorial complexity.
-This is witnessed in their evaluation: DiracDec is not good at dealing with ``computational examples'', which has many AC symbols and are therefore time consuming.
+Recently, Xu et al.~\cite{diracdec} proposed a theory for deciding the equivalence of Dirac notation, alongside a prototype implementation in Mathematica called DiracDec.
+They demonstrated that the equivalence of basic Dirac notation is decidable.
+Their algorithm, based on a pure term rewriting system, has been proven to be confluent and terminating.
+Despite this, there remains a gap between DiracDec and a practical solver for Dirac notation equivalence.
 
-Usability is another problem.
-DiracDec does not support labelled Dirac notation, which use registers to denote the subsystem, and express the state locally.
-The typing of DiracDec only deals with Dirac notation, which is not sufficient for a working scenario, where we want to have defined symbols and a context for the typing assumptions of variables.
-Moreover, to avoid type checking during term rewriting, DiracDec separate the same symbol (e.g. multiplication) for different types to disambiguate, which makes the system bloated.
-Also, it is hard to integrate the Mathematica implementation as a solvers into other projects.
+One challenge is the efficiency of the algorithm when dealing with equivalences beyond the scope of term rewriting.
+DiracDec decides the entire equational theory by rewriting modulo \( E \), where \( E \) represents a set of axioms that cannot be decided by normalization alone, such as the associativity and commutativity (AC) of certain function symbols.
+DiracDec uses a direct but inefficient algorithm to decide these axioms, which searches through all possible permutations and exhibits factorial complexity.
+This inefficiency becomes particularly evident in "computational examples" containing many AC symbols, which are time-consuming to process.
 
-The system design involves a trade-off between simplicity and efficiency.
-The simplicity of term rewriting in DiracDec allows them to reach satisfying theoretical results, but it is also a constraint for optimization.
-Based on DiracDec, this work aims at building a practical solver.
-It transforms the term rewriting system into a hybrid algorithm, and overcomes the problems mentioned above.
+Usability is another area where DiracDec falls short.
+It does not support labelled Dirac notation, which uses registers to denote subsystems and express states locally.
+Additionally, DiracDec's typing system only does not provide context for variable typing assumptions or the definition of symbols, which are required in practical scenarios.
+To avoid type checking during term rewriting, DiracDec separates symbols (e.g., multiplication) for different types, leading to unnecessary complexity.
+Moreover, integrating the Mathematica implementation into other projects as a solver is challenging.
+
+The system design of DiracDec reflects a trade-off between simplicity and efficiency.
+While the simplicity of term rewriting allows for strong theoretical results, it limits optimization opportunities.
+Building on DiracDec, this work aims to develop a practical solver.
+We transform the term rewriting system into a hybrid algorithm, overcoming the challenges mentioned above.
 Our main technical contributions are:
 \begin{itemize}
-    \item An efficient algorithm to decide the equational theories in $E$.
-The core idea is to perform equivalence checking through normalization, with the normal form for $E$ being obtained by sorting.
-    \item The support for constant register labels. The equivalence decision of labelled Dirac notation is reduced to the unlabelled situation.
-    \item A more usable \CC\ solver Dirace with abstract language and typing. We also support to define symbols (e.g., transpose and trace) using the functions syntax.
+    \item An efficient algorithm for deciding the equational theories in \( E \), based on equivalence checking through normalization, with the normal form for \( E \) being obtained via sorting.
+    \item Support for constant register labels, and reducing the equivalence decision of labelled Dirac notation to the unlabelled case.
+    \item A more user-friendly \CC\ solver, Dirace, featuring an abstract language and typing system. We also support the definition of symbols (e.g., transpose and trace) using function syntax.
 \end{itemize}
 
-As an evaluation, we tested Dirace on the benchmark of DiracDec and new examples for labelled Dirac notation. 
-The result shows significant improvement in decidability and efficiency over DiracDec.
+We evaluated Dirace against the DiracDec benchmark and new examples involving labelled Dirac notation.
+The results show significant improvements in both decidability and efficiency compared to DiracDec.
+
 % Dirace successfully decides all the examples that are expressable in its language, including those failed by DiracDec because of complexity or insufficient decision power.
 
-% \yx{mention the dilemma of simplicity and efficiency}
+\section{Motivation and Preliminary}
 
-\section{Motivation and Preliminary}
-One interesting fact about the quantum world is that for two maximally entangled states, applying one operator $M$ on one subsystem is equivalent to applying $M^T$, the transpose of $M$, on the other.
-This property holds no matter how far the two systems are separated, and is expressed as an equation of Dirac notation.
+An interesting property of quantum mechanics is that for two maximally entangled states, applying a quantum operator \( M \) to one subsystem is equivalent to applying \( M^T \) (the transpose of \( M \)) to the other subsystem.
+This relationship holds regardless of the spatial separation between the two systems, and it can be expressed as an equation in Dirac notation.
 \begin{example}
     \label{ex: motivating}
-    Let $q$ and $r$ be two quantum systems on $\mathcal{H}_T$ space. Let $M$ be a quantum operation on $\mathcal{H}_T$,
-    and $\ket{\Phi} = \sum_{i \in T} \ket{i}\otimes \ket{i}$ be the maximally entangled state, then we have
+    Let \( q \) and \( r \) represent two quantum systems in the Hilbert space \( \mathcal{H}_T \).
+Let \( M \) be a quantum operation acting on \( \mathcal{H}_T \), and let \( \ket{\Phi} = \sum_{i \in T} \ket{i} \otimes \ket{i} \) be the maximally entangled state.
+Then, we have the following equation:
     \[
@@
 \end{example}
-Here $q$ and $r$ are called labels, denoting the system for Dirac notation.
-To automate reasoning about such equations, we need to formalize the language and develop a proof system.
-% The following paragraphs briefly introduce the Dirac notation and the universal algebra framework we use.
 
-\subsubsection{Dirac notation}
-Quantum states live in complex Hilbert spaces. We use vectors in the space to descirbe pure quantum states, and operations are described by linear transformations. 
-It uses the ket $\ket{i}$ and the bra $\bra{i}$ to indicate bases of the space and the dual space.
-Together with other variable symbols, they are composed with each other in sequence, and the composition will be interpreted into different operations, depending on the type of operands.
-For example, $\braket{i|j}$ represents the inner product of $\bra{i}$ and $\ket{j}$, which is a scalar, while $\ket{i}\bra{j}$ represents the outer product, resulting in an operator.
-Dirac notation also use $\otimes$ to indicate the vectors and operators in the tensor product space.
+In this equation, \( q \) and \( r \) are labels denoting the respective subsystems in the Dirac notation.
+To automate reasoning about such equations, we must formalize the language and develop a proof system to handle it.
 
-Dirac notation further enjoys the property that the interpretation is independent on the order of composition, thus parentheses can be omitted.
-For example, the formula \(\bra{i}\ket{\phi}\bra{\psi}\ket{j}\) can be understood as
+\subsubsection{Dirac Notation}
+
+Quantum states are represented as vectors in complex Hilbert spaces, and operations on these states are described by linear transformations.
+Dirac notation uses the ket \( \ket{i} \) and the bra \( \bra{i} \) to denote basis vectors in a Hilbert space and its dual space, respectively.
+These symbols can be composed together in various ways to form more complex expressions.
+The meaning of these compositions depends on the types of the operands involved.
+For example, the inner product \( \braket{i|j} \) represents the scalar result of the dot product between \( \bra{i} \) and \( \ket{j} \), while the outer product \( \ket{i}\bra{j} \) represents an operator.
+Additionally, Dirac notation uses the tensor product symbol \( \otimes \) to describe the combined space of multiple quantum systems.
+
+One of the key features of Dirac notation is its order-independent interpretation.
+This means that the composition of terms can be written without parentheses, as the interpretation is unaffected by the order of multiplication.
+For instance, the formula \( \bra{i}\ket{\phi}\bra{\psi}\ket{j} \) can be understood as:
 \[
@@
 \]
-and they are equivalent for all variables.
+and these expressions are equivalent for all variables involved.
 
-% With the concrete basis, Dirac notation can be interpreted as matrices.
-For example, the ket of two basis of qubit states are column vectors $\ket{0} = \begin{bmatrix} 1 \\ 0 \end{bmatrix}$ and $\ket{1} = \begin{bmatrix} 0 \\ 1 \end{bmatrix}$, and the corresponding bras are row vectors in the dual space $\bra{0} = \begin{bmatrix} 1 & 0 \end{bmatrix}$ and $\bra{1} = \begin{bmatrix} 0 & 1 \end{bmatrix}$.
-% The calculation between Dirac notation then corresponds to matrix operatoins.
-For example, inner product $\braket{0| 1} = \begin{bmatrix} 0 & 1 \end{bmatrix} \begin{bmatrix} 1 \\ 0 \end{bmatrix} = 0$, and tensor product $\ket{0} \otimes \ket{1} = \ket{(0, 1)} = \begin{bmatrix} 1 \\ 0 \end{bmatrix} \otimes \begin{bmatrix} 0 \\ 1 \end{bmatrix} = \begin{bmatrix} 0 \\ 1 \\ 0 \\ 0 \end{bmatrix}$.
+In practice, Dirac notation is often combined with other syntactic elements, such as the summation symbol \( \sum_{i \in S} A \), to enhance expressiveness.
+Furthermore, labelled Dirac notation, such as \( \ket{i}_q \otimes \ket{i}_r \), is used to denote quantum systems in consideration, where subscripts (registers) are added to distinguish different subsystems.
 
-In practice, Dirac notation are combined with other syntax, such as the big operator sum $\sum_{i \in S} A$, for better expressiveness.
-We also use labelled Dirac notation like $\ket{i}_q \otimes \ket{i}_r$ to denote the quantum system in consideration, where Dirac notation are subscripted with registers.
-
-\subsubsection{Universal algebra}
+\subsubsection{Universal Algebra}
 We use universal algebra and equational logic to formally represent Dirac notation and the reasoning procedure.
-A universal algebra defines a signature of function symbols.
-Terms in the algebra are constructed by constants, variables or function application on terms. 
+A universal algebra defines a signature of function symbols, with terms constructed from constants, variables, and function applications.
 Other basic concepts like substitution of variables or pattern matching are also defined.
 In our case of Dirac notation, the signature consists of constructors and operations like $\ket{i}$ or $A \otimes B$.
-Reasonings of Dirac notation terms follow the equational logic.
-It defines an equaitonal relation compatible with substitution and term construction, which essentially formalizes the common sense of equivalence in algebra.
+The reasoning process is guided by equational logic, which defines an equivalence relation that is compatible with substitution and term construction.
+This relation formalizes the intuitive concept of equivalence in algebra.
 
 \section{Language, Typing and Semantics}
-The first step is to formally pin down the language for Dirac notation.
-As a reference, the language in DiracDec has a concrete design, meaning that the same syntax for different types corresponds to different symbols.
-Our intention is to move from the concrete design to a more abstract one. This is closer to the casual Dirac notation we use, and simplifies the term rewriting system.
+The first step is to formalize the language for Dirac notation.
+In DiracDec, the language has a concrete design, where the same syntax for different types corresponds to distinct symbols.
+Our goal is to transition from this concrete design to a more abstract one, which aligns more closely with the conventional Dirac notation we use and simplifies the term rewriting system.
 
-For this purpose, our language involves three layers: the index, the type and the term.
-Terms describe the concrete instances like ket, bra and operator, which will be checked and typed.
-Index represents classical datatypes, and they appear in type expressions to distinguish between different Hilbert spaces and sets.
-\begin{definition}[index syntax]
-    The syntax for type indices is
+To achieve this, we organize our language into three layers: the index, the type, and the term.
+Terms represent concrete instances such as kets, bras, and operators, which will be typed and checked.
+The index represents classical data types and appears in type expressions to differentiate between various Hilbert spaces and sets.
+
+\begin{definition}[Index Syntax]
+    The syntax for type indices is:
     \begin{align*}
@@
 \end{definition}
-Here $x$ is a variable. $\sigma_1 \times \sigma_2$ is the product type for tensor product space or Cartesian product sets.
+Here, \( x \) is a variable, and \( \sigma_1 \times \sigma_2 \) represents the product type for tensor product spaces or Cartesian product sets.
 
-\begin{definition}[type syntax]
-    The syntax for Dirac notation types is
+\begin{definition}[Type Syntax]
+    The syntax for Dirac notation types is:
     \begin{align*}
-        T ::=\ & \Basis(\sigma) \mid \SType \mid \KType(\sigma) \mid \BType(\sigma) \mid \OType(\sigma_1, \sigma_2) \mid T_1 \to T_2 \mid \forall x.T \mid \SET(\sigma). \\
+        T ::=\ & \Basis(\sigma) \mid \SType \mid \KType(\sigma) \mid \BType(\sigma) \mid \OType(\sigma_1, \sigma_2) \mid T_1 \to T_2 \mid \forall x.T \mid \SET(\sigma).
     \end{align*}
 \end{definition}
-$\Basis(\sigma)$ is the type for basis in index $\sigma$.
-$\SType$ indicates scalars, $\KType(\sigma)$ and $\BType(\sigma)$ indicates ket and bra terms in Hilbert space $\sigma$, and $\OType(\sigma_1, \sigma_2)$ indicates linear operators with Hilbert space $\sigma_2$ as domain and $\sigma_1$ as codomain.
-$\mathsf{Set}(\sigma)$ is the type of subsets in $\sigma$ index, and is used to denote the values of bound variables in summation.
-The remaining two generators are types for functions.
-$T_1 \to T_2$ denotes normal functions, which accepts a $T_1$ type argument and results in a $T_2$ type term.
-$\forall x.
-T$ denotes index functions, which accepts an index argument $x$ and results in a $T$ type term, where $T$ can depend on $x$.
-Index functions are necessary for defining transformations that are polymorphic on Hilbert spaces.
+\( \Basis(\sigma) \) denotes the type for basis elements in the index \( \sigma \).
+\( \SType \) represents scalars, while \( \KType(\sigma) \) and \( \BType(\sigma) \) refer to ket and bra types in the Hilbert space \( \sigma \), respectively.
+\( \OType(\sigma_1, \sigma_2) \) represents linear operators with \( \sigma_2 \) as the domain and \( \sigma_1 \) as the codomain.
+\( \SET(\sigma) \) refers to the type of subsets of \( \sigma \), used to denote the values of bound variables in summations.
+The remaining two constructs define function types: \( T_1 \to T_2 \) represents normal functions that take a \( T_1 \)-type argument and return a \( T_2 \)-type term, while \( \forall x.
+T \) represents index functions that take an index argument \( x \) and produce a \( T \)-type term, which may depend on \( x \).
+Index functions are essential for defining polymorphic transformations across Hilbert spaces.
 
-\begin{definition}[term syntax]
-    The syntax for Dirac notation terms is
+\begin{definition}[Term Syntax]
+    The syntax for Dirac notation terms is:
     \begin{align*}
@@
 \end{definition}
-The generators are explained in order.
-$\lambda x : T.e$ is the abstraction for normal functions, and $\mu x.e$ is the abstraction for index functions.
-$e_1\ e_2$ is function application. 
-% $e_1 \circ e_2$ is the composition operator of Dirac notation.
-% $\hat{0}$ and $\hat{1}$ are constant basis of $\Basis(\mathsf{bool})$ type.
-$(e_1, e_2)$ is the basis pair for product types. $0$, $1$, $\ADDS$, $e_1 \times \cdots \times e_n$ and $e^*$ are symbols for scalars. 
-The next line includes constant symbols for ket, bra and operator.
-$e^\dagger$ denotes the conjugate transpose of $e$. $e_1.e_2$ denotes scaling the term $e_2$ with scalar $e_1$.
-$\mathbf{U}(e)$ is the universal set with index $e$.
-$e_1 \star e_2$ is the Cartesian product.
-$\sum_{e_1} e_2$ is the big operator sum, which is modelled by folding the function $e_2$ over value sets $e_1$.
-Usually, the sum body is specified by an abstraction.
-Therefore we use the notation $\sum_{x \in s} X$ to denote $\sum_{s} \lambda x : T .
-X$ as well.
-Here $\ADDS$ and $\ADD$ are two different AC symbols representing the scalar addition and the linear algebra addition respectively. 
-% They will be denoted as $a_1 + \cdots + a_n$ and $X_1 + \cdots + X_n$.
-There are five kinds of linear algebra multiplications among ket, bra and operator.
-We encode the typing information by using five different symbols, namely $\DOT$, $\MULK$, $\MULB$, $\OUTER$ and $\MULO$.
+The terms above are explained in order:
+\( \lambda x : T.e \) represents the abstraction for normal functions, and \( \mu x.e \) represents the abstraction for index functions.
+\( e_1\ e_2 \) denotes function application.
+\( (e_1, e_2) \) is the basis pair for product types.
+The constants \( 0 \), \( 1 \), \( \ADDS \), \( e_1 \times \cdots \times e_n \), and \( e^* \) are symbols for scalars.
+The next line includes constant symbols for kets, bras, and operators.
+\( \ket{e} \) is a ket, \( \bra{t} \) is a bra, and \( e^\dagger \) denotes the conjugate transpose of \( e \).
+\( e_1.e_2 \) represents scaling the term \( e_2 \) by scalar \( e_1 \).
+\( \mathbf{U}(e) \) denotes the universal set with index \( e \).
+\( e_1 \star e_2 \) represents the Cartesian product of \( e_1 \) and \( e_2 \).
+\( \sum_{e_1} e_2 \) is the big operator sum, modeled by folding the function \( e_2 \) over the value set \( e_1 \).
+Typically, the sum's body is given by an abstraction.
+For convenience, we also use the notation \( \sum_{x \in s} X \) to represent \( \sum_{s} \lambda x : T .
+X \).
+Additionally, \( \ADDS \) and \( \ADD \) are distinct AC symbols: \( \ADDS \) is used for scalar addition and \( \ADD \) for linear algebra addition.
+There are five kinds of linear algebraic multiplications among kets, bras, and operators.
+These are encoded using five different symbols: \( \DOT \), \( \MULK \), \( \MULB \), \( \OUTER \), and \( \MULO \).
 
-Compared to DiracDec, the addition, adjoint, scaling and tensor product symbols are merged together.
-They are denoted as $B\cdot K$, $K_1 \cdot K_2$, $B_1 \cdot B_2$, $K \cdot B$ and $O_1 \cdot O_2$, respectively.
-The multiplications are still separated because their properties do not have to much overlap.
+Compared to DiracDec, the symbols for addition, adjoint, scaling, and tensor products have been consolidated.
+The multiplications are still kept separate due to their differing properties.
+These operations are denoted as \( B \cdot K \), \( K_1 \cdot K_2 \), \( B_1 \cdot B_2 \), \( K \cdot B \), and \( O_1 \cdot O_2 \), respectively.
 
@@
 
-Typing is the procedure to classify the term through a proof system in a typing context. 
-Our context consists of the environment $E$ for definitions and assumptions, and the context $\Gamma$ for bound variables. 
-$E$ and $\Gamma$ are sequences of assumptions $x : T$ or definitions $x := t : T$.
-\begin{definition}[environment and context]
+The typing system is responsible for classifying terms within a proof system, using a context that specifies the types of variables and definitions.
+The context is divided into two components: the environment \( E \), which includes definitions and assumptions, and the context \( \Gamma \), which tracks bound variables.
+Both \( E \) and \( \Gamma \) consist of sequences of assumptions \( x : T \) or definitions \( x := t : T \).
+\begin{definition}[Environment and Context]
+    The syntax for \( E \) and \( \Gamma \) is:
     \begin{align*}
-        E ::=\ & [] \mid E; x : \Index \mid E; x : T \mid E; x := t : T. \\
-        \Gamma ::=\ & [] \mid \Gamma; x : \Index \mid \Gamma; x : T.
+        E ::= &\ [] \mid E; x : \Index \mid E; x : T \mid E; x := t : T. \\
+        \Gamma ::= &\ [] \mid \Gamma; x : \Index \mid \Gamma; x : T.
     \end{align*}
 \end{definition}
-Definitions refer to the symbols that can be unfolded. Definition symbols usually have abstract meanings, and here we use them to encode more operations in Dirac notation, such as transpose and trace.
-Assumptions are the declarations of the type for variables. In the scenario of equational proofs, variable assumptions are universally quantified implicitly.
+Definitions refer to symbols that can be expanded or unfolded, and typically represent abstract concepts, such as transpose or trace in Dirac notation.
+Assumptions, on the other hand, define the types of variables, and variable assumptions are implicitly universally quantified in the case of equational proofs.
 
-The type checking of our language involves maintaining a well-formed environment and context $E[\Gamma]$.
-We say an expression $t$ has type $X$ in context $E[\Gamma]$, if the typing judgement $E[\Gamma] \vdash t : X$ can be proved through the typing rules in~\Cref{sec: full typing rules}.
-Here we present and explain the rules selectively.
-Firstly, well-formed contexts $\WF(E)[\Gamma]$ are built in the incremental way, e.g.:
+Type checking in our language involves maintaining a well-formed environment and context \( E[\Gamma] \).
+We say an expression \( t \) has type \( X \) in context \( E[\Gamma] \) if the typing judgment \( E[\Gamma] \vdash t : X \) can be proven through the rules in~\Cref{sec: full typing rules}.
+The well-formedness of the context \( \WF(E)[\Gamma] \) is built incrementally:
 \[
@@
     \qquad
-    \frac{E[]\vdash t:T \qquad x \notin E}{\WF(E; x:=t:T)[]}.
+    \frac{E[]\vdash t:T \qquad x \notin E}{\WF(E; x := t:T)[]}.
 \]
-Starting from an empty context, we can assume new index symbols, and assume or define symbols with checked types.
-Based on the well-formed context, typing judgements can be proved by information from $E[\Gamma]$, or built inductively.
-In the following rules, for example, the condition $x : \Index \in E[\Gamma]$ is true if $E$ or $\Gamma$ has the assumption in their sequences.
-$\KType(\sigma)$ will be a valid type for kets, if the argument $\sigma$ is typed as the index.
+Starting with an empty context, new index symbols can be introduced, and symbols with checked types can be defined.
+Typing judgments are then made inductively using the information from \( E[\Gamma] \).
+For instance, the rule \( x : \Index \in E[\Gamma] \) holds if \( E \) or \( \Gamma \) contains an assumption for \( x \).
+And \(\KType(\sigma)\) is a valid type for kets, if the argument $\sigma$ is typed as an index.
 \begin{gather*}
@@
 
-The Dirac notation will then be typed accordingly.
-For example, the ket syntax $\ket{t}$ has type $\KType(\sigma)$, if $t$ is typed as a basis term of index $\sigma$.
-Also, the inner product of a bra and a ket with the same type index $\sigma$ is typed as the scalar.
-This corresponds to the constraint of inner products that vectors should be in the same Hilbert space.
+Terms are then typed accordingly.
+For example, the ket \( \ket{t} \) will have the type \( \KType(\sigma) \) if \( t \) is a basis term of index \( \sigma \).
+Similarly, the inner product between a bra and a ket of the same index \( \sigma \) is typed as a scalar.
+It corresponds to the constraint of inner product that vectors should be from the same Hilbert space.
 \begin{gather*}
@@
 
-The typing for functions and applications follow the common practice. 
-Take the index function $\mu x.
-t$ as an example: here $x$ is a bound variable typed as $\Index$, and the type $U\{x/u\}$ of application $(t\ u)$ is obtained by replacing $x$ with the index instance $u$.
+The typing for functions and applications follows the standard practice.
+For example, the index function \( \mu x.
+t \) is typed with \( x \) as a bound variable of type \( \Index \), and the application \( (t\ u) \) has the type \( U\{x/u\} \), obtained by replacing \( x \) with the index instance \( u \).
 \begin{gather*}
@@
     \qquad
-    \frac{E[\Gamma] \vdash t:U \to T \qquad E[\Gamma] \vdash u:U}{E[\Gamma] \vdash (t\ u):T} \\
-    \\
+    \frac{E[\Gamma] \vdash t:U \to T \qquad E[\Gamma] \vdash u:U}{E[\Gamma] \vdash (t\ u):T} 
+    \\[1em]
     \frac{E[\Gamma; x : \Index] \vdash t : U}{E[\Gamma] \vdash (\mu x.t) : \forall x.U}
@@
 
-Lastly, the big operator sum is modelled by folding a function on a set, therefore the typing rule is as follows:
+Finally, the big operator sum is modeled by folding a function over a set, with the typing rule as follows:
 \[
@@
 
-% And lastly we have the typing rules for composition $x \circ y$. Similar to casual Dirac notation, the typing of composition depends on the types of operands.
-% \begin{gather*}
-%     \frac{E[\Gamma] \vdash x : \SType \qquad E[\Gamma] \vdash y : \KType(\sigma)}{E[\Gamma] \vdash x \circ y : \KType(\sigma)}
-%     \qquad
-%     \frac{E[\Gamma] \vdash x : \OType(\sigma, \tau) \qquad E[\Gamma] \vdash y : \KType(\tau)}{E[\Gamma] \vdash x \circ y : \KType(\sigma)}
-% \end{gather*}
+\begin{lemma}
+    The typing of expressions is both decidable and unique.
+\end{lemma}
 
-\begin{lemma}
-    The typing of expressions are decidable and unique.
-\end{lemma}
 \begin{proof}
-    The type can be calculated in the recursive manner. There is at most one typing rule for the same function symbol and argument types, which leads to the uniqueness of typing.
+    The type of an expression can be determined recursively. For any given function symbol and argument types, there is at most one typing rule, ensuring the uniqueness of typing.
 \end{proof}
@@
 
-% Semantics define the meanings of expressions, and the goal of the desirable algorithm is to decide whether
-% two expressions have the equivalent semantics. We can give the semantics by axiomatic equations and the denotational interpretations.
+The semantics of a language define the meaning of its expressions.
+In this context, the objective of our algorithm is to determine whether two expressions are semantically equivalent.
+We define the semantics in a denotational manner, mapping syntax to set-theoretic objects.
 
-To give meanings to the syntax, we assign the semantics in the denotational way, which is a mapping from syntax to set-theoretical objects.
+\subsubsection{Denotational Semantics}
+The denotational semantics interpret every expression as an object in linear algebra, according to a valuation mapping \( v \), which assigns values to the variables in the expression.
+The semantics of an expression \( e \) with a given valuation \( v \) is denoted as \( \sem{e}_v \).
+Two expressions \( e_1 \) and \( e_2 \) are considered equivalent if their semantics are equal for all valuations, i.e., \( \sem{e_1}_v = \sem{e_2}_v \) for all \( v \).
 
-\subsubsection{Denotational semantics} 
-The denotational semantics interpret every expression as an object in linear algebra, according to a valuation mapping $v$ from variables to their values. 
-The semantics of $e$ with valuation $v$ is written as $\sem{e}_v$, and two expressions $e_1$ and $e_2$ are equivalence if they have equal semantics in the mathematical sense for all valuations, written as $\sem{e_1} = \sem{e_2}$.
+The complete interpretation of terms and types is provided in~\Cref{sec: full denotational sem}.
+Variables typed with \( \Index \) are interpreted as finite sets, and the product of two indices \( \sem{\sigma_1 \times \sigma_2} \) is defined as the Cartesian product of the sets \( \sem{\sigma_1} \) and \( \sem{\sigma_2} \).
+More generally, each type is interpreted as a set.
+For example, the scalar type \( \sem{\SType} \) is interpreted as the set of complex numbers \( \mathbb{C} \), and the ket and bra types \( \sem{\KType(\sigma)} \) and \( \sem{\BType(\sigma)} \) are interpreted as the Hilbert space \( \mathcal{H}_{\sem{\sigma}} \) and its dual \( \mathcal{H}_{\sem{\sigma}}^* \), respectively.
+Terms are explained as the set elements.
+For example, the semantics of ket tensor product $\sem{K_1 \otimes K_2} \equiv \sem{K_1} \otimes \sem{K_2}$, is obtained by first calculating the semantics $\sem{K_1}$ and $\sem{K_2}$ as vectors, and then take the vector tensor product as result.
 
-The complete interpretation is listed in~\Cref{sec: full denotational sem}.
-Variables typed with $\Index$ are interpreted as finite sets, and index product $\sem{\sigma_1 \times \sigma_2} = \sem{\sigma_1} \times \sem{\sigma_2}$ are defined as Cartesian product of sets.
-Following on, every type is interpreted as a set.
-For example, the scalar type $\sem{\SType} = \mathbb{C}$ is interpreted as complex number set, the the ket and bra type $\sem{\KType(\sigma)} = \mathcal{H}_{\sem{\sigma}}$, $\sem{\KType(\sigma)} = \mathcal{H}_{\sem{\sigma}}^*$ are interpreted as the Hilbert space and dual space depending on $\sigma$.
-One special case is the delta symbol
-$\sem{\delta_{s, t}} = 
-\left \{  
-    \begin{array}{ll}
-        1, & \text{where } \sem{s} = \sem{t}, \\
-        0, & \text{where } \sem{s} \neq \sem{t}.
-    \end{array} 
-\right .
-$.
+% One special case is the delta function \( \delta_{s,t} \), which is interpreted as:
+% \[
+%     \sem{\delta_{s,t}} =
+%     \begin{cases}
+%         1, & \text{if } \sem{s} = \sem{t}, \\
+%         0, & \text{if } \sem{s} \neq \sem{t}.
+%     \end{cases}
+% \]d
 
-The idea behind the interpretation of types and terms is to characterize the typing relation with set theory inclusion. Therefore we have the following lemma.
+The idea behind the interpretation of types and terms is to formalize the typing relation using set-theoretic inclusion.
+Specifically, for a well-formed context \( E[\Gamma] \), term \( t \), and type \( T \), if \( E[\Gamma] \vdash t : T \), then for any valuation \( v \), the semantics of \( t \) must lie within the semantics of \( T \).
+
 \begin{lemma}
-    For all well-formed context $E[\Gamma]$, term $t$ and type $T$, if $E[\Gamma]\vdash t : T$, then for all valuations $v$, $\sem{t}_v \in \sem{T}_v$.
+    For any well-formed context \( E[\Gamma] \), term \( t \), and type \( T \), if \( E[\Gamma] \vdash t : T \), then for all valuations \( v \), \( \sem{t}_v \in \sem{T}_v \).
+\end{lemma}
+
+\begin{proof}
+    The proof follows directly by checking each case.
+\end{proof}
+
+This interpretation formalizes the standard understanding of Dirac notation and provides the foundation for the algorithm.
+However, computers cannot directly reason about equivalence through mathematical interpretations.
+Therefore, we proceed by defining a proof system that abstracts these concepts.
+
+\subsubsection{Axiomatic semantics}
+
+The proof system for equivalence is based on equational logic, together with axioms that describe the properties of Dirac notation.
+A full list of these axioms can be found in~\Cref{sec: full axioms}.
+The axioms cover fundamental aspects of linear spaces, as well as other structures like the tensor and inner products.
+For example, we have the absorption law for zero symbols:
+\(X \cdot \mathbf{0} = \mathbf{0},\)
+and the bilinearity of the dot product:
+\[
+(a.X) \cdot Y = a \cdot (X \cdot Y), \quad X \cdot (Y_1 + Y_2) = X \cdot Y_1 + X \cdot Y_2, 
+\]
+\[
+X \cdot (a.Y) = a \cdot (X \cdot Y), \quad (X_1 + X_2) \cdot Y = X_1 \cdot Y + X_2 \cdot Y.
+\]
+
+As mentioned in the introduction, a subset \( E \) of these axioms cannot be decided by term rewriting alone. The axioms in \( E \) include:
+\[
+\begin{aligned}
+    \text{AC-equivalence} &\quad \text{e.g.,} \quad X + Y = Y + X, \quad (X + Y) + Z = X + (Y + Z), \\
+    \alpha\text{-equivalence} &\quad \lambda x . A = \lambda y . A\{x/y\}, \\
+    \text{SUM-SWAP} &\quad \sum_{i \in s_1} \sum_{j \in s_2} A = \sum_{j \in s_2} \sum_{i \in s_1} A, \\
+    \text{scalar theories} &\quad \text{e.g.,} \quad a + 0 = a, \quad a \times (b + c) = a \times b + a \times c.
+\end{aligned}
+\]
+The scalar theories are treated separately as a module and are not considered in this work. In the implementation, we use the Mathematica kernel to decide scalar equivalences.
+These equational axioms provide an operable theory for the proof automation algorithm.
+Denotational semantics can be seen as one model for this theory, meaning that equivalences derived from the axioms always imply equivalence in the interpretations.
+\begin{lemma}
+    For all well-formed contexts \( E[\Gamma] \) and terms \( e_1, e_2 \), if \( \vdash e_1 = e_2 \), then \( \sem{e_1} = \sem{e_2} \).
 \end{lemma}
 \begin{proof}
-    Directly check all the cases.
+    This follows directly from checking all cases.
 \end{proof}
 
-This explanation formalizes the common explanation of Dirac notation, and best describes the target of the algorithm.
-However, computers cannot reason about equivalence through mathematical interpretations. We move on define the proof system as an abstraction.
+Thus, the axioms are sound with respect to denotational semantics.
+However, the reverse does not hold: there exist equivalences in denotational semantics that cannot be captured by these axioms.
+Nevertheless, our work focuses on solving practical examples, which are fully covered by the axioms in the experiments.
 
-\subsubsection{Axiomatic semantics} 
-The proof system for equivalence consists of equational logic as the framework and axioms describing the properties of Dirac notation.
-The full list is in~\Cref{sec: full axioms}.
-There are axioms for linear space, and other structures like tensor product and inner product.
-% From the operational view, each equation declares a valid rewriting operation, and two expressions are equiavlent if and only if they can be rewritten into the same form using the axioms.
-For example, we have the absorption law of zero symbols $X \cdot \mathbf{0} = \mathbf{0}$, and the bilinearity of dot product
+To conclude this section, we demonstrate the formalization of the symbols used in the motivating example.
+\begin{example}[Formalizing the Motivating Example]
+    \label{ex: formalizing motivating}
+    Definitions and assumptions in the environment \( E \) are formalized as follows:
     \begin{align*}
-    (a.X) \cdot Y = a.(X \cdot Y) \qquad X \cdot (Y_1 + Y_2) = X \cdot Y_1 + X \cdot Y_2 \\
-    X \cdot (a.Y) = a.(X \cdot Y) \qquad (X_1 + X_2) \cdot Y = X_1 \cdot Y + X_2 \cdot Y
+        & \text{TPO} && := \mu T_1. \mu T_2. \lambda O : \OType(T_1, T_2). \sum_{i \in \mathbf{U}(T_1)} \sum_{j \in \mathbf{U}(T_2)} \bra{i} O \ket{j} . \ket{j}\bra{i} \\
+        & &&\quad : \forall T_1. \forall T_2. \OType(T_1, T_2) \to \OType(T_2, T_1); \\
+        &\text{phi} &&:= \mu T. \sum_{i \in \mathbf{U}(T)} \sum_{j \in \mathbf{U}(T)} \ket{(i, j)} : \forall T.\KType(T \times T); \\
+        & T && : \Index; \\
+        & M && : \OType(T, T).
     \end{align*}
-
-As is mentioned in the introduction, a subset $E$ of the axioms is not decided by term rewriting. The set consists of:
-\begin{align*}
-    & \textrm{AC-equivalence} 
-    && \textrm{e.g.} \qquad X + Y = Y + X \qquad (X + Y) + Z = X + (Y + Z) \\
-    & \textrm{$\alpha$-equivalence}
-    && \lambda x . A = \lambda y . A\{x/y\} \\
-    & \textrm{SUM-SWAP}
-    && \sum_{i \in s_1} \sum_{j \in s_2} A = \sum_{j \in s_2} \sum_{i \in s_1} A \\
-    & \textrm{scalar theories} 
-    && \textrm{e.g.} \qquad a + 0 = a \qquad a \times (b + c) = a \times b + a \times c
-\end{align*}
-The equational theories for scalars are treated as a moduloe, and are not considered here. In the implementation, we use the Mathematica kernel to decide scalar equivalences.
-
-The equational axioms provide an operable theory for the proof automation algorithm,
-and the denotational semantics can be considered as one model for the theory.
-In this sense, the equivalence derived by axioms always imply the equivalence in interpretations.
-\begin{lemma}
-    For all well-formed context $E[\Gamma]$ and terms $e_1$, $e_2$, if $AX \vdash e_1 = e_2$, then $\sem{e_1} = \sem{e_2}$.
-\end{lemma}
-\begin{proof}
-    Direct check all cases.
-\end{proof}
-
-Therefore, the axioms are sound with respect to denotational semantics. 
-Unfortunately, the inverse does not hold: there are equivalences in denotational semantics that cannot be captured by these axioms. 
-However, our work focuses on solving the practical examples, which are covered by our axioms in the experiments.
-
-To conclude the design introduction, we demonstrate the formalization of symbols used in the motivating example.
-\begin{example}[formalizing the motivating example]
-    \label{ex: formalizing motivating}
-    Definitions and assumptions in the environment $E$:
-    \begin{align*}
-        & \textrm{TPO} && := \mu T_1. \mu T_2. \lambda O : \OType(T_1, T_2). \sum_{i \in \mathbf{U}(T_1)} \sum_{j \in \mathbf{U}(T_2)} \bra{i} O \ket{j} . \ket{j}\bra{i} \\
-        & && : \forall T_1. \forall T_2. \OType(T_1, T_2) \to \OType(T_2, T_1); \\
-        & \textrm{phi} && := \mu T. \sum_{i \in \mathbf{U}(T)} \sum_{j \in \mathbf{U}(T)} \ket{(i, j)} : \forall T.\KType(T \times T); \\
-        & \textrm{T} && : \Index; \\
-        & \textrm{M} && : \OType(T, T).
-    \end{align*}
-    The symbol TPO is the encoding of operator transpose, which is polymorphic on the Hilbert spaces by $T_1$ and $T_2$.
-    phi takes the index $T$ and defines the maximally entangled states. It sums over all basis in $T$, indicated by the universal set $\mathbf{U}(T)$.
-    With the assumption of index T and operator M, we can write the equivalence in the non-label version, formally as 
+    The symbol \( \text{TPO} \) represents the transpose of an operator, polymorphic on the Hilbert spaces \( T_1 \) and \( T_2 \).
+The symbol \( \text{phi} \) takes the index \( T \) and defines the maximally entangled states, summing over all basis elements in \( T \), as indicated by the universal set \( \mathbf{U}(T) \).
+    With the assumption of the index \( T \) and operator \( M \), we can express the equivalence in the non-labelled version as:
     \[
@@
 
-\section{Algorithm for Deciding Dirac notation equations}
+\section{Algorithm for Deciding Dirac Notation Equations}
 \label{sec: decide}
 
-We use the term rewriting technique to decide the equivalence of Dirac notation implied by its axiomatic semantics.
-Term rewriting decides equivalence by normalizing the two terms, and checking whether they have the same syntax.
-The normalization is powered by matching and substitution in universal algebra, using a set of rewriting rules.
-As is mentioned in the sematnics, the axioms in $E$ cannot be decided by term rewriting.
-DiracDec decides $E$ by examining all possible permutations in the rewriting result.
-In this work, the general idea is to carry through the normalization procedure using sorting algorithms. 
-The procedure of the normalization is displayed below. 
-The first three steps use term rewriting to work on the structure of Dirac notation, while the last three steps deals with the equational theory $E$.
+We utilize the term rewriting technique to decide the equivalence of Dirac notation, based on its axiomatic semantics.
+The process involves normalizing two terms and checking whether they have identical syntax.
+This normalization is achieved through matching and substitution in universal algebra, using a set of predefined rewriting rules.
+
+As mentioned in the semantics section, the axioms in \( E \) cannot be fully decided through term rewriting.
+DiracDec decides \( E \) by examining all possible permutations in the rewriting result.
+In this work, we introduce a more efficient approach by incorporating sorting algorithms into the normalization procedure.
+
+The normalization procedure consists of the following steps.
 \begin{enumerate}
-    \item \textbf{first rewritings}: expand definitions and simplify,
-    \item \textbf{variable expansion}: consider scalars expressions for completeness,
-    \item \textbf{second rewritings}: normalize modulo $E$,
-    \item \textbf{sorting without bound variables}: normalize AC-equivalence,
-    \item \textbf{swap successive summation}: normalize SUM-SWAP,
-    \item \textbf{de Bruijn normalization}: normalize $\alpha$-equivalence.
+    \item \textbf{First Rewritings}: Expand definitions and simplify expressions.
+    \item \textbf{Variable Expansion}: Consider scalar expressions to ensure completeness.
+    \item \textbf{Second Rewritings}: Normalize terms modulo \( E \).
+    \item \textbf{Sorting Without Bound Variables}: Normalize AC-equivalence.
+    \item \textbf{Swap Successive Summations}: Normalize SUM-SWAP equivalences.
+    \item \textbf{De Bruijn Normalization}: Normalize \( \alpha \)-equivalence.
 \end{enumerate}
 
-% \input{flowchart.tex}
+\subsection{Normalization Modulo \( E \) by Term Rewriting}
 
-\subsection{Normalization modulo $E$ by Term Rewriting}
-Term rewriting rules, written as $l \to r$, are used to normalize terms by recursively matching the subterms of the term with the left hand side $l$, and replace them with the corresponding right hand side $r$.
-The procedure terminates when no more rewritings can be made.
-The full list of rewriting rules are in~\Cref{sec: rewriting rules}, and we present some of them to illustrate the design idea.
+Term rewriting rules, represented as \( l \ \reduce\ r \), are applied recursively to normalize terms.
+In each step, subterms matching the left-hand side \( l \) of a rule are replaced with the corresponding right-hand side \( r \).
+The procedure terminates when no further rewritings can be made.
+A comprehensive list of rewriting rules can be found in~\Cref{sec: rewriting rules}. Below are some key examples to illustrate the design:
 
-Firstly, we use the following flattening rule to decide the associative with AC symbols, because we allow functions with indefinite arities.
+One of our optimizations is using functions with indefinite arities. 
+Therefore, we use a flattening rule to handle associativity with AC symbols:
 \[
-a_1 + \cdots + (b_1 + \cdots + b_m) + \cdots + a_n
-\ \reduce\ a_1 + \cdots + b_1 + \cdots + b_m + \cdots + a_n
+a_1 + \cdots + (b_1 + \cdots + b_m) + \cdots + a_n \ \reduce\ a_1 + \cdots + b_1 + \cdots + b_m + \cdots + a_n.
 \]
-Commutativity is left for the sorting algorithm later on. 
-Many rewriting rules are obtained by assigning directions to the equational axioms, for example:
+Commutativity is handled later in the sorting step. Many of the rewriting rules are directly derived from the equational axioms, such as:
 \begin{align*}
-    &\textrm{(R-DOT6)} && \bra{s} \cdot \ket{t} \ \reduce \delta_{s, t} \\
-    &\textrm{(R-DELTA0)} && \delta_{s, s} \ \reduce\ 1 \\
-    &\textrm{(R-MULK1)} && O : \OType(\sigma, \tau) \Rightarrow O \cdot \mathbf{0}_{\KType(\tau)} \ \reduce \mathbf{0}_{\KType(\sigma)} \\
-    &\textrm{(R-MULK11)} && (O_1 \otimes O_2) \cdot (K_1 \otimes K_2)\ \reduce\ (O_1 \cdot K_1) \otimes (O_2 \cdot K_2)
+    & \textrm{(R-DOT6)} && \bra{s} \cdot \ket{t} \ \reduce\  \delta_{s, t}, \\
+    & \textrm{(R-DELTA0)} && \delta_{s, s} \ \reduce\  1, \\
+    & \textrm{(R-MULK1)} && O : \OType(\sigma, \tau) \Rightarrow O \cdot \mathbf{0}_{\KType(\tau)} \ \reduce\  \mathbf{0}_{\KType(\sigma)}, \\
+    & \textrm{(R-MULK11)} && (O_1 \otimes O_2) \cdot (K_1 \otimes K_2) \ \reduce\  (O_1 \cdot K_1) \otimes (O_2 \cdot K_2).
 \end{align*}
-Some of them are obvious. The (R-DOT6) says that inner product of two basis will be evaluated to a delta expression,
-and (R-DELTA0) further transforms delta on identical basis to scalar $1$.
-(R-MULK1) corresponds to the axiom that multiplication on zero vectors results in zero. It has a condition on typing, which is calculated during rewriting.
-Some rules require the intuition. For example, (R-MULK11) shows that we prefer tensor product than multiplication.
+Some of these directions are obvious.
+For example, (R-DOT6) states that the inner product of two basis vectors is reduced to a delta expression, and (R-DELTA0) transforms a delta function on identical basis to a scalar \( 1 \).
+The rule (R-MULK1) reflects the axiom that multiplying a zero vector results in zero.
+This rule is conditional on typing, which is checked during rewriting.
+Some rules, like (R-MULK11), require a deeper understanding, such as the preference for tensor products over multiplication.
 
-As a reference, the term rewriting system for DiracDec is proved complete for the axioms, except the sum symbol.
-The result is obtained by checking confluence and termination of the system.
-Our rewriting rules are translations from DiracDec into the typed and abstract language, therefore completeness for corresponding symbols is also expected to hold.
+As a reference, the term rewriting system in DiracDec has been proven complete for all axioms, except for the sum symbol.
+The completeness result is derived from checking the confluence and termination of the system.
+Our rewriting rules are translations from DiracDec into the typed and abstract language, ensuring that the corresponding symbols in our system are also complete.
 
-%%%%%%%%%%% It's better to avoid mentioning completion rules.
-% Other part of rules assign the rewriting diraction to equations according to the property of symbols and intuitive observations of their relations.
-% When different rules rewrites the same term to different forms, rules for confluence purpose are added.
-% \begin{align*}
-%     %
-%     & \text{(R-SCR1)} 
-%     && a.(b.X) \ \reduce\ (a \times b).X \\
-%     %
-%     & \text{(R-DOT10)}
-%     && (B \cdot O) \cdot K \ \reduce\ B \cdot (O \cdot K), \\
-%     %
-%     & \text{(R-DOT11)}
-%     && \bra{(s, t)} \cdot ((O_1 \otimes O_2) \cdot K) \ \reduce\ ((\bra{s} \cdot O_1) \otimes (\bra{t} \cdot O_2)) \cdot K, \\
-%     %
-%     & \text{(R-MULB10)}
-%     && \bra{(s, t)} \cdot (O_1 \otimes O_2)\ \reduce\ (\bra{s} \cdot O_1) \otimes (\bra{t} \cdot O_2),
-% \end{align*}
-% The (R-SCR1) rule follows the idea to reduce linear algebra scaling into scalar multiplications,
-% and (R-MULB10) transform inner product into tensor product. 
-% As an example of the completetion rule, here the normalization of term $(\bra{(s, t)} \cdot (O_1 \otimes O_2)) \cdot K$ have two rewriting paths: (a).
-apply (R-MULB10) and get $((\bra{s} \cdot O_1) \otimes (\bra{t} \cdot O_2)) \cdot K$, or
-% (b).
-first apply (R-DOT10) and sort the term into $\bra{(s, t)} \cdot ((O_1 \otimes O_2) \cdot K)$, and then apply (R-DOT11) to get the same result.
-Here (R-DOT11) is for completeness of cases similar to this example.
+We also have additional rules that handle summations. For example:
+\begin{align*}
+    & \quad i \text{ free in } t \Rightarrow \sum_{i \in \mathbf{U}(\sigma)} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (\delta_{i, t}.A) \ \reduce\  \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} A\{i/t\}, \\
+    & \quad \left( \sum_{i \in M} B \right) \cdot K \ \reduce\  \sum_{i \in M} (B \cdot K).
+\end{align*}
+The first rule eliminates delta expressions in summations, while the second rule pushes summations outside of inner products.
+While there is no guarantee of completeness for these rules, they work effectively in practice.
 
-We have other rules dealing with big operator sum. For example, 
-\begin{align*}
-    & \textrm{(R-SUM-ELIM2)}
-    && i \text{ free in } t \Rightarrow \sum_{i \in \mathbf{U}(\sigma)} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (\delta_{i, t}.A) \\
-    & && \reduce\ \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} A\{i/t\} \\
-    & \textrm{(R-SUM-PUSH5)}
-    && (\sum_{i \in M} B)\cdot K \ \reduce\ \sum_{i \in M}(B \cdot K)
-\end{align*}
-The (R-SUM-ELIM2) rule will try to eliminate the delta expression in summations.
-And the (R-SUM-PUSH5) rule will lift sum to the outside of inner product.
-There are no completeness guarantee, but the rules work well in practice.
-
-One technique revealed in the DiracDec work is the expansion of variables, critical for proofs with sum:
-\begin{gather*}
-    \frac{E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash K \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)}(\bra{i}\cdot K).\ket{i}}
-    \qquad \qquad
-    \frac{E[\Gamma] \vdash B : \BType(\sigma)}{E[\Gamma] \vdash B \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)}(B \cdot \ket{i}).\bra{i}} \\
-    \\
-    \frac{E[\Gamma] \vdash O : \OType(\sigma, \tau)}{E[\Gamma] \vdash O \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)} \sum_{j \in \mathbf{U}(\tau)}(\bra{i} \cdot O \cdot \ket{j}).(\ket{i} \cdot \bra{j})}
-\end{gather*}
-These rules transform variables to the symbolic summation of their decomposition on the basis.
+\subsubsection{Variable Expansion}
+One important technique, revealed in the DiracDec work, is the expansion of variables, which is critical for proofs involving summation. For example:
+\[
+\frac{E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash K \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)}(\bra{i} \cdot K).\ket{i}} \quad \quad
+\frac{E[\Gamma] \vdash B : \BType(\sigma)}{E[\Gamma] \vdash B \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)}(B \cdot \ket{i}).\bra{i}},
+\]
+\[
+\frac{E[\Gamma] \vdash O : \OType(\sigma, \tau)}{E[\Gamma] \vdash O \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)} \sum_{j \in \mathbf{U}(\tau)}(\bra{i} \cdot O \cdot \ket{j}).(\ket{i} \cdot \bra{j})}.
+\]
+These rules transform variables into their symbolic summations based on their decomposition over the basis. 
 The rules are not terminating, therefore is applied recrusively once in the second step called \textit{variable expansion}.
-On the other hand, we discovered that doing the expansion on all variables only once is already sufficient.
+Nevertheless, we have found that applying the expansion only once for all variables is sufficient for normalization.
 \begin{lemma}
-    Let $\textrm{expand}(e)$ indicate the result of expanding all variables in $e$ once.
-    For all well-typed term $e$ in $E[\Gamma]$, $\textrm{expand}(\textrm{expand}(e))$ and $\textrm{expand}(e)$ have the same normal form.
+    Let \( \textrm{expand}(e) \) denote the result of expanding all variables in \( e \) once.
+For all well-typed terms \( e \) in \( E[\Gamma] \), \( \textrm{expand}(\textrm{expand}(e)) \) and \( \textrm{expand}(e) \) have the same normal form.
 \end{lemma}
 \begin{proof}
-    If we expand a ket variable twice, the structure will be reduced to
+    Expanding a ket variable twice, for example, results in the following transformation:
     \[
-        \sum_{i \in \mathbf{U}(\sigma)}(\bra{i}\cdot \sum_{j \in \mathbf{U}(\sigma)}(\bra{j}\cdot K).\ket{j}).\ket{i}
-        \ \reduce\ 
-        \sum_{i \in \mathbf{U}(\sigma)}\sum_{j \in \mathbf{U}(\sigma)}(\bra{j}\cdot K \times \braket{i|j}).\ket{i},
+    \sum_{i \in \mathbf{U}(\sigma)} (\bra{i} \cdot \sum_{j \in \mathbf{U}(\sigma)} (\bra{j} \cdot K) \cdot \ket{j}) \cdot \ket{i} \ \reduce\  \sum_{i \in \mathbf{U}(\sigma)} \sum_{j \in \mathbf{U}(\sigma)} (\bra{j} \cdot K \cdot \braket{i|j}) \cdot \ket{i},
     \]
-    the elimination rule of delta symbol returns the term to the one-expansion result. Similar for bra and operator.
+    where the delta symbol elimination rule returns the term to its original form. The same holds for bra and operator terms.
 \end{proof}
 
-\section{Deciding Equational Theory $E$}
+\subsection{Deciding Equational Theory \( E \)}
 
-In the Mathematica implementation of DiracDec, the equational theory $E$ is decided by a unification, which tries to find a substitution of summation bound variables that makes the two expressions syntactically equivalent.
-The unification iterates through all permutations of AC symbol arguments, and the complexity is factorial in the maximum number of AC symbol arguments.
+In the Mathematica implementation of DiracDec, the equational theory \( E \) is decided using unification, which attempts to find a substitution for the summation bound variables that makes two expressions syntactically equivalent.
+This unification process iterates through all permutations of AC symbol arguments, and its complexity is factorial in the maximum number of AC symbol arguments.
 
-As the first improvement, we check the $\alpha$-equivalence by de Bruijn index~\cite{deBruijn1972lambda}, where references to bound variable names are replaced by the distance of the lambda abstraction to the variable.
-For example, the nominal lambda abstraction $\lambda x.
-x$ is transformed into $\lambda .
-0$, while $\lambda x.
+As the first improvement, we check \( \alpha \)-equivalence using the de Bruijn index~\cite{deBruijn1972lambda}, which replaces references to bound variable names with the distance from the lambda abstraction to the variable.
+For instance, the nominal lambda abstraction \( \lambda x.
+x \) is transformed into \( \lambda .
+0 \), while \( \lambda x.
 \lambda y.
-(x\ (y\ x))$ will is transformed into $\lambda.\lambda.
-(1\ (0\ 1))$.
-Transformation into de Bruijn index is at the last step of the normalization to check the equivalence of terms with different bound variable names.
+(x\ (y\ x)) \) is transformed into \( \lambda.\lambda.
+(1\ (0\ 1)) \).
+This transformation into the de Bruijn index is performed as the final step in the normalization process to check equivalence between terms with different bound variable names.
 
-The remaining axioms of AC-equivalence and SUM-SWAP claims equivalence under permutations.
-A standard approach to decide is normalization by sorting in a given order.
-For example, given the dictionary order $a < b < c$, the term $b + c + a$ (and any other AC equivalent ones) will be normalized into $a + b + c$.
-For our setting, there are two difficulties intertwined together: how to assign such an order to all terms in our language, and how to sort for the two axioms simultaneously.
-As a typical example, the two terms 
+The remaining axioms, such as AC-equivalence and SUM-SWAP, assert equivalence under permutations.
+A standard approach for deciding such equivalences is to normalize terms by sorting in a predefined order.
+For example, given the dictionary order \( a < b < c \), the term \( b + c + a \) (and any other AC-equivalent term) is normalized into \( a + b + c \).
+However, in our setting, two intertwined difficulties arise: how to assign an order to all terms in the language, and how to simultaneously sort for both axioms.
+
+Consider the following two equivalent terms:
 \[
-    \sum_{i \in s_1} \sum_{j \in s_2} \bra{i}A\ket{j} \times \bra{j}B\ket{i}
+\sum_{i \in s_1} \sum_{j \in s_2} \bra{i} A \ket{j} \times \bra{j} B \ket{i}
 = 
-    \sum_{i \in s_2} \sum_{j \in s_1} \bra{i}B\ket{j} \times \bra{j}A\ket{i}
+\sum_{i \in s_2} \sum_{j \in s_1} \bra{i} B \ket{j} \times \bra{j} A \ket{i}
 \]
-are equivalent. However, directly sorting elemnts of scalar multiplication in the lexical order will not lead to the same form.
+While these two terms are equivalent, directly sorting the elements of scalar multiplication using lexical order does not yield the same form.
 
-We propose an algorithm to solve the problem by sorting without bound variables.
-The observation is that in a successive sum expression $\sum_{i \in s_1}\cdots\sum_{j \in s_n}A$, the name and order of bound variables $i, \cdots, j$ can be permuted freely.
-Therefore, all bound variables should be treated uniformly during sorting, and the order of sum can then be decided according to the position of their bound variables.
+To address this issue, we propose an algorithm that sorts terms without considering bound variables.
+The key observation is that in a successive sum expression \( \sum_{i \in s_1} \cdots \sum_{j \in s_n} A \), the names and order of the bound variables \( i, \dots, j \) can be freely permuted.
+Therefore, all bound variables should be treated uniformly during sorting, and the order of summation can then be determined based on the position of the bound variables.
 
-In the above example, we ignore the bound variables first, and sort the sum body into $\bra{\bullet}A\ket{\bullet} \times \bra{\bullet}B\ket{\bullet}$.
-Then we swap the sum, so that the bound variable at first $\bullet$ position appears at the outmost position.
-The results will have the same de Bruijn normal form, namely $\sum_{s_1} \sum_{s_2} \bra{\$1} A \ket{\$0} \times \bra{\$0} B \ket{\$1}$.
+In the example above, we first ignore the bound variables and sort the sum body into \( \bra{\bullet} A \ket{\bullet} \times \bra{\bullet} B \ket{\bullet} \).
+Then, we swap the summations such that the bound variable at the first \( \bullet \) position appears at the outermost position.
+The results will have the same de Bruijn normal form, namely \( \sum_{s_1} \sum_{s_2} \bra{\$1} A \ket{\$0} \times \bra{\$0} B \ket{\$1} \).
 
-To describe the algorithm in the following, we introduce two notations: for a term $e = f(a_1, a_2, \cdots, a_n)$, $\textrm{head}(e)$ indicates the function symbol $f$,
-and $\textrm{arg}(e, i)$ indicates the $i$-th argument $a_i$. 
-% Variables and constants are considered as functions with zero arguments.
-\begin{definition}[order without bound variables]
-    Assume we have a total order of all symbols.
-    Let $\mathcal{B}$ be the set of bound variables.
-    For simplicity, we assume all bound variables are unique.
-    The relation $e_1 =_\mathcal{B} e_2$ holds when
+To describe the algorithm in the following, we introduce two key notations.
+For a term \( e = f(a_1, a_2, \dots, a_n) \), \( \textrm{head}(e) \) denotes the function symbol \( f \), while \( \textrm{arg}(e, i) \) refers to the \( i \)-th argument \( a_i \) of the term.
+In this context, variables and constants are treated as functions with zero arguments.
+
+\begin{definition}[Order Without Bound Variables]
+Let \( \mathcal{B} \) represent the set of bound variables, with the assumption that all bound variables are unique.
+We also assume that a total order exists over all symbols.
+The relation \( e_1 =_\mathcal{B} e_2 \) holds if:
 \begin{itemize}
-        \item $head(e_1) = head(e_2)$ and $arg(e_1, i) =_\mathcal{B} arg(e_2, i)$ for all $i$, or
-        \item $e_1 \in \mathcal{B}$ and $e_2 \in \mathcal{B}$.
+    \item \( \textrm{head}(e_1) = \textrm{head}(e_2) \), and for all \( i \), \( \textrm{arg}(e_1, i) =_\mathcal{B} \textrm{arg}(e_2, i) \), or
+    \item \( e_1 \in \mathcal{B} \) and \(e_2 \in \mathcal{B}\).
 \end{itemize}
 
-    The relation $e_1 <_\mathcal{B} e_2$ between the two terms holds when:
+The relation \( e_1 <_\mathcal{B} e_2 \) holds between two terms if:
 \begin{itemize}
@@
 \end{definition}
-It can be checked that $e_1 =_\mathcal{B} e_2$ if and only if neither $e_1 <_\mathcal{B} e_2$ or $e_2 <_\mathcal{B} e_1$ holds.
-The idea is to compare the function symbols in the top down order, and omit bound variable occurance. 
-With the order, the sort function normalize the term with respect to AC equivalence.
-\begin{definition}[sort transformation]
+It can be shown that \( e_1 =_\mathcal{B} e_2 \) if and only if neither \( e_1 <_\mathcal{B} e_2 \) nor \( e_2 <_\mathcal{B} e_1 \) holds.
+The purpose of this ordering is to compare function symbols in a top-down manner while ignoring bound variables.
+The sorted order enables normalization of terms in terms of AC equivalence.
+\begin{definition}[Sort Transformation]
     For a term $e$ with bound variable set $\mathcal{B}$,
@@
 
-The order of bound variables can be decided according to the result of the sort transformation, and then successive summations can be normalized using the order.
-\begin{definition}[swap transformation]
-    For a term $e$ with sorting result sort$(e)$, we can obtain a total order of all bound variables appearing in the term $e$.
-Specifically, they are ordered by their first appearances, except in function definitions $\lambda x$ and $\mu x$.
-    The swap transformation then arrange successive summation according to this order.
+After sorting, the next step is the \textit{swap transformation}, which arranges successive summations based on the order of bound variables.
+\begin{definition}[Swap Transformation]
+For a term \( e \) with a sorting result \( \textrm{sort}(e) \), the swap transformation proceeds by ordering all bound variables according to their first appearances, except in function definitions \( \lambda x \) and \( \mu x \).
+The swap transformation then reorders the successive summations accordingly.
 \end{definition}
 
-The algorithm succeeds in all equivalences of our benchmark, and we formalize its completeness as a conjecture.
-\begin{conjecture}[completeness of transformation]
-    For any two terms $e_1$ and $e_2$ that is equivalent under AC equivalence, SUM-SWAP and $\alpha$-equivalence, they will have the same form after sort, swap and de Bruijn transformations.
+The algorithm that applies the sort and swap transformations, followed by de Bruijn normalization, successfully handles all equivalences in our benchmark.
+We formalize the completeness of this transformation in the following conjecture:
+
+\begin{conjecture}[Completeness of Transformation]
+    For any two terms \( e_1 \) and \( e_2 \) that are equivalent under AC equivalence, SUM-SWAP, and \( \alpha \)-equivalence, they will have the same form after sort, swap, and de Bruijn transformations.
 \end{conjecture}
@@
 
-Lastly, we can prove that equivalence by this normalization procedure is sound with respect to the semantics. 
-\begin{theorem}[soundness]
-    For any well-formed context $E[\Gamma]$ and well-typed expressions $e_1$ and $e_2$, if $e_1\downarrow = e_2\downarrow$, then $\sem{e_1} = \sem{e_2}$.
+Lastly, we can prove that the equivalence established by this normalization procedure is sound with respect to the semantics.
+
+\begin{theorem}[Soundness]
+    For any well-formed context \( E[\Gamma] \) and well-typed expressions \( e_1 \) and \( e_2 \), if  $e_1$ and $e_2$ have the same normal form, then \( \sem{e_1} = \sem{e_2} \).
 \end{theorem}
+
 \begin{proof}
-    The soundness of term rewriting holds because each rewriting rule preserves the equivalence.
-The operations in sort and swap transformation follows AC-equivalence and SUM-SWAP axioms.
-The de Bruijn normal form is sound for $\alpha$-equivalence.
+    The soundness of the term rewriting procedure follows from the fact that each rewriting rule preserves equivalence.
+Furthermore, the operations in the sort and swap transformations respect the AC-equivalence and SUM-SWAP axioms.
+Finally, the de Bruijn normalization ensures soundness for \( \alpha \)-equivalence.
 \end{proof}

### 2025-01-27 main.tex
2025/1/27

@@
         & |\ \MULK(e_1, e_2) \mid \MULB(e_1, e_2) \mid \OUTER(e_1, e_2) \mid \MULO(e_1, e_2) \\
-        & |\ \mathbf{U}(e) \mid e_1 \star e_2 \mid \sum_{e_1} e_2.
+        & |\ \mathbf{U}(\sigma) \mid e_1 \star e_2 \mid \sum_{e_1} e_2.
     \end{align*}
@@
 \( e_1.e_2 \) represents scaling the term \( e_2 \) by scalar \( e_1 \).
-\( \mathbf{U}(e) \) denotes the universal set with index \( e \).
+\( \mathbf{U}(\sigma) \) denotes the universal set with index \( \sigma \).
 \( e_1 \star e_2 \) represents the Cartesian product of \( e_1 \) and \( e_2 \).
@@
 
-We also built an example benchmark for labelled Dirac notation.
-\yx{We need examples.}
+We also built an example benchmark for labelled Dirac notation, see~\Cref{sec: examples for labelled}.
 

### 2025-01-27 01intro_preliminary.tex
splited

--- 2025-01-25 main.tex
@@
-% This is samplepaper.tex, a sample chapter demonstrating the
-% LLNCS macro package for Springer Computer Science proceedings;
-% Version 2.21 of 2022/01/12
-%
-\documentclass[runningheads]{llncs}
-
-% \usepackage[utf8]{inputenc}
-% \usepackage[margin=0.8in]{geometry}
-
-\usepackage{color}
-\usepackage{amsmath}
-%\usepackage{amssymb}
-\usepackage{graphicx}
-% \usepackage{amsthm}
-\usepackage{stmaryrd}
-\usepackage[all]{xy}
-\usepackage{multirow}
-\usepackage{paralist}
-\usepackage{hhline}
-\usepackage{bm}
-\usepackage{longtable}
-\usepackage{makecell}
-\usepackage{mdframed}
-
-\usepackage{wasysym}
-\usepackage{extarrows}
-\usepackage{tikz}
-
-% add new commands for comments here
-\newcommand{\yx}[1]{\textit{\color{blue}[YX] : #1}}
-\newcommand{\lz}[1]{\textit{\color{red}[LZ] : #1}}
-
-\newcommand{\modify}[1]{{\color{red}#1}}
-
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{stmaryrd}
-\usepackage{hyperref}
-
-\usepackage{braket}
-\usepackage{cleveref}
-
-\input{lstlisting}
-
-\usepackage{tikz}
-\usetikzlibrary{shapes.geometric, arrows}
-\usetikzlibrary{trees}
-\tikzstyle{arrow} = [thick,->,>=stealth]
-
-\newenvironment{ruletable}[1]
-{
-    \begin{longtable}{cl}
-    \caption{#1}\\
-    \hline
-    \textbf{Rule} & \textbf{Description} \\
-    \hline
-    \endfirsthead
-
-    \hline
-    \textbf{Rule} & \textbf{Description} \\
-    \hline
-    \endhead
-
-    % \hline
-    % \multicolumn{2}{r}{\textit{Continued on the next page}} \\
-    \hline
-    \endfoot
-
-    \hline
-    \endlastfoot
-}
-{
-    \end{longtable}
-}
-
-% define C++ logo
-\newcommand{\CC}{C\nolinebreak\hspace{-.05em}\raisebox{.4ex}{\tiny\bf +}\nolinebreak\hspace{-.10em}\raisebox{.4ex}{\tiny\bf +}}
-\def\CC{{C\nolinebreak[4]\hspace{-.05em}\raisebox{.4ex}{\tiny\bf ++}}}
-
-%
-\usepackage[T1]{fontenc}
-% T1 fonts will be used to generate the final print and online PDFs,
-% so please use T1 fonts in your manuscript whenever possible.
-% Other font encondings may result in incorrect characters.
-%
-\usepackage{graphicx}
-% Used for displaying a sample figure. If possible, figure files should
-% be included in EPS format.
-%
-% If you use the hyperref package, please uncomment the following two lines
-% to display URLs in blue roman font according to Springer's eBook style:
-\usepackage{color}
-\renewcommand\UrlFont{\color{blue}\rmfamily}
-\urlstyle{rm}
-%
-\begin{document}
-%
-\title{Dirace: Practical Proof Automation of Dirac Notation Equations}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% %
-% %\titlerunning{Abbreviated paper title}
-% % If the paper title is too long for the running head, you can set
-% % an abbreviated paper title here
-% %
-% \author{First Author\inst{1}\orcidID{0000-1111-2222-3333} \and
-% Second Author\inst{2,3}\orcidID{1111-2222-3333-4444} \and
-% Third Author\inst{3}\orcidID{2222--3333-4444-5555}}
-% %
-% \authorrunning{F. Author et al.}
-% % First names are abbreviated in the running head.
-% % If there are more than two authors, 'et al.' is used.
-% %
-% \institute{Princeton University, Princeton NJ 08544, USA \and
-% Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
-% \email{lncs@springer.com}\\
-% \url{http://www.springer.com/gp/computer-science/lncs} \and
-% ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
-% \email{\{abc,lncs\}@uni-heidelberg.de}}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%
-\maketitle              % typeset the header of the contribution
-%
-\begin{abstract}
-Dirac notation is a fundamental language in quantum computation and quantum information.
-Recently, the term rewriting system DiracDec~\cite{diracdec} was introduced to automate equational reasoning with Dirac notation, a critical yet time-intensive task.
-Building upon DiracDec, this work aims to develop a solver optimized for practical applications.
-Enhancements include an improved typing system, a simplified language and rewriting system, more efficient algorithms, and added support for labeled Dirac notation.
-Our solver, Dirace, is implemented in \CC\ with a Mathematica backend, demonstrating significant improvements in decision-making power and time efficiency.
-
-% \keywords{First keyword  \and Second keyword \and Another keyword.}
-\end{abstract}
-%
-%
-%
-
-%%%%%%%%%%%%%%%%
-\newcommand*{\sem}[1]{{\llbracket #1 \rrbracket}}
-\newcommand{\DiracDec}{\textsf{DiracDec}}
-
-\newcommand{\reduce}{\triangleright}
-
-\newcommand{\Sort}{\mathsf{Sort}}
-\newcommand{\WF}{\mathcal{WF}}
-
-\newcommand{\Index}{\mathsf{Index}}
-\newcommand{\Type}{\mathsf{Type}}
-\newcommand{\Basis}{\mathsf{Basis}}
-
-\newcommand{\SType}{\mathcal{S}}
-\newcommand{\KType}{\mathcal{K}}
-\newcommand{\BType}{\mathcal{B}}
-\newcommand{\OType}{\mathcal{O}}
-\newcommand{\SET}{\mathsf{Set}}
-
-\newcommand{\ZEROK}{\mathbf{0}_\mathcal{K}}
-\newcommand{\ZEROB}{\mathbf{0}_\mathcal{B}}
-\newcommand{\ZEROO}{\mathbf{0}_\mathcal{O}}
-
-\newcommand{\PAIR}{\mathsf{PAIR}}
-
-\newcommand{\ZERO}{\mathsf{0}}
-\newcommand{\ONE}{\mathsf{1}}
-\newcommand{\ADDS}{\mathsf{ADDS}}
-\newcommand{\ADD}{\mathsf{ADD}}
-\newcommand{\MULS}{\mathsf{MULS}}
-\newcommand{\MUL}{\mathsf{MUL}}
-\newcommand{\CONJ}{\mathsf{CONJ}}
-\newcommand{\CJG}{\mathsf{CJG}}
-\newcommand{\ADJ}{\mathsf{ADJ}}
-\newcommand{\DELTA}{\mathsf{DELTA}}
-\newcommand{\DOT}{\mathsf{DOT}}
-\newcommand{\SCR}{\mathsf{SCR}}
-\newcommand{\TSR}{\mathsf{TSR}}
-\newcommand{\KET}{\mathsf{KET}}
-\newcommand{\BRA}{\mathsf{BRA}}
-\newcommand{\ONEO}{\mathbf{1}_\mathcal{O}}
-\newcommand{\OUTER}{\mathsf{OUTER}}
-\newcommand{\MULK}{\mathsf{MULK}}
-\newcommand{\MULB}{\mathsf{MULB}}
-\newcommand{\MULO}{\mathsf{MULO}}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \section{Introduction}
@@
     \item Support for constant register labels, and reducing the equivalence decision of labelled Dirac notation to the unlabelled case.
-    \item A more user-friendly \CC\ solver, Dirace, featuring an abstract language and typing system. We also support the definition of symbols (e.g., transpose and trace) using function syntax.
+    \item A more user-friendly \CC\ solver, D-Hammer, featuring an abstract language and typing system. We also support the definition of symbols (e.g., transpose and trace) using function syntax.
 \end{itemize}
 
-We evaluated Dirace against the DiracDec benchmark and new examples involving labelled Dirac notation.
+We evaluated D-Hammer against the DiracDec benchmark and new examples involving labelled Dirac notation.
 The results show significant improvements in both decidability and efficiency compared to DiracDec.
 
-% Dirace successfully decides all the examples that are expressable in its language, including those failed by DiracDec because of complexity or insufficient decision power.
+% D-Hammer successfully decides all the examples that are expressable in its language, including those failed by DiracDec because of complexity or insufficient decision power.
 
@@
 This relation formalizes the intuitive concept of equivalence in algebra.
-
-\section{Language, Typing and Semantics}
-The first step is to formalize the language for Dirac notation.
-In DiracDec, the language has a concrete design, where the same syntax for different types corresponds to distinct symbols.
-Our goal is to transition from this concrete design to a more abstract one, which aligns more closely with the conventional Dirac notation we use and simplifies the term rewriting system.
-
-To achieve this, we organize our language into three layers: the index, the type, and the term.
-Terms represent concrete instances such as kets, bras, and operators, which will be typed and checked.
-The index represents classical data types and appears in type expressions to differentiate between various Hilbert spaces and sets.
-
-\begin{definition}[Index Syntax]
-    The syntax for type indices is:
-    \begin{align*}
-        \sigma ::=\ & x \mid \sigma_1 \times \sigma_2.
-    \end{align*}
-\end{definition}
-Here, \( x \) is a variable, and \( \sigma_1 \times \sigma_2 \) represents the product type for tensor product spaces or Cartesian product sets.
-
-\begin{definition}[Type Syntax]
-    The syntax for Dirac notation types is:
-    \begin{align*}
-        T ::=\ & \Basis(\sigma) \mid \SType \mid \KType(\sigma) \mid \BType(\sigma) \mid \OType(\sigma_1, \sigma_2) \mid T_1 \to T_2 \mid \forall x.T \mid \SET(\sigma).
-    \end{align*}
-\end{definition}
-\( \Basis(\sigma) \) denotes the type for basis elements in the index \( \sigma \).
-\( \SType \) represents scalars, while \( \KType(\sigma) \) and \( \BType(\sigma) \) refer to ket and bra types in the Hilbert space \( \sigma \), respectively.
-\( \OType(\sigma_1, \sigma_2) \) represents linear operators with \( \sigma_2 \) as the domain and \( \sigma_1 \) as the codomain.
-\( \SET(\sigma) \) refers to the type of subsets of \( \sigma \), used to denote the values of bound variables in summations.
-The remaining two constructs define function types: \( T_1 \to T_2 \) represents normal functions that take a \( T_1 \)-type argument and return a \( T_2 \)-type term, while \( \forall x.
-T \) represents index functions that take an index argument \( x \) and produce a \( T \)-type term, which may depend on \( x \).
-Index functions are essential for defining polymorphic transformations across Hilbert spaces.
-
-\begin{definition}[Term Syntax]
-    The syntax for Dirac notation terms is:
-    \begin{align*}
-        e ::=\ & x \mid \lambda x : T.e \mid \mu x.e \mid e_1\ e_2 \mid (e_1, e_2) \\
-        & |\ 0 \mid 1 \mid \ADDS(e_1, \cdots, e_n) \mid e_1 \times \cdots \times e_n \mid e^* \mid \delta_{e_1, e_2} \mid \DOT(e_1, e_2) \\
-        & |\ \ZEROK(\sigma) \mid \ZEROB(\sigma) \mid \ZEROO(\sigma_1, \sigma_2) \mid \ONEO(\sigma) \\
-        & |\ \ket{e} \mid \bra{t} \mid e^\dagger \mid e_1.e_2 \mid \ADD(e_1, \cdots, e_n) \mid e_1 \otimes e_2 \\
-        & |\ \MULK(e_1, e_2) \mid \MULB(e_1, e_2) \mid \OUTER(e_1, e_2) \mid \MULO(e_1, e_2) \\
-        & |\ \mathbf{U}(e) \mid e_1 \star e_2 \mid \sum_{e_1} e_2.
-    \end{align*}
-\end{definition}
-The terms above are explained in order:
-\( \lambda x : T.e \) represents the abstraction for normal functions, and \( \mu x.e \) represents the abstraction for index functions.
-\( e_1\ e_2 \) denotes function application.
-\( (e_1, e_2) \) is the basis pair for product types.
-The constants \( 0 \), \( 1 \), \( \ADDS \), \( e_1 \times \cdots \times e_n \), and \( e^* \) are symbols for scalars.
-The next line includes constant symbols for kets, bras, and operators.
-\( \ket{e} \) is a ket, \( \bra{t} \) is a bra, and \( e^\dagger \) denotes the conjugate transpose of \( e \).
-\( e_1.e_2 \) represents scaling the term \( e_2 \) by scalar \( e_1 \).
-\( \mathbf{U}(e) \) denotes the universal set with index \( e \).
-\( e_1 \star e_2 \) represents the Cartesian product of \( e_1 \) and \( e_2 \).
-\( \sum_{e_1} e_2 \) is the big operator sum, modeled by folding the function \( e_2 \) over the value set \( e_1 \).
-Typically, the sum's body is given by an abstraction.
-For convenience, we also use the notation \( \sum_{x \in s} X \) to represent \( \sum_{s} \lambda x : T .
-X \).
-Additionally, \( \ADDS \) and \( \ADD \) are distinct AC symbols: \( \ADDS \) is used for scalar addition and \( \ADD \) for linear algebra addition.
-There are five kinds of linear algebraic multiplications among kets, bras, and operators.
-These are encoded using five different symbols: \( \DOT \), \( \MULK \), \( \MULB \), \( \OUTER \), and \( \MULO \).
-
-Compared to DiracDec, the symbols for addition, adjoint, scaling, and tensor products have been consolidated.
-The multiplications are still kept separate due to their differing properties.
-These operations are denoted as \( B \cdot K \), \( K_1 \cdot K_2 \), \( B_1 \cdot B_2 \), \( K \cdot B \), and \( O_1 \cdot O_2 \), respectively.
-
-\subsection{Typing System}
-
-The typing system is responsible for classifying terms within a proof system, using a context that specifies the types of variables and definitions.
-The context is divided into two components: the environment \( E \), which includes definitions and assumptions, and the context \( \Gamma \), which tracks bound variables.
-Both \( E \) and \( \Gamma \) consist of sequences of assumptions \( x : T \) or definitions \( x := t : T \).
-\begin{definition}[Environment and Context]
-    The syntax for \( E \) and \( \Gamma \) is:
-    \begin{align*}
-        E ::= &\ [] \mid E; x : \Index \mid E; x : T \mid E; x := t : T. \\
-        \Gamma ::= &\ [] \mid \Gamma; x : \Index \mid \Gamma; x : T.
-    \end{align*}
-\end{definition}
-Definitions refer to symbols that can be expanded or unfolded, and typically represent abstract concepts, such as transpose or trace in Dirac notation.
-Assumptions, on the other hand, define the types of variables, and variable assumptions are implicitly universally quantified in the case of equational proofs.
-
-Type checking in our language involves maintaining a well-formed environment and context \( E[\Gamma] \).
-We say an expression \( t \) has type \( X \) in context \( E[\Gamma] \) if the typing judgment \( E[\Gamma] \vdash t : X \) can be proven through the rules in~\Cref{sec: full typing rules}.
-The well-formedness of the context \( \WF(E)[\Gamma] \) is built incrementally:
-\[
-    \frac{}{\WF([])[]}
-    \qquad
-    \frac{\WF(E)[] \qquad x \notin E}{\WF(E; x : \Index)[]}
-    \qquad
-    \frac{E[]\vdash t:T \qquad x \notin E}{\WF(E; x := t:T)[]}.
-\]
-Starting with an empty context, new index symbols can be introduced, and symbols with checked types can be defined.
-Typing judgments are then made inductively using the information from \( E[\Gamma] \).
-For instance, the rule \( x : \Index \in E[\Gamma] \) holds if \( E \) or \( \Gamma \) contains an assumption for \( x \).
-And \(\KType(\sigma)\) is a valid type for kets, if the argument $\sigma$ is typed as an index.
-\begin{gather*}
-    \frac{\WF(E)[\Gamma] \qquad x : \Index \in E[\Gamma]}{E[\Gamma] \vdash x : \Index}
-    \qquad
-    \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \KType(\sigma) : \Type}
-\end{gather*}
-
-Terms are then typed accordingly.
-For example, the ket \( \ket{t} \) will have the type \( \KType(\sigma) \) if \( t \) is a basis term of index \( \sigma \).
-Similarly, the inner product between a bra and a ket of the same index \( \sigma \) is typed as a scalar.
-It corresponds to the constraint of inner product that vectors should be from the same Hilbert space.
-\begin{gather*}
-    \frac{E[\Gamma]\vdash t : \Basis(\sigma)}{E[\Gamma] \vdash \ket{t} : \KType(\sigma)}
-    \qquad
-    \frac{E[\Gamma]\vdash B : \BType(\sigma) \qquad E[\Gamma]\vdash K : \KType(\sigma)}{E[\Gamma] \vdash B \cdot K : \SType}
-\end{gather*}
-
-The typing for functions and applications follows the standard practice.
-For example, the index function \( \mu x.
-t \) is typed with \( x \) as a bound variable of type \( \Index \), and the application \( (t\ u) \) has the type \( U\{x/u\} \), obtained by replacing \( x \) with the index instance \( u \).
-\begin{gather*}
-    \frac{E[\Gamma; x : T] \vdash t : U}{E[\Gamma] \vdash (\lambda x : T . t) : T \to U}
-    \qquad
-    \frac{E[\Gamma] \vdash t:U \to T \qquad E[\Gamma] \vdash u:U}{E[\Gamma] \vdash (t\ u):T} 
-    \\[1em]
-    \frac{E[\Gamma; x : \Index] \vdash t : U}{E[\Gamma] \vdash (\mu x.t) : \forall x.U}
-    \qquad
-    \frac{E[\Gamma] \vdash t:\forall x.U \qquad E[\Gamma] \vdash u : \Index}{E[\Gamma] \vdash (t\ u):U\{x/u\}}
-\end{gather*}
-
-Finally, the big operator sum is modeled by folding a function over a set, with the typing rule as follows:
-\[
-    \frac{E[\Gamma] \vdash s : \SET(\sigma) \qquad E[\Gamma] \vdash f : \Basis(\sigma) \to \KType(\tau)}{E[\Gamma] \vdash \sum_{s} f : \KType(\tau)}.
-\]
-
-\begin{lemma}
-    The typing of expressions is both decidable and unique.
-\end{lemma}
-
-\begin{proof}
-    The type of an expression can be determined recursively. For any given function symbol and argument types, there is at most one typing rule, ensuring the uniqueness of typing.
-\end{proof}
-
-\subsection{Semantics}
-
-The semantics of a language define the meaning of its expressions.
-In this context, the objective of our algorithm is to determine whether two expressions are semantically equivalent.
-We define the semantics in a denotational manner, mapping syntax to set-theoretic objects.
-
-\subsubsection{Denotational Semantics}
-The denotational semantics interpret every expression as an object in linear algebra, according to a valuation mapping \( v \), which assigns values to the variables in the expression.
-The semantics of an expression \( e \) with a given valuation \( v \) is denoted as \( \sem{e}_v \).
-Two expressions \( e_1 \) and \( e_2 \) are considered equivalent if their semantics are equal for all valuations, i.e., \( \sem{e_1}_v = \sem{e_2}_v \) for all \( v \).
-
-The complete interpretation of terms and types is provided in~\Cref{sec: full denotational sem}.
-Variables typed with \( \Index \) are interpreted as finite sets, and the product of two indices \( \sem{\sigma_1 \times \sigma_2} \) is defined as the Cartesian product of the sets \( \sem{\sigma_1} \) and \( \sem{\sigma_2} \).
-More generally, each type is interpreted as a set.
-For example, the scalar type \( \sem{\SType} \) is interpreted as the set of complex numbers \( \mathbb{C} \), and the ket and bra types \( \sem{\KType(\sigma)} \) and \( \sem{\BType(\sigma)} \) are interpreted as the Hilbert space \( \mathcal{H}_{\sem{\sigma}} \) and its dual \( \mathcal{H}_{\sem{\sigma}}^* \), respectively.
-Terms are explained as the set elements.
-For example, the semantics of ket tensor product $\sem{K_1 \otimes K_2} \equiv \sem{K_1} \otimes \sem{K_2}$, is obtained by first calculating the semantics $\sem{K_1}$ and $\sem{K_2}$ as vectors, and then take the vector tensor product as result.
-
-% One special case is the delta function \( \delta_{s,t} \), which is interpreted as:
-% \[
-%     \sem{\delta_{s,t}} =
-%     \begin{cases}
-%         1, & \text{if } \sem{s} = \sem{t}, \\
-%         0, & \text{if } \sem{s} \neq \sem{t}.
-%     \end{cases}
-% \]d
-
-The idea behind the interpretation of types and terms is to formalize the typing relation using set-theoretic inclusion.
-Specifically, for a well-formed context \( E[\Gamma] \), term \( t \), and type \( T \), if \( E[\Gamma] \vdash t : T \), then for any valuation \( v \), the semantics of \( t \) must lie within the semantics of \( T \).
-
-\begin{lemma}
-    For any well-formed context \( E[\Gamma] \), term \( t \), and type \( T \), if \( E[\Gamma] \vdash t : T \), then for all valuations \( v \), \( \sem{t}_v \in \sem{T}_v \).
-\end{lemma}
-
-\begin{proof}
-    The proof follows directly by checking each case.
-\end{proof}
-
-This interpretation formalizes the standard understanding of Dirac notation and provides the foundation for the algorithm.
-However, computers cannot directly reason about equivalence through mathematical interpretations.
-Therefore, we proceed by defining a proof system that abstracts these concepts.
-
-\subsubsection{Axiomatic semantics}
-
-The proof system for equivalence is based on equational logic, together with axioms that describe the properties of Dirac notation.
-A full list of these axioms can be found in~\Cref{sec: full axioms}.
-The axioms cover fundamental aspects of linear spaces, as well as other structures like the tensor and inner products.
-For example, we have the absorption law for zero symbols:
-\(X \cdot \mathbf{0} = \mathbf{0},\)
-and the bilinearity of the dot product:
-\[
-(a.X) \cdot Y = a \cdot (X \cdot Y), \quad X \cdot (Y_1 + Y_2) = X \cdot Y_1 + X \cdot Y_2, 
-\]
-\[
-X \cdot (a.Y) = a \cdot (X \cdot Y), \quad (X_1 + X_2) \cdot Y = X_1 \cdot Y + X_2 \cdot Y.
-\]
-
-As mentioned in the introduction, a subset \( E \) of these axioms cannot be decided by term rewriting alone. The axioms in \( E \) include:
-\[
-\begin{aligned}
-    \text{AC-equivalence} &\quad \text{e.g.,} \quad X + Y = Y + X, \quad (X + Y) + Z = X + (Y + Z), \\
-    \alpha\text{-equivalence} &\quad \lambda x . A = \lambda y . A\{x/y\}, \\
-    \text{SUM-SWAP} &\quad \sum_{i \in s_1} \sum_{j \in s_2} A = \sum_{j \in s_2} \sum_{i \in s_1} A, \\
-    \text{scalar theories} &\quad \text{e.g.,} \quad a + 0 = a, \quad a \times (b + c) = a \times b + a \times c.
-\end{aligned}
-\]
-The scalar theories are treated separately as a module and are not considered in this work. In the implementation, we use the Mathematica kernel to decide scalar equivalences.
-These equational axioms provide an operable theory for the proof automation algorithm.
-Denotational semantics can be seen as one model for this theory, meaning that equivalences derived from the axioms always imply equivalence in the interpretations.
-\begin{lemma}
-    For all well-formed contexts \( E[\Gamma] \) and terms \( e_1, e_2 \), if \( \vdash e_1 = e_2 \), then \( \sem{e_1} = \sem{e_2} \).
-\end{lemma}
-\begin{proof}
-    This follows directly from checking all cases.
-\end{proof}
-
-Thus, the axioms are sound with respect to denotational semantics.
-However, the reverse does not hold: there exist equivalences in denotational semantics that cannot be captured by these axioms.
-Nevertheless, our work focuses on solving practical examples, which are fully covered by the axioms in the experiments.
-
-To conclude this section, we demonstrate the formalization of the symbols used in the motivating example.
-\begin{example}[Formalizing the Motivating Example]
-    \label{ex: formalizing motivating}
-    Definitions and assumptions in the environment \( E \) are formalized as follows:
-    \begin{align*}
-        & \text{TPO} && := \mu T_1. \mu T_2. \lambda O : \OType(T_1, T_2). \sum_{i \in \mathbf{U}(T_1)} \sum_{j \in \mathbf{U}(T_2)} \bra{i} O \ket{j} . \ket{j}\bra{i} \\
-        & &&\quad : \forall T_1. \forall T_2. \OType(T_1, T_2) \to \OType(T_2, T_1); \\
-        &\text{phi} &&:= \mu T. \sum_{i \in \mathbf{U}(T)} \sum_{j \in \mathbf{U}(T)} \ket{(i, j)} : \forall T.\KType(T \times T); \\
-        & T && : \Index; \\
-        & M && : \OType(T, T).
-    \end{align*}
-    The symbol \( \text{TPO} \) represents the transpose of an operator, polymorphic on the Hilbert spaces \( T_1 \) and \( T_2 \).
-The symbol \( \text{phi} \) takes the index \( T \) and defines the maximally entangled states, summing over all basis elements in \( T \), as indicated by the universal set \( \mathbf{U}(T) \).
-    With the assumption of the index \( T \) and operator \( M \), we can express the equivalence in the non-labelled version as:
-    \[
-    (\textrm{M} \otimes \mathbf{1}_\mathcal{O}(\textrm{T})) \cdot (\textrm{phi T}) = (\mathbf{1}_\mathcal{O}(\textrm{T}) \otimes (\textrm{TPO T T M})) \cdot (\textrm{phi T}).
-    \]
-\end{example}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Algorithm for Deciding Dirac Notation Equations}
-\label{sec: decide}
-
-We utilize the term rewriting technique to decide the equivalence of Dirac notation, based on its axiomatic semantics.
-The process involves normalizing two terms and checking whether they have identical syntax.
-This normalization is achieved through matching and substitution in universal algebra, using a set of predefined rewriting rules.
-
-As mentioned in the semantics section, the axioms in \( E \) cannot be fully decided through term rewriting.
-DiracDec decides \( E \) by examining all possible permutations in the rewriting result.
-In this work, we introduce a more efficient approach by incorporating sorting algorithms into the normalization procedure.
-
-The normalization procedure consists of the following steps.
-\begin{enumerate}
-    \item \textbf{First Rewritings}: Expand definitions and simplify expressions.
-    \item \textbf{Variable Expansion}: Consider scalar expressions to ensure completeness.
-    \item \textbf{Second Rewritings}: Normalize terms modulo \( E \).
-    \item \textbf{Sorting Without Bound Variables}: Normalize AC-equivalence.
-    \item \textbf{Swap Successive Summations}: Normalize SUM-SWAP equivalences.
-    \item \textbf{De Bruijn Normalization}: Normalize \( \alpha \)-equivalence.
-\end{enumerate}
-
-\subsection{Normalization Modulo \( E \) by Term Rewriting}
-
-Term rewriting rules, represented as \( l \ \reduce\ r \), are applied recursively to normalize terms.
-In each step, subterms matching the left-hand side \( l \) of a rule are replaced with the corresponding right-hand side \( r \).
-The procedure terminates when no further rewritings can be made.
-A comprehensive list of rewriting rules can be found in~\Cref{sec: rewriting rules}. Below are some key examples to illustrate the design:
-
-One of our optimizations is using functions with indefinite arities. 
-Therefore, we use a flattening rule to handle associativity with AC symbols:
-\[
-a_1 + \cdots + (b_1 + \cdots + b_m) + \cdots + a_n \ \reduce\ a_1 + \cdots + b_1 + \cdots + b_m + \cdots + a_n.
-\]
-Commutativity is handled later in the sorting step. Many of the rewriting rules are directly derived from the equational axioms, such as:
-\begin{align*}
-    & \textrm{(R-DOT6)} && \bra{s} \cdot \ket{t} \ \reduce\  \delta_{s, t}, \\
-    & \textrm{(R-DELTA0)} && \delta_{s, s} \ \reduce\  1, \\
-    & \textrm{(R-MULK1)} && O : \OType(\sigma, \tau) \Rightarrow O \cdot \mathbf{0}_{\KType(\tau)} \ \reduce\  \mathbf{0}_{\KType(\sigma)}, \\
-    & \textrm{(R-MULK11)} && (O_1 \otimes O_2) \cdot (K_1 \otimes K_2) \ \reduce\  (O_1 \cdot K_1) \otimes (O_2 \cdot K_2).
-\end{align*}
-Some of these directions are obvious.
-For example, (R-DOT6) states that the inner product of two basis vectors is reduced to a delta expression, and (R-DELTA0) transforms a delta function on identical basis to a scalar \( 1 \).
-The rule (R-MULK1) reflects the axiom that multiplying a zero vector results in zero.
-This rule is conditional on typing, which is checked during rewriting.
-Some rules, like (R-MULK11), require a deeper understanding, such as the preference for tensor products over multiplication.
-
-As a reference, the term rewriting system in DiracDec has been proven complete for all axioms, except for the sum symbol.
-The completeness result is derived from checking the confluence and termination of the system.
-Our rewriting rules are translations from DiracDec into the typed and abstract language, ensuring that the corresponding symbols in our system are also complete.
-
-We also have additional rules that handle summations. For example:
-\begin{align*}
-    & \quad i \text{ free in } t \Rightarrow \sum_{i \in \mathbf{U}(\sigma)} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (\delta_{i, t}.A) \ \reduce\  \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} A\{i/t\}, \\
-    & \quad \left( \sum_{i \in M} B \right) \cdot K \ \reduce\  \sum_{i \in M} (B \cdot K).
-\end{align*}
-The first rule eliminates delta expressions in summations, while the second rule pushes summations outside of inner products.
-While there is no guarantee of completeness for these rules, they work effectively in practice.
-
-\subsubsection{Variable Expansion}
-One important technique, revealed in the DiracDec work, is the expansion of variables, which is critical for proofs involving summation. For example:
-\[
-\frac{E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash K \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)}(\bra{i} \cdot K).\ket{i}} \quad \quad
-\frac{E[\Gamma] \vdash B : \BType(\sigma)}{E[\Gamma] \vdash B \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)}(B \cdot \ket{i}).\bra{i}},
-\]
-\[
-\frac{E[\Gamma] \vdash O : \OType(\sigma, \tau)}{E[\Gamma] \vdash O \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)} \sum_{j \in \mathbf{U}(\tau)}(\bra{i} \cdot O \cdot \ket{j}).(\ket{i} \cdot \bra{j})}.
-\]
-These rules transform variables into their symbolic summations based on their decomposition over the basis. 
-The rules are not terminating, therefore is applied recrusively once in the second step called \textit{variable expansion}.
-Nevertheless, we have found that applying the expansion only once for all variables is sufficient for normalization.
-\begin{lemma}
-    Let \( \textrm{expand}(e) \) denote the result of expanding all variables in \( e \) once.
-For all well-typed terms \( e \) in \( E[\Gamma] \), \( \textrm{expand}(\textrm{expand}(e)) \) and \( \textrm{expand}(e) \) have the same normal form.
-\end{lemma}
-\begin{proof}
-    Expanding a ket variable twice, for example, results in the following transformation:
-    \[
-    \sum_{i \in \mathbf{U}(\sigma)} (\bra{i} \cdot \sum_{j \in \mathbf{U}(\sigma)} (\bra{j} \cdot K) \cdot \ket{j}) \cdot \ket{i} \ \reduce\  \sum_{i \in \mathbf{U}(\sigma)} \sum_{j \in \mathbf{U}(\sigma)} (\bra{j} \cdot K \cdot \braket{i|j}) \cdot \ket{i},
-    \]
-    where the delta symbol elimination rule returns the term to its original form. The same holds for bra and operator terms.
-\end{proof}
-
-\subsection{Deciding Equational Theory \( E \)}
-
-In the Mathematica implementation of DiracDec, the equational theory \( E \) is decided using unification, which attempts to find a substitution for the summation bound variables that makes two expressions syntactically equivalent.
-This unification process iterates through all permutations of AC symbol arguments, and its complexity is factorial in the maximum number of AC symbol arguments.
-
-As the first improvement, we check \( \alpha \)-equivalence using the de Bruijn index~\cite{deBruijn1972lambda}, which replaces references to bound variable names with the distance from the lambda abstraction to the variable.
-For instance, the nominal lambda abstraction \( \lambda x.
-x \) is transformed into \( \lambda .
-0 \), while \( \lambda x.
-\lambda y.
-(x\ (y\ x)) \) is transformed into \( \lambda.\lambda.
-(1\ (0\ 1)) \).
-This transformation into the de Bruijn index is performed as the final step in the normalization process to check equivalence between terms with different bound variable names.
-
-The remaining axioms, such as AC-equivalence and SUM-SWAP, assert equivalence under permutations.
-A standard approach for deciding such equivalences is to normalize terms by sorting in a predefined order.
-For example, given the dictionary order \( a < b < c \), the term \( b + c + a \) (and any other AC-equivalent term) is normalized into \( a + b + c \).
-However, in our setting, two intertwined difficulties arise: how to assign an order to all terms in the language, and how to simultaneously sort for both axioms.
-
-Consider the following two equivalent terms:
-\[
-\sum_{i \in s_1} \sum_{j \in s_2} \bra{i} A \ket{j} \times \bra{j} B \ket{i}
-= 
-\sum_{i \in s_2} \sum_{j \in s_1} \bra{i} B \ket{j} \times \bra{j} A \ket{i}
-\]
-While these two terms are equivalent, directly sorting the elements of scalar multiplication using lexical order does not yield the same form.
-
-To address this issue, we propose an algorithm that sorts terms without considering bound variables.
-The key observation is that in a successive sum expression \( \sum_{i \in s_1} \cdots \sum_{j \in s_n} A \), the names and order of the bound variables \( i, \dots, j \) can be freely permuted.
-Therefore, all bound variables should be treated uniformly during sorting, and the order of summation can then be determined based on the position of the bound variables.
-
-In the example above, we first ignore the bound variables and sort the sum body into \( \bra{\bullet} A \ket{\bullet} \times \bra{\bullet} B \ket{\bullet} \).
-Then, we swap the summations such that the bound variable at the first \( \bullet \) position appears at the outermost position.
-The results will have the same de Bruijn normal form, namely \( \sum_{s_1} \sum_{s_2} \bra{\$1} A \ket{\$0} \times \bra{\$0} B \ket{\$1} \).
-
-To describe the algorithm in the following, we introduce two key notations.
-For a term \( e = f(a_1, a_2, \dots, a_n) \), \( \textrm{head}(e) \) denotes the function symbol \( f \), while \( \textrm{arg}(e, i) \) refers to the \( i \)-th argument \( a_i \) of the term.
-In this context, variables and constants are treated as functions with zero arguments.
-
-\begin{definition}[Order Without Bound Variables]
-Let \( \mathcal{B} \) represent the set of bound variables, with the assumption that all bound variables are unique.
-We also assume that a total order exists over all symbols.
-The relation \( e_1 =_\mathcal{B} e_2 \) holds if:
-\begin{itemize}
-    \item \( \textrm{head}(e_1) = \textrm{head}(e_2) \), and for all \( i \), \( \textrm{arg}(e_1, i) =_\mathcal{B} \textrm{arg}(e_2, i) \), or
-    \item \( e_1 \in \mathcal{B} \) and \(e_2 \in \mathcal{B}\).
-\end{itemize}
-
-The relation \( e_1 <_\mathcal{B} e_2 \) holds between two terms if:
-\begin{itemize}
-    \item $e_1 \notin \mathcal{B}$ and $e_2 \in \mathcal{B}$, or
-    \item $head(e_1) < head(e_2)$, or
-    \item $head(e_1) = head(e_2)$, and there exists $n$ with $arg(e_1, n) <_\mathcal{B} arg(e_2, n)$, where $arg(e_1, i) =_\mathcal{B} arg(e_2, i)$ for all $i < n$.
-\end{itemize}
-\end{definition}
-It can be shown that \( e_1 =_\mathcal{B} e_2 \) if and only if neither \( e_1 <_\mathcal{B} e_2 \) nor \( e_2 <_\mathcal{B} e_1 \) holds.
-The purpose of this ordering is to compare function symbols in a top-down manner while ignoring bound variables.
-The sorted order enables normalization of terms in terms of AC equivalence.
-\begin{definition}[Sort Transformation]
-    For a term $e$ with bound variable set $\mathcal{B}$,
-    The sort transformation sort(e) is defined as
-    \begin{itemize}
-        \item $e$, if $e$ is a variable or constant;
-        \item $\lambda x:T. \textrm{sort}(e')$, if $e \equiv \lambda x : T. e'$;
-        \item $\mu x. \textrm{sort}(e')$, if $e \equiv \mu x. \textrm{sort}(e')$; or
-        \item $f(\textrm{sort}(a_1), \cdots, \textrm{sort}(a_n))$, if $e \equiv f(a_1, \cdots, a_n)$.
-If $f$ is an AC symbol, then the order of sort($a_i$) is sorted ascendingly according to $<_\mathcal{B}$.
-    \end{itemize}
-\end{definition}
-
-After sorting, the next step is the \textit{swap transformation}, which arranges successive summations based on the order of bound variables.
-\begin{definition}[Swap Transformation]
-For a term \( e \) with a sorting result \( \textrm{sort}(e) \), the swap transformation proceeds by ordering all bound variables according to their first appearances, except in function definitions \( \lambda x \) and \( \mu x \).
-The swap transformation then reorders the successive summations accordingly.
-\end{definition}
-
-The algorithm that applies the sort and swap transformations, followed by de Bruijn normalization, successfully handles all equivalences in our benchmark.
-We formalize the completeness of this transformation in the following conjecture:
-
-\begin{conjecture}[Completeness of Transformation]
-    For any two terms \( e_1 \) and \( e_2 \) that are equivalent under AC equivalence, SUM-SWAP, and \( \alpha \)-equivalence, they will have the same form after sort, swap, and de Bruijn transformations.
-\end{conjecture}
-
-% The order depends on there occurances in the last sorting result. If no occurance, then the order will depend on the set (for sum) and the type (for lambda abstraction only).
-
-%%%%%%%%%%%%%%%%%%%%%%%
-
-% The idea is to assign an order to terms, which is independent on the bound variables. Because we can have terms with nested AC symbols.
-
-Lastly, we can prove that the equivalence established by this normalization procedure is sound with respect to the semantics.
-
-\begin{theorem}[Soundness]
-    For any well-formed context \( E[\Gamma] \) and well-typed expressions \( e_1 \) and \( e_2 \), if  $e_1$ and $e_2$ have the same normal form, then \( \sem{e_1} = \sem{e_2} \).
-\end{theorem}
-
-\begin{proof}
-    The soundness of the term rewriting procedure follows from the fact that each rewriting rule preserves equivalence.
-Furthermore, the operations in the sort and swap transformations respect the AC-equivalence and SUM-SWAP axioms.
-Finally, the de Bruijn normalization ensures soundness for \( \alpha \)-equivalence.
-\end{proof}
-
-% Labelled Dirac Notation
-
-\input{labelled}
-
-\section{Implementation and Case Study}
-
-The main purpose of this work is to build a practical tool that works well in checking Dirac notation equations.
-The refinements and extensions above concludes in our implementation called Dirace, a solver written in \CC.
-It has a parser built by ANTLR4, and scalar reasonings are powered by a Mathematica kernel.
-The user can use commands to make definitions and assumptions in the maintained context, conduct the normalization and equivalence checking, and obtain the rewriting trace output.
-This implementation
-is tested on the benchmark of the DiracDec work, and succeeds in proving most of them efficiently. 
-It can be used from the command line interactively, or can be integrated into other \CC\ projects as a library.
-
-\subsubsection{Project Structure}
-The project structure is illustrated in~\Cref{fig: dirace structure}.
-\texttt{ualg} is the module for universal algebra, defining basic concepts like terms and substitutions.
-It serves as the library for \texttt{dirace}, which are then utilized in the example benchmarks and the toplevel command line application.
-The components of \texttt{dirace} are as follows:
-\begin{itemize}
-    \item \texttt{symbols.cpp}: the reserved symbols and AC symbols;
-    \item \texttt{syntax\_theory.cpp}: syntax related algorithms, such as de Bruijn normalization and freeness of variables;
-    \item \texttt{calculus.cpp}: type checker and intergration with Mathematica;
-    \item \texttt{reduction.cpp}: all the rewriting rules and transformations;
-    \item \texttt{dirac\_parser.cpp}: parser for Dirac notation and Dirace commands;
-    \item \texttt{prover.cpp}: the prover that maintains the context and process commands like definition or equivalence checking.
-\end{itemize}
-
-\input{dirace.tex}
-
-The internal data structure for terms is a pointer-based syntax tree following the function application style:
-\[
-    \texttt{
-        term ::= ID | ID [term (, term)*].
-    }
-\]
-The syntax tree can be an identifier, or an application with an identifier as the function head, and several syntax trees as arguments.
-There are several Dirac notation terms and their corresponding syntax trees.
-\footnotesize{
-\begin{align*}
-    & X_1 + X_2 + X_3 && \texttt{ADD[X1, X2, X3]} 
-    \\
-    & \lambda x: \OType(T_1,  T_2). x^\dagger && \texttt{FUN[x, OTYPE[T1, T2], ADJ[x]]}
-    \\
-    & \sum_{i \in \mathbf{U}(T)} \ket{i} \bra{i} && \texttt{SUM[USET[T], FUN[i, BASIS[T], OUTER[KET[i], BRA[i]]]]}
-\end{align*}
-}
-The syntax tree structure is also compatible with the datatype of Mathematica. This improves the interoperability between Dirace and the Mathematica system, enabling them to work interleavingly.
-To improve usability, Dirace also supports many special syntacies for terms,  and most Dirac notation terms will be encoded in the natural way.
-Here are some examples for the parsing syntax.
-
-\begin{figure}
-    \center
-\begin{tabular}{c >{\centering\arraybackslash}p{4cm} l}
-    \hline
-    syntax & parsing result & explanation \\
-    \hline
-    \texttt{|e>} & \texttt{KET[e]} & the ket basis\\
-    \texttt{e1 + ... + en} & \texttt{ADD[e1, ..., en]} & the addition\\
-    \texttt{e1\ e2} & \texttt{COMPO[e1, e2]} & composition in Dirac notation \\
-    \texttt{e1\^{}*} & \texttt{CONJ[e1]} & scalar conjugation \\
-    \texttt{fun i : T => X} & \texttt{FUN[i, T, X]} & lambda abstraction \\
-    \hline
-\end{tabular}
-\end{figure}
-
-Finally, Dirace uses a prover to host the computation.
-The prover maintains a well-formed context $E[\Gamma]$, and processes commands to modify the context and conduct calculations.
-The commands are listed below.
-\begin{itemize}
-    \item \texttt{\textcolor{NavyBlue}{Def} ID := term.} It defines the \texttt{ID} as the \texttt{term}, using the \textbf{W-Def} typing rule.
-    \item \texttt{\textcolor{NavyBlue}{Var} ID := term.} It make an assumption of \texttt{ID} with the \texttt{term} as type, using the \textbf{W-AssumeE} typing rules.
-    \item \texttt{\textcolor{NavyBlue}{Check} term.} Type checking the \texttt{term} and output the result.
-    \item \texttt{\textcolor{NavyBlue}{Normalize} term.} Normalize the \texttt{term} using the algorithm introduced in~\Cref{sec: decide}.
-    \item \texttt{\textcolor{NavyBlue}{CheckEq} term \textcolor{NavyBlue}{with} term.} Check the equivalence of the two terms calculating and comparing their normal forms.
-\end{itemize}
-The prover will type check the terms for each command.
-We can also use \texttt{\textcolor{NavyBlue}{Normalize} term \textcolor{NavyBlue}{with trace}.} to output the proof trace during normalization.
-The proof trace is a sequence of records, including the rule or transformation appied, the position of application, and the pre- and post-transformation terms.
-The record helps understand the normalization procedure better, and can be turned into verified proofs in theorem provers in the future.
-
-\subsubsection{Use Case}
-Here we encode the motivating~\Cref{ex: motivating}, examine and explain how it is checked in Dirace. The encoding is shown below.
-
-    \begin{lstlisting}[style=dirace]
-Var T : INDEX. Var M : OTYPE[T, T].
-Def phi := idx T => Sum nv in USET[T], |(nv, nv)>.
-Var r1 : REG[T]. Var r2 : REG[T].
-CheckEq M_r1;r1 (phi T)_(r1, r2) with (TPO T T M)_r2;r2 (phi T)_(r1, r2).
-    \end{lstlisting}
-
-The first three lines use the \texttt{\textcolor{NavyBlue}{Var}} and \texttt{\textcolor{NavyBlue}{Def}} commands to set up the context for the Dirac notation.
-\texttt{T} is a type index, representing arbitrary Hilbert space types.
-\texttt{M} is assume to be an operator in the Hilbert space with type \texttt{T}.
-\texttt{phi} is defined as the maximally entangled state, depending on the bound variable \texttt{T} as index.
-\texttt{r1} and \texttt{r2} are register names for the two subsystems.
-
-In the left hand side of \texttt{\textcolor{NavyBlue}{CheckEq}} command, \texttt{M\_r1;r1} denotes the labelled notation $M_{r_1; r_1}$, and \texttt{(phi T)\_(r1, r2)} denotes the entangled state $\ket{\Phi}_{(r_1, r_2)}$.
-They are connected by a white space, which is parsed into the composition of Dirac notation, and will be reduced into the operator-ket multiplication after typing.
-The right hand side is interpreted similarly, except the defined symbol \texttt{TPO} in the environment:
-
-\begin{lstlisting}[style=dirace]
-Def TPO := idx sigma => idx tau => fun O : OTYPE[sigma, tau] => Sum i in USET[sigma], Sum j in USET[tau], (<i| O |j>).(|j> <i|).
-\end{lstlisting}
-
-The \texttt{TPO} symbol represents the transpose of operators, and encodes the formalization in~\Cref{ex: formalizing motivating}.
-Thanks to the design of environment and functions, many other commonly used symbols in Dirac notation are encoded and provided as defined symbols in Dirace.
-
-Within one second, the prover reports the result of equivalence with their common normal form:
-    \begin{lstlisting}[style=dirace]
-The two terms are equal.
-[Normalized Term] SUM[USET[T], FUN[BASIS[T], SUM[USET[T], FUN[BASIS[T], SCR[DOT[BRA[$\texttt{\$1}$], MULK[M, KET[$\texttt{\$0}$]]], LTSR[LKET[$\texttt{\$1}$, r1], LKET[$\texttt{\$0}$, r2]]]]]]] : DTYPE[RSET[r1, r2], RSET]
-    \end{lstlisting}
-
-The normal form is in the internal syntax tree format mentioned above. A more readable interpretation is:
-\[
-\sum_{\mathbf{U}(T)} \sum_{\mathbf{U}(T)} \bra{\$1}M\ket{\$ 0} . \ket{\$1}_{r_1} \otimes \ket{\$0}_{r_2} : \DType(\{r_1, r_2\}, \emptyset).
-\]
-Here $\$0$ and $\$1$ are de Bruijn indices. The result is a ket on the $\{r_1, r_2\}$ system as expected, and follows pattern proposed in~\Cref{sec: labelled}.
-
-\subsubsection{Benchmark performance}
-
-To evaluate Dirace, we first test the examples from DiracDec benchmark and make a comparison.
-The experiments are carried out using a MacBook Pro with M3 Max chip.
-
-\begin{figure}
-    \center
-    \begin{tabular}{c|c c c|c c c}
-        \hline
-        \multirow{2}{*}{source} & \multicolumn{3}{c|}{DiracDec} & \multicolumn{3}{c}{Dirace} \\
-        \cline{2-7}
-                                 & expressable & success & time(s)           & expressable & success & time(s)                 \\
-        \hline
-        textbook(QCQI)          & 18          & 18        &    1.02        &    18      & 18          &   0.82      \\
-        CoqQ                    & 162          & 156       &    48.69       &   158     &  158   &     9.74     \\
-        circuits                 & 2          & 2       &    17.67       &   3     &  2   &     1.4     \\
-        research paper                & 4          & 4         &  59.53       &   4    & 4       &  0.73     \\
-        \hline
-    \end{tabular}        
-    \caption{For DiracDec, examples that cannot be decided within 60 seconds are not included.}
-\end{figure}
-
-The timing of Dirace does not include the initialization of Mathematica kernel link, which takes about 3 seconds in the beginning.
-As to expressibility, the language for DiracDec has the support for projectors $\texttt{fst}$ and $\texttt{snd}$ on basis pairs, satisfying $\texttt{fst} (s, t) = s$ and $\texttt{snd} (s, t) = t$.
-We found this feature is almost not used, so we removed the support.
-As a result, Dirace encodes 158 examples for the CoqQ part, 4 less than DiracDec.
-For decidability, we improved the rewriting rules about sum for Dirace, and proved several more examples failed by DiracDec.
-The main difference is about their time efficiency.
-Because of our algorithm to decide AC-equivalence and SUM-SWAP, Dirace has a significant efficiency improvement, especially on those ``computational examples'' mentioned in the DiracDec paper.
-One typical example comes from the paper by Jens about the equivalence of operators for qubits.
-The system has to decompose the term on the concrete $\ket{0}$ and $\ket{1}$ basis, resulting a lot of addition elements.
-It takes DiracDec about one minute, but Dirace solves it within one second.
-
-We also built an example benchmark for labelled Dirac notation.
-\yx{We need examples.}
-
-\section{Related Work}
-
-Automated theorem proving has seen significant advancements in recent years, particularly through the development of satisfiability modulo theories (SMT) solvers,
-% SMT solvers extend propositional satisfiability by integrating theories such as arithmetic, arrays, and bit-vectors, making them suitable for solving a wide range of verification problems in both hardware and software domains.
-including prominent tools like Z3. These solvers have become essential in various fields such as formal verification, synthesis, and model checking. 
-Equational reasoning is another crucial area of research within automated theorem proving, which focuses on solving problems that involve equations between terms in an algebraic structure. 
-Equational provers, such as Vampire and E, have played a pivotal role in addressing the challenge of proving equations in first-order logic, employing sophisticated algorithms like superposition and term rewriting.
-
-Formal verification of quantum computation is receiving increasing attentions during these years. See~\cite{Lewis2023} for a comprehensive review. 
-Verification frameworks in Coq include the foundational formalization CoqQ and quantum circuit language QWIRE. 
-Verifications of quantum programs are also considered, such as the Hoare logic based methods [] and model checking based methods [].
-The equational reasoning of Dirac notation is crucial for establishing property proofs in these works.
-Verification theoreis and tools based on other languages are also proposed, such as PyZX for reasoning and simplification of ZX-calculus.
-
-\section{Conclusion and Future Work}
-Based on the first equational reasoning tool for Dirac notation called DiracDec, this work improves and extends the theory for practical applications, and provides the solver Dirace.
-Experiments show that the tool demonstrates advantages in decidability, efficiency and usability.
-
-We expect Dirace to have applications in areas like quantum program verification or proofs of post-quantum cryptography protocols in the future.
-One promising following up is to connect Dirace with theorem provers like Coq.
-It involves transforming theorem prover expressions into Dirace, and verify the proof trace of Dirace in theorem provers.
-Besids, most quantum program verifiers nowadays depend on matrix calculations.
-Dirace can serve as the replacement for matrix methods to enable symbolic deductions.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% \begin{credits}
-%     \subsubsection{\ackname} A bold run-in heading in small font size at the end of the paper is
-%     used for general acknowledgments, for example: This study was funded
-%     by X (grant number Y).
-% \end{credits}
-
-%
-% ---- Bibliography ----
-%
-% BibTeX users should specify bibliography style 'splncs04'.
-% References will then be sorted and formatted in the correct style.
-%
-\bibliographystyle{splncs04}
-\bibliography{ref}
-%
-
-\appendix
-\include{appendix}
-
-\end{document}

### 2025-01-27 02techniques.tex
splited

--- 2025-01-27 main.tex
@@
-% This is samplepaper.tex, a sample chapter demonstrating the
-% LLNCS macro package for Springer Computer Science proceedings;
-% Version 2.21 of 2022/01/12
-%
-\documentclass[runningheads]{llncs}
-
-% \usepackage[utf8]{inputenc}
-% \usepackage[margin=0.8in]{geometry}
-
-\usepackage{color}
-\usepackage{amsmath}
-%\usepackage{amssymb}
-\usepackage{graphicx}
-% \usepackage{amsthm}
-\usepackage{stmaryrd}
-\usepackage[all]{xy}
-\usepackage{multirow}
-\usepackage{paralist}
-\usepackage{hhline}
-\usepackage{bm}
-\usepackage{longtable}
-\usepackage{makecell}
-\usepackage{mdframed}
-
-\usepackage{wasysym}
-\usepackage{extarrows}
-\usepackage{tikz}
-
-% add new commands for comments here
-\newcommand{\yx}[1]{\textit{\color{blue}[YX] : #1}}
-\newcommand{\lz}[1]{\textit{\color{red}[LZ] : #1}}
-
-\newcommand{\modify}[1]{{\color{red}#1}}
-
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{stmaryrd}
-\usepackage{hyperref}
-
-\usepackage{braket}
-\usepackage{cleveref}
-
-\input{lstlisting}
-
-\usepackage{tikz}
-\usetikzlibrary{shapes.geometric, arrows}
-\usetikzlibrary{trees}
-\tikzstyle{arrow} = [thick,->,>=stealth]
-
-\newenvironment{ruletable}[1]
-{
-    \begin{longtable}{cl}
-    \caption{#1}\\
-    \hline
-    \textbf{Rule} & \textbf{Description} \\
-    \hline
-    \endfirsthead
-
-    \hline
-    \textbf{Rule} & \textbf{Description} \\
-    \hline
-    \endhead
-
-    % \hline
-    % \multicolumn{2}{r}{\textit{Continued on the next page}} \\
-    \hline
-    \endfoot
-
-    \hline
-    \endlastfoot
-}
-{
-    \end{longtable}
-}
-
-% define C++ logo
-\newcommand{\CC}{C\nolinebreak\hspace{-.05em}\raisebox{.4ex}{\tiny\bf +}\nolinebreak\hspace{-.10em}\raisebox{.4ex}{\tiny\bf +}}
-\def\CC{{C\nolinebreak[4]\hspace{-.05em}\raisebox{.4ex}{\tiny\bf ++}}}
-
-%
-\usepackage[T1]{fontenc}
-% T1 fonts will be used to generate the final print and online PDFs,
-% so please use T1 fonts in your manuscript whenever possible.
-% Other font encondings may result in incorrect characters.
-%
-\usepackage{graphicx}
-% Used for displaying a sample figure. If possible, figure files should
-% be included in EPS format.
-%
-% If you use the hyperref package, please uncomment the following two lines
-% to display URLs in blue roman font according to Springer's eBook style:
-\usepackage{color}
-\renewcommand\UrlFont{\color{blue}\rmfamily}
-\urlstyle{rm}
-%
-\begin{document}
-%
-\title{Dirace: Practical Proof Automation of Dirac Notation Equations}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% %
-% %\titlerunning{Abbreviated paper title}
-% % If the paper title is too long for the running head, you can set
-% % an abbreviated paper title here
-% %
-% \author{First Author\inst{1}\orcidID{0000-1111-2222-3333} \and
-% Second Author\inst{2,3}\orcidID{1111-2222-3333-4444} \and
-% Third Author\inst{3}\orcidID{2222--3333-4444-5555}}
-% %
-% \authorrunning{F. Author et al.}
-% % First names are abbreviated in the running head.
-% % If there are more than two authors, 'et al.' is used.
-% %
-% \institute{Princeton University, Princeton NJ 08544, USA \and
-% Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
-% \email{lncs@springer.com}\\
-% \url{http://www.springer.com/gp/computer-science/lncs} \and
-% ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
-% \email{\{abc,lncs\}@uni-heidelberg.de}}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%
-\maketitle              % typeset the header of the contribution
-%
-\begin{abstract}
-Dirac notation is a fundamental language in quantum computation and quantum information.
-Recently, the term rewriting system DiracDec~\cite{diracdec} was introduced to automate equational reasoning with Dirac notation, a critical yet time-intensive task.
-Building upon DiracDec, this work aims to develop a solver optimized for practical applications.
-Enhancements include an improved typing system, a simplified language and rewriting system, more efficient algorithms, and added support for labeled Dirac notation.
-Our solver, Dirace, is implemented in \CC\ with a Mathematica backend, demonstrating significant improvements in decision-making power and time efficiency.
-
-% \keywords{First keyword  \and Second keyword \and Another keyword.}
-\end{abstract}
-%
-%
-%
-
-%%%%%%%%%%%%%%%%
-\newcommand*{\sem}[1]{{\llbracket #1 \rrbracket}}
-\newcommand{\DiracDec}{\textsf{DiracDec}}
-
-\newcommand{\reduce}{\triangleright}
-
-\newcommand{\Sort}{\mathsf{Sort}}
-\newcommand{\WF}{\mathcal{WF}}
-
-\newcommand{\Index}{\mathsf{Index}}
-\newcommand{\Type}{\mathsf{Type}}
-\newcommand{\Basis}{\mathsf{Basis}}
-
-\newcommand{\SType}{\mathcal{S}}
-\newcommand{\KType}{\mathcal{K}}
-\newcommand{\BType}{\mathcal{B}}
-\newcommand{\OType}{\mathcal{O}}
-\newcommand{\SET}{\mathsf{Set}}
-
-\newcommand{\ZEROK}{\mathbf{0}_\mathcal{K}}
-\newcommand{\ZEROB}{\mathbf{0}_\mathcal{B}}
-\newcommand{\ZEROO}{\mathbf{0}_\mathcal{O}}
-
-\newcommand{\PAIR}{\mathsf{PAIR}}
-
-\newcommand{\ZERO}{\mathsf{0}}
-\newcommand{\ONE}{\mathsf{1}}
-\newcommand{\ADDS}{\mathsf{ADDS}}
-\newcommand{\ADD}{\mathsf{ADD}}
-\newcommand{\MULS}{\mathsf{MULS}}
-\newcommand{\MUL}{\mathsf{MUL}}
-\newcommand{\CONJ}{\mathsf{CONJ}}
-\newcommand{\CJG}{\mathsf{CJG}}
-\newcommand{\ADJ}{\mathsf{ADJ}}
-\newcommand{\DELTA}{\mathsf{DELTA}}
-\newcommand{\DOT}{\mathsf{DOT}}
-\newcommand{\SCR}{\mathsf{SCR}}
-\newcommand{\TSR}{\mathsf{TSR}}
-\newcommand{\KET}{\mathsf{KET}}
-\newcommand{\BRA}{\mathsf{BRA}}
-\newcommand{\ONEO}{\mathbf{1}_\mathcal{O}}
-\newcommand{\OUTER}{\mathsf{OUTER}}
-\newcommand{\MULK}{\mathsf{MULK}}
-\newcommand{\MULB}{\mathsf{MULB}}
-\newcommand{\MULO}{\mathsf{MULO}}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Introduction}
-
-In 1939, Dirac introduced his notation for quantum mechanics~\cite{dirac1939new}, designed to represent linear algebraic formulas in a compact and convenient form.
-For example, the expression \( a\ket{\psi} + b\ket{\phi} \) represents the superposition of two quantum states, \( \ket{\psi} \) and \( \ket{\phi} \).
-Today, Dirac notation is widely accepted as the standard language in quantum computation and quantum information.
-Its reasoning forms the foundation of research and applications, much like boolean and integer logic do in classical computer science.
-
-In quantum algorithm formalizations and quantum programming languages, Dirac notation is frequently used in equational proofs, which are critical, repetitive, and often time-intensive.
-These notations also play a key role in defining program states, operations, and assertions in quantum programming languages.
-To automate the verification of these programs, we need tools that can simplify and check the equivalence of preconditions.
-However, unlike the well-established SAT and SMT solvers for classical logic, a practical solver for Dirac notation equivalence remains an unmet need, creating a barrier for progress in several fields.
-
-Recently, Xu et al.~\cite{diracdec} proposed a theory for deciding the equivalence of Dirac notation, alongside a prototype implementation in Mathematica called DiracDec.
-They demonstrated that the equivalence of basic Dirac notation is decidable.
-Their algorithm, based on a pure term rewriting system, has been proven to be confluent and terminating.
-Despite this, there remains a gap between DiracDec and a practical solver for Dirac notation equivalence.
-
-One challenge is the efficiency of the algorithm when dealing with equivalences beyond the scope of term rewriting.
-DiracDec decides the entire equational theory by rewriting modulo \( E \), where \( E \) represents a set of axioms that cannot be decided by normalization alone, such as the associativity and commutativity (AC) of certain function symbols.
-DiracDec uses a direct but inefficient algorithm to decide these axioms, which searches through all possible permutations and exhibits factorial complexity.
-This inefficiency becomes particularly evident in "computational examples" containing many AC symbols, which are time-consuming to process.
-
-Usability is another area where DiracDec falls short.
-It does not support labelled Dirac notation, which uses registers to denote subsystems and express states locally.
-Additionally, DiracDec's typing system only does not provide context for variable typing assumptions or the definition of symbols, which are required in practical scenarios.
-To avoid type checking during term rewriting, DiracDec separates symbols (e.g., multiplication) for different types, leading to unnecessary complexity.
-Moreover, integrating the Mathematica implementation into other projects as a solver is challenging.
-
-The system design of DiracDec reflects a trade-off between simplicity and efficiency.
-While the simplicity of term rewriting allows for strong theoretical results, it limits optimization opportunities.
-Building on DiracDec, this work aims to develop a practical solver.
-We transform the term rewriting system into a hybrid algorithm, overcoming the challenges mentioned above.
-Our main technical contributions are:
-\begin{itemize}
-    \item An efficient algorithm for deciding the equational theories in \( E \), based on equivalence checking through normalization, with the normal form for \( E \) being obtained via sorting.
-    \item Support for constant register labels, and reducing the equivalence decision of labelled Dirac notation to the unlabelled case.
-    \item A more user-friendly \CC\ solver, Dirace, featuring an abstract language and typing system. We also support the definition of symbols (e.g., transpose and trace) using function syntax.
-\end{itemize}
-
-We evaluated Dirace against the DiracDec benchmark and new examples involving labelled Dirac notation.
-The results show significant improvements in both decidability and efficiency compared to DiracDec.
-
-% Dirace successfully decides all the examples that are expressable in its language, including those failed by DiracDec because of complexity or insufficient decision power.
-
-\section{Motivation and Preliminary}
-
-An interesting property of quantum mechanics is that for two maximally entangled states, applying a quantum operator \( M \) to one subsystem is equivalent to applying \( M^T \) (the transpose of \( M \)) to the other subsystem.
-This relationship holds regardless of the spatial separation between the two systems, and it can be expressed as an equation in Dirac notation.
-\begin{example}
-    \label{ex: motivating}
-    Let \( q \) and \( r \) represent two quantum systems in the Hilbert space \( \mathcal{H}_T \).
-Let \( M \) be a quantum operation acting on \( \mathcal{H}_T \), and let \( \ket{\Phi} = \sum_{i \in T} \ket{i} \otimes \ket{i} \) be the maximally entangled state.
-Then, we have the following equation:
-    \[
-    M_q \ket{\Phi}_{q; r} = M_r^T \ket{\Phi}_{q; r}.
-    \]
-\end{example}
-
-In this equation, \( q \) and \( r \) are labels denoting the respective subsystems in the Dirac notation.
-To automate reasoning about such equations, we must formalize the language and develop a proof system to handle it.
-
-\subsubsection{Dirac Notation}
-
-Quantum states are represented as vectors in complex Hilbert spaces, and operations on these states are described by linear transformations.
-Dirac notation uses the ket \( \ket{i} \) and the bra \( \bra{i} \) to denote basis vectors in a Hilbert space and its dual space, respectively.
-These symbols can be composed together in various ways to form more complex expressions.
-The meaning of these compositions depends on the types of the operands involved.
-For example, the inner product \( \braket{i|j} \) represents the scalar result of the dot product between \( \bra{i} \) and \( \ket{j} \), while the outer product \( \ket{i}\bra{j} \) represents an operator.
-Additionally, Dirac notation uses the tensor product symbol \( \otimes \) to describe the combined space of multiple quantum systems.
-
-One of the key features of Dirac notation is its order-independent interpretation.
-This means that the composition of terms can be written without parentheses, as the interpretation is unaffected by the order of multiplication.
-For instance, the formula \( \bra{i}\ket{\phi}\bra{\psi}\ket{j} \) can be understood as:
-\[
-    \braket{i|\phi}\braket{\psi|j} = \bra{i} (\ket{\phi}\bra{\psi}) \ket{j},
-\]
-and these expressions are equivalent for all variables involved.
-
-In practice, Dirac notation is often combined with other syntactic elements, such as the summation symbol \( \sum_{i \in S} A \), to enhance expressiveness.
-Furthermore, labelled Dirac notation, such as \( \ket{i}_q \otimes \ket{i}_r \), is used to denote quantum systems in consideration, where subscripts (registers) are added to distinguish different subsystems.
-
-\subsubsection{Universal Algebra}
-We use universal algebra and equational logic to formally represent Dirac notation and the reasoning procedure.
-A universal algebra defines a signature of function symbols, with terms constructed from constants, variables, and function applications.
-Other basic concepts like substitution of variables or pattern matching are also defined.
-In our case of Dirac notation, the signature consists of constructors and operations like $\ket{i}$ or $A \otimes B$.
-The reasoning process is guided by equational logic, which defines an equivalence relation that is compatible with substitution and term construction.
-This relation formalizes the intuitive concept of equivalence in algebra.
-
 \section{Language, Typing and Semantics}
@@
 
-\input{labelled}
+\section{Labelled Dirac Notation}
+\label{sec: labelled}
+Labelled Dirac notation uses register names to indicate the quantum system of vectors and operators.
+This allows us to express and reason aboud the states and operations locally, without referring to the whole system.
+For instance, assume $Q$ and $R$ are two registers, we have
+\[
+    M_Q \cdot \ket{\Phi}_{(Q, R)} = ((M \otimes I) \cdot \ket{\Phi})_{(Q, R)}.
+\]
+On the left hand side, the state of two subsystems is $\ket{\Phi}$, and we apply quantum operation $M$ on the system $Q$.
+It is equivalent to extend the operation using identity operators on other subsystems (i.e.
+the cylinder extension), and consider the application in the whole system.
 
-\section{Implementation and Case Study}
+In this section, we introduce the language and typing of registers and labelled Dirac notation, and demonstrate how to transform the equivalence problem into that for the Dirac notation studied above.
+The first new notion is quantum registers, and we assume they are constructed from a set $\cR$.
+\begin{definition}[quantum registers]
+  \begin{align*}
+    R ::= r\in\cR \mid (R, R).
+  \end{align*}
+\end{definition}
 
-The main purpose of this work is to build a practical tool that works well in checking Dirac notation equations.
-The refinements and extensions above concludes in our implementation called Dirace, a solver written in \CC.
-It has a parser built by ANTLR4, and scalar reasonings are powered by a Mathematica kernel.
-The user can use commands to make definitions and assumptions in the maintained context, conduct the normalization and equivalence checking, and obtain the rewriting trace output.
-This implementation
-is tested on the benchmark of the DiracDec work, and succeeds in proving most of them efficiently. 
-It can be used from the command line interactively, or can be integrated into other \CC\ projects as a library.
+Registers can be composed by pairs of $(R, R)$, and this structure corresponds to the structure of tensor product spaces in Dirac notation.
+To reason about the registers, we define their variable set as the enumeration of all register symbols involved.
 
-\subsubsection{Project Structure}
-The project structure is illustrated in~\Cref{fig: dirace structure}.
-\texttt{ualg} is the module for universal algebra, defining basic concepts like terms and substitutions.
-It serves as the library for \texttt{dirace}, which are then utilized in the example benchmarks and the toplevel command line application.
-The components of \texttt{dirace} are as follows:
+\begin{definition}[register variable set]
+The variable set of a register is defined inductively by:
 \begin{itemize}
-    \item \texttt{symbols.cpp}: the reserved symbols and AC symbols;
-    \item \texttt{syntax\_theory.cpp}: syntax related algorithms, such as de Bruijn normalization and freeness of variables;
-    \item \texttt{calculus.cpp}: type checker and intergration with Mathematica;
-    \item \texttt{reduction.cpp}: all the rewriting rules and transformations;
-    \item \texttt{dirac\_parser.cpp}: parser for Dirac notation and Dirace commands;
-    \item \texttt{prover.cpp}: the prover that maintains the context and process commands like definition or equivalence checking.
+    \item $\var(R) = \{r\}$, if $R\equiv r$; or
+    \item $\var(R) = \var(R_1) \cup\var(R_2)$, if $R\equiv (R_1,R_2)$.
 \end{itemize}
+\end{definition}
 
-\input{dirace.tex}
+% \textbf{Remark: } Set operations: $\cup$ for union; $\cap$ for intersection; $\setminus$ for difference. So,
+% % \begin{itemize}
+% $ S_1 \cap S_2 \equiv S_1 \cup S_2 \setminus (S_1 \setminus S_2) \setminus (S_2 \setminus S_1) $.
+% \end{itemize}
 
-The internal data structure for terms is a pointer-based syntax tree following the function application style:
+% The labelled Dirac notation is an extension on the existing type and syntax.
+\begin{definition}[labelled Dirac notation]
+  The \textbf{labelled Dirac notation} includes all Dirac notation symbols and the generators defined below.
+  Here, $s\subseteq \cR$ is a register variable set.
+  \begin{align*}
+    T & ::= \DType(s,s) \mid \mathsf{Reg}(\sigma) \\
+    e & ::= R \mid |i\>_r \mid {}_r\<i| \mid e_R \mid e_{R;R} \mid
+    e \otimes e \otimes \cdots \otimes e \mid e \cdot e
+  \end{align*}
+\end{definition}
+$\DType(s_1, s_2)$ is the unified type for all labelled Dirac notation, where $s_1$ indicates the codomain systems and $s_2$ indicates the domain systems.
+For instance, labelled ket has type $\DType(s_1, \emptyset)$, and labelled bra has type $\DType(\emptyset, s_2)$.
+$\reg(\sigma)$ are types for registers $R$, and the index $\sigma$ indicates the type of Hilbert space represented by the register.
+Terms also include the labelled notation $e_R$ for bra, ket and $e_{R;R}$ for operators.
+We introduce new dot and tensor product symbols for labelled Dirac notation.
+In labelled Dirac notation, the structure of tensor product does not matter, therefore $\otimes$ is an AC symbol.
+Following the unified type $\DType(s,s)$, all kinds of multiplications are represented by the same dot product $e \cdot e$.
+Finally, $\ket{i}_r$ and ${}_r\bra{i}$ are labelled basis for the normal form of labelled Dirac notation, where $r$ are registers symbols in $\mathcal{R}$.
+
+\subsubsection{Typing rules}
+Some typing rules are introduced here.
+The rule for $\DType(s_1, s_2)$ requires that all registers in variable set $s_1$ and $s_2$ are well-typed.
+The rule for $K_R$ demonstrates how a register label is added to the Dirac notation, and the rule for $D^\dagger$ shows that labelled Dirac notation also have symbols for calculation, such as the adjoint.
 \[
-    \texttt{
-        term ::= ID | ID [term (, term)*].
+    \frac{E[\Gamma] \vdash  \sigma : \Index}{E[\Gamma] \vdash \reg(\sigma) : \Type}
+    \qquad
+    \frac{E[\Gamma] \vdash r : \reg(\sigma_r) \text{ for all $r$ in $s_1$ and $s_2$} }{E[\Gamma] \vdash \DType(s_1, s_2) : \textsf{Type}} 
+\]
+\[
+    \frac{E[\Gamma] \vdash R : \reg(\sigma)\qquad E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash K_R : \DType(\var R, \emptyset)}
+    \qquad
+    \frac{E[\Gamma] \vdash D : \DType(s_1,s_2)}{E[\Gamma] \vdash D^\dagger : \DType(s_2,s_1)}
+\]
+The dot and the tensor product symbols are different from those in unlabelled Dirac notation.
+Since the goal of labels is to replace the order and structure of tensor products by the reference to registers, the tensor product becomes an AC symbol.
+The typing still checks whether the component subsystems are disjoint with each other.
+\[
+    \frac{
+        E[\Gamma] \vdash D_i : \DType(s_i,s_i') \qquad
+        \bigcap_i s_i = \emptyset \qquad
+        \bigcap_i s_i' = \emptyset
     }
+    {E[\Gamma] \vdash D_1 \otimes \cdots \otimes D_i : \DType(\bigcup_i s_i, \bigcup_i s_i')}.
 \]
-The syntax tree can be an identifier, or an application with an identifier as the function head, and several syntax trees as arguments.
-There are several Dirac notation terms and their corresponding syntax trees.
-\footnotesize{
+As for the dot product, the disjointness is considered except registers contracted by multiplication.
+\[
+    \qquad
+    \frac{
+        \begin{aligned}
+            E[\Gamma] \vdash D_1 : \DType(s_1,s_1') \\
+            E[\Gamma] \vdash D_2 : \DType(s_2,s_2')
+        \end{aligned}
+        \qquad 
+        \begin{aligned}
+            s_1 \cap s_2 \backslash s_1' = \emptyset \\
+            s_2' \cap s_1' \backslash s_2 = \emptyset
+        \end{aligned}
+    }
+    {E[\Gamma] \vdash D_1\cdot D_2 : \DType(s_1 \cup (s_2\backslash s_1'), s_2' \cup (s_1'\backslash s_2))}.
+\]
+
+\subsubsection{Normalization}
+The normalization procedure of Dirac notation is extended to check equivalence with labels.
+We add rules to the term rewriting system, which in general try to represent labelled Dirac notation with labelled basis and scalar coefficients.
+The first step is the label elimination.
+Take operator as an example:
 \begin{align*}
-    & X_1 + X_2 + X_3 && \texttt{ADD[X1, X2, X3]} 
-    \\
-    & \lambda x: \OType(T_1,  T_2). x^\dagger && \texttt{FUN[x, OTYPE[T1, T2], ADJ[x]]}
-    \\
-    & \sum_{i \in \mathbf{U}(T)} \ket{i} \bra{i} && \texttt{SUM[USET[T], FUN[i, BASIS[T], OUTER[KET[i], BRA[i]]]]}
+    & O_{R,R'} \ \reduce\ \sum_{i_{r_1}\in\bU(\sigma_{r_1})}\cdots \sum_{i_{r_n}\in\bU(\sigma_{r_n})}
+    \sum_{i_{r'_1}\in\bU(\sigma_{r'_1})}\cdots \sum_{i_{r'_{n'}}\in\bU(\sigma_{r'_{n'}})} \\
+    & \qquad (\<i_R|\cdot O\cdot |i_{R'}\>).(|i_{r_1}\>_{r_i}\otimes\cdots\otimes|i_{r_n}\>_{r_n} \otimes {}_{r'_1}\<i_{r'_1}|\otimes\cdots\otimes{}_{r'_{n'}}\<i_{r'_{n'}}|).
 \end{align*}
-}
-The syntax tree structure is also compatible with the datatype of Mathematica. This improves the interoperability between Dirace and the Mathematica system, enabling them to work interleavingly.
-To improve usability, Dirace also supports many special syntacies for terms,  and most Dirac notation terms will be encoded in the natural way.
-Here are some examples for the parsing syntax.
+The rules for $e_R$ (ket and bra) are similar. This step reduces all labelled terms $e_R$ or $e_{R;R}$.
+Other symbols on labelled Dirac notation are also reduced by rules like $(D_1 \cdot D_2)^\dagger \ \reduce\ D_2^\dagger \cdot D_1^\dagger$.
+The final step operates on sum and dot product. They will lift summation to the outside, and eliminate the bra-ket pairs whenever possible.
+\begin{align*}
+    & \textrm{(R-SUM-PUSHD0)}
+    && X_1 \otimes \cdots (\sum_{i \in M} D) \cdots \otimes X_2\ \reduce\ \sum_{i \in M} (X_1 \otimes \cdots D \cdots \otimes X_n) \\
+    % %
+    % & \textrm{(R-SUM-PUSHD1)}
+    % && (\sum_{i \in M} D_1) \cdot D_2 \ \reduce\ \sum_{i \in M} (D_1 \cdot D_2) \\
+    %
+    & \textrm{(R-L-SORT0)}
+    && A : \DType(s_1, s_2), B : \DType(s_1', s_2'), s_2 \cap s_1'=\emptyset \Rightarrow A \cdot B \ \reduce\ A \otimes B \\
+    %
+    & \textrm{(R-L-SORT1)}
+    && {}_r\bra{i}\cdot\ket{j}_r \ \reduce\ \delta_{i, j} \\
+    %
+    & \textrm{(R-L-SORT2)}
+    && {}_r\bra{i}\cdot(Y_1 \otimes \cdots \otimes \ket{j}_r \otimes \cdots \otimes Y_m) \ \reduce\ \delta_{i, j}.(Y_1  \otimes \cdots \otimes Y_m)
+    %
+\end{align*}
+% These rules are added to the rewriting system in~\Cref{sec: decide} and executed together.
+In the end, if there are no variables of $\DType(s_1, s_2)$, the expression will be reduced to the addition of big operator sum, and each sum body is labelled basis with Dirac notation scalar coefficients:
+\[
+    \sum_{i}\cdots\sum_{j} a_1 . (\ket{i}_{p} \otimes \cdots \otimes \bra{j}_{q})
+    + \cdots +
+    \sum_{k}\cdots\sum_{l} a_m . (\ket{k}_{r} \otimes \cdots \otimes \bra{l}_{s})
+\]
+In this stage, we only need to check the scalars to decide the equivalence of labelled Dirac notation.
 
-\begin{figure}
-    \center
-\begin{tabular}{c >{\centering\arraybackslash}p{4cm} l}
-    \hline
-    syntax & parsing result & explanation \\
-    \hline
-    \texttt{|e>} & \texttt{KET[e]} & the ket basis\\
-    \texttt{e1 + ... + en} & \texttt{ADD[e1, ..., en]} & the addition\\
-    \texttt{e1\ e2} & \texttt{COMPO[e1, e2]} & composition in Dirac notation \\
-    \texttt{e1\^{}*} & \texttt{CONJ[e1]} & scalar conjugation \\
-    \texttt{fun i : T => X} & \texttt{FUN[i, T, X]} & lambda abstraction \\
-    \hline
-\end{tabular}
-\end{figure}
-
-Finally, Dirace uses a prover to host the computation.
-The prover maintains a well-formed context $E[\Gamma]$, and processes commands to modify the context and conduct calculations.
-The commands are listed below.
-\begin{itemize}
-    \item \texttt{\textcolor{NavyBlue}{Def} ID := term.} It defines the \texttt{ID} as the \texttt{term}, using the \textbf{W-Def} typing rule.
-    \item \texttt{\textcolor{NavyBlue}{Var} ID := term.} It make an assumption of \texttt{ID} with the \texttt{term} as type, using the \textbf{W-AssumeE} typing rules.
-    \item \texttt{\textcolor{NavyBlue}{Check} term.} Type checking the \texttt{term} and output the result.
-    \item \texttt{\textcolor{NavyBlue}{Normalize} term.} Normalize the \texttt{term} using the algorithm introduced in~\Cref{sec: decide}.
-    \item \texttt{\textcolor{NavyBlue}{CheckEq} term \textcolor{NavyBlue}{with} term.} Check the equivalence of the two terms calculating and comparing their normal forms.
-\end{itemize}
-The prover will type check the terms for each command.
-We can also use \texttt{\textcolor{NavyBlue}{Normalize} term \textcolor{NavyBlue}{with trace}.} to output the proof trace during normalization.
-The proof trace is a sequence of records, including the rule or transformation appied, the position of application, and the pre- and post-transformation terms.
-The record helps understand the normalization procedure better, and can be turned into verified proofs in theorem provers in the future.
-
-\subsubsection{Use Case}
-Here we encode the motivating~\Cref{ex: motivating}, examine and explain how it is checked in Dirace. The encoding is shown below.
-
-    \begin{lstlisting}[style=dirace]
-Var T : INDEX. Var M : OTYPE[T, T].
-Def phi := idx T => Sum nv in USET[T], |(nv, nv)>.
-Var r1 : REG[T]. Var r2 : REG[T].
-CheckEq M_r1;r1 (phi T)_(r1, r2) with (TPO T T M)_r2;r2 (phi T)_(r1, r2).
-    \end{lstlisting}
-
-The first three lines use the \texttt{\textcolor{NavyBlue}{Var}} and \texttt{\textcolor{NavyBlue}{Def}} commands to set up the context for the Dirac notation.
-\texttt{T} is a type index, representing arbitrary Hilbert space types.
-\texttt{M} is assume to be an operator in the Hilbert space with type \texttt{T}.
-\texttt{phi} is defined as the maximally entangled state, depending on the bound variable \texttt{T} as index.
-\texttt{r1} and \texttt{r2} are register names for the two subsystems.
-
-In the left hand side of \texttt{\textcolor{NavyBlue}{CheckEq}} command, \texttt{M\_r1;r1} denotes the labelled notation $M_{r_1; r_1}$, and \texttt{(phi T)\_(r1, r2)} denotes the entangled state $\ket{\Phi}_{(r_1, r_2)}$.
-They are connected by a white space, which is parsed into the composition of Dirac notation, and will be reduced into the operator-ket multiplication after typing.
-The right hand side is interpreted similarly, except the defined symbol \texttt{TPO} in the environment:
-
-\begin{lstlisting}[style=dirace]
-Def TPO := idx sigma => idx tau => fun O : OTYPE[sigma, tau] => Sum i in USET[sigma], Sum j in USET[tau], (<i| O |j>).(|j> <i|).
-\end{lstlisting}
-
-The \texttt{TPO} symbol represents the transpose of operators, and encodes the formalization in~\Cref{ex: formalizing motivating}.
-Thanks to the design of environment and functions, many other commonly used symbols in Dirac notation are encoded and provided as defined symbols in Dirace.
-
-Within one second, the prover reports the result of equivalence with their common normal form:
-    \begin{lstlisting}[style=dirace]
-The two terms are equal.
-[Normalized Term] SUM[USET[T], FUN[BASIS[T], SUM[USET[T], FUN[BASIS[T], SCR[DOT[BRA[$\texttt{\$1}$], MULK[M, KET[$\texttt{\$0}$]]], LTSR[LKET[$\texttt{\$1}$, r1], LKET[$\texttt{\$0}$, r2]]]]]]] : DTYPE[RSET[r1, r2], RSET]
-    \end{lstlisting}
-
-The normal form is in the internal syntax tree format mentioned above. A more readable interpretation is:
-\[
-\sum_{\mathbf{U}(T)} \sum_{\mathbf{U}(T)} \bra{\$1}M\ket{\$ 0} . \ket{\$1}_{r_1} \otimes \ket{\$0}_{r_2} : \DType(\{r_1, r_2\}, \emptyset).
-\]
-Here $\$0$ and $\$1$ are de Bruijn indices. The result is a ket on the $\{r_1, r_2\}$ system as expected, and follows pattern proposed in~\Cref{sec: labelled}.
-
-\subsubsection{Benchmark performance}
-
-To evaluate Dirace, we first test the examples from DiracDec benchmark and make a comparison.
-The experiments are carried out using a MacBook Pro with M3 Max chip.
-
-\begin{figure}
-    \center
-    \begin{tabular}{c|c c c|c c c}
-        \hline
-        \multirow{2}{*}{source} & \multicolumn{3}{c|}{DiracDec} & \multicolumn{3}{c}{Dirace} \\
-        \cline{2-7}
-                                 & expressable & success & time(s)           & expressable & success & time(s)                 \\
-        \hline
-        textbook(QCQI)          & 18          & 18        &    1.02        &    18      & 18          &   0.82      \\
-        CoqQ                    & 162          & 156       &    48.69       &   158     &  158   &     9.74     \\
-        circuits                 & 2          & 2       &    17.67       &   3     &  2   &     1.4     \\
-        research paper                & 4          & 4         &  59.53       &   4    & 4       &  0.73     \\
-        \hline
-    \end{tabular}        
-    \caption{For DiracDec, examples that cannot be decided within 60 seconds are not included.}
-\end{figure}
-
-The timing of Dirace does not include the initialization of Mathematica kernel link, which takes about 3 seconds in the beginning.
-As to expressibility, the language for DiracDec has the support for projectors $\texttt{fst}$ and $\texttt{snd}$ on basis pairs, satisfying $\texttt{fst} (s, t) = s$ and $\texttt{snd} (s, t) = t$.
-We found this feature is almost not used, so we removed the support.
-As a result, Dirace encodes 158 examples for the CoqQ part, 4 less than DiracDec.
-For decidability, we improved the rewriting rules about sum for Dirace, and proved several more examples failed by DiracDec.
-The main difference is about their time efficiency.
-Because of our algorithm to decide AC-equivalence and SUM-SWAP, Dirace has a significant efficiency improvement, especially on those ``computational examples'' mentioned in the DiracDec paper.
-One typical example comes from the paper by Jens about the equivalence of operators for qubits.
-The system has to decompose the term on the concrete $\ket{0}$ and $\ket{1}$ basis, resulting a lot of addition elements.
-It takes DiracDec about one minute, but Dirace solves it within one second.
-
-We also built an example benchmark for labelled Dirac notation, see~\Cref{sec: examples for labelled}.
-
-\section{Related Work}
-
-Automated theorem proving has seen significant advancements in recent years, particularly through the development of satisfiability modulo theories (SMT) solvers,
-% SMT solvers extend propositional satisfiability by integrating theories such as arithmetic, arrays, and bit-vectors, making them suitable for solving a wide range of verification problems in both hardware and software domains.
-including prominent tools like Z3. These solvers have become essential in various fields such as formal verification, synthesis, and model checking. 
-Equational reasoning is another crucial area of research within automated theorem proving, which focuses on solving problems that involve equations between terms in an algebraic structure. 
-Equational provers, such as Vampire and E, have played a pivotal role in addressing the challenge of proving equations in first-order logic, employing sophisticated algorithms like superposition and term rewriting.
-
-Formal verification of quantum computation is receiving increasing attentions during these years. See~\cite{Lewis2023} for a comprehensive review. 
-Verification frameworks in Coq include the foundational formalization CoqQ and quantum circuit language QWIRE. 
-Verifications of quantum programs are also considered, such as the Hoare logic based methods [] and model checking based methods [].
-The equational reasoning of Dirac notation is crucial for establishing property proofs in these works.
-Verification theoreis and tools based on other languages are also proposed, such as PyZX for reasoning and simplification of ZX-calculus.
-
-\section{Conclusion and Future Work}
-Based on the first equational reasoning tool for Dirac notation called DiracDec, this work improves and extends the theory for practical applications, and provides the solver Dirace.
-Experiments show that the tool demonstrates advantages in decidability, efficiency and usability.
-
-We expect Dirace to have applications in areas like quantum program verification or proofs of post-quantum cryptography protocols in the future.
-One promising following up is to connect Dirace with theorem provers like Coq.
-It involves transforming theorem prover expressions into Dirace, and verify the proof trace of Dirace in theorem provers.
-Besids, most quantum program verifiers nowadays depend on matrix calculations.
-Dirace can serve as the replacement for matrix methods to enable symbolic deductions.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% \begin{credits}
-%     \subsubsection{\ackname} A bold run-in heading in small font size at the end of the paper is
-%     used for general acknowledgments, for example: This study was funded
-%     by X (grant number Y).
-% \end{credits}
-
-%
-% ---- Bibliography ----
-%
-% BibTeX users should specify bibliography style 'splncs04'.
-% References will then be sorted and formatted in the correct style.
-%
-\bibliographystyle{splncs04}
-\bibliography{ref}
-%
-
-\appendix
-\include{appendix}
-
-\end{document}
+% \begin{lemma}[normal form]
+%     \label{lem: labelled normal form}
+%     For a well-typed term $e$ in $E[\Gamma]$, if $e$ does not contain variables of $\DType(s_1, s_2)$ type, the normal form of $e$ will be
+%     \[
+%     \sum_{i}\cdots\sum_{j} a_1 . (\ket{i}_{p} \otimes \cdots \otimes \bra{j}_{q})
+%     + \cdots +
+%     \sum_{k}\cdots\sum_{l} a_m . (\ket{k}_{r} \otimes \cdots \otimes \bra{l}_{s})
+%     \]
+%     Where $a_i$ are scalar typed Dirac notation.
+% \end{lemma}
+% \begin{proof}
+%     Labelled Dirac notation of $e_R$, $e_R;R$, 
+% \end{proof}

### 2025-01-27 03tool_eval.tex
splited

--- 2025-01-27 main.tex
@@
-% This is samplepaper.tex, a sample chapter demonstrating the
-% LLNCS macro package for Springer Computer Science proceedings;
-% Version 2.21 of 2022/01/12
-%
-\documentclass[runningheads]{llncs}
-
-% \usepackage[utf8]{inputenc}
-% \usepackage[margin=0.8in]{geometry}
-
-\usepackage{color}
-\usepackage{amsmath}
-%\usepackage{amssymb}
-\usepackage{graphicx}
-% \usepackage{amsthm}
-\usepackage{stmaryrd}
-\usepackage[all]{xy}
-\usepackage{multirow}
-\usepackage{paralist}
-\usepackage{hhline}
-\usepackage{bm}
-\usepackage{longtable}
-\usepackage{makecell}
-\usepackage{mdframed}
-
-\usepackage{wasysym}
-\usepackage{extarrows}
-\usepackage{tikz}
-
-% add new commands for comments here
-\newcommand{\yx}[1]{\textit{\color{blue}[YX] : #1}}
-\newcommand{\lz}[1]{\textit{\color{red}[LZ] : #1}}
-
-\newcommand{\modify}[1]{{\color{red}#1}}
-
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{stmaryrd}
-\usepackage{hyperref}
-
-\usepackage{braket}
-\usepackage{cleveref}
-
-\input{lstlisting}
-
-\usepackage{tikz}
-\usetikzlibrary{shapes.geometric, arrows}
-\usetikzlibrary{trees}
-\tikzstyle{arrow} = [thick,->,>=stealth]
-
-\newenvironment{ruletable}[1]
-{
-    \begin{longtable}{cl}
-    \caption{#1}\\
-    \hline
-    \textbf{Rule} & \textbf{Description} \\
-    \hline
-    \endfirsthead
-
-    \hline
-    \textbf{Rule} & \textbf{Description} \\
-    \hline
-    \endhead
-
-    % \hline
-    % \multicolumn{2}{r}{\textit{Continued on the next page}} \\
-    \hline
-    \endfoot
-
-    \hline
-    \endlastfoot
-}
-{
-    \end{longtable}
-}
-
-% define C++ logo
-\newcommand{\CC}{C\nolinebreak\hspace{-.05em}\raisebox{.4ex}{\tiny\bf +}\nolinebreak\hspace{-.10em}\raisebox{.4ex}{\tiny\bf +}}
-\def\CC{{C\nolinebreak[4]\hspace{-.05em}\raisebox{.4ex}{\tiny\bf ++}}}
-
-%
-\usepackage[T1]{fontenc}
-% T1 fonts will be used to generate the final print and online PDFs,
-% so please use T1 fonts in your manuscript whenever possible.
-% Other font encondings may result in incorrect characters.
-%
-\usepackage{graphicx}
-% Used for displaying a sample figure. If possible, figure files should
-% be included in EPS format.
-%
-% If you use the hyperref package, please uncomment the following two lines
-% to display URLs in blue roman font according to Springer's eBook style:
-\usepackage{color}
-\renewcommand\UrlFont{\color{blue}\rmfamily}
-\urlstyle{rm}
-%
-\begin{document}
-%
-\title{Dirace: Practical Proof Automation of Dirac Notation Equations}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% %
-% %\titlerunning{Abbreviated paper title}
-% % If the paper title is too long for the running head, you can set
-% % an abbreviated paper title here
-% %
-% \author{First Author\inst{1}\orcidID{0000-1111-2222-3333} \and
-% Second Author\inst{2,3}\orcidID{1111-2222-3333-4444} \and
-% Third Author\inst{3}\orcidID{2222--3333-4444-5555}}
-% %
-% \authorrunning{F. Author et al.}
-% % First names are abbreviated in the running head.
-% % If there are more than two authors, 'et al.' is used.
-% %
-% \institute{Princeton University, Princeton NJ 08544, USA \and
-% Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
-% \email{lncs@springer.com}\\
-% \url{http://www.springer.com/gp/computer-science/lncs} \and
-% ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
-% \email{\{abc,lncs\}@uni-heidelberg.de}}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%
-\maketitle              % typeset the header of the contribution
-%
-\begin{abstract}
-Dirac notation is a fundamental language in quantum computation and quantum information.
-Recently, the term rewriting system DiracDec~\cite{diracdec} was introduced to automate equational reasoning with Dirac notation, a critical yet time-intensive task.
-Building upon DiracDec, this work aims to develop a solver optimized for practical applications.
-Enhancements include an improved typing system, a simplified language and rewriting system, more efficient algorithms, and added support for labeled Dirac notation.
-Our solver, Dirace, is implemented in \CC\ with a Mathematica backend, demonstrating significant improvements in decision-making power and time efficiency.
-
-% \keywords{First keyword  \and Second keyword \and Another keyword.}
-\end{abstract}
-%
-%
-%
-
-%%%%%%%%%%%%%%%%
-\newcommand*{\sem}[1]{{\llbracket #1 \rrbracket}}
-\newcommand{\DiracDec}{\textsf{DiracDec}}
-
-\newcommand{\reduce}{\triangleright}
-
-\newcommand{\Sort}{\mathsf{Sort}}
-\newcommand{\WF}{\mathcal{WF}}
-
-\newcommand{\Index}{\mathsf{Index}}
-\newcommand{\Type}{\mathsf{Type}}
-\newcommand{\Basis}{\mathsf{Basis}}
-
-\newcommand{\SType}{\mathcal{S}}
-\newcommand{\KType}{\mathcal{K}}
-\newcommand{\BType}{\mathcal{B}}
-\newcommand{\OType}{\mathcal{O}}
-\newcommand{\SET}{\mathsf{Set}}
-
-\newcommand{\ZEROK}{\mathbf{0}_\mathcal{K}}
-\newcommand{\ZEROB}{\mathbf{0}_\mathcal{B}}
-\newcommand{\ZEROO}{\mathbf{0}_\mathcal{O}}
-
-\newcommand{\PAIR}{\mathsf{PAIR}}
-
-\newcommand{\ZERO}{\mathsf{0}}
-\newcommand{\ONE}{\mathsf{1}}
-\newcommand{\ADDS}{\mathsf{ADDS}}
-\newcommand{\ADD}{\mathsf{ADD}}
-\newcommand{\MULS}{\mathsf{MULS}}
-\newcommand{\MUL}{\mathsf{MUL}}
-\newcommand{\CONJ}{\mathsf{CONJ}}
-\newcommand{\CJG}{\mathsf{CJG}}
-\newcommand{\ADJ}{\mathsf{ADJ}}
-\newcommand{\DELTA}{\mathsf{DELTA}}
-\newcommand{\DOT}{\mathsf{DOT}}
-\newcommand{\SCR}{\mathsf{SCR}}
-\newcommand{\TSR}{\mathsf{TSR}}
-\newcommand{\KET}{\mathsf{KET}}
-\newcommand{\BRA}{\mathsf{BRA}}
-\newcommand{\ONEO}{\mathbf{1}_\mathcal{O}}
-\newcommand{\OUTER}{\mathsf{OUTER}}
-\newcommand{\MULK}{\mathsf{MULK}}
-\newcommand{\MULB}{\mathsf{MULB}}
-\newcommand{\MULO}{\mathsf{MULO}}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Introduction}
-
-In 1939, Dirac introduced his notation for quantum mechanics~\cite{dirac1939new}, designed to represent linear algebraic formulas in a compact and convenient form.
-For example, the expression \( a\ket{\psi} + b\ket{\phi} \) represents the superposition of two quantum states, \( \ket{\psi} \) and \( \ket{\phi} \).
-Today, Dirac notation is widely accepted as the standard language in quantum computation and quantum information.
-Its reasoning forms the foundation of research and applications, much like boolean and integer logic do in classical computer science.
-
-In quantum algorithm formalizations and quantum programming languages, Dirac notation is frequently used in equational proofs, which are critical, repetitive, and often time-intensive.
-These notations also play a key role in defining program states, operations, and assertions in quantum programming languages.
-To automate the verification of these programs, we need tools that can simplify and check the equivalence of preconditions.
-However, unlike the well-established SAT and SMT solvers for classical logic, a practical solver for Dirac notation equivalence remains an unmet need, creating a barrier for progress in several fields.
-
-Recently, Xu et al.~\cite{diracdec} proposed a theory for deciding the equivalence of Dirac notation, alongside a prototype implementation in Mathematica called DiracDec.
-They demonstrated that the equivalence of basic Dirac notation is decidable.
-Their algorithm, based on a pure term rewriting system, has been proven to be confluent and terminating.
-Despite this, there remains a gap between DiracDec and a practical solver for Dirac notation equivalence.
-
-One challenge is the efficiency of the algorithm when dealing with equivalences beyond the scope of term rewriting.
-DiracDec decides the entire equational theory by rewriting modulo \( E \), where \( E \) represents a set of axioms that cannot be decided by normalization alone, such as the associativity and commutativity (AC) of certain function symbols.
-DiracDec uses a direct but inefficient algorithm to decide these axioms, which searches through all possible permutations and exhibits factorial complexity.
-This inefficiency becomes particularly evident in "computational examples" containing many AC symbols, which are time-consuming to process.
-
-Usability is another area where DiracDec falls short.
-It does not support labelled Dirac notation, which uses registers to denote subsystems and express states locally.
-Additionally, DiracDec's typing system only does not provide context for variable typing assumptions or the definition of symbols, which are required in practical scenarios.
-To avoid type checking during term rewriting, DiracDec separates symbols (e.g., multiplication) for different types, leading to unnecessary complexity.
-Moreover, integrating the Mathematica implementation into other projects as a solver is challenging.
-
-The system design of DiracDec reflects a trade-off between simplicity and efficiency.
-While the simplicity of term rewriting allows for strong theoretical results, it limits optimization opportunities.
-Building on DiracDec, this work aims to develop a practical solver.
-We transform the term rewriting system into a hybrid algorithm, overcoming the challenges mentioned above.
-Our main technical contributions are:
-\begin{itemize}
-    \item An efficient algorithm for deciding the equational theories in \( E \), based on equivalence checking through normalization, with the normal form for \( E \) being obtained via sorting.
-    \item Support for constant register labels, and reducing the equivalence decision of labelled Dirac notation to the unlabelled case.
-    \item A more user-friendly \CC\ solver, Dirace, featuring an abstract language and typing system. We also support the definition of symbols (e.g., transpose and trace) using function syntax.
-\end{itemize}
-
-We evaluated Dirace against the DiracDec benchmark and new examples involving labelled Dirac notation.
-The results show significant improvements in both decidability and efficiency compared to DiracDec.
-
-% Dirace successfully decides all the examples that are expressable in its language, including those failed by DiracDec because of complexity or insufficient decision power.
-
-\section{Motivation and Preliminary}
-
-An interesting property of quantum mechanics is that for two maximally entangled states, applying a quantum operator \( M \) to one subsystem is equivalent to applying \( M^T \) (the transpose of \( M \)) to the other subsystem.
-This relationship holds regardless of the spatial separation between the two systems, and it can be expressed as an equation in Dirac notation.
-\begin{example}
-    \label{ex: motivating}
-    Let \( q \) and \( r \) represent two quantum systems in the Hilbert space \( \mathcal{H}_T \).
-Let \( M \) be a quantum operation acting on \( \mathcal{H}_T \), and let \( \ket{\Phi} = \sum_{i \in T} \ket{i} \otimes \ket{i} \) be the maximally entangled state.
-Then, we have the following equation:
-    \[
-    M_q \ket{\Phi}_{q; r} = M_r^T \ket{\Phi}_{q; r}.
-    \]
-\end{example}
-
-In this equation, \( q \) and \( r \) are labels denoting the respective subsystems in the Dirac notation.
-To automate reasoning about such equations, we must formalize the language and develop a proof system to handle it.
-
-\subsubsection{Dirac Notation}
-
-Quantum states are represented as vectors in complex Hilbert spaces, and operations on these states are described by linear transformations.
-Dirac notation uses the ket \( \ket{i} \) and the bra \( \bra{i} \) to denote basis vectors in a Hilbert space and its dual space, respectively.
-These symbols can be composed together in various ways to form more complex expressions.
-The meaning of these compositions depends on the types of the operands involved.
-For example, the inner product \( \braket{i|j} \) represents the scalar result of the dot product between \( \bra{i} \) and \( \ket{j} \), while the outer product \( \ket{i}\bra{j} \) represents an operator.
-Additionally, Dirac notation uses the tensor product symbol \( \otimes \) to describe the combined space of multiple quantum systems.
-
-One of the key features of Dirac notation is its order-independent interpretation.
-This means that the composition of terms can be written without parentheses, as the interpretation is unaffected by the order of multiplication.
-For instance, the formula \( \bra{i}\ket{\phi}\bra{\psi}\ket{j} \) can be understood as:
-\[
-    \braket{i|\phi}\braket{\psi|j} = \bra{i} (\ket{\phi}\bra{\psi}) \ket{j},
-\]
-and these expressions are equivalent for all variables involved.
-
-In practice, Dirac notation is often combined with other syntactic elements, such as the summation symbol \( \sum_{i \in S} A \), to enhance expressiveness.
-Furthermore, labelled Dirac notation, such as \( \ket{i}_q \otimes \ket{i}_r \), is used to denote quantum systems in consideration, where subscripts (registers) are added to distinguish different subsystems.
-
-\subsubsection{Universal Algebra}
-We use universal algebra and equational logic to formally represent Dirac notation and the reasoning procedure.
-A universal algebra defines a signature of function symbols, with terms constructed from constants, variables, and function applications.
-Other basic concepts like substitution of variables or pattern matching are also defined.
-In our case of Dirac notation, the signature consists of constructors and operations like $\ket{i}$ or $A \otimes B$.
-The reasoning process is guided by equational logic, which defines an equivalence relation that is compatible with substitution and term construction.
-This relation formalizes the intuitive concept of equivalence in algebra.
-
-\section{Language, Typing and Semantics}
-The first step is to formalize the language for Dirac notation.
-In DiracDec, the language has a concrete design, where the same syntax for different types corresponds to distinct symbols.
-Our goal is to transition from this concrete design to a more abstract one, which aligns more closely with the conventional Dirac notation we use and simplifies the term rewriting system.
-
-To achieve this, we organize our language into three layers: the index, the type, and the term.
-Terms represent concrete instances such as kets, bras, and operators, which will be typed and checked.
-The index represents classical data types and appears in type expressions to differentiate between various Hilbert spaces and sets.
-
-\begin{definition}[Index Syntax]
-    The syntax for type indices is:
-    \begin{align*}
-        \sigma ::=\ & x \mid \sigma_1 \times \sigma_2.
-    \end{align*}
-\end{definition}
-Here, \( x \) is a variable, and \( \sigma_1 \times \sigma_2 \) represents the product type for tensor product spaces or Cartesian product sets.
-
-\begin{definition}[Type Syntax]
-    The syntax for Dirac notation types is:
-    \begin{align*}
-        T ::=\ & \Basis(\sigma) \mid \SType \mid \KType(\sigma) \mid \BType(\sigma) \mid \OType(\sigma_1, \sigma_2) \mid T_1 \to T_2 \mid \forall x.T \mid \SET(\sigma).
-    \end{align*}
-\end{definition}
-\( \Basis(\sigma) \) denotes the type for basis elements in the index \( \sigma \).
-\( \SType \) represents scalars, while \( \KType(\sigma) \) and \( \BType(\sigma) \) refer to ket and bra types in the Hilbert space \( \sigma \), respectively.
-\( \OType(\sigma_1, \sigma_2) \) represents linear operators with \( \sigma_2 \) as the domain and \( \sigma_1 \) as the codomain.
-\( \SET(\sigma) \) refers to the type of subsets of \( \sigma \), used to denote the values of bound variables in summations.
-The remaining two constructs define function types: \( T_1 \to T_2 \) represents normal functions that take a \( T_1 \)-type argument and return a \( T_2 \)-type term, while \( \forall x.
-T \) represents index functions that take an index argument \( x \) and produce a \( T \)-type term, which may depend on \( x \).
-Index functions are essential for defining polymorphic transformations across Hilbert spaces.
-
-\begin{definition}[Term Syntax]
-    The syntax for Dirac notation terms is:
-    \begin{align*}
-        e ::=\ & x \mid \lambda x : T.e \mid \mu x.e \mid e_1\ e_2 \mid (e_1, e_2) \\
-        & |\ 0 \mid 1 \mid \ADDS(e_1, \cdots, e_n) \mid e_1 \times \cdots \times e_n \mid e^* \mid \delta_{e_1, e_2} \mid \DOT(e_1, e_2) \\
-        & |\ \ZEROK(\sigma) \mid \ZEROB(\sigma) \mid \ZEROO(\sigma_1, \sigma_2) \mid \ONEO(\sigma) \\
-        & |\ \ket{e} \mid \bra{t} \mid e^\dagger \mid e_1.e_2 \mid \ADD(e_1, \cdots, e_n) \mid e_1 \otimes e_2 \\
-        & |\ \MULK(e_1, e_2) \mid \MULB(e_1, e_2) \mid \OUTER(e_1, e_2) \mid \MULO(e_1, e_2) \\
-        & |\ \mathbf{U}(\sigma) \mid e_1 \star e_2 \mid \sum_{e_1} e_2.
-    \end{align*}
-\end{definition}
-The terms above are explained in order:
-\( \lambda x : T.e \) represents the abstraction for normal functions, and \( \mu x.e \) represents the abstraction for index functions.
-\( e_1\ e_2 \) denotes function application.
-\( (e_1, e_2) \) is the basis pair for product types.
-The constants \( 0 \), \( 1 \), \( \ADDS \), \( e_1 \times \cdots \times e_n \), and \( e^* \) are symbols for scalars.
-The next line includes constant symbols for kets, bras, and operators.
-\( \ket{e} \) is a ket, \( \bra{t} \) is a bra, and \( e^\dagger \) denotes the conjugate transpose of \( e \).
-\( e_1.e_2 \) represents scaling the term \( e_2 \) by scalar \( e_1 \).
-\( \mathbf{U}(\sigma) \) denotes the universal set with index \( \sigma \).
-\( e_1 \star e_2 \) represents the Cartesian product of \( e_1 \) and \( e_2 \).
-\( \sum_{e_1} e_2 \) is the big operator sum, modeled by folding the function \( e_2 \) over the value set \( e_1 \).
-Typically, the sum's body is given by an abstraction.
-For convenience, we also use the notation \( \sum_{x \in s} X \) to represent \( \sum_{s} \lambda x : T .
-X \).
-Additionally, \( \ADDS \) and \( \ADD \) are distinct AC symbols: \( \ADDS \) is used for scalar addition and \( \ADD \) for linear algebra addition.
-There are five kinds of linear algebraic multiplications among kets, bras, and operators.
-These are encoded using five different symbols: \( \DOT \), \( \MULK \), \( \MULB \), \( \OUTER \), and \( \MULO \).
-
-Compared to DiracDec, the symbols for addition, adjoint, scaling, and tensor products have been consolidated.
-The multiplications are still kept separate due to their differing properties.
-These operations are denoted as \( B \cdot K \), \( K_1 \cdot K_2 \), \( B_1 \cdot B_2 \), \( K \cdot B \), and \( O_1 \cdot O_2 \), respectively.
-
-\subsection{Typing System}
-
-The typing system is responsible for classifying terms within a proof system, using a context that specifies the types of variables and definitions.
-The context is divided into two components: the environment \( E \), which includes definitions and assumptions, and the context \( \Gamma \), which tracks bound variables.
-Both \( E \) and \( \Gamma \) consist of sequences of assumptions \( x : T \) or definitions \( x := t : T \).
-\begin{definition}[Environment and Context]
-    The syntax for \( E \) and \( \Gamma \) is:
-    \begin{align*}
-        E ::= &\ [] \mid E; x : \Index \mid E; x : T \mid E; x := t : T. \\
-        \Gamma ::= &\ [] \mid \Gamma; x : \Index \mid \Gamma; x : T.
-    \end{align*}
-\end{definition}
-Definitions refer to symbols that can be expanded or unfolded, and typically represent abstract concepts, such as transpose or trace in Dirac notation.
-Assumptions, on the other hand, define the types of variables, and variable assumptions are implicitly universally quantified in the case of equational proofs.
-
-Type checking in our language involves maintaining a well-formed environment and context \( E[\Gamma] \).
-We say an expression \( t \) has type \( X \) in context \( E[\Gamma] \) if the typing judgment \( E[\Gamma] \vdash t : X \) can be proven through the rules in~\Cref{sec: full typing rules}.
-The well-formedness of the context \( \WF(E)[\Gamma] \) is built incrementally:
-\[
-    \frac{}{\WF([])[]}
-    \qquad
-    \frac{\WF(E)[] \qquad x \notin E}{\WF(E; x : \Index)[]}
-    \qquad
-    \frac{E[]\vdash t:T \qquad x \notin E}{\WF(E; x := t:T)[]}.
-\]
-Starting with an empty context, new index symbols can be introduced, and symbols with checked types can be defined.
-Typing judgments are then made inductively using the information from \( E[\Gamma] \).
-For instance, the rule \( x : \Index \in E[\Gamma] \) holds if \( E \) or \( \Gamma \) contains an assumption for \( x \).
-And \(\KType(\sigma)\) is a valid type for kets, if the argument $\sigma$ is typed as an index.
-\begin{gather*}
-    \frac{\WF(E)[\Gamma] \qquad x : \Index \in E[\Gamma]}{E[\Gamma] \vdash x : \Index}
-    \qquad
-    \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \KType(\sigma) : \Type}
-\end{gather*}
-
-Terms are then typed accordingly.
-For example, the ket \( \ket{t} \) will have the type \( \KType(\sigma) \) if \( t \) is a basis term of index \( \sigma \).
-Similarly, the inner product between a bra and a ket of the same index \( \sigma \) is typed as a scalar.
-It corresponds to the constraint of inner product that vectors should be from the same Hilbert space.
-\begin{gather*}
-    \frac{E[\Gamma]\vdash t : \Basis(\sigma)}{E[\Gamma] \vdash \ket{t} : \KType(\sigma)}
-    \qquad
-    \frac{E[\Gamma]\vdash B : \BType(\sigma) \qquad E[\Gamma]\vdash K : \KType(\sigma)}{E[\Gamma] \vdash B \cdot K : \SType}
-\end{gather*}
-
-The typing for functions and applications follows the standard practice.
-For example, the index function \( \mu x.
-t \) is typed with \( x \) as a bound variable of type \( \Index \), and the application \( (t\ u) \) has the type \( U\{x/u\} \), obtained by replacing \( x \) with the index instance \( u \).
-\begin{gather*}
-    \frac{E[\Gamma; x : T] \vdash t : U}{E[\Gamma] \vdash (\lambda x : T . t) : T \to U}
-    \qquad
-    \frac{E[\Gamma] \vdash t:U \to T \qquad E[\Gamma] \vdash u:U}{E[\Gamma] \vdash (t\ u):T} 
-    \\[1em]
-    \frac{E[\Gamma; x : \Index] \vdash t : U}{E[\Gamma] \vdash (\mu x.t) : \forall x.U}
-    \qquad
-    \frac{E[\Gamma] \vdash t:\forall x.U \qquad E[\Gamma] \vdash u : \Index}{E[\Gamma] \vdash (t\ u):U\{x/u\}}
-\end{gather*}
-
-Finally, the big operator sum is modeled by folding a function over a set, with the typing rule as follows:
-\[
-    \frac{E[\Gamma] \vdash s : \SET(\sigma) \qquad E[\Gamma] \vdash f : \Basis(\sigma) \to \KType(\tau)}{E[\Gamma] \vdash \sum_{s} f : \KType(\tau)}.
-\]
-
-\begin{lemma}
-    The typing of expressions is both decidable and unique.
-\end{lemma}
-
-\begin{proof}
-    The type of an expression can be determined recursively. For any given function symbol and argument types, there is at most one typing rule, ensuring the uniqueness of typing.
-\end{proof}
-
-\subsection{Semantics}
-
-The semantics of a language define the meaning of its expressions.
-In this context, the objective of our algorithm is to determine whether two expressions are semantically equivalent.
-We define the semantics in a denotational manner, mapping syntax to set-theoretic objects.
-
-\subsubsection{Denotational Semantics}
-The denotational semantics interpret every expression as an object in linear algebra, according to a valuation mapping \( v \), which assigns values to the variables in the expression.
-The semantics of an expression \( e \) with a given valuation \( v \) is denoted as \( \sem{e}_v \).
-Two expressions \( e_1 \) and \( e_2 \) are considered equivalent if their semantics are equal for all valuations, i.e., \( \sem{e_1}_v = \sem{e_2}_v \) for all \( v \).
-
-The complete interpretation of terms and types is provided in~\Cref{sec: full denotational sem}.
-Variables typed with \( \Index \) are interpreted as finite sets, and the product of two indices \( \sem{\sigma_1 \times \sigma_2} \) is defined as the Cartesian product of the sets \( \sem{\sigma_1} \) and \( \sem{\sigma_2} \).
-More generally, each type is interpreted as a set.
-For example, the scalar type \( \sem{\SType} \) is interpreted as the set of complex numbers \( \mathbb{C} \), and the ket and bra types \( \sem{\KType(\sigma)} \) and \( \sem{\BType(\sigma)} \) are interpreted as the Hilbert space \( \mathcal{H}_{\sem{\sigma}} \) and its dual \( \mathcal{H}_{\sem{\sigma}}^* \), respectively.
-Terms are explained as the set elements.
-For example, the semantics of ket tensor product $\sem{K_1 \otimes K_2} \equiv \sem{K_1} \otimes \sem{K_2}$, is obtained by first calculating the semantics $\sem{K_1}$ and $\sem{K_2}$ as vectors, and then take the vector tensor product as result.
-
-% One special case is the delta function \( \delta_{s,t} \), which is interpreted as:
-% \[
-%     \sem{\delta_{s,t}} =
-%     \begin{cases}
-%         1, & \text{if } \sem{s} = \sem{t}, \\
-%         0, & \text{if } \sem{s} \neq \sem{t}.
-%     \end{cases}
-% \]d
-
-The idea behind the interpretation of types and terms is to formalize the typing relation using set-theoretic inclusion.
-Specifically, for a well-formed context \( E[\Gamma] \), term \( t \), and type \( T \), if \( E[\Gamma] \vdash t : T \), then for any valuation \( v \), the semantics of \( t \) must lie within the semantics of \( T \).
-
-\begin{lemma}
-    For any well-formed context \( E[\Gamma] \), term \( t \), and type \( T \), if \( E[\Gamma] \vdash t : T \), then for all valuations \( v \), \( \sem{t}_v \in \sem{T}_v \).
-\end{lemma}
-
-\begin{proof}
-    The proof follows directly by checking each case.
-\end{proof}
-
-This interpretation formalizes the standard understanding of Dirac notation and provides the foundation for the algorithm.
-However, computers cannot directly reason about equivalence through mathematical interpretations.
-Therefore, we proceed by defining a proof system that abstracts these concepts.
-
-\subsubsection{Axiomatic semantics}
-
-The proof system for equivalence is based on equational logic, together with axioms that describe the properties of Dirac notation.
-A full list of these axioms can be found in~\Cref{sec: full axioms}.
-The axioms cover fundamental aspects of linear spaces, as well as other structures like the tensor and inner products.
-For example, we have the absorption law for zero symbols:
-\(X \cdot \mathbf{0} = \mathbf{0},\)
-and the bilinearity of the dot product:
-\[
-(a.X) \cdot Y = a \cdot (X \cdot Y), \quad X \cdot (Y_1 + Y_2) = X \cdot Y_1 + X \cdot Y_2, 
-\]
-\[
-X \cdot (a.Y) = a \cdot (X \cdot Y), \quad (X_1 + X_2) \cdot Y = X_1 \cdot Y + X_2 \cdot Y.
-\]
-
-As mentioned in the introduction, a subset \( E \) of these axioms cannot be decided by term rewriting alone. The axioms in \( E \) include:
-\[
-\begin{aligned}
-    \text{AC-equivalence} &\quad \text{e.g.,} \quad X + Y = Y + X, \quad (X + Y) + Z = X + (Y + Z), \\
-    \alpha\text{-equivalence} &\quad \lambda x . A = \lambda y . A\{x/y\}, \\
-    \text{SUM-SWAP} &\quad \sum_{i \in s_1} \sum_{j \in s_2} A = \sum_{j \in s_2} \sum_{i \in s_1} A, \\
-    \text{scalar theories} &\quad \text{e.g.,} \quad a + 0 = a, \quad a \times (b + c) = a \times b + a \times c.
-\end{aligned}
-\]
-The scalar theories are treated separately as a module and are not considered in this work. In the implementation, we use the Mathematica kernel to decide scalar equivalences.
-These equational axioms provide an operable theory for the proof automation algorithm.
-Denotational semantics can be seen as one model for this theory, meaning that equivalences derived from the axioms always imply equivalence in the interpretations.
-\begin{lemma}
-    For all well-formed contexts \( E[\Gamma] \) and terms \( e_1, e_2 \), if \( \vdash e_1 = e_2 \), then \( \sem{e_1} = \sem{e_2} \).
-\end{lemma}
-\begin{proof}
-    This follows directly from checking all cases.
-\end{proof}
-
-Thus, the axioms are sound with respect to denotational semantics.
-However, the reverse does not hold: there exist equivalences in denotational semantics that cannot be captured by these axioms.
-Nevertheless, our work focuses on solving practical examples, which are fully covered by the axioms in the experiments.
-
-To conclude this section, we demonstrate the formalization of the symbols used in the motivating example.
-\begin{example}[Formalizing the Motivating Example]
-    \label{ex: formalizing motivating}
-    Definitions and assumptions in the environment \( E \) are formalized as follows:
-    \begin{align*}
-        & \text{TPO} && := \mu T_1. \mu T_2. \lambda O : \OType(T_1, T_2). \sum_{i \in \mathbf{U}(T_1)} \sum_{j \in \mathbf{U}(T_2)} \bra{i} O \ket{j} . \ket{j}\bra{i} \\
-        & &&\quad : \forall T_1. \forall T_2. \OType(T_1, T_2) \to \OType(T_2, T_1); \\
-        &\text{phi} &&:= \mu T. \sum_{i \in \mathbf{U}(T)} \sum_{j \in \mathbf{U}(T)} \ket{(i, j)} : \forall T.\KType(T \times T); \\
-        & T && : \Index; \\
-        & M && : \OType(T, T).
-    \end{align*}
-    The symbol \( \text{TPO} \) represents the transpose of an operator, polymorphic on the Hilbert spaces \( T_1 \) and \( T_2 \).
-The symbol \( \text{phi} \) takes the index \( T \) and defines the maximally entangled states, summing over all basis elements in \( T \), as indicated by the universal set \( \mathbf{U}(T) \).
-    With the assumption of the index \( T \) and operator \( M \), we can express the equivalence in the non-labelled version as:
-    \[
-    (\textrm{M} \otimes \mathbf{1}_\mathcal{O}(\textrm{T})) \cdot (\textrm{phi T}) = (\mathbf{1}_\mathcal{O}(\textrm{T}) \otimes (\textrm{TPO T T M})) \cdot (\textrm{phi T}).
-    \]
-\end{example}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Algorithm for Deciding Dirac Notation Equations}
-\label{sec: decide}
-
-We utilize the term rewriting technique to decide the equivalence of Dirac notation, based on its axiomatic semantics.
-The process involves normalizing two terms and checking whether they have identical syntax.
-This normalization is achieved through matching and substitution in universal algebra, using a set of predefined rewriting rules.
-
-As mentioned in the semantics section, the axioms in \( E \) cannot be fully decided through term rewriting.
-DiracDec decides \( E \) by examining all possible permutations in the rewriting result.
-In this work, we introduce a more efficient approach by incorporating sorting algorithms into the normalization procedure.
-
-The normalization procedure consists of the following steps.
-\begin{enumerate}
-    \item \textbf{First Rewritings}: Expand definitions and simplify expressions.
-    \item \textbf{Variable Expansion}: Consider scalar expressions to ensure completeness.
-    \item \textbf{Second Rewritings}: Normalize terms modulo \( E \).
-    \item \textbf{Sorting Without Bound Variables}: Normalize AC-equivalence.
-    \item \textbf{Swap Successive Summations}: Normalize SUM-SWAP equivalences.
-    \item \textbf{De Bruijn Normalization}: Normalize \( \alpha \)-equivalence.
-\end{enumerate}
-
-\subsection{Normalization Modulo \( E \) by Term Rewriting}
-
-Term rewriting rules, represented as \( l \ \reduce\ r \), are applied recursively to normalize terms.
-In each step, subterms matching the left-hand side \( l \) of a rule are replaced with the corresponding right-hand side \( r \).
-The procedure terminates when no further rewritings can be made.
-A comprehensive list of rewriting rules can be found in~\Cref{sec: rewriting rules}. Below are some key examples to illustrate the design:
-
-One of our optimizations is using functions with indefinite arities. 
-Therefore, we use a flattening rule to handle associativity with AC symbols:
-\[
-a_1 + \cdots + (b_1 + \cdots + b_m) + \cdots + a_n \ \reduce\ a_1 + \cdots + b_1 + \cdots + b_m + \cdots + a_n.
-\]
-Commutativity is handled later in the sorting step. Many of the rewriting rules are directly derived from the equational axioms, such as:
-\begin{align*}
-    & \textrm{(R-DOT6)} && \bra{s} \cdot \ket{t} \ \reduce\  \delta_{s, t}, \\
-    & \textrm{(R-DELTA0)} && \delta_{s, s} \ \reduce\  1, \\
-    & \textrm{(R-MULK1)} && O : \OType(\sigma, \tau) \Rightarrow O \cdot \mathbf{0}_{\KType(\tau)} \ \reduce\  \mathbf{0}_{\KType(\sigma)}, \\
-    & \textrm{(R-MULK11)} && (O_1 \otimes O_2) \cdot (K_1 \otimes K_2) \ \reduce\  (O_1 \cdot K_1) \otimes (O_2 \cdot K_2).
-\end{align*}
-Some of these directions are obvious.
-For example, (R-DOT6) states that the inner product of two basis vectors is reduced to a delta expression, and (R-DELTA0) transforms a delta function on identical basis to a scalar \( 1 \).
-The rule (R-MULK1) reflects the axiom that multiplying a zero vector results in zero.
-This rule is conditional on typing, which is checked during rewriting.
-Some rules, like (R-MULK11), require a deeper understanding, such as the preference for tensor products over multiplication.
-
-As a reference, the term rewriting system in DiracDec has been proven complete for all axioms, except for the sum symbol.
-The completeness result is derived from checking the confluence and termination of the system.
-Our rewriting rules are translations from DiracDec into the typed and abstract language, ensuring that the corresponding symbols in our system are also complete.
-
-We also have additional rules that handle summations. For example:
-\begin{align*}
-    & \quad i \text{ free in } t \Rightarrow \sum_{i \in \mathbf{U}(\sigma)} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (\delta_{i, t}.A) \ \reduce\  \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} A\{i/t\}, \\
-    & \quad \left( \sum_{i \in M} B \right) \cdot K \ \reduce\  \sum_{i \in M} (B \cdot K).
-\end{align*}
-The first rule eliminates delta expressions in summations, while the second rule pushes summations outside of inner products.
-While there is no guarantee of completeness for these rules, they work effectively in practice.
-
-\subsubsection{Variable Expansion}
-One important technique, revealed in the DiracDec work, is the expansion of variables, which is critical for proofs involving summation. For example:
-\[
-\frac{E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash K \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)}(\bra{i} \cdot K).\ket{i}} \quad \quad
-\frac{E[\Gamma] \vdash B : \BType(\sigma)}{E[\Gamma] \vdash B \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)}(B \cdot \ket{i}).\bra{i}},
-\]
-\[
-\frac{E[\Gamma] \vdash O : \OType(\sigma, \tau)}{E[\Gamma] \vdash O \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)} \sum_{j \in \mathbf{U}(\tau)}(\bra{i} \cdot O \cdot \ket{j}).(\ket{i} \cdot \bra{j})}.
-\]
-These rules transform variables into their symbolic summations based on their decomposition over the basis. 
-The rules are not terminating, therefore is applied recrusively once in the second step called \textit{variable expansion}.
-Nevertheless, we have found that applying the expansion only once for all variables is sufficient for normalization.
-\begin{lemma}
-    Let \( \textrm{expand}(e) \) denote the result of expanding all variables in \( e \) once.
-For all well-typed terms \( e \) in \( E[\Gamma] \), \( \textrm{expand}(\textrm{expand}(e)) \) and \( \textrm{expand}(e) \) have the same normal form.
-\end{lemma}
-\begin{proof}
-    Expanding a ket variable twice, for example, results in the following transformation:
-    \[
-    \sum_{i \in \mathbf{U}(\sigma)} (\bra{i} \cdot \sum_{j \in \mathbf{U}(\sigma)} (\bra{j} \cdot K) \cdot \ket{j}) \cdot \ket{i} \ \reduce\  \sum_{i \in \mathbf{U}(\sigma)} \sum_{j \in \mathbf{U}(\sigma)} (\bra{j} \cdot K \cdot \braket{i|j}) \cdot \ket{i},
-    \]
-    where the delta symbol elimination rule returns the term to its original form. The same holds for bra and operator terms.
-\end{proof}
-
-\subsection{Deciding Equational Theory \( E \)}
-
-In the Mathematica implementation of DiracDec, the equational theory \( E \) is decided using unification, which attempts to find a substitution for the summation bound variables that makes two expressions syntactically equivalent.
-This unification process iterates through all permutations of AC symbol arguments, and its complexity is factorial in the maximum number of AC symbol arguments.
-
-As the first improvement, we check \( \alpha \)-equivalence using the de Bruijn index~\cite{deBruijn1972lambda}, which replaces references to bound variable names with the distance from the lambda abstraction to the variable.
-For instance, the nominal lambda abstraction \( \lambda x.
-x \) is transformed into \( \lambda .
-0 \), while \( \lambda x.
-\lambda y.
-(x\ (y\ x)) \) is transformed into \( \lambda.\lambda.
-(1\ (0\ 1)) \).
-This transformation into the de Bruijn index is performed as the final step in the normalization process to check equivalence between terms with different bound variable names.
-
-The remaining axioms, such as AC-equivalence and SUM-SWAP, assert equivalence under permutations.
-A standard approach for deciding such equivalences is to normalize terms by sorting in a predefined order.
-For example, given the dictionary order \( a < b < c \), the term \( b + c + a \) (and any other AC-equivalent term) is normalized into \( a + b + c \).
-However, in our setting, two intertwined difficulties arise: how to assign an order to all terms in the language, and how to simultaneously sort for both axioms.
-
-Consider the following two equivalent terms:
-\[
-\sum_{i \in s_1} \sum_{j \in s_2} \bra{i} A \ket{j} \times \bra{j} B \ket{i}
-= 
-\sum_{i \in s_2} \sum_{j \in s_1} \bra{i} B \ket{j} \times \bra{j} A \ket{i}
-\]
-While these two terms are equivalent, directly sorting the elements of scalar multiplication using lexical order does not yield the same form.
-
-To address this issue, we propose an algorithm that sorts terms without considering bound variables.
-The key observation is that in a successive sum expression \( \sum_{i \in s_1} \cdots \sum_{j \in s_n} A \), the names and order of the bound variables \( i, \dots, j \) can be freely permuted.
-Therefore, all bound variables should be treated uniformly during sorting, and the order of summation can then be determined based on the position of the bound variables.
-
-In the example above, we first ignore the bound variables and sort the sum body into \( \bra{\bullet} A \ket{\bullet} \times \bra{\bullet} B \ket{\bullet} \).
-Then, we swap the summations such that the bound variable at the first \( \bullet \) position appears at the outermost position.
-The results will have the same de Bruijn normal form, namely \( \sum_{s_1} \sum_{s_2} \bra{\$1} A \ket{\$0} \times \bra{\$0} B \ket{\$1} \).
-
-To describe the algorithm in the following, we introduce two key notations.
-For a term \( e = f(a_1, a_2, \dots, a_n) \), \( \textrm{head}(e) \) denotes the function symbol \( f \), while \( \textrm{arg}(e, i) \) refers to the \( i \)-th argument \( a_i \) of the term.
-In this context, variables and constants are treated as functions with zero arguments.
-
-\begin{definition}[Order Without Bound Variables]
-Let \( \mathcal{B} \) represent the set of bound variables, with the assumption that all bound variables are unique.
-We also assume that a total order exists over all symbols.
-The relation \( e_1 =_\mathcal{B} e_2 \) holds if:
-\begin{itemize}
-    \item \( \textrm{head}(e_1) = \textrm{head}(e_2) \), and for all \( i \), \( \textrm{arg}(e_1, i) =_\mathcal{B} \textrm{arg}(e_2, i) \), or
-    \item \( e_1 \in \mathcal{B} \) and \(e_2 \in \mathcal{B}\).
-\end{itemize}
-
-The relation \( e_1 <_\mathcal{B} e_2 \) holds between two terms if:
-\begin{itemize}
-    \item $e_1 \notin \mathcal{B}$ and $e_2 \in \mathcal{B}$, or
-    \item $head(e_1) < head(e_2)$, or
-    \item $head(e_1) = head(e_2)$, and there exists $n$ with $arg(e_1, n) <_\mathcal{B} arg(e_2, n)$, where $arg(e_1, i) =_\mathcal{B} arg(e_2, i)$ for all $i < n$.
-\end{itemize}
-\end{definition}
-It can be shown that \( e_1 =_\mathcal{B} e_2 \) if and only if neither \( e_1 <_\mathcal{B} e_2 \) nor \( e_2 <_\mathcal{B} e_1 \) holds.
-The purpose of this ordering is to compare function symbols in a top-down manner while ignoring bound variables.
-The sorted order enables normalization of terms in terms of AC equivalence.
-\begin{definition}[Sort Transformation]
-    For a term $e$ with bound variable set $\mathcal{B}$,
-    The sort transformation sort(e) is defined as
-    \begin{itemize}
-        \item $e$, if $e$ is a variable or constant;
-        \item $\lambda x:T. \textrm{sort}(e')$, if $e \equiv \lambda x : T. e'$;
-        \item $\mu x. \textrm{sort}(e')$, if $e \equiv \mu x. \textrm{sort}(e')$; or
-        \item $f(\textrm{sort}(a_1), \cdots, \textrm{sort}(a_n))$, if $e \equiv f(a_1, \cdots, a_n)$.
-If $f$ is an AC symbol, then the order of sort($a_i$) is sorted ascendingly according to $<_\mathcal{B}$.
-    \end{itemize}
-\end{definition}
-
-After sorting, the next step is the \textit{swap transformation}, which arranges successive summations based on the order of bound variables.
-\begin{definition}[Swap Transformation]
-For a term \( e \) with a sorting result \( \textrm{sort}(e) \), the swap transformation proceeds by ordering all bound variables according to their first appearances, except in function definitions \( \lambda x \) and \( \mu x \).
-The swap transformation then reorders the successive summations accordingly.
-\end{definition}
-
-The algorithm that applies the sort and swap transformations, followed by de Bruijn normalization, successfully handles all equivalences in our benchmark.
-We formalize the completeness of this transformation in the following conjecture:
-
-\begin{conjecture}[Completeness of Transformation]
-    For any two terms \( e_1 \) and \( e_2 \) that are equivalent under AC equivalence, SUM-SWAP, and \( \alpha \)-equivalence, they will have the same form after sort, swap, and de Bruijn transformations.
-\end{conjecture}
-
-% The order depends on there occurances in the last sorting result. If no occurance, then the order will depend on the set (for sum) and the type (for lambda abstraction only).
-
-%%%%%%%%%%%%%%%%%%%%%%%
-
-% The idea is to assign an order to terms, which is independent on the bound variables. Because we can have terms with nested AC symbols.
-
-Lastly, we can prove that the equivalence established by this normalization procedure is sound with respect to the semantics.
-
-\begin{theorem}[Soundness]
-    For any well-formed context \( E[\Gamma] \) and well-typed expressions \( e_1 \) and \( e_2 \), if  $e_1$ and $e_2$ have the same normal form, then \( \sem{e_1} = \sem{e_2} \).
-\end{theorem}
-
-\begin{proof}
-    The soundness of the term rewriting procedure follows from the fact that each rewriting rule preserves equivalence.
-Furthermore, the operations in the sort and swap transformations respect the AC-equivalence and SUM-SWAP axioms.
-Finally, the de Bruijn normalization ensures soundness for \( \alpha \)-equivalence.
-\end{proof}
-
-% Labelled Dirac Notation
-
-\input{labelled}
-
 \section{Implementation and Case Study}
@@
 The main purpose of this work is to build a practical tool that works well in checking Dirac notation equations.
-The refinements and extensions above concludes in our implementation called Dirace, a solver written in \CC.
+The refinements and extensions above concludes in our implementation called D-Hammer, a solver written in \CC.
 It has a parser built by ANTLR4, and scalar reasonings are powered by a Mathematica kernel.
@@
 \subsubsection{Project Structure}
-The project structure is illustrated in~\Cref{fig: dirace structure}.
+The project structure is illustrated in~\Cref{fig: dhammer structure}.
 \texttt{ualg} is the module for universal algebra, defining basic concepts like terms and substitutions.
-It serves as the library for \texttt{dirace}, which are then utilized in the example benchmarks and the toplevel command line application.
-The components of \texttt{dirace} are as follows:
+It serves as the library for \texttt{dhammer}, which are then utilized in the example benchmarks and the toplevel command line application.
+The components of \texttt{dhammer} are as follows:
 \begin{itemize}
@@
     \item \texttt{reduction.cpp}: all the rewriting rules and transformations;
-    \item \texttt{dirac\_parser.cpp}: parser for Dirac notation and Dirace commands;
+    \item \texttt{dirac\_parser.cpp}: parser for Dirac notation and D-Hammer commands;
     \item \texttt{prover.cpp}: the prover that maintains the context and process commands like definition or equivalence checking.
@@
 
-\input{dirace.tex}
+\input{fig/proj_structure.tex}
 
@@
 }
-The syntax tree structure is also compatible with the datatype of Mathematica. This improves the interoperability between Dirace and the Mathematica system, enabling them to work interleavingly.
-To improve usability, Dirace also supports many special syntacies for terms,  and most Dirac notation terms will be encoded in the natural way.
+The syntax tree structure is also compatible with the datatype of Mathematica. This improves the interoperability between D-Hammer and the Mathematica system, enabling them to work interleavingly.
+To improve usability, D-Hammer also supports many special syntacies for terms,  and most Dirac notation terms will be encoded in the natural way.
 Here are some examples for the parsing syntax.
@@
 
-Finally, Dirace uses a prover to host the computation.
+Finally, D-Hammer uses a prover to host the computation.
 The prover maintains a well-formed context $E[\Gamma]$, and processes commands to modify the context and conduct calculations.
@@
 \subsubsection{Use Case}
-Here we encode the motivating~\Cref{ex: motivating}, examine and explain how it is checked in Dirace. The encoding is shown below.
+Here we encode the motivating~\Cref{ex: motivating}, examine and explain how it is checked in D-Hammer. The encoding is shown below.
 
-    \begin{lstlisting}[style=dirace]
+    \begin{lstlisting}[style=dhammer]
 Var T : INDEX. Var M : OTYPE[T, T].
@@
 
-\begin{lstlisting}[style=dirace]
+\begin{lstlisting}[style=dhammer]
 Def TPO := idx sigma => idx tau => fun O : OTYPE[sigma, tau] => Sum i in USET[sigma], Sum j in USET[tau], (<i| O |j>).(|j> <i|).
@@
 The \texttt{TPO} symbol represents the transpose of operators, and encodes the formalization in~\Cref{ex: formalizing motivating}.
-Thanks to the design of environment and functions, many other commonly used symbols in Dirac notation are encoded and provided as defined symbols in Dirace.
+Thanks to the design of environment and functions, many other commonly used symbols in Dirac notation are encoded and provided as defined symbols in D-Hammer.
 
 Within one second, the prover reports the result of equivalence with their common normal form:
-    \begin{lstlisting}[style=dirace]
+    \begin{lstlisting}[style=dhammer]
 The two terms are equal.
@@
 
-To evaluate Dirace, we first test the examples from DiracDec benchmark and make a comparison.
+To evaluate D-Hammer, we first test the examples from DiracDec benchmark and make a comparison.
 The experiments are carried out using a MacBook Pro with M3 Max chip.
@@
         \hline
-        \multirow{2}{*}{source} & \multicolumn{3}{c|}{DiracDec} & \multicolumn{3}{c}{Dirace} \\
+        \multirow{2}{*}{source} & \multicolumn{3}{c|}{DiracDec} & \multicolumn{3}{c}{D-Hammer} \\
         \cline{2-7}
@@
 
-The timing of Dirace does not include the initialization of Mathematica kernel link, which takes about 3 seconds in the beginning.
+The timing of D-Hammer does not include the initialization of Mathematica kernel link, which takes about 3 seconds in the beginning.
 As to expressibility, the language for DiracDec has the support for projectors $\texttt{fst}$ and $\texttt{snd}$ on basis pairs, satisfying $\texttt{fst} (s, t) = s$ and $\texttt{snd} (s, t) = t$.
 We found this feature is almost not used, so we removed the support.
-As a result, Dirace encodes 158 examples for the CoqQ part, 4 less than DiracDec.
-For decidability, we improved the rewriting rules about sum for Dirace, and proved several more examples failed by DiracDec.
+As a result, D-Hammer encodes 158 examples for the CoqQ part, 4 less than DiracDec.
+For decidability, we improved the rewriting rules about sum for D-Hammer, and proved several more examples failed by DiracDec.
 The main difference is about their time efficiency.
-Because of our algorithm to decide AC-equivalence and SUM-SWAP, Dirace has a significant efficiency improvement, especially on those ``computational examples'' mentioned in the DiracDec paper.
+Because of our algorithm to decide AC-equivalence and SUM-SWAP, D-Hammer has a significant efficiency improvement, especially on those ``computational examples'' mentioned in the DiracDec paper.
 One typical example comes from the paper by Jens about the equivalence of operators for qubits.
 The system has to decompose the term on the concrete $\ket{0}$ and $\ket{1}$ basis, resulting a lot of addition elements.
-It takes DiracDec about one minute, but Dirace solves it within one second.
+It takes DiracDec about one minute, but D-Hammer solves it within one second.
 
 We also built an example benchmark for labelled Dirac notation, see~\Cref{sec: examples for labelled}.
-
-\section{Related Work}
-
-Automated theorem proving has seen significant advancements in recent years, particularly through the development of satisfiability modulo theories (SMT) solvers,
-% SMT solvers extend propositional satisfiability by integrating theories such as arithmetic, arrays, and bit-vectors, making them suitable for solving a wide range of verification problems in both hardware and software domains.
-including prominent tools like Z3. These solvers have become essential in various fields such as formal verification, synthesis, and model checking. 
-Equational reasoning is another crucial area of research within automated theorem proving, which focuses on solving problems that involve equations between terms in an algebraic structure. 
-Equational provers, such as Vampire and E, have played a pivotal role in addressing the challenge of proving equations in first-order logic, employing sophisticated algorithms like superposition and term rewriting.
-
-Formal verification of quantum computation is receiving increasing attentions during these years. See~\cite{Lewis2023} for a comprehensive review. 
-Verification frameworks in Coq include the foundational formalization CoqQ and quantum circuit language QWIRE. 
-Verifications of quantum programs are also considered, such as the Hoare logic based methods [] and model checking based methods [].
-The equational reasoning of Dirac notation is crucial for establishing property proofs in these works.
-Verification theoreis and tools based on other languages are also proposed, such as PyZX for reasoning and simplification of ZX-calculus.
-
-\section{Conclusion and Future Work}
-Based on the first equational reasoning tool for Dirac notation called DiracDec, this work improves and extends the theory for practical applications, and provides the solver Dirace.
-Experiments show that the tool demonstrates advantages in decidability, efficiency and usability.
-
-We expect Dirace to have applications in areas like quantum program verification or proofs of post-quantum cryptography protocols in the future.
-One promising following up is to connect Dirace with theorem provers like Coq.
-It involves transforming theorem prover expressions into Dirace, and verify the proof trace of Dirace in theorem provers.
-Besids, most quantum program verifiers nowadays depend on matrix calculations.
-Dirace can serve as the replacement for matrix methods to enable symbolic deductions.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% \begin{credits}
-%     \subsubsection{\ackname} A bold run-in heading in small font size at the end of the paper is
-%     used for general acknowledgments, for example: This study was funded
-%     by X (grant number Y).
-% \end{credits}
-
-%
-% ---- Bibliography ----
-%
-% BibTeX users should specify bibliography style 'splncs04'.
-% References will then be sorted and formatted in the correct style.
-%
-\bibliographystyle{splncs04}
-\bibliography{ref}
-%
-
-\appendix
-\include{appendix}
-
-\end{document}

### 2025-01-27 fig/proj_structure.tex
splited

--- 2025-01-23 dirace.tex
@@
             \node (toplevel) [file, right of=examples, xshift=1cm] {toplevel.cpp};
-            \node (dirace) [module, below right of=examples, xshift=0.3cm] {dirace};
-            \node (ualg) [module, below left of=dirace] {ualg};
+            \node (dhammer) [module, below right of=examples, xshift=0.3cm] {dhammer};
+            \node (ualg) [module, below left of=dhammer] {ualg};
             \node (ANTLR4) [thirdparty, below of =ualg] {ANTLR4};
-            \node (WSTP) [thirdparty, below right of =dirace, xshift=1cm, yshift=-0.3cm, align=center] {WSTP interface\\(Mathematica)};
+            \node (WSTP) [thirdparty, below right of =dhammer, xshift=1cm, yshift=-0.3cm, align=center] {WSTP interface\\(Mathematica)};
 
-            \draw [arrow] (examples) -- (dirace);
-            \draw [arrow] (toplevel) -- (dirace);
-            \draw [arrow] (dirace) -- (ualg);
+            \draw [arrow] (examples) -- (dhammer);
+            \draw [arrow] (toplevel) -- (dhammer);
+            \draw [arrow] (dhammer) -- (ualg);
             \draw [arrow] (ualg) -- (ANTLR4);
-            \draw [arrow] (dirace) -- (WSTP);
+            \draw [arrow] (dhammer) -- (WSTP);
         \end{tikzpicture}
@@
     \rmfamily
-    \caption{The structure of Dirace.
-The left figure illustrates the whole system, and the right figure illustrates the \texttt{dirace} module.
+    \caption{The structure of D-Hammer.
+The left figure illustrates the whole system, and the right figure illustrates the \texttt{dhammer} module.
 Blue nodes denote modules, red nodes denote third-party libraries, and green nodes denote files.
 Arrows represent dependency.}
-    \label{fig: dirace structure}
+    \label{fig: dhammer structure}
 \end{figure}

### 2025-01-27 main.tex
splited

--- 2025-01-25 main.tex
@@
 %
-\title{Dirace: Practical Proof Automation of Dirac Notation Equations}
+\title{D-Hammer: Practical Proof Automation of Dirac Notation Equations}
 
@@
 Enhancements include an improved typing system, a simplified language and rewriting system, more efficient algorithms, and added support for labeled Dirac notation.
-Our solver, Dirace, is implemented in \CC\ with a Mathematica backend, demonstrating significant improvements in decision-making power and time efficiency.
+Our solver, D-Hammer, is implemented in \CC\ with a Mathematica backend, demonstrating significant improvements in decision-making power and time efficiency.
 
@@
 
+\newcommand{\var}{\mathsf{var}}
+\newcommand{\reg}{\mathsf{Reg}}
+\newcommand{\DType}{\mathcal{D}}
+\newcommand{\cR}{\mathcal{R}}
+\newcommand{\cN}{\mathcal{N}}
+\newcommand{\tD}{\tilde{D}}
+\newcommand{\te}{\tilde{e}}
+\newcommand{\tT}{\tilde{T}}
+\newcommand{\tADD}{\widetilde{ADD}}
+\newcommand{\bU}{\mathbf{U}}
+\newcommand{\<}{\langle}
+\newcommand{\simp}{\mathsf{Simp}}
+\newcommand{\List}{\mathsf{list}}
+\renewcommand{\>}{\rangle}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\section{Introduction}
+\input{01intro_preliminary.tex}
 
-In 1939, Dirac introduced his notation for quantum mechanics~\cite{dirac1939new}, designed to represent linear algebraic formulas in a compact and convenient form.
-For example, the expression \( a\ket{\psi} + b\ket{\phi} \) represents the superposition of two quantum states, \( \ket{\psi} \) and \( \ket{\phi} \).
-Today, Dirac notation is widely accepted as the standard language in quantum computation and quantum information.
-Its reasoning forms the foundation of research and applications, much like boolean and integer logic do in classical computer science.
+\input{02techniques.tex}
 
-In quantum algorithm formalizations and quantum programming languages, Dirac notation is frequently used in equational proofs, which are critical, repetitive, and often time-intensive.
-These notations also play a key role in defining program states, operations, and assertions in quantum programming languages.
-To automate the verification of these programs, we need tools that can simplify and check the equivalence of preconditions.
-However, unlike the well-established SAT and SMT solvers for classical logic, a practical solver for Dirac notation equivalence remains an unmet need, creating a barrier for progress in several fields.
+\input{03tool_eval.tex}
 
-Recently, Xu et al.~\cite{diracdec} proposed a theory for deciding the equivalence of Dirac notation, alongside a prototype implementation in Mathematica called DiracDec.
-They demonstrated that the equivalence of basic Dirac notation is decidable.
-Their algorithm, based on a pure term rewriting system, has been proven to be confluent and terminating.
-Despite this, there remains a gap between DiracDec and a practical solver for Dirac notation equivalence.
-
-One challenge is the efficiency of the algorithm when dealing with equivalences beyond the scope of term rewriting.
-DiracDec decides the entire equational theory by rewriting modulo \( E \), where \( E \) represents a set of axioms that cannot be decided by normalization alone, such as the associativity and commutativity (AC) of certain function symbols.
-DiracDec uses a direct but inefficient algorithm to decide these axioms, which searches through all possible permutations and exhibits factorial complexity.
-This inefficiency becomes particularly evident in "computational examples" containing many AC symbols, which are time-consuming to process.
-
-Usability is another area where DiracDec falls short.
-It does not support labelled Dirac notation, which uses registers to denote subsystems and express states locally.
-Additionally, DiracDec's typing system only does not provide context for variable typing assumptions or the definition of symbols, which are required in practical scenarios.
-To avoid type checking during term rewriting, DiracDec separates symbols (e.g., multiplication) for different types, leading to unnecessary complexity.
-Moreover, integrating the Mathematica implementation into other projects as a solver is challenging.
-
-The system design of DiracDec reflects a trade-off between simplicity and efficiency.
-While the simplicity of term rewriting allows for strong theoretical results, it limits optimization opportunities.
-Building on DiracDec, this work aims to develop a practical solver.
-We transform the term rewriting system into a hybrid algorithm, overcoming the challenges mentioned above.
-Our main technical contributions are:
-\begin{itemize}
-    \item An efficient algorithm for deciding the equational theories in \( E \), based on equivalence checking through normalization, with the normal form for \( E \) being obtained via sorting.
-    \item Support for constant register labels, and reducing the equivalence decision of labelled Dirac notation to the unlabelled case.
-    \item A more user-friendly \CC\ solver, Dirace, featuring an abstract language and typing system. We also support the definition of symbols (e.g., transpose and trace) using function syntax.
-\end{itemize}
-
-We evaluated Dirace against the DiracDec benchmark and new examples involving labelled Dirac notation.
-The results show significant improvements in both decidability and efficiency compared to DiracDec.
-
-% Dirace successfully decides all the examples that are expressable in its language, including those failed by DiracDec because of complexity or insufficient decision power.
-
-\section{Motivation and Preliminary}
-
-An interesting property of quantum mechanics is that for two maximally entangled states, applying a quantum operator \( M \) to one subsystem is equivalent to applying \( M^T \) (the transpose of \( M \)) to the other subsystem.
-This relationship holds regardless of the spatial separation between the two systems, and it can be expressed as an equation in Dirac notation.
-\begin{example}
-    \label{ex: motivating}
-    Let \( q \) and \( r \) represent two quantum systems in the Hilbert space \( \mathcal{H}_T \).
-Let \( M \) be a quantum operation acting on \( \mathcal{H}_T \), and let \( \ket{\Phi} = \sum_{i \in T} \ket{i} \otimes \ket{i} \) be the maximally entangled state.
-Then, we have the following equation:
-    \[
-    M_q \ket{\Phi}_{q; r} = M_r^T \ket{\Phi}_{q; r}.
-    \]
-\end{example}
-
-In this equation, \( q \) and \( r \) are labels denoting the respective subsystems in the Dirac notation.
-To automate reasoning about such equations, we must formalize the language and develop a proof system to handle it.
-
-\subsubsection{Dirac Notation}
-
-Quantum states are represented as vectors in complex Hilbert spaces, and operations on these states are described by linear transformations.
-Dirac notation uses the ket \( \ket{i} \) and the bra \( \bra{i} \) to denote basis vectors in a Hilbert space and its dual space, respectively.
-These symbols can be composed together in various ways to form more complex expressions.
-The meaning of these compositions depends on the types of the operands involved.
-For example, the inner product \( \braket{i|j} \) represents the scalar result of the dot product between \( \bra{i} \) and \( \ket{j} \), while the outer product \( \ket{i}\bra{j} \) represents an operator.
-Additionally, Dirac notation uses the tensor product symbol \( \otimes \) to describe the combined space of multiple quantum systems.
-
-One of the key features of Dirac notation is its order-independent interpretation.
-This means that the composition of terms can be written without parentheses, as the interpretation is unaffected by the order of multiplication.
-For instance, the formula \( \bra{i}\ket{\phi}\bra{\psi}\ket{j} \) can be understood as:
-\[
-    \braket{i|\phi}\braket{\psi|j} = \bra{i} (\ket{\phi}\bra{\psi}) \ket{j},
-\]
-and these expressions are equivalent for all variables involved.
-
-In practice, Dirac notation is often combined with other syntactic elements, such as the summation symbol \( \sum_{i \in S} A \), to enhance expressiveness.
-Furthermore, labelled Dirac notation, such as \( \ket{i}_q \otimes \ket{i}_r \), is used to denote quantum systems in consideration, where subscripts (registers) are added to distinguish different subsystems.
-
-\subsubsection{Universal Algebra}
-We use universal algebra and equational logic to formally represent Dirac notation and the reasoning procedure.
-A universal algebra defines a signature of function symbols, with terms constructed from constants, variables, and function applications.
-Other basic concepts like substitution of variables or pattern matching are also defined.
-In our case of Dirac notation, the signature consists of constructors and operations like $\ket{i}$ or $A \otimes B$.
-The reasoning process is guided by equational logic, which defines an equivalence relation that is compatible with substitution and term construction.
-This relation formalizes the intuitive concept of equivalence in algebra.
-
-\section{Language, Typing and Semantics}
-The first step is to formalize the language for Dirac notation.
-In DiracDec, the language has a concrete design, where the same syntax for different types corresponds to distinct symbols.
-Our goal is to transition from this concrete design to a more abstract one, which aligns more closely with the conventional Dirac notation we use and simplifies the term rewriting system.
-
-To achieve this, we organize our language into three layers: the index, the type, and the term.
-Terms represent concrete instances such as kets, bras, and operators, which will be typed and checked.
-The index represents classical data types and appears in type expressions to differentiate between various Hilbert spaces and sets.
-
-\begin{definition}[Index Syntax]
-    The syntax for type indices is:
-    \begin{align*}
-        \sigma ::=\ & x \mid \sigma_1 \times \sigma_2.
-    \end{align*}
-\end{definition}
-Here, \( x \) is a variable, and \( \sigma_1 \times \sigma_2 \) represents the product type for tensor product spaces or Cartesian product sets.
-
-\begin{definition}[Type Syntax]
-    The syntax for Dirac notation types is:
-    \begin{align*}
-        T ::=\ & \Basis(\sigma) \mid \SType \mid \KType(\sigma) \mid \BType(\sigma) \mid \OType(\sigma_1, \sigma_2) \mid T_1 \to T_2 \mid \forall x.T \mid \SET(\sigma).
-    \end{align*}
-\end{definition}
-\( \Basis(\sigma) \) denotes the type for basis elements in the index \( \sigma \).
-\( \SType \) represents scalars, while \( \KType(\sigma) \) and \( \BType(\sigma) \) refer to ket and bra types in the Hilbert space \( \sigma \), respectively.
-\( \OType(\sigma_1, \sigma_2) \) represents linear operators with \( \sigma_2 \) as the domain and \( \sigma_1 \) as the codomain.
-\( \SET(\sigma) \) refers to the type of subsets of \( \sigma \), used to denote the values of bound variables in summations.
-The remaining two constructs define function types: \( T_1 \to T_2 \) represents normal functions that take a \( T_1 \)-type argument and return a \( T_2 \)-type term, while \( \forall x.
-T \) represents index functions that take an index argument \( x \) and produce a \( T \)-type term, which may depend on \( x \).
-Index functions are essential for defining polymorphic transformations across Hilbert spaces.
-
-\begin{definition}[Term Syntax]
-    The syntax for Dirac notation terms is:
-    \begin{align*}
-        e ::=\ & x \mid \lambda x : T.e \mid \mu x.e \mid e_1\ e_2 \mid (e_1, e_2) \\
-        & |\ 0 \mid 1 \mid \ADDS(e_1, \cdots, e_n) \mid e_1 \times \cdots \times e_n \mid e^* \mid \delta_{e_1, e_2} \mid \DOT(e_1, e_2) \\
-        & |\ \ZEROK(\sigma) \mid \ZEROB(\sigma) \mid \ZEROO(\sigma_1, \sigma_2) \mid \ONEO(\sigma) \\
-        & |\ \ket{e} \mid \bra{t} \mid e^\dagger \mid e_1.e_2 \mid \ADD(e_1, \cdots, e_n) \mid e_1 \otimes e_2 \\
-        & |\ \MULK(e_1, e_2) \mid \MULB(e_1, e_2) \mid \OUTER(e_1, e_2) \mid \MULO(e_1, e_2) \\
-        & |\ \mathbf{U}(e) \mid e_1 \star e_2 \mid \sum_{e_1} e_2.
-    \end{align*}
-\end{definition}
-The terms above are explained in order:
-\( \lambda x : T.e \) represents the abstraction for normal functions, and \( \mu x.e \) represents the abstraction for index functions.
-\( e_1\ e_2 \) denotes function application.
-\( (e_1, e_2) \) is the basis pair for product types.
-The constants \( 0 \), \( 1 \), \( \ADDS \), \( e_1 \times \cdots \times e_n \), and \( e^* \) are symbols for scalars.
-The next line includes constant symbols for kets, bras, and operators.
-\( \ket{e} \) is a ket, \( \bra{t} \) is a bra, and \( e^\dagger \) denotes the conjugate transpose of \( e \).
-\( e_1.e_2 \) represents scaling the term \( e_2 \) by scalar \( e_1 \).
-\( \mathbf{U}(e) \) denotes the universal set with index \( e \).
-\( e_1 \star e_2 \) represents the Cartesian product of \( e_1 \) and \( e_2 \).
-\( \sum_{e_1} e_2 \) is the big operator sum, modeled by folding the function \( e_2 \) over the value set \( e_1 \).
-Typically, the sum's body is given by an abstraction.
-For convenience, we also use the notation \( \sum_{x \in s} X \) to represent \( \sum_{s} \lambda x : T .
-X \).
-Additionally, \( \ADDS \) and \( \ADD \) are distinct AC symbols: \( \ADDS \) is used for scalar addition and \( \ADD \) for linear algebra addition.
-There are five kinds of linear algebraic multiplications among kets, bras, and operators.
-These are encoded using five different symbols: \( \DOT \), \( \MULK \), \( \MULB \), \( \OUTER \), and \( \MULO \).
-
-Compared to DiracDec, the symbols for addition, adjoint, scaling, and tensor products have been consolidated.
-The multiplications are still kept separate due to their differing properties.
-These operations are denoted as \( B \cdot K \), \( K_1 \cdot K_2 \), \( B_1 \cdot B_2 \), \( K \cdot B \), and \( O_1 \cdot O_2 \), respectively.
-
-\subsection{Typing System}
-
-The typing system is responsible for classifying terms within a proof system, using a context that specifies the types of variables and definitions.
-The context is divided into two components: the environment \( E \), which includes definitions and assumptions, and the context \( \Gamma \), which tracks bound variables.
-Both \( E \) and \( \Gamma \) consist of sequences of assumptions \( x : T \) or definitions \( x := t : T \).
-\begin{definition}[Environment and Context]
-    The syntax for \( E \) and \( \Gamma \) is:
-    \begin{align*}
-        E ::= &\ [] \mid E; x : \Index \mid E; x : T \mid E; x := t : T. \\
-        \Gamma ::= &\ [] \mid \Gamma; x : \Index \mid \Gamma; x : T.
-    \end{align*}
-\end{definition}
-Definitions refer to symbols that can be expanded or unfolded, and typically represent abstract concepts, such as transpose or trace in Dirac notation.
-Assumptions, on the other hand, define the types of variables, and variable assumptions are implicitly universally quantified in the case of equational proofs.
-
-Type checking in our language involves maintaining a well-formed environment and context \( E[\Gamma] \).
-We say an expression \( t \) has type \( X \) in context \( E[\Gamma] \) if the typing judgment \( E[\Gamma] \vdash t : X \) can be proven through the rules in~\Cref{sec: full typing rules}.
-The well-formedness of the context \( \WF(E)[\Gamma] \) is built incrementally:
-\[
-    \frac{}{\WF([])[]}
-    \qquad
-    \frac{\WF(E)[] \qquad x \notin E}{\WF(E; x : \Index)[]}
-    \qquad
-    \frac{E[]\vdash t:T \qquad x \notin E}{\WF(E; x := t:T)[]}.
-\]
-Starting with an empty context, new index symbols can be introduced, and symbols with checked types can be defined.
-Typing judgments are then made inductively using the information from \( E[\Gamma] \).
-For instance, the rule \( x : \Index \in E[\Gamma] \) holds if \( E \) or \( \Gamma \) contains an assumption for \( x \).
-And \(\KType(\sigma)\) is a valid type for kets, if the argument $\sigma$ is typed as an index.
-\begin{gather*}
-    \frac{\WF(E)[\Gamma] \qquad x : \Index \in E[\Gamma]}{E[\Gamma] \vdash x : \Index}
-    \qquad
-    \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \KType(\sigma) : \Type}
-\end{gather*}
-
-Terms are then typed accordingly.
-For example, the ket \( \ket{t} \) will have the type \( \KType(\sigma) \) if \( t \) is a basis term of index \( \sigma \).
-Similarly, the inner product between a bra and a ket of the same index \( \sigma \) is typed as a scalar.
-It corresponds to the constraint of inner product that vectors should be from the same Hilbert space.
-\begin{gather*}
-    \frac{E[\Gamma]\vdash t : \Basis(\sigma)}{E[\Gamma] \vdash \ket{t} : \KType(\sigma)}
-    \qquad
-    \frac{E[\Gamma]\vdash B : \BType(\sigma) \qquad E[\Gamma]\vdash K : \KType(\sigma)}{E[\Gamma] \vdash B \cdot K : \SType}
-\end{gather*}
-
-The typing for functions and applications follows the standard practice.
-For example, the index function \( \mu x.
-t \) is typed with \( x \) as a bound variable of type \( \Index \), and the application \( (t\ u) \) has the type \( U\{x/u\} \), obtained by replacing \( x \) with the index instance \( u \).
-\begin{gather*}
-    \frac{E[\Gamma; x : T] \vdash t : U}{E[\Gamma] \vdash (\lambda x : T . t) : T \to U}
-    \qquad
-    \frac{E[\Gamma] \vdash t:U \to T \qquad E[\Gamma] \vdash u:U}{E[\Gamma] \vdash (t\ u):T} 
-    \\[1em]
-    \frac{E[\Gamma; x : \Index] \vdash t : U}{E[\Gamma] \vdash (\mu x.t) : \forall x.U}
-    \qquad
-    \frac{E[\Gamma] \vdash t:\forall x.U \qquad E[\Gamma] \vdash u : \Index}{E[\Gamma] \vdash (t\ u):U\{x/u\}}
-\end{gather*}
-
-Finally, the big operator sum is modeled by folding a function over a set, with the typing rule as follows:
-\[
-    \frac{E[\Gamma] \vdash s : \SET(\sigma) \qquad E[\Gamma] \vdash f : \Basis(\sigma) \to \KType(\tau)}{E[\Gamma] \vdash \sum_{s} f : \KType(\tau)}.
-\]
-
-\begin{lemma}
-    The typing of expressions is both decidable and unique.
-\end{lemma}
-
-\begin{proof}
-    The type of an expression can be determined recursively. For any given function symbol and argument types, there is at most one typing rule, ensuring the uniqueness of typing.
-\end{proof}
-
-\subsection{Semantics}
-
-The semantics of a language define the meaning of its expressions.
-In this context, the objective of our algorithm is to determine whether two expressions are semantically equivalent.
-We define the semantics in a denotational manner, mapping syntax to set-theoretic objects.
-
-\subsubsection{Denotational Semantics}
-The denotational semantics interpret every expression as an object in linear algebra, according to a valuation mapping \( v \), which assigns values to the variables in the expression.
-The semantics of an expression \( e \) with a given valuation \( v \) is denoted as \( \sem{e}_v \).
-Two expressions \( e_1 \) and \( e_2 \) are considered equivalent if their semantics are equal for all valuations, i.e., \( \sem{e_1}_v = \sem{e_2}_v \) for all \( v \).
-
-The complete interpretation of terms and types is provided in~\Cref{sec: full denotational sem}.
-Variables typed with \( \Index \) are interpreted as finite sets, and the product of two indices \( \sem{\sigma_1 \times \sigma_2} \) is defined as the Cartesian product of the sets \( \sem{\sigma_1} \) and \( \sem{\sigma_2} \).
-More generally, each type is interpreted as a set.
-For example, the scalar type \( \sem{\SType} \) is interpreted as the set of complex numbers \( \mathbb{C} \), and the ket and bra types \( \sem{\KType(\sigma)} \) and \( \sem{\BType(\sigma)} \) are interpreted as the Hilbert space \( \mathcal{H}_{\sem{\sigma}} \) and its dual \( \mathcal{H}_{\sem{\sigma}}^* \), respectively.
-Terms are explained as the set elements.
-For example, the semantics of ket tensor product $\sem{K_1 \otimes K_2} \equiv \sem{K_1} \otimes \sem{K_2}$, is obtained by first calculating the semantics $\sem{K_1}$ and $\sem{K_2}$ as vectors, and then take the vector tensor product as result.
-
-% One special case is the delta function \( \delta_{s,t} \), which is interpreted as:
-% \[
-%     \sem{\delta_{s,t}} =
-%     \begin{cases}
-%         1, & \text{if } \sem{s} = \sem{t}, \\
-%         0, & \text{if } \sem{s} \neq \sem{t}.
-%     \end{cases}
-% \]d
-
-The idea behind the interpretation of types and terms is to formalize the typing relation using set-theoretic inclusion.
-Specifically, for a well-formed context \( E[\Gamma] \), term \( t \), and type \( T \), if \( E[\Gamma] \vdash t : T \), then for any valuation \( v \), the semantics of \( t \) must lie within the semantics of \( T \).
-
-\begin{lemma}
-    For any well-formed context \( E[\Gamma] \), term \( t \), and type \( T \), if \( E[\Gamma] \vdash t : T \), then for all valuations \( v \), \( \sem{t}_v \in \sem{T}_v \).
-\end{lemma}
-
-\begin{proof}
-    The proof follows directly by checking each case.
-\end{proof}
-
-This interpretation formalizes the standard understanding of Dirac notation and provides the foundation for the algorithm.
-However, computers cannot directly reason about equivalence through mathematical interpretations.
-Therefore, we proceed by defining a proof system that abstracts these concepts.
-
-\subsubsection{Axiomatic semantics}
-
-The proof system for equivalence is based on equational logic, together with axioms that describe the properties of Dirac notation.
-A full list of these axioms can be found in~\Cref{sec: full axioms}.
-The axioms cover fundamental aspects of linear spaces, as well as other structures like the tensor and inner products.
-For example, we have the absorption law for zero symbols:
-\(X \cdot \mathbf{0} = \mathbf{0},\)
-and the bilinearity of the dot product:
-\[
-(a.X) \cdot Y = a \cdot (X \cdot Y), \quad X \cdot (Y_1 + Y_2) = X \cdot Y_1 + X \cdot Y_2, 
-\]
-\[
-X \cdot (a.Y) = a \cdot (X \cdot Y), \quad (X_1 + X_2) \cdot Y = X_1 \cdot Y + X_2 \cdot Y.
-\]
-
-As mentioned in the introduction, a subset \( E \) of these axioms cannot be decided by term rewriting alone. The axioms in \( E \) include:
-\[
-\begin{aligned}
-    \text{AC-equivalence} &\quad \text{e.g.,} \quad X + Y = Y + X, \quad (X + Y) + Z = X + (Y + Z), \\
-    \alpha\text{-equivalence} &\quad \lambda x . A = \lambda y . A\{x/y\}, \\
-    \text{SUM-SWAP} &\quad \sum_{i \in s_1} \sum_{j \in s_2} A = \sum_{j \in s_2} \sum_{i \in s_1} A, \\
-    \text{scalar theories} &\quad \text{e.g.,} \quad a + 0 = a, \quad a \times (b + c) = a \times b + a \times c.
-\end{aligned}
-\]
-The scalar theories are treated separately as a module and are not considered in this work. In the implementation, we use the Mathematica kernel to decide scalar equivalences.
-These equational axioms provide an operable theory for the proof automation algorithm.
-Denotational semantics can be seen as one model for this theory, meaning that equivalences derived from the axioms always imply equivalence in the interpretations.
-\begin{lemma}
-    For all well-formed contexts \( E[\Gamma] \) and terms \( e_1, e_2 \), if \( \vdash e_1 = e_2 \), then \( \sem{e_1} = \sem{e_2} \).
-\end{lemma}
-\begin{proof}
-    This follows directly from checking all cases.
-\end{proof}
-
-Thus, the axioms are sound with respect to denotational semantics.
-However, the reverse does not hold: there exist equivalences in denotational semantics that cannot be captured by these axioms.
-Nevertheless, our work focuses on solving practical examples, which are fully covered by the axioms in the experiments.
-
-To conclude this section, we demonstrate the formalization of the symbols used in the motivating example.
-\begin{example}[Formalizing the Motivating Example]
-    \label{ex: formalizing motivating}
-    Definitions and assumptions in the environment \( E \) are formalized as follows:
-    \begin{align*}
-        & \text{TPO} && := \mu T_1. \mu T_2. \lambda O : \OType(T_1, T_2). \sum_{i \in \mathbf{U}(T_1)} \sum_{j \in \mathbf{U}(T_2)} \bra{i} O \ket{j} . \ket{j}\bra{i} \\
-        & &&\quad : \forall T_1. \forall T_2. \OType(T_1, T_2) \to \OType(T_2, T_1); \\
-        &\text{phi} &&:= \mu T. \sum_{i \in \mathbf{U}(T)} \sum_{j \in \mathbf{U}(T)} \ket{(i, j)} : \forall T.\KType(T \times T); \\
-        & T && : \Index; \\
-        & M && : \OType(T, T).
-    \end{align*}
-    The symbol \( \text{TPO} \) represents the transpose of an operator, polymorphic on the Hilbert spaces \( T_1 \) and \( T_2 \).
-The symbol \( \text{phi} \) takes the index \( T \) and defines the maximally entangled states, summing over all basis elements in \( T \), as indicated by the universal set \( \mathbf{U}(T) \).
-    With the assumption of the index \( T \) and operator \( M \), we can express the equivalence in the non-labelled version as:
-    \[
-    (\textrm{M} \otimes \mathbf{1}_\mathcal{O}(\textrm{T})) \cdot (\textrm{phi T}) = (\mathbf{1}_\mathcal{O}(\textrm{T}) \otimes (\textrm{TPO T T M})) \cdot (\textrm{phi T}).
-    \]
-\end{example}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Algorithm for Deciding Dirac Notation Equations}
-\label{sec: decide}
-
-We utilize the term rewriting technique to decide the equivalence of Dirac notation, based on its axiomatic semantics.
-The process involves normalizing two terms and checking whether they have identical syntax.
-This normalization is achieved through matching and substitution in universal algebra, using a set of predefined rewriting rules.
-
-As mentioned in the semantics section, the axioms in \( E \) cannot be fully decided through term rewriting.
-DiracDec decides \( E \) by examining all possible permutations in the rewriting result.
-In this work, we introduce a more efficient approach by incorporating sorting algorithms into the normalization procedure.
-
-The normalization procedure consists of the following steps.
-\begin{enumerate}
-    \item \textbf{First Rewritings}: Expand definitions and simplify expressions.
-    \item \textbf{Variable Expansion}: Consider scalar expressions to ensure completeness.
-    \item \textbf{Second Rewritings}: Normalize terms modulo \( E \).
-    \item \textbf{Sorting Without Bound Variables}: Normalize AC-equivalence.
-    \item \textbf{Swap Successive Summations}: Normalize SUM-SWAP equivalences.
-    \item \textbf{De Bruijn Normalization}: Normalize \( \alpha \)-equivalence.
-\end{enumerate}
-
-\subsection{Normalization Modulo \( E \) by Term Rewriting}
-
-Term rewriting rules, represented as \( l \ \reduce\ r \), are applied recursively to normalize terms.
-In each step, subterms matching the left-hand side \( l \) of a rule are replaced with the corresponding right-hand side \( r \).
-The procedure terminates when no further rewritings can be made.
-A comprehensive list of rewriting rules can be found in~\Cref{sec: rewriting rules}. Below are some key examples to illustrate the design:
-
-One of our optimizations is using functions with indefinite arities. 
-Therefore, we use a flattening rule to handle associativity with AC symbols:
-\[
-a_1 + \cdots + (b_1 + \cdots + b_m) + \cdots + a_n \ \reduce\ a_1 + \cdots + b_1 + \cdots + b_m + \cdots + a_n.
-\]
-Commutativity is handled later in the sorting step. Many of the rewriting rules are directly derived from the equational axioms, such as:
-\begin{align*}
-    & \textrm{(R-DOT6)} && \bra{s} \cdot \ket{t} \ \reduce\  \delta_{s, t}, \\
-    & \textrm{(R-DELTA0)} && \delta_{s, s} \ \reduce\  1, \\
-    & \textrm{(R-MULK1)} && O : \OType(\sigma, \tau) \Rightarrow O \cdot \mathbf{0}_{\KType(\tau)} \ \reduce\  \mathbf{0}_{\KType(\sigma)}, \\
-    & \textrm{(R-MULK11)} && (O_1 \otimes O_2) \cdot (K_1 \otimes K_2) \ \reduce\  (O_1 \cdot K_1) \otimes (O_2 \cdot K_2).
-\end{align*}
-Some of these directions are obvious.
-For example, (R-DOT6) states that the inner product of two basis vectors is reduced to a delta expression, and (R-DELTA0) transforms a delta function on identical basis to a scalar \( 1 \).
-The rule (R-MULK1) reflects the axiom that multiplying a zero vector results in zero.
-This rule is conditional on typing, which is checked during rewriting.
-Some rules, like (R-MULK11), require a deeper understanding, such as the preference for tensor products over multiplication.
-
-As a reference, the term rewriting system in DiracDec has been proven complete for all axioms, except for the sum symbol.
-The completeness result is derived from checking the confluence and termination of the system.
-Our rewriting rules are translations from DiracDec into the typed and abstract language, ensuring that the corresponding symbols in our system are also complete.
-
-We also have additional rules that handle summations. For example:
-\begin{align*}
-    & \quad i \text{ free in } t \Rightarrow \sum_{i \in \mathbf{U}(\sigma)} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (\delta_{i, t}.A) \ \reduce\  \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} A\{i/t\}, \\
-    & \quad \left( \sum_{i \in M} B \right) \cdot K \ \reduce\  \sum_{i \in M} (B \cdot K).
-\end{align*}
-The first rule eliminates delta expressions in summations, while the second rule pushes summations outside of inner products.
-While there is no guarantee of completeness for these rules, they work effectively in practice.
-
-\subsubsection{Variable Expansion}
-One important technique, revealed in the DiracDec work, is the expansion of variables, which is critical for proofs involving summation. For example:
-\[
-\frac{E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash K \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)}(\bra{i} \cdot K).\ket{i}} \quad \quad
-\frac{E[\Gamma] \vdash B : \BType(\sigma)}{E[\Gamma] \vdash B \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)}(B \cdot \ket{i}).\bra{i}},
-\]
-\[
-\frac{E[\Gamma] \vdash O : \OType(\sigma, \tau)}{E[\Gamma] \vdash O \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)} \sum_{j \in \mathbf{U}(\tau)}(\bra{i} \cdot O \cdot \ket{j}).(\ket{i} \cdot \bra{j})}.
-\]
-These rules transform variables into their symbolic summations based on their decomposition over the basis. 
-The rules are not terminating, therefore is applied recrusively once in the second step called \textit{variable expansion}.
-Nevertheless, we have found that applying the expansion only once for all variables is sufficient for normalization.
-\begin{lemma}
-    Let \( \textrm{expand}(e) \) denote the result of expanding all variables in \( e \) once.
-For all well-typed terms \( e \) in \( E[\Gamma] \), \( \textrm{expand}(\textrm{expand}(e)) \) and \( \textrm{expand}(e) \) have the same normal form.
-\end{lemma}
-\begin{proof}
-    Expanding a ket variable twice, for example, results in the following transformation:
-    \[
-    \sum_{i \in \mathbf{U}(\sigma)} (\bra{i} \cdot \sum_{j \in \mathbf{U}(\sigma)} (\bra{j} \cdot K) \cdot \ket{j}) \cdot \ket{i} \ \reduce\  \sum_{i \in \mathbf{U}(\sigma)} \sum_{j \in \mathbf{U}(\sigma)} (\bra{j} \cdot K \cdot \braket{i|j}) \cdot \ket{i},
-    \]
-    where the delta symbol elimination rule returns the term to its original form. The same holds for bra and operator terms.
-\end{proof}
-
-\subsection{Deciding Equational Theory \( E \)}
-
-In the Mathematica implementation of DiracDec, the equational theory \( E \) is decided using unification, which attempts to find a substitution for the summation bound variables that makes two expressions syntactically equivalent.
-This unification process iterates through all permutations of AC symbol arguments, and its complexity is factorial in the maximum number of AC symbol arguments.
-
-As the first improvement, we check \( \alpha \)-equivalence using the de Bruijn index~\cite{deBruijn1972lambda}, which replaces references to bound variable names with the distance from the lambda abstraction to the variable.
-For instance, the nominal lambda abstraction \( \lambda x.
-x \) is transformed into \( \lambda .
-0 \), while \( \lambda x.
-\lambda y.
-(x\ (y\ x)) \) is transformed into \( \lambda.\lambda.
-(1\ (0\ 1)) \).
-This transformation into the de Bruijn index is performed as the final step in the normalization process to check equivalence between terms with different bound variable names.
-
-The remaining axioms, such as AC-equivalence and SUM-SWAP, assert equivalence under permutations.
-A standard approach for deciding such equivalences is to normalize terms by sorting in a predefined order.
-For example, given the dictionary order \( a < b < c \), the term \( b + c + a \) (and any other AC-equivalent term) is normalized into \( a + b + c \).
-However, in our setting, two intertwined difficulties arise: how to assign an order to all terms in the language, and how to simultaneously sort for both axioms.
-
-Consider the following two equivalent terms:
-\[
-\sum_{i \in s_1} \sum_{j \in s_2} \bra{i} A \ket{j} \times \bra{j} B \ket{i}
-= 
-\sum_{i \in s_2} \sum_{j \in s_1} \bra{i} B \ket{j} \times \bra{j} A \ket{i}
-\]
-While these two terms are equivalent, directly sorting the elements of scalar multiplication using lexical order does not yield the same form.
-
-To address this issue, we propose an algorithm that sorts terms without considering bound variables.
-The key observation is that in a successive sum expression \( \sum_{i \in s_1} \cdots \sum_{j \in s_n} A \), the names and order of the bound variables \( i, \dots, j \) can be freely permuted.
-Therefore, all bound variables should be treated uniformly during sorting, and the order of summation can then be determined based on the position of the bound variables.
-
-In the example above, we first ignore the bound variables and sort the sum body into \( \bra{\bullet} A \ket{\bullet} \times \bra{\bullet} B \ket{\bullet} \).
-Then, we swap the summations such that the bound variable at the first \( \bullet \) position appears at the outermost position.
-The results will have the same de Bruijn normal form, namely \( \sum_{s_1} \sum_{s_2} \bra{\$1} A \ket{\$0} \times \bra{\$0} B \ket{\$1} \).
-
-To describe the algorithm in the following, we introduce two key notations.
-For a term \( e = f(a_1, a_2, \dots, a_n) \), \( \textrm{head}(e) \) denotes the function symbol \( f \), while \( \textrm{arg}(e, i) \) refers to the \( i \)-th argument \( a_i \) of the term.
-In this context, variables and constants are treated as functions with zero arguments.
-
-\begin{definition}[Order Without Bound Variables]
-Let \( \mathcal{B} \) represent the set of bound variables, with the assumption that all bound variables are unique.
-We also assume that a total order exists over all symbols.
-The relation \( e_1 =_\mathcal{B} e_2 \) holds if:
-\begin{itemize}
-    \item \( \textrm{head}(e_1) = \textrm{head}(e_2) \), and for all \( i \), \( \textrm{arg}(e_1, i) =_\mathcal{B} \textrm{arg}(e_2, i) \), or
-    \item \( e_1 \in \mathcal{B} \) and \(e_2 \in \mathcal{B}\).
-\end{itemize}
-
-The relation \( e_1 <_\mathcal{B} e_2 \) holds between two terms if:
-\begin{itemize}
-    \item $e_1 \notin \mathcal{B}$ and $e_2 \in \mathcal{B}$, or
-    \item $head(e_1) < head(e_2)$, or
-    \item $head(e_1) = head(e_2)$, and there exists $n$ with $arg(e_1, n) <_\mathcal{B} arg(e_2, n)$, where $arg(e_1, i) =_\mathcal{B} arg(e_2, i)$ for all $i < n$.
-\end{itemize}
-\end{definition}
-It can be shown that \( e_1 =_\mathcal{B} e_2 \) if and only if neither \( e_1 <_\mathcal{B} e_2 \) nor \( e_2 <_\mathcal{B} e_1 \) holds.
-The purpose of this ordering is to compare function symbols in a top-down manner while ignoring bound variables.
-The sorted order enables normalization of terms in terms of AC equivalence.
-\begin{definition}[Sort Transformation]
-    For a term $e$ with bound variable set $\mathcal{B}$,
-    The sort transformation sort(e) is defined as
-    \begin{itemize}
-        \item $e$, if $e$ is a variable or constant;
-        \item $\lambda x:T. \textrm{sort}(e')$, if $e \equiv \lambda x : T. e'$;
-        \item $\mu x. \textrm{sort}(e')$, if $e \equiv \mu x. \textrm{sort}(e')$; or
-        \item $f(\textrm{sort}(a_1), \cdots, \textrm{sort}(a_n))$, if $e \equiv f(a_1, \cdots, a_n)$.
-If $f$ is an AC symbol, then the order of sort($a_i$) is sorted ascendingly according to $<_\mathcal{B}$.
-    \end{itemize}
-\end{definition}
-
-After sorting, the next step is the \textit{swap transformation}, which arranges successive summations based on the order of bound variables.
-\begin{definition}[Swap Transformation]
-For a term \( e \) with a sorting result \( \textrm{sort}(e) \), the swap transformation proceeds by ordering all bound variables according to their first appearances, except in function definitions \( \lambda x \) and \( \mu x \).
-The swap transformation then reorders the successive summations accordingly.
-\end{definition}
-
-The algorithm that applies the sort and swap transformations, followed by de Bruijn normalization, successfully handles all equivalences in our benchmark.
-We formalize the completeness of this transformation in the following conjecture:
-
-\begin{conjecture}[Completeness of Transformation]
-    For any two terms \( e_1 \) and \( e_2 \) that are equivalent under AC equivalence, SUM-SWAP, and \( \alpha \)-equivalence, they will have the same form after sort, swap, and de Bruijn transformations.
-\end{conjecture}
-
-% The order depends on there occurances in the last sorting result. If no occurance, then the order will depend on the set (for sum) and the type (for lambda abstraction only).
-
-%%%%%%%%%%%%%%%%%%%%%%%
-
-% The idea is to assign an order to terms, which is independent on the bound variables. Because we can have terms with nested AC symbols.
-
-Lastly, we can prove that the equivalence established by this normalization procedure is sound with respect to the semantics.
-
-\begin{theorem}[Soundness]
-    For any well-formed context \( E[\Gamma] \) and well-typed expressions \( e_1 \) and \( e_2 \), if  $e_1$ and $e_2$ have the same normal form, then \( \sem{e_1} = \sem{e_2} \).
-\end{theorem}
-
-\begin{proof}
-    The soundness of the term rewriting procedure follows from the fact that each rewriting rule preserves equivalence.
-Furthermore, the operations in the sort and swap transformations respect the AC-equivalence and SUM-SWAP axioms.
-Finally, the de Bruijn normalization ensures soundness for \( \alpha \)-equivalence.
-\end{proof}
-
-% Labelled Dirac Notation
-
-\input{labelled}
-
-\section{Implementation and Case Study}
-
-The main purpose of this work is to build a practical tool that works well in checking Dirac notation equations.
-The refinements and extensions above concludes in our implementation called Dirace, a solver written in \CC.
-It has a parser built by ANTLR4, and scalar reasonings are powered by a Mathematica kernel.
-The user can use commands to make definitions and assumptions in the maintained context, conduct the normalization and equivalence checking, and obtain the rewriting trace output.
-This implementation
-is tested on the benchmark of the DiracDec work, and succeeds in proving most of them efficiently. 
-It can be used from the command line interactively, or can be integrated into other \CC\ projects as a library.
-
-\subsubsection{Project Structure}
-The project structure is illustrated in~\Cref{fig: dirace structure}.
-\texttt{ualg} is the module for universal algebra, defining basic concepts like terms and substitutions.
-It serves as the library for \texttt{dirace}, which are then utilized in the example benchmarks and the toplevel command line application.
-The components of \texttt{dirace} are as follows:
-\begin{itemize}
-    \item \texttt{symbols.cpp}: the reserved symbols and AC symbols;
-    \item \texttt{syntax\_theory.cpp}: syntax related algorithms, such as de Bruijn normalization and freeness of variables;
-    \item \texttt{calculus.cpp}: type checker and intergration with Mathematica;
-    \item \texttt{reduction.cpp}: all the rewriting rules and transformations;
-    \item \texttt{dirac\_parser.cpp}: parser for Dirac notation and Dirace commands;
-    \item \texttt{prover.cpp}: the prover that maintains the context and process commands like definition or equivalence checking.
-\end{itemize}
-
-\input{dirace.tex}
-
-The internal data structure for terms is a pointer-based syntax tree following the function application style:
-\[
-    \texttt{
-        term ::= ID | ID [term (, term)*].
-    }
-\]
-The syntax tree can be an identifier, or an application with an identifier as the function head, and several syntax trees as arguments.
-There are several Dirac notation terms and their corresponding syntax trees.
-\footnotesize{
-\begin{align*}
-    & X_1 + X_2 + X_3 && \texttt{ADD[X1, X2, X3]} 
-    \\
-    & \lambda x: \OType(T_1,  T_2). x^\dagger && \texttt{FUN[x, OTYPE[T1, T2], ADJ[x]]}
-    \\
-    & \sum_{i \in \mathbf{U}(T)} \ket{i} \bra{i} && \texttt{SUM[USET[T], FUN[i, BASIS[T], OUTER[KET[i], BRA[i]]]]}
-\end{align*}
-}
-The syntax tree structure is also compatible with the datatype of Mathematica. This improves the interoperability between Dirace and the Mathematica system, enabling them to work interleavingly.
-To improve usability, Dirace also supports many special syntacies for terms,  and most Dirac notation terms will be encoded in the natural way.
-Here are some examples for the parsing syntax.
-
-\begin{figure}
-    \center
-\begin{tabular}{c >{\centering\arraybackslash}p{4cm} l}
-    \hline
-    syntax & parsing result & explanation \\
-    \hline
-    \texttt{|e>} & \texttt{KET[e]} & the ket basis\\
-    \texttt{e1 + ... + en} & \texttt{ADD[e1, ..., en]} & the addition\\
-    \texttt{e1\ e2} & \texttt{COMPO[e1, e2]} & composition in Dirac notation \\
-    \texttt{e1\^{}*} & \texttt{CONJ[e1]} & scalar conjugation \\
-    \texttt{fun i : T => X} & \texttt{FUN[i, T, X]} & lambda abstraction \\
-    \hline
-\end{tabular}
-\end{figure}
-
-Finally, Dirace uses a prover to host the computation.
-The prover maintains a well-formed context $E[\Gamma]$, and processes commands to modify the context and conduct calculations.
-The commands are listed below.
-\begin{itemize}
-    \item \texttt{\textcolor{NavyBlue}{Def} ID := term.} It defines the \texttt{ID} as the \texttt{term}, using the \textbf{W-Def} typing rule.
-    \item \texttt{\textcolor{NavyBlue}{Var} ID := term.} It make an assumption of \texttt{ID} with the \texttt{term} as type, using the \textbf{W-AssumeE} typing rules.
-    \item \texttt{\textcolor{NavyBlue}{Check} term.} Type checking the \texttt{term} and output the result.
-    \item \texttt{\textcolor{NavyBlue}{Normalize} term.} Normalize the \texttt{term} using the algorithm introduced in~\Cref{sec: decide}.
-    \item \texttt{\textcolor{NavyBlue}{CheckEq} term \textcolor{NavyBlue}{with} term.} Check the equivalence of the two terms calculating and comparing their normal forms.
-\end{itemize}
-The prover will type check the terms for each command.
-We can also use \texttt{\textcolor{NavyBlue}{Normalize} term \textcolor{NavyBlue}{with trace}.} to output the proof trace during normalization.
-The proof trace is a sequence of records, including the rule or transformation appied, the position of application, and the pre- and post-transformation terms.
-The record helps understand the normalization procedure better, and can be turned into verified proofs in theorem provers in the future.
-
-\subsubsection{Use Case}
-Here we encode the motivating~\Cref{ex: motivating}, examine and explain how it is checked in Dirace. The encoding is shown below.
-
-    \begin{lstlisting}[style=dirace]
-Var T : INDEX. Var M : OTYPE[T, T].
-Def phi := idx T => Sum nv in USET[T], |(nv, nv)>.
-Var r1 : REG[T]. Var r2 : REG[T].
-CheckEq M_r1;r1 (phi T)_(r1, r2) with (TPO T T M)_r2;r2 (phi T)_(r1, r2).
-    \end{lstlisting}
-
-The first three lines use the \texttt{\textcolor{NavyBlue}{Var}} and \texttt{\textcolor{NavyBlue}{Def}} commands to set up the context for the Dirac notation.
-\texttt{T} is a type index, representing arbitrary Hilbert space types.
-\texttt{M} is assume to be an operator in the Hilbert space with type \texttt{T}.
-\texttt{phi} is defined as the maximally entangled state, depending on the bound variable \texttt{T} as index.
-\texttt{r1} and \texttt{r2} are register names for the two subsystems.
-
-In the left hand side of \texttt{\textcolor{NavyBlue}{CheckEq}} command, \texttt{M\_r1;r1} denotes the labelled notation $M_{r_1; r_1}$, and \texttt{(phi T)\_(r1, r2)} denotes the entangled state $\ket{\Phi}_{(r_1, r_2)}$.
-They are connected by a white space, which is parsed into the composition of Dirac notation, and will be reduced into the operator-ket multiplication after typing.
-The right hand side is interpreted similarly, except the defined symbol \texttt{TPO} in the environment:
-
-\begin{lstlisting}[style=dirace]
-Def TPO := idx sigma => idx tau => fun O : OTYPE[sigma, tau] => Sum i in USET[sigma], Sum j in USET[tau], (<i| O |j>).(|j> <i|).
-\end{lstlisting}
-
-The \texttt{TPO} symbol represents the transpose of operators, and encodes the formalization in~\Cref{ex: formalizing motivating}.
-Thanks to the design of environment and functions, many other commonly used symbols in Dirac notation are encoded and provided as defined symbols in Dirace.
-
-Within one second, the prover reports the result of equivalence with their common normal form:
-    \begin{lstlisting}[style=dirace]
-The two terms are equal.
-[Normalized Term] SUM[USET[T], FUN[BASIS[T], SUM[USET[T], FUN[BASIS[T], SCR[DOT[BRA[$\texttt{\$1}$], MULK[M, KET[$\texttt{\$0}$]]], LTSR[LKET[$\texttt{\$1}$, r1], LKET[$\texttt{\$0}$, r2]]]]]]] : DTYPE[RSET[r1, r2], RSET]
-    \end{lstlisting}
-
-The normal form is in the internal syntax tree format mentioned above. A more readable interpretation is:
-\[
-\sum_{\mathbf{U}(T)} \sum_{\mathbf{U}(T)} \bra{\$1}M\ket{\$ 0} . \ket{\$1}_{r_1} \otimes \ket{\$0}_{r_2} : \DType(\{r_1, r_2\}, \emptyset).
-\]
-Here $\$0$ and $\$1$ are de Bruijn indices. The result is a ket on the $\{r_1, r_2\}$ system as expected, and follows pattern proposed in~\Cref{sec: labelled}.
-
-\subsubsection{Benchmark performance}
-
-To evaluate Dirace, we first test the examples from DiracDec benchmark and make a comparison.
-The experiments are carried out using a MacBook Pro with M3 Max chip.
-
-\begin{figure}
-    \center
-    \begin{tabular}{c|c c c|c c c}
-        \hline
-        \multirow{2}{*}{source} & \multicolumn{3}{c|}{DiracDec} & \multicolumn{3}{c}{Dirace} \\
-        \cline{2-7}
-                                 & expressable & success & time(s)           & expressable & success & time(s)                 \\
-        \hline
-        textbook(QCQI)          & 18          & 18        &    1.02        &    18      & 18          &   0.82      \\
-        CoqQ                    & 162          & 156       &    48.69       &   158     &  158   &     9.74     \\
-        circuits                 & 2          & 2       &    17.67       &   3     &  2   &     1.4     \\
-        research paper                & 4          & 4         &  59.53       &   4    & 4       &  0.73     \\
-        \hline
-    \end{tabular}        
-    \caption{For DiracDec, examples that cannot be decided within 60 seconds are not included.}
-\end{figure}
-
-The timing of Dirace does not include the initialization of Mathematica kernel link, which takes about 3 seconds in the beginning.
-As to expressibility, the language for DiracDec has the support for projectors $\texttt{fst}$ and $\texttt{snd}$ on basis pairs, satisfying $\texttt{fst} (s, t) = s$ and $\texttt{snd} (s, t) = t$.
-We found this feature is almost not used, so we removed the support.
-As a result, Dirace encodes 158 examples for the CoqQ part, 4 less than DiracDec.
-For decidability, we improved the rewriting rules about sum for Dirace, and proved several more examples failed by DiracDec.
-The main difference is about their time efficiency.
-Because of our algorithm to decide AC-equivalence and SUM-SWAP, Dirace has a significant efficiency improvement, especially on those ``computational examples'' mentioned in the DiracDec paper.
-One typical example comes from the paper by Jens about the equivalence of operators for qubits.
-The system has to decompose the term on the concrete $\ket{0}$ and $\ket{1}$ basis, resulting a lot of addition elements.
-It takes DiracDec about one minute, but Dirace solves it within one second.
-
-We also built an example benchmark for labelled Dirac notation.
-\yx{We need examples.}
-
-\section{Related Work}
-
-Automated theorem proving has seen significant advancements in recent years, particularly through the development of satisfiability modulo theories (SMT) solvers,
-% SMT solvers extend propositional satisfiability by integrating theories such as arithmetic, arrays, and bit-vectors, making them suitable for solving a wide range of verification problems in both hardware and software domains.
-including prominent tools like Z3. These solvers have become essential in various fields such as formal verification, synthesis, and model checking. 
-Equational reasoning is another crucial area of research within automated theorem proving, which focuses on solving problems that involve equations between terms in an algebraic structure. 
-Equational provers, such as Vampire and E, have played a pivotal role in addressing the challenge of proving equations in first-order logic, employing sophisticated algorithms like superposition and term rewriting.
-
-Formal verification of quantum computation is receiving increasing attentions during these years. See~\cite{Lewis2023} for a comprehensive review. 
-Verification frameworks in Coq include the foundational formalization CoqQ and quantum circuit language QWIRE. 
-Verifications of quantum programs are also considered, such as the Hoare logic based methods [] and model checking based methods [].
-The equational reasoning of Dirac notation is crucial for establishing property proofs in these works.
-Verification theoreis and tools based on other languages are also proposed, such as PyZX for reasoning and simplification of ZX-calculus.
-
-\section{Conclusion and Future Work}
-Based on the first equational reasoning tool for Dirac notation called DiracDec, this work improves and extends the theory for practical applications, and provides the solver Dirace.
-Experiments show that the tool demonstrates advantages in decidability, efficiency and usability.
-
-We expect Dirace to have applications in areas like quantum program verification or proofs of post-quantum cryptography protocols in the future.
-One promising following up is to connect Dirace with theorem provers like Coq.
-It involves transforming theorem prover expressions into Dirace, and verify the proof trace of Dirace in theorem provers.
-Besids, most quantum program verifiers nowadays depend on matrix calculations.
-Dirace can serve as the replacement for matrix methods to enable symbolic deductions.
+\input{04related_conclusion.tex}
 

### 2025-01-27 01intro_preliminary.tex
rearranged

@@
 
+\subsubsection{Use Case}
+\yx{simplify to fit in the motivating section}
+Here we encode the motivating~\Cref{ex: motivating}, examine and explain how it is checked in D-Hammer. The encoding is shown below.
+
+    \begin{lstlisting}[style=dhammer]
+Var T : INDEX. Var M : OTYPE[T, T].
+Def phi := idx T => Sum nv in USET[T], |(nv, nv)>.
+Var r1 : REG[T]. Var r2 : REG[T].
+CheckEq M_r1;r1 (phi T)_(r1, r2) with (TPO T T M)_r2;r2 (phi T)_(r1, r2).
+    \end{lstlisting}
+
+The first three lines use the \texttt{\textcolor{NavyBlue}{Var}} and \texttt{\textcolor{NavyBlue}{Def}} commands to set up the context for the Dirac notation.
+\texttt{T} is a type index, representing arbitrary Hilbert space types.
+\texttt{M} is assume to be an operator in the Hilbert space with type \texttt{T}.
+\texttt{phi} is defined as the maximally entangled state, depending on the bound variable \texttt{T} as index.
+\texttt{r1} and \texttt{r2} are register names for the two subsystems.
+
+In the left hand side of \texttt{\textcolor{NavyBlue}{CheckEq}} command, \texttt{M\_r1;r1} denotes the labelled notation $M_{r_1; r_1}$, and \texttt{(phi T)\_(r1, r2)} denotes the entangled state $\ket{\Phi}_{(r_1, r_2)}$.
+They are connected by a white space, which is parsed into the composition of Dirac notation, and will be reduced into the operator-ket multiplication after typing.
+The right hand side is interpreted similarly, except the defined symbol \texttt{TPO} in the environment:
+
+\begin{lstlisting}[style=dhammer]
+Def TPO := idx sigma => idx tau => fun O : OTYPE[sigma, tau] => Sum i in USET[sigma], Sum j in USET[tau], (<i| O |j>).(|j> <i|).
+\end{lstlisting}
+
+The \texttt{TPO} symbol represents the transpose of operators, and encodes the formalization in~\Cref{ex: formalizing motivating}.
+Thanks to the design of environment and functions, many other commonly used symbols in Dirac notation are encoded and provided as defined symbols in D-Hammer.
+
+Within one second, the prover reports the result of equivalence with their common normal form:
+    \begin{lstlisting}[style=dhammer]
+The two terms are equal.
+[Normalized Term] SUM[USET[T], FUN[BASIS[T], SUM[USET[T], FUN[BASIS[T], SCR[DOT[BRA[$\texttt{\$1}$], MULK[M, KET[$\texttt{\$0}$]]], LTSR[LKET[$\texttt{\$1}$, r1], LKET[$\texttt{\$0}$, r2]]]]]]] : DTYPE[RSET[r1, r2], RSET]
+    \end{lstlisting}
+
+The normal form is in the internal syntax tree format mentioned above. A more readable interpretation is:
+\[
+\sum_{\mathbf{U}(T)} \sum_{\mathbf{U}(T)} \bra{\$1}M\ket{\$ 0} . \ket{\$1}_{r_1} \otimes \ket{\$0}_{r_2} : \DType(\{r_1, r_2\}, \emptyset).
+\]
+Here $\$0$ and $\$1$ are de Bruijn indices. The result is a ket on the $\{r_1, r_2\}$ system as expected, and follows pattern proposed in~\Cref{sec: labelled}.
+
 \subsubsection{Dirac Notation}

### 2025-01-27 02language_design.tex
rearranged

--- 2025-01-27 02techniques.tex
@@
 \end{example}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Algorithm for Deciding Dirac Notation Equations}
-\label{sec: decide}
-
-We utilize the term rewriting technique to decide the equivalence of Dirac notation, based on its axiomatic semantics.
-The process involves normalizing two terms and checking whether they have identical syntax.
-This normalization is achieved through matching and substitution in universal algebra, using a set of predefined rewriting rules.
-
-As mentioned in the semantics section, the axioms in \( E \) cannot be fully decided through term rewriting.
-DiracDec decides \( E \) by examining all possible permutations in the rewriting result.
-In this work, we introduce a more efficient approach by incorporating sorting algorithms into the normalization procedure.
-
-The normalization procedure consists of the following steps.
-\begin{enumerate}
-    \item \textbf{First Rewritings}: Expand definitions and simplify expressions.
-    \item \textbf{Variable Expansion}: Consider scalar expressions to ensure completeness.
-    \item \textbf{Second Rewritings}: Normalize terms modulo \( E \).
-    \item \textbf{Sorting Without Bound Variables}: Normalize AC-equivalence.
-    \item \textbf{Swap Successive Summations}: Normalize SUM-SWAP equivalences.
-    \item \textbf{De Bruijn Normalization}: Normalize \( \alpha \)-equivalence.
-\end{enumerate}
-
-\subsection{Normalization Modulo \( E \) by Term Rewriting}
-
-Term rewriting rules, represented as \( l \ \reduce\ r \), are applied recursively to normalize terms.
-In each step, subterms matching the left-hand side \( l \) of a rule are replaced with the corresponding right-hand side \( r \).
-The procedure terminates when no further rewritings can be made.
-A comprehensive list of rewriting rules can be found in~\Cref{sec: rewriting rules}. Below are some key examples to illustrate the design:
-
-One of our optimizations is using functions with indefinite arities. 
-Therefore, we use a flattening rule to handle associativity with AC symbols:
-\[
-a_1 + \cdots + (b_1 + \cdots + b_m) + \cdots + a_n \ \reduce\ a_1 + \cdots + b_1 + \cdots + b_m + \cdots + a_n.
-\]
-Commutativity is handled later in the sorting step. Many of the rewriting rules are directly derived from the equational axioms, such as:
-\begin{align*}
-    & \textrm{(R-DOT6)} && \bra{s} \cdot \ket{t} \ \reduce\  \delta_{s, t}, \\
-    & \textrm{(R-DELTA0)} && \delta_{s, s} \ \reduce\  1, \\
-    & \textrm{(R-MULK1)} && O : \OType(\sigma, \tau) \Rightarrow O \cdot \mathbf{0}_{\KType(\tau)} \ \reduce\  \mathbf{0}_{\KType(\sigma)}, \\
-    & \textrm{(R-MULK11)} && (O_1 \otimes O_2) \cdot (K_1 \otimes K_2) \ \reduce\  (O_1 \cdot K_1) \otimes (O_2 \cdot K_2).
-\end{align*}
-Some of these directions are obvious.
-For example, (R-DOT6) states that the inner product of two basis vectors is reduced to a delta expression, and (R-DELTA0) transforms a delta function on identical basis to a scalar \( 1 \).
-The rule (R-MULK1) reflects the axiom that multiplying a zero vector results in zero.
-This rule is conditional on typing, which is checked during rewriting.
-Some rules, like (R-MULK11), require a deeper understanding, such as the preference for tensor products over multiplication.
-
-As a reference, the term rewriting system in DiracDec has been proven complete for all axioms, except for the sum symbol.
-The completeness result is derived from checking the confluence and termination of the system.
-Our rewriting rules are translations from DiracDec into the typed and abstract language, ensuring that the corresponding symbols in our system are also complete.
-
-We also have additional rules that handle summations. For example:
-\begin{align*}
-    & \quad i \text{ free in } t \Rightarrow \sum_{i \in \mathbf{U}(\sigma)} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (\delta_{i, t}.A) \ \reduce\  \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} A\{i/t\}, \\
-    & \quad \left( \sum_{i \in M} B \right) \cdot K \ \reduce\  \sum_{i \in M} (B \cdot K).
-\end{align*}
-The first rule eliminates delta expressions in summations, while the second rule pushes summations outside of inner products.
-While there is no guarantee of completeness for these rules, they work effectively in practice.
-
-\subsubsection{Variable Expansion}
-One important technique, revealed in the DiracDec work, is the expansion of variables, which is critical for proofs involving summation. For example:
-\[
-\frac{E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash K \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)}(\bra{i} \cdot K).\ket{i}} \quad \quad
-\frac{E[\Gamma] \vdash B : \BType(\sigma)}{E[\Gamma] \vdash B \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)}(B \cdot \ket{i}).\bra{i}},
-\]
-\[
-\frac{E[\Gamma] \vdash O : \OType(\sigma, \tau)}{E[\Gamma] \vdash O \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)} \sum_{j \in \mathbf{U}(\tau)}(\bra{i} \cdot O \cdot \ket{j}).(\ket{i} \cdot \bra{j})}.
-\]
-These rules transform variables into their symbolic summations based on their decomposition over the basis. 
-The rules are not terminating, therefore is applied recrusively once in the second step called \textit{variable expansion}.
-Nevertheless, we have found that applying the expansion only once for all variables is sufficient for normalization.
-\begin{lemma}
-    Let \( \textrm{expand}(e) \) denote the result of expanding all variables in \( e \) once.
-For all well-typed terms \( e \) in \( E[\Gamma] \), \( \textrm{expand}(\textrm{expand}(e)) \) and \( \textrm{expand}(e) \) have the same normal form.
-\end{lemma}
-\begin{proof}
-    Expanding a ket variable twice, for example, results in the following transformation:
-    \[
-    \sum_{i \in \mathbf{U}(\sigma)} (\bra{i} \cdot \sum_{j \in \mathbf{U}(\sigma)} (\bra{j} \cdot K) \cdot \ket{j}) \cdot \ket{i} \ \reduce\  \sum_{i \in \mathbf{U}(\sigma)} \sum_{j \in \mathbf{U}(\sigma)} (\bra{j} \cdot K \cdot \braket{i|j}) \cdot \ket{i},
-    \]
-    where the delta symbol elimination rule returns the term to its original form. The same holds for bra and operator terms.
-\end{proof}
-
-\subsection{Deciding Equational Theory \( E \)}
-
-In the Mathematica implementation of DiracDec, the equational theory \( E \) is decided using unification, which attempts to find a substitution for the summation bound variables that makes two expressions syntactically equivalent.
-This unification process iterates through all permutations of AC symbol arguments, and its complexity is factorial in the maximum number of AC symbol arguments.
-
-As the first improvement, we check \( \alpha \)-equivalence using the de Bruijn index~\cite{deBruijn1972lambda}, which replaces references to bound variable names with the distance from the lambda abstraction to the variable.
-For instance, the nominal lambda abstraction \( \lambda x.
-x \) is transformed into \( \lambda .
-0 \), while \( \lambda x.
-\lambda y.
-(x\ (y\ x)) \) is transformed into \( \lambda.\lambda.
-(1\ (0\ 1)) \).
-This transformation into the de Bruijn index is performed as the final step in the normalization process to check equivalence between terms with different bound variable names.
-
-The remaining axioms, such as AC-equivalence and SUM-SWAP, assert equivalence under permutations.
-A standard approach for deciding such equivalences is to normalize terms by sorting in a predefined order.
-For example, given the dictionary order \( a < b < c \), the term \( b + c + a \) (and any other AC-equivalent term) is normalized into \( a + b + c \).
-However, in our setting, two intertwined difficulties arise: how to assign an order to all terms in the language, and how to simultaneously sort for both axioms.
-
-Consider the following two equivalent terms:
-\[
-\sum_{i \in s_1} \sum_{j \in s_2} \bra{i} A \ket{j} \times \bra{j} B \ket{i}
-= 
-\sum_{i \in s_2} \sum_{j \in s_1} \bra{i} B \ket{j} \times \bra{j} A \ket{i}
-\]
-While these two terms are equivalent, directly sorting the elements of scalar multiplication using lexical order does not yield the same form.
-
-To address this issue, we propose an algorithm that sorts terms without considering bound variables.
-The key observation is that in a successive sum expression \( \sum_{i \in s_1} \cdots \sum_{j \in s_n} A \), the names and order of the bound variables \( i, \dots, j \) can be freely permuted.
-Therefore, all bound variables should be treated uniformly during sorting, and the order of summation can then be determined based on the position of the bound variables.
-
-In the example above, we first ignore the bound variables and sort the sum body into \( \bra{\bullet} A \ket{\bullet} \times \bra{\bullet} B \ket{\bullet} \).
-Then, we swap the summations such that the bound variable at the first \( \bullet \) position appears at the outermost position.
-The results will have the same de Bruijn normal form, namely \( \sum_{s_1} \sum_{s_2} \bra{\$1} A \ket{\$0} \times \bra{\$0} B \ket{\$1} \).
-
-To describe the algorithm in the following, we introduce two key notations.
-For a term \( e = f(a_1, a_2, \dots, a_n) \), \( \textrm{head}(e) \) denotes the function symbol \( f \), while \( \textrm{arg}(e, i) \) refers to the \( i \)-th argument \( a_i \) of the term.
-In this context, variables and constants are treated as functions with zero arguments.
-
-\begin{definition}[Order Without Bound Variables]
-Let \( \mathcal{B} \) represent the set of bound variables, with the assumption that all bound variables are unique.
-We also assume that a total order exists over all symbols.
-The relation \( e_1 =_\mathcal{B} e_2 \) holds if:
-\begin{itemize}
-    \item \( \textrm{head}(e_1) = \textrm{head}(e_2) \), and for all \( i \), \( \textrm{arg}(e_1, i) =_\mathcal{B} \textrm{arg}(e_2, i) \), or
-    \item \( e_1 \in \mathcal{B} \) and \(e_2 \in \mathcal{B}\).
-\end{itemize}
-
-The relation \( e_1 <_\mathcal{B} e_2 \) holds between two terms if:
-\begin{itemize}
-    \item $e_1 \notin \mathcal{B}$ and $e_2 \in \mathcal{B}$, or
-    \item $head(e_1) < head(e_2)$, or
-    \item $head(e_1) = head(e_2)$, and there exists $n$ with $arg(e_1, n) <_\mathcal{B} arg(e_2, n)$, where $arg(e_1, i) =_\mathcal{B} arg(e_2, i)$ for all $i < n$.
-\end{itemize}
-\end{definition}
-It can be shown that \( e_1 =_\mathcal{B} e_2 \) if and only if neither \( e_1 <_\mathcal{B} e_2 \) nor \( e_2 <_\mathcal{B} e_1 \) holds.
-The purpose of this ordering is to compare function symbols in a top-down manner while ignoring bound variables.
-The sorted order enables normalization of terms in terms of AC equivalence.
-\begin{definition}[Sort Transformation]
-    For a term $e$ with bound variable set $\mathcal{B}$,
-    The sort transformation sort(e) is defined as
-    \begin{itemize}
-        \item $e$, if $e$ is a variable or constant;
-        \item $\lambda x:T. \textrm{sort}(e')$, if $e \equiv \lambda x : T. e'$;
-        \item $\mu x. \textrm{sort}(e')$, if $e \equiv \mu x. \textrm{sort}(e')$; or
-        \item $f(\textrm{sort}(a_1), \cdots, \textrm{sort}(a_n))$, if $e \equiv f(a_1, \cdots, a_n)$.
-If $f$ is an AC symbol, then the order of sort($a_i$) is sorted ascendingly according to $<_\mathcal{B}$.
-    \end{itemize}
-\end{definition}
-
-After sorting, the next step is the \textit{swap transformation}, which arranges successive summations based on the order of bound variables.
-\begin{definition}[Swap Transformation]
-For a term \( e \) with a sorting result \( \textrm{sort}(e) \), the swap transformation proceeds by ordering all bound variables according to their first appearances, except in function definitions \( \lambda x \) and \( \mu x \).
-The swap transformation then reorders the successive summations accordingly.
-\end{definition}
-
-The algorithm that applies the sort and swap transformations, followed by de Bruijn normalization, successfully handles all equivalences in our benchmark.
-We formalize the completeness of this transformation in the following conjecture:
-
-\begin{conjecture}[Completeness of Transformation]
-    For any two terms \( e_1 \) and \( e_2 \) that are equivalent under AC equivalence, SUM-SWAP, and \( \alpha \)-equivalence, they will have the same form after sort, swap, and de Bruijn transformations.
-\end{conjecture}
-
-% The order depends on there occurances in the last sorting result. If no occurance, then the order will depend on the set (for sum) and the type (for lambda abstraction only).
-
-%%%%%%%%%%%%%%%%%%%%%%%
-
-% The idea is to assign an order to terms, which is independent on the bound variables. Because we can have terms with nested AC symbols.
-
-Lastly, we can prove that the equivalence established by this normalization procedure is sound with respect to the semantics.
-
-\begin{theorem}[Soundness]
-    For any well-formed context \( E[\Gamma] \) and well-typed expressions \( e_1 \) and \( e_2 \), if  $e_1$ and $e_2$ have the same normal form, then \( \sem{e_1} = \sem{e_2} \).
-\end{theorem}
-
-\begin{proof}
-    The soundness of the term rewriting procedure follows from the fact that each rewriting rule preserves equivalence.
-Furthermore, the operations in the sort and swap transformations respect the AC-equivalence and SUM-SWAP axioms.
-Finally, the de Bruijn normalization ensures soundness for \( \alpha \)-equivalence.
-\end{proof}
-
-% Labelled Dirac Notation
-
-\section{Labelled Dirac Notation}
-\label{sec: labelled}
-Labelled Dirac notation uses register names to indicate the quantum system of vectors and operators.
-This allows us to express and reason aboud the states and operations locally, without referring to the whole system.
-For instance, assume $Q$ and $R$ are two registers, we have
-\[
-    M_Q \cdot \ket{\Phi}_{(Q, R)} = ((M \otimes I) \cdot \ket{\Phi})_{(Q, R)}.
-\]
-On the left hand side, the state of two subsystems is $\ket{\Phi}$, and we apply quantum operation $M$ on the system $Q$.
-It is equivalent to extend the operation using identity operators on other subsystems (i.e.
-the cylinder extension), and consider the application in the whole system.
-
-In this section, we introduce the language and typing of registers and labelled Dirac notation, and demonstrate how to transform the equivalence problem into that for the Dirac notation studied above.
-The first new notion is quantum registers, and we assume they are constructed from a set $\cR$.
-\begin{definition}[quantum registers]
-  \begin{align*}
-    R ::= r\in\cR \mid (R, R).
-  \end{align*}
-\end{definition}
-
-Registers can be composed by pairs of $(R, R)$, and this structure corresponds to the structure of tensor product spaces in Dirac notation.
-To reason about the registers, we define their variable set as the enumeration of all register symbols involved.
-
-\begin{definition}[register variable set]
-The variable set of a register is defined inductively by:
-\begin{itemize}
-    \item $\var(R) = \{r\}$, if $R\equiv r$; or
-    \item $\var(R) = \var(R_1) \cup\var(R_2)$, if $R\equiv (R_1,R_2)$.
-\end{itemize}
-\end{definition}
-
-% \textbf{Remark: } Set operations: $\cup$ for union; $\cap$ for intersection; $\setminus$ for difference. So,
-% % \begin{itemize}
-% $ S_1 \cap S_2 \equiv S_1 \cup S_2 \setminus (S_1 \setminus S_2) \setminus (S_2 \setminus S_1) $.
-% \end{itemize}
-
-% The labelled Dirac notation is an extension on the existing type and syntax.
-\begin{definition}[labelled Dirac notation]
-  The \textbf{labelled Dirac notation} includes all Dirac notation symbols and the generators defined below.
-  Here, $s\subseteq \cR$ is a register variable set.
-  \begin{align*}
-    T & ::= \DType(s,s) \mid \mathsf{Reg}(\sigma) \\
-    e & ::= R \mid |i\>_r \mid {}_r\<i| \mid e_R \mid e_{R;R} \mid
-    e \otimes e \otimes \cdots \otimes e \mid e \cdot e
-  \end{align*}
-\end{definition}
-$\DType(s_1, s_2)$ is the unified type for all labelled Dirac notation, where $s_1$ indicates the codomain systems and $s_2$ indicates the domain systems.
-For instance, labelled ket has type $\DType(s_1, \emptyset)$, and labelled bra has type $\DType(\emptyset, s_2)$.
-$\reg(\sigma)$ are types for registers $R$, and the index $\sigma$ indicates the type of Hilbert space represented by the register.
-Terms also include the labelled notation $e_R$ for bra, ket and $e_{R;R}$ for operators.
-We introduce new dot and tensor product symbols for labelled Dirac notation.
-In labelled Dirac notation, the structure of tensor product does not matter, therefore $\otimes$ is an AC symbol.
-Following the unified type $\DType(s,s)$, all kinds of multiplications are represented by the same dot product $e \cdot e$.
-Finally, $\ket{i}_r$ and ${}_r\bra{i}$ are labelled basis for the normal form of labelled Dirac notation, where $r$ are registers symbols in $\mathcal{R}$.
-
-\subsubsection{Typing rules}
-Some typing rules are introduced here.
-The rule for $\DType(s_1, s_2)$ requires that all registers in variable set $s_1$ and $s_2$ are well-typed.
-The rule for $K_R$ demonstrates how a register label is added to the Dirac notation, and the rule for $D^\dagger$ shows that labelled Dirac notation also have symbols for calculation, such as the adjoint.
-\[
-    \frac{E[\Gamma] \vdash  \sigma : \Index}{E[\Gamma] \vdash \reg(\sigma) : \Type}
-    \qquad
-    \frac{E[\Gamma] \vdash r : \reg(\sigma_r) \text{ for all $r$ in $s_1$ and $s_2$} }{E[\Gamma] \vdash \DType(s_1, s_2) : \textsf{Type}} 
-\]
-\[
-    \frac{E[\Gamma] \vdash R : \reg(\sigma)\qquad E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash K_R : \DType(\var R, \emptyset)}
-    \qquad
-    \frac{E[\Gamma] \vdash D : \DType(s_1,s_2)}{E[\Gamma] \vdash D^\dagger : \DType(s_2,s_1)}
-\]
-The dot and the tensor product symbols are different from those in unlabelled Dirac notation.
-Since the goal of labels is to replace the order and structure of tensor products by the reference to registers, the tensor product becomes an AC symbol.
-The typing still checks whether the component subsystems are disjoint with each other.
-\[
-    \frac{
-        E[\Gamma] \vdash D_i : \DType(s_i,s_i') \qquad
-        \bigcap_i s_i = \emptyset \qquad
-        \bigcap_i s_i' = \emptyset
-    }
-    {E[\Gamma] \vdash D_1 \otimes \cdots \otimes D_i : \DType(\bigcup_i s_i, \bigcup_i s_i')}.
-\]
-As for the dot product, the disjointness is considered except registers contracted by multiplication.
-\[
-    \qquad
-    \frac{
-        \begin{aligned}
-            E[\Gamma] \vdash D_1 : \DType(s_1,s_1') \\
-            E[\Gamma] \vdash D_2 : \DType(s_2,s_2')
-        \end{aligned}
-        \qquad 
-        \begin{aligned}
-            s_1 \cap s_2 \backslash s_1' = \emptyset \\
-            s_2' \cap s_1' \backslash s_2 = \emptyset
-        \end{aligned}
-    }
-    {E[\Gamma] \vdash D_1\cdot D_2 : \DType(s_1 \cup (s_2\backslash s_1'), s_2' \cup (s_1'\backslash s_2))}.
-\]
-
-\subsubsection{Normalization}
-The normalization procedure of Dirac notation is extended to check equivalence with labels.
-We add rules to the term rewriting system, which in general try to represent labelled Dirac notation with labelled basis and scalar coefficients.
-The first step is the label elimination.
-Take operator as an example:
-\begin{align*}
-    & O_{R,R'} \ \reduce\ \sum_{i_{r_1}\in\bU(\sigma_{r_1})}\cdots \sum_{i_{r_n}\in\bU(\sigma_{r_n})}
-    \sum_{i_{r'_1}\in\bU(\sigma_{r'_1})}\cdots \sum_{i_{r'_{n'}}\in\bU(\sigma_{r'_{n'}})} \\
-    & \qquad (\<i_R|\cdot O\cdot |i_{R'}\>).(|i_{r_1}\>_{r_i}\otimes\cdots\otimes|i_{r_n}\>_{r_n} \otimes {}_{r'_1}\<i_{r'_1}|\otimes\cdots\otimes{}_{r'_{n'}}\<i_{r'_{n'}}|).
-\end{align*}
-The rules for $e_R$ (ket and bra) are similar. This step reduces all labelled terms $e_R$ or $e_{R;R}$.
-Other symbols on labelled Dirac notation are also reduced by rules like $(D_1 \cdot D_2)^\dagger \ \reduce\ D_2^\dagger \cdot D_1^\dagger$.
-The final step operates on sum and dot product. They will lift summation to the outside, and eliminate the bra-ket pairs whenever possible.
-\begin{align*}
-    & \textrm{(R-SUM-PUSHD0)}
-    && X_1 \otimes \cdots (\sum_{i \in M} D) \cdots \otimes X_2\ \reduce\ \sum_{i \in M} (X_1 \otimes \cdots D \cdots \otimes X_n) \\
-    % %
-    % & \textrm{(R-SUM-PUSHD1)}
-    % && (\sum_{i \in M} D_1) \cdot D_2 \ \reduce\ \sum_{i \in M} (D_1 \cdot D_2) \\
-    %
-    & \textrm{(R-L-SORT0)}
-    && A : \DType(s_1, s_2), B : \DType(s_1', s_2'), s_2 \cap s_1'=\emptyset \Rightarrow A \cdot B \ \reduce\ A \otimes B \\
-    %
-    & \textrm{(R-L-SORT1)}
-    && {}_r\bra{i}\cdot\ket{j}_r \ \reduce\ \delta_{i, j} \\
-    %
-    & \textrm{(R-L-SORT2)}
-    && {}_r\bra{i}\cdot(Y_1 \otimes \cdots \otimes \ket{j}_r \otimes \cdots \otimes Y_m) \ \reduce\ \delta_{i, j}.(Y_1  \otimes \cdots \otimes Y_m)
-    %
-\end{align*}
-% These rules are added to the rewriting system in~\Cref{sec: decide} and executed together.
-In the end, if there are no variables of $\DType(s_1, s_2)$, the expression will be reduced to the addition of big operator sum, and each sum body is labelled basis with Dirac notation scalar coefficients:
-\[
-    \sum_{i}\cdots\sum_{j} a_1 . (\ket{i}_{p} \otimes \cdots \otimes \bra{j}_{q})
-    + \cdots +
-    \sum_{k}\cdots\sum_{l} a_m . (\ket{k}_{r} \otimes \cdots \otimes \bra{l}_{s})
-\]
-In this stage, we only need to check the scalars to decide the equivalence of labelled Dirac notation.
-
-% \begin{lemma}[normal form]
-%     \label{lem: labelled normal form}
-%     For a well-typed term $e$ in $E[\Gamma]$, if $e$ does not contain variables of $\DType(s_1, s_2)$ type, the normal form of $e$ will be
-%     \[
-%     \sum_{i}\cdots\sum_{j} a_1 . (\ket{i}_{p} \otimes \cdots \otimes \bra{j}_{q})
-%     + \cdots +
-%     \sum_{k}\cdots\sum_{l} a_m . (\ket{k}_{r} \otimes \cdots \otimes \bra{l}_{s})
-%     \]
-%     Where $a_i$ are scalar typed Dirac notation.
-% \end{lemma}
-% \begin{proof}
-%     Labelled Dirac notation of $e_R$, $e_R;R$, 
-% \end{proof}

### 2025-01-27 03labelled.tex
rearranged

--- 2025-01-27 02techniques.tex
@@
-\section{Language, Typing and Semantics}
-The first step is to formalize the language for Dirac notation.
-In DiracDec, the language has a concrete design, where the same syntax for different types corresponds to distinct symbols.
-Our goal is to transition from this concrete design to a more abstract one, which aligns more closely with the conventional Dirac notation we use and simplifies the term rewriting system.
-
-To achieve this, we organize our language into three layers: the index, the type, and the term.
-Terms represent concrete instances such as kets, bras, and operators, which will be typed and checked.
-The index represents classical data types and appears in type expressions to differentiate between various Hilbert spaces and sets.
-
-\begin{definition}[Index Syntax]
-    The syntax for type indices is:
-    \begin{align*}
-        \sigma ::=\ & x \mid \sigma_1 \times \sigma_2.
-    \end{align*}
-\end{definition}
-Here, \( x \) is a variable, and \( \sigma_1 \times \sigma_2 \) represents the product type for tensor product spaces or Cartesian product sets.
-
-\begin{definition}[Type Syntax]
-    The syntax for Dirac notation types is:
-    \begin{align*}
-        T ::=\ & \Basis(\sigma) \mid \SType \mid \KType(\sigma) \mid \BType(\sigma) \mid \OType(\sigma_1, \sigma_2) \mid T_1 \to T_2 \mid \forall x.T \mid \SET(\sigma).
-    \end{align*}
-\end{definition}
-\( \Basis(\sigma) \) denotes the type for basis elements in the index \( \sigma \).
-\( \SType \) represents scalars, while \( \KType(\sigma) \) and \( \BType(\sigma) \) refer to ket and bra types in the Hilbert space \( \sigma \), respectively.
-\( \OType(\sigma_1, \sigma_2) \) represents linear operators with \( \sigma_2 \) as the domain and \( \sigma_1 \) as the codomain.
-\( \SET(\sigma) \) refers to the type of subsets of \( \sigma \), used to denote the values of bound variables in summations.
-The remaining two constructs define function types: \( T_1 \to T_2 \) represents normal functions that take a \( T_1 \)-type argument and return a \( T_2 \)-type term, while \( \forall x.
-T \) represents index functions that take an index argument \( x \) and produce a \( T \)-type term, which may depend on \( x \).
-Index functions are essential for defining polymorphic transformations across Hilbert spaces.
-
-\begin{definition}[Term Syntax]
-    The syntax for Dirac notation terms is:
-    \begin{align*}
-        e ::=\ & x \mid \lambda x : T.e \mid \mu x.e \mid e_1\ e_2 \mid (e_1, e_2) \\
-        & |\ 0 \mid 1 \mid \ADDS(e_1, \cdots, e_n) \mid e_1 \times \cdots \times e_n \mid e^* \mid \delta_{e_1, e_2} \mid \DOT(e_1, e_2) \\
-        & |\ \ZEROK(\sigma) \mid \ZEROB(\sigma) \mid \ZEROO(\sigma_1, \sigma_2) \mid \ONEO(\sigma) \\
-        & |\ \ket{e} \mid \bra{t} \mid e^\dagger \mid e_1.e_2 \mid \ADD(e_1, \cdots, e_n) \mid e_1 \otimes e_2 \\
-        & |\ \MULK(e_1, e_2) \mid \MULB(e_1, e_2) \mid \OUTER(e_1, e_2) \mid \MULO(e_1, e_2) \\
-        & |\ \mathbf{U}(\sigma) \mid e_1 \star e_2 \mid \sum_{e_1} e_2.
-    \end{align*}
-\end{definition}
-The terms above are explained in order:
-\( \lambda x : T.e \) represents the abstraction for normal functions, and \( \mu x.e \) represents the abstraction for index functions.
-\( e_1\ e_2 \) denotes function application.
-\( (e_1, e_2) \) is the basis pair for product types.
-The constants \( 0 \), \( 1 \), \( \ADDS \), \( e_1 \times \cdots \times e_n \), and \( e^* \) are symbols for scalars.
-The next line includes constant symbols for kets, bras, and operators.
-\( \ket{e} \) is a ket, \( \bra{t} \) is a bra, and \( e^\dagger \) denotes the conjugate transpose of \( e \).
-\( e_1.e_2 \) represents scaling the term \( e_2 \) by scalar \( e_1 \).
-\( \mathbf{U}(\sigma) \) denotes the universal set with index \( \sigma \).
-\( e_1 \star e_2 \) represents the Cartesian product of \( e_1 \) and \( e_2 \).
-\( \sum_{e_1} e_2 \) is the big operator sum, modeled by folding the function \( e_2 \) over the value set \( e_1 \).
-Typically, the sum's body is given by an abstraction.
-For convenience, we also use the notation \( \sum_{x \in s} X \) to represent \( \sum_{s} \lambda x : T .
-X \).
-Additionally, \( \ADDS \) and \( \ADD \) are distinct AC symbols: \( \ADDS \) is used for scalar addition and \( \ADD \) for linear algebra addition.
-There are five kinds of linear algebraic multiplications among kets, bras, and operators.
-These are encoded using five different symbols: \( \DOT \), \( \MULK \), \( \MULB \), \( \OUTER \), and \( \MULO \).
-
-Compared to DiracDec, the symbols for addition, adjoint, scaling, and tensor products have been consolidated.
-The multiplications are still kept separate due to their differing properties.
-These operations are denoted as \( B \cdot K \), \( K_1 \cdot K_2 \), \( B_1 \cdot B_2 \), \( K \cdot B \), and \( O_1 \cdot O_2 \), respectively.
-
-\subsection{Typing System}
-
-The typing system is responsible for classifying terms within a proof system, using a context that specifies the types of variables and definitions.
-The context is divided into two components: the environment \( E \), which includes definitions and assumptions, and the context \( \Gamma \), which tracks bound variables.
-Both \( E \) and \( \Gamma \) consist of sequences of assumptions \( x : T \) or definitions \( x := t : T \).
-\begin{definition}[Environment and Context]
-    The syntax for \( E \) and \( \Gamma \) is:
-    \begin{align*}
-        E ::= &\ [] \mid E; x : \Index \mid E; x : T \mid E; x := t : T. \\
-        \Gamma ::= &\ [] \mid \Gamma; x : \Index \mid \Gamma; x : T.
-    \end{align*}
-\end{definition}
-Definitions refer to symbols that can be expanded or unfolded, and typically represent abstract concepts, such as transpose or trace in Dirac notation.
-Assumptions, on the other hand, define the types of variables, and variable assumptions are implicitly universally quantified in the case of equational proofs.
-
-Type checking in our language involves maintaining a well-formed environment and context \( E[\Gamma] \).
-We say an expression \( t \) has type \( X \) in context \( E[\Gamma] \) if the typing judgment \( E[\Gamma] \vdash t : X \) can be proven through the rules in~\Cref{sec: full typing rules}.
-The well-formedness of the context \( \WF(E)[\Gamma] \) is built incrementally:
-\[
-    \frac{}{\WF([])[]}
-    \qquad
-    \frac{\WF(E)[] \qquad x \notin E}{\WF(E; x : \Index)[]}
-    \qquad
-    \frac{E[]\vdash t:T \qquad x \notin E}{\WF(E; x := t:T)[]}.
-\]
-Starting with an empty context, new index symbols can be introduced, and symbols with checked types can be defined.
-Typing judgments are then made inductively using the information from \( E[\Gamma] \).
-For instance, the rule \( x : \Index \in E[\Gamma] \) holds if \( E \) or \( \Gamma \) contains an assumption for \( x \).
-And \(\KType(\sigma)\) is a valid type for kets, if the argument $\sigma$ is typed as an index.
-\begin{gather*}
-    \frac{\WF(E)[\Gamma] \qquad x : \Index \in E[\Gamma]}{E[\Gamma] \vdash x : \Index}
-    \qquad
-    \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \KType(\sigma) : \Type}
-\end{gather*}
-
-Terms are then typed accordingly.
-For example, the ket \( \ket{t} \) will have the type \( \KType(\sigma) \) if \( t \) is a basis term of index \( \sigma \).
-Similarly, the inner product between a bra and a ket of the same index \( \sigma \) is typed as a scalar.
-It corresponds to the constraint of inner product that vectors should be from the same Hilbert space.
-\begin{gather*}
-    \frac{E[\Gamma]\vdash t : \Basis(\sigma)}{E[\Gamma] \vdash \ket{t} : \KType(\sigma)}
-    \qquad
-    \frac{E[\Gamma]\vdash B : \BType(\sigma) \qquad E[\Gamma]\vdash K : \KType(\sigma)}{E[\Gamma] \vdash B \cdot K : \SType}
-\end{gather*}
-
-The typing for functions and applications follows the standard practice.
-For example, the index function \( \mu x.
-t \) is typed with \( x \) as a bound variable of type \( \Index \), and the application \( (t\ u) \) has the type \( U\{x/u\} \), obtained by replacing \( x \) with the index instance \( u \).
-\begin{gather*}
-    \frac{E[\Gamma; x : T] \vdash t : U}{E[\Gamma] \vdash (\lambda x : T . t) : T \to U}
-    \qquad
-    \frac{E[\Gamma] \vdash t:U \to T \qquad E[\Gamma] \vdash u:U}{E[\Gamma] \vdash (t\ u):T} 
-    \\[1em]
-    \frac{E[\Gamma; x : \Index] \vdash t : U}{E[\Gamma] \vdash (\mu x.t) : \forall x.U}
-    \qquad
-    \frac{E[\Gamma] \vdash t:\forall x.U \qquad E[\Gamma] \vdash u : \Index}{E[\Gamma] \vdash (t\ u):U\{x/u\}}
-\end{gather*}
-
-Finally, the big operator sum is modeled by folding a function over a set, with the typing rule as follows:
-\[
-    \frac{E[\Gamma] \vdash s : \SET(\sigma) \qquad E[\Gamma] \vdash f : \Basis(\sigma) \to \KType(\tau)}{E[\Gamma] \vdash \sum_{s} f : \KType(\tau)}.
-\]
-
-\begin{lemma}
-    The typing of expressions is both decidable and unique.
-\end{lemma}
-
-\begin{proof}
-    The type of an expression can be determined recursively. For any given function symbol and argument types, there is at most one typing rule, ensuring the uniqueness of typing.
-\end{proof}
-
-\subsection{Semantics}
-
-The semantics of a language define the meaning of its expressions.
-In this context, the objective of our algorithm is to determine whether two expressions are semantically equivalent.
-We define the semantics in a denotational manner, mapping syntax to set-theoretic objects.
-
-\subsubsection{Denotational Semantics}
-The denotational semantics interpret every expression as an object in linear algebra, according to a valuation mapping \( v \), which assigns values to the variables in the expression.
-The semantics of an expression \( e \) with a given valuation \( v \) is denoted as \( \sem{e}_v \).
-Two expressions \( e_1 \) and \( e_2 \) are considered equivalent if their semantics are equal for all valuations, i.e., \( \sem{e_1}_v = \sem{e_2}_v \) for all \( v \).
-
-The complete interpretation of terms and types is provided in~\Cref{sec: full denotational sem}.
-Variables typed with \( \Index \) are interpreted as finite sets, and the product of two indices \( \sem{\sigma_1 \times \sigma_2} \) is defined as the Cartesian product of the sets \( \sem{\sigma_1} \) and \( \sem{\sigma_2} \).
-More generally, each type is interpreted as a set.
-For example, the scalar type \( \sem{\SType} \) is interpreted as the set of complex numbers \( \mathbb{C} \), and the ket and bra types \( \sem{\KType(\sigma)} \) and \( \sem{\BType(\sigma)} \) are interpreted as the Hilbert space \( \mathcal{H}_{\sem{\sigma}} \) and its dual \( \mathcal{H}_{\sem{\sigma}}^* \), respectively.
-Terms are explained as the set elements.
-For example, the semantics of ket tensor product $\sem{K_1 \otimes K_2} \equiv \sem{K_1} \otimes \sem{K_2}$, is obtained by first calculating the semantics $\sem{K_1}$ and $\sem{K_2}$ as vectors, and then take the vector tensor product as result.
-
-% One special case is the delta function \( \delta_{s,t} \), which is interpreted as:
-% \[
-%     \sem{\delta_{s,t}} =
-%     \begin{cases}
-%         1, & \text{if } \sem{s} = \sem{t}, \\
-%         0, & \text{if } \sem{s} \neq \sem{t}.
-%     \end{cases}
-% \]d
-
-The idea behind the interpretation of types and terms is to formalize the typing relation using set-theoretic inclusion.
-Specifically, for a well-formed context \( E[\Gamma] \), term \( t \), and type \( T \), if \( E[\Gamma] \vdash t : T \), then for any valuation \( v \), the semantics of \( t \) must lie within the semantics of \( T \).
-
-\begin{lemma}
-    For any well-formed context \( E[\Gamma] \), term \( t \), and type \( T \), if \( E[\Gamma] \vdash t : T \), then for all valuations \( v \), \( \sem{t}_v \in \sem{T}_v \).
-\end{lemma}
-
-\begin{proof}
-    The proof follows directly by checking each case.
-\end{proof}
-
-This interpretation formalizes the standard understanding of Dirac notation and provides the foundation for the algorithm.
-However, computers cannot directly reason about equivalence through mathematical interpretations.
-Therefore, we proceed by defining a proof system that abstracts these concepts.
-
-\subsubsection{Axiomatic semantics}
-
-The proof system for equivalence is based on equational logic, together with axioms that describe the properties of Dirac notation.
-A full list of these axioms can be found in~\Cref{sec: full axioms}.
-The axioms cover fundamental aspects of linear spaces, as well as other structures like the tensor and inner products.
-For example, we have the absorption law for zero symbols:
-\(X \cdot \mathbf{0} = \mathbf{0},\)
-and the bilinearity of the dot product:
-\[
-(a.X) \cdot Y = a \cdot (X \cdot Y), \quad X \cdot (Y_1 + Y_2) = X \cdot Y_1 + X \cdot Y_2, 
-\]
-\[
-X \cdot (a.Y) = a \cdot (X \cdot Y), \quad (X_1 + X_2) \cdot Y = X_1 \cdot Y + X_2 \cdot Y.
-\]
-
-As mentioned in the introduction, a subset \( E \) of these axioms cannot be decided by term rewriting alone. The axioms in \( E \) include:
-\[
-\begin{aligned}
-    \text{AC-equivalence} &\quad \text{e.g.,} \quad X + Y = Y + X, \quad (X + Y) + Z = X + (Y + Z), \\
-    \alpha\text{-equivalence} &\quad \lambda x . A = \lambda y . A\{x/y\}, \\
-    \text{SUM-SWAP} &\quad \sum_{i \in s_1} \sum_{j \in s_2} A = \sum_{j \in s_2} \sum_{i \in s_1} A, \\
-    \text{scalar theories} &\quad \text{e.g.,} \quad a + 0 = a, \quad a \times (b + c) = a \times b + a \times c.
-\end{aligned}
-\]
-The scalar theories are treated separately as a module and are not considered in this work. In the implementation, we use the Mathematica kernel to decide scalar equivalences.
-These equational axioms provide an operable theory for the proof automation algorithm.
-Denotational semantics can be seen as one model for this theory, meaning that equivalences derived from the axioms always imply equivalence in the interpretations.
-\begin{lemma}
-    For all well-formed contexts \( E[\Gamma] \) and terms \( e_1, e_2 \), if \( \vdash e_1 = e_2 \), then \( \sem{e_1} = \sem{e_2} \).
-\end{lemma}
-\begin{proof}
-    This follows directly from checking all cases.
-\end{proof}
-
-Thus, the axioms are sound with respect to denotational semantics.
-However, the reverse does not hold: there exist equivalences in denotational semantics that cannot be captured by these axioms.
-Nevertheless, our work focuses on solving practical examples, which are fully covered by the axioms in the experiments.
-
-To conclude this section, we demonstrate the formalization of the symbols used in the motivating example.
-\begin{example}[Formalizing the Motivating Example]
-    \label{ex: formalizing motivating}
-    Definitions and assumptions in the environment \( E \) are formalized as follows:
-    \begin{align*}
-        & \text{TPO} && := \mu T_1. \mu T_2. \lambda O : \OType(T_1, T_2). \sum_{i \in \mathbf{U}(T_1)} \sum_{j \in \mathbf{U}(T_2)} \bra{i} O \ket{j} . \ket{j}\bra{i} \\
-        & &&\quad : \forall T_1. \forall T_2. \OType(T_1, T_2) \to \OType(T_2, T_1); \\
-        &\text{phi} &&:= \mu T. \sum_{i \in \mathbf{U}(T)} \sum_{j \in \mathbf{U}(T)} \ket{(i, j)} : \forall T.\KType(T \times T); \\
-        & T && : \Index; \\
-        & M && : \OType(T, T).
-    \end{align*}
-    The symbol \( \text{TPO} \) represents the transpose of an operator, polymorphic on the Hilbert spaces \( T_1 \) and \( T_2 \).
-The symbol \( \text{phi} \) takes the index \( T \) and defines the maximally entangled states, summing over all basis elements in \( T \), as indicated by the universal set \( \mathbf{U}(T) \).
-    With the assumption of the index \( T \) and operator \( M \), we can express the equivalence in the non-labelled version as:
-    \[
-    (\textrm{M} \otimes \mathbf{1}_\mathcal{O}(\textrm{T})) \cdot (\textrm{phi T}) = (\mathbf{1}_\mathcal{O}(\textrm{T}) \otimes (\textrm{TPO T T M})) \cdot (\textrm{phi T}).
-    \]
-\end{example}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Algorithm for Deciding Dirac Notation Equations}
-\label{sec: decide}
-
-We utilize the term rewriting technique to decide the equivalence of Dirac notation, based on its axiomatic semantics.
-The process involves normalizing two terms and checking whether they have identical syntax.
-This normalization is achieved through matching and substitution in universal algebra, using a set of predefined rewriting rules.
-
-As mentioned in the semantics section, the axioms in \( E \) cannot be fully decided through term rewriting.
-DiracDec decides \( E \) by examining all possible permutations in the rewriting result.
-In this work, we introduce a more efficient approach by incorporating sorting algorithms into the normalization procedure.
-
-The normalization procedure consists of the following steps.
-\begin{enumerate}
-    \item \textbf{First Rewritings}: Expand definitions and simplify expressions.
-    \item \textbf{Variable Expansion}: Consider scalar expressions to ensure completeness.
-    \item \textbf{Second Rewritings}: Normalize terms modulo \( E \).
-    \item \textbf{Sorting Without Bound Variables}: Normalize AC-equivalence.
-    \item \textbf{Swap Successive Summations}: Normalize SUM-SWAP equivalences.
-    \item \textbf{De Bruijn Normalization}: Normalize \( \alpha \)-equivalence.
-\end{enumerate}
-
-\subsection{Normalization Modulo \( E \) by Term Rewriting}
-
-Term rewriting rules, represented as \( l \ \reduce\ r \), are applied recursively to normalize terms.
-In each step, subterms matching the left-hand side \( l \) of a rule are replaced with the corresponding right-hand side \( r \).
-The procedure terminates when no further rewritings can be made.
-A comprehensive list of rewriting rules can be found in~\Cref{sec: rewriting rules}. Below are some key examples to illustrate the design:
-
-One of our optimizations is using functions with indefinite arities. 
-Therefore, we use a flattening rule to handle associativity with AC symbols:
-\[
-a_1 + \cdots + (b_1 + \cdots + b_m) + \cdots + a_n \ \reduce\ a_1 + \cdots + b_1 + \cdots + b_m + \cdots + a_n.
-\]
-Commutativity is handled later in the sorting step. Many of the rewriting rules are directly derived from the equational axioms, such as:
-\begin{align*}
-    & \textrm{(R-DOT6)} && \bra{s} \cdot \ket{t} \ \reduce\  \delta_{s, t}, \\
-    & \textrm{(R-DELTA0)} && \delta_{s, s} \ \reduce\  1, \\
-    & \textrm{(R-MULK1)} && O : \OType(\sigma, \tau) \Rightarrow O \cdot \mathbf{0}_{\KType(\tau)} \ \reduce\  \mathbf{0}_{\KType(\sigma)}, \\
-    & \textrm{(R-MULK11)} && (O_1 \otimes O_2) \cdot (K_1 \otimes K_2) \ \reduce\  (O_1 \cdot K_1) \otimes (O_2 \cdot K_2).
-\end{align*}
-Some of these directions are obvious.
-For example, (R-DOT6) states that the inner product of two basis vectors is reduced to a delta expression, and (R-DELTA0) transforms a delta function on identical basis to a scalar \( 1 \).
-The rule (R-MULK1) reflects the axiom that multiplying a zero vector results in zero.
-This rule is conditional on typing, which is checked during rewriting.
-Some rules, like (R-MULK11), require a deeper understanding, such as the preference for tensor products over multiplication.
-
-As a reference, the term rewriting system in DiracDec has been proven complete for all axioms, except for the sum symbol.
-The completeness result is derived from checking the confluence and termination of the system.
-Our rewriting rules are translations from DiracDec into the typed and abstract language, ensuring that the corresponding symbols in our system are also complete.
-
-We also have additional rules that handle summations. For example:
-\begin{align*}
-    & \quad i \text{ free in } t \Rightarrow \sum_{i \in \mathbf{U}(\sigma)} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (\delta_{i, t}.A) \ \reduce\  \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} A\{i/t\}, \\
-    & \quad \left( \sum_{i \in M} B \right) \cdot K \ \reduce\  \sum_{i \in M} (B \cdot K).
-\end{align*}
-The first rule eliminates delta expressions in summations, while the second rule pushes summations outside of inner products.
-While there is no guarantee of completeness for these rules, they work effectively in practice.
-
-\subsubsection{Variable Expansion}
-One important technique, revealed in the DiracDec work, is the expansion of variables, which is critical for proofs involving summation. For example:
-\[
-\frac{E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash K \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)}(\bra{i} \cdot K).\ket{i}} \quad \quad
-\frac{E[\Gamma] \vdash B : \BType(\sigma)}{E[\Gamma] \vdash B \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)}(B \cdot \ket{i}).\bra{i}},
-\]
-\[
-\frac{E[\Gamma] \vdash O : \OType(\sigma, \tau)}{E[\Gamma] \vdash O \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)} \sum_{j \in \mathbf{U}(\tau)}(\bra{i} \cdot O \cdot \ket{j}).(\ket{i} \cdot \bra{j})}.
-\]
-These rules transform variables into their symbolic summations based on their decomposition over the basis. 
-The rules are not terminating, therefore is applied recrusively once in the second step called \textit{variable expansion}.
-Nevertheless, we have found that applying the expansion only once for all variables is sufficient for normalization.
-\begin{lemma}
-    Let \( \textrm{expand}(e) \) denote the result of expanding all variables in \( e \) once.
-For all well-typed terms \( e \) in \( E[\Gamma] \), \( \textrm{expand}(\textrm{expand}(e)) \) and \( \textrm{expand}(e) \) have the same normal form.
-\end{lemma}
-\begin{proof}
-    Expanding a ket variable twice, for example, results in the following transformation:
-    \[
-    \sum_{i \in \mathbf{U}(\sigma)} (\bra{i} \cdot \sum_{j \in \mathbf{U}(\sigma)} (\bra{j} \cdot K) \cdot \ket{j}) \cdot \ket{i} \ \reduce\  \sum_{i \in \mathbf{U}(\sigma)} \sum_{j \in \mathbf{U}(\sigma)} (\bra{j} \cdot K \cdot \braket{i|j}) \cdot \ket{i},
-    \]
-    where the delta symbol elimination rule returns the term to its original form. The same holds for bra and operator terms.
-\end{proof}
-
-\subsection{Deciding Equational Theory \( E \)}
-
-In the Mathematica implementation of DiracDec, the equational theory \( E \) is decided using unification, which attempts to find a substitution for the summation bound variables that makes two expressions syntactically equivalent.
-This unification process iterates through all permutations of AC symbol arguments, and its complexity is factorial in the maximum number of AC symbol arguments.
-
-As the first improvement, we check \( \alpha \)-equivalence using the de Bruijn index~\cite{deBruijn1972lambda}, which replaces references to bound variable names with the distance from the lambda abstraction to the variable.
-For instance, the nominal lambda abstraction \( \lambda x.
-x \) is transformed into \( \lambda .
-0 \), while \( \lambda x.
-\lambda y.
-(x\ (y\ x)) \) is transformed into \( \lambda.\lambda.
-(1\ (0\ 1)) \).
-This transformation into the de Bruijn index is performed as the final step in the normalization process to check equivalence between terms with different bound variable names.
-
-The remaining axioms, such as AC-equivalence and SUM-SWAP, assert equivalence under permutations.
-A standard approach for deciding such equivalences is to normalize terms by sorting in a predefined order.
-For example, given the dictionary order \( a < b < c \), the term \( b + c + a \) (and any other AC-equivalent term) is normalized into \( a + b + c \).
-However, in our setting, two intertwined difficulties arise: how to assign an order to all terms in the language, and how to simultaneously sort for both axioms.
-
-Consider the following two equivalent terms:
-\[
-\sum_{i \in s_1} \sum_{j \in s_2} \bra{i} A \ket{j} \times \bra{j} B \ket{i}
-= 
-\sum_{i \in s_2} \sum_{j \in s_1} \bra{i} B \ket{j} \times \bra{j} A \ket{i}
-\]
-While these two terms are equivalent, directly sorting the elements of scalar multiplication using lexical order does not yield the same form.
-
-To address this issue, we propose an algorithm that sorts terms without considering bound variables.
-The key observation is that in a successive sum expression \( \sum_{i \in s_1} \cdots \sum_{j \in s_n} A \), the names and order of the bound variables \( i, \dots, j \) can be freely permuted.
-Therefore, all bound variables should be treated uniformly during sorting, and the order of summation can then be determined based on the position of the bound variables.
-
-In the example above, we first ignore the bound variables and sort the sum body into \( \bra{\bullet} A \ket{\bullet} \times \bra{\bullet} B \ket{\bullet} \).
-Then, we swap the summations such that the bound variable at the first \( \bullet \) position appears at the outermost position.
-The results will have the same de Bruijn normal form, namely \( \sum_{s_1} \sum_{s_2} \bra{\$1} A \ket{\$0} \times \bra{\$0} B \ket{\$1} \).
-
-To describe the algorithm in the following, we introduce two key notations.
-For a term \( e = f(a_1, a_2, \dots, a_n) \), \( \textrm{head}(e) \) denotes the function symbol \( f \), while \( \textrm{arg}(e, i) \) refers to the \( i \)-th argument \( a_i \) of the term.
-In this context, variables and constants are treated as functions with zero arguments.
-
-\begin{definition}[Order Without Bound Variables]
-Let \( \mathcal{B} \) represent the set of bound variables, with the assumption that all bound variables are unique.
-We also assume that a total order exists over all symbols.
-The relation \( e_1 =_\mathcal{B} e_2 \) holds if:
-\begin{itemize}
-    \item \( \textrm{head}(e_1) = \textrm{head}(e_2) \), and for all \( i \), \( \textrm{arg}(e_1, i) =_\mathcal{B} \textrm{arg}(e_2, i) \), or
-    \item \( e_1 \in \mathcal{B} \) and \(e_2 \in \mathcal{B}\).
-\end{itemize}
-
-The relation \( e_1 <_\mathcal{B} e_2 \) holds between two terms if:
-\begin{itemize}
-    \item $e_1 \notin \mathcal{B}$ and $e_2 \in \mathcal{B}$, or
-    \item $head(e_1) < head(e_2)$, or
-    \item $head(e_1) = head(e_2)$, and there exists $n$ with $arg(e_1, n) <_\mathcal{B} arg(e_2, n)$, where $arg(e_1, i) =_\mathcal{B} arg(e_2, i)$ for all $i < n$.
-\end{itemize}
-\end{definition}
-It can be shown that \( e_1 =_\mathcal{B} e_2 \) if and only if neither \( e_1 <_\mathcal{B} e_2 \) nor \( e_2 <_\mathcal{B} e_1 \) holds.
-The purpose of this ordering is to compare function symbols in a top-down manner while ignoring bound variables.
-The sorted order enables normalization of terms in terms of AC equivalence.
-\begin{definition}[Sort Transformation]
-    For a term $e$ with bound variable set $\mathcal{B}$,
-    The sort transformation sort(e) is defined as
-    \begin{itemize}
-        \item $e$, if $e$ is a variable or constant;
-        \item $\lambda x:T. \textrm{sort}(e')$, if $e \equiv \lambda x : T. e'$;
-        \item $\mu x. \textrm{sort}(e')$, if $e \equiv \mu x. \textrm{sort}(e')$; or
-        \item $f(\textrm{sort}(a_1), \cdots, \textrm{sort}(a_n))$, if $e \equiv f(a_1, \cdots, a_n)$.
-If $f$ is an AC symbol, then the order of sort($a_i$) is sorted ascendingly according to $<_\mathcal{B}$.
-    \end{itemize}
-\end{definition}
-
-After sorting, the next step is the \textit{swap transformation}, which arranges successive summations based on the order of bound variables.
-\begin{definition}[Swap Transformation]
-For a term \( e \) with a sorting result \( \textrm{sort}(e) \), the swap transformation proceeds by ordering all bound variables according to their first appearances, except in function definitions \( \lambda x \) and \( \mu x \).
-The swap transformation then reorders the successive summations accordingly.
-\end{definition}
-
-The algorithm that applies the sort and swap transformations, followed by de Bruijn normalization, successfully handles all equivalences in our benchmark.
-We formalize the completeness of this transformation in the following conjecture:
-
-\begin{conjecture}[Completeness of Transformation]
-    For any two terms \( e_1 \) and \( e_2 \) that are equivalent under AC equivalence, SUM-SWAP, and \( \alpha \)-equivalence, they will have the same form after sort, swap, and de Bruijn transformations.
-\end{conjecture}
-
-% The order depends on there occurances in the last sorting result. If no occurance, then the order will depend on the set (for sum) and the type (for lambda abstraction only).
-
-%%%%%%%%%%%%%%%%%%%%%%%
-
-% The idea is to assign an order to terms, which is independent on the bound variables. Because we can have terms with nested AC symbols.
-
-Lastly, we can prove that the equivalence established by this normalization procedure is sound with respect to the semantics.
-
-\begin{theorem}[Soundness]
-    For any well-formed context \( E[\Gamma] \) and well-typed expressions \( e_1 \) and \( e_2 \), if  $e_1$ and $e_2$ have the same normal form, then \( \sem{e_1} = \sem{e_2} \).
-\end{theorem}
-
-\begin{proof}
-    The soundness of the term rewriting procedure follows from the fact that each rewriting rule preserves equivalence.
-Furthermore, the operations in the sort and swap transformations respect the AC-equivalence and SUM-SWAP axioms.
-Finally, the de Bruijn normalization ensures soundness for \( \alpha \)-equivalence.
-\end{proof}
-
 % Labelled Dirac Notation

### 2025-01-27 04efficient_alg.tex
rearranged

--- 2025-01-27 02techniques.tex
@@
-\section{Language, Typing and Semantics}
-The first step is to formalize the language for Dirac notation.
-In DiracDec, the language has a concrete design, where the same syntax for different types corresponds to distinct symbols.
-Our goal is to transition from this concrete design to a more abstract one, which aligns more closely with the conventional Dirac notation we use and simplifies the term rewriting system.
-
-To achieve this, we organize our language into three layers: the index, the type, and the term.
-Terms represent concrete instances such as kets, bras, and operators, which will be typed and checked.
-The index represents classical data types and appears in type expressions to differentiate between various Hilbert spaces and sets.
-
-\begin{definition}[Index Syntax]
-    The syntax for type indices is:
-    \begin{align*}
-        \sigma ::=\ & x \mid \sigma_1 \times \sigma_2.
-    \end{align*}
-\end{definition}
-Here, \( x \) is a variable, and \( \sigma_1 \times \sigma_2 \) represents the product type for tensor product spaces or Cartesian product sets.
-
-\begin{definition}[Type Syntax]
-    The syntax for Dirac notation types is:
-    \begin{align*}
-        T ::=\ & \Basis(\sigma) \mid \SType \mid \KType(\sigma) \mid \BType(\sigma) \mid \OType(\sigma_1, \sigma_2) \mid T_1 \to T_2 \mid \forall x.T \mid \SET(\sigma).
-    \end{align*}
-\end{definition}
-\( \Basis(\sigma) \) denotes the type for basis elements in the index \( \sigma \).
-\( \SType \) represents scalars, while \( \KType(\sigma) \) and \( \BType(\sigma) \) refer to ket and bra types in the Hilbert space \( \sigma \), respectively.
-\( \OType(\sigma_1, \sigma_2) \) represents linear operators with \( \sigma_2 \) as the domain and \( \sigma_1 \) as the codomain.
-\( \SET(\sigma) \) refers to the type of subsets of \( \sigma \), used to denote the values of bound variables in summations.
-The remaining two constructs define function types: \( T_1 \to T_2 \) represents normal functions that take a \( T_1 \)-type argument and return a \( T_2 \)-type term, while \( \forall x.
-T \) represents index functions that take an index argument \( x \) and produce a \( T \)-type term, which may depend on \( x \).
-Index functions are essential for defining polymorphic transformations across Hilbert spaces.
-
-\begin{definition}[Term Syntax]
-    The syntax for Dirac notation terms is:
-    \begin{align*}
-        e ::=\ & x \mid \lambda x : T.e \mid \mu x.e \mid e_1\ e_2 \mid (e_1, e_2) \\
-        & |\ 0 \mid 1 \mid \ADDS(e_1, \cdots, e_n) \mid e_1 \times \cdots \times e_n \mid e^* \mid \delta_{e_1, e_2} \mid \DOT(e_1, e_2) \\
-        & |\ \ZEROK(\sigma) \mid \ZEROB(\sigma) \mid \ZEROO(\sigma_1, \sigma_2) \mid \ONEO(\sigma) \\
-        & |\ \ket{e} \mid \bra{t} \mid e^\dagger \mid e_1.e_2 \mid \ADD(e_1, \cdots, e_n) \mid e_1 \otimes e_2 \\
-        & |\ \MULK(e_1, e_2) \mid \MULB(e_1, e_2) \mid \OUTER(e_1, e_2) \mid \MULO(e_1, e_2) \\
-        & |\ \mathbf{U}(\sigma) \mid e_1 \star e_2 \mid \sum_{e_1} e_2.
-    \end{align*}
-\end{definition}
-The terms above are explained in order:
-\( \lambda x : T.e \) represents the abstraction for normal functions, and \( \mu x.e \) represents the abstraction for index functions.
-\( e_1\ e_2 \) denotes function application.
-\( (e_1, e_2) \) is the basis pair for product types.
-The constants \( 0 \), \( 1 \), \( \ADDS \), \( e_1 \times \cdots \times e_n \), and \( e^* \) are symbols for scalars.
-The next line includes constant symbols for kets, bras, and operators.
-\( \ket{e} \) is a ket, \( \bra{t} \) is a bra, and \( e^\dagger \) denotes the conjugate transpose of \( e \).
-\( e_1.e_2 \) represents scaling the term \( e_2 \) by scalar \( e_1 \).
-\( \mathbf{U}(\sigma) \) denotes the universal set with index \( \sigma \).
-\( e_1 \star e_2 \) represents the Cartesian product of \( e_1 \) and \( e_2 \).
-\( \sum_{e_1} e_2 \) is the big operator sum, modeled by folding the function \( e_2 \) over the value set \( e_1 \).
-Typically, the sum's body is given by an abstraction.
-For convenience, we also use the notation \( \sum_{x \in s} X \) to represent \( \sum_{s} \lambda x : T .
-X \).
-Additionally, \( \ADDS \) and \( \ADD \) are distinct AC symbols: \( \ADDS \) is used for scalar addition and \( \ADD \) for linear algebra addition.
-There are five kinds of linear algebraic multiplications among kets, bras, and operators.
-These are encoded using five different symbols: \( \DOT \), \( \MULK \), \( \MULB \), \( \OUTER \), and \( \MULO \).
-
-Compared to DiracDec, the symbols for addition, adjoint, scaling, and tensor products have been consolidated.
-The multiplications are still kept separate due to their differing properties.
-These operations are denoted as \( B \cdot K \), \( K_1 \cdot K_2 \), \( B_1 \cdot B_2 \), \( K \cdot B \), and \( O_1 \cdot O_2 \), respectively.
-
-\subsection{Typing System}
-
-The typing system is responsible for classifying terms within a proof system, using a context that specifies the types of variables and definitions.
-The context is divided into two components: the environment \( E \), which includes definitions and assumptions, and the context \( \Gamma \), which tracks bound variables.
-Both \( E \) and \( \Gamma \) consist of sequences of assumptions \( x : T \) or definitions \( x := t : T \).
-\begin{definition}[Environment and Context]
-    The syntax for \( E \) and \( \Gamma \) is:
-    \begin{align*}
-        E ::= &\ [] \mid E; x : \Index \mid E; x : T \mid E; x := t : T. \\
-        \Gamma ::= &\ [] \mid \Gamma; x : \Index \mid \Gamma; x : T.
-    \end{align*}
-\end{definition}
-Definitions refer to symbols that can be expanded or unfolded, and typically represent abstract concepts, such as transpose or trace in Dirac notation.
-Assumptions, on the other hand, define the types of variables, and variable assumptions are implicitly universally quantified in the case of equational proofs.
-
-Type checking in our language involves maintaining a well-formed environment and context \( E[\Gamma] \).
-We say an expression \( t \) has type \( X \) in context \( E[\Gamma] \) if the typing judgment \( E[\Gamma] \vdash t : X \) can be proven through the rules in~\Cref{sec: full typing rules}.
-The well-formedness of the context \( \WF(E)[\Gamma] \) is built incrementally:
-\[
-    \frac{}{\WF([])[]}
-    \qquad
-    \frac{\WF(E)[] \qquad x \notin E}{\WF(E; x : \Index)[]}
-    \qquad
-    \frac{E[]\vdash t:T \qquad x \notin E}{\WF(E; x := t:T)[]}.
-\]
-Starting with an empty context, new index symbols can be introduced, and symbols with checked types can be defined.
-Typing judgments are then made inductively using the information from \( E[\Gamma] \).
-For instance, the rule \( x : \Index \in E[\Gamma] \) holds if \( E \) or \( \Gamma \) contains an assumption for \( x \).
-And \(\KType(\sigma)\) is a valid type for kets, if the argument $\sigma$ is typed as an index.
-\begin{gather*}
-    \frac{\WF(E)[\Gamma] \qquad x : \Index \in E[\Gamma]}{E[\Gamma] \vdash x : \Index}
-    \qquad
-    \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \KType(\sigma) : \Type}
-\end{gather*}
-
-Terms are then typed accordingly.
-For example, the ket \( \ket{t} \) will have the type \( \KType(\sigma) \) if \( t \) is a basis term of index \( \sigma \).
-Similarly, the inner product between a bra and a ket of the same index \( \sigma \) is typed as a scalar.
-It corresponds to the constraint of inner product that vectors should be from the same Hilbert space.
-\begin{gather*}
-    \frac{E[\Gamma]\vdash t : \Basis(\sigma)}{E[\Gamma] \vdash \ket{t} : \KType(\sigma)}
-    \qquad
-    \frac{E[\Gamma]\vdash B : \BType(\sigma) \qquad E[\Gamma]\vdash K : \KType(\sigma)}{E[\Gamma] \vdash B \cdot K : \SType}
-\end{gather*}
-
-The typing for functions and applications follows the standard practice.
-For example, the index function \( \mu x.
-t \) is typed with \( x \) as a bound variable of type \( \Index \), and the application \( (t\ u) \) has the type \( U\{x/u\} \), obtained by replacing \( x \) with the index instance \( u \).
-\begin{gather*}
-    \frac{E[\Gamma; x : T] \vdash t : U}{E[\Gamma] \vdash (\lambda x : T . t) : T \to U}
-    \qquad
-    \frac{E[\Gamma] \vdash t:U \to T \qquad E[\Gamma] \vdash u:U}{E[\Gamma] \vdash (t\ u):T} 
-    \\[1em]
-    \frac{E[\Gamma; x : \Index] \vdash t : U}{E[\Gamma] \vdash (\mu x.t) : \forall x.U}
-    \qquad
-    \frac{E[\Gamma] \vdash t:\forall x.U \qquad E[\Gamma] \vdash u : \Index}{E[\Gamma] \vdash (t\ u):U\{x/u\}}
-\end{gather*}
-
-Finally, the big operator sum is modeled by folding a function over a set, with the typing rule as follows:
-\[
-    \frac{E[\Gamma] \vdash s : \SET(\sigma) \qquad E[\Gamma] \vdash f : \Basis(\sigma) \to \KType(\tau)}{E[\Gamma] \vdash \sum_{s} f : \KType(\tau)}.
-\]
-
-\begin{lemma}
-    The typing of expressions is both decidable and unique.
-\end{lemma}
-
-\begin{proof}
-    The type of an expression can be determined recursively. For any given function symbol and argument types, there is at most one typing rule, ensuring the uniqueness of typing.
-\end{proof}
-
-\subsection{Semantics}
-
-The semantics of a language define the meaning of its expressions.
-In this context, the objective of our algorithm is to determine whether two expressions are semantically equivalent.
-We define the semantics in a denotational manner, mapping syntax to set-theoretic objects.
-
-\subsubsection{Denotational Semantics}
-The denotational semantics interpret every expression as an object in linear algebra, according to a valuation mapping \( v \), which assigns values to the variables in the expression.
-The semantics of an expression \( e \) with a given valuation \( v \) is denoted as \( \sem{e}_v \).
-Two expressions \( e_1 \) and \( e_2 \) are considered equivalent if their semantics are equal for all valuations, i.e., \( \sem{e_1}_v = \sem{e_2}_v \) for all \( v \).
-
-The complete interpretation of terms and types is provided in~\Cref{sec: full denotational sem}.
-Variables typed with \( \Index \) are interpreted as finite sets, and the product of two indices \( \sem{\sigma_1 \times \sigma_2} \) is defined as the Cartesian product of the sets \( \sem{\sigma_1} \) and \( \sem{\sigma_2} \).
-More generally, each type is interpreted as a set.
-For example, the scalar type \( \sem{\SType} \) is interpreted as the set of complex numbers \( \mathbb{C} \), and the ket and bra types \( \sem{\KType(\sigma)} \) and \( \sem{\BType(\sigma)} \) are interpreted as the Hilbert space \( \mathcal{H}_{\sem{\sigma}} \) and its dual \( \mathcal{H}_{\sem{\sigma}}^* \), respectively.
-Terms are explained as the set elements.
-For example, the semantics of ket tensor product $\sem{K_1 \otimes K_2} \equiv \sem{K_1} \otimes \sem{K_2}$, is obtained by first calculating the semantics $\sem{K_1}$ and $\sem{K_2}$ as vectors, and then take the vector tensor product as result.
-
-% One special case is the delta function \( \delta_{s,t} \), which is interpreted as:
-% \[
-%     \sem{\delta_{s,t}} =
-%     \begin{cases}
-%         1, & \text{if } \sem{s} = \sem{t}, \\
-%         0, & \text{if } \sem{s} \neq \sem{t}.
-%     \end{cases}
-% \]d
-
-The idea behind the interpretation of types and terms is to formalize the typing relation using set-theoretic inclusion.
-Specifically, for a well-formed context \( E[\Gamma] \), term \( t \), and type \( T \), if \( E[\Gamma] \vdash t : T \), then for any valuation \( v \), the semantics of \( t \) must lie within the semantics of \( T \).
-
-\begin{lemma}
-    For any well-formed context \( E[\Gamma] \), term \( t \), and type \( T \), if \( E[\Gamma] \vdash t : T \), then for all valuations \( v \), \( \sem{t}_v \in \sem{T}_v \).
-\end{lemma}
-
-\begin{proof}
-    The proof follows directly by checking each case.
-\end{proof}
-
-This interpretation formalizes the standard understanding of Dirac notation and provides the foundation for the algorithm.
-However, computers cannot directly reason about equivalence through mathematical interpretations.
-Therefore, we proceed by defining a proof system that abstracts these concepts.
-
-\subsubsection{Axiomatic semantics}
-
-The proof system for equivalence is based on equational logic, together with axioms that describe the properties of Dirac notation.
-A full list of these axioms can be found in~\Cref{sec: full axioms}.
-The axioms cover fundamental aspects of linear spaces, as well as other structures like the tensor and inner products.
-For example, we have the absorption law for zero symbols:
-\(X \cdot \mathbf{0} = \mathbf{0},\)
-and the bilinearity of the dot product:
-\[
-(a.X) \cdot Y = a \cdot (X \cdot Y), \quad X \cdot (Y_1 + Y_2) = X \cdot Y_1 + X \cdot Y_2, 
-\]
-\[
-X \cdot (a.Y) = a \cdot (X \cdot Y), \quad (X_1 + X_2) \cdot Y = X_1 \cdot Y + X_2 \cdot Y.
-\]
-
-As mentioned in the introduction, a subset \( E \) of these axioms cannot be decided by term rewriting alone. The axioms in \( E \) include:
-\[
-\begin{aligned}
-    \text{AC-equivalence} &\quad \text{e.g.,} \quad X + Y = Y + X, \quad (X + Y) + Z = X + (Y + Z), \\
-    \alpha\text{-equivalence} &\quad \lambda x . A = \lambda y . A\{x/y\}, \\
-    \text{SUM-SWAP} &\quad \sum_{i \in s_1} \sum_{j \in s_2} A = \sum_{j \in s_2} \sum_{i \in s_1} A, \\
-    \text{scalar theories} &\quad \text{e.g.,} \quad a + 0 = a, \quad a \times (b + c) = a \times b + a \times c.
-\end{aligned}
-\]
-The scalar theories are treated separately as a module and are not considered in this work. In the implementation, we use the Mathematica kernel to decide scalar equivalences.
-These equational axioms provide an operable theory for the proof automation algorithm.
-Denotational semantics can be seen as one model for this theory, meaning that equivalences derived from the axioms always imply equivalence in the interpretations.
-\begin{lemma}
-    For all well-formed contexts \( E[\Gamma] \) and terms \( e_1, e_2 \), if \( \vdash e_1 = e_2 \), then \( \sem{e_1} = \sem{e_2} \).
-\end{lemma}
-\begin{proof}
-    This follows directly from checking all cases.
-\end{proof}
-
-Thus, the axioms are sound with respect to denotational semantics.
-However, the reverse does not hold: there exist equivalences in denotational semantics that cannot be captured by these axioms.
-Nevertheless, our work focuses on solving practical examples, which are fully covered by the axioms in the experiments.
-
-To conclude this section, we demonstrate the formalization of the symbols used in the motivating example.
-\begin{example}[Formalizing the Motivating Example]
-    \label{ex: formalizing motivating}
-    Definitions and assumptions in the environment \( E \) are formalized as follows:
-    \begin{align*}
-        & \text{TPO} && := \mu T_1. \mu T_2. \lambda O : \OType(T_1, T_2). \sum_{i \in \mathbf{U}(T_1)} \sum_{j \in \mathbf{U}(T_2)} \bra{i} O \ket{j} . \ket{j}\bra{i} \\
-        & &&\quad : \forall T_1. \forall T_2. \OType(T_1, T_2) \to \OType(T_2, T_1); \\
-        &\text{phi} &&:= \mu T. \sum_{i \in \mathbf{U}(T)} \sum_{j \in \mathbf{U}(T)} \ket{(i, j)} : \forall T.\KType(T \times T); \\
-        & T && : \Index; \\
-        & M && : \OType(T, T).
-    \end{align*}
-    The symbol \( \text{TPO} \) represents the transpose of an operator, polymorphic on the Hilbert spaces \( T_1 \) and \( T_2 \).
-The symbol \( \text{phi} \) takes the index \( T \) and defines the maximally entangled states, summing over all basis elements in \( T \), as indicated by the universal set \( \mathbf{U}(T) \).
-    With the assumption of the index \( T \) and operator \( M \), we can express the equivalence in the non-labelled version as:
-    \[
-    (\textrm{M} \otimes \mathbf{1}_\mathcal{O}(\textrm{T})) \cdot (\textrm{phi T}) = (\mathbf{1}_\mathcal{O}(\textrm{T}) \otimes (\textrm{TPO T T M})) \cdot (\textrm{phi T}).
-    \]
-\end{example}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\section{Algorithm for Deciding Dirac Notation Equations}
+\section{Efficient Algorithm for Equivalence Checking}
 \label{sec: decide}
@@
 
-\subsection{Normalization Modulo \( E \) by Term Rewriting}
+% \subsection{Normalization Modulo \( E \) by Term Rewriting}
 
-Term rewriting rules, represented as \( l \ \reduce\ r \), are applied recursively to normalize terms.
+% Term rewriting rules, represented as \( l \ \reduce\ r \), are applied recursively to normalize terms.
 In each step, subterms matching the left-hand side \( l \) of a rule are replaced with the corresponding right-hand side \( r \).
 The procedure terminates when no further rewritings can be made.
-A comprehensive list of rewriting rules can be found in~\Cref{sec: rewriting rules}. Below are some key examples to illustrate the design:
+% A comprehensive list of rewriting rules can be found in~\Cref{sec: rewriting rules}. Below are some key examples to illustrate the design:
 
-One of our optimizations is using functions with indefinite arities. 
-Therefore, we use a flattening rule to handle associativity with AC symbols:
-\[
-a_1 + \cdots + (b_1 + \cdots + b_m) + \cdots + a_n \ \reduce\ a_1 + \cdots + b_1 + \cdots + b_m + \cdots + a_n.
-\]
-Commutativity is handled later in the sorting step. Many of the rewriting rules are directly derived from the equational axioms, such as:
-\begin{align*}
-    & \textrm{(R-DOT6)} && \bra{s} \cdot \ket{t} \ \reduce\  \delta_{s, t}, \\
-    & \textrm{(R-DELTA0)} && \delta_{s, s} \ \reduce\  1, \\
-    & \textrm{(R-MULK1)} && O : \OType(\sigma, \tau) \Rightarrow O \cdot \mathbf{0}_{\KType(\tau)} \ \reduce\  \mathbf{0}_{\KType(\sigma)}, \\
-    & \textrm{(R-MULK11)} && (O_1 \otimes O_2) \cdot (K_1 \otimes K_2) \ \reduce\  (O_1 \cdot K_1) \otimes (O_2 \cdot K_2).
-\end{align*}
-Some of these directions are obvious.
+% One of our optimizations is using functions with indefinite arities. 
+% Therefore, we use a flattening rule to handle associativity with AC symbols:
+% \[
+% a_1 + \cdots + (b_1 + \cdots + b_m) + \cdots + a_n \ \reduce\ a_1 + \cdots + b_1 + \cdots + b_m + \cdots + a_n.
+% \]
+% Commutativity is handled later in the sorting step. Many of the rewriting rules are directly derived from the equational axioms, such as:
+% \begin{align*}
+%     & \textrm{(R-DOT6)} && \bra{s} \cdot \ket{t} \ \reduce\  \delta_{s, t}, \\
+%     & \textrm{(R-DELTA0)} && \delta_{s, s} \ \reduce\  1, \\
+%     & \textrm{(R-MULK1)} && O : \OType(\sigma, \tau) \Rightarrow O \cdot \mathbf{0}_{\KType(\tau)} \ \reduce\  \mathbf{0}_{\KType(\sigma)}, \\
+%     & \textrm{(R-MULK11)} && (O_1 \otimes O_2) \cdot (K_1 \otimes K_2) \ \reduce\  (O_1 \cdot K_1) \otimes (O_2 \cdot K_2).
+% \end{align*}
+% Some of these directions are obvious.
 For example, (R-DOT6) states that the inner product of two basis vectors is reduced to a delta expression, and (R-DELTA0) transforms a delta function on identical basis to a scalar \( 1 \).
@@
 
-As a reference, the term rewriting system in DiracDec has been proven complete for all axioms, except for the sum symbol.
+% As a reference, the term rewriting system in DiracDec has been proven complete for all axioms, except for the sum symbol.
 The completeness result is derived from checking the confluence and termination of the system.
@@
 
-We also have additional rules that handle summations. For example:
-\begin{align*}
-    & \quad i \text{ free in } t \Rightarrow \sum_{i \in \mathbf{U}(\sigma)} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (\delta_{i, t}.A) \ \reduce\  \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} A\{i/t\}, \\
-    & \quad \left( \sum_{i \in M} B \right) \cdot K \ \reduce\  \sum_{i \in M} (B \cdot K).
-\end{align*}
-The first rule eliminates delta expressions in summations, while the second rule pushes summations outside of inner products.
+% We also have additional rules that handle summations. For example:
+% \begin{align*}
+%     & \quad i \text{ free in } t \Rightarrow \sum_{i \in \mathbf{U}(\sigma)} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (\delta_{i, t}.A) \ \reduce\  \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} A\{i/t\}, \\
+%     & \quad \left( \sum_{i \in M} B \right) \cdot K \ \reduce\  \sum_{i \in M} (B \cdot K).
+% \end{align*}
+% The first rule eliminates delta expressions in summations, while the second rule pushes summations outside of inner products.
 While there is no guarantee of completeness for these rules, they work effectively in practice.
 
-\subsubsection{Variable Expansion}
-One important technique, revealed in the DiracDec work, is the expansion of variables, which is critical for proofs involving summation. For example:
-\[
-\frac{E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash K \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)}(\bra{i} \cdot K).\ket{i}} \quad \quad
-\frac{E[\Gamma] \vdash B : \BType(\sigma)}{E[\Gamma] \vdash B \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)}(B \cdot \ket{i}).\bra{i}},
-\]
-\[
-\frac{E[\Gamma] \vdash O : \OType(\sigma, \tau)}{E[\Gamma] \vdash O \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)} \sum_{j \in \mathbf{U}(\tau)}(\bra{i} \cdot O \cdot \ket{j}).(\ket{i} \cdot \bra{j})}.
-\]
-These rules transform variables into their symbolic summations based on their decomposition over the basis. 
-The rules are not terminating, therefore is applied recrusively once in the second step called \textit{variable expansion}.
-Nevertheless, we have found that applying the expansion only once for all variables is sufficient for normalization.
-\begin{lemma}
-    Let \( \textrm{expand}(e) \) denote the result of expanding all variables in \( e \) once.
+% \subsubsection{Variable Expansion}
+% One important technique, revealed in the DiracDec work, is the expansion of variables, which is critical for proofs involving summation. For example:
+% \[
+% \frac{E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash K \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)}(\bra{i} \cdot K).\ket{i}} \quad \quad
+% \frac{E[\Gamma] \vdash B : \BType(\sigma)}{E[\Gamma] \vdash B \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)}(B \cdot \ket{i}).\bra{i}},
+% \]
+% \[
+% \frac{E[\Gamma] \vdash O : \OType(\sigma, \tau)}{E[\Gamma] \vdash O \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)} \sum_{j \in \mathbf{U}(\tau)}(\bra{i} \cdot O \cdot \ket{j}).(\ket{i} \cdot \bra{j})}.
+% \]
+% These rules transform variables into their symbolic summations based on their decomposition over the basis. 
+% The rules are not terminating, therefore is applied recrusively once in the second step called \textit{variable expansion}.
+% Nevertheless, we have found that applying the expansion only once for all variables is sufficient for normalization.
+% \begin{lemma}
+%     Let \( \textrm{expand}(e) \) denote the result of expanding all variables in \( e \) once.
 For all well-typed terms \( e \) in \( E[\Gamma] \), \( \textrm{expand}(\textrm{expand}(e)) \) and \( \textrm{expand}(e) \) have the same normal form.
-\end{lemma}
-\begin{proof}
-    Expanding a ket variable twice, for example, results in the following transformation:
-    \[
-    \sum_{i \in \mathbf{U}(\sigma)} (\bra{i} \cdot \sum_{j \in \mathbf{U}(\sigma)} (\bra{j} \cdot K) \cdot \ket{j}) \cdot \ket{i} \ \reduce\  \sum_{i \in \mathbf{U}(\sigma)} \sum_{j \in \mathbf{U}(\sigma)} (\bra{j} \cdot K \cdot \braket{i|j}) \cdot \ket{i},
-    \]
-    where the delta symbol elimination rule returns the term to its original form. The same holds for bra and operator terms.
-\end{proof}
+% \end{lemma}
+% \begin{proof}
+%     Expanding a ket variable twice, for example, results in the following transformation:
+%     \[
+%     \sum_{i \in \mathbf{U}(\sigma)} (\bra{i} \cdot \sum_{j \in \mathbf{U}(\sigma)} (\bra{j} \cdot K) \cdot \ket{j}) \cdot \ket{i} \ \reduce\  \sum_{i \in \mathbf{U}(\sigma)} \sum_{j \in \mathbf{U}(\sigma)} (\bra{j} \cdot K \cdot \braket{i|j}) \cdot \ket{i},
+%     \]
+%     where the delta symbol elimination rule returns the term to its original form. The same holds for bra and operator terms.
+% \end{proof}
 
@@
 
-As the first improvement, we check \( \alpha \)-equivalence using the de Bruijn index~\cite{deBruijn1972lambda}, which replaces references to bound variable names with the distance from the lambda abstraction to the variable.
+\yx{mention de Bruijn normal form.}
+
+% As the first improvement, we check \( \alpha \)-equivalence using the de Bruijn index~\cite{deBruijn1972lambda}, which replaces references to bound variable names with the distance from the lambda abstraction to the variable.
 For instance, the nominal lambda abstraction \( \lambda x.
@@
 The sorted order enables normalization of terms in terms of AC equivalence.
+\yx{rephrase in pseudo code}
 \begin{definition}[Sort Transformation]
@@
 \end{proof}
-
-% Labelled Dirac Notation
-
-\section{Labelled Dirac Notation}
-\label{sec: labelled}
-Labelled Dirac notation uses register names to indicate the quantum system of vectors and operators.
-This allows us to express and reason aboud the states and operations locally, without referring to the whole system.
-For instance, assume $Q$ and $R$ are two registers, we have
-\[
-    M_Q \cdot \ket{\Phi}_{(Q, R)} = ((M \otimes I) \cdot \ket{\Phi})_{(Q, R)}.
-\]
-On the left hand side, the state of two subsystems is $\ket{\Phi}$, and we apply quantum operation $M$ on the system $Q$.
-It is equivalent to extend the operation using identity operators on other subsystems (i.e.
-the cylinder extension), and consider the application in the whole system.
-
-In this section, we introduce the language and typing of registers and labelled Dirac notation, and demonstrate how to transform the equivalence problem into that for the Dirac notation studied above.
-The first new notion is quantum registers, and we assume they are constructed from a set $\cR$.
-\begin{definition}[quantum registers]
-  \begin{align*}
-    R ::= r\in\cR \mid (R, R).
-  \end{align*}
-\end{definition}
-
-Registers can be composed by pairs of $(R, R)$, and this structure corresponds to the structure of tensor product spaces in Dirac notation.
-To reason about the registers, we define their variable set as the enumeration of all register symbols involved.
-
-\begin{definition}[register variable set]
-The variable set of a register is defined inductively by:
-\begin{itemize}
-    \item $\var(R) = \{r\}$, if $R\equiv r$; or
-    \item $\var(R) = \var(R_1) \cup\var(R_2)$, if $R\equiv (R_1,R_2)$.
-\end{itemize}
-\end{definition}
-
-% \textbf{Remark: } Set operations: $\cup$ for union; $\cap$ for intersection; $\setminus$ for difference. So,
-% % \begin{itemize}
-% $ S_1 \cap S_2 \equiv S_1 \cup S_2 \setminus (S_1 \setminus S_2) \setminus (S_2 \setminus S_1) $.
-% \end{itemize}
-
-% The labelled Dirac notation is an extension on the existing type and syntax.
-\begin{definition}[labelled Dirac notation]
-  The \textbf{labelled Dirac notation} includes all Dirac notation symbols and the generators defined below.
-  Here, $s\subseteq \cR$ is a register variable set.
-  \begin{align*}
-    T & ::= \DType(s,s) \mid \mathsf{Reg}(\sigma) \\
-    e & ::= R \mid |i\>_r \mid {}_r\<i| \mid e_R \mid e_{R;R} \mid
-    e \otimes e \otimes \cdots \otimes e \mid e \cdot e
-  \end{align*}
-\end{definition}
-$\DType(s_1, s_2)$ is the unified type for all labelled Dirac notation, where $s_1$ indicates the codomain systems and $s_2$ indicates the domain systems.
-For instance, labelled ket has type $\DType(s_1, \emptyset)$, and labelled bra has type $\DType(\emptyset, s_2)$.
-$\reg(\sigma)$ are types for registers $R$, and the index $\sigma$ indicates the type of Hilbert space represented by the register.
-Terms also include the labelled notation $e_R$ for bra, ket and $e_{R;R}$ for operators.
-We introduce new dot and tensor product symbols for labelled Dirac notation.
-In labelled Dirac notation, the structure of tensor product does not matter, therefore $\otimes$ is an AC symbol.
-Following the unified type $\DType(s,s)$, all kinds of multiplications are represented by the same dot product $e \cdot e$.
-Finally, $\ket{i}_r$ and ${}_r\bra{i}$ are labelled basis for the normal form of labelled Dirac notation, where $r$ are registers symbols in $\mathcal{R}$.
-
-\subsubsection{Typing rules}
-Some typing rules are introduced here.
-The rule for $\DType(s_1, s_2)$ requires that all registers in variable set $s_1$ and $s_2$ are well-typed.
-The rule for $K_R$ demonstrates how a register label is added to the Dirac notation, and the rule for $D^\dagger$ shows that labelled Dirac notation also have symbols for calculation, such as the adjoint.
-\[
-    \frac{E[\Gamma] \vdash  \sigma : \Index}{E[\Gamma] \vdash \reg(\sigma) : \Type}
-    \qquad
-    \frac{E[\Gamma] \vdash r : \reg(\sigma_r) \text{ for all $r$ in $s_1$ and $s_2$} }{E[\Gamma] \vdash \DType(s_1, s_2) : \textsf{Type}} 
-\]
-\[
-    \frac{E[\Gamma] \vdash R : \reg(\sigma)\qquad E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash K_R : \DType(\var R, \emptyset)}
-    \qquad
-    \frac{E[\Gamma] \vdash D : \DType(s_1,s_2)}{E[\Gamma] \vdash D^\dagger : \DType(s_2,s_1)}
-\]
-The dot and the tensor product symbols are different from those in unlabelled Dirac notation.
-Since the goal of labels is to replace the order and structure of tensor products by the reference to registers, the tensor product becomes an AC symbol.
-The typing still checks whether the component subsystems are disjoint with each other.
-\[
-    \frac{
-        E[\Gamma] \vdash D_i : \DType(s_i,s_i') \qquad
-        \bigcap_i s_i = \emptyset \qquad
-        \bigcap_i s_i' = \emptyset
-    }
-    {E[\Gamma] \vdash D_1 \otimes \cdots \otimes D_i : \DType(\bigcup_i s_i, \bigcup_i s_i')}.
-\]
-As for the dot product, the disjointness is considered except registers contracted by multiplication.
-\[
-    \qquad
-    \frac{
-        \begin{aligned}
-            E[\Gamma] \vdash D_1 : \DType(s_1,s_1') \\
-            E[\Gamma] \vdash D_2 : \DType(s_2,s_2')
-        \end{aligned}
-        \qquad 
-        \begin{aligned}
-            s_1 \cap s_2 \backslash s_1' = \emptyset \\
-            s_2' \cap s_1' \backslash s_2 = \emptyset
-        \end{aligned}
-    }
-    {E[\Gamma] \vdash D_1\cdot D_2 : \DType(s_1 \cup (s_2\backslash s_1'), s_2' \cup (s_1'\backslash s_2))}.
-\]
-
-\subsubsection{Normalization}
-The normalization procedure of Dirac notation is extended to check equivalence with labels.
-We add rules to the term rewriting system, which in general try to represent labelled Dirac notation with labelled basis and scalar coefficients.
-The first step is the label elimination.
-Take operator as an example:
-\begin{align*}
-    & O_{R,R'} \ \reduce\ \sum_{i_{r_1}\in\bU(\sigma_{r_1})}\cdots \sum_{i_{r_n}\in\bU(\sigma_{r_n})}
-    \sum_{i_{r'_1}\in\bU(\sigma_{r'_1})}\cdots \sum_{i_{r'_{n'}}\in\bU(\sigma_{r'_{n'}})} \\
-    & \qquad (\<i_R|\cdot O\cdot |i_{R'}\>).(|i_{r_1}\>_{r_i}\otimes\cdots\otimes|i_{r_n}\>_{r_n} \otimes {}_{r'_1}\<i_{r'_1}|\otimes\cdots\otimes{}_{r'_{n'}}\<i_{r'_{n'}}|).
-\end{align*}
-The rules for $e_R$ (ket and bra) are similar. This step reduces all labelled terms $e_R$ or $e_{R;R}$.
-Other symbols on labelled Dirac notation are also reduced by rules like $(D_1 \cdot D_2)^\dagger \ \reduce\ D_2^\dagger \cdot D_1^\dagger$.
-The final step operates on sum and dot product. They will lift summation to the outside, and eliminate the bra-ket pairs whenever possible.
-\begin{align*}
-    & \textrm{(R-SUM-PUSHD0)}
-    && X_1 \otimes \cdots (\sum_{i \in M} D) \cdots \otimes X_2\ \reduce\ \sum_{i \in M} (X_1 \otimes \cdots D \cdots \otimes X_n) \\
-    % %
-    % & \textrm{(R-SUM-PUSHD1)}
-    % && (\sum_{i \in M} D_1) \cdot D_2 \ \reduce\ \sum_{i \in M} (D_1 \cdot D_2) \\
-    %
-    & \textrm{(R-L-SORT0)}
-    && A : \DType(s_1, s_2), B : \DType(s_1', s_2'), s_2 \cap s_1'=\emptyset \Rightarrow A \cdot B \ \reduce\ A \otimes B \\
-    %
-    & \textrm{(R-L-SORT1)}
-    && {}_r\bra{i}\cdot\ket{j}_r \ \reduce\ \delta_{i, j} \\
-    %
-    & \textrm{(R-L-SORT2)}
-    && {}_r\bra{i}\cdot(Y_1 \otimes \cdots \otimes \ket{j}_r \otimes \cdots \otimes Y_m) \ \reduce\ \delta_{i, j}.(Y_1  \otimes \cdots \otimes Y_m)
-    %
-\end{align*}
-% These rules are added to the rewriting system in~\Cref{sec: decide} and executed together.
-In the end, if there are no variables of $\DType(s_1, s_2)$, the expression will be reduced to the addition of big operator sum, and each sum body is labelled basis with Dirac notation scalar coefficients:
-\[
-    \sum_{i}\cdots\sum_{j} a_1 . (\ket{i}_{p} \otimes \cdots \otimes \bra{j}_{q})
-    + \cdots +
-    \sum_{k}\cdots\sum_{l} a_m . (\ket{k}_{r} \otimes \cdots \otimes \bra{l}_{s})
-\]
-In this stage, we only need to check the scalars to decide the equivalence of labelled Dirac notation.
-
-% \begin{lemma}[normal form]
-%     \label{lem: labelled normal form}
-%     For a well-typed term $e$ in $E[\Gamma]$, if $e$ does not contain variables of $\DType(s_1, s_2)$ type, the normal form of $e$ will be
-%     \[
-%     \sum_{i}\cdots\sum_{j} a_1 . (\ket{i}_{p} \otimes \cdots \otimes \bra{j}_{q})
-%     + \cdots +
-%     \sum_{k}\cdots\sum_{l} a_m . (\ket{k}_{r} \otimes \cdots \otimes \bra{l}_{s})
-%     \]
-%     Where $a_i$ are scalar typed Dirac notation.
-% \end{lemma}
-% \begin{proof}
-%     Labelled Dirac notation of $e_R$, $e_R;R$, 
-% \end{proof}

### 2025-01-27 05tool_eval.tex
rearranged

--- 2025-01-27 03tool_eval.tex
@@
 
-\subsubsection{Use Case}
-Here we encode the motivating~\Cref{ex: motivating}, examine and explain how it is checked in D-Hammer. The encoding is shown below.
-
-    \begin{lstlisting}[style=dhammer]
-Var T : INDEX. Var M : OTYPE[T, T].
-Def phi := idx T => Sum nv in USET[T], |(nv, nv)>.
-Var r1 : REG[T]. Var r2 : REG[T].
-CheckEq M_r1;r1 (phi T)_(r1, r2) with (TPO T T M)_r2;r2 (phi T)_(r1, r2).
-    \end{lstlisting}
-
-The first three lines use the \texttt{\textcolor{NavyBlue}{Var}} and \texttt{\textcolor{NavyBlue}{Def}} commands to set up the context for the Dirac notation.
-\texttt{T} is a type index, representing arbitrary Hilbert space types.
-\texttt{M} is assume to be an operator in the Hilbert space with type \texttt{T}.
-\texttt{phi} is defined as the maximally entangled state, depending on the bound variable \texttt{T} as index.
-\texttt{r1} and \texttt{r2} are register names for the two subsystems.
-
-In the left hand side of \texttt{\textcolor{NavyBlue}{CheckEq}} command, \texttt{M\_r1;r1} denotes the labelled notation $M_{r_1; r_1}$, and \texttt{(phi T)\_(r1, r2)} denotes the entangled state $\ket{\Phi}_{(r_1, r_2)}$.
-They are connected by a white space, which is parsed into the composition of Dirac notation, and will be reduced into the operator-ket multiplication after typing.
-The right hand side is interpreted similarly, except the defined symbol \texttt{TPO} in the environment:
-
-\begin{lstlisting}[style=dhammer]
-Def TPO := idx sigma => idx tau => fun O : OTYPE[sigma, tau] => Sum i in USET[sigma], Sum j in USET[tau], (<i| O |j>).(|j> <i|).
-\end{lstlisting}
-
-The \texttt{TPO} symbol represents the transpose of operators, and encodes the formalization in~\Cref{ex: formalizing motivating}.
-Thanks to the design of environment and functions, many other commonly used symbols in Dirac notation are encoded and provided as defined symbols in D-Hammer.
-
-Within one second, the prover reports the result of equivalence with their common normal form:
-    \begin{lstlisting}[style=dhammer]
-The two terms are equal.
-[Normalized Term] SUM[USET[T], FUN[BASIS[T], SUM[USET[T], FUN[BASIS[T], SCR[DOT[BRA[$\texttt{\$1}$], MULK[M, KET[$\texttt{\$0}$]]], LTSR[LKET[$\texttt{\$1}$, r1], LKET[$\texttt{\$0}$, r2]]]]]]] : DTYPE[RSET[r1, r2], RSET]
-    \end{lstlisting}
-
-The normal form is in the internal syntax tree format mentioned above. A more readable interpretation is:
-\[
-\sum_{\mathbf{U}(T)} \sum_{\mathbf{U}(T)} \bra{\$1}M\ket{\$ 0} . \ket{\$1}_{r_1} \otimes \ket{\$0}_{r_2} : \DType(\{r_1, r_2\}, \emptyset).
-\]
-Here $\$0$ and $\$1$ are de Bruijn indices. The result is a ket on the $\{r_1, r_2\}$ system as expected, and follows pattern proposed in~\Cref{sec: labelled}.
-
 \subsubsection{Benchmark performance}
@@
 We also built an example benchmark for labelled Dirac notation, see~\Cref{sec: examples for labelled}.
+
+\yx{Mention that our tool can do more complicated examples.}

### 2025-01-27 06related_conclusion.tex
rearranged

\section{Related Work}

Automated theorem proving has seen significant advancements in recent years, particularly through the development of satisfiability modulo theories (SMT) solvers,
% SMT solvers extend propositional satisfiability by integrating theories such as arithmetic, arrays, and bit-vectors, making them suitable for solving a wide range of verification problems in both hardware and software domains. 
including prominent tools like Z3. These solvers have become essential in various fields such as formal verification, synthesis, and model checking. 
Equational reasoning is another crucial area of research within automated theorem proving, which focuses on solving problems that involve equations between terms in an algebraic structure. 
Equational provers, such as Vampire and E, have played a pivotal role in addressing the challenge of proving equations in first-order logic, employing sophisticated algorithms like superposition and term rewriting.

Formal verification of quantum computation is receiving increasing attentions during these years. See~\cite{Lewis2023} for a comprehensive review. 
Verification frameworks in Coq include the foundational formalization CoqQ and quantum circuit language QWIRE. 
Verifications of quantum programs are also considered, such as the Hoare logic based methods [] and model checking based methods [].
The equational reasoning of Dirac notation is crucial for establishing property proofs in these works.
Verification theoreis and tools based on other languages are also proposed, such as PyZX for reasoning and simplification of ZX-calculus.

\section{Conclusion and Future Work}
Based on the first equational reasoning tool for Dirac notation called DiracDec, this work improves and extends the theory for practical applications, and provides the solver D-Hammer. Experiments show that the tool demonstrates advantages in decidability, efficiency and usability.

We expect D-Hammer to have applications in areas like quantum program verification or proofs of post-quantum cryptography protocols in the future.
One promising following up is to connect D-Hammer with theorem provers like Coq. It involves transforming theorem prover expressions into D-Hammer, and verify the proof trace of D-Hammer in theorem provers. Besids, most quantum program verifiers nowadays depend on matrix calculations. D-Hammer can serve as the replacement for matrix methods to enable symbolic deductions.

### 2025-01-28 01intro_preliminary.tex
.

@@
 
-In 1939, Dirac introduced his notation for quantum mechanics~\cite{dirac1939new}, designed to represent linear algebraic formulas in a compact and convenient form.
+Dirac notation~\cite{dirac1939new}, a.k.a.\, bra-ket notation, is a
+mathematical formalism for representing quantum states using linear
+algebra notation: for instance, Dirac notation uses the linear
+combination \( a\ket{\psi} + b\ket{\phi} \) to represent the
+superposition of the two quantum states \( \ket{\psi} \) and
+\( \ket{\phi} \). Another essential ingredient of Dirac notation
+is the tensor product $\otimes$, which is used to describe composite
+states: for instance, Dirac notation uses the tensor expression
+\( \ket{\psi} \otimes \ket{\phi} \) to denote the composition of
+the two quantum states \( \ket{\psi} \) and \( \ket{\phi} \). It is
+also common to use a variant of Dirac notation, called labelled Dirac
+notation, to describe composite quantum states. In labelled Dirac
+notation, bras and kets are tagged with labels that identify the
+subsystems on which they operate. For instance, the labelled tensor
+expression
+\( \ket{\psi}_{S'} \otimes \ket{\phi}_{S} \)
+states that $\ket{\psi}_S$ and $\ket{\psi}_{S'}$ describe two quantum
+states over some subsystem $S$ and $S'$ respectively. By considering
+the relationship between $S$ and $S'$, one can obtain identities for
+free, e.g.\, 
+%
+$$\ket{\phi}_S \otimes \ket{\psi}_{T} = \ket{\psi}_{T} \otimes \ket{\phi}_S
+\mbox{if}~S\cap S'=\emptyset$$
+%
+In turn, commutativity of the tensor product ensures that one can
+reason locally about quantum systems, and contributes to making
+labelled Dirac notation a convenient, compositional formalism to
+reason about quantum states---in a way that is similar to the way
+bunched logics support compositional reasoning about mutable states.
+
+Labelled Dirac notation is also commonly used to express assertions
+about quantum programs. Specifically, many quantum Hoare logics use
+labelled Dirac notation to express program assertions, including
+pre-conditions and post-conditions. These logics also use implicitly
+equational reasoning between labelled Dirac expressions, to glue
+applications of proof rules---similar to the use of the rule of
+consequence in the setting of classical program. Therefore, it is
+essential for the verification of quantum programs to have automated
+means of proving equality of two complex expressions based on labelled
+Dirac notation.
+
+\subsubsection*{Contributions}
+This paper presents an automated tool, called D-Hammer, for reasoning
+about labelled Dirac notation. D-Hammer uses a rich, dependently typed
+language, to formalize labelled Dirac notation, and supports common
+idioms used to describe quantum systems, including big operators of
+the form $\sum_{i\in I} a_i \ket{\phi_i}_S$ to represent indexed
+superpositions of states. The semantics of typable expressions is
+given in terms of Hilbert spaces, tailored to interpreting tensor
+product as an associative and commutative operator. We leverage this
+interpretation to define a rich equational theory for labelled Dirac
+notation, and prove its soundness with respect to our denotational
+semantics. Finally, we define an efficient normalization procedure to
+prove equivalence of two expressions. We then evaluate our procedure
+with respect to examples from the literature. Our evaluation covers
+examples based on (vanilla, unlabelled) Dirac notation, as well as
+examples based on labelled Dirac notation. The main conclusions are
+that our approach outperforms prior work by~\cite{diracdec} and is
+able to prove complex examples of reasoning with labelled Dirac
+notation from the literature, including examples from prior work on
+quantum separation logic~\cite{DBLP:conf/lics/ZhouBHYY21}.
+
+%% In 1939, Dirac introduced his notation for quantum mechanics~\cite{dirac1939new}, designed to represent linear algebraic formulas in a compact and convenient form.
 For example, the expression \( a\ket{\psi} + b\ket{\phi} \) represents the superposition of two quantum states, \( \ket{\psi} \) and \( \ket{\phi} \).
@@
 
-In quantum algorithm formalizations and quantum programming languages, Dirac notation is frequently used in equational proofs, which are critical, repetitive, and often time-intensive.
+%% In quantum algorithm formalizations and quantum programming languages, Dirac notation is frequently used in equational proofs, which are critical, repetitive, and often time-intensive.
 These notations also play a key role in defining program states, operations, and assertions in quantum programming languages.
@@
 
-Recently, Xu et al.~\cite{diracdec} proposed a theory for deciding the equivalence of Dirac notation, alongside a prototype implementation in Mathematica called DiracDec.
+%% Recently, Xu et al.~\cite{diracdec} proposed a theory for deciding the equivalence of Dirac notation, alongside a prototype implementation in Mathematica called DiracDec.
 They demonstrated that the equivalence of basic Dirac notation is decidable.
@@
 
-One challenge is the efficiency of the algorithm when dealing with equivalences beyond the scope of term rewriting.
+%% One challenge is the efficiency of the algorithm when dealing with equivalences beyond the scope of term rewriting.
 DiracDec decides the entire equational theory by rewriting modulo \( E \), where \( E \) represents a set of axioms that cannot be decided by normalization alone, such as the associativity and commutativity (AC) of certain function symbols.
@@
 
-Usability is another area where DiracDec falls short.
+%% Usability is another area where DiracDec falls short.
 It does not support labelled Dirac notation, which uses registers to denote subsystems and express states locally.
@@
 
-The system design of DiracDec reflects a trade-off between simplicity and efficiency.
+%% The system design of DiracDec reflects a trade-off between simplicity and efficiency.
 While the simplicity of term rewriting allows for strong theoretical results, it limits optimization opportunities.
@@
 Our main technical contributions are:
-\begin{itemize}
-    \item An efficient algorithm for deciding the equational theories in \( E \), based on equivalence checking through normalization, with the normal form for \( E \) being obtained via sorting.
-    \item Support for constant register labels, and reducing the equivalence decision of labelled Dirac notation to the unlabelled case.
-    \item A more user-friendly \CC\ solver, D-Hammer, featuring an abstract language and typing system. We also support the definition of symbols (e.g., transpose and trace) using function syntax.
-\end{itemize}
+%% \begin{itemize}
+%%     \item An efficient algorithm for deciding the equational theories in \( E \), based on equivalence checking through normalization, with the normal form for \( E \) being obtained via sorting.
+%%     \item Support for constant register labels, and reducing the equivalence decision of labelled Dirac notation to the unlabelled case.
+%%     \item A more user-friendly \CC\ solver, D-Hammer, featuring an abstract language and typing system. We also support the definition of symbols (e.g., transpose and trace) using function syntax.
+%% \end{itemize}
 
-We evaluated D-Hammer against the DiracDec benchmark and new examples involving labelled Dirac notation.
+%% We evaluated D-Hammer against the DiracDec benchmark and new examples involving labelled Dirac notation.
 The results show significant improvements in both decidability and efficiency compared to DiracDec.
 
-% D-Hammer successfully decides all the examples that are expressable in its language, including those failed by DiracDec because of complexity or insufficient decision power.
+%% % D-Hammer successfully decides all the examples that are expressable in its language, including those failed by DiracDec because of complexity or insufficient decision power.
 
-\section{Motivation and Preliminary}
+\section{Motivation and Preliminaries}
 

### 2025-01-28 02language_design.tex
language improved

@@
 \section{Language, Typing and Semantics}
-The first step is to formalize the language for Dirac notation.
-In DiracDec, the language has a concrete design, where the same syntax for different types corresponds to distinct symbols.
-Our goal is to transition from this concrete design to a more abstract one, which aligns more closely with the conventional Dirac notation we use and simplifies the term rewriting system.
+We use a typed abstract syntax to represent Dirac notation. The overall system design leads to the following advantages. 
+\begin{itemize}
+    \item Dirac notation can be divided into four types, namely scalar, ket, bra and operator.
+Many function symbols have constraints on types.
+It is important to have a typing system and check the validity of expressions.
+    \item It has functions and higher-order typing. It allows to formally encode defined symbols like transpose or trace, which are usually used to reprensent the term in an abstract manner. 
+    \item We use an abstract syntax for better and simpler representations.
+It follows the conventional Dirac notation, where we extensively use the composition symbol to denote calculations, and symbols like scalar multiplications for different types are unified as one.
+    \item The structures are friendly to optimizations for efficiency. For example, associative and commutative (AC) symbols have indefinite arities, which enables normalization by sorting.
+\end{itemize}
 
-To achieve this, we organize our language into three layers: the index, the type, and the term.
+Since a Hilbert space $\mathcal{H}_V$ is dependent on the basis set $V$,
+types for Dirac notation also depends on the type index.
+Therefore, the language is organized into three layers: the index, the type, and the term. 
 Terms represent concrete instances such as kets, bras, and operators, which will be typed and checked.
@@
         e ::=\ & x \mid \lambda x : T.e \mid \mu x.e \mid e_1\ e_2 \mid (e_1, e_2) \\
-        & |\ 0 \mid 1 \mid \ADDS(e_1, \cdots, e_n) \mid e_1 \times \cdots \times e_n \mid e^* \mid \delta_{e_1, e_2} \mid \DOT(e_1, e_2) \\
+        & |\ 0 \mid 1 \mid e_1 \times \cdots \times e_n \mid e^* \mid \delta_{e_1, e_2} \\
         & |\ \ZEROK(\sigma) \mid \ZEROB(\sigma) \mid \ZEROO(\sigma_1, \sigma_2) \mid \ONEO(\sigma) \\
-        & |\ \ket{e} \mid \bra{t} \mid e^\dagger \mid e_1.e_2 \mid \ADD(e_1, \cdots, e_n) \mid e_1 \otimes e_2 \\
-        & |\ \MULK(e_1, e_2) \mid \MULB(e_1, e_2) \mid \OUTER(e_1, e_2) \mid \MULO(e_1, e_2) \\
+        & |\ \ket{e} \mid \bra{t} \mid e^\dagger \mid e_1.e_2 \mid e_1 + \cdots + e_n \mid e_1 \otimes e_2 \mid e_1 \cdot e_2 \\
         & |\ \mathbf{U}(\sigma) \mid e_1 \star e_2 \mid \sum_{e_1} e_2.
@@
 \end{definition}
-The terms above are explained in order:
-\( \lambda x : T.e \) represents the abstraction for normal functions, and \( \mu x.e \) represents the abstraction for index functions.
+The terms above are explained in five lines.
+\begin{enumerate}
+    \item \textbf{function and basis}: \( \lambda x : T.e \) represents the abstraction for normal functions, and \( \mu x.e \) represents the abstraction for index functions.
     \( e_1\ e_2 \) denotes function application.
     \( (e_1, e_2) \) is the basis pair for product types.
-The constants \( 0 \), \( 1 \), \( \ADDS \), \( e_1 \times \cdots \times e_n \), and \( e^* \) are symbols for scalars.
-The next line includes constant symbols for kets, bras, and operators.
-\( \ket{e} \) is a ket, \( \bra{t} \) is a bra, and \( e^\dagger \) denotes the conjugate transpose of \( e \).
+    \item \textbf{scalar}: \( 0 \), \( 1 \), \( e_1 \times \cdots \times e_n \) and \( e^* \) are symbols for scalars.
+    \( \delta_{e_1, e_2} \) compares where two basis are the same and evaluates to $1$ or $0$ accordingly.
+    \item \textbf{Dirac constant}: zero ket, zero bra, zero operator and identity operator.
+    \item \textbf{Dirac function}: \( \ket{e} \) is a ket, \( \bra{t} \) is a bra, and \( e^\dagger \) denotes the conjugate transpose of \( e \).
 \( e_1.e_2 \) represents scaling the term \( e_2 \) by scalar \( e_1 \).
-\( \mathbf{U}(\sigma) \) denotes the universal set with index \( \sigma \).
+\(e_1 + \cdots + e_n\) is the addition.
+\( e_1 \otimes e_2 \) represents tensor product, and \( e_1 \cdot e_2 \) represents the multiplication.
+    \item \textbf{summation}: \( \mathbf{U}(\sigma) \) denotes the universal set with index \( \sigma \).
 \( e_1 \star e_2 \) represents the Cartesian product of \( e_1 \) and \( e_2 \).
@@
 X \).
-Additionally, \( \ADDS \) and \( \ADD \) are distinct AC symbols: \( \ADDS \) is used for scalar addition and \( \ADD \) for linear algebra addition.
-There are five kinds of linear algebraic multiplications among kets, bras, and operators.
-These are encoded using five different symbols: \( \DOT \), \( \MULK \), \( \MULB \), \( \OUTER \), and \( \MULO \).
+\end{enumerate}
 
-Compared to DiracDec, the symbols for addition, adjoint, scaling, and tensor products have been consolidated.
-The multiplications are still kept separate due to their differing properties.
-These operations are denoted as \( B \cdot K \), \( K_1 \cdot K_2 \), \( B_1 \cdot B_2 \), \( K \cdot B \), and \( O_1 \cdot O_2 \), respectively.
+The scalar multiplication $\times$ and addition $+$ are AC symbols, and they have indefinite arity. 
+Because we use abstract syntax, the Dirac functions have different interpertations for different types. 
+We use letters like $a, b, c$ to represent scalar variables, $K$ and $B$ to represent ket and bra variables, and $O$ for operators.
+Therefore, $O \cdot K$ is interpreted as the operator-ket multiplication, and scalars can also be constructed from inner products $B \cdot K$.
 

### 2025-01-28 06related_conclusion.tex
language improved

@@
 \section{Related Work}
+
+The first step is to formalize the language for Dirac notation.
+In DiracDec, the language has a concrete design, where the same syntax for different types corresponds to distinct symbols.
+Our goal is to transition from this concrete design to a more abstract one, which aligns more closely with the conventional Dirac notation we use and simplifies the term rewriting system.
 

### 2025-01-28 01intro_preliminary.tex

@@
 labelled Dirac notation to express program assertions, including
-pre-conditions and post-conditions. These logics also use implicitly
-equational reasoning between labelled Dirac expressions, to glue
-applications of proof rules---similar to the use of the rule of
+pre-conditions and post-conditions~\cite{TODO}. These logics also use
+implicitly equational reasoning between labelled Dirac expressions, to
+glue applications of proof rules---similar to the use of the rule of
 consequence in the setting of classical program. Therefore, it is

### 2025-01-28 02language_design.tex
2025/1/28

@@
 
-The typing system is responsible for classifying terms within a proof system, using a context that specifies the types of variables and definitions.
-The context is divided into two components: the environment \( E \), which includes definitions and assumptions, and the context \( \Gamma \), which tracks bound variables.
+The typing system is responsible for classifying terms within a proof system, according to the types of variables and definitions.
+We use an environment \( E \) and context \( \Gamma \) for typing.
+\( E \) includes definitions and assumptions, and the context \( \Gamma \) tracks bound variables during the proof.
 Both \( E \) and \( \Gamma \) consist of sequences of assumptions \( x : T \) or definitions \( x := t : T \).
@@
 
-Type checking in our language involves maintaining a well-formed environment and context \( E[\Gamma] \).
 We say an expression \( t \) has type \( X \) in context \( E[\Gamma] \) if the typing judgment \( E[\Gamma] \vdash t : X \) can be proven through the rules in~\Cref{sec: full typing rules}. 
-The well-formedness of the context \( \WF(E)[\Gamma] \) is built incrementally:
+It involves maintaining a well-formed environment and context \( E[\Gamma] \), which is built incrementally:
 \[
@@
 The idea behind the interpretation of types and terms is to formalize the typing relation using set-theoretic inclusion.
-Specifically, for a well-formed context \( E[\Gamma] \), term \( t \), and type \( T \), if \( E[\Gamma] \vdash t : T \), then for any valuation \( v \), the semantics of \( t \) must lie within the semantics of \( T \).
+Specifically, for a well-formed context \( E[\Gamma] \), term \( t \), and type \( T \), if \( E[\Gamma] \vdash t : T \), then for any valuation \( v \), the semantics of \( t \) must belong to the semantics of \( T \).
 
@@
 and the bilinearity of the dot product:
-\[
-(a.X) \cdot Y = a \cdot (X \cdot Y), \quad X \cdot (Y_1 + Y_2) = X \cdot Y_1 + X \cdot Y_2, 
-\]
-\[
+\begin{align*}
+(a.X) \cdot Y = a \cdot (X \cdot Y), \quad X \cdot (Y_1 + Y_2) = X \cdot Y_1 + X \cdot Y_2, \\
 X \cdot (a.Y) = a \cdot (X \cdot Y), \quad (X_1 + X_2) \cdot Y = X_1 \cdot Y + X_2 \cdot Y.
-\]
-
-As mentioned in the introduction, a subset \( E \) of these axioms cannot be decided by term rewriting alone. The axioms in \( E \) include:
-\[
-\begin{aligned}
+\end{align*}
+The entire axioms are separated into $R$ and $F$ two sets.
+During equivalence checking, most of the axioms are decided by term rewriting, which form the set $R$. However, there are axioms requiring special algorithms, which are collected in the set $F$.
+\begin{definition}[axiom set F]
+\label{def: axiom F}
+\begin{align*}
     \text{AC-equivalence} &\quad \text{e.g.,} \quad X + Y = Y + X, \quad (X + Y) + Z = X + (Y + Z), \\
@@
     \text{scalar theories} &\quad \text{e.g.,} \quad a + 0 = a, \quad a \times (b + c) = a \times b + a \times c.
-\end{aligned}
-\]
+\end{align*}
+\end{definition}
+
 The scalar theories are treated separately as a module and are not considered in this work. In the implementation, we use the Mathematica kernel to decide scalar equivalences.
+
 These equational axioms provide an operable theory for the proof automation algorithm.
@@
 
-Thus, the axioms are sound with respect to denotational semantics.
-However, the reverse does not hold: there exist equivalences in denotational semantics that cannot be captured by these axioms.
-Nevertheless, our work focuses on solving practical examples, which are fully covered by the axioms in the experiments.
-
-To conclude this section, we demonstrate the formalization of the symbols used in the motivating example.
+With these syntax, typing and semantics introduced, we demonstrate the formalization of the symbols used in the motivating example to conclude this section.
 \begin{example}[Formalizing the Motivating Example]

### 2025-01-28 04efficient_alg.tex
2025/1/28

@@
 \label{sec: decide}
+The equivalence of Dirac notations are decided by normalization, which transforms equivalent expressions under a set of axioms into the same syntax.
+For the entire axiom set $R \cup F$ of Dirac notation, the previous algorithm conducts partial normalization on $R$ by term rewriting modulo $F$, and axioms in $F$ are checked by iterating through all permutations.
+This direct solution has factorial compleixty on maximum number of AC symbol arguemnts, and is observed to be critical to the low efficiency.
+Here we propose an efficient algorithm to perform total nomralization on $R \cup F$, and their comparison are illustrated below.
+\[
+\text{(partial)}\qquad e_1 \mathop{\twoheadrightarrow}^R e_1' \mathop{=}^F e_2' \mathop{\twoheadleftarrow}^R e_2
+\qquad
+\text{(total)}\qquad e_1 \mathop{\twoheadrightarrow}^{R \cup F} e \equiv e \mathop{\twoheadleftarrow}^{R \cup F} e_2
+\]
 
-We utilize the term rewriting technique to decide the equivalence of Dirac notation, based on its axiomatic semantics.
-The process involves normalizing two terms and checking whether they have identical syntax.
-This normalization is achieved through matching and substitution in universal algebra, using a set of predefined rewriting rules.
+% The problem is that, the axiom set $F$ in~\Cref{def: axiom F} cannot be decided through term rewriting.
+A direct solution is to decide \( E \) by examining all possible permutations in the rewriting result, which has high complexity.
+% In this work, we introduce a more efficient approach by incorporating sorting algorithms into the normalization procedure.
 
-As mentioned in the semantics section, the axioms in \( E \) cannot be fully decided through term rewriting.
-DiracDec decides \( E \) by examining all possible permutations in the rewriting result.
-In this work, we introduce a more efficient approach by incorporating sorting algorithms into the normalization procedure.
-
-The normalization procedure consists of the following steps.
+The normalization procedure consists of the following steps. Steps 1 - 3 are term rewriting steps for $R$, and steps 4 - 6 are our algorithm to further normalize axiom set $F$.
 \begin{enumerate}
     \item \textbf{First Rewritings}: Expand definitions and simplify expressions.
-    \item \textbf{Variable Expansion}: Consider scalar expressions to ensure completeness.
-    \item \textbf{Second Rewritings}: Normalize terms modulo \( E \).
+    \item \textbf{Variable Expansion}: Convert to scalar expressions for better decidability.
+    \item \textbf{Second Rewritings}: Normalize terms on \( R \) modulo \( F \).
     \item \textbf{Sorting Without Bound Variables}: Normalize AC-equivalence.
@@
 \end{enumerate}
-
-% \subsection{Normalization Modulo \( E \) by Term Rewriting}
 
@@
 
-\subsection{Deciding Equational Theory \( E \)}
-
-In the Mathematica implementation of DiracDec, the equational theory \( E \) is decided using unification, which attempts to find a substitution for the summation bound variables that makes two expressions syntactically equivalent.
-This unification process iterates through all permutations of AC symbol arguments, and its complexity is factorial in the maximum number of AC symbol arguments.
-
-\yx{mention de Bruijn normal form.}
-
-% As the first improvement, we check \( \alpha \)-equivalence using the de Bruijn index~\cite{deBruijn1972lambda}, which replaces references to bound variable names with the distance from the lambda abstraction to the variable.
+Now we analyse the axioms in $F$ to understand the difficulty and solution for normalization. 
+For $\alpha$-equivalence, we want to rule out the influence of bound variable names.
+Therefore we use de Bruijn notation~\cite{deBruijn1972lambda}, which replaces the name with the distance from the lambda abstraction to the variable.
 For instance, the nominal lambda abstraction \( \lambda x.
 x \) is transformed into \( \lambda .
-0 \), while \( \lambda x.
+\$0 \), while \( \lambda x.
 \lambda y.
 (x\ (y\ x)) \) is transformed into \( \lambda.\lambda.
-(1\ (0\ 1)) \).
-This transformation into the de Bruijn index is performed as the final step in the normalization process to check equivalence between terms with different bound variable names.
+(\$1\ (\$0\ \$1)) \).
 
@@
 
-To address this issue, we propose an algorithm that sorts terms without considering bound variables.
+To address this issue, we propose an algorithm to sort in two steps.
 The key observation is that in a successive sum expression \( \sum_{i \in s_1} \cdots \sum_{j \in s_n} A \), the names and order of the bound variables \( i, \dots, j \) can be freely permuted.
-Therefore, all bound variables should be treated uniformly during sorting, and the order of summation can then be determined based on the position of the bound variables.
+Therefore, a good idea is to normalize AC-equivalence first, where all bound variables are treated uniformly.
+Afterwards, the order of summation can then be determined based on the position of the bound variables.
+
+% \begin{align*}
+%     \sum_{i \in s_1} \sum_{j \in s_2} \bra{i} A \ket{j} \times \bra{j} B \ket{i} 
+%     \qquad
+%     \sum_{i \in s_2} \sum_{j \in s_1} \bra{i} B \ket{j} \times \bra{j} A \ket{i} \\
+%     \sum_{i \in s_1} \sum_{j \in s_2} \bra{i} A \ket{j} \times \bra{j} B \ket{i} 
+%     \qquad 
+%     \sum_{i \in s_2} \sum_{j \in s_1} \bra{j} A \ket{i} \times \bra{i} B \ket{j} \\
+%     \sum_{i \in s_1} \sum_{j \in s_2} \bra{i} A \ket{j} \times \bra{j} B \ket{i} \\
+%     \sum_{s_1} \sum_{s_2} \bra{\$1} A \ket{\$0} \times \bra{\$0} B \ket{\$1}
+% \end{align*}
 
@@
 In this context, variables and constants are treated as functions with zero arguments.
-
 \begin{definition}[Order Without Bound Variables]
@@
 The purpose of this ordering is to compare function symbols in a top-down manner while ignoring bound variables.
-The sorted order enables normalization of terms in terms of AC equivalence.
-\yx{rephrase in pseudo code}
+This order enables normalization of terms for AC equivalence.
 \begin{definition}[Sort Transformation]
     For a term $e$ with bound variable set $\mathcal{B}$,
-    The sort transformation sort(e) is defined as
-    \begin{itemize}
-        \item $e$, if $e$ is a variable or constant;
-        \item $\lambda x:T. \textrm{sort}(e')$, if $e \equiv \lambda x : T. e'$;
-        \item $\mu x. \textrm{sort}(e')$, if $e \equiv \mu x. \textrm{sort}(e')$; or
-        \item $f(\textrm{sort}(a_1), \cdots, \textrm{sort}(a_n))$, if $e \equiv f(a_1, \cdots, a_n)$.
-If $f$ is an AC symbol, then the order of sort($a_i$) is sorted ascendingly according to $<_\mathcal{B}$.
-    \end{itemize}
+    The sort transformation is defined in~\Cref{alg: sort}.
 \end{definition}
+
+\begin{algorithm}
+    \caption{Sort Transformation}
+    \label{alg: sort}
+    \begin{algorithmic}[1]
+        \Procedure{Sort}{$e, \mathcal{B}$}
+            \If{$e \equiv \lambda x : T . e'$}
+                \State \Return $\lambda x : T . \textsc{Sort}(e')$
+            \ElsIf{$e \equiv \mu x. e'$}
+                \State \Return $\mu x . \textsc{Sort}(e')$
+            \ElsIf{$e \equiv f(a_1, \cdots, a_n)$}
+                \State $ls := \textsc{Sort}(a_1), \cdots, \textsc{Sort}(a_n)$
+                \State $ls := ls$ sorted by $<_\mathcal{B}$
+                \State \Return $f(ls)$
+            \EndIf
+        \EndProcedure
+    \end{algorithmic}
+\end{algorithm}
 
@@
 \begin{definition}[Swap Transformation]
-For a term \( e \) with a sorting result \( \textrm{sort}(e) \), the swap transformation proceeds by ordering all bound variables according to their first appearances, except in function definitions \( \lambda x \) and \( \mu x \).
+For a term \( e \) with a sorting result \( \textsc{Sort}(e) \), the swap transformation proceeds by ordering all bound variables according to their first appearances, except in function definitions \( \lambda x \) and \( \mu x \).
 The swap transformation then reorders the successive summations accordingly.
@@
 
-The algorithm that applies the sort and swap transformations, followed by de Bruijn normalization, successfully handles all equivalences in our benchmark.
-We formalize the completeness of this transformation in the following conjecture:
-
-\begin{conjecture}[Completeness of Transformation]
-    For any two terms \( e_1 \) and \( e_2 \) that are equivalent under AC equivalence, SUM-SWAP, and \( \alpha \)-equivalence, they will have the same form after sort, swap, and de Bruijn transformations.
-\end{conjecture}
+Take the term $\sum_{i \in s_2} \sum_{j \in s_1} \bra{i} B \ket{j} \times \bra{j} A \ket{i}$ as an example.
+Its bound variable set $\mathcal{B} = \{i, j\}$.
+Assume we have $A < B$, then the sorting result will be $\sum_{i \in s_2} \sum_{j \in s_1} \bra{j} A \ket{i} \times \bra{i} B \ket{j}$.
+Then we decide that the order for bound variables will be $j < i$ because $j$ appears first in the body.
+Using the swap transformation, the sorted result will be $\sum_{j \in s_1} \sum_{i \in s_2} \bra{j} A \ket{i} \times \bra{i} B \ket{j}$.
 

### 2025-01-29 03labelled.tex
.

@@
+\def\>{\ensuremath{\rangle}}
+\def\<{\ensuremath{\langle}}
+
+\newcommand {\cnot } {{\mathsf{CNOT}}}
+\newcommand {\swap } {{\mathsf{SWAP}}}
+
 % Labelled Dirac Notation
@@
 \label{sec: labelled}
+
+\subsection{Why Labelled Dirac Notation Important?}
+
+Dirac notation provides an intuitive and concise mathematical expression for describing quantum states and system evolution.
+However, its description of local evolution in multi-body systems can still be further simplified.
+For example, when we attempt to describe the local evolution of the (unnormalized) GHZ entangled state on three qubits $p,q,r$ using standard Dirac notation, specifically applying the CNOT gate to $p,r$, we use the formula:
+$$(\cnot\otimes I)(|000\>+|111\>)$$
+and follow the conventions for writing and interpreting it as follows: 1.
+Arrange $p,q,r$ in the conventional order, 2.
+Lift the local operation to the global level, i.e., $\cnot\otimes I$ because CNOT acts on $p,r$ (the first and second systems) in sequence, and the operation on $q$ (the third system) is equivalent to identity operation $I$.
+This is also why we arrange them as $p,r,q$ instead of $p,q,r$.
+We quickly realize that this method becomes involved for complex situations since the conventional order cannot guarantee that 1.
+the order of all local operations is consistent with it, 2.
+the local operations involve adjacent qubits in the conventional order, as illustrated below.
+When we continue to apply the CNOT operation to $q,r$ and $p,q$, we need to write it as:
+$${\color{blue}(I\otimes \swap)(\cnot\otimes I)(I\otimes \swap)}{\color{red}(I\otimes (\swap\cdot \cnot\cdot \swap))}(\cnot\otimes I)(|000\>+|111\>),$$
+since: $q,r$ are reversely ordered, so the lift of CNOT on $q,r$ is not $I\otimes \cnot$, but $I\otimes (\swap\cdot \cnot\cdot \swap)$, where SWAP is used to temporarily swap the $q,r$ systems; $p,q$ are not adjacent, and their lift cannot be simply described as the tensor product of CNOT and $I$.
+Instead, we need to use SWAP to temporarily swap $q,r$, $(I\otimes \swap)(\cnot\otimes I)(I\otimes \swap)$, where $(I\otimes \swap)$ indicates the temporary swap of $q,r$.
+
+This significantly limits the application of standard Dirac notation in multi-body systems.
+To address this, physicists introduce labels (or subscripts) to indicate the systems on which quantum states or operations are applied, thereby avoiding unnecessary lifting and employment of swap gates.
+For example, rewriting the above example using labels:
+$$\cnot_{p,q}\cnot_{q,r}\cnot_{p,r}(|000\>_{pqr}+|111\>_{pqr}),$$
+without the need of determining the conventional order nor lifting using additional SWAPs. 
+The introduction of labels allows us to avoid deliberately maintaining the order of qubits.
+Tensor products become AC symbols, so we can rearrange qubits as needed to perform calculations.
+For example, we can perform the calculation as follows:
+\begin{align*}
+  &\cnot_{p,q}\cnot_{q,r}\cnot_{p,r}(|000\>_{pqr}+|111\>_{pqr}) \\
+  =\ &\cnot_{p,q}\cnot_{q,r}((\cnot|00\>)_{pr}|0\>_q + (\cnot|11\>)_{pr}|1\>_r) \\
+  =\ &\cnot_{p,q}\cnot_{q,r}(|000\>_{pqr} + |110\>_{pqr}) \\
+  =\ &\cnot_{p,q}((\cnot|00\>)_{qr}|0\>_p + (\cnot|10\>)_{qr}|1\>_p) \\
+  =\ &\cnot_{p,q}(|000\>_{pqr} + |111\>_{pqr}) \\
+  =\ &((\cnot|00\>)_{pq}|0\>_r + (\cnot|11\>)_{pq}|1\>_r) \\
+  =\ &|000\>_{pqr} + |101\>_{pqr}
+\end{align*}
+In addition, it can very freely describe local measurements, partial traces (the state or evolution of subsystems in multi-body systems), and partial inner products (the representation of partial traces in pure states), which are sufficient to handle the mathematical formulas of quantum mechanics in multi-body systems.
+
+Labelled Dirac notation is not only ubiquitous in the description of multi-body systems, but also plays an important role in program logic -- just as in classical program logic, variable names are used to construct logical formulas instead of describing them as global memory functions, in quantum program logic, variable names are used to label the subsystems on which quantum gates act, rather than lifting them to the global system -- the latter would cause exponential length of the formula w.r.t the number of variables, as discussed in \cite{}.
+
+Related Work:
+Labelled Dirac notation vs. graph representation.
+Therefore, in the circuit model, graph structures such as DAG or ZX calculus are often used to describe the evolution of two-qubit gates in multi-qubit systems.
+
+what is the labelled dirac notation?
+
+the semantics -- cylinder extension -- uniqueness
+
+\subsection{Syntax, Typing and Rewriting Rules}
+
+syntax -- typing -- rules
+
+\subsection{Normalization}
+
+normalization -- algorithm that eliminate labels
+
 Labelled Dirac notation uses register names to indicate the quantum system of vectors and operators.

### 2025-01-29 01intro_preliminary.tex
section 2 modified

@@
 An interesting property of quantum mechanics is that for two maximally entangled states, applying a quantum operator \( M \) to one subsystem is equivalent to applying \( M^T \) (the transpose of \( M \)) to the other subsystem.
-This relationship holds regardless of the spatial separation between the two systems, and it can be expressed as an equation in Dirac notation.
+This relationship holds regardless of the spatial separation between the two systems, and it can be expressed as an equation in labelled Dirac notation.
 \begin{example}
@@
     \[
-    M_q \ket{\Phi}_{q; r} = M_r^T \ket{\Phi}_{q; r}.
+    M_q \ket{\Phi}_{(q, r)} = M_r^T \ket{\Phi}_{(q, r)}.
     \]
@@
 
+A little explanation on the notation and terms.
+Quantum states are represented as vectors in complex Hilbert spaces, and operations on these states are described by linear transformations, or operators.
+Dirac notation uses the ket \( \ket{i} \) and the bra \( \bra{i} \) to denote basis vectors in a Hilbert space and its dual space, respectively.
+These symbols can be composed together in various ways to form more complex expressions.
+For example, $M \ket{i}$ represents the operator-vector multiplication, and the tensor product \(\ket{i}\otimes\ket{i}\) describes the product state on multiple quantum systems.
+
 In this equation, \( q \) and \( r \) are labels denoting the respective subsystems in the Dirac notation.
-To automate reasoning about such equations, we must formalize the language and develop a proof system to handle it.
+$M_q$ means that the operation $M$ is applied on system $q$, and $\ket{\Phi}_{(q, r)}$ denotes the quantum state $\ket{\Phi}$ on the (ordered) composed system $(q, r)$.
+To reason about their equivalence, we can remove the labels by extending them to the global system $(q, r)$:
+\begin{align}
+    (M \otimes I) \ket{\Phi} = (I \otimes M^T) \ket{\Phi}.
+    \label{eq: motivating unlabelled}
+\end{align}
+Here the identity operator $I$ are inserted in suitable positions to extend the operator $M$, a special case of extending to the global system called cylinder extension.
 
@@
 
-\subsubsection{Dirac Notation}
-
-Quantum states are represented as vectors in complex Hilbert spaces, and operations on these states are described by linear transformations.
-Dirac notation uses the ket \( \ket{i} \) and the bra \( \bra{i} \) to denote basis vectors in a Hilbert space and its dual space, respectively.
-These symbols can be composed together in various ways to form more complex expressions.
-The meaning of these compositions depends on the types of the operands involved.
-For example, the inner product \( \braket{i|j} \) represents the scalar result of the dot product between \( \bra{i} \) and \( \ket{j} \), while the outer product \( \ket{i}\bra{j} \) represents an operator.
-Additionally, Dirac notation uses the tensor product symbol \( \otimes \) to describe the combined space of multiple quantum systems.
-
-One of the key features of Dirac notation is its order-independent interpretation.
-This means that the composition of terms can be written without parentheses, as the interpretation is unaffected by the order of multiplication.
-For instance, the formula \( \bra{i}\ket{\phi}\bra{\psi}\ket{j} \) can be understood as:
-\[
-    \braket{i|\phi}\braket{\psi|j} = \bra{i} (\ket{\phi}\bra{\psi}) \ket{j},
-\]
-and these expressions are equivalent for all variables involved.
-
-In practice, Dirac notation is often combined with other syntactic elements, such as the summation symbol \( \sum_{i \in S} A \), to enhance expressiveness.
-Furthermore, labelled Dirac notation, such as \( \ket{i}_q \otimes \ket{i}_r \), is used to denote quantum systems in consideration, where subscripts (registers) are added to distinguish different subsystems.
-
 \subsubsection{Universal Algebra}

### 2025-01-29 01intro_preliminary.tex
rearranged

@@
 
-\subsubsection{Use Case}
-\yx{simplify to fit in the motivating section}
-Here we encode the motivating~\Cref{ex: motivating}, examine and explain how it is checked in D-Hammer. The encoding is shown below.
-
-    \begin{lstlisting}[style=dhammer]
-Var T : INDEX. Var M : OTYPE[T, T].
-Def phi := idx T => Sum nv in USET[T], |(nv, nv)>.
-Var r1 : REG[T]. Var r2 : REG[T].
-CheckEq M_r1;r1 (phi T)_(r1, r2) with (TPO T T M)_r2;r2 (phi T)_(r1, r2).
-    \end{lstlisting}
-
-The first three lines use the \texttt{\textcolor{NavyBlue}{Var}} and \texttt{\textcolor{NavyBlue}{Def}} commands to set up the context for the Dirac notation.
-\texttt{T} is a type index, representing arbitrary Hilbert space types.
-\texttt{M} is assume to be an operator in the Hilbert space with type \texttt{T}.
-\texttt{phi} is defined as the maximally entangled state, depending on the bound variable \texttt{T} as index.
-\texttt{r1} and \texttt{r2} are register names for the two subsystems.
-
-In the left hand side of \texttt{\textcolor{NavyBlue}{CheckEq}} command, \texttt{M\_r1;r1} denotes the labelled notation $M_{r_1; r_1}$, and \texttt{(phi T)\_(r1, r2)} denotes the entangled state $\ket{\Phi}_{(r_1, r_2)}$.
-They are connected by a white space, which is parsed into the composition of Dirac notation, and will be reduced into the operator-ket multiplication after typing.
-The right hand side is interpreted similarly, except the defined symbol \texttt{TPO} in the environment:
-
-\begin{lstlisting}[style=dhammer]
-Def TPO := idx sigma => idx tau => fun O : OTYPE[sigma, tau] => Sum i in USET[sigma], Sum j in USET[tau], (<i| O |j>).(|j> <i|).
-\end{lstlisting}
-
-The \texttt{TPO} symbol represents the transpose of operators, and encodes the formalization in~\Cref{ex: formalizing motivating}.
-Thanks to the design of environment and functions, many other commonly used symbols in Dirac notation are encoded and provided as defined symbols in D-Hammer.
-
-Within one second, the prover reports the result of equivalence with their common normal form:
-    \begin{lstlisting}[style=dhammer]
-The two terms are equal.
-[Normalized Term] SUM[USET[T], FUN[BASIS[T], SUM[USET[T], FUN[BASIS[T], SCR[DOT[BRA[$\texttt{\$1}$], MULK[M, KET[$\texttt{\$0}$]]], LTSR[LKET[$\texttt{\$1}$, r1], LKET[$\texttt{\$0}$, r2]]]]]]] : DTYPE[RSET[r1, r2], RSET]
-    \end{lstlisting}
-
-The normal form is in the internal syntax tree format mentioned above. A more readable interpretation is:
-\[
-\sum_{\mathbf{U}(T)} \sum_{\mathbf{U}(T)} \bra{\$1}M\ket{\$ 0} . \ket{\$1}_{r_1} \otimes \ket{\$0}_{r_2} : \DType(\{r_1, r_2\}, \emptyset).
-\]
-Here $\$0$ and $\$1$ are de Bruijn indices. The result is a ket on the $\{r_1, r_2\}$ system as expected, and follows pattern proposed in~\Cref{sec: labelled}.
-
 \subsubsection{Universal Algebra}

### 2025-01-29 main.tex
rearranged
.
2025/1/28

@@
 \newcommand{\yx}[1]{\textit{\color{blue}[YX] : #1}}
+\newcommand{\gb}[1]{\textit{\color{green}[GB] : #1}}
 \newcommand{\lz}[1]{\textit{\color{red}[LZ] : #1}}
@@
 \usepackage{cleveref}
+
+\usepackage{algorithm}
+\usepackage{algpseudocode}
 
@@
 \begin{abstract}
-Dirac notation is a fundamental language in quantum computation and quantum information.
-Recently, the term rewriting system DiracDec~\cite{diracdec} was introduced to automate equational reasoning with Dirac notation, a critical yet time-intensive task.
-Building upon DiracDec, this work aims to develop a solver optimized for practical applications.
-Enhancements include an improved typing system, a simplified language and rewriting system, more efficient algorithms, and added support for labeled Dirac notation.
-Our solver, D-Hammer, is implemented in \CC\ with a Mathematica backend, demonstrating significant improvements in decision-making power and time efficiency.
+Labelled Dirac notation is a formalism commonly used by physicists to
+represent many-body quantum systems and computer scientists to assert
+properties of quantum programs. Labelled Dirac notation is supported
+by a rich equational theory for proving equality between expressions
+of the language. These proofs are typically carried on pen-and-paper,
+and can be exceedingly long and error-prone. We introduce D-Hammer,
+the first tool to supported automated equational proof for labelled
+Dirac notation. The salient features of D-Hammer are: an expressive,
+higher-order, dependently-typed language for labelled Dirac notation;
+an efficient algorithm to normalize expressions (modulo associativity
+and commutativity); and an optimized \CC\ implementation. We evaluate
+the implementation on representative examples from plain and labelled
+Dirac notation. In the case of plain Dirac notation, we show that our
+implementation significantly outperforms DiracDec (Xu et al, POPL'25).
 
@@
 
-\input{02techniques.tex}
+\input{02language_design.tex}
 
-\input{03tool_eval.tex}
+\input{03labelled.tex}
 
-\input{04related_conclusion.tex}
+\input{04efficient_alg.tex}
+
+\input{05tool_eval.tex}
+
+\input{06related_conclusion.tex}
 

### 2025-01-29 06related_conclusion.tex
.

--- 2025-01-27 06related_conclusion.tex
@@
 \section{Related Work}
+Categorical quantum mechanics (CQM)~\cite{DBLP:conf/lics/AbramskyC04}
+is a category-theoretic foundation for quantum mechanics. CQM is the
+basis of the ZX calculus~\cite{DBLP:conf/icalp/CoeckeD08}, a popular
+graphical calculus for quantum states~\cite{vandewetering2020zx}. A
+key aspect of the ZX calculus and related graphical calculi is that
+they can be equipped with powerful rewriting systems. These systems
+support a form of graph rewriting, i.e.\, they exploit the graph
+representation of quantum states. There exists several completeness
+results for ZX-like rewriting systems, as well as some practical
+tools, such as PyZX~\cite{kissinger2019pyzx}. Similarly, there exist
+completeness results and tools for the sum-over-paths representation
+of quantum states~\cite{amy2018towards,amy2023complete}.
 
-Automated theorem proving has seen significant advancements in recent years, particularly through the development of satisfiability modulo theories (SMT) solvers,
-% SMT solvers extend propositional satisfiability by integrating theories such as arithmetic, arrays, and bit-vectors, making them suitable for solving a wide range of verification problems in both hardware and software domains.
-including prominent tools like Z3. These solvers have become essential in various fields such as formal verification, synthesis, and model checking. 
-Equational reasoning is another crucial area of research within automated theorem proving, which focuses on solving problems that involve equations between terms in an algebraic structure. 
-Equational provers, such as Vampire and E, have played a pivotal role in addressing the challenge of proving equations in first-order logic, employing sophisticated algorithms like superposition and term rewriting.
+\paragraph*{Comparison with DiracDec~\cite{diracdec}}
+Xu \emph{et al}~\cite{diracdec} define a language and an associate and
+commutative rewriting system for Dirac notation, and implement their
+language and rewriting system in Mathematica. Our approach follows a
+similar pattern. However, there are significant differences in terms
+of scope, expressiveness, and efficiency. We review these differences
+below.
 
-Formal verification of quantum computation is receiving increasing attentions during these years. See~\cite{Lewis2023} for a comprehensive review. 
-Verification frameworks in Coq include the foundational formalization CoqQ and quantum circuit language QWIRE. 
-Verifications of quantum programs are also considered, such as the Hoare logic based methods [] and model checking based methods [].
-The equational reasoning of Dirac notation is crucial for establishing property proofs in these works.
-Verification theoreis and tools based on other languages are also proposed, such as PyZX for reasoning and simplification of ZX-calculus.
+The most obvious difference is that D-Hammer targets labelled Dirac
+notation, whereas DiracDec targets plain Dirac notation. As already
+mentioned, the use of labelled Dirac notation is essential in many
+applications; in particular, labelled Dirac notation can simplify
+notation and proofs, and is in general better suited for writing and
+reasoning about complex, many-body, quantum states. However, there are
+several other important differences. First, our language leverages the
+power of higher-order dependent type theory. In particular, we
+leverage higher-order functions to provide a compact and expressive
+representation of big operators. Second, our language supports AC
+symbols with indefinite arities, which leads to compact representation
+of terms, and eases AC reasoning. Third, the relation with Mathematica
+is fundamentally different. DiracDec is implemented in Mathematica; as
+a consequence, its behavior, and in particular, its typing rules are
+constrained by the type system of Mathematica. In contrast, D-Hammer
+is designed as a separate tool; as a consequence, it benefits from an
+improved representation of terms (e.g.\, AX operators with indefinite
+arity), a more expressive type system (e.g.\, dependent types), and a
+more efficient rewriting engine. D-Hammer still relies on Mathematica
+to reason about functions that are not natively supported by rewriting
+rules, but these interactions are constrained and do not have negative
+effects on the overall efficiency of the system. This is reflected in
+our experimental comparison of D-Hammer and DiracDec, which shows how
+the former outperforms the latter.
+
+%% The first step is to formalize the language for Dirac notation.
+In DiracDec, the language has a concrete design, where the same syntax for different types corresponds to distinct symbols.
+Our goal is to transition from this concrete design to a more abstract one, which aligns more closely with the conventional Dirac notation we use and simplifies the term rewriting system.
+
+%% %% Automated theorem proving has seen significant advancements in recent years, particularly through the development of satisfiability modulo theories (SMT) solvers,
+%% %% % SMT solvers extend propositional satisfiability by integrating theories such as arithmetic, arrays, and bit-vectors, making them suitable for solving a wide range of verification problems in both hardware and software domains.
+%% %% including prominent tools like Z3. These solvers have become essential in various fields such as formal verification, synthesis, and model checking. 
+%% %% Equational reasoning is another crucial area of research within automated theorem proving, which focuses on solving problems that involve equations between terms in an algebraic structure. 
+%% %% Equational provers, such as Vampire and E, have played a pivotal role in addressing the challenge of proving equations in first-order logic, employing sophisticated algorithms like superposition and term rewriting.
+
+%% %% Formal verification of quantum computation is receiving increasing attentions during these years. See~\cite{Lewis2023} for a comprehensive review. 
+%% %% Verification frameworks in Coq include the foundational formalization CoqQ and quantum circuit language QWIRE. 
+%% %% Verifications of quantum programs are also considered, such as the Hoare logic based methods [] and model checking based methods [].
+%% %% The equational reasoning of Dirac notation is crucial for establishing property proofs in these works.
+%% %% Verification theoreis and tools based on other languages are also proposed, such as PyZX for reasoning and simplification of ZX-calculus.
 
 \section{Conclusion and Future Work}
-Based on the first equational reasoning tool for Dirac notation called DiracDec, this work improves and extends the theory for practical applications, and provides the solver D-Hammer.
+We have designed and implemented D-Hammer, a dependently typed
+higher-order language and proof system for labelled Dirac
+notation. D-Hammer benefits from an optimized implementation in
+\CC\ and a tight integration with Mathematica to reason about a broad
+range of mathematical functions, including trigonometric and
+exponential functions, that are commonly used in quantum physics.
+There are three important directions for future work. The first
+direction is to enhance D-Hammer with support for big tensor notation,
+i.e.\, to support expressions of the form $\bigotimes_{i\in I}
+e_i$. The second direction is to extend D-Hammer with a mechanism to
+generate independently verifiable certificates.  There is a large body
+of work on producing certificates for automated tools, in particular
+SMT solvers; see e.g.~\cite{DBLP:journals/cacm/BarbosaBCDKLNNOPRTZ23}
+for a recent overview. One potential option would be to integrate
+D-Hammer with the Coq or Lean proof assistants; in the first case, one
+would benefit from the formalization of labelled Dirac notation in
+CoqQ~\cite{Zhou2023}, whereas in the second case, one would benefit
+from powerful mechanisms to integrate rewriting procedures into the
+Lean proof assistant. Finally, the third direction is to connect
+D-Hammer with quantum program verifiers. Two potential applications
+are automating equational proofs for tools that already use Dirac
+notation, and to substitute numerical methods for tools that use
+matrices instead of symbolic assertions.
+
+%% Based on the first equational reasoning tool for Dirac notation called DiracDec, this work improves and extends the theory for practical applications, and provides the solver D-Hammer.
 Experiments show that the tool demonstrates advantages in decidability, efficiency and usability.
 
-We expect D-Hammer to have applications in areas like quantum program verification or proofs of post-quantum cryptography protocols in the future.
-One promising following up is to connect D-Hammer with theorem provers like Coq.
+%% We expect D-Hammer to have applications in areas like quantum program verification or proofs of post-quantum cryptography protocols in the future.
+%% One promising following up is to connect D-Hammer with theorem provers like Coq.
 It involves transforming theorem prover expressions into D-Hammer, and verify the proof trace of D-Hammer in theorem provers.

### 2025-01-29 01intro_preliminary.tex
.

@@
 about quantum programs. Specifically, many quantum Hoare logics use
-labelled Dirac notation to express program assertions, including
-pre-conditions and post-conditions~\cite{TODO}. These logics also use
+labelled Dirac notation to express program assertions~\cite{DBLP:conf/lics/ZhouBHYY21,Zhou2023}. These logics also use
 implicitly equational reasoning between labelled Dirac expressions, to
@@
 given in terms of Hilbert spaces, tailored to interpreting tensor
-product as an associative and commutative operator. We leverage this
-interpretation to define a rich equational theory for labelled Dirac
-notation, and prove its soundness with respect to our denotational
-semantics. Finally, we define an efficient normalization procedure to
-prove equivalence of two expressions. We then evaluate our procedure
-with respect to examples from the literature. Our evaluation covers
-examples based on (vanilla, unlabelled) Dirac notation, as well as
-examples based on labelled Dirac notation. The main conclusions are
-that our approach outperforms prior work by~\cite{diracdec} and is
-able to prove complex examples of reasoning with labelled Dirac
-notation from the literature, including examples from prior work on
-quantum separation logic~\cite{DBLP:conf/lics/ZhouBHYY21}.
+product as an AC symbol. We leverage this interpretation to define a
+rich equational theory for labelled Dirac notation, and prove its
+soundness with respect to our denotational semantics. Finally, we
+define an efficient normalization procedure to prove equivalence of
+two expressions. We then evaluate our procedure with respect to
+examples from the literature. Our evaluation covers examples based on
+(vanilla, unlabelled) Dirac notation, as well as examples based on
+labelled Dirac notation. The main conclusions are that our approach
+outperforms prior work by~\cite{diracdec} and is able to prove complex
+examples of reasoning with labelled Dirac notation from the
+literature, including examples from prior work on quantum separation
+logic~\cite{DBLP:conf/lics/ZhouBHYY21}.
 

### 2025-01-29 05tool_eval.tex
rearranged
update evaluation

--- 2025-01-27 03tool_eval.tex
@@
 
-\subsubsection{Benchmark performance}
+\subsection{Evaluation}
+We evaluate D-Hammer on several example sets, and make a comparison with the previous tool DiracDec.
+The experiments are carried out using a MacBook Pro with M3 Max chip. See~\Cref{fig: all examples} for a summary.
 
-To evaluate D-Hammer, we first test the examples from DiracDec benchmark and make a comparison.
-The experiments are carried out using a MacBook Pro with M3 Max chip.
-
-\begin{figure}
+\begin{figure}[h]
     \center
@@
         CoqQ                    & 162          & 156       &    48.69       &   158     &  158   &     9.74     \\
-        circuits                 & 2          & 2       &    17.67       &   3     &  2   &     1.4     \\
-        research paper                & 4          & 4         &  59.53       &   4    & 4       &  0.73     \\
+        others                  & 7          &  6         &   77.20    &    7        &   6     &  2.13     \\
+        % circuits                 & 2          & 2       &    17.67       &   3     &  2   &     1.4     \\
+        % research paper                & 4          & 4         &  59.53       &   4    & 4       &  0.73     \\
+        labelled Dirac notation      &   -         &   -          &     -           &      -      &        13       &     6.69    \\
         \hline
     \end{tabular}        
-    \caption{For DiracDec, examples that cannot be decided within 60 seconds are not included.}
+    \caption{Summary of all examples evaluated.}
+    \label{fig: all examples}
 \end{figure}
 
-The timing of D-Hammer does not include the initialization of Mathematica kernel link, which takes about 3 seconds in the beginning.
-As to expressibility, the language for DiracDec has the support for projectors $\texttt{fst}$ and $\texttt{snd}$ on basis pairs, satisfying $\texttt{fst} (s, t) = s$ and $\texttt{snd} (s, t) = t$.
-We found this feature is almost not used, so we removed the support.
-As a result, D-Hammer encodes 158 examples for the CoqQ part, 4 less than DiracDec.
+\subsubsection{Textbook (QCQI)}
+The 18 examples of QCQI textbook equations from DiracDec are all encoded and solved. The running time are short and close because these examples are simple and direct.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.8\textwidth]{fig/coqq.pdf}
+    \caption{The running time comparison between DiracDec and D-Hammer on the CoqQ benchmark.}
+    \label{fig: CoqQ plot}
+\end{figure}
+
+\subsubsection{CoqQ}
+As to expressibility, D-Hammer encodes 158 examples for the CoqQ part, 4 less than DiracDec.
+This is because the language for DiracDec has the support for projectors $\texttt{fst}$ and $\texttt{snd}$ on basis pairs, satisfying $\texttt{fst} (s, t) = s$ and $\texttt{snd} (s, t) = t$.
+We found this feature is almost not used and removed the support.
 For decidability, we improved the rewriting rules about sum for D-Hammer, and proved several more examples failed by DiracDec.
-The main difference is about their time efficiency.
-Because of our algorithm to decide AC-equivalence and SUM-SWAP, D-Hammer has a significant efficiency improvement, especially on those ``computational examples'' mentioned in the DiracDec paper.
-One typical example comes from the paper by Jens about the equivalence of operators for qubits.
-The system has to decompose the term on the concrete $\ket{0}$ and $\ket{1}$ basis, resulting a lot of addition elements.
+\Cref{fig: CoqQ plot} shows a direct comparison of their time efficiency.
+We can observe that for small examples, D-Hammer is slower because of the marginal overhead. 
+When the running time goes up, D-Hammer provides 2 to 40 times of acceleration in average.
+The relationship between efficiency improvement and example structure further supports the necessity of our algorithm. 
+We find that examples with great efficiency improvement, e.g. COQQ-129 and COQQ-148 shown in~\Cref{fig: CoqQ plot}, tend to use deeply nested summations, where our algorithm has an effect.
+
+\subsubsection{Others}
+Other examples are taken from quantum circuits and one recent paper [TODO].
+Because these examples involve decomposition on concrete $\ket{0}$ and $\ket{1}$ qubit basis, resulting a lot of addition elements, our algorithm dealing with AC symbols brings along significant improvement.
+One typical example is
+\begin{gather*}
+    (I \otimes P) \cdot U \cdot (I \otimes P) \cdot U^\dagger = \ket{0}\bra{0} \otimes I + \ket{1}\bra{1}\otimes (P \cdot P), \\
+    \textrm{where } P \triangleq e^{-i\theta/2} \ket{0}\bra{0} + e^{i\theta/2}\ket{1}\bra{1}, \qquad U \triangleq \ket{0}\bra{0} \otimes X + \ket{1}\bra{1} \otimes I.
+\end{gather*}
 It takes DiracDec about one minute, but D-Hammer solves it within one second.
 
-We also built an example benchmark for labelled Dirac notation, see~\Cref{sec: examples for labelled}.
+\subsubsection{Labelled Dirac Notation}
+We have a new set of examples for labelled Dirac notation, see~\Cref{sec: examples for labelled}.
+These examples use label to express multiplication and summation flexibly.
+D-Hammer extends these labelled notations to the global system using the algorithm in~\Cref{sec: labelled}, and decide their equivalence within reasonable time.
+In particular, D-Hammer solved a very sophisticated long example with 7 registers from [LICS...], which is challenging even for experts.

### 2025-01-29 06related_conclusion.tex

@@
 is designed as a separate tool; as a consequence, it benefits from an
-improved representation of terms (e.g.\, AX operators with indefinite
+improved representation of terms (e.g.\, AC operators with indefinite
 arity), a more expressive type system (e.g.\, dependent types), and a

### 2025-01-29 01intro_preliminary.tex
update preliminary

@@
 \section{Motivation and Preliminaries}
+\subsection{Standard Dirac Notation}
 
-An interesting property of quantum mechanics is that for two maximally entangled states, applying a quantum operator \( M \) to one subsystem is equivalent to applying \( M^T \) (the transpose of \( M \)) to the other subsystem.
+Dirac notation, also known as bra-ket notation, provides an intuitive and concise mathematical framework for describing quantum states and operations in quantum mechanics. 
+We write $\cH$ for a Hilbert space, i.e., a vector space equipped with an standard inner produce $\<\bu,\bv\> \in \bC$ for $\bu,\bv\in\cH$.
+Dirac notation consists of the following components that reflect the basic postulates of quantum mechanics:
+\begin{itemize}
+  \item Ket $|u\>$ is a column vector that denotes a quantum state $\bu$ in the state Hilbert space $\cH$.
+For example, the computational bases of qubit system are commonly written as $|0\> = \matrix{1 \\ 0}$ and $|1\> = \matrix{0 \\ 1}$.
+  \item Bra $\<u|$ is a row vector, the conjugate transpose of $|u\>$, that denotes the dual state of $|u\>$. 
+  %It is alternative to interpret as a linear mapping from $\cH$ to $\bC$ defined as $\<u| : |v\> \mapsto \<\bu,\bv\>$. It is the conjugate transpose of $|u\>$.
+  For exmaple, $\<0| = \matrix{1 & 0}$ and $\<1| = \matrix{0 & 1}$.
+  \item Inner produce $\<u|v\> \triangleq \<\bu,\bv\>$ which indicates the probability amplitude for $|u\>$ to collapse into $|v\>$. 
+  In convention, it is computed by matrix multiplication of two state, e.g., 
+  $\<0|1\> = \matrix{1 & 0} \matrix{0 \\ 1} = 0.$
+  \item Outer product $|u\>\<v| \triangleq |w\> \mapsto (\<v|w\>) |u\>$.
+Any linear map, such as unitary tranformation, measurement operators, etc, can be decomposed as the sum of outer products.
+It is also computed by matrix multiplication, e.g.,
+  $|1\>\<0| = \matrix{0 \\ 1} \matrix{1 & 0}  = \matrix{0 & 0 \\ 1 & 0}.$
+  \item Tensor product $|u\>\otimes|v\>$ (or simply $|u\>|v\>$ or $|uv\>$), $\<u|\otimes\<v|$ (or simply $\<u|\<v|$ or $\<uv|$) for describing the state, dual state and linear map of composite systems respectively.
+It is computed by the the Kronecker product of matrices, e.g., $\<0|\<1| = {\color{blue}\matrix{1 & 0}}\otimes {\color{red}\matrix{1 & 0}} = \matrix{({\color{blue}1}*{\color{red}1}) & ({\color{blue}1}*{\color{red}0}) & ({\color{blue}0}*{\color{red}1}) & ({\color{blue}0}*{\color{red}0})} = \matrix{1&0&0&0}$.
+\end{itemize}
+
+\subsection{Labelled Dirac Notation and Motivating Example}
+
+Unfortunately, the standard Dirac notation becomes unexpectedly tedious for describing multi-body systems. To see this, consider the following example:
+\begin{example}
+  \label{example1}
+  Let $p,q,r$ be three qubits and initially in the (unnormalized) $\ghz$ state $|\ghz\> \triangleq |000\>+|111\>$.
+Show that applying 3-qubit Toffoli gate ($\ccnot$) with control qubits $p,r$ and target qubit $q$ is equivalent to applying 2-qubit $\cnot$ gate with control qubit $r$ or $p$ and target qubit $q$.
+\end{example}
+We formalize the conclusion using standard Dirac notation as
+\begin{align*}
+  \ccnot|\ghz\> &= (I\otimes \cnot)|\ghz\> \\
+  &= (\swap\otimes I)(I\otimes \cnot)(\swap\otimes I)|\ghz\>
+\end{align*}
+following the steps: 1.
+Arrange qubits in the conventional order, here we choose $p,r,q$ to simplify the representation of $\ccnot$; 2.
+Lift the local operation $\cnot$ to the global system.
+For $\cnot$ acting on $r,q$ is easier, since $r,q$ consistent with the chosen order, and we only need to tensoring it by an identity operator $I$ on $p$, i.e., $(I\otimes \cnot)$.
+For $\cnot$ acting on $p,q$, realize that $p,q$ is not adjacent in the chosen order, we additionally need to use $\swap$ to temporarily exchange the qubits $p$ and $r$, i.e., globally, the $(\swap\otimes I)$ before and after $(I\otimes \cnot)$ (lifting of $\cnot$ on $r,q$).
+
+Roughly speaking, encoding in standard Dirac notation needs to tensoring identity operators and employing additional $\swap$ gates since generally the conventional order cannot guarantee 1.
+the order of all local operations is consistent with it, 2.
+the local operations involve only adjacent qubits in the conventional order.
+
+To address this, physicists introduce labels (or subscripts) to indicate the systems on which quantum states or operations are applied, thereby avoiding unnecessary lifting and swap gates.
+For example, rewriting the above example using labels:
+\begin{align*}
+  \ccnot_{prq}|\ghz\>_{pqr} = 
+  \cnot_{rq}|\ghz\>_{pqr}
+  = \cnot_{pq}|\ghz\>_{pqr}
+\end{align*}
+without determining the conventional order nor lifting using additional $I$ and $\swap$s. 
+The use of labels allows us to avoid deliberately maintaining the order of qubits.
+Tensor product become an AC symbol, so we can rearrange qubits as needed to perform calculations.
+For example, we can perform the calculation as follows:
+\begin{align*}
+  &\ccnot_{prq}|\ghz\>_{pqr}
+  = (\ccnot|000\> + \ccnot|111\>)_{prq}
+  = (|000\> + |110\>)_{prq}\\
+  &\cnot_{rq}|\ghz\>_{pqr}
+  = (\cnot|00\>)_{rq}|0\>_p + (\cnot|11\>)_{rq}|1\>_p
+  =(|000\> + |101\>)_{rqp}\\
+  &\cnot_{pq}|\ghz\>_{pqr}
+  = (\cnot|00\>)_{pq}|0\>_r + (\cnot|11\>)_{pq}|1\>_r
+  = (|000\> + |101\>)_{pqr}
+\end{align*}
+where the RHS of each line are equivalent.
+In addition, it can very freely describe local measurements, partial traces (the state or evolution of subsystems in multi-body systems), and partial inner products (the representation of partial traces in pure states), which are sufficient to handle the mathematical formulas of quantum mechanics in multi-body systems.
+
+Labelled Dirac notation is not only ubiquitous in the description of multi-body systems, but also plays an important role in program logic -- just as in classical program logic, variable names are used to construct logical formulas instead of describing them as global memory functions, in quantum program logic, variable names are used to label the subsystems on which quantum gates act, rather than lifting them to the global system -- the latter would cause exponential length of the formula w.r.t the number of variables, as discussed in \cite{}.
+
+We generalize Example \ref{example1} as the motivating example in the following:
+\begin{example}[Apply multiplexer to $\ghz$ state]
+  \label{ex: motivating}
+  Let $p,q,r$ be three quantum systems with the same Hilbert space \( \mathcal{H}_T \) (i.e., with computational bases $\{|i\>\}_{i\in T}$) and GHZ state $\ghz = \sum_i|iii\>\<iii|$.
+Consider the linear operator sets $U_{ij}$ and the corresponding multiplexers $M\triangleq \sum_{ij}|ij\>\<ij|\otimes U_{ij}$ and $N\triangleq \sum_i|i\>\<i|\otimes U_{ii}$.
+Show that:
+  $$M_{prq}\ghz_{prq} = N_{rq}|\ghz\>_{pqr} = N_{pq}|\ghz\>_{pqr}.$$
+\end{example}
+
+% An interesting property of quantum mechanics is that for two maximally entangled states, applying a quantum operator \( M \) to one subsystem is equivalent to applying \( M^T \) (the transpose of \( M \)) to the other subsystem.
 This relationship holds regardless of the spatial separation between the two systems, and it can be expressed as an equation in labelled Dirac notation.
-\begin{example}
-    \label{ex: motivating}
-    Let \( q \) and \( r \) represent two quantum systems in the Hilbert space \( \mathcal{H}_T \).
+% \begin{example}
+%     \label{ex: motivating}
+%     Let \( q \) and \( r \) represent two quantum systems in the Hilbert space \( \mathcal{H}_T \).
 Let \( M \) be a quantum operation acting on \( \mathcal{H}_T \), and let \( \ket{\Phi} = \sum_{i \in T} \ket{i} \otimes \ket{i} \) be the maximally entangled state.
 Then, we have the following equation:
-    \[
-    M_q \ket{\Phi}_{(q, r)} = M_r^T \ket{\Phi}_{(q, r)}.
-    \]
-\end{example}
+%     \[
+%     M_q \ket{\Phi}_{(q, r)} = M_r^T \ket{\Phi}_{(q, r)}.
+%     \]
+% \end{example}
 
-A little explanation on the notation and terms.
-Quantum states are represented as vectors in complex Hilbert spaces, and operations on these states are described by linear transformations, or operators.
+% A little explanation on the notation and terms.
+% Quantum states are represented as vectors in complex Hilbert spaces, and operations on these states are described by linear transformations, or operators.
 Dirac notation uses the ket \( \ket{i} \) and the bra \( \bra{i} \) to denote basis vectors in a Hilbert space and its dual space, respectively.
 These symbols can be composed together in various ways to form more complex expressions.
-For example, $M \ket{i}$ represents the operator-vector multiplication, and the tensor product \(\ket{i}\otimes\ket{i}\) describes the product state on multiple quantum systems.
+% For example, $M \ket{i}$ represents the operator-vector multiplication, and the tensor product \(\ket{i}\otimes\ket{i}\) describes the product state on multiple quantum systems.
 
-In this equation, \( q \) and \( r \) are labels denoting the respective subsystems in the Dirac notation.
+% In this equation, \( q \) and \( r \) are labels denoting the respective subsystems in the Dirac notation.
 $M_q$ means that the operation $M$ is applied on system $q$, and $\ket{\Phi}_{(q, r)}$ denotes the quantum state $\ket{\Phi}$ on the (ordered) composed system $(q, r)$.
-To reason about their equivalence, we can remove the labels by extending them to the global system $(q, r)$:
-\begin{align}
-    (M \otimes I) \ket{\Phi} = (I \otimes M^T) \ket{\Phi}.
-    \label{eq: motivating unlabelled}
-\end{align}
-Here the identity operator $I$ are inserted in suitable positions to extend the operator $M$, a special case of extending to the global system called cylinder extension.
+% To reason about their equivalence, we can remove the labels by extending them to the global system $(q, r)$:
+% \begin{align}
+%     (M \otimes I) \ket{\Phi} = (I \otimes M^T) \ket{\Phi}.
+%     \label{eq: motivating unlabelled}
+% \end{align}
+% Here the identity operator $I$ are inserted in suitable positions to extend the operator $M$, a special case of extending to the global system called cylinder extension.
 

### 2025-01-29 01intro_preliminary.tex
.

@@
 Dirac notation, also known as bra-ket notation, provides an intuitive and concise mathematical framework for describing quantum states and operations in quantum mechanics. 
-We write $\cH$ for a Hilbert space, i.e., a vector space equipped with an standard inner produce $\<\bu,\bv\> \in \bC$ for $\bu,\bv\in\cH$.
+We write $\cH$ for a Hilbert space, i.e., a vector space equipped with an standard inner produce $\langle\bu,\bv\rangle \in \bC$ for $\bu,\bv\in\cH$.
 Dirac notation consists of the following components that reflect the basic postulates of quantum mechanics:
 \begin{itemize}
-  \item Ket $|u\>$ is a column vector that denotes a quantum state $\bu$ in the state Hilbert space $\cH$.
-For example, the computational bases of qubit system are commonly written as $|0\> = \matrix{1 \\ 0}$ and $|1\> = \matrix{0 \\ 1}$.
-  \item Bra $\<u|$ is a row vector, the conjugate transpose of $|u\>$, that denotes the dual state of $|u\>$. 
-  %It is alternative to interpret as a linear mapping from $\cH$ to $\bC$ defined as $\<u| : |v\> \mapsto \<\bu,\bv\>$. It is the conjugate transpose of $|u\>$.
-  For exmaple, $\<0| = \matrix{1 & 0}$ and $\<1| = \matrix{0 & 1}$.
-  \item Inner produce $\<u|v\> \triangleq \<\bu,\bv\>$ which indicates the probability amplitude for $|u\>$ to collapse into $|v\>$. 
+  \item Ket $|u\rangle$ is a column vector that denotes a quantum state $\bu$ in the state Hilbert space $\cH$.
+For example, the computational bases of qubit system are commonly written as $|0\rangle = \matrix{1 \\ 0}$ and $|1\rangle = \matrix{0 \\ 1}$.
+  \item Bra $\langle u|$ is a row vector, the conjugate transpose of $|u\rangle$, that denotes the dual state of $|u\rangle$. 
+  %It is alternative to interpret as a linear mapping from $\cH$ to $\bC$ defined as $\langle u| : |v\rangle \mapsto \langle \bu,\bv\rangle$. It is the conjugate transpose of $|u\rangle$.
+  For exmaple, $\langle 0| = \matrix{1 & 0}$ and $\langle 1| = \matrix{0 & 1}$.
+  \item Inner produce $\langle u|v\rangle \triangleq \langle \bu,\bv\rangle$ which indicates the probability amplitude for $|u\rangle$ to collapse into $|v\rangle$. 
   In convention, it is computed by matrix multiplication of two state, e.g., 
-  $\<0|1\> = \matrix{1 & 0} \matrix{0 \\ 1} = 0.$
-  \item Outer product $|u\>\<v| \triangleq |w\> \mapsto (\<v|w\>) |u\>$.
+  $\langle 0|1\rangle = \matrix{1 & 0} \matrix{0 \\ 1} = 0.$
+  \item Outer product $|u\rangle\langle v| \triangleq |w\rangle \mapsto (\langle v|w\rangle) |u\rangle$.
 Any linear map, such as unitary tranformation, measurement operators, etc, can be decomposed as the sum of outer products.
 It is also computed by matrix multiplication, e.g.,
-  $|1\>\<0| = \matrix{0 \\ 1} \matrix{1 & 0}  = \matrix{0 & 0 \\ 1 & 0}.$
-  \item Tensor product $|u\>\otimes|v\>$ (or simply $|u\>|v\>$ or $|uv\>$), $\<u|\otimes\<v|$ (or simply $\<u|\<v|$ or $\<uv|$) for describing the state, dual state and linear map of composite systems respectively.
-It is computed by the the Kronecker product of matrices, e.g., $\<0|\<1| = {\color{blue}\matrix{1 & 0}}\otimes {\color{red}\matrix{1 & 0}} = \matrix{({\color{blue}1}*{\color{red}1}) & ({\color{blue}1}*{\color{red}0}) & ({\color{blue}0}*{\color{red}1}) & ({\color{blue}0}*{\color{red}0})} = \matrix{1&0&0&0}$.
+  $|1\rangle\langle 0| = \matrix{0 \\ 1} \matrix{1 & 0}  = \matrix{0 & 0 \\ 1 & 0}.$
+  \item Tensor product $|u\rangle\otimes|v\rangle$ (or simply $|u\rangle|v\rangle$ or $|uv\rangle$), $\langle u|\otimes\langle v|$ (or simply $\langle u|\langle v|$ or $\langle uv|$) for describing the state, dual state and linear map of composite systems respectively.
+It is computed by the the Kronecker product of matrices, e.g., $\langle 0|\langle 1| = {\color{blue}\matrix{1 & 0}}\otimes {\color{red}\matrix{1 & 0}} = \matrix{({\color{blue}1}*{\color{red}1}) & ({\color{blue}1}*{\color{red}0}) & ({\color{blue}0}*{\color{red}1}) & ({\color{blue}0}*{\color{red}0})} = \matrix{1&0&0&0}$.
 \end{itemize}
@@
   \label{example1}
-  Let $p,q,r$ be three qubits and initially in the (unnormalized) $\ghz$ state $|\ghz\> \triangleq |000\>+|111\>$.
+  Let $p,q,r$ be three qubits and initially in the (unnormalized) $\ghz$ state $|\ghz\rangle \triangleq |000\rangle+|111\rangle$.
 Show that applying 3-qubit Toffoli gate ($\ccnot$) with control qubits $p,r$ and target qubit $q$ is equivalent to applying 2-qubit $\cnot$ gate with control qubit $r$ or $p$ and target qubit $q$.
@@
 \begin{align*}
-  \ccnot|\ghz\> &= (I\otimes \cnot)|\ghz\> \\
-  &= (\swap\otimes I)(I\otimes \cnot)(\swap\otimes I)|\ghz\>
+  \ccnot|\ghz\rangle &= (I\otimes \cnot)|\ghz\rangle \\
+  &= (\swap\otimes I)(I\otimes \cnot)(\swap\otimes I)|\ghz\rangle
 \end{align*}
@@
 \begin{align*}
-  \ccnot_{prq}|\ghz\>_{pqr} = 
-  \cnot_{rq}|\ghz\>_{pqr}
-  = \cnot_{pq}|\ghz\>_{pqr}
+  \ccnot_{prq}|\ghz\rangle_{pqr} = 
+  \cnot_{rq}|\ghz\rangle_{pqr}
+  = \cnot_{pq}|\ghz\rangle_{pqr}
 \end{align*}
@@
 \begin{align*}
-  &\ccnot_{prq}|\ghz\>_{pqr}
-  = (\ccnot|000\> + \ccnot|111\>)_{prq}
-  = (|000\> + |110\>)_{prq}\\
-  &\cnot_{rq}|\ghz\>_{pqr}
-  = (\cnot|00\>)_{rq}|0\>_p + (\cnot|11\>)_{rq}|1\>_p
-  =(|000\> + |101\>)_{rqp}\\
-  &\cnot_{pq}|\ghz\>_{pqr}
-  = (\cnot|00\>)_{pq}|0\>_r + (\cnot|11\>)_{pq}|1\>_r
-  = (|000\> + |101\>)_{pqr}
+  &\ccnot_{prq}|\ghz\rangle_{pqr}
+  = (\ccnot|000\rangle + \ccnot|111\rangle)_{prq}
+  = (|000\rangle + |110\rangle)_{prq}\\
+  &\cnot_{rq}|\ghz\rangle_{pqr}
+  = (\cnot|00\rangle)_{rq}|0\rangle_p + (\cnot|11\rangle)_{rq}|1\rangle_p
+  =(|000\rangle + |101\rangle)_{rqp}\\
+  &\cnot_{pq}|\ghz\rangle_{pqr}
+  = (\cnot|00\rangle)_{pq}|0\rangle_r + (\cnot|11\rangle)_{pq}|1\rangle_r
+  = (|000\rangle + |101\rangle)_{pqr}
 \end{align*}
@@
   \label{ex: motivating}
-  Let $p,q,r$ be three quantum systems with the same Hilbert space \( \mathcal{H}_T \) (i.e., with computational bases $\{|i\>\}_{i\in T}$) and GHZ state $\ghz = \sum_i|iii\>\<iii|$.
-Consider the linear operator sets $U_{ij}$ and the corresponding multiplexers $M\triangleq \sum_{ij}|ij\>\<ij|\otimes U_{ij}$ and $N\triangleq \sum_i|i\>\<i|\otimes U_{ii}$.
+  Let $p,q,r$ be three quantum systems with the same Hilbert space \( \mathcal{H}_T \) (i.e., with computational bases $\{|i\rangle\}_{i\in T}$) and GHZ state $\ghz = \sum_i|iii\rangle\langle iii|$.
+Consider the linear operator sets $U_{ij}$ and the corresponding multiplexers $M\triangleq \sum_{ij}|ij\rangle\langle ij|\otimes U_{ij}$ and $N\triangleq \sum_i|i\rangle\langle i|\otimes U_{ii}$.
 Show that:
-  $$M_{prq}\ghz_{prq} = N_{rq}|\ghz\>_{pqr} = N_{pq}|\ghz\>_{pqr}.$$
+  $$M_{prq}\ghz_{prq} = N_{rq}|\ghz\rangle_{pqr} = N_{pq}|\ghz\rangle_{pqr}.$$
 \end{example}

### 2025-01-29 01intro_preliminary.tex
update

--- 2025-01-29 01intro_preliminary.tex
@@
 The use of labels allows us to avoid deliberately maintaining the order of qubits.
-Tensor product become an AC symbol, so we can rearrange qubits as needed to perform calculations.
+Tensor product become an associative and commutative (AC) symbol, so we can rearrange qubits as needed to perform calculations.
 For example, we can perform the calculation as follows:
@@
 
-Labelled Dirac notation is not only ubiquitous in the description of multi-body systems, but also plays an important role in program logic -- just as in classical program logic, variable names are used to construct logical formulas instead of describing them as global memory functions, in quantum program logic, variable names are used to label the subsystems on which quantum gates act, rather than lifting them to the global system -- the latter would cause exponential length of the formula w.r.t the number of variables, as discussed in \cite{}.
+Labelled Dirac notation is not only ubiquitous in the description of multi-body systems, but also plays an important role in program logic -- just as in classical program logic, variable names are used to construct logical formulas instead of describing them as global memory functions, in quantum program logic, variable names are used to label the subsystems on which quantum gates act, rather than lifting them to the global system -- the latter would cause exponential length of the formula w.r.t the number of variables, as discussed in [cite].
 
@@
   \label{ex: motivating}
-  Let $p,q,r$ be three quantum systems with the same Hilbert space \( \mathcal{H}_T \) (i.e., with computational bases $\{|i\>\}_{i\in T}$) and GHZ state $\ghz = \sum_i|iii\>\<iii|$.
+  Let $p,q,r$ be three quantum systems with the same Hilbert space \( \mathcal{H}_T \) (i.e., with computational bases $\{|i\>\}_{i\in T}$) and GHZ state $|\ghz\> = \sum_i|iii\>\<iii|$.
 Consider the linear operator sets $U_{ij}$ and the corresponding multiplexers $M\triangleq \sum_{ij}|ij\>\<ij|\otimes U_{ij}$ and $N\triangleq \sum_i|i\>\<i|\otimes U_{ii}$.
 Show that:
-  $$M_{prq}\ghz_{prq} = N_{rq}|\ghz\>_{pqr} = N_{pq}|\ghz\>_{pqr}.$$
+  $$M_{prq}|\ghz\>_{prq} = N_{rq}|\ghz\>_{pqr} = N_{pq}|\ghz\>_{pqr}.$$
 \end{example}
@@
 
-\subsubsection{Universal Algebra}
-We use universal algebra and equational logic to formally represent Dirac notation and the reasoning procedure.
+% \subsubsection{Universal Algebra}
+% We use universal algebra and equational logic to formally represent Dirac notation and the reasoning procedure.
 A universal algebra defines a signature of function symbols, with terms constructed from constants, variables, and function applications.
-Other basic concepts like substitution of variables or pattern matching are also defined.
-In our case of Dirac notation, the signature consists of constructors and operations like $\ket{i}$ or $A \otimes B$.
-The reasoning process is guided by equational logic, which defines an equivalence relation that is compatible with substitution and term construction.
+% Other basic concepts like substitution of variables or pattern matching are also defined.
+% In our case of Dirac notation, the signature consists of constructors and operations like $\ket{i}$ or $A \otimes B$.
+% The reasoning process is guided by equational logic, which defines an equivalence relation that is compatible with substitution and term construction.
 This relation formalizes the intuitive concept of equivalence in algebra.

### 2025-01-29 02language_design.tex
rearranged
update

@@
     \item It has functions and higher-order typing. It allows to formally encode defined symbols like transpose or trace, which are usually used to reprensent the term in an abstract manner. 
-    \item We use an abstract syntax for better and simpler representations.
-It follows the conventional Dirac notation, where we extensively use the composition symbol to denote calculations, and symbols like scalar multiplications for different types are unified as one.
     \item The structures are friendly to optimizations for efficiency. For example, associative and commutative (AC) symbols have indefinite arities, which enables normalization by sorting.
@@
     \begin{align*}
-        e ::=\ & x \mid \lambda x : T.e \mid \mu x.e \mid e_1\ e_2 \mid (e_1, e_2) \\
+        e ::=\ & x \mid \lambda x : T.e \mid \lambda x : \Index.e \mid e_1\ e_2 \mid (e_1, e_2) \\
         & |\ 0 \mid 1 \mid e_1 \times \cdots \times e_n \mid e^* \mid \delta_{e_1, e_2} \\
@@
 \begin{enumerate}
-    \item \textbf{function and basis}: \( \lambda x : T.e \) represents the abstraction for normal functions, and \( \mu x.e \) represents the abstraction for index functions.
+    \item \textbf{function and basis}: \( \lambda x : T.e \) represents the abstraction for normal functions, and \( \lambda x : \Index.e \) represents the abstraction for index functions.
     \( e_1\ e_2 \) denotes function application.
@@
 \subsection{Typing System}
-
 The typing system is responsible for classifying terms within a proof system, according to the types of variables and definitions. 
-We use an environment \( E \) and context \( \Gamma \) for typing.
-\( E \) includes definitions and assumptions, and the context \( \Gamma \) tracks bound variables during the proof.
-Both \( E \) and \( \Gamma \) consist of sequences of assumptions \( x : T \) or definitions \( x := t : T \).
-\begin{definition}[Environment and Context]
-    The syntax for \( E \) and \( \Gamma \) is:
+We use a context \( E \) to preserve the assumptions \( x : T \) and definitions \( x := t : T \).
+\begin{definition}[Context]
+    The syntax for context \( E \) is:
     \begin{align*}
-        E ::= &\ [] \mid E; x : \Index \mid E; x : T \mid E; x := t : T. \\
-        \Gamma ::= &\ [] \mid \Gamma; x : \Index \mid \Gamma; x : T.
+        E ::= &\ [] \mid E; x : \Index \mid E; x : T \mid E; x := t : T.
     \end{align*}
 \end{definition}
-Definitions refer to symbols that can be expanded or unfolded, and typically represent abstract concepts, such as transpose or trace in Dirac notation.
-Assumptions, on the other hand, define the types of variables, and variable assumptions are implicitly universally quantified in the case of equational proofs.
-
-We say an expression \( t \) has type \( X \) in context \( E[\Gamma] \) if the typing judgment \( E[\Gamma] \vdash t : X \) can be proven through the rules in~\Cref{sec: full typing rules}. 
-It involves maintaining a well-formed environment and context \( E[\Gamma] \), which is built incrementally:
-\[
-    \frac{}{\WF([])[]}
+Definitions refer to symbols that can be expanded or unfolded, and typically represent abstract concepts.
+such as transpose or trace in Dirac notation.
+Assumptions, on the other hand, define the types of variables.
+We say an expression \( t \) has type \( X \) in context \( E \) if the typing judgment \( E \vdash t : X \) can be proven through the rules in~\Cref{sec: full typing rules}. These are two instances:
+\begin{gather*}
+    \frac{E \vdash t : \Basis(\sigma)}{E \vdash \ket{t} : \KType(\sigma)},
     \qquad
-    \frac{\WF(E)[] \qquad x \notin E}{\WF(E; x : \Index)[]}
-    \qquad
-    \frac{E[]\vdash t:T \qquad x \notin E}{\WF(E; x := t:T)[]}.
-\]
-Starting with an empty context, new index symbols can be introduced, and symbols with checked types can be defined. 
-Typing judgments are then made inductively using the information from \( E[\Gamma] \).
-For instance, the rule \( x : \Index \in E[\Gamma] \) holds if \( E \) or \( \Gamma \) contains an assumption for \( x \).
-And \(\KType(\sigma)\) is a valid type for kets, if the argument $\sigma$ is typed as an index.
-\begin{gather*}
-    \frac{\WF(E)[\Gamma] \qquad x : \Index \in E[\Gamma]}{E[\Gamma] \vdash x : \Index}
-    \qquad
-    \frac{E[\Gamma] \vdash \sigma : \Index}{E[\Gamma] \vdash \KType(\sigma) : \Type}
+    \frac{E \vdash B : \BType(\sigma) \qquad E \vdash K : \KType(\sigma)}{E \vdash B \cdot K : \SType}.
 \end{gather*}
-
-Terms are then typed accordingly.
-For example, the ket \( \ket{t} \) will have the type \( \KType(\sigma) \) if \( t \) is a basis term of index \( \sigma \).
+The ket \( \ket{t} \) will have the type \( \KType(\sigma) \) if \( t \) is a basis term of index \( \sigma \).
 Similarly, the inner product between a bra and a ket of the same index \( \sigma \) is typed as a scalar.
 It corresponds to the constraint of inner product that vectors should be from the same Hilbert space.
-\begin{gather*}
-    \frac{E[\Gamma]\vdash t : \Basis(\sigma)}{E[\Gamma] \vdash \ket{t} : \KType(\sigma)}
-    \qquad
-    \frac{E[\Gamma]\vdash B : \BType(\sigma) \qquad E[\Gamma]\vdash K : \KType(\sigma)}{E[\Gamma] \vdash B \cdot K : \SType}
-\end{gather*}
-
-The typing for functions and applications follows the standard practice.
-For example, the index function \( \mu x.
-t \) is typed with \( x \) as a bound variable of type \( \Index \), and the application \( (t\ u) \) has the type \( U\{x/u\} \), obtained by replacing \( x \) with the index instance \( u \).
-\begin{gather*}
-    \frac{E[\Gamma; x : T] \vdash t : U}{E[\Gamma] \vdash (\lambda x : T . t) : T \to U}
-    \qquad
-    \frac{E[\Gamma] \vdash t:U \to T \qquad E[\Gamma] \vdash u:U}{E[\Gamma] \vdash (t\ u):T} 
-    \\[1em]
-    \frac{E[\Gamma; x : \Index] \vdash t : U}{E[\Gamma] \vdash (\mu x.t) : \forall x.U}
-    \qquad
-    \frac{E[\Gamma] \vdash t:\forall x.U \qquad E[\Gamma] \vdash u : \Index}{E[\Gamma] \vdash (t\ u):U\{x/u\}}
-\end{gather*}
-
-Finally, the big operator sum is modeled by folding a function over a set, with the typing rule as follows:
+Especially, the big operator sum is modeled by folding a function over a set, with the typing rule as follows:
 \[
-    \frac{E[\Gamma] \vdash s : \SET(\sigma) \qquad E[\Gamma] \vdash f : \Basis(\sigma) \to \KType(\tau)}{E[\Gamma] \vdash \sum_{s} f : \KType(\tau)}.
+    \frac{E \vdash s : \SET(\sigma) \qquad E \vdash f : \Basis(\sigma) \to \KType(\tau)}{E \vdash \sum_{s} f : \KType(\tau)}.
 \]
 
-\begin{lemma}
-    The typing of expressions is both decidable and unique.
-\end{lemma}
+% \begin{lemma}
+%     The typing of expressions is both decidable and unique.
+% \end{lemma}
 
-\begin{proof}
-    The type of an expression can be determined recursively. For any given function symbol and argument types, there is at most one typing rule, ensuring the uniqueness of typing.
-\end{proof}
+% \begin{proof}
+%     The type of an expression can be determined recursively. For any given function symbol and argument types, there is at most one typing rule, ensuring the uniqueness of typing.
+% \end{proof}
 
@@
 The idea behind the interpretation of types and terms is to formalize the typing relation using set-theoretic inclusion.
-Specifically, for a well-formed context \( E[\Gamma] \), term \( t \), and type \( T \), if \( E[\Gamma] \vdash t : T \), then for any valuation \( v \), the semantics of \( t \) must belong to the semantics of \( T \).
+Specifically, for a well-formed context \( E \), term \( t \), and type \( T \), if \( E \vdash t : T \), then for any valuation \( v \), the semantics of \( t \) must belong to the semantics of \( T \).
 
 \begin{lemma}
-    For any well-formed context \( E[\Gamma] \), term \( t \), and type \( T \), if \( E[\Gamma] \vdash t : T \), then for all valuations \( v \), \( \sem{t}_v \in \sem{T}_v \).
+    For any well-formed context \( E \), term \( t \), and type \( T \), if \( E \vdash t : T \), then for all valuations \( v \), \( \sem{t}_v \in \sem{T}_v \).
 \end{lemma}
@@
 \begin{lemma}
-    For all well-formed contexts \( E[\Gamma] \) and terms \( e_1, e_2 \), if \( \vdash e_1 = e_2 \), then \( \sem{e_1} = \sem{e_2} \).
+    For all well-formed contexts \( E \) and terms \( e_1, e_2 \), if \( \vdash e_1 = e_2 \), then \( \sem{e_1} = \sem{e_2} \).
 \end{lemma}
@@
 
-With these syntax, typing and semantics introduced, we demonstrate the formalization of the symbols used in the motivating example to conclude this section.
-\begin{example}[Formalizing the Motivating Example]
-    \label{ex: formalizing motivating}
-    Definitions and assumptions in the environment \( E \) are formalized as follows:
-    \begin{align*}
-        & \text{TPO} && := \mu T_1. \mu T_2. \lambda O : \OType(T_1, T_2). \sum_{i \in \mathbf{U}(T_1)} \sum_{j \in \mathbf{U}(T_2)} \bra{i} O \ket{j} . \ket{j}\bra{i} \\
-        & &&\quad : \forall T_1. \forall T_2. \OType(T_1, T_2) \to \OType(T_2, T_1); \\
-        &\text{phi} &&:= \mu T. \sum_{i \in \mathbf{U}(T)} \sum_{j \in \mathbf{U}(T)} \ket{(i, j)} : \forall T.\KType(T \times T); \\
-        & T && : \Index; \\
-        & M && : \OType(T, T).
-    \end{align*}
-    The symbol \( \text{TPO} \) represents the transpose of an operator, polymorphic on the Hilbert spaces \( T_1 \) and \( T_2 \).
+% With these syntax, typing and semantics introduced, we demonstrate the formalization of the symbols used in the motivating example to conclude this section.
+% \begin{example}[Formalizing the Motivating Example]
+%     \label{ex: formalizing motivating}
+%     Definitions and assumptions in the environment \( E \) are formalized as follows:
+%     \begin{align*}
+%         & \text{TPO} && := \lambda T_1 : \Index. \lambda T_2 : \Index. \lambda O : \OType(T_1, T_2). \sum_{i \in \mathbf{U}(T_1)} \sum_{j \in \mathbf{U}(T_2)} \bra{i} O \ket{j} . \ket{j}\bra{i} \\
+%         & &&\quad : \forall T_1. \forall T_2. \OType(T_1, T_2) \to \OType(T_2, T_1); \\
+%         &\text{phi} &&:= \lambda T : \Index. \sum_{i \in \mathbf{U}(T)} \sum_{j \in \mathbf{U}(T)} \ket{(i, j)} : \forall T.\KType(T \times T); \\
+%         & T && : \Index; \\
+%         & M && : \OType(T, T).
+%     \end{align*}
+%     The symbol \( \text{TPO} \) represents the transpose of an operator, polymorphic on the Hilbert spaces \( T_1 \) and \( T_2 \).
 The symbol \( \text{phi} \) takes the index \( T \) and defines the maximally entangled states, summing over all basis elements in \( T \), as indicated by the universal set \( \mathbf{U}(T) \).
-    With the assumption of the index \( T \) and operator \( M \), we can express the equivalence in the non-labelled version as:
-    \[
-    (\textrm{M} \otimes \mathbf{1}_\mathcal{O}(\textrm{T})) \cdot (\textrm{phi T}) = (\mathbf{1}_\mathcal{O}(\textrm{T}) \otimes (\textrm{TPO T T M})) \cdot (\textrm{phi T}).
-    \]
-\end{example}
+%     With the assumption of the index \( T \) and operator \( M \), we can express the equivalence in the non-labelled version as:
+%     \[
+%     (\textrm{M} \otimes \mathbf{1}_\mathcal{O}(\textrm{T})) \cdot (\textrm{phi T}) = (\mathbf{1}_\mathcal{O}(\textrm{T}) \otimes (\textrm{TPO T T M})) \cdot (\textrm{phi T}).
+%     \]
+% \end{example}

### 2025-01-29 04efficient_alg.tex
update

@@
                 \State \Return $\lambda x : T . \textsc{Sort}(e')$
-            \ElsIf{$e \equiv \mu x. e'$}
-                \State \Return $\mu x . \textsc{Sort}(e')$
+            \ElsIf{$e \equiv \lambda x : \Index. e'$}
+                \State \Return $\lambda x : \Index . \textsc{Sort}(e')$
             \ElsIf{$e \equiv f(a_1, \cdots, a_n)$}
@@
 \begin{definition}[Swap Transformation]
-For a term \( e \) with a sorting result \( \textsc{Sort}(e) \), the swap transformation proceeds by ordering all bound variables according to their first appearances, except in function definitions \( \lambda x \) and \( \mu x \).
+For a term \( e \) with a sorting result \( \textsc{Sort}(e) \), the swap transformation proceeds by ordering all bound variables according to their first appearances, except in function definitions \( \lambda x \).
 The swap transformation then reorders the successive summations accordingly.
@@
 \begin{theorem}[Soundness]
-    For any well-formed context \( E[\Gamma] \) and well-typed expressions \( e_1 \) and \( e_2 \), if  $e_1$ and $e_2$ have the same normal form, then \( \sem{e_1} = \sem{e_2} \).
+    For any well-formed context \( E \) and well-typed expressions \( e_1 \) and \( e_2 \), if  $e_1$ and $e_2$ have the same normal form, then \( \sem{e_1} = \sem{e_2} \).
 \end{theorem}

### 2025-01-29 05tool_eval.tex
update

@@
 \section{Implementation and Case Study}
-
-The main purpose of this work is to build a practical tool that works well in checking Dirac notation equations.
 The refinements and extensions above concludes in our implementation called D-Hammer, a solver written in \CC.
@@
 The user can use commands to make definitions and assumptions in the maintained context, conduct the normalization and equivalence checking, and obtain the rewriting trace output.
-This implementation
-is tested on the benchmark of the DiracDec work, and succeeds in proving most of them efficiently. 
 It can be used from the command line interactively, or can be integrated into other \CC\ projects as a library.
+This implementation is tested on the benchmark of the previous work and new examples for labelled Dirac notation, demonstrating significant improvement in expressivity and efficiency.
 
@@
 The syntax tree structure is also compatible with the datatype of Mathematica. This improves the interoperability between D-Hammer and the Mathematica system, enabling them to work interleavingly.
-To improve usability, D-Hammer also supports many special syntacies for terms,  and most Dirac notation terms will be encoded in the natural way.
+To improve usability, D-Hammer also supports many special notations for terms,  and most Dirac notation terms will be encoded in the natural way.
 Here are some examples for the parsing syntax.
@@
 Finally, D-Hammer uses a prover to host the computation.
-The prover maintains a well-formed context $E[\Gamma]$, and processes commands to modify the context and conduct calculations.
+The prover maintains a well-formed context $E$, and processes commands to modify the context and conduct calculations.
 The commands are listed below.
@@
     \item \texttt{\textcolor{NavyBlue}{Def} ID := term.} It defines the \texttt{ID} as the \texttt{term}, using the \textbf{W-Def} typing rule.
-    \item \texttt{\textcolor{NavyBlue}{Var} ID := term.} It make an assumption of \texttt{ID} with the \texttt{term} as type, using the \textbf{W-AssumeE} typing rules.
+    \item \texttt{\textcolor{NavyBlue}{Var} ID := term.} It make an assumption of \texttt{ID} with the \texttt{term} as type, using the \textbf{W-Assume} typing rules.
     \item \texttt{\textcolor{NavyBlue}{Check} term.} Type checking the \texttt{term} and output the result.
@@
 \subsection{Evaluation}
-We evaluate D-Hammer on several example sets, and make a comparison with the previous tool DiracDec.
+We evaluate D-Hammer on several example sets, and make a comparison with the previous tool DiracDec~\cite{diracdec}.
 The experiments are carried out using a MacBook Pro with M3 Max chip. See~\Cref{fig: all examples} for a summary.
@@
         CoqQ                    & 162          & 156       &    48.69       &   158     &  158   &     9.74     \\
-        others                  & 7          &  6         &   77.20    &    7        &   6     &  2.13     \\
+        % others                  & 7          &  6         &   77.20    &    7        &   6     &  2.13     \\
         % circuits                 & 2          & 2       &    17.67       &   3     &  2   &     1.4     \\
@@
     \includegraphics[width=0.8\textwidth]{fig/coqq.pdf}
-    \caption{The running time comparison between DiracDec and D-Hammer on the CoqQ benchmark.}
+    \caption{Time comparison between DiracDec and D-Hammer on the CoqQ benchmark.}
     \label{fig: CoqQ plot}
@@
 
-\subsubsection{Others}
-Other examples are taken from quantum circuits and one recent paper [TODO].
+% \subsubsection{Others}
+% Other examples are taken from quantum circuits and one recent paper [TODO].
 Because these examples involve decomposition on concrete $\ket{0}$ and $\ket{1}$ qubit basis, resulting a lot of addition elements, our algorithm dealing with AC symbols brings along significant improvement.
-One typical example is
-\begin{gather*}
-    (I \otimes P) \cdot U \cdot (I \otimes P) \cdot U^\dagger = \ket{0}\bra{0} \otimes I + \ket{1}\bra{1}\otimes (P \cdot P), \\
-    \textrm{where } P \triangleq e^{-i\theta/2} \ket{0}\bra{0} + e^{i\theta/2}\ket{1}\bra{1}, \qquad U \triangleq \ket{0}\bra{0} \otimes X + \ket{1}\bra{1} \otimes I.
-\end{gather*}
-It takes DiracDec about one minute, but D-Hammer solves it within one second.
+% One typical example is
+% \begin{gather*}
+%     (I \otimes P) \cdot U \cdot (I \otimes P) \cdot U^\dagger = \ket{0}\bra{0} \otimes I + \ket{1}\bra{1}\otimes (P \cdot P), \\
+%     \textrm{where } P \triangleq e^{-i\theta/2} \ket{0}\bra{0} + e^{i\theta/2}\ket{1}\bra{1}, \qquad U \triangleq \ket{0}\bra{0} \otimes X + \ket{1}\bra{1} \otimes I.
+% \end{gather*}
+% It takes DiracDec about one minute, but D-Hammer solves it within one second.
 
@@
 We have a new set of examples for labelled Dirac notation, see~\Cref{sec: examples for labelled}.
+These examples comes from multiple sources, including 
 These examples use label to express multiplication and summation flexibly.

### 2025-01-29 main.tex
.
update evaluation
update preliminary
update

@@
 {
-    \begin{longtable}{cl}
+    \begin{longtable}{p{3cm} l}
     \caption{#1}\\
@@
 
+\def\>{\ensuremath{\rangle}}
+\def\<{\ensuremath{\langle}}
+
+\newcommand {\cH } {{\mathcal{H}}}
+\newcommand {\bu } {{\mathbf{u}}}
+\newcommand {\bv } {{\mathbf{v}}}
+\newcommand {\bC } {{\mathbb{C}}}
+\newcommand {\ghz } {{\mathrm{GHZ}}}
+\newcommand {\cnot } {{\mathsf{CNOT}}}
+\newcommand {\ccnot } {{\mathsf{CCNOT}}}
+\newcommand {\swap } {{\mathsf{SWAP}}}
+\renewcommand{\matrix}[1]{\begin{bmatrix}#1\end{bmatrix}}
+
 %
@@
 %
-\title{D-Hammer: Practical Proof Automation of Dirac Notation Equations}
+\title{D-Hammer: Efficient Equational Reasoning for Labelled Dirac Notation}
+\author{}
+\institute{}
 
@@
 higher-order, dependently-typed language for labelled Dirac notation;
-an efficient algorithm to normalize expressions (modulo associativity
+an efficient algorithm for normalization (incorporating associativity
 and commutativity); and an optimized \CC\ implementation. We evaluate
@@
 \newcommand{\bU}{\mathbf{U}}
-\newcommand{\<}{\langle}
+\renewcommand{\<}{\langle}
 \newcommand{\simp}{\mathsf{Simp}}

### 2025-01-29 01intro_preliminary.tex

--- 2025-01-29 01intro_preliminary.tex
@@
   \label{ex: motivating}
-  Let $p,q,r$ be three quantum systems with the same Hilbert space \( \mathcal{H}_T \) (i.e., with computational bases $\{|i\rangle\}_{i\in T}$) and GHZ state $\ghz = \sum_i|iii\rangle\langle iii|$.
-Consider the linear operator sets $U_{ij}$ and the corresponding multiplexers $M\triangleq \sum_{ij}|ij\rangle\langle ij|\otimes U_{ij}$ and $N\triangleq \sum_i|i\rangle\langle i|\otimes U_{ii}$.
+  Let $p,q,r$ be three quantum systems with the same Hilbert space \( \mathcal{H}_T \) (i.e., with computational bases $\{|i\>\}_{i\in T}$) and GHZ state $|\ghz\> = \sum_i|iii\>\<iii|$.
+Consider the linear operator sets $U_{ij}$ and the corresponding multiplexers $M\triangleq \sum_{ij}|ij\>\<ij|\otimes U_{ij}$ and $N\triangleq \sum_i|i\>\<i|\otimes U_{ii}$.
 Show that:
-  $$M_{prq}\ghz_{prq} = N_{rq}|\ghz\rangle_{pqr} = N_{pq}|\ghz\rangle_{pqr}.$$
+  $$M_{prq}|\ghz\>_{prq} = N_{rq}|\ghz\>_{pqr} = N_{pq}|\ghz\>_{pqr}.$$
 \end{example}

### 2025-01-29 main.tex
.

@@
 higher-order, dependently-typed language for labelled Dirac notation;
-an efficient algorithm for normalization (incorporating associativity
-and commutativity); and an optimized \CC\ implementation. We evaluate
-the implementation on representative examples from plain and labelled
-Dirac notation. In the case of plain Dirac notation, we show that our
-implementation significantly outperforms DiracDec (Xu et al, POPL'25).
+an efficient algorithm for normalization; and an optimized
+\CC\ implementation. We evaluate the implementation on representative
+examples from plain and labelled Dirac notation. In the case of plain
+Dirac notation, we show that our implementation significantly
+outperforms DiracDec (Xu et al, POPL'25).
 

### 2025-01-30 01intro_preliminary.tex
.

@@
 
-Dirac notation, also known as bra-ket notation, provides an intuitive and concise mathematical framework for describing quantum states and operations in quantum mechanics. 
-We write $\cH$ for a Hilbert space, i.e., a vector space equipped with an standard inner produce $\langle\bu,\bv\rangle \in \bC$ for $\bu,\bv\in\cH$.
-Dirac notation consists of the following components that reflect the basic postulates of quantum mechanics:
+Dirac notation, also known as bra-ket notation, provides an intuitive
+and concise mathematical framework for describing quantum states and
+operations in quantum mechanics.  We write $\cH$ for a Hilbert space,
+i.e., a vector space equipped with an standard inner product
+$\langle\bu,\bv\rangle \in \bC$ for $\bu,\bv\in\cH$.  Dirac notation
+consists of the following components that reflect the basic postulates
+of quantum mechanics:
 \begin{itemize}
@@
   %It is alternative to interpret as a linear mapping from $\cH$ to $\bC$ defined as $\langle u| : |v\rangle \mapsto \langle \bu,\bv\rangle$. It is the conjugate transpose of $|u\rangle$.
-  For exmaple, $\langle 0| = \matrix{1 & 0}$ and $\langle 1| = \matrix{0 & 1}$.
-  \item Inner produce $\langle u|v\rangle \triangleq \langle \bu,\bv\rangle$ which indicates the probability amplitude for $|u\rangle$ to collapse into $|v\rangle$. 
-  In convention, it is computed by matrix multiplication of two state, e.g., 
+  For example, $\langle 0| = \matrix{1 & 0}$ and $\langle 1| = \matrix{0 & 1}$.
+  \item Inner product $\langle u|v\rangle \triangleq \langle \bu,\bv\rangle$ which indicates the probability amplitude for $|u\rangle$ to collapse into $|v\rangle$. 
+  By convention, it is computed by matrix multiplication of two states, e.g., 
   $\langle 0|1\rangle = \matrix{1 & 0} \matrix{0 \\ 1} = 0.$
   \item Outer product $|u\rangle\langle v| \triangleq |w\rangle \mapsto (\langle v|w\rangle) |u\rangle$.
-Any linear map, such as unitary tranformation, measurement operators, etc, can be decomposed as the sum of outer products.
+Any linear map, such as unitary tranformation, measurement operator, etc, can be decomposed as the sum of outer products.
 It is also computed by matrix multiplication, e.g.,
@@
 \subsection{Labelled Dirac Notation and Motivating Example}
-
-Unfortunately, the standard Dirac notation becomes unexpectedly tedious for describing multi-body systems. To see this, consider the following example:
+Labelled Dirac notation is a generalization of Dirac notation for
+describing multi-body quantum systems. The need for labelled Dirac
+notation is motivated by the following example:
 \begin{example}
   \label{example1}
-  Let $p,q,r$ be three qubits and initially in the (unnormalized) $\ghz$ state $|\ghz\rangle \triangleq |000\rangle+|111\rangle$.
-Show that applying 3-qubit Toffoli gate ($\ccnot$) with control qubits $p,r$ and target qubit $q$ is equivalent to applying 2-qubit $\cnot$ gate with control qubit $r$ or $p$ and target qubit $q$.
-\end{example}
-We formalize the conclusion using standard Dirac notation as
+  Let $p,q,r$ be three qubits and initially in the (unnormalized)
+  $\ghz$ state $|\ghz\rangle \triangleq |000\rangle+|111\rangle$.
+  Applying the 3-qubit Toffoli gate ($\ccnot$) with control qubits
+  $p,r$ and target qubit $q$ to GHZ is equivalent to applying 2-qubit
+  $\cnot$ gate with control qubit $r$ or $p$ and target qubit $q$.
+  Using Dirac notation, the identity is written as:
 \begin{align*}
@@
 \end{align*}
-following the steps: 1.
-Arrange qubits in the conventional order, here we choose $p,r,q$ to simplify the representation of $\ccnot$; 2.
-Lift the local operation $\cnot$ to the global system.
-For $\cnot$ acting on $r,q$ is easier, since $r,q$ consistent with the chosen order, and we only need to tensoring it by an identity operator $I$ on $p$, i.e., $(I\otimes \cnot)$.
-For $\cnot$ acting on $p,q$, realize that $p,q$ is not adjacent in the chosen order, we additionally need to use $\swap$ to temporarily exchange the qubits $p$ and $r$, i.e., globally, the $(\swap\otimes I)$ before and after $(I\otimes \cnot)$ (lifting of $\cnot$ on $r,q$).
+Formalizing the statement in Dirac notation requires the following
+steps: 1. Arrange qubits in the conventional order, here we choose
+$p,r,q$ to simplify the representation of $\ccnot$; 2. Lift the local
+operation $\cnot$ to the global system. For $\cnot$ acting on $r,q$ is
+easier, since $r,q$ consistent with the chosen order, and we only need
+to tensoring it by an identity operator $I$ on $p$, i.e., $(I\otimes
+\cnot)$.  For $\cnot$ acting on $p,q$, realize that $p,q$ is not
+adjacent in the chosen order, we additionally need to use $\swap$ to
+temporarily exchange the qubits $p$ and $r$, i.e., globally, the
+$(\swap\otimes I)$ before and after $(I\otimes \cnot)$ (lifting of
+$\cnot$ on $r,q$).
 
-Roughly speaking, encoding in standard Dirac notation needs to tensoring identity operators and employing additional $\swap$ gates since generally the conventional order cannot guarantee 1.
-the order of all local operations is consistent with it, 2.
-the local operations involve only adjacent qubits in the conventional order.
-
-To address this, physicists introduce labels (or subscripts) to indicate the systems on which quantum states or operations are applied, thereby avoiding unnecessary lifting and swap gates.
-For example, rewriting the above example using labels:
+Roughly speaking, encoding in standard Dirac notation needs to
+tensoring identity operators and employing additional $\swap$ gates
+since generally the conventional order cannot guarantee 1. the order
+of all local operations is consistent with it, 2. the local operations
+involve only adjacent qubits in the conventional order.
+\end{example}
+To address the limitations of Dirac notations, physicists routinely
+use labels (or subscripts) to indicate the systems on which quantum
+states or operations are applied, thereby avoiding unnecessary lifting
+and swap gates. For example, rewriting the above example using labels
+yields:
 \begin{align*}
@@
 \end{align*}
-without determining the conventional order nor lifting using additional $I$ and $\swap$s. 
-The use of labels allows us to avoid deliberately maintaining the order of qubits.
-Tensor product become an associative and commutative (AC) symbol, so we can rearrange qubits as needed to perform calculations.
-For example, we can perform the calculation as follows:
+This formalization avoids determining and maintaining the conventional
+order of qubits. nor lifting using additional $I$ and $\swap$s. In
+this setting, the tensor products become an associative and
+commutative (AC) symbol, so we can rearrange qubits as needed to
+perform calculations. For our example, we can perform the calculation as
+follows:
 \begin{align*}
@@
 \end{align*}
-where the RHS of each line are equivalent.
-In addition, it can very freely describe local measurements, partial traces (the state or evolution of subsystems in multi-body systems), and partial inner products (the representation of partial traces in pure states), which are sufficient to handle the mathematical formulas of quantum mechanics in multi-body systems.
+where the RHS of each line are equivalent.  In addition, labelled
+Dirac notation can conveniently describe local measurements, partial
+traces (the state or evolution of subsystems in multi-body systems),
+and partial inner products (the representation of partial traces in
+pure states), which are sufficient to handle the mathematical formulas
+of quantum mechanics in multi-body systems.
 
-Labelled Dirac notation is not only ubiquitous in the description of multi-body systems, but also plays an important role in program logic -- just as in classical program logic, variable names are used to construct logical formulas instead of describing them as global memory functions, in quantum program logic, variable names are used to label the subsystems on which quantum gates act, rather than lifting them to the global system -- the latter would cause exponential length of the formula w.r.t the number of variables, as discussed in [cite].
+Labelled Dirac notation is not only ubiquitous in the description of multi-body systems, but also plays an important role in program logics.
+Just as in classical program logic, variable names are used to construct logical formulas instead of describing them as global memory functions, in quantum program logic, variable names are used to label the subsystems on which quantum gates act, rather than lifting them to the global system.
+Indeed, the latter would cause exponential length of the formula w.r.t the number of variables, as discussed in [cite].
 
-We generalize Example \ref{example1} as the motivating example in the following:
+The next sections introduce labelled Dirac notation, and an automated
+procedure for proving equalities. We will use the following
+generalization of Example \ref{example1} as motivating example:
 \begin{example}[Apply multiplexer to $\ghz$ state]
@@
 Consider the linear operator sets $U_{ij}$ and the corresponding multiplexers $M\triangleq \sum_{ij}|ij\>\<ij|\otimes U_{ij}$ and $N\triangleq \sum_i|i\>\<i|\otimes U_{ii}$.
-Show that:
+We have:
   $$M_{prq}|\ghz\>_{prq} = N_{rq}|\ghz\>_{pqr} = N_{pq}|\ghz\>_{pqr}.$$

### 2025-01-30 01intro_preliminary.tex
change back the motivating example

@@
 The next sections introduce labelled Dirac notation, and an automated
-procedure for proving equalities. We will use the following
-generalization of Example \ref{example1} as motivating example:
-\begin{example}[Apply multiplexer to $\ghz$ state]
+procedure for proving equalities. 
+Instead of GHZ state in Example \ref{example1}, we will use a similar but simpler example on Bell state as motivating example:
+\begin{example}
     \label{ex: motivating}
-  Let $p,q,r$ be three quantum systems with the same Hilbert space \( \mathcal{H}_T \) (i.e., with computational bases $\{|i\>\}_{i\in T}$) and GHZ state $|\ghz\> = \sum_i|iii\>\<iii|$.
+    Let \( q \) and \( r \) represent two quantum systems in the Hilbert space \( \mathcal{H}_T \).
+Let \( M \) be a quantum operation acting on \( \mathcal{H}_T \), and let \( \ket{\Phi} = \sum_{i \in T} \ket{i} \otimes \ket{i} \) be the maximally entangled state.
+Then, it holds that
+    \[
+    M_q \ket{\Phi}_{(q, r)} = M_r^T \ket{\Phi}_{(q, r)}.
+    \]
+\end{example}
+
+% We will use the following
+% generalization of Example \ref{example1} as motivating example:
+% \begin{example}[Apply multiplexer to $\ghz$ state]
+%   \label{ex: motivating}
+%   Let $p,q,r$ be three quantum systems with the same Hilbert space \( \mathcal{H}_T \) (i.e., with computational bases $\{|i\>\}_{i\in T}$) and GHZ state $|\ghz\> = \sum_i|iii\>\<iii|$.
 Consider the linear operator sets $U_{ij}$ and the corresponding multiplexers $M\triangleq \sum_{ij}|ij\>\<ij|\otimes U_{ij}$ and $N\triangleq \sum_i|i\>\<i|\otimes U_{ii}$.
 We have:
-  $$M_{prq}|\ghz\>_{prq} = N_{rq}|\ghz\>_{pqr} = N_{pq}|\ghz\>_{pqr}.$$
-\end{example}
+%   $$M_{prq}|\ghz\>_{prq} = N_{rq}|\ghz\>_{pqr} = N_{pq}|\ghz\>_{pqr}.$$
+% \end{example}
 

### 2025-01-30 03labelled.tex
update preliminary
update
change back the motivating example

--- 2025-01-27 03labelled.tex
@@
 \label{sec: labelled}
-Labelled Dirac notation uses register names to indicate the quantum system of vectors and operators.
-This allows us to express and reason aboud the states and operations locally, without referring to the whole system.
-For instance, assume $Q$ and $R$ are two registers, we have
-\[
-    M_Q \cdot \ket{\Phi}_{(Q, R)} = ((M \otimes I) \cdot \ket{\Phi})_{(Q, R)}.
-\]
-On the left hand side, the state of two subsystems is $\ket{\Phi}$, and we apply quantum operation $M$ on the system $Q$.
-It is equivalent to extend the operation using identity operators on other subsystems (i.e.
-the cylinder extension), and consider the application in the whole system.
 
-In this section, we introduce the language and typing of registers and labelled Dirac notation, and demonstrate how to transform the equivalence problem into that for the Dirac notation studied above.
-The first new notion is quantum registers, and we assume they are constructed from a set $\cR$.
-\begin{definition}[quantum registers]
+% what is the labelled dirac notation?
+
+% the semantics -- cylinder extension -- uniqueness
+
+% \subsection{Syntax, Typing and Rewriting Rules}
+
+% syntax -- typing -- rules
+
+% \subsection{Normalization}
+% normalization -- algorithm that eliminate labels
+
+In this section, we extend the language by allowing quantum variables to indicate the quantum system of vectors and operators it act. 
+As discussed, this allows us to express and reason aboud the states and operations locally, without referring to the whole system. 
+We further demonstrate how to transform the equivalence problem into that for the Dirac notation studied above.
+
+\subsection{Syntax, Typing, Semantics and Rewriting Rules}
+We first introduce the notation of quantum registers for structured variable combinations.
+This is necessary because, unlike assignments for classical variables, unitary transformations on composite systems -- a quantum version of assignments -- cannot generally be decomposed into separate unitary transformations on individual subsystems.
+Let $\cR$ be the set of quantum variables.
+%The first new notion is quantum registers, and we assume they are constructed from a set $\cR$.
+\begin{definition}[Quantum Register] Register $R$ is inductively generated by
   \begin{align*}
@@
 \end{definition}
+For simplicity, we only allow pairing the registers, which corresponds to the structure of tensor product, i.e., $\cH_{(R_1,R_2)}$ is isomorphic to $\cH_{R_1}\otimes \cH_{R_2}$, which allows us to view the tensor product space as the space of paired registers.
 
-Registers can be composed by pairs of $(R, R)$, and this structure corresponds to the structure of tensor product spaces in Dirac notation.
-To reason about the registers, we define their variable set as the enumeration of all register symbols involved.
-
-\begin{definition}[register variable set]
+The no-cloning theorem, a fundamental property of quantum computing, prevents us from copying an unknown quantum state.
+This requires an additional check on the valid registers -- they should not include repeated quantum variables -- which is often handled in programming languages via linear types.
+Here, we start from the basic definition of \emph{order-free} variable set of a register:
+\begin{definition}
 The variable set of a register is defined inductively by:
@@
 % The labelled Dirac notation is an extension on the existing type and syntax.
-\begin{definition}[labelled Dirac notation]
+Now, we are ready to extend the type syntax and term syntax as follows:
+\begin{definition}[Labelled Dirac Notation]
   The \textbf{labelled Dirac notation} includes all Dirac notation symbols and the generators defined below.
-  Here, $s\subseteq \cR$ is a register variable set.
+  Here, $s\subseteq \cR$ is a quantum variable set.
   \begin{align*}
@@
     e & ::= R \mid |i\>_r \mid {}_r\<i| \mid e_R \mid e_{R;R} \mid
-    e \otimes e \otimes \cdots \otimes e \mid e \cdot e
+    e \otimes e \otimes \cdots \otimes e \mid e \cdot e.
   \end{align*}
@@
 $\DType(s_1, s_2)$ is the unified type for all labelled Dirac notation, where $s_1$ indicates the codomain systems and $s_2$ indicates the domain systems.
-For instance, labelled ket has type $\DType(s_1, \emptyset)$, and labelled bra has type $\DType(\emptyset, s_2)$.
+For instance, labelled ket $|i\>_r$ has type $\DType(\{r\}, \emptyset)$, and labelled bra ${}_r\<i|$ has type $\DType(\emptyset, \{r\})$.
 $\reg(\sigma)$ are types for registers $R$, and the index $\sigma$ indicates the type of Hilbert space represented by the register.
-Terms also include the labelled notation $e_R$ for bra, ket and $e_{R;R}$ for operators.
-We introduce new dot and tensor product symbols for labelled Dirac notation.
-In labelled Dirac notation, the structure of tensor product does not matter, therefore $\otimes$ is an AC symbol.
-Following the unified type $\DType(s,s)$, all kinds of multiplications are represented by the same dot product $e \cdot e$.
-Finally, $\ket{i}_r$ and ${}_r\bra{i}$ are labelled basis for the normal form of labelled Dirac notation, where $r$ are registers symbols in $\mathcal{R}$.
+It is allowed to lift a Dirac notation associate with corresponding quantum variables or registers, e.g., $\ket{i}_r$ and ${}_r\bra{i}$ are labelled basis, 
+$e_R$ for bra, ket and $e_{R;R}$ for operators. 
+We further introduce new $\cdot$ for generalized composition (unified for all kinds of multiplications between kets, bras and operators) and $\otimes$ for labelled tensor product since they do not share the same properties v.s.
+its counterpart in Dirac notation, i.e., generalized composition is not associative and labelled tensor is indeed an AC symbol.
+% In labelled Dirac notation, the structure of tensor product does not matter, therefore $\otimes$ is an AC symbol.
+% Following the unified type $\DType(s,s)$, all kinds of multiplications are represented by the same dot product $e \cdot e$.
+% Finally, $\ket{i}_r$ and ${}_r\bra{i}$ are labelled basis for the normal form of labelled Dirac notation, where $r$ are registers symbols in $\mathcal{R}$.
 
-\subsubsection{Typing rules}
-Some typing rules are introduced here.
-The rule for $\DType(s_1, s_2)$ requires that all registers in variable set $s_1$ and $s_2$ are well-typed.
-The rule for $K_R$ demonstrates how a register label is added to the Dirac notation, and the rule for $D^\dagger$ shows that labelled Dirac notation also have symbols for calculation, such as the adjoint.
-\[
-    \frac{E[\Gamma] \vdash  \sigma : \Index}{E[\Gamma] \vdash \reg(\sigma) : \Type}
+\paragraph*{Typing rules.}
+There are various rules for computing types and checking vadility of registers and labelled terms. Here we display some of the rules and see Appendix for all the rest.
+\begin{gather*}
+  \frac{
+      E \vdash R : \reg(\sigma) \quad
+      E \vdash Q : \reg(\tau)
+      \quad \var(R) \cap \var(Q) = \emptyset
+  }{E \vdash (R,Q) : \reg(\sigma \times \tau)} \\[0.2cm]
+  \frac{
+          E \vdash r : \reg(\sigma) \quad
+          E \vdash i : \Basis(\sigma)}
+  {E \vdash |i\>_r : \DType(\{r\}, \emptyset)}
   \qquad
-    \frac{E[\Gamma] \vdash r : \reg(\sigma_r) \text{ for all $r$ in $s_1$ and $s_2$} }{E[\Gamma] \vdash \DType(s_1, s_2) : \textsf{Type}} 
-\]
-\[
-    \frac{E[\Gamma] \vdash R : \reg(\sigma)\qquad E[\Gamma] \vdash K : \KType(\sigma)}{E[\Gamma] \vdash K_R : \DType(\var R, \emptyset)}
-    \qquad
-    \frac{E[\Gamma] \vdash D : \DType(s_1,s_2)}{E[\Gamma] \vdash D^\dagger : \DType(s_2,s_1)}
-\]
-The dot and the tensor product symbols are different from those in unlabelled Dirac notation.
+  \frac{E \vdash R : \reg(\sigma)\qquad E \vdash K : \KType(\sigma)}{E \vdash K_R : \DType(\var(R), \emptyset)} \\[0.2cm]
+    \frac{
+        E \vdash D_i : \DType(s_i,s_i') \qquad
+        \forall\,i\neq j.\ s_i\cap s_j = \emptyset \qquad
+        \forall\,i\neq j.\ s'_i\cap s'_j = \emptyset
+    }
+    {E \vdash D_1 \otimes \cdots \otimes D_i : \DType(\bigcup_i s_i, \bigcup_i s_i')}.
+\end{gather*}
+The first rule says that a paired register is of the product type and its components should be disjoint.
+Once we lift a Dirac notation to the labelled one (line 2), we enforce that the term and register are of the same indices, reflecting the fact the state should be consistant to the corresponding subsystems.
+The third line gives the typing of labelled tensoring with a check whether the component subsystems are disjoint with each other.
+
+% \[
+%     \frac{
+%         \begin{aligned}
+%             E \vdash D_1 : \DType(s_1,s_1') \\
+%             E \vdash D_2 : \DType(s_2,s_2')
+%         \end{aligned}
+%         \qquad 
+%         \begin{aligned}
+%             s_1 \cap s_2 \backslash s_1' = \emptyset \\
+%             s_2' \cap s_1' \backslash s_2 = \emptyset
+%         \end{aligned}
+%     }
+%     {E \vdash D_1\cdot D_2 : \DType(s_1 \cup (s_2\backslash s_1'), s_2' \cup (s_1'\backslash s_2))}.
+% \]
+
+% Some typing rules are introduced here.
+% The rule for $\DType(s_1, s_2)$ requires that all registers in variable set $s_1$ and $s_2$ are well-typed.
+% The rule for $K_R$ demonstrates how a register label is added to the Dirac notation, and the rule for $D^\dagger$ shows that labelled Dirac notation also have symbols for calculation, such as the adjoint.
+% \[
+%     \frac{E \vdash  \sigma : \Index}{E \vdash \reg(\sigma) : \Type}
+%     \qquad
+%     \frac{E \vdash r : \reg(\sigma_r) \text{ for all $r$ in $s_1$ and $s_2$} }{E \vdash \DType(s_1, s_2) : \textsf{Type}} 
+% \]
+% \[
+%     \frac{E \vdash R : \reg(\sigma)\qquad E \vdash K : \KType(\sigma)}{E \vdash K_R : \DType(\var R, \emptyset)}
+%     \qquad
+%     \frac{E \vdash D : \DType(s_1,s_2)}{E \vdash D^\dagger : \DType(s_2,s_1)}
+% \]
+% The dot and the tensor product symbols are different from those in unlabelled Dirac notation.
 Since the goal of labels is to replace the order and structure of tensor products by the reference to registers, the tensor product becomes an AC symbol.
 The typing still checks whether the component subsystems are disjoint with each other.
-\[
-    \frac{
-        E[\Gamma] \vdash D_i : \DType(s_i,s_i') \qquad
-        \bigcap_i s_i = \emptyset \qquad
-        \bigcap_i s_i' = \emptyset
-    }
-    {E[\Gamma] \vdash D_1 \otimes \cdots \otimes D_i : \DType(\bigcup_i s_i, \bigcup_i s_i')}.
-\]
-As for the dot product, the disjointness is considered except registers contracted by multiplication.
-\[
-    \qquad
-    \frac{
-        \begin{aligned}
-            E[\Gamma] \vdash D_1 : \DType(s_1,s_1') \\
-            E[\Gamma] \vdash D_2 : \DType(s_2,s_2')
-        \end{aligned}
-        \qquad 
-        \begin{aligned}
-            s_1 \cap s_2 \backslash s_1' = \emptyset \\
-            s_2' \cap s_1' \backslash s_2 = \emptyset
-        \end{aligned}
-    }
-    {E[\Gamma] \vdash D_1\cdot D_2 : \DType(s_1 \cup (s_2\backslash s_1'), s_2' \cup (s_1'\backslash s_2))}.
-\]
+% \[
+%     \frac{
+%         E \vdash D_i : \DType(s_i,s_i') \qquad
+%         \bigcap_i s_i = \emptyset \qquad
+%         \bigcap_i s_i' = \emptyset
+%     }
+%     {E \vdash D_1 \otimes \cdots \otimes D_i : \DType(\bigcup_i s_i, \bigcup_i s_i')}.
+% \]
+% As for the dot product, the disjointness is considered except registers contracted by multiplication.
+% \[
+%     \qquad
+%     \frac{
+%         \begin{aligned}
+%             E \vdash D_1 : \DType(s_1,s_1') \\
+%             E \vdash D_2 : \DType(s_2,s_2')
+%         \end{aligned}
+%         \qquad 
+%         \begin{aligned}
+%             s_1 \cap s_2 \backslash s_1' = \emptyset \\
+%             s_2' \cap s_1' \backslash s_2 = \emptyset
+%         \end{aligned}
+%     }
+%     {E \vdash D_1\cdot D_2 : \DType(s_1 \cup (s_2\backslash s_1'), s_2' \cup (s_1'\backslash s_2))}.
+% \]
+
+\paragraph*{Semantics} While the labelled Dirac notation avoid unnecessary lifting and ordering, its semantics faithfully capture these details.
+The key points are 1.
+cylindrical extension that lift a ket or bra or operator to any larger domain and codomain; 2.
+general composition further employs cylindrical extension obeys the principle of ``localizing objects as much as possible''~\cite{Zhou2023}.
 

### 2025-01-30 02language_design.tex
update evaluation

@@
 However, computers cannot directly reason about equivalence through mathematical interpretations.
-Therefore, we proceed by defining a proof system that abstracts these concepts.
+We proceed by defining a proof system that abstracts these concepts.
 
@@
 \begin{lemma}
-    For all well-formed contexts \( E \) and terms \( e_1, e_2 \), if \( \vdash e_1 = e_2 \), then \( \sem{e_1} = \sem{e_2} \).
+    For all well-formed contexts \( E \) and terms \( e_1, e_2 \), if \( E \vdash e_1 = e_2 \), then \( \sem{e_1} = \sem{e_2} \).
 \end{lemma}
@@
 
-% With these syntax, typing and semantics introduced, we demonstrate the formalization of the symbols used in the motivating example to conclude this section.
-% \begin{example}[Formalizing the Motivating Example]
-%     \label{ex: formalizing motivating}
-%     Definitions and assumptions in the environment \( E \) are formalized as follows:
-%     \begin{align*}
-%         & \text{TPO} && := \lambda T_1 : \Index. \lambda T_2 : \Index. \lambda O : \OType(T_1, T_2). \sum_{i \in \mathbf{U}(T_1)} \sum_{j \in \mathbf{U}(T_2)} \bra{i} O \ket{j} . \ket{j}\bra{i} \\
-%         & &&\quad : \forall T_1. \forall T_2. \OType(T_1, T_2) \to \OType(T_2, T_1); \\
-%         &\text{phi} &&:= \lambda T : \Index. \sum_{i \in \mathbf{U}(T)} \sum_{j \in \mathbf{U}(T)} \ket{(i, j)} : \forall T.\KType(T \times T); \\
-%         & T && : \Index; \\
-%         & M && : \OType(T, T).
-%     \end{align*}
-%     The symbol \( \text{TPO} \) represents the transpose of an operator, polymorphic on the Hilbert spaces \( T_1 \) and \( T_2 \).
+With these syntax, typing and semantics introduced, we demonstrate the formalization of the symbols used in the motivating example to conclude this section.
+\begin{example}[Formalizing the Motivating Example]
+    \label{ex: formalizing motivating}
+    Definitions and assumptions in the environment \( E \) are formalized as follows:
+    \begin{align*}
+        & \text{TPO} && := \lambda T_1 : \Index. \lambda T_2 : \Index. \lambda O : \OType(T_1, T_2). \sum_{i \in \mathbf{U}(T_1)} \sum_{j \in \mathbf{U}(T_2)} \bra{i} O \ket{j} . \ket{j}\bra{i} \\
+        & &&\quad : \forall T_1. \forall T_2. \OType(T_1, T_2) \to \OType(T_2, T_1); \\
+        &\text{phi} &&:= \lambda T : \Index. \sum_{i \in \mathbf{U}(T)} \sum_{j \in \mathbf{U}(T)} \ket{(i, j)} : \forall T.\KType(T \times T); \\
+        & T && : \Index; \\
+        & M && : \OType(T, T).
+    \end{align*}
+    Notice how the functions and higher-order typing helps to formalize the abstract concepts here.
+    The symbol \( \text{TPO} \) represents the transpose of an operator, polymorphic on the Hilbert spaces \( T_1 \) and \( T_2 \).
+And its type \(\forall T_1.
+\forall T_2.
+\OType(T_1, T_2) \to \OType(T_2, T_1)\) suggests that TPO is a function from operators to operators, depending on the index $T_1$ and $T_2$.
     The symbol \( \text{phi} \) takes the index \( T \) and defines the maximally entangled states, summing over all basis elements in \( T \), as indicated by the universal set \( \mathbf{U}(T) \).
-%     With the assumption of the index \( T \) and operator \( M \), we can express the equivalence in the non-labelled version as:
-%     \[
-%     (\textrm{M} \otimes \mathbf{1}_\mathcal{O}(\textrm{T})) \cdot (\textrm{phi T}) = (\mathbf{1}_\mathcal{O}(\textrm{T}) \otimes (\textrm{TPO T T M})) \cdot (\textrm{phi T}).
-%     \]
-% \end{example}
+    With the assumption of the index \( T \) and operator \( M \), we can express the equivalence in the plain Dirac notation as:
+    \[
+    (\textrm{M} \otimes \mathbf{1}_\mathcal{O}(\textrm{T})) \cdot (\textrm{phi T}) = (\mathbf{1}_\mathcal{O}(\textrm{T}) \otimes (\textrm{TPO T T M})) \cdot (\textrm{phi T}).
+    \]
+\end{example}

### 2025-01-30 05tool_eval.tex
update evaluation

@@
 
-\subsubsection{Use Case}
-Here we encode the motivating~\Cref{ex: motivating}, examine and explain how it is checked in D-Hammer. The encoding is shown below.
+\subsection{Equivalence Checking Algorithm}
+
+The equivalence of Dirac notations are decided by normalization, which transforms equivalent expressions under a set of axioms into the same syntax.
+We use an efficient algorithm to perform the normalization totally on $R \cup F$.
+
+\begin{enumerate}
+    \item \textbf{First Rewritings}: Expand definitions and simplify expressions.
+    \item \textbf{Variable Expansion}: Convert to abstract element-wise representation.
+    \item \textbf{Second Rewritings}: Normalize terms on \( R \) modulo \( F \).
+    \item \textbf{Sorting Without Bound Variables}: Normalize AC-equivalence.
+    \item \textbf{Swap Successive Summations}: Normalize SUM-SWAP equivalence.
+    \item \textbf{De Bruijn Normalization}: Normalize \( \alpha \)-equivalence.
+\end{enumerate}
+Here steps 1 - 3 are term rewriting for $R$, and steps 4 - 6 are special algorithms to further normalize the axiom set $F$.
+Step 2 expands variables like $K$ to their abstract element-wise representation $\sum_{i} (\bra{i} \cdot K).
+\ket{i}$, which improves the decidability with sum.
+For step 4 and 5, the key observation is that in a successive sum expression \( \sum_{i \in s_1} \cdots \sum_{j \in s_n} A \), the names and order of the bound variables \( i, \dots, j \) can be freely permuted.
+Therefore we first ignore bound variables and normalize AC-equivalence by sorting.
+Then decide the order of summation accordingly.
+The last step converts functions to de Bruijn indeices~\cite{deBruijn1972lambda} to decide $\alpha$-equivalence.
+
+As a comparison, the previous work conducts normalization paritally on $R$, and decide axioms in $F$ by iterating through all possible permutations.
+Our algorithm totally normalize the term and significantly improves the efficiency.
+\[
+\text{(partial)}\qquad e_1 \mathop{\twoheadrightarrow}^R e_1' \mathop{=}^F e_2' \mathop{\twoheadleftarrow}^R e_2
+\qquad
+\text{(total)}\qquad e_1 \mathop{\twoheadrightarrow}^{R \cup F} e \equiv e \mathop{\twoheadleftarrow}^{R \cup F} e_2
+\]
+
+\yx{we can add a normalization example here.}
+
+\subsection{Use Case}
+Here we encode the motivating~\Cref{ex: motivating}, examine and explain how to check it using D-Hammer. The encoding is shown below.
 
@@
 Var r1 : REG[T]. Var r2 : REG[T].
-CheckEq M_r1;r1 (phi T)_(r1, r2) with (TPO T T M)_r2;r2 (phi T)_(r1, r2).
+CheckEq M_r1 (phi T)_(r1, r2) with (TPO T T M)_r2 (phi T)_(r1, r2).
     \end{lstlisting}
@@
 
-In the left hand side of \texttt{\textcolor{NavyBlue}{CheckEq}} command, \texttt{M\_r1;r1} denotes the labelled notation $M_{r_1; r_1}$, and \texttt{(phi T)\_(r1, r2)} denotes the entangled state $\ket{\Phi}_{(r_1, r_2)}$.
+In the left hand side of \texttt{\textcolor{NavyBlue}{CheckEq}} command, \texttt{M\_r1} denotes the labelled notation $M_{r_1}$, and \texttt{(phi T)\_(r1, r2)} denotes the entangled state $\ket{\Phi}_{(r_1, r_2)}$.
 They are connected by a white space, which is parsed into the composition of Dirac notation, and will be reduced into the operator-ket multiplication after typing.
@@
 
-\subsection{Evaluation}
+\section{Evaluation}
 We evaluate D-Hammer on several example sets, and make a comparison with the previous tool DiracDec~\cite{diracdec}.
-The experiments are carried out using a MacBook Pro with M3 Max chip. See~\Cref{fig: all examples} for a summary.
+The experiments are carried out using a MacBook Pro with M3 Max chip. Results are summarized as follows.
 
-\begin{figure}[h]
+{
     \center
@@
         % research paper                & 4          & 4         &  59.53       &   4    & 4       &  0.73     \\
-        labelled Dirac notation      &   -         &   -          &     -           &      -      &        13       &     6.69    \\
+        labelled Dirac notation      &   -         &   -          &     -           &      -      &        18       &     22.72    \\
         \hline
     \end{tabular}        
-    \caption{Summary of all examples evaluated.}
-    \label{fig: all examples}
-\end{figure}
+}
 
@@
 \subsubsection{Labelled Dirac Notation}
-We have a new set of examples for labelled Dirac notation, see~\Cref{sec: examples for labelled}.
-These examples comes from multiple sources, including 
-These examples use label to express multiplication and summation flexibly.
-D-Hammer extends these labelled notations to the global system using the algorithm in~\Cref{sec: labelled}, and decide their equivalence within reasonable time.
-In particular, D-Hammer solved a very sophisticated long example with 7 registers from [LICS...], which is challenging even for experts.
+We present a new set of examples for labelled Dirac notation (LDN), as detailed in~\Cref{sec: examples for labelled}.
+These examples, illustrated in~\Cref{fig: labelled examples}, include six representative cases drawn from a variety of sources, such as well-established theorems, research paper results, quantum circuit equivalence, and quantum program verification.
+D-Hammer normalizes them and successfully checks their equivalence using the algorithm in~\Cref{sec: labelled}.
+Among the examples, LDN-12 shows the flexibility in combining labelled Dirac notations.
+Example LDN-14 and LDN-18 are about circuits or programs on qubits.
+LDN-14 shows how to calculate controlled-not gate in different ways.
+In LDN-18, $q, q_1$ and $q_2$ represent variables for quantum programs, and the equation shows the change of the state after applying $X$ gate on $q_2$.
+Experiments show that our algorithm is relatively inefficient for labelled Dirac notations based on qubits.
+
+It is a noteworthy result that D-Hammer solved LDN-10, a highly complex and lengthy example.
+It is a theorem on quantum separation logic from~\cite{DBLP:conf/lics/ZhouBHYY21}, and is challenging to prove even for experts.
+Notice that 7 registers are involved here, and it would be impractical to organize and referring to the subsystems without using labels.
+
+\begin{figure}[h]
+    \center
+    % \scriptsize
+    \setlength{\extrarowheight}{2pt}
+    \begin{tabular}{l c c l}
+        \hline
+        example & source & time(s) & equation \\
+        \hline
+        LDN-4 & theorem & 0.03 & \( \bra{\Phi}_{(x, y)} M_y \cdot N_x \ket{\Phi}_{(x, y)} = \mathrm{tr}(M^T N) \) \\
+        LDN-10 & paper~\cite{DBLP:conf/lics/ZhouBHYY21} & 5.17 & \( \mathrm{tr}_{((a',(b,b')),c')}\Big[\mathrm{tr}_{r}\Big(U_{(r,(a,b))} \cdot\Big(|s\>_r\<s|\otimes 
+        \Big[V_{((a',(b,b')),c')} \cdots \)\\
+        LDN-11 & paper~\cite{PALSBERG2024206} & 0.12 & \( U_{(a,b)}\cdot W_{(b,c)}\cdot V_{(a,c)} = 
+        \sum_i |i\>_a\<i|\otimes \big((P_i)_c\cdot W_{(b,c)}\cdot (Q_i)_c\big)\) \\
+        LDN-12 & circuit & 0.07 & \( \ket{i}_{a;b}\bra{j} \cdot C_{(b,c)} \cdot D_{(c,d)} = {}_b\bra{j} \cdot C_{(b,c)} \cdot D_{(c,d)} \cdot \ket{i}_{a}\) \\
+        LDN-14 & circuit & 7.37 & \scriptsize{\( \textsf{CNOT}_{rq}\ket{\textrm{GHZ}}_{pqr} = (\textsf{CNOT}\ket{00})_{rq}\ket{0}_p + (\textsf{CNOT}\ket{11})_{rq}\ket{1}_p \)} \\
+        LDN-18 & program & 1.01 & \( -\ket{0}_q\ket{+-}_{q_1, q_2} = X_{q_2} \ket{0}_q\ket{+-}_{q_1, q_2} \) \\
+        \hline
+    \end{tabular}
+    \caption{Part of examples for labelled Dirac notations.}
+    \label{fig: labelled examples}
+\end{figure}

### 2025-01-30 01intro_preliminary.tex
update motivating example

@@
 
-The next sections introduce labelled Dirac notation, and an automated
-procedure for proving equalities. 
-Instead of GHZ state in Example \ref{example1}, we will use a similar but simpler example on Bell state as motivating example:
+The next sections introduce the formal language, labelled Dirac notation and the systematical way to reduce labels and prove equalities. 
+For this purpose, we use the following motivating example as the main role in our explanations.
+Instead of GHZ states, it is a similar but simpler example on Bell states:
 \begin{example}
@@
 \end{example}
+Following the introduction on labels, we can consider the global system $(q,r)$ and transform the equation above into the plain Dirac notation:
+\begin{align}
+    (M \otimes I) \ket{\Phi} = (I \otimes M^T) \ket{\Phi}.
+    \label{eq: motivating plain}
+\end{align}
+Together with corresponding explanations for this example, we show how an automated system is built and used to solve equalities likewise.
 

### 2025-01-30 03labelled.tex
update labelled section (not finished yet)

@@
+\newcommand {\cD } {{\mathcal{D}}}
+\newcommand {\cl } {{\mathit{cl}}}
+
 % Labelled Dirac Notation
@@
 
-\subsection{Syntax, Typing, Semantics and Rewriting Rules}
+\subsection{Syntax, Typing and Semantics}
 We first introduce the notation of quantum registers for structured variable combinations.
@@
 
-\paragraph*{Semantics} While the labelled Dirac notation avoid unnecessary lifting and ordering, its semantics faithfully capture these details.
+\paragraph*{Semantics.} While the labelled Dirac notation avoids unnecessary lifting and ordering, its semantics faithfully capture these details.
 The key points are 1.
-cylindrical extension that lift a ket or bra or operator to any larger domain and codomain; 2.
-general composition further employs cylindrical extension obeys the principle of ``localizing objects as much as possible''~\cite{Zhou2023}.
+cylindrical extension that lift a ket or bra or operator to larger domain and codomain; 2.
+general composition further employs cylindrical extension that obeys the principle of ``localizing objects as much as possible''~\cite{Zhou2023}.
 
-\subsubsection{Normalization}
-The normalization procedure of Dirac notation is extended to check equivalence with labels.
-We add rules to the term rewriting system, which in general try to represent labelled Dirac notation with labelled basis and scalar coefficients.
-The first step is the label elimination.
+\begin{definition}[Cylindrical Extension]
+  For any $D : \cD(s_1,s_2)$ and $s$ that disjoint with both $s_1$ and $s_2$, we define $\cl(D,s) \triangleq D\otimes \mathbf{1}_{s}$ of type $\cD(s_1\cup s, s_2\cup s)$.
+\end{definition}
+Formally, we equip $\cR$ with a default order, say for example the alphabetical order of names.
+For any valid register $R$, there exists the operator $\swap_R$ that sorts $R$; for example, if $E\vdash p : \reg(\sigma_p)$ and $E\vdash q : \reg(\sigma_q)$, then $\swap_{(q,p)} \triangleq \sum_{i\in \bU(\sigma_p)}\sum_{j\in \bU(\sigma_q)}|i\>\<j|\otimes |j\>\<i|$.
+We can further define $\swap_{s_1,s_2,\cdots}$ for merging disjoint sets orderly.
+
+For any interpretation $\sem{\cdot}$, and write $\cH_{\sem{s}} \triangleq \bigotimes_{r\in s}\cH_{\sem{\sigma_r}}$ for any $s\subseteq \cR$, where $E\vdash r : \reg(\sigma_r)$ for all $r\in\cR$.
+We interpret labelled Dirac notations inductively as (see details in Appendix) : assume $E\vdash D_i : \cD(s_i,s'_i)$ in the following,
+\begin{itemize}
+  \item $\sem{|i\>_r} = |\sem{i}\>$; $\sem{K_R} = \swap_R \cdot \sem{K}$; $\sem{O_{R_1,R_2}} = \swap_{R_1}\cdot \sem{O}\cdot \swap_{R_2}$;
+  \item $\sem{D_1\otimes \cdots \otimes D_n} = \swap_{s_1,\cdot,s_n}\cdot (\sem{D_1}\otimes \cdots \otimes \sem{D_1})\cdot \swap_{s'_1,\cdot,s'_n}$;
+  \item $\sem{D_1\cdot D_2} = \sem{\cl(D_1, s_2\backslash s'_1)}\cdot \sem{\cl(D_2, s'_1\backslash s_2)}$.
+Note that $s_2\backslash s'_1$ and $s'_1\backslash s_2$ are the minimal extension that make it interpretable.
+E.g., to interpret ${}_p\<i|\cdot |j\>_{p,q}$, we at least need to extend ${}_p\<i|$ to ${}_p\<i|\otimes I_{q}$.
+\end{itemize}
+Thus, $\sem{D}$ is a linear map from $\cH_{\sem{s'}}$ to $\cH_{\sem{s}}$ given $E\vdash D : \cD(s,s')$.
+Following the semantics, labelled tensor is independent of its order, i.e., $\sem{D_1\otimes D_2} = \sem{D_2\otimes D_1}$ and $\sem{D_1\otimes (D_2\otimes D_3)} = \sem{(D_1\otimes D_2)\otimes D_3}$, which ensures the soundness of treating labelled tensor as an AC symbol.
+
+\subsection{Normalization}
+The normalization procedure of labelled Dirac notation allows us to eliminate all labels and thus translate the equality of labelled terms to the equality of standard Dirac terms.
+This is achieved by the following three steps:
+
+\paragraph*{1. Elimination of $e_R$ or $e_{R;R}$.}
+We decompose all $e_R$ or $e_{R;R}$ to the labelled basis with scalar coefficients. 
+%We add rules to the term rewriting system, which in general try to represent labelled Dirac notation with labelled basis and scalar coefficients. The first step is the label elimination. 
 Take operator as an example:
@@
 \end{align*}
-The rules for $e_R$ (ket and bra) are similar. This step reduces all labelled terms $e_R$ or $e_{R;R}$.
-Other symbols on labelled Dirac notation are also reduced by rules like $(D_1 \cdot D_2)^\dagger \ \reduce\ D_2^\dagger \cdot D_1^\dagger$.
+The rules for $e_R$ (ket and bra) are similar.
+%This step reduces all labelled terms $e_R$ or $e_{R;R}$.
+
+\paragraph*{2. Rewriting to the normal form.}
+We add three types of rules for dealing with operators on labeled terms, 1) recursively applying them to subterms, 2) push big operators out and 3) eliminate bra-ket pairs. For example,
+%Other symbols on labelled Dirac notation are also 
+Take the rule for conjugate $(D_1 \cdot D_2)^\dagger \ \reduce\ D_2^\dagger \cdot D_1^\dagger$ as an example of 1).
+For 2), 
 The final step operates on sum and dot product. They will lift summation to the outside, and eliminate the bra-ket pairs whenever possible.

### 2025-01-30 03labelled.tex
update labelled section

@@
     \sum_{i_{r'_1}\in\bU(\sigma_{r'_1})}\cdots \sum_{i_{r'_{n'}}\in\bU(\sigma_{r'_{n'}})} \\
-    & \qquad (\<i_R|\cdot O\cdot |i_{R'}\>).(|i_{r_1}\>_{r_i}\otimes\cdots\otimes|i_{r_n}\>_{r_n} \otimes {}_{r'_1}\<i_{r'_1}|\otimes\cdots\otimes{}_{r'_{n'}}\<i_{r'_{n'}}|).
+    & \qquad (\<i_R|\cdot O\cdot |i_{R'}\>).(|i_{r_1}\>_{r_1}\otimes\cdots\otimes|i_{r_n}\>_{r_n} \otimes {}_{r'_1}\<i_{r'_1}|\otimes\cdots\otimes{}_{r'_{n'}}\<i_{r'_{n'}}|).
 \end{align*}
-The rules for $e_R$ (ket and bra) are similar.
+where $|i_R\>$ and $\<i_{R'}|$ is constructed by tensoring the basis according to the structure of $R$. The rules for $e_R$ (ket and bra) are similar.
 %This step reduces all labelled terms $e_R$ or $e_{R;R}$.
@@
 \paragraph*{2. Rewriting to the normal form.}
-We add three types of rules for dealing with operators on labeled terms, 1) recursively applying them to subterms, 2) push big operators out and 3) eliminate bra-ket pairs. For example,
+We add three types of rules for dealing with operators on labeled terms, 1) recursively applying them to subterms, 2) push big operators out and 3) eliminate generalized composition and bra-ket pairs.
 %Other symbols on labelled Dirac notation are also 
 Take the rule for conjugate $(D_1 \cdot D_2)^\dagger \ \reduce\ D_2^\dagger \cdot D_1^\dagger$ as an example of 1).
-For 2), 
-The final step operates on sum and dot product. They will lift summation to the outside, and eliminate the bra-ket pairs whenever possible.
+For 2), we use distributivity rules for scaling, labelled tensor and generalized composition. For example, rule
+\[
+% \textrm{(R-SUM-PUSHD0)}
+X_1 \otimes \cdots (\sum_{i \in M} D) \cdots \otimes X_2\ \reduce\ \sum_{i \in M} (X_1 \otimes \cdots D \cdots \otimes X_n)
+\]
+will lift summation to the outside.
+Extra rules for 3) are established including:
+% The final step operates on sum and dot product. They will lift summation to the outside, and eliminate the bra-ket pairs whenever possible.
 \begin{align*}
-    & \textrm{(R-SUM-PUSHD0)}
-    && X_1 \otimes \cdots (\sum_{i \in M} D) \cdots \otimes X_2\ \reduce\ \sum_{i \in M} (X_1 \otimes \cdots D \cdots \otimes X_n) \\
+    % & \textrm{(R-SUM-PUSHD0)}
+    % && X_1 \otimes \cdots (\sum_{i \in M} D) \cdots \otimes X_2\ \reduce\ \sum_{i \in M} (X_1 \otimes \cdots D \cdots \otimes X_n) \\
     % %
@@
 % These rules are added to the rewriting system in~\Cref{sec: decide} and executed together.
-In the end, if there are no variables of $\DType(s_1, s_2)$, the expression will be reduced to the addition of big operator sum, and each sum body is labelled basis with Dirac notation scalar coefficients:
-\[
+Assuming no variables of $\DType(s_1, s_2)$, 
+repeating the application of the above rules yield the normal form -- the addition of big operator sum: 
+\begin{equation}
+  \label{eqn:Labelled normal}
   \sum_{i}\cdots\sum_{j} a_1 . (\ket{i}_{p} \otimes \cdots \otimes \bra{j}_{q})
@@
   \sum_{k}\cdots\sum_{l} a_m . (\ket{k}_{r} \otimes \cdots \otimes \bra{l}_{s})
-\]
-In this stage, we only need to check the scalars to decide the equivalence of labelled Dirac notation.
+\end{equation}
+where each sum body is a tensor of labelled basis with Dirac notation scalar coefficients.
+\paragraph*{3. Ordering and elimination of quantum variables.} 
+We further sort tensors of labelled basis in every sum body of Eqn.
+(\ref{eqn:Labelled normal}) by 1.
+ket first and 2.
+the default order of variables.
+This yield the same shape of every subterm, e.g.,
+\begin{equation*}
+  % \label{eqn:Labelled normal}
+  \sum_{i}\cdots\sum_{j}\cdots \sum_{i'}\cdots\sum_{j'} a_1 . ((\ket{i}_p \otimes \cdots \otimes \ket{j}_q) \otimes (\bra{i'}_{p'}\otimes\cdots\otimes \bra{j'}_{q'}))
+\end{equation*}
+and thus it is equivalent to decide the equivalence of additions of subterms of:
+\begin{equation}
+  \label{eqn:Labelled normal1}
+  \sum_{i}\cdots\sum_{j}\cdots \sum_{i'}\cdots\sum_{j'} a_1 . ((\ket{i} \otimes \cdots \otimes \ket{j}) \otimes (\bra{i'}\otimes\cdots\otimes \bra{j'}))
+\end{equation}
+which do not involve any labels.
+
+\begin{theorem}[Soundness]
+  All rewriting rules for labelled Dirac notations listed here and in Appendix are sound, i.e., for any interpretation $\sem{\cdot}$, $D\reduce D'$ implies $\sem{D} = \sem{D'}$.
+\end{theorem}
+\begin{proof}
+  By checking each rules and its semantics.
+\end{proof}
+
+\begin{theorem}[Completeness for Dealing with Labels]
+  Above normalization procedure translates every well-typed equality of labelled Dirac notations to equality of Dirac notation.
+\end{theorem}
+\begin{proof}
+  It is sufficient to check if each step achieves its goal.
+Step 1 is trivial.
+For step 2, it can be further checked that if it is not in form (\ref{eqn:Labelled normal}), there must exists one of the rules can be further applied.
+  For step 3, realize that all rewriting rules perserve the types, i.e., $D\reduce D'$ then $E\vdash D : \cD(s,s')$ and $E\vdash D : \cD(s,s')$ for some $s,s'\subseteq \cR$, which ensure the same shape of every subterm after sorting.
+\end{proof}
+% , we claim that tensor of labelled basis in every sum body of Eqn.
+(\ref{eqn:Labelled normal}) are of the same type.
+As a consequence, by sort the labelled tensors according to a default order of variables, they
+
+% In this stage, we only need to check the scalars to decide the equivalence of labelled Dirac notation.
 

### 2025-01-30 02language_design.tex
alg moved to sec 3

@@
 \begin{lemma}
+    \label{lem: axiom sound}
     For all well-formed contexts \( E \) and terms \( e_1, e_2 \), if \( E \vdash e_1 = e_2 \), then \( \sem{e_1} = \sem{e_2} \).
@@
 
-With these syntax, typing and semantics introduced, we demonstrate the formalization of the symbols used in the motivating example to conclude this section.
-\begin{example}[Formalizing the Motivating Example]
+With these syntax, typing and semantics introduced, we can formalize~\Cref{eq: motivating plain}, the motivating~\Cref{ex: motivating} in plain Dirac notation.
+\begin{example}[Motivating Example Formalization]
     \label{ex: formalizing motivating}
-    Definitions and assumptions in the environment \( E \) are formalized as follows:
+    Definitions and assumptions in the context \( E \) are formalized as follows:
     \begin{align*}
@@
 \end{example}
+
+\subsection{Algorithm to Decide Equivalence}
+
+The equivalence of Dirac notations are decided by normalization, which transforms equivalent expressions under a set of axioms into the same syntax.
+We use an efficient algorithm to perform the normalization totally on $R \cup F$.
+
+\begin{enumerate}
+    \item \textbf{First Rewritings}: Expand definitions and simplify expressions.
+    \item \textbf{Variable Expansion}: Convert to abstract element-wise representation.
+    \item \textbf{Second Rewritings}: Normalize terms on \( R \) modulo \( F \).
+    \item \textbf{Sorting Without Bound Variables}: Normalize AC-equivalence.
+    \item \textbf{Swap Successive Summations}: Normalize SUM-SWAP equivalence.
+    \item \textbf{De Bruijn Normalization}: Normalize \( \alpha \)-equivalence.
+\end{enumerate}
+
+Here steps 1 - 3 are term rewriting for $R$. 
+Term rewriting reduces a term repeatedly using a set of rules in the form of $l\ \reduce\ r$.
+The reduction proceeds by matching the subterms with the left-hand side, and replacing it with the right-hand side.
+For example, the rule $a.K + K\ \reduce\ (a + 1).K$ matches the term $(x\times y) .
+\ket{t} + \ket{t}$ by the substitution $\{a \to  x\times y, K \to \ket{t}\}$, and is rewritten into $(x\times y + 1) .
+\ket{t}$.
+Step 1 and 3 uses the same set of rewriting rules, see~\Cref{sec: rewriting rules}.
+Step 2 expands variables like the ket $K$ to their abstract element-wise representation $\sum_{i} (\bra{i} \cdot K). \ket{i}$, which improves the decidability with sum.
+
+Steps 4 - 6 are special algorithms to further normalize the axiom set $F$.
+Here the main difficulty is the coexistence of AC-equivalence and SUM-SWAP, so that naive sorting will not convert equivalent terms to the same form.
+For step 4 and 5, the key observation is that in a successive sum expression \( \sum_{i \in s_1} \cdots \sum_{j \in s_n} A \), the names and order of the bound variables \( i, \dots, j \) can be freely permuted.
+Therefore we first ignore bound variables and normalize AC-equivalence by sorting.
+Then the order of summation can be decided accordingly.
+The last step converts functions to de Bruijn indices~\cite{deBruijn1972lambda} to decide $\alpha$-equivalence.
+For detailed descriptions, see~\Cref{sec: decide}.
+
+As a comparison, the previous work conducts normalization paritally on $R$, and decide axioms in $F$ by iterating through all possible permutations.
+Our algorithm totally normalize the term, as is illustrated below.
+\[
+\text{(partial)}\qquad e_1 \mathop{\twoheadrightarrow}^R e_1' \mathop{=}^F e_2' \mathop{\twoheadleftarrow}^R e_2
+\qquad
+\text{(total)}\qquad e_1 \mathop{\twoheadrightarrow}^{R \cup F} e \equiv e \mathop{\twoheadleftarrow}^{R \cup F} e_2
+\]
+While the permutation checking algorithm has factorial complexity on the number of AC symbol arguments, our algorithm uses normalization by sorting, and the complexity is polynomial on the term size.
+This improvement in efficiency is also confirmed by experiments.
+
+% Lastly, we can prove that the equivalence established by this normalization procedure is sound with respect to the semantics.
+
+% \begin{theorem}[Soundness]
+%     For any well-formed context \( E \) and well-typed expressions \( e_1 \) and \( e_2 \), if  $e_1$ and $e_2$ have the same normal form, then \( \sem{e_1} = \sem{e_2} \).
+% \end{theorem}
+% \begin{proof}
+%     Because of~\Cref{lem: axiom sound}, we only need to prove soundness with respect to axioms. 
+%     This is because the term rewriting procedure follows from the fact that each rewriting rule preserves equivalence.
+Furthermore, the operations in the sort and swap transformations respect the AC-equivalence and SUM-SWAP axioms.
+Finally, the de Bruijn normalization ensures soundness for \( \alpha \)-equivalence.
+% \end{proof}

### 2025-01-30 05tool_eval.tex
alg moved to sec 3

@@
 \section{Implementation and Case Study}
-The refinements and extensions above concludes in our implementation called D-Hammer, a solver written in \CC.
+The improvement and extensions above concludes in our implementation called D-Hammer, a solver written in \CC.
 It has a parser built by ANTLR4, and scalar reasonings are powered by a Mathematica kernel.
@@
 This implementation is tested on the benchmark of the previous work and new examples for labelled Dirac notation, demonstrating significant improvement in expressivity and efficiency. 
+The tool and example encodings will be open-source and public available.
 
@@
 
-\subsection{Equivalence Checking Algorithm}
-
-The equivalence of Dirac notations are decided by normalization, which transforms equivalent expressions under a set of axioms into the same syntax.
-We use an efficient algorithm to perform the normalization totally on $R \cup F$.
-
-\begin{enumerate}
-    \item \textbf{First Rewritings}: Expand definitions and simplify expressions.
-    \item \textbf{Variable Expansion}: Convert to abstract element-wise representation.
-    \item \textbf{Second Rewritings}: Normalize terms on \( R \) modulo \( F \).
-    \item \textbf{Sorting Without Bound Variables}: Normalize AC-equivalence.
-    \item \textbf{Swap Successive Summations}: Normalize SUM-SWAP equivalence.
-    \item \textbf{De Bruijn Normalization}: Normalize \( \alpha \)-equivalence.
-\end{enumerate}
-Here steps 1 - 3 are term rewriting for $R$, and steps 4 - 6 are special algorithms to further normalize the axiom set $F$.
-Step 2 expands variables like $K$ to their abstract element-wise representation $\sum_{i} (\bra{i} \cdot K).
-\ket{i}$, which improves the decidability with sum.
-For step 4 and 5, the key observation is that in a successive sum expression \( \sum_{i \in s_1} \cdots \sum_{j \in s_n} A \), the names and order of the bound variables \( i, \dots, j \) can be freely permuted.
-Therefore we first ignore bound variables and normalize AC-equivalence by sorting.
-Then decide the order of summation accordingly.
-The last step converts functions to de Bruijn indeices~\cite{deBruijn1972lambda} to decide $\alpha$-equivalence.
-
-As a comparison, the previous work conducts normalization paritally on $R$, and decide axioms in $F$ by iterating through all possible permutations.
-Our algorithm totally normalize the term and significantly improves the efficiency.
-\[
-\text{(partial)}\qquad e_1 \mathop{\twoheadrightarrow}^R e_1' \mathop{=}^F e_2' \mathop{\twoheadleftarrow}^R e_2
-\qquad
-\text{(total)}\qquad e_1 \mathop{\twoheadrightarrow}^{R \cup F} e \equiv e \mathop{\twoheadleftarrow}^{R \cup F} e_2
-\]
-
-\yx{we can add a normalization example here.}
-
-\subsection{Use Case}
-Here we encode the motivating~\Cref{ex: motivating}, examine and explain how to check it using D-Hammer. The encoding is shown below.
+\subsubsection{Use Case}
+As a tutorial, we encode the motivating~\Cref{ex: motivating}, examine and explain how to check it using D-Hammer. The encoding is shown below.
 
@@
 
-{
-    \center
+\begin{center}
     \begin{tabular}{c|c c c|c c c}
@@
         % research paper                & 4          & 4         &  59.53       &   4    & 4       &  0.73     \\
-        labelled Dirac notation      &   -         &   -          &     -           &      -      &        18       &     22.72    \\
+        % labelled Dirac notation      &   -         &   -          &     -           &      -      &        18       &     22.72    \\
         \hline
     \end{tabular}        
-}
+\end{center}
 
@@
 When the running time goes up, D-Hammer provides 2 to 40 times of acceleration in average.
-The relationship between efficiency improvement and example structure further supports the necessity of our algorithm. 
+The relationship between efficiency improvement and example characteristics further supports the necessity of our algorithm. 
 We find that examples with great efficiency improvement, e.g. COQQ-129 and COQQ-148 shown in~\Cref{fig: CoqQ plot}, tend to use deeply nested summations, where our algorithm has an effect.

### 2025-01-30 01intro_preliminary.tex
add figure

@@
 notation is motivated by the following example:
+
+% \begin{figure}[t]
+% \begin{quantikz}
+%   & \gate{X} & \ctrl{1} & \\
+%   & & \targ{} &
+%   \end{quantikz}
+%   =\begin{quantikz}
+%   & \ctrl{1} & \gate{X} & \\
+%   & \targ{} & \gate{X} &
+%   \end{quantikz}
+%   \end{figure}
+
 \begin{example}
@@
   \ccnot|\ghz\rangle &= (I\otimes \cnot)|\ghz\rangle \\
-  &= (\swap\otimes I)(I\otimes \cnot)(\swap\otimes I)|\ghz\rangle
+  &= (\Swap\otimes I)(I\otimes \cnot)(\Swap\otimes I)|\ghz\rangle
 \end{align*}
+which might be illustrated by the following circuit models.
+\begin{figure}[h]
+  % \begin{minipage}{0.48\linewidth}
+      \centering
+      \vspace{-0.5cm}
+      \begin{quantikz}[column sep=0.1em, row sep=0em]
+      \lstick{$p\qquad\qquad $} & \lstick{} &[0.2cm] & \ctrl{1} & [0.2cm] &  \quad \ \ \;\! \quad\qquad\quad & 
+      &[0.2cm] & & [0.2cm] & \quad \ \ \;\! \quad\qquad\quad & 
+      &[0.2cm] & \gate[2,swap,style={dashed}]{} & [0.2cm] & [0.2cm]\gate[2,swap,style={dashed}]{} & [0.2cm] & \\[-0.1cm]
+      \lstick{$r\qquad\qquad $} & \lstick{$|\ghz\>$} & & \ctrl{1} & & \quad = \quad\qquad\quad & \lstick{$|\ghz\>$} 
+      & & \ctrl{1} & & \quad = \quad\qquad\quad & \lstick{$|\ghz\>$}
+      & & & \ctrl{1} & & & \\
+      \lstick{$q\qquad\qquad $} & \lstick{} & & \targ{} & & \quad \ \ \;\! \quad\qquad\quad & 
+      & & \targ{} & & \quad \ \ \;\! \quad\qquad\quad &
+      & & & \targ{} & & & \\
+      \end{quantikz}
+      \vspace{-0.5cm}
+      % \vspace{-0.7em}
+      % \caption{Circuit computing the last bit of $a+(011)_2$.}
+      % \label{fig:addition}
+  %  \end{minipage}
+  \end{figure}
 Formalizing the statement in Dirac notation requires the following
@@
 \cnot)$.  For $\cnot$ acting on $p,q$, realize that $p,q$ is not
-adjacent in the chosen order, we additionally need to use $\swap$ to
+adjacent in the chosen order, we additionally need to use $\Swap$ to
 temporarily exchange the qubits $p$ and $r$, i.e., globally, the
-$(\swap\otimes I)$ before and after $(I\otimes \cnot)$ (lifting of
+$(\Swap\otimes I)$ before and after $(I\otimes \cnot)$ (lifting of
 $\cnot$ on $r,q$).
@@
 Roughly speaking, encoding in standard Dirac notation needs to
-tensoring identity operators and employing additional $\swap$ gates
+tensoring identity operators and employing additional $\Swap$ gates
 since generally the conventional order cannot guarantee 1. the order
@@
 This formalization avoids determining and maintaining the conventional
-order of qubits. nor lifting using additional $I$ and $\swap$s. In
+order of qubits, nor lifting using additional $I$ and $\Swap$s. In
 this setting, the tensor products become an associative and

### 2025-01-30 03labelled.tex
add figure

@@
 Formally, we equip $\cR$ with a default order, say for example the alphabetical order of names.
-For any valid register $R$, there exists the operator $\swap_R$ that sorts $R$; for example, if $E\vdash p : \reg(\sigma_p)$ and $E\vdash q : \reg(\sigma_q)$, then $\swap_{(q,p)} \triangleq \sum_{i\in \bU(\sigma_p)}\sum_{j\in \bU(\sigma_q)}|i\>\<j|\otimes |j\>\<i|$.
-We can further define $\swap_{s_1,s_2,\cdots}$ for merging disjoint sets orderly.
+For any valid register $R$, there exists the operator $\Swap_R$ that sorts $R$; for example, if $E\vdash p : \reg(\sigma_p)$ and $E\vdash q : \reg(\sigma_q)$, then $\Swap_{(q,p)} \triangleq \sum_{i\in \bU(\sigma_p)}\sum_{j\in \bU(\sigma_q)}|i\>\<j|\otimes |j\>\<i|$.
+We can further define $\Swap_{s_1,s_2,\cdots}$ for merging disjoint sets orderly.
 
@@
 \begin{itemize}
-  \item $\sem{|i\>_r} = |\sem{i}\>$; $\sem{K_R} = \swap_R \cdot \sem{K}$; $\sem{O_{R_1,R_2}} = \swap_{R_1}\cdot \sem{O}\cdot \swap_{R_2}$;
-  \item $\sem{D_1\otimes \cdots \otimes D_n} = \swap_{s_1,\cdot,s_n}\cdot (\sem{D_1}\otimes \cdots \otimes \sem{D_1})\cdot \swap_{s'_1,\cdot,s'_n}$;
+  \item $\sem{|i\>_r} = |\sem{i}\>$; $\sem{K_R} = \Swap_R \cdot \sem{K}$; $\sem{O_{R_1,R_2}} = \Swap_{R_1}\cdot \sem{O}\cdot \Swap_{R_2}$;
+  \item $\sem{D_1\otimes \cdots \otimes D_n} = \Swap_{s_1,\cdot,s_n}\cdot (\sem{D_1}\otimes \cdots \otimes \sem{D_1})\cdot \Swap_{s'_1,\cdot,s'_n}$;
   \item $\sem{D_1\cdot D_2} = \sem{\cl(D_1, s_2\backslash s'_1)}\cdot \sem{\cl(D_2, s'_1\backslash s_2)}$.

### 2025-01-31 01intro_preliminary.tex
minor

@@
 labelled Dirac notation a convenient, compositional formalism to
-reason about quantum states---in a way that is similar to the way
+reason about quantum states --- in a way that is similar to the way
 bunched logics support compositional reasoning about mutable states.
@@
 implicitly equational reasoning between labelled Dirac expressions, to
-glue applications of proof rules---similar to the use of the rule of
+glue applications of proof rules --- similar to the use of the rule of
 consequence in the setting of classical program. Therefore, it is
@@
 \section{Motivation and Preliminaries}
-\subsection{Standard Dirac Notation}
+\subsection{Plain Dirac Notation}
 
@@
 Labelled Dirac notation is a generalization of Dirac notation for
-describing multi-body quantum systems. The need for labelled Dirac
+describing many-body quantum systems. The need for labelled Dirac
 notation is motivated by the following example:
@@
 to tensoring it by an identity operator $I$ on $p$, i.e., $(I\otimes
-\cnot)$.  For $\cnot$ acting on $p,q$, realize that $p,q$ is not
+\cnot)$.  For $\cnot$ acting on $p,q$, realize that $p,q$ are not
 adjacent in the chosen order, we additionally need to use $\Swap$ to
@@
 
-Roughly speaking, encoding in standard Dirac notation needs to
+Roughly speaking, encoding in plain Dirac notation needs to
 tensoring identity operators and employing additional $\Swap$ gates
@@
 Dirac notation can conveniently describe local measurements, partial
-traces (the state or evolution of subsystems in multi-body systems),
+traces (the state or evolution of subsystems in many-body systems),
 and partial inner products (the representation of partial traces in
 pure states), which are sufficient to handle the mathematical formulas
-of quantum mechanics in multi-body systems.
+of quantum mechanics in many-body systems.
 
-Labelled Dirac notation is not only ubiquitous in the description of multi-body systems, but also plays an important role in program logics.
+Labelled Dirac notation is not only ubiquitous in the description of many-body systems, but also plays an important role in program logics.
 Just as in classical program logic, variable names are used to construct logical formulas instead of describing them as global memory functions, in quantum program logic, variable names are used to label the subsystems on which quantum gates act, rather than lifting them to the global system.

### 2025-01-31 03labelled.tex
minor

@@
 As discussed, this allows us to express and reason aboud the states and operations locally, without referring to the whole system. 
-We further demonstrate how to transform the equivalence problem into that for the Dirac notation studied above.
+We further demonstrate how to transform the equivalence problem into that for the plain Dirac notation studied above.
 
@@
 We first introduce the notation of quantum registers for structured variable combinations.
-This is necessary because, unlike assignments for classical variables, unitary transformations on composite systems -- a quantum version of assignments -- cannot generally be decomposed into separate unitary transformations on individual subsystems.
+This is necessary because, unlike assignments for classical variables, unitary transformations on composite systems --- a quantum version of assignments --- cannot generally be decomposed into separate unitary transformations on individual subsystems.
 Let $\cR$ be the set of quantum variables.
@@
 The no-cloning theorem, a fundamental property of quantum computing, prevents us from copying an unknown quantum state.
-This requires an additional check on the valid registers -- they should not include repeated quantum variables -- which is often handled in programming languages via linear types.
+This requires an additional check on the valid registers --- they should not include repeated quantum variables --- which is often handled in programming languages via linear types.
 Here, we start from the basic definition of \emph{order-free} variable set of a register:
@@
 \begin{definition}[Labelled Dirac Notation]
-  The \textbf{labelled Dirac notation} includes all Dirac notation symbols and the generators defined below.
+  The \textbf{labelled Dirac notation} includes all plain Dirac notation symbols and the generators defined below.
   Here, $s\subseteq \cR$ is a quantum variable set.
@@
 $\reg(\sigma)$ are types for registers $R$, and the index $\sigma$ indicates the type of Hilbert space represented by the register.
-It is allowed to lift a Dirac notation associate with corresponding quantum variables or registers, e.g., $\ket{i}_r$ and ${}_r\bra{i}$ are labelled basis, 
+It is allowed to lift a plain Dirac notation associate with corresponding quantum variables or registers, e.g., $\ket{i}_r$ and ${}_r\bra{i}$ are labelled basis, 
 $e_R$ for bra, ket and $e_{R;R}$ for operators. 
 We further introduce new $\cdot$ for generalized composition (unified for all kinds of multiplications between kets, bras and operators) and $\otimes$ for labelled tensor product since they do not share the same properties v.s.
-its counterpart in Dirac notation, i.e., generalized composition is not associative and labelled tensor is indeed an AC symbol.
+its counterpart in plain Dirac notation, i.e., generalized composition is not associative and labelled tensor is indeed an AC symbol.
 % In labelled Dirac notation, the structure of tensor product does not matter, therefore $\otimes$ is an AC symbol.
@@
 The first rule says that a paired register is of the product type and its components should be disjoint.
-Once we lift a Dirac notation to the labelled one (line 2), we enforce that the term and register are of the same indices, reflecting the fact the state should be consistant to the corresponding subsystems.
+To lift a plain Dirac notation into the labelled one (line 2), we enforce that the term and register are of the same indices, reflecting the fact the state should be consistant to the corresponding subsystems.
 The third line gives the typing of labelled tensoring with a check whether the component subsystems are disjoint with each other.
@@
 \subsection{Normalization}
-The normalization procedure of labelled Dirac notation allows us to eliminate all labels and thus translate the equality of labelled terms to the equality of standard Dirac terms.
+The normalization procedure of labelled Dirac notation allows us to eliminate all labels and thus translate the equality of labelled terms to the equality of plain Dirac terms.
 This is achieved by the following three steps:
@@
 Assuming no variables of $\DType(s_1, s_2)$, 
-repeating the application of the above rules yield the normal form -- the addition of big operator sum: 
+repeating the application of the above rules yield the normal form --- the addition of big operator sum: 
 \begin{equation}
@@
 \end{equation}
-where each sum body is a tensor of labelled basis with Dirac notation scalar coefficients.
+where each sum body is a tensor of labelled basis with scalar coefficients in plain Dirac notation.
 \paragraph*{3. Ordering and elimination of quantum variables.} 
@@
 \begin{theorem}[Completeness for Dealing with Labels]
-  Above normalization procedure translates every well-typed equality of labelled Dirac notations to equality of Dirac notation.
+  Above normalization procedure translates every well-typed equality of labelled Dirac notations to equality of plain Dirac notation.
 \end{theorem}

### 2025-01-31 01intro_preliminary.tex
minor

@@
 labelled Dirac notation a convenient, compositional formalism to
-reason about quantum states --- in a way that is similar to the way
+reason about quantum states--in a way that is similar to the way
 bunched logics support compositional reasoning about mutable states.
@@
 about quantum programs. Specifically, many quantum Hoare logics use
-labelled Dirac notation to express program assertions~\cite{DBLP:conf/lics/ZhouBHYY21,Zhou2023}. These logics also use
+labelled Dirac notation and its variants to express program assertions (see for example~\cite{DBLP:conf/lics/ZhouBHYY21,Zhou2023,qRHL_unruh2019,QSL_Le_2022,Zhong2024-np,incorrectness_2022,qafny2024}).
+These logics also use
 implicitly equational reasoning between labelled Dirac expressions, to
-glue applications of proof rules --- similar to the use of the rule of
+glue applications of proof rules--similar to the use of the rule of
 consequence in the setting of classical program. Therefore, it is

### 2025-01-31 03labelled.tex
minor

@@
 We first introduce the notation of quantum registers for structured variable combinations.
-This is necessary because, unlike assignments for classical variables, unitary transformations on composite systems --- a quantum version of assignments --- cannot generally be decomposed into separate unitary transformations on individual subsystems.
+This is necessary because, unlike assignments for classical variables, unitary transformations on composite systems--a quantum version of assignments--cannot generally be decomposed into separate unitary transformations on individual subsystems.
 Let $\cR$ be the set of quantum variables.
@@
 The no-cloning theorem, a fundamental property of quantum computing, prevents us from copying an unknown quantum state.
-This requires an additional check on the valid registers --- they should not include repeated quantum variables --- which is often handled in programming languages via linear types.
+This requires an additional check on the valid registers--they should not include repeated quantum variables--which is often handled in programming languages via linear types.
 Here, we start from the basic definition of \emph{order-free} variable set of a register:
@@
 Assuming no variables of $\DType(s_1, s_2)$, 
-repeating the application of the above rules yield the normal form --- the addition of big operator sum: 
+repeating the application of the above rules yield the normal form--the addition of big operator sum: 
 \begin{equation}

### 2025-01-31 02language_design.tex
.

@@
-\section{Language, Typing and Semantics}
-We use a typed abstract syntax to represent Dirac notation. The overall system design leads to the following advantages. 
+\section{Dirac notation}
+This section introduces the language of Dirac notation, its
+denotational and axiomatic semantics, and describes D-Hammer approach
+to equational reasoning. Two main ingredients of our language are:
 \begin{itemize}
-    \item Dirac notation can be divided into four types, namely scalar, ket, bra and operator.
-Many function symbols have constraints on types.
-It is important to have a typing system and check the validity of expressions.
-    \item It has functions and higher-order typing. It allows to formally encode defined symbols like transpose or trace, which are usually used to reprensent the term in an abstract manner. 
-    \item The structures are friendly to optimizations for efficiency. For example, associative and commutative (AC) symbols have indefinite arities, which enables normalization by sorting.
+\item a rich typing discipline that distinguishes between scalars,
+  kets, bras and operators, but supports sufficient overloading to
+  remain close to standard Dirac notation;
+\item higher-order, indexed (a.k.a.\, weak dependent) types. It allows
+  to formally encode defined symbols like transpose or trace, which
+  are usually used to represent the term in an abstract manner;
+\item operators with indefinite arities. Indefinite arities are
+  instrumental for reasoning efficiently about associative and
+  commutative (AC) symbols have indefinite arities, as they enable
+  normalization by sorting.
 \end{itemize}
 
-Since a Hilbert space $\mathcal{H}_V$ is dependent on the basis set $V$,
-types for Dirac notation also depends on the type index.
-Therefore, the language is organized into three layers: the index, the type, and the term. 
-Terms represent concrete instances such as kets, bras, and operators, which will be typed and checked.
-The index represents classical data types and appears in type expressions to differentiate between various Hilbert spaces and sets.
+\subsection{Language}
+Since a Hilbert space $\mathcal{H}_V$ is dependent on the basis set
+$V$, types for Dirac notation also depends on the type index.
+Therefore, the language is organized into three layers: the index, the
+type, and the term.  Terms represent concrete instances such as kets,
+bras, and operators, which will be typed and checked. The index
+represents classical data types and appears in type expressions to
+differentiate between various Hilbert spaces and sets.
 
@@
 \subsubsection{Denotational Semantics}
-The denotational semantics interpret every expression as an object in linear algebra, according to a valuation mapping \( v \), which assigns values to the variables in the expression.
-The semantics of an expression \( e \) with a given valuation \( v \) is denoted as \( \sem{e}_v \).
-Two expressions \( e_1 \) and \( e_2 \) are considered equivalent if their semantics are equal for all valuations, i.e., \( \sem{e_1}_v = \sem{e_2}_v \) for all \( v \).
+The denotational semantics interpret every expression as an object in
+linear algebra, according to a valuation mapping \( v \), which
+assigns values to the variables in the expression. Concretely, we
+first define an interpretation $\sem{T}$ for each type.{\textcolor{red}{TODO}}
+
+The semantics of an expression \( e \) with a given valuation \( v \)
+is denoted as \( \sem{e}_v \). Two expressions \( e_1 \) and \( e_2 \)
+are semantically equivalent if their semantics are equal for all
+valuations, i.e., \( \sem{e_1}_v = \sem{e_2}_v \) for all \( v \).
 

### 2025-01-31 01intro_preliminary.tex
update section 3

@@
 
-The next sections introduce the formal language, labelled Dirac notation and the systematical way to reduce labels and prove equalities. 
-For this purpose, we use the following motivating example as the main role in our explanations.
+In the following,we use the motivating example as the main role in our demonstrations.
 Instead of GHZ states, it is a similar but simpler example on Bell states:
@@
 \end{example}
-Following the introduction on labels, we can consider the global system $(q,r)$ and transform the equation above into the plain Dirac notation:
+Following the explanation on labels, we can consider the global system $(q,r)$ and transform the equation above into the plain Dirac notation:
 \begin{align}
@@
 \end{align}
-Together with corresponding explanations for this example, we show how an automated system is built and used to solve equalities likewise.
+The next sections introduce the formal language, labelled Dirac notation and the systematical way to reduce labels.
+We show how an automated system is built and used to solve equalities likewise.
 

### 2025-01-31 02language_design.tex
update section 3

@@
 
+\Cref{fig: normalization demo} shows the normalization outline for $(M \otimes \mathbf{1}_\mathcal{O}(T)) \cdot (\text{phi}\ T)$.
+
+\begin{figure}
+    \scriptsize
+    \begin{align*}
+        & (\eqnmarkbox[blue]{mvar}{M} \otimes \eqnmarkbox[red]{identity}{\mathbf{1}_\mathcal{O}(T)})(\eqnmarkbox[red]{phi-definition}{\textrm{phi}\ T})
+        \Rightarrow
+        (\sum_{k,l\in T} \bra{k} M \ket{l}.\ket{k}\bra{l}
+        \eqnmarkbox[red]{sumpush1}{\otimes \sum_{i \in T}} \ket{i}\bra{i})(\eqnmarkbox[red]{sumpush2}{\sum_{j\in T}}\ket{j, j}) 
+        \\
+        & \Rightarrow
+        \sum_{i,j,k,l\in T} \bra{k} M \ket{l} . \ket{k, i}\eqnmarkbox[red]{braket-dot}{\bra{l, i} \cdot \ket{j,j}}
+        \\
+        & \Rightarrow
+        \eqnmarkbox[red]{delta-elim-sum}{\sum_{i,j,k,l\in T}} (\eqnmarkbox[red]{delta-elim-delta}{\delta_{l,j} \times \delta_{i,j}} \times \bra{k} M \ket{l}) . \ket{k, i}
+        \Rightarrow
+        \eqnmarkbox[brown]{swapsum}{\sum_{j, k \in T}} \bra{k} M \ket{j} . \ket{k, j}
+        \\[1em]
+        & \Rightarrow
+        \sum_{k, j \in T} \bra{k} M \ket{j} . \ket{k, j}
+        \ \eqnmarkbox[brown]{deBruijn}{\xRightarrow{\text{de Bruijn}}}\ 
+        \sum_T\sum_T\bra{\$1} M \ket{\$0} . \ket{\$1, \$0}
+    \end{align*}
+    \annotate[yshift=1em]{right}{mvar}{\scriptsize{variable expansion}}
+    \annotate[yshift=0.5em, xshift=1cm]{right}{identity}{\scriptsize{identity operator}}
+    \annotate[yshift=0em]{below,left}{phi-definition}{\scriptsize{definition}}
+    \annotatetwo[yshift=0em]{below}{sumpush1}{sumpush2}{\scriptsize{sum lifting}}
+    \annotate[yshift=0em]{below}{braket-dot}{\scriptsize{inner product}}
+    \annotatetwo[yshift=0em,xshift=2cm]{below}{delta-elim-sum}{delta-elim-delta}{\scriptsize{$\delta$-elimination}}
+    \annotate[yshift=0em]{below}{swapsum}{\scriptsize{sum swapping}}
+    \caption{A nomralization outline for the left-hand side of~\Cref{eq: motivating plain}.
+Matched subterms are marked with colors.
+Blue marking represents variable expansion, red marking represents rule applications, and brown marking represents normalization step 4-6.}
+    \label{fig: normalization demo}
+\end{figure}
+
+% \begin{gather*}
+%     \eqnmarkbox[blue]{node1}{GEBV} = \sum_{i}^{n} \eqnmarkbox[red]{node2}{X_{i}} b_{i} \\
+%     \eqnmarkbox[green]{node3}{ssssss}
+% \end{gather*}
+
+%     \annotate[yshift=1em]{left}{node1}{Genomic Estimated Breeding Value}
+%     \annotate[yshift=-1em]{below}{node2}{Genotype at locus i}
+%     \annotatetwo[yshift=2em]{above}{node2}{node3}{var 1}
+
 % Lastly, we can prove that the equivalence established by this normalization procedure is sound with respect to the semantics.

### 2025-01-31 01intro_preliminary.tex
minor

@@
 examples from the literature. Our evaluation covers examples based on
-(vanilla, unlabelled) Dirac notation, as well as examples based on
+plain Dirac notation, as well as examples based on
 labelled Dirac notation. The main conclusions are that our approach
@@
 Labelled Dirac notation is a generalization of Dirac notation for
-describing many-body quantum systems. The need for labelled Dirac
-notation is motivated by the following example:
+describing many-body quantum systems. The following example shows the necessity of labels:
 
@@
 order of qubits, nor lifting using additional $I$ and $\Swap$s. In
-this setting, the tensor products become an associative and
-commutative (AC) symbol, so we can rearrange qubits as needed to
+this setting, the tensor products become associative and
+commutative, so we can rearrange qubits as needed to
 perform calculations. For our example, we can perform the calculation as
@@
 \end{align}
-The next sections introduce the formal language, labelled Dirac notation and the systematical way to reduce labels.
-We show how an automated system is built and used to solve equalities likewise.
+The next sections introduce the formal language, labelled Dirac notation and the systematical way to reduce labels, showing how an automated system is built and used to solve equalities likewise.
 

### 2025-01-31 03labelled.tex
minor

@@
 In this section, we extend the language by allowing quantum variables to indicate the quantum system of vectors and operators it act. 
-As discussed, this allows us to express and reason aboud the states and operations locally, without referring to the whole system. 
+As discussed, this allows us to express and reason about the states and operations locally, without referring to the whole system. 
 We further demonstrate how to transform the equivalence problem into that for the plain Dirac notation studied above.
@@
 
-\paragraph*{Semantics.} While the labelled Dirac notation avoids unnecessary lifting and ordering, its semantics faithfully capture these details.
+\paragraph*{Semantics.} The labelled Dirac notation handles lifting and ordering for us, and its semantics faithfully capture these details.
 The key points are 1.

### 2025-01-31 05tool_eval.tex
minor

@@
 They are connected by a white space, which is parsed into the composition of Dirac notation, and will be reduced into the operator-ket multiplication after typing.
-The right hand side is interpreted similarly, except the defined symbol \texttt{TPO} in the environment:
+The right hand side is interpreted similarly, except the defined symbol \texttt{TPO} in the context:
 
@@
 The \texttt{TPO} symbol represents the transpose of operators, and encodes the formalization in~\Cref{ex: formalizing motivating}.
-Thanks to the design of environment and functions, many other commonly used symbols in Dirac notation are encoded and provided as defined symbols in D-Hammer.
+Other commonly used concepts in Dirac notation are encoded and provided as defined symbols in D-Hammer.
 

### 2025-01-31 06related_conclusion.tex
minor
related work on tree automata

@@
 of quantum states~\cite{amy2018towards,amy2023complete}.
+
+Automata-based approaches has also been successfully applied for reasoning about quantum circuit~\cite{AutoQ_pldi_2023}.
+Instead of representing a set of  $n$-qubit states in exponential size, \cite{AutoQ_pldi_2023} encode it as (finite) tree automata with much less (often $\mathit{poly}(n)$) transitions.
+They further established transformers to implement the semantics of quantum gates over this representation, providing better scalability compared to other approaches.
+It is developed as a practical tool \textsc{autoq} with the capabilities for symbolic reasoning~\cite{AutoQ2023}.
+More recently, \cite{AutoQ_popl2025} further explore level-synchronized tree automata to encoding state, a promising model for parameterized verification.
 

### 2025-01-31 02language_design.tex
minor
.

@@
 denotational and axiomatic semantics, and describes D-Hammer approach
-to equational reasoning. Two main ingredients of our language are:
+to equational reasoning. Three main ingredients of our language are:
 \begin{itemize}
@@
 \end{definition}
-\( \Basis(\sigma) \) denotes the type for basis elements in the index \( \sigma \).
-\( \SType \) represents scalars, while \( \KType(\sigma) \) and \( \BType(\sigma) \) refer to ket and bra types in the Hilbert space \( \sigma \), respectively.
-\( \OType(\sigma_1, \sigma_2) \) represents linear operators with \( \sigma_2 \) as the domain and \( \sigma_1 \) as the codomain.
-\( \SET(\sigma) \) refers to the type of subsets of \( \sigma \), used to denote the values of bound variables in summations.
-The remaining two constructs define function types: \( T_1 \to T_2 \) represents normal functions that take a \( T_1 \)-type argument and return a \( T_2 \)-type term, while \( \forall x.
-T \) represents index functions that take an index argument \( x \) and produce a \( T \)-type term, which may depend on \( x \).
-Index functions are essential for defining polymorphic transformations across Hilbert spaces.
+\( \Basis(\sigma) \) denotes the type for basis elements in the index
+\( \sigma \).  \( \SType \) represents scalars, while \(
+\KType(\sigma) \) and \( \BType(\sigma) \) refer to ket and bra types
+in the Hilbert space \( \sigma \), respectively.  \( \OType(\sigma_1,
+\sigma_2) \) represents linear operators with \( \sigma_2 \) as the
+domain and \( \sigma_1 \) as the codomain.  \( \SET(\sigma) \) refers
+to the type of subsets of \( \sigma \), used to denote the values of
+bound variables in summations.  The remaining two constructs define
+function types: \( T_1 \to T_2 \) represent the set of functions that
+take a \( T_1 \)-type argument and return a \( T_2 \)-type term, while
+\( \forall x. T \) represents the dependently typed functions that
+take an index argument \( x :\Index \) and produce a \( T \)-type
+term, where $T$ may depend on \( x \).  Index functions are essential
+for defining polymorphic transformations over Hilbert spaces.
 
@@
 
-The scalar multiplication $\times$ and addition $+$ are AC symbols, and they have indefinite arity. 
-Because we use abstract syntax, the Dirac functions have different interpertations for different types. 
-We use letters like $a, b, c$ to represent scalar variables, $K$ and $B$ to represent ket and bra variables, and $O$ for operators.
-Therefore, $O \cdot K$ is interpreted as the operator-ket multiplication, and scalars can also be constructed from inner products $B \cdot K$.
+The scalar multiplication $\times$ and addition $+$ are AC symbols,
+and they have indefinite arity.  We use letters like $a, b, c$ to
+represent scalar variables, $K$ and $B$ to represent ket and bra
+variables, and $O$ for operators.  Therefore, $O \cdot K$ is
+interpreted as the operator-ket multiplication, and scalars can also
+be constructed from inner products $B \cdot K$.
 
@@
 \subsubsection{Denotational Semantics}
-The denotational semantics interpret every expression as an object in
-linear algebra, according to a valuation mapping \( v \), which
-assigns values to the variables in the expression. Concretely, we
-first define an interpretation $\sem{T}$ for each type.{\textcolor{red}{TODO}}
+The denotational semantics maps types to sets, and expressions and
+indices to values in the interpretation of types and indices
+respectively. As for other dependently typed systems, all
+interpretations are parametrized by a valuation mapping \( v \), which
+assigns values to variables and indices. We let \( \sem{e}_v \) denote
+the interpretation of an expression $e$ w.r.t.\, a valuation \( v \);
+we use similar notations for types and indices. As usual, we say that
+a valuation \( v \) is valid w.r.t.\, a context $\Gamma$ if for every
+variable declaration $x:T$, we have \( \sem{x}_v \in\sem{T}_v \) and
+for every definition $x:= t: T$, we have  \( \sem{x}_v=\sem{t}_v \).
 
-The semantics of an expression \( e \) with a given valuation \( v \)
-is denoted as \( \sem{e}_v \). Two expressions \( e_1 \) and \( e_2 \)
-are semantically equivalent if their semantics are equal for all
-valuations, i.e., \( \sem{e_1}_v = \sem{e_2}_v \) for all \( v \).
-
-The complete interpretation of terms and types is provided in~\Cref{sec: full denotational sem}.
-Variables typed with \( \Index \) are interpreted as finite sets, and the product of two indices \( \sem{\sigma_1 \times \sigma_2} \) is defined as the Cartesian product of the sets \( \sem{\sigma_1} \) and \( \sem{\sigma_2} \).
-More generally, each type is interpreted as a set.
-For example, the scalar type \( \sem{\SType} \) is interpreted as the set of complex numbers \( \mathbb{C} \), and the ket and bra types \( \sem{\KType(\sigma)} \) and \( \sem{\BType(\sigma)} \) are interpreted as the Hilbert space \( \mathcal{H}_{\sem{\sigma}} \) and its dual \( \mathcal{H}_{\sem{\sigma}}^* \), respectively.
-Terms are explained as the set elements.
-For example, the semantics of ket tensor product $\sem{K_1 \otimes K_2} \equiv \sem{K_1} \otimes \sem{K_2}$, is obtained by first calculating the semantics $\sem{K_1}$ and $\sem{K_2}$ as vectors, and then take the vector tensor product as result.
+In more detail, variables typed with \( \Index \) are interpreted as
+finite sets, and the product of two indices \( \sem{\sigma_1 \times
+  \sigma_2} \) is defined as the Cartesian product of the sets \(
+\sem{\sigma_1} \) and \( \sem{\sigma_2} \). More generally, each type
+is interpreted as a set. For example, the scalar type \( \sem{\SType}
+\) is interpreted as the set of complex numbers \( \mathbb{C} \), and
+the ket and bra types \( \sem{\KType(\sigma)} \) and \(
+\sem{\BType(\sigma)} \) are interpreted as the Hilbert space \(
+\mathcal{H}_{\sem{\sigma}} \) and its dual \(
+\mathcal{H}_{\sem{\sigma}}^* \), respectively. Terms are explained as
+the set elements. For example, the semantics of ket tensor product
+$\sem{K_1 \otimes K_2} \equiv \sem{K_1} \otimes \sem{K_2}$, is
+obtained by first calculating the semantics $\sem{K_1}$ and
+$\sem{K_2}$ as vectors, and then take the vector tensor product as
+result.  The complete interpretation of terms and types is provided
+in~\Cref{sec: full denotational sem}.
 
@@
 
-The idea behind the interpretation of types and terms is to formalize the typing relation using set-theoretic inclusion.
-Specifically, for a well-formed context \( E \), term \( t \), and type \( T \), if \( E \vdash t : T \), then for any valuation \( v \), the semantics of \( t \) must belong to the semantics of \( T \).
+The type system is sound w.r.t.\, the denotational semantics of
+expressions.
+Specifically, for a well-formed context \( E \), term \( t \), and type \( T \), if \( E \vdash t : T \), then for any valuation \( v \) that is valid for $\Gamma$, the interpretation of \( t \)  w.r.t.\, $v$ is an element of the interpretation of \( T \) w.r.t.\, $v$.
 
-\begin{lemma}
+\begin{lemma}[Soundness of type system]
     For any well-formed context \( E \), term \( t \), and type \( T \), if \( E \vdash t : T \), then for all valuations \( v \), \( \sem{t}_v \in \sem{T}_v \).
 \end{lemma}
-
 \begin{proof}
-    The proof follows directly by checking each case.
+By induction on the derivation.
 \end{proof}
-
-This interpretation formalizes the standard understanding of Dirac notation and provides the foundation for the algorithm.
-However, computers cannot directly reason about equivalence through mathematical interpretations.
-We proceed by defining a proof system that abstracts these concepts.
+This interpretation formalizes the standard understanding of Dirac
+notation and provides the foundation for the algorithm. However,
+computers cannot directly reason about equivalence through
+mathematical interpretations. We proceed by defining a proof system
+that abstracts these concepts.
 
@@
 \end{align*}
-The entire axioms are separated into $R$ and $F$ two sets.
+The entire axioms are separated into  two sets $R$ and $F$.
 During equivalence checking, most of the axioms are decided by term rewriting, which form the set $R$. However, there are axioms requiring special algorithms, which are collected in the set $F$.
@@
 \begin{align*}
-    \text{AC-equivalence} &\quad \text{e.g.,} \quad X + Y = Y + X, \quad (X + Y) + Z = X + (Y + Z), \\
-    \alpha\text{-equivalence} &\quad \lambda x . A = \lambda y . A\{x/y\}, \\
-    \text{SUM-SWAP} &\quad \sum_{i \in s_1} \sum_{j \in s_2} A = \sum_{j \in s_2} \sum_{i \in s_1} A, \\
-    \text{scalar theories} &\quad \text{e.g.,} \quad a + 0 = a, \quad a \times (b + c) = a \times b + a \times c.
+    \textup{(AC-equivalence)} &\ \text{e.g.,} \quad X + Y = Y + X, \quad (X + Y) + Z = X + (Y + Z), \\
+    \textup{($\alpha$-equivalence)} &\ \lambda x . A = \lambda y . A\{x/y\},
+    \quad
+    \textup{(SUM-SWAP)} \ \sum_{i \in s_1} \sum_{j \in s_2} A = \sum_{j \in s_2} \sum_{i \in s_1} A, \\
+    \textup{(scalar theories)} &\ \text{e.g.,} \quad a + 0 = a, \quad a \times (b + c) = a \times b + a \times c.
 \end{align*}
@@
 
-The scalar theories are treated separately as a module and are not considered in this work. In the implementation, we use the Mathematica kernel to decide scalar equivalences.
+\gb{these two paragraphs do not make too much sense.
+We also need to define \( E \vdash e_1 = e_2 \)}The scalar theories are treated separately as a module and are not considered in this work.
+In the implementation, we use the Mathematica kernel to decide scalar equivalences.
 
@@
 Denotational semantics can be seen as one model for this theory, meaning that equivalences derived from the axioms always imply equivalence in the interpretations.
-\begin{lemma}
-    \label{lem: axiom sound}
-    For all well-formed contexts \( E \) and terms \( e_1, e_2 \), if \( E \vdash e_1 = e_2 \), then \( \sem{e_1} = \sem{e_2} \).
+
+An equation \( \Gamma \vdash e_1 = e_2 :T \) is valid, written \(
+\Gamma \vDash e_1 = e_2 :T \), if \( \sem{e_1}_v = \sem{e_2}_v \) for
+all valuations \( v \) that are valid w.r.t.\, \( \Gamma \).
+
+\begin{lemma}[Soundness of equational theory]\label{lem: axiom sound}
+If \( E \vdash e_1 = e_2 :T \) then \( E \vDash e_1 = e_2\).
 \end{lemma}
 \begin{proof}
-    This follows directly from checking all cases.
+By induction on the structure of the derivation
 \end{proof}
-
-With these syntax, typing and semantics introduced, we can formalize~\Cref{eq: motivating plain}, the motivating~\Cref{ex: motivating} in plain Dirac notation.
+Next, we formalize the motivating example~\Cref{ex: motivating} in
+plain Dirac notation.
 \begin{example}[Motivating Example Formalization]
@@
 
-As a comparison, the previous work conducts normalization paritally on $R$, and decide axioms in $F$ by iterating through all possible permutations.
-Our algorithm totally normalize the term, as is illustrated below.
-\[
-\text{(partial)}\qquad e_1 \mathop{\twoheadrightarrow}^R e_1' \mathop{=}^F e_2' \mathop{\twoheadleftarrow}^R e_2
-\qquad
-\text{(total)}\qquad e_1 \mathop{\twoheadrightarrow}^{R \cup F} e \equiv e \mathop{\twoheadleftarrow}^{R \cup F} e_2
-\]
-While the permutation checking algorithm has factorial complexity on the number of AC symbol arguments, our algorithm uses normalization by sorting, and the complexity is polynomial on the term size.
-This improvement in efficiency is also confirmed by experiments.
-
 \Cref{fig: normalization demo} shows the normalization outline for $(M \otimes \mathbf{1}_\mathcal{O}(T)) \cdot (\text{phi}\ T)$.
-
-\begin{figure}
+\begin{figure}[h]
     \scriptsize
@@
 
+As a comparison, the previous work conducts normalization paritally on $R$, and decide axioms in $F$ by iterating through all possible permutations.
+Our algorithm totally normalize the term, as is illustrated below.
+\[
+\text{(partial)}\qquad e_1 \mathop{\twoheadrightarrow}^R e_1' \mathop{=}^F e_2' \mathop{\twoheadleftarrow}^R e_2
+\qquad
+\text{(total)}\qquad e_1 \mathop{\twoheadrightarrow}^{R \cup F} e \equiv e \mathop{\twoheadleftarrow}^{R \cup F} e_2
+\]
+While the permutation checking algorithm has factorial complexity on the number of AC symbol arguments, our algorithm uses normalization by sorting, and the complexity is polynomial on the term size.
+This improvement in efficiency is also confirmed by experiments.
+
 % \begin{gather*}

### 2025-01-31 02language_design.tex
.

@@
 \gb{these two paragraphs do not make too much sense.
-We also need to define \( E \vdash e_1 = e_2 \)}The scalar theories are treated separately as a module and are not considered in this work.
+We also need to define what is an equation and what is the meaning of \( E \vdash e_1 = e_2 \)}The scalar theories are treated separately as a module and are not considered in this work.
 In the implementation, we use the Mathematica kernel to decide scalar equivalences.
@@
 \subsection{Algorithm to Decide Equivalence}
+\gb{I still think that the names first rewritings, second rewriting and de bruijn normalization are not good, in particular, de bruijn normalization is non-standard terminology that will make people familiar with the field cringe.}
 

### 2025-01-31 02language_design.tex
.

@@
 \begin{lemma}[Soundness of type system]
-    For any well-formed context \( E \), term \( t \), and type \( T \), if \( E \vdash t : T \), then for all valuations \( v \), \( \sem{t}_v \in \sem{T}_v \).
+  If \( E \vdash t : T \), then for all valuations \( v \) valid w.r.t.\,
+  $\Gamma$, we have \( \sem{t}_v \in \sem{T}_v \).
 \end{lemma}

### 2025-01-31 03labelled.tex
update Sec. 4

@@
 This requires an additional check on the valid registers--they should not include repeated quantum variables--which is often handled in programming languages via linear types.
-Here, we start from the basic definition of \emph{order-free} variable set of a register:
-\begin{definition}
-The variable set of a register is defined inductively by:
-\begin{itemize}
-    \item $\var(R) = \{r\}$, if $R\equiv r$; or
-    \item $\var(R) = \var(R_1) \cup\var(R_2)$, if $R\equiv (R_1,R_2)$.
-\end{itemize}
-\end{definition}
+As such, we define the \emph{order-free} variable set of a register inductively by: 1) $\var(R) \triangleq \{r\}$, if $R\equiv r$ and 2) $\var(R) \triangleq \var(R_1) \cup\var(R_2)$, if $R\equiv (R_1,R_2)$.
+We use the variable set to establish side conditions of typing valid registers and employ it as the type parameter for annotating labelled Dirac terms.
+
+% \begin{definition}
+% The variable set of a register is defined inductively by:
+% \begin{itemize}
+%     \item $\var(R) = \{r\}$, if $R\equiv r$; or
+%     \item $\var(R) = \var(R_1) \cup\var(R_2)$, if $R\equiv (R_1,R_2)$.
+% \end{itemize}
+% \end{definition}
 
@@
 $\DType(s_1, s_2)$ is the unified type for all labelled Dirac notation, where $s_1$ indicates the codomain systems and $s_2$ indicates the domain systems.
+Roughly speaking, we define Hilbert space $\cH_s\triangleq\bigotimes_{r\in s}\cH_r$ for each set $s$, so that $\cH_{\emptyset}$ is a one-dimensional space isomorphic to complex numbers.
+Then the function view of ket and bra~\cite{Zhou2023} provides an alternative way, i.e., a ket on subsystem $s$ as a linear map from $\cH_{\emptyset}$ to $\cH_s$, a bra as a linear map from $\cH_s$ to $\cH_{\emptyset}$, to unify the type of kets, bras, and operators.
 For instance, labelled ket $|i\>_r$ has type $\DType(\{r\}, \emptyset)$, and labelled bra ${}_r\<i|$ has type $\DType(\emptyset, \{r\})$.
 $\reg(\sigma)$ are types for registers $R$, and the index $\sigma$ indicates the type of Hilbert space represented by the register.
-It is allowed to lift a plain Dirac notation associate with corresponding quantum variables or registers, e.g., $\ket{i}_r$ and ${}_r\bra{i}$ are labelled basis, 
-$e_R$ for bra, ket and $e_{R;R}$ for operators. 
+It is allowed to lift a plain Dirac notation associate with corresponding quantum variables or registers, e.g., $\ket{i}_r$ and ${}_r\bra{i}$ are labelled basis, $e_R$ for bra, ket, and $e_{R;R'}$ for operators which additionally allows different domain $R'$ and codomain $R$.
 We further introduce new $\cdot$ for generalized composition (unified for all kinds of multiplications between kets, bras and operators) and $\otimes$ for labelled tensor product since they do not share the same properties v.s.
@@
 
-For any interpretation $\sem{\cdot}$, and write $\cH_{\sem{s}} \triangleq \bigotimes_{r\in s}\cH_{\sem{\sigma_r}}$ for any $s\subseteq \cR$, where $E\vdash r : \reg(\sigma_r)$ for all $r\in\cR$.
-We interpret labelled Dirac notations inductively as (see details in Appendix) : assume $E\vdash D_i : \cD(s_i,s'_i)$ in the following,
+For any valuation $v$, and write $\cH_{\sem{s}_v} \triangleq \bigotimes_{r\in s}\cH_{\sem{\sigma_r}_v}$ for any $s\subseteq \cR$, where $E\vdash r : \reg(\sigma_r)$ for all $r\in\cR$.
+We interpret labelled Dirac notations inductively as: assume $E\vdash D_i : \cD(s_i,s'_i)$ in the following,
 \begin{itemize}
-  \item $\sem{|i\>_r} = |\sem{i}\>$; $\sem{K_R} = \Swap_R \cdot \sem{K}$; $\sem{O_{R_1,R_2}} = \Swap_{R_1}\cdot \sem{O}\cdot \Swap_{R_2}$;
-  \item $\sem{D_1\otimes \cdots \otimes D_n} = \Swap_{s_1,\cdot,s_n}\cdot (\sem{D_1}\otimes \cdots \otimes \sem{D_1})\cdot \Swap_{s'_1,\cdot,s'_n}$;
-  \item $\sem{D_1\cdot D_2} = \sem{\cl(D_1, s_2\backslash s'_1)}\cdot \sem{\cl(D_2, s'_1\backslash s_2)}$.
+  \item $\sem{|i\>_r}_v = |\sem{i}_v\>$;\quad 
+        $\sem{\<i|_r}_v = \<\sem{i}_v|$;\quad  
+        $\sem{K_R}_v = \sem{\Swap_R}_v \cdot \sem{K}_v$; \\
+        $\sem{B_R}_v = \sem{B}_v\cdot \sem{\Swap_R}_v$;\quad
+        $\sem{O_{R_1,R_2}} = \sem{\Swap_{R_1}}_v\cdot \sem{O}_v\cdot \sem{\Swap_{R_2}}_v$;
+  \item $\sem{D_1\otimes \cdots \otimes D_n}_v = \sem{\Swap_{s_1,\cdot,s_n}}_v\cdot (\sem{D_1}_v\otimes \cdots \otimes \sem{D_1}_v)\cdot \sem{\Swap_{s'_1,\cdot,s'_n}}_v$;
+  \item $\sem{D_1\cdot D_2}_v = \sem{\cl(D_1, s_2\backslash s'_1)}_v\cdot \sem{\cl(D_2, s'_1\backslash s_2)}_v$.
 Note that $s_2\backslash s'_1$ and $s'_1\backslash s_2$ are the minimal extension that make it interpretable.
@@
 \end{itemize}
-Thus, $\sem{D}$ is a linear map from $\cH_{\sem{s'}}$ to $\cH_{\sem{s}}$ given $E\vdash D : \cD(s,s')$.
-Following the semantics, labelled tensor is independent of its order, i.e., $\sem{D_1\otimes D_2} = \sem{D_2\otimes D_1}$ and $\sem{D_1\otimes (D_2\otimes D_3)} = \sem{(D_1\otimes D_2)\otimes D_3}$, which ensures the soundness of treating labelled tensor as an AC symbol.
+Thus, $\sem{D}_v$ is a linear map from $\cH_{\sem{s'}_v}$ to $\cH_{\sem{s}_v}$ given $E\vdash D : \cD(s,s')$.
+Following the semantics, labelled tensor is independent of its order, i.e., $\sem{D_1\otimes D_2}_v = \sem{D_2\otimes D_1}_v$ and $\sem{D_1\otimes (D_2\otimes D_3)}_v = \sem{(D_1\otimes D_2)\otimes D_3}_v$, which ensures the soundness of treating labelled tensor as an AC symbol.
 
@@
 Assuming no variables of $\DType(s_1, s_2)$, 
-repeating the application of the above rules yield the normal form--the addition of big operator sum: 
+repeating the application of the above rules yield the normal form of both sides of the equality--the addition of big operators: 
 \begin{equation}
@@
 the default order of variables.
-This yield the same shape of every subterm, e.g.,
+This yield the same shape of every subterm on both sides of the equality, e.g.,
 \begin{equation*}

### 2025-01-31 03labelled.tex
update Sec. 4

@@
 
-For any valuation $v$, and write $\cH_{\sem{s}_v} \triangleq \bigotimes_{r\in s}\cH_{\sem{\sigma_r}_v}$ for any $s\subseteq \cR$, where $E\vdash r : \reg(\sigma_r)$ for all $r\in\cR$.
-We interpret labelled Dirac notations inductively as: assume $E\vdash D_i : \cD(s_i,s'_i)$ in the following,
+For given context $E$ and set $s = \{r_1,\cdots,r_n\}\subseteq\cR$ ($r_i$ is sorted) with $E\vdash r_i : \sigma_i$, we define $\sigma_s\triangleq \sigma_1\times\cdots\times\sigma_n$ if $s$ is non-empty.
+\label{context: sigma s}
+For any valuation $v$, we further write $\cH_{\sem{s}_v} \triangleq \bigotimes_{i}\cH_{\sem{\sigma_i}_v}$ which equals to $\cH_{\sem{\sigma_s}_v}$ for non-empty $s$.
+% for any $s\subseteq \cR$, where $E\vdash r : \reg(\sigma_r)$ for all $r\in\cR$. 
+We interpret labelled Dirac notations inductively as (assume $E\vdash D_i : \cD(s_i,s'_i)$) :
 \begin{itemize}
@@
 
-\begin{theorem}[Soundness]
-  All rewriting rules for labelled Dirac notations listed here and in Appendix are sound, i.e., for any interpretation $\sem{\cdot}$, $D\reduce D'$ implies $\sem{D} = \sem{D'}$.
+Define $T_{s,s'}$ as $\SType$ if $s=s'=\emptyset$, $\KType(\sigma_s)$ if $s'=\emptyset$, $\BType(\sigma_{s'})$ if $s=\emptyset$ and $\OType(\sigma_s,\sigma_{s'})$ otherwise (see Sec.
+\ref{context: sigma s}).
+We claim the soundness of our normalization as:
+% Then if the original equality is $E\vdash D_1 = D_2 : \cD(s,s')$, what remains to be check $E\vdash e_1 = e_2 : T_{s,s'}$.
+
+\begin{theorem}[Soundness and Completeness of Normalization]
+  Assuming $E\vdash D_1 : \cD(s,s')$, $E\vdash D_2 : \cD(s,s')$, 
+  and no variables of $\DType(\cdot,\cdot)$ in $D_1,D_2$.
+  Let $e_1 = e_2$ be obtained by above normalization procedure on $D_1 = D_2$. Then:
+  \begin{itemize}
+    \item (Soundness) $E\vdash e_1 = e_2 : T_{s,s'}$ implies $E\vDash D_1 = D_2$;
+    \item (Completeness) $E\vDash D_1 = D_2$ implies $E\vDash e_1 = e_2$
+  \end{itemize}
+  %All rewriting rules for labelled Dirac notations listed here and in Appendix are sound, i.e., for any interpretation $\sem{\cdot}$, $D\reduce D'$ implies $\sem{D} = \sem{D'}$.
 \end{theorem}
 \begin{proof}
-  By checking each rules and its semantics.
+  Notice that all rewriting rules and ordering perserve the types, so the final step eliminating labels is both sound and complete. 
+  It is routinely to check that each rewriting rules and ordering step are sound based on the semantics.
+This directly leads to the soundness.
+For completeness, we need further check if the normalization procedure always provide $e_1 = e_2$, i.e., it never gets stuck and terminates.
+Step 1 is trivial.
+For step 2, it can be further checked that 1.
+there is no circular rules and thus terminates, and 2.
+if it is not in form (\ref{eqn:Labelled normal}), there must exists one of the rules can be further applied.
+For step 3, ordering is always applicable.
+This completes the proof.
 \end{proof}
 
-\begin{theorem}[Completeness for Dealing with Labels]
-  Above normalization procedure translates every well-typed equality of labelled Dirac notations to equality of plain Dirac notation.
-\end{theorem}
-\begin{proof}
-  It is sufficient to check if each step achieves its goal.
+% \begin{theorem}[Completeness of normalization]
+%   $E\vDash D_1 = D_2$ only if $E\vDash e_1 = e_2$, assuming that $E\vdash D_1 : \cD(s,s')$, $E\vdash D_2 : \cD(s,s')$, and $e_1 = e_2$ are obtained by above normalization procedure on $D_1 = D_2$.
+% \end{theorem}
+% \begin{proof}
+%   It is sufficient to check if each step achieves its goal.
 Step 1 is trivial.
 For step 2, it can be further checked that if it is not in form (\ref{eqn:Labelled normal}), there must exists one of the rules can be further applied.
-  For step 3, realize that all rewriting rules perserve the types, i.e., $D\reduce D'$ then $E\vdash D : \cD(s,s')$ and $E\vdash D : \cD(s,s')$ for some $s,s'\subseteq \cR$, which ensure the same shape of every subterm after sorting.
-\end{proof}
+%   For step 3, realize that all rewriting rules perserve the types, i.e., $D\reduce D'$ then $E\vdash D : \cD(s,s')$ and $E\vdash D : \cD(s,s')$ for some $s,s'\subseteq \cR$, which ensure the same shape of every subterm after sorting.
+% \end{proof}
 % , we claim that tensor of labelled basis in every sum body of Eqn.

### 2025-01-31 01intro_preliminary.tex
update

@@
 about quantum programs. Specifically, many quantum Hoare logics use
-labelled Dirac notation and its variants to express program assertions (see for example~\cite{DBLP:conf/lics/ZhouBHYY21,Zhou2023,qRHL_unruh2019,QSL_Le_2022,Zhong2024-np,incorrectness_2022,qafny2024}).
+labelled Dirac notation and its variants to express program assertions (see for example \cite{DBLP:conf/lics/ZhouBHYY21} \cite{Zhou2023} \cite{qRHL_unruh2019} \cite{QSL_Le_2022} \cite{Zhong2024-np} \cite{incorrectness_2022} \cite{qafny2024}).
 These logics also use
@@
 
-Roughly speaking, encoding in plain Dirac notation needs to
-tensoring identity operators and employing additional $\Swap$ gates
+Roughly speaking, encoding in plain Dirac notation requires
+tensoring identity operators and employing additional $\Swap$ gates,
 since generally the conventional order cannot guarantee 1. the order

### 2025-01-31 02language_design.tex
update

--- 2025-01-31 02language_design.tex
@@
 The typing system is responsible for classifying terms within a proof system, according to the types of variables and definitions. 
-We use a context \( E \) to preserve the assumptions \( x : T \) and definitions \( x := t : T \).
+We use a context \( \Gamma \) to preserve the assumptions \( x : T \) and definitions \( x := t : T \).
 \begin{definition}[Context]
-    The syntax for context \( E \) is:
+    The syntax for context \( \Gamma \) is:
     \begin{align*}
-        E ::= &\ [] \mid E; x : \Index \mid E; x : T \mid E; x := t : T.
+        \Gamma ::= &\ [] \mid \Gamma; x : \Index \mid \Gamma; x : T \mid \Gamma; x := t : T.
     \end{align*}
@@
 Assumptions, on the other hand, define the types of variables.
-We say an expression \( t \) has type \( X \) in context \( E \) if the typing judgment \( E \vdash t : X \) can be proven through the rules in~\Cref{sec: full typing rules}. These are two instances:
+We say an expression \( t \) has type \( X \) in context \( \Gamma \) if the typing judgment \( \Gamma \vdash t : X \) can be proven through the rules in~\Cref{sec: full typing rules}.
+These are two instances:
 \begin{gather*}
-    \frac{E \vdash t : \Basis(\sigma)}{E \vdash \ket{t} : \KType(\sigma)},
+    \frac{\Gamma \vdash t : \Basis(\sigma)}{\Gamma \vdash \ket{t} : \KType(\sigma)},
     \qquad
-    \frac{E \vdash B : \BType(\sigma) \qquad E \vdash K : \KType(\sigma)}{E \vdash B \cdot K : \SType}.
+    \frac{\Gamma \vdash B : \BType(\sigma) \qquad \Gamma \vdash K : \KType(\sigma)}{\Gamma \vdash B \cdot K : \SType}.
 \end{gather*}
@@
 \[
-    \frac{E \vdash s : \SET(\sigma) \qquad E \vdash f : \Basis(\sigma) \to \KType(\tau)}{E \vdash \sum_{s} f : \KType(\tau)}.
+    \frac{\Gamma \vdash s : \SET(\sigma) \qquad \Gamma \vdash f : \Basis(\sigma) \to \KType(\tau)}{\Gamma \vdash \sum_{s} f : \KType(\tau)}.
 \]
@@
 expressions.
-Specifically, for a well-formed context \( E \), term \( t \), and type \( T \), if \( E \vdash t : T \), then for any valuation \( v \) that is valid for $\Gamma$, the interpretation of \( t \)  w.r.t.\, $v$ is an element of the interpretation of \( T \) w.r.t.\, $v$.
+Specifically, for a well-formed context \( \Gamma \), term \( t \), and type \( T \), if \( \Gamma \vdash t : T \), then for any valuation \( v \) that is valid for $\Gamma$, the interpretation of \( t \)  w.r.t.\, $v$ is an element of the interpretation of \( T \) w.r.t.\, $v$.
 
 \begin{lemma}[Soundness of type system]
-    For any well-formed context \( E \), term \( t \), and type \( T \), if \( E \vdash t : T \), then for all valuations \( v \), \( \sem{t}_v \in \sem{T}_v \).
+  If \( \Gamma \vdash t : T \), then for all valuations \( v \) valid w.r.t.\,
+  $\Gamma$, we have \( \sem{t}_v \in \sem{T}_v \).
 \end{lemma}
-\begin{proof}
-By induction on the derivation.
-\end{proof}
+% \begin{proof}
+% By induction on the derivation.
+% \end{proof}
 This interpretation formalizes the standard understanding of Dirac
@@
 \end{align*}
-The entire axioms are separated into  two sets $R$ and $F$.
-During equivalence checking, most of the axioms are decided by term rewriting, which form the set $R$. However, there are axioms requiring special algorithms, which are collected in the set $F$.
-\begin{definition}[axiom set F]
-\label{def: axiom F}
+The entire axioms are separated into  two sets $R$ and $E$.
+$R$ contains the axioms normalized by term rewriting. There are axioms requiring special algorithms, which are collected in the set $E$.
+\begin{definition}[axiom set E]
+\label{def: axiom E}
 \begin{align*}
@@
 
-\gb{these two paragraphs do not make too much sense.
-We also need to define \( E \vdash e_1 = e_2 \)}The scalar theories are treated separately as a module and are not considered in this work.
-In the implementation, we use the Mathematica kernel to decide scalar equivalences.
-
-These equational axioms provide an operable theory for the proof automation algorithm.
+% These equational axioms provide an operable theory for the proof automation algorithm.
 Denotational semantics can be seen as one model for this theory, meaning that equivalences derived from the axioms always imply equivalence in the interpretations.
 
-An equation \( \Gamma \vdash e_1 = e_2 :T \) is valid, written \(
-\Gamma \vDash e_1 = e_2 :T \), if \( \sem{e_1}_v = \sem{e_2}_v \) for
+We say an equation $e_1 = e_2$ is provable, written as $\Gamma \vdash e_1 = e_2 : T$, if $\Gamma \vdash e_1 : T$ and $\Gamma \vdash e_2 : T$ are provable, and $e_1 = e_2$ can be deduced in $\Gamma$ using the axioms and equational logic.
+An equation \( e_1 = e_2 \) is valid in context $\Gamma$, written \(
+\Gamma \vDash e_1 = e_2 \), if \( \sem{e_1}_v = \sem{e_2}_v \) for
 all valuations \( v \) that are valid w.r.t.\, \( \Gamma \).
@@
 \begin{lemma}[Soundness of equational theory]\label{lem: axiom sound}
-If \( E \vdash e_1 = e_2 :T \) then \( E \vDash e_1 = e_2\).
+If \( \Gamma \vdash e_1 = e_2 :T \) then \( \Gamma \vDash e_1 = e_2\).
 \end{lemma}
-\begin{proof}
-By induction on the structure of the derivation
-\end{proof}
+% \begin{proof}
+% By induction on the structure of the derivation
+% \end{proof}
 Next, we formalize the motivating example~\Cref{ex: motivating} in
@@
     \label{ex: formalizing motivating}
-    Definitions and assumptions in the context \( E \) are formalized as follows:
+    Definitions and assumptions in the context \( \Gamma \) are formalized as follows:
     \begin{align*}
@@
         &\text{phi} &&:= \lambda T : \Index. \sum_{i \in \mathbf{U}(T)} \sum_{j \in \mathbf{U}(T)} \ket{(i, j)} : \forall T.\KType(T \times T); \\
-        & T && : \Index; \\
-        & M && : \OType(T, T).
+        & T && : \Index;  \qquad \qquad \qquad \qquad \qquad M \qquad : \OType(T, T).
     \end{align*}
@@
     The symbol \( \text{TPO} \) represents the transpose of an operator, polymorphic on the Hilbert spaces \( T_1 \) and \( T_2 \). 
-And its type \(\forall T_1.
-\forall T_2.
-\OType(T_1, T_2) \to \OType(T_2, T_1)\) suggests that TPO is a function from operators to operators, depending on the index $T_1$ and $T_2$.
     The symbol \( \text{phi} \) takes the index \( T \) and defines the maximally entangled states, summing over all basis elements in \( T \), as indicated by the universal set \( \mathbf{U}(T) \).
@@
 \subsection{Algorithm to Decide Equivalence}
+% \gb{I still think that the names first rewritings, second rewriting and de bruijn normalization are not good, in particular, de bruijn normalization is non-standard terminology that will make people familiar with the field cringe.}
 
 The equivalence of Dirac notations are decided by normalization, which transforms equivalent expressions under a set of axioms into the same syntax.
-We use an efficient algorithm to perform the normalization totally on $R \cup F$.
+We use an efficient algorithm to perform the normalization totally on $R \cup E$.
 
 \begin{enumerate}
-    \item \textbf{First Rewritings}: Expand definitions and simplify expressions.
+    \item \textbf{Term Rewriting by rules}: Expand definitions and simplify expressions.
     \item \textbf{Variable Expansion}: Convert to abstract element-wise representation.
-    \item \textbf{Second Rewritings}: Normalize terms on \( R \) modulo \( F \).
+    \item \textbf{Term Rewriting by rules}: Normalize terms on \( R \) modulo \( E \).
     \item \textbf{Sorting Without Bound Variables}: Normalize AC-equivalence.
     \item \textbf{Swap Successive Summations}: Normalize SUM-SWAP equivalence.
-    \item \textbf{De Bruijn Normalization}: Normalize \( \alpha \)-equivalence.
+    \item \textbf{Use de Bruijn Index}: Normalize \( \alpha \)-equivalence.
 \end{enumerate}
@@
 The reduction proceeds by matching the subterms with the left-hand side, and replacing it with the right-hand side.
-For example, the rule $a.K + K\ \reduce\ (a + 1).K$ matches the term $(x\times y) .
-\ket{t} + \ket{t}$ by the substitution $\{a \to  x\times y, K \to \ket{t}\}$, and is rewritten into $(x\times y + 1) .
-\ket{t}$.
-Step 1 and 3 uses the same set of rewriting rules, see~\Cref{sec: rewriting rules}.
-Step 2 expands variables like the ket $K$ to their abstract element-wise representation $\sum_{i} (\bra{i} \cdot K). \ket{i}$, which improves the decidability with sum.
+For example, the term $(x\times y) . \ket{t} + \ket{t}$ is matched by the rule $a.K + K\ \reduce\ (a + 1).K$, and is rewritten into $(x\times y + 1) . \ket{t}$.
+Step 1 and 3 uses the same set of rewriting rules in~\Cref{sec: rewriting rules}.
+Step 2 expands variables to their abstract element-wise representation, e.g., $K \ \reduce\ \sum_{i} (\bra{i} \cdot K). \ket{i}$. It improves the decidability with sum.
 
-Steps 4 - 6 are special algorithms to further normalize the axiom set $F$.
+Steps 4 - 6 are special algorithms to further normalize the axiom set $E$.
 Here the main difficulty is the coexistence of AC-equivalence and SUM-SWAP, so that naive sorting will not convert equivalent terms to the same form.
@@
 
-As a comparison, the previous work conducts normalization paritally on $R$, and decide axioms in $F$ by iterating through all possible permutations.
+As a comparison, the previous work conducts normalization paritally on $R$, and decide axioms in $E$ by iterating through all possible permutations.
 Our algorithm totally normalize the term, as is illustrated below.
 \[
-\text{(partial)}\qquad e_1 \mathop{\twoheadrightarrow}^R e_1' \mathop{=}^F e_2' \mathop{\twoheadleftarrow}^R e_2
+\text{(partial)}\qquad e_1 \mathop{\twoheadrightarrow}^R e_1' \mathop{=}^E e_2' \mathop{\twoheadleftarrow}^R e_2
 \qquad
-\text{(total)}\qquad e_1 \mathop{\twoheadrightarrow}^{R \cup F} e \equiv e \mathop{\twoheadleftarrow}^{R \cup F} e_2
+\text{(total)}\qquad e_1 \mathop{\twoheadrightarrow}^{R \cup E} e \equiv e \mathop{\twoheadleftarrow}^{R \cup E} e_2
 \]
-While the permutation checking algorithm has factorial complexity on the number of AC symbol arguments, our algorithm uses normalization by sorting, and the complexity is polynomial on the term size.
-This improvement in efficiency is also confirmed by experiments.
+% While the permutation checking algorithm has factorial complexity on the number of AC symbol arguments, our algorithm uses normalization by sorting, and the complexity is polynomial on the term size.
+% This improvement in efficiency is also confirmed by experiments.
 

### 2025-01-31 03labelled.tex
update

@@
 This requires an additional check on the valid registers--they should not include repeated quantum variables--which is often handled in programming languages via linear types.
-As such, we define the \emph{order-free} variable set of a register inductively by: 1) $\var(R) \triangleq \{r\}$, if $R\equiv r$ and 2) $\var(R) \triangleq \var(R_1) \cup\var(R_2)$, if $R\equiv (R_1,R_2)$.
+As such, we define the \emph{order-free} variable set of a register as all quantum variables appearing in the register.
 We use the variable set to establish side conditions of typing valid registers and employ it as the type parameter for annotating labelled Dirac terms.
@@
   \frac{
-      E \vdash R : \reg(\sigma) \quad
-      E \vdash Q : \reg(\tau)
+      \Gamma \vdash R : \reg(\sigma) \quad
+      \Gamma \vdash Q : \reg(\tau)
       \quad \var(R) \cap \var(Q) = \emptyset
-  }{E \vdash (R,Q) : \reg(\sigma \times \tau)} \\[0.2cm]
+  }{\Gamma \vdash (R,Q) : \reg(\sigma \times \tau)} \\[0.2cm]
   \frac{
-          E \vdash r : \reg(\sigma) \quad
-          E \vdash i : \Basis(\sigma)}
-  {E \vdash |i\>_r : \DType(\{r\}, \emptyset)}
+          \Gamma \vdash r : \reg(\sigma) \quad
+          \Gamma \vdash i : \Basis(\sigma)}
+  {\Gamma \vdash |i\>_r : \DType(\{r\}, \emptyset)}
   \qquad
-  \frac{E \vdash R : \reg(\sigma)\qquad E \vdash K : \KType(\sigma)}{E \vdash K_R : \DType(\var(R), \emptyset)} \\[0.2cm]
+  \frac{\Gamma \vdash R : \reg(\sigma)\qquad \Gamma \vdash K : \KType(\sigma)}{\Gamma \vdash K_R : \DType(\var(R), \emptyset)} \\[0.2cm]
     \frac{
-        E \vdash D_i : \DType(s_i,s_i') \qquad
+        \Gamma \vdash D_i : \DType(s_i,s_i') \qquad
         \forall\,i\neq j.\ s_i\cap s_j = \emptyset \qquad
@@
     }
-    {E \vdash D_1 \otimes \cdots \otimes D_i : \DType(\bigcup_i s_i, \bigcup_i s_i')}.
+    {\Gamma \vdash D_1 \otimes \cdots \otimes D_i : \DType(\bigcup_i s_i, \bigcup_i s_i')}.
 \end{gather*}
@@
 Formally, we equip $\cR$ with a default order, say for example the alphabetical order of names.
-For any valid register $R$, there exists the operator $\Swap_R$ that sorts $R$; for example, if $E\vdash p : \reg(\sigma_p)$ and $E\vdash q : \reg(\sigma_q)$, then $\Swap_{(q,p)} \triangleq \sum_{i\in \bU(\sigma_p)}\sum_{j\in \bU(\sigma_q)}|i\>\<j|\otimes |j\>\<i|$.
+For any valid register $R$, there exists the operator $\Swap_R$ that sorts $R$; for example, if $\Gamma\vdash p : \reg(\sigma_p)$ and $\Gamma\vdash q : \reg(\sigma_q)$, then $\Swap_{(q,p)} \triangleq \sum_{i\in \bU(\sigma_p)}\sum_{j\in \bU(\sigma_q)}|i\>\<j|\otimes |j\>\<i|$.
 We can further define $\Swap_{s_1,s_2,\cdots}$ for merging disjoint sets orderly.
 
-For given context $E$ and set $s = \{r_1,\cdots,r_n\}\subseteq\cR$ ($r_i$ is sorted) with $E\vdash r_i : \sigma_i$, we define $\sigma_s\triangleq \sigma_1\times\cdots\times\sigma_n$ if $s$ is non-empty.
+For given context $\Gamma$ and set $s = \{r_1,\cdots,r_n\}\subseteq\cR$ ($r_i$ is sorted) with $\Gamma\vdash r_i : \sigma_i$, we define $\sigma_s\triangleq \sigma_1\times\cdots\times\sigma_n$ if $s$ is non-empty.
 \label{context: sigma s}
 For any valuation $v$, we further write $\cH_{\sem{s}_v} \triangleq \bigotimes_{i}\cH_{\sem{\sigma_i}_v}$ which equals to $\cH_{\sem{\sigma_s}_v}$ for non-empty $s$.
-% for any $s\subseteq \cR$, where $E\vdash r : \reg(\sigma_r)$ for all $r\in\cR$. 
-We interpret labelled Dirac notations inductively as (assume $E\vdash D_i : \cD(s_i,s'_i)$) :
+% for any $s\subseteq \cR$, where $\Gamma\vdash r : \reg(\sigma_r)$ for all $r\in\cR$. 
+We interpret labelled Dirac notations inductively as (assume $\Gamma\vdash D_i : \cD(s_i,s'_i)$) :
 \begin{itemize}
@@
 \end{itemize}
-Thus, $\sem{D}_v$ is a linear map from $\cH_{\sem{s'}_v}$ to $\cH_{\sem{s}_v}$ given $E\vdash D : \cD(s,s')$.
+Thus, $\sem{D}_v$ is a linear map from $\cH_{\sem{s'}_v}$ to $\cH_{\sem{s}_v}$ given $\Gamma\vdash D : \cD(s,s')$.
 Following the semantics, labelled tensor is independent of its order, i.e., $\sem{D_1\otimes D_2}_v = \sem{D_2\otimes D_1}_v$ and $\sem{D_1\otimes (D_2\otimes D_3)}_v = \sem{(D_1\otimes D_2)\otimes D_3}_v$, which ensures the soundness of treating labelled tensor as an AC symbol.
@@
 and thus it is equivalent to decide the equivalence of additions of subterms of:
-\begin{equation}
-  \label{eqn:Labelled normal1}
+\begin{equation*}
   \sum_{i}\cdots\sum_{j}\cdots \sum_{i'}\cdots\sum_{j'} a_1 . ((\ket{i} \otimes \cdots \otimes \ket{j}) \otimes (\bra{i'}\otimes\cdots\otimes \bra{j'}))
-\end{equation}
+\end{equation*}
 which do not involve any labels.
@@
 We claim the soundness of our normalization as:
-% Then if the original equality is $E\vdash D_1 = D_2 : \cD(s,s')$, what remains to be check $E\vdash e_1 = e_2 : T_{s,s'}$.
+% Then if the original equality is $\Gamma\vdash D_1 = D_2 : \cD(s,s')$, what remains to be check $\Gamma\vdash e_1 = e_2 : T_{s,s'}$.
 
 \begin{theorem}[Soundness and Completeness of Normalization]
-  Assuming $E\vdash D_1 : \cD(s,s')$, $E\vdash D_2 : \cD(s,s')$, 
+  Assuming $\Gamma\vdash D_1 : \cD(s,s')$, $\Gamma\vdash D_2 : \cD(s,s')$, 
   and no variables of $\DType(\cdot,\cdot)$ in $D_1,D_2$.
@@
   \begin{itemize}
-    \item (Soundness) $E\vdash e_1 = e_2 : T_{s,s'}$ implies $E\vDash D_1 = D_2$;
-    \item (Completeness) $E\vDash D_1 = D_2$ implies $E\vDash e_1 = e_2$
+    \item (Soundness) $\Gamma\vdash e_1 = e_2 : T_{s,s'}$ implies $\Gamma\vDash D_1 = D_2$;
+    \item (Completeness) $\Gamma\vDash D_1 = D_2$ implies $\Gamma\vDash e_1 = e_2$
   \end{itemize}
@@
 \end{theorem}
-\begin{proof}
-  Notice that all rewriting rules and ordering perserve the types, so the final step eliminating labels is both sound and complete. 
-  It is routinely to check that each rewriting rules and ordering step are sound based on the semantics.
+% \begin{proof}
+%   Notice that all rewriting rules and ordering perserve the types, so the final step eliminating labels is both sound and complete. 
+%   It is routinely to check that each rewriting rules and ordering step are sound based on the semantics.
 This directly leads to the soundness.
@@
 This completes the proof.
-\end{proof}
+% \end{proof}
 

### 2025-01-31 05tool_eval.tex
update

@@
 The improvement and extensions above concludes in our implementation called D-Hammer, a solver written in \CC.
-It has a parser built by ANTLR4, and scalar reasonings are powered by a Mathematica kernel.
+It has a parser built by ANTLR4, and scalar reasonings are powered by Mathematica Engine.
 The user can use commands to make definitions and assumptions in the maintained context, conduct the normalization and equivalence checking, and obtain the rewriting trace output.
@@
 This implementation is tested on the benchmark of the previous work and new examples for labelled Dirac notation, demonstrating significant improvement in expressivity and efficiency. 
-The tool and example encodings will be open-source and public available.
+The tool and example encodings will be open-source and publicly available.
 
-\subsubsection{Project Structure}
-The project structure is illustrated in~\Cref{fig: dhammer structure}.
-\texttt{ualg} is the module for universal algebra, defining basic concepts like terms and substitutions.
+\subsubsection{Structure and Mechanism} 
+The project consists of the following components:
+\begin{itemize}
+    \item \texttt{antlr4}: the third-party library for building the parser.
+    \item \texttt{WSTP interface}: the wrapper to link with Mathematica Engine.
+    \item \texttt{ualg}: the framework module for universal algebra, defining basic concepts like terms and substitutions.
+    \item \texttt{dhammer}: the main module containing symbols definitions, type checking, rewriting rules, normalization algorithm and the prover.
+    \item \texttt{example}: an example benchmark for evaluation.
+    \item \texttt{toplevel}: the command line application.
+\end{itemize}
+% The project structure is illustrated in~\Cref{fig: dhammer structure}.
+% \texttt{ualg} is the module for universal algebra, defining basic concepts like terms and substitutions.
 It serves as the library for \texttt{dhammer}, which are then utilized in the example benchmarks and the toplevel command line application.
-The components of \texttt{dhammer} are as follows:
-\begin{itemize}
-    \item \texttt{symbols.cpp}: the reserved symbols and AC symbols;
-    \item \texttt{syntax\_theory.cpp}: syntax related algorithms, such as de Bruijn normalization and freeness of variables;
-    \item \texttt{calculus.cpp}: type checker and intergration with Mathematica;
-    \item \texttt{reduction.cpp}: all the rewriting rules and transformations;
-    \item \texttt{dirac\_parser.cpp}: parser for Dirac notation and D-Hammer commands;
-    \item \texttt{prover.cpp}: the prover that maintains the context and process commands like definition or equivalence checking.
-\end{itemize}
+Important components of \texttt{dhammer} are:
 
-\input{fig/proj_structure.tex}
+% \input{fig/proj_structure.tex}
 
@@
 }
-The syntax tree structure is also compatible with the datatype of Mathematica. This improves the interoperability between D-Hammer and the Mathematica system, enabling them to work interleavingly.
+% The syntax tree structure is also compatible with the datatype of Mathematica. This improves the interoperability between D-Hammer and the Mathematica system, enabling them to work interleavingly.
 To improve usability, D-Hammer also supports many special notations for terms,  and most Dirac notation terms will be encoded in the natural way.
@@
 Finally, D-Hammer uses a prover to host the computation.
-The prover maintains a well-formed context $E$, and processes commands to modify the context and conduct calculations.
+The prover maintains a well-formed context $\Gamma$, and processes commands to modify the context and conduct calculations.
 The commands are listed below.
@@
 
-\begin{figure}[h]
+\begin{figure}
     \centering
-    \includegraphics[width=0.8\textwidth]{fig/coqq.pdf}
+    \includegraphics[width=0.65\textwidth]{fig/coqq.pdf}
     \caption{Time comparison between DiracDec and D-Hammer on the CoqQ benchmark.}
@@
 \subsubsection{Labelled Dirac Notation}
-We present a new set of examples for labelled Dirac notation (LDN), as detailed in~\Cref{sec: examples for labelled}.
+We present a new set of examples for labelled Dirac notation (LDN).
 These examples, illustrated in~\Cref{fig: labelled examples}, include six representative cases drawn from a variety of sources, such as well-established theorems, research paper results, quantum circuit equivalence, and quantum program verification.
@@
 In LDN-18, $q, q_1$ and $q_2$ represent variables for quantum programs, and the equation shows the change of the state after applying $X$ gate on $q_2$.
-Experiments show that our algorithm is relatively inefficient for labelled Dirac notations based on qubits.
+% Experiments show that our algorithm is relatively inefficient for labelled Dirac notations based on qubits.
 
@@
     \end{tabular}
-    \caption{Part of examples for labelled Dirac notations.}
+    \caption{Part of examples for labelled Dirac notations. See~\Cref{sec: examples for labelled} for the full list.}
     \label{fig: labelled examples}

### 2025-01-31 06related_conclusion.tex
update

@@
 exponential functions, that are commonly used in quantum physics.
-There are three important directions for future work. The first
-direction is to enhance D-Hammer with support for big tensor notation,
-i.e.\, to support expressions of the form $\bigotimes_{i\in I}
-e_i$. The second direction is to extend D-Hammer with a mechanism to
+There are two important directions for future work. 
+% The first
+% direction is to enhance D-Hammer with support for big tensor notation,
+% i.e.\, to support expressions of the form $\bigotimes_{i\in I}
+% e_i$. 
+The first direction is to extend D-Hammer with a mechanism to
 generate independently verifiable certificates.  There is a large body
@@
 from powerful mechanisms to integrate rewriting procedures into the
-Lean proof assistant. Finally, the third direction is to connect
+Lean proof assistant. The second direction is to connect
 D-Hammer with quantum program verifiers. Two potential applications

### 2025-01-31 draw.py

@@ -1,0 +1,88 @@
+import pandas as pd
+import matplotlib.pyplot as plt
+
+# Read CSV file (assuming no header)
+file_path = 'data/coqq.csv'  # Change this to your actual file path
+df = pd.read_csv(file_path, header=None)
+
+# Assign column names (assuming three columns: Name, DiracDec, D-Hammer)
+df.columns = ['Name', 'DiracDec', 'D-Hammer']
+
+# set the Name column as the index
+df.set_index('Name', inplace=True)
+
+# Create the figure and axis
+fig, ax = plt.subplots()
+
+# Set log scale for both axes
+ax.set_xscale('log')
+ax.set_yscale('log')
+
+# Add light gray background and grid
+plt.style.use('seaborn-v0_8-dark')
+ax.grid(color='gray', linestyle='--', linewidth=0.5)
+
+# Set the background color of the plot area (rectangle data region)
+ax.set_facecolor('lightgray')
+
+# Create scatter plot
+ax.scatter(df['DiracDec'], df['D-Hammer'], color='blue', marker='o', s=20, label='Examples')
+
+# Set the scope of the plot:
+ax.set_ylim(10**(-2.7), 10**(0.2))
+
+# Add x = y line
+x_min, x_max = ax.get_xlim()  # Get current x-axis limits
+y_min, y_max = ax.get_ylim()  # Get current y-axis limits
+xy_min = min(x_min, y_min)  # Take the lower bound for diagonal start
+xy_max = max(x_max, y_max)  # Take the upper bound for diagonal end
+ax.plot([xy_min, xy_max], [xy_min, xy_max], color='red', linestyle='--', linewidth=1.5, label="Equal time")
+
+# Labels and title
+ax.set_xlabel('DiracDec time (s)')
+ax.set_ylabel('D-Hammer time (s)')
+
+
+# Add a label to a specific data point
+highlight_name = 'COQQ-129 dualsoK'
+highlight_x = df.loc[highlight_name, 'DiracDec']
+highlight_y = df.loc[highlight_name, 'D-Hammer']
+highlight_label = highlight_name
+
+ax.annotate(highlight_label, 
+            (highlight_x, highlight_y), 
+            textcoords="offset points",
+            xytext=(50,-20),  # Move label further right and up
+            ha='right', 
+            fontsize=10, 
+            color='black', 
+            arrowprops=dict(arrowstyle="->", color='black'),
[…]

### 2025-01-31 02language_design.tex
18-page version

@@
 \begin{enumerate}
-    \item \textbf{Term Rewriting by rules}: Expand definitions and simplify expressions.
-    \item \textbf{Variable Expansion}: Convert to abstract element-wise representation.
-    \item \textbf{Term Rewriting by rules}: Normalize terms on \( R \) modulo \( E \).
-    \item \textbf{Sorting Without Bound Variables}: Normalize AC-equivalence.
-    \item \textbf{Swap Successive Summations}: Normalize SUM-SWAP equivalence.
-    \item \textbf{Use de Bruijn Index}: Normalize \( \alpha \)-equivalence.
+    \item \textbf{Rule based term rewriting}: Expand definitions and simplify expressions.
+    \item \textbf{Variable expansion}: Convert to abstract element-wise representation.
+    \item \textbf{Rule based term rewriting}: Normalize terms on \( R \) modulo \( E \).
+    \item \textbf{Sorting without bound variables}: Normalize AC-equivalence.
+    \item \textbf{Swapping successive summations}: Normalize SUM-SWAP equivalence.
+    \item \textbf{Use de Bruijn index}: Normalize \( \alpha \)-equivalence.
 \end{enumerate}

### 2025-01-31 05tool_eval.tex
18-page version

@@
 \texttt{T} is a type index, representing arbitrary Hilbert space types.
-\texttt{M} is assume to be an operator in the Hilbert space with type \texttt{T}.
+\texttt{M} is assumed to be an operator in the Hilbert space with type \texttt{T}.
 \texttt{phi} is defined as the maximally entangled state, depending on the bound variable \texttt{T} as index.
@@
 Among the examples, LDN-12 shows the flexibility in combining labelled Dirac notations.
-Example LDN-14 and LDN-18 are about circuits or programs on qubits.
 LDN-14 shows how to calculate controlled-not gate in different ways. 
-In LDN-18, $q, q_1$ and $q_2$ represent variables for quantum programs, and the equation shows the change of the state after applying $X$ gate on $q_2$.
+%In LDN-18, $q, q_1$ and $q_2$ represent variables for quantum programs, and the equation shows the change of the state after applying $X$ gate on $q_2$. 
 % Experiments show that our algorithm is relatively inefficient for labelled Dirac notations based on qubits.

### 2025-01-31 06related_conclusion.tex
18-page version

@@
 leverage higher-order functions to provide a compact and expressive
-representation of big operators. Second, our language supports AC
+representation of big operators. Second, our language adopts AC
 symbols with indefinite arities, which leads to compact representation
@@
 a consequence, its behavior, and in particular, its typing rules are
-constrained by the type system of Mathematica. In contrast, D-Hammer
+constrained by the lack of typing in Mathematica. In contrast, D-Hammer
 is designed as a separate tool; as a consequence, it benefits from an

### 2025-01-31 02language_design.tex
sound and complete for normalization

@@
 \begin{lemma}[Soundness of type system]
+  \label{lemma:sound type system}
   If \( \Gamma \vdash t : T \), then for all valuations \( v \) valid w.r.t.\,

### 2025-01-31 03labelled.tex
sound and complete for normalization

@@
 The third line gives the typing of labelled tensoring with a check whether the component subsystems are disjoint with each other.
+Since $\cR$ is given, let $\sigma_r : \Index$ denote type of $r$, i.e.,  $\Gamma\vdash r : \reg(\sigma_r)$, for simplicity.
 
@@
 Formally, we equip $\cR$ with a default order, say for example the alphabetical order of names.
-For any valid register $R$, there exists the operator $\Swap_R$ that sorts $R$; for example, if $\Gamma\vdash p : \reg(\sigma_p)$ and $\Gamma\vdash q : \reg(\sigma_q)$, then $\Swap_{(q,p)} \triangleq \sum_{i\in \bU(\sigma_p)}\sum_{j\in \bU(\sigma_q)}|i\>\<j|\otimes |j\>\<i|$.
+For any valid register $R$, there exists the operator $\Swap_R$ that sorts $R$; for example, $\Swap_{(q,p)} \triangleq \sum_{i\in \bU(\sigma_p)}\sum_{j\in \bU(\sigma_q)}|i\>\<j|\otimes |j\>\<i|$.
 We can further define $\Swap_{s_1,s_2,\cdots}$ for merging disjoint sets orderly.
 
-For given context $\Gamma$ and set $s = \{r_1,\cdots,r_n\}\subseteq\cR$ ($r_i$ is sorted) with $\Gamma\vdash r_i : \sigma_i$, we define $\sigma_s\triangleq \sigma_1\times\cdots\times\sigma_n$ if $s$ is non-empty.
-\label{context: sigma s}
-For any valuation $v$, we further write $\cH_{\sem{s}_v} \triangleq \bigotimes_{i}\cH_{\sem{\sigma_i}_v}$ which equals to $\cH_{\sem{\sigma_s}_v}$ for non-empty $s$.
+For given context $\Gamma$ and any valuation $v$,
+we interpret 
+$$\mbox{$\sem{\cD(s,s')} \triangleq \mathcal{L}(\bigotimes_{j}\cH_{\sem{\sigma_{r'_j}}_v}, \bigotimes_{i}\cH_{\sem{\sigma_{r_i}}_v})$}$$ where $\mathcal{L}$ denotes the set of linear maps, 
+the sets $s = \{r_1,\cdots,r_n\}$ and $s' = \{r'_1,\cdots,r'_m\}$ ($r_i$ and $r'_j$ are sorted).
+% we define $\sigma_s\triangleq \sigma_1\times\cdots\times\sigma_n$ if $s$ is non-empty. \label{context: sigma s}
+% For any valuation $v$, we further write $\cH_{\sem{s}_v} \triangleq \bigotimes_{i}\cH_{\sem{\sigma_i}_v}$ which equals to $\cH_{\sem{\sigma_s}_v}$ for non-empty $s$.
+% We interpret $\sem{\cD(s,s')} \triangleq \mathcal{L}(\cH_{\sem{s'}_v}, \cH_{\sem{s}_v})$ where $\mathcal{L}$ denotes the set of linear maps.
 % for any $s\subseteq \cR$, where $\Gamma\vdash r : \reg(\sigma_r)$ for all $r\in\cR$. 
@@
 \end{itemize}
-Thus, $\sem{D}_v$ is a linear map from $\cH_{\sem{s'}_v}$ to $\cH_{\sem{s}_v}$ given $\Gamma\vdash D : \cD(s,s')$.
+It can be shown that Lemma \ref{lemma:sound type system} also holds for labelled terms, i.e., $\sem{D}_v \in \sem{\cD(s,s')}_v$ given $\Gamma\vdash D : \cD(s,s')$.
 Following the semantics, labelled tensor is independent of its order, i.e., $\sem{D_1\otimes D_2}_v = \sem{D_2\otimes D_1}_v$ and $\sem{D_1\otimes (D_2\otimes D_3)}_v = \sem{(D_1\otimes D_2)\otimes D_3}_v$, which ensures the soundness of treating labelled tensor as an AC symbol.
@@
 
-Define $T_{s,s'}$ as $\SType$ if $s=s'=\emptyset$, $\KType(\sigma_s)$ if $s'=\emptyset$, $\BType(\sigma_{s'})$ if $s=\emptyset$ and $\OType(\sigma_s,\sigma_{s'})$ otherwise (see Sec.
+%Define $T_{s,s'}$ as $\SType$ if $s=s'=\emptyset$, $\KType(\sigma_s)$ if $s'=\emptyset$, $\BType(\sigma_{s'})$ if $s=\emptyset$ and $\OType(\sigma_s,\sigma_{s'})$ otherwise (see Sec.
 \ref{context: sigma s}).
@@
 
+By routinely showing that Lemma \ref{lem: axiom sound} holds for the extra rules mentioned above and in Appendix, we further claim:
 \begin{theorem}[Soundness and Completeness of Normalization]
@@
   and no variables of $\DType(\cdot,\cdot)$ in $D_1,D_2$.
-  Let $e_1 = e_2$ be obtained by above normalization procedure on $D_1 = D_2$. Then:
-  \begin{itemize}
-    \item (Soundness) $\Gamma\vdash e_1 = e_2 : T_{s,s'}$ implies $\Gamma\vDash D_1 = D_2$;
-    \item (Completeness) $\Gamma\vDash D_1 = D_2$ implies $\Gamma\vDash e_1 = e_2$
-  \end{itemize}
+  Let $e_1 = e_2$ be obtained by above normalization procedure on $D_1 = D_2$. Then $\Gamma\vDash D_1 = D_2$ if and only if $\Gamma\vDash e_1 = e_2$.
   %All rewriting rules for labelled Dirac notations listed here and in Appendix are sound, i.e., for any interpretation $\sem{\cdot}$, $D\reduce D'$ implies $\sem{D} = \sem{D'}$.

### 2025-02-01 03labelled.tex
update

@@
 We can further define $\Swap_{s_1,s_2,\cdots}$ for merging disjoint sets orderly.
+See Appendix \ref{sec: full denotational sem} for details.
 
@@
 we interpret 
-$$\mbox{$\sem{\cD(s,s')} \triangleq \mathcal{L}(\bigotimes_{j}\cH_{\sem{\sigma_{r'_j}}_v}, \bigotimes_{i}\cH_{\sem{\sigma_{r_i}}_v})$}$$ where $\mathcal{L}$ denotes the set of linear maps, 
+$$\mbox{$\sem{\cD(s,s')}_v \triangleq \mathcal{L}(\bigotimes_{j}\cH_{\sem{\sigma_{r'_j}}_v}, \bigotimes_{i}\cH_{\sem{\sigma_{r_i}}_v})$}$$ where $\mathcal{L}$ denotes the set of linear maps, 
 the sets $s = \{r_1,\cdots,r_n\}$ and $s' = \{r'_1,\cdots,r'_m\}$ ($r_i$ and $r'_j$ are sorted).
@@
         $\sem{B_R}_v = \sem{B}_v\cdot \sem{\Swap_R}_v$;\quad
-        $\sem{O_{R_1,R_2}} = \sem{\Swap_{R_1}}_v\cdot \sem{O}_v\cdot \sem{\Swap_{R_2}}_v$;
+        $\sem{O_{R_1,R_2}}_v = \sem{\Swap_{R_1}}_v\cdot \sem{O}_v\cdot \sem{\Swap_{R_2}}_v$;
   \item $\sem{D_1\otimes \cdots \otimes D_n}_v = \sem{\Swap_{s_1,\cdot,s_n}}_v\cdot (\sem{D_1}_v\otimes \cdots \otimes \sem{D_1}_v)\cdot \sem{\Swap_{s'_1,\cdot,s'_n}}_v$;
@@
 \end{align*}
-where $|i_R\>$ and $\<i_{R'}|$ is constructed by tensoring the basis according to the structure of $R$. The rules for $e_R$ (ket and bra) are similar.
+where $|i_R\>$ and $\<i_{R'}|$ are constructed by tensoring the basis according to the structure of $R$ (see Appendix \ref{sec: full denotational sem}). The rules for $e_R$ (ket and bra) are similar.
 %This step reduces all labelled terms $e_R$ or $e_{R;R}$.
@@
 % Then if the original equality is $\Gamma\vdash D_1 = D_2 : \cD(s,s')$, what remains to be check $\Gamma\vdash e_1 = e_2 : T_{s,s'}$.
-
-By routinely showing that Lemma \ref{lem: axiom sound} holds for the extra rules mentioned above and in Appendix, we further claim:
+\paragraph*{}
+By routinely showing that Lemma \ref{lem: axiom sound} holds for the extra rules mentioned above and in Appendix \ref{sec: rewriting rules}, we further claim:
 \begin{theorem}[Soundness and Completeness of Normalization]

### 2025-02-01 03labelled.tex
update the proof in appendix

@@
 This yield the same shape of every subterm on both sides of the equality, e.g.,
-\begin{equation*}
-  % \label{eqn:Labelled normal}
+\begin{equation}
+  \label{eqn:Labelled normal1}
   \sum_{i}\cdots\sum_{j}\cdots \sum_{i'}\cdots\sum_{j'} a_1 . ((\ket{i}_p \otimes \cdots \otimes \ket{j}_q) \otimes (\bra{i'}_{p'}\otimes\cdots\otimes \bra{j'}_{q'}))
-\end{equation*}
+\end{equation}
 and thus it is equivalent to decide the equivalence of additions of subterms of:
-\begin{equation*}
-  \sum_{i}\cdots\sum_{j}\cdots \sum_{i'}\cdots\sum_{j'} a_1 . ((\ket{i} \otimes \cdots \otimes \ket{j}) \otimes (\bra{i'}\otimes\cdots\otimes \bra{j'}))
-\end{equation*}
+\begin{equation}
+  \label{eqn:Labelled normal2}
+  \sum_{i}\cdots\sum_{j}\cdots \sum_{i'}\cdots\sum_{j'} a_1 . ((\ket{i} \otimes \cdots \otimes \ket{j}) \cdot (\bra{i'}\otimes\cdots\otimes \bra{j'}))
+\end{equation}
 which do not involve any labels.
@@
 \begin{theorem}[Soundness and Completeness of Normalization]
+  \label{thm: normalization}
   Assuming $\Gamma\vdash D_1 : \cD(s,s')$, $\Gamma\vdash D_2 : \cD(s,s')$, 

### 2025-02-01 03labelled.tex
.

@@
 \paragraph*{}
-By routinely showing that Lemma \ref{lem: axiom sound} holds for the extra rules mentioned above and in Appendix \ref{sec: rewriting rules}, we further claim:
+By routinely showing that Lemma \ref{lem: axiom sound} holds for the extra rules mentioned above and in Appendix \ref{sec: rewriting rules} (if we regard them as equational theory), we further claim:
 \begin{theorem}[Soundness and Completeness of Normalization]

### 2025-02-01 01intro_preliminary.tex
.

@@
 \end{align*}
-where the RHS of each line are equivalent.  In addition, labelled
+where the RHS of each line is equivalent.  In addition, labelled
 Dirac notation can conveniently describe local measurements, partial
@@
 Just as in classical program logic, variable names are used to construct logical formulas instead of describing them as global memory functions, in quantum program logic, variable names are used to label the subsystems on which quantum gates act, rather than lifting them to the global system.
-Indeed, the latter would cause exponential length of the formula w.r.t the number of variables, as discussed in [cite].
+Indeed, the latter would cause exponential length of the formula w.r.t the number of variables, as discussed in \cite{QSL_Le_2022}.
 

### 2025-02-01 01intro_preliminary.tex
update. section 1 - 2 writing checked.

@@
 
-Dirac notation~\cite{dirac1939new}, a.k.a.\, bra-ket notation, is a
-mathematical formalism for representing quantum states using linear
-algebra notation: for instance, Dirac notation uses the linear
-combination \( a\ket{\psi} + b\ket{\phi} \) to represent the
-superposition of the two quantum states \( \ket{\psi} \) and
-\( \ket{\phi} \). Another essential ingredient of Dirac notation
-is the tensor product $\otimes$, which is used to describe composite
-states: for instance, Dirac notation uses the tensor expression
-\( \ket{\psi} \otimes \ket{\phi} \) to denote the composition of
-the two quantum states \( \ket{\psi} \) and \( \ket{\phi} \). It is
-also common to use a variant of Dirac notation, called labelled Dirac
-notation, to describe composite quantum states. In labelled Dirac
-notation, bras and kets are tagged with labels that identify the
-subsystems on which they operate. For instance, the labelled tensor
-expression
-\( \ket{\psi}_{S'} \otimes \ket{\phi}_{S} \)
-states that $\ket{\psi}_S$ and $\ket{\psi}_{S'}$ describe two quantum
-states over some subsystem $S$ and $S'$ respectively. By considering
-the relationship between $S$ and $S'$, one can obtain identities for
-free, e.g.\, 
+Dirac notation~\cite{dirac1939new}, also known as bra-ket notation, is a mathematical formalism for representing quantum states using linear algebra notation.
+For example, Dirac notation uses the linear combination \( a\ket{\psi} + b\ket{\phi} \) to represent the superposition of the quantum states \( \ket{\psi} \) and \( \ket{\phi} \).
+Another essential ingredient of Dirac notation is the tensor product $\otimes$, which is used to describe composite states.
+For instance, the tensor expression \( \ket{\psi} \otimes \ket{\phi} \) denotes the composition of the two quantum states \( \ket{\psi} \) and \( \ket{\phi} \).
+A variant of Dirac notation, called labelled Dirac notation, is often used to describe composite quantum states.
+In labelled Dirac notation, bras and kets are tagged with labels to identify the subsystems they operate on.
+For example, the labelled tensor expression \( \ket{\psi}_{S'} \otimes \ket{\phi}_{S} \) indicates that \( \ket{\psi}_{S} \) and \( \ket{\phi}_{S'} \) describe two quantum states over some subsystem $S$ and $S'$, respectively.
+By considering the relationship between $S$ and $S'$, one can obtain identities for free, e.g.\,
 %
-$$\ket{\phi}_S \otimes \ket{\psi}_{T} = \ket{\psi}_{T} \otimes \ket{\phi}_S
-\mbox{if}~S\cap S'=\emptyset$$
+$$\ket{\phi}_S \otimes \ket{\psi}_{T} = \ket{\psi}_{T} \otimes \ket{\phi}_S \quad \text{if}~S \cap S' = \emptyset$$
 %
 In turn, commutativity of the tensor product ensures that one can
-reason locally about quantum systems, and contributes to making
-labelled Dirac notation a convenient, compositional formalism to
-reason about quantum states--in a way that is similar to the way
-bunched logics support compositional reasoning about mutable states.
+reason locally about quantum systems, and contributes to making labelled Dirac notation a convenient, compositional formalism for reasoning about quantum states, akin to how bunched logics support compositional reasoning about mutable states.
 
-Labelled Dirac notation is also commonly used to express assertions
-about quantum programs. Specifically, many quantum Hoare logics use
-labelled Dirac notation and its variants to express program assertions (see for example \cite{DBLP:conf/lics/ZhouBHYY21} \cite{Zhou2023} \cite{qRHL_unruh2019} \cite{QSL_Le_2022} \cite{Zhong2024-np} \cite{incorrectness_2022} \cite{qafny2024}).
-These logics also use
-implicitly equational reasoning between labelled Dirac expressions, to
-glue applications of proof rules--similar to the use of the rule of
-consequence in the setting of classical program. Therefore, it is
+Labelled Dirac notation is also widely used to express assertions in quantum programs.
+Specifically, many quantum Hoare logics rely on labelled Dirac notation and its variants to express program assertions (see, for example, \cite{DBLP:conf/lics/ZhouBHYY21} \cite{Zhou2023} \cite{qRHL_unruh2019} \cite{QSL_Le_2022} \cite{Zhong2024-np} \cite{incorrectness_2022} \cite{qafny2024}).
+These logics also employ implicit equational reasoning between labelled Dirac expressions to glue applications of proof rule--similar to the rule of consequence in the setting of classical program verification.
+Consequently, it is
 essential for the verification of quantum programs to have automated
@@
 \subsubsection*{Contributions}
-This paper presents an automated tool, called D-Hammer, for reasoning
-about labelled Dirac notation. D-Hammer uses a rich, dependently typed
-language, to formalize labelled Dirac notation, and supports common
-idioms used to describe quantum systems, including big operators of
-the form $\sum_{i\in I} a_i \ket{\phi_i}_S$ to represent indexed
-superpositions of states. The semantics of typable expressions is
-given in terms of Hilbert spaces, tailored to interpreting tensor
-product as an AC symbol. We leverage this interpretation to define a
-rich equational theory for labelled Dirac notation, and prove its
-soundness with respect to our denotational semantics. Finally, we
-define an efficient normalization procedure to prove equivalence of
-two expressions. We then evaluate our procedure with respect to
-examples from the literature. Our evaluation covers examples based on
-plain Dirac notation, as well as examples based on
-labelled Dirac notation. The main conclusions are that our approach
-outperforms prior work by~\cite{diracdec} and is able to prove complex
-examples of reasoning with labelled Dirac notation from the
-literature, including examples from prior work on quantum separation
-logic~\cite{DBLP:conf/lics/ZhouBHYY21}.
+This paper presents D-Hammer, an automated tool for reasoning about labelled Dirac notation.
+D-Hammer uses a rich, dependently typed language to formalize labelled Dirac notation and supports common idioms for describing quantum systems, including big operators of the form \( \sum_{i \in I} a_i \ket{\phi_i}_S \) to represent indexed superpositions of states.
+The semantics of typable expressions are given in terms of Hilbert spaces, tailored to interpret the tensor product as an AC symbol.
+We leverage this interpretation to define a rich equational theory for labelled Dirac notation and prove its soundness with respect to our denotational semantics.
+Finally, we define an efficient normalization procedure to prove equivalence between two expressions.
+We then evaluate our procedure with respect to examples from the literature.
+Our evaluation covers examples based on plain Dirac notation, as well as examples based on
+labelled Dirac notation.
+The main conclusions are that our approach outperforms prior work by~\cite{diracdec} and is capable of proving complex examples from the literature, including those from prior work on quantum separation logic~\cite{DBLP:conf/lics/ZhouBHYY21}.
 
@@
 
-% \begin{figure}[t]
-% \begin{quantikz}
-%   & \gate{X} & \ctrl{1} & \\
-%   & & \targ{} &
-%   \end{quantikz}
-%   =\begin{quantikz}
-%   & \ctrl{1} & \gate{X} & \\
-%   & \targ{} & \gate{X} &
-%   \end{quantikz}
-%   \end{figure}
-
 \begin{example}
@@
   \ccnot|\ghz\rangle &= (I\otimes \cnot)|\ghz\rangle \\
-  &= (\Swap\otimes I)(I\otimes \cnot)(\Swap\otimes I)|\ghz\rangle
+  &= (\Swap\otimes I)(I\otimes \cnot)(\Swap\otimes I)|\ghz\rangle,
 \end{align*}
@@
   \end{figure}
+\end{example}
+
 Formalizing the statement in Dirac notation requires the following
-steps: 1. Arrange qubits in the conventional order, here we choose
+steps: 1. Arrange the qubits in a conventional order, here we choose
 $p,r,q$ to simplify the representation of $\ccnot$; 2. Lift the local
-operation $\cnot$ to the global system. For $\cnot$ acting on $r,q$ is
-easier, since $r,q$ consistent with the chosen order, and we only need
-to tensoring it by an identity operator $I$ on $p$, i.e., $(I\otimes
-\cnot)$.  For $\cnot$ acting on $p,q$, realize that $p,q$ are not
-adjacent in the chosen order, we additionally need to use $\Swap$ to
-temporarily exchange the qubits $p$ and $r$, i.e., globally, the
-$(\Swap\otimes I)$ before and after $(I\otimes \cnot)$ (lifting of
-$\cnot$ on $r,q$).
+operation $\cnot$ to the global system. When $\cnot$ acts on $r,q$, it is
+straightforward, as $r,q$ consistent with the chosen order. We only need
+to tensor it with an identity operator $I$ on $p$, i.e., $(I\otimes
+\cnot)$.  For $\cnot$ acting on $p,q$, note that $p,q$ are not
+adjacent in the chosen order. Thus, we additionally need the $\Swap$ gate to
+temporarily exchange the qubits $p$ and $r$, i.e., globally, we apply
+$(\Swap\otimes I)$ before and after $(I\otimes \cnot)$ to lift $\cnot$ on $r,q$.
 
 Roughly speaking, encoding in plain Dirac notation requires
-tensoring identity operators and employing additional $\Swap$ gates,
-since generally the conventional order cannot guarantee 1. the order
+tensoring identity operators and using additional $\Swap$ gates,
+since the conventional order does not generally guarantee the following: 1. the order
 of all local operations is consistent with it, 2. the local operations
-involve only adjacent qubits in the conventional order.
-\end{example}
+only involve adjacent qubits in the conventional order.
+
 To address the limitations of Dirac notations, physicists routinely
@@
 states or operations are applied, thereby avoiding unnecessary lifting
-and swap gates. For example, rewriting the above example using labels
-yields:
+and swap gates. For example, rewriting the previous example using labels,
+we obtain:
 \begin{align*}
@@
 this setting, the tensor products become associative and
-commutative, so we can rearrange qubits as needed to
-perform calculations. For our example, we can perform the calculation as
+commutative, allowing us to rearrange qubits as needed for
+calculations. For our example, we can perform the calculation as
 follows:

### 2025-02-01 03labelled.tex
.
update. section 1 - 2 writing checked.

--- 2025-02-01 03labelled.tex
@@
 \paragraph*{}
-By routinely showing that Lemma \ref{lem: axiom sound} holds for the extra rules mentioned above and in Appendix \ref{sec: rewriting rules}, we further claim:
+By routinely showing that Lemma \ref{lem: axiom sound} holds for the extra rules mentioned above and in Appendix \ref{sec: rewriting rules} (regard rules as equational theory), we further claim:
 \begin{theorem}[Soundness and Completeness of Normalization]

### 2025-02-01 01intro_preliminary.tex
.

--- 2025-02-01 01intro_preliminary.tex
@@
 labelled Dirac notation. The main conclusions are that our approach
-outperforms prior work by~\cite{diracdec} and is able to prove complex
+outperforms DiracDec~\cite{diracdec} and is able to prove complex
 examples of reasoning with labelled Dirac notation from the

### 2025-02-01 01intro_preliminary.tex

--- 2025-02-01 01intro_preliminary.tex
@@
 labelled Dirac notation.
-The main conclusions are that our approach outperforms prior work by~\cite{diracdec} and is capable of proving complex examples from the literature, including those from prior work on quantum separation logic~\cite{DBLP:conf/lics/ZhouBHYY21}.
+The main conclusions are that our approach outperforms DiracDec~\cite{diracdec} to reason about plain Dirac notation and is able of proving complex examples from the literature on labelled Dirac notation, including examples from prior work on quantum separation logic~\cite{DBLP:conf/lics/ZhouBHYY21}.
 

### 2025-02-01 02language_design.tex

@@
 
-\subsection{Algorithm to Decide Equivalence}
+\subsection{Algorithm to Prove Equivalence}
 % \gb{I still think that the names first rewritings, second rewriting and de bruijn normalization are not good, in particular, de bruijn normalization is non-standard terminology that will make people familiar with the field cringe.}
 
-The equivalence of Dirac notations are decided by normalization, which transforms equivalent expressions under a set of axioms into the same syntax.
+The equivalence of Dirac notations are proved by normalization, which transforms equivalent expressions under a set of axioms into the same syntax.
 We use an efficient algorithm to perform the normalization totally on $R \cup E$.
@@
 
-As a comparison, the previous work conducts normalization paritally on $R$, and decide axioms in $E$ by iterating through all possible permutations.
+As a comparison, the previous work conducts normalization paritally on $R$, and equivalence by $E$ is proved through checking all possible permutations.
 Our algorithm totally normalize the term, as is illustrated below.

### 2025-02-01 05tool_eval.tex

@@
         \hline
-        LDN-4 & theorem & 0.03 & \( \bra{\Phi}_{(x, y)} M_y \cdot N_x \ket{\Phi}_{(x, y)} = \mathrm{tr}(M^T N) \) \\
+        LDN-4 & theorem & 0.03 & \( M_{r_1}\sum_{i}\ket{(i,i)}_{(r_1,r_2)} = M^T_{r_2}\sum_{i}\ket{(i,i)}_{(r_1, r_2)} \)\\
         LDN-10 & paper~\cite{DBLP:conf/lics/ZhouBHYY21} & 5.17 & \( \mathrm{tr}_{((a',(b,b')),c')}\Big[\mathrm{tr}_{r}\Big(U_{(r,(a,b))} \cdot\Big(|s\>_r\<s|\otimes 
@@
         LDN-14 & circuit & 7.37 & \scriptsize{\( \textsf{CNOT}_{rq}\ket{\textrm{GHZ}}_{pqr} = (\textsf{CNOT}\ket{00})_{rq}\ket{0}_p + (\textsf{CNOT}\ket{11})_{rq}\ket{1}_p \)} \\
-        LDN-18 & program & 1.01 & \( -\ket{0}_q\ket{+-}_{q_1, q_2} = X_{q_2} \ket{0}_q\ket{+-}_{q_1, q_2} \) \\
+        LDN-16 & circuit & 0.08 & \scriptsize{\( M_{prq}\ket{\textrm{GHZ}}_{prq} = N_{rq} \ket{\textrm{GHZ}}_{pqr}, M \triangleq \sum_{ij}\ket{ij}\bra{ij}\otimes U_{ij} \cdots \)} \\
         \hline

### 2025-02-01 main.tex
add figure
update section 3

@@
 \usepackage{tikz}
+\usetikzlibrary{tikzmark}
+\usepackage{annotate-equations}
 
@@
 \usepackage{algpseudocode}
+
+\usepackage{tikz}
+\usetikzlibrary{quantikz2}
 
@@
 \newcommand {\ccnot } {{\mathsf{CCNOT}}}
-\newcommand {\swap } {{\mathsf{SWAP}}}
+\newcommand {\Swap } {{\mathsf{SWAP}}}
 \renewcommand{\matrix}[1]{\begin{bmatrix}#1\end{bmatrix}}
@@
 \begin{abstract}
-Labelled Dirac notation is a formalism commonly used by physicists to
-represent many-body quantum systems and computer scientists to assert
-properties of quantum programs. Labelled Dirac notation is supported
-by a rich equational theory for proving equality between expressions
-of the language. These proofs are typically carried on pen-and-paper,
-and can be exceedingly long and error-prone. We introduce D-Hammer,
-the first tool to supported automated equational proof for labelled
-Dirac notation. The salient features of D-Hammer are: an expressive,
-higher-order, dependently-typed language for labelled Dirac notation;
-an efficient algorithm for normalization; and an optimized
-\CC\ implementation. We evaluate the implementation on representative
-examples from plain and labelled Dirac notation. In the case of plain
-Dirac notation, we show that our implementation significantly
-outperforms DiracDec (Xu et al, POPL'25).
+    Labelled Dirac notation is a formalism commonly used by physicists to represent many-body quantum systems and by computer scientists to assert properties of quantum programs.
+It is supported by a rich equational theory for proving equality between expressions in the language.
+These proofs are typically carried on pen-and-paper, and can be exceedingly long and error-prone.
+We introduce D-Hammer, the first tool to support automated equational proof for labelled Dirac notation.
+The salient features of D-Hammer include: an expressive, higher-order, dependently-typed language for labelled Dirac notation; an efficient normalization algorithm; and an optimized \CC\ implementation.
+We evaluate the implementation on representative examples from both plain and labelled Dirac notation.
+In the case of plain Dirac notation, we show that our implementation significantly outperforms DiracDec (Xu et al., POPL'25).
 

### 2025-02-01 01intro_preliminary.tex
writing improved

--- 2025-02-01 01intro_preliminary.tex
@@
 \end{align*}
-where the RHS of each line is equivalent.  In addition, labelled
-Dirac notation can conveniently describe local measurements, partial
-traces (the state or evolution of subsystems in many-body systems),
-and partial inner products (the representation of partial traces in
-pure states), which are sufficient to handle the mathematical formulas
-of quantum mechanics in many-body systems.
+The right-hand side (RHS) of each line is equivalent, as shown.  
+In addition, labelled
+Dirac notation can conveniently describe local measurements, partial traces (representing the state or evolution of subsystems in many-body systems), and partial inner products (which correspond to partial traces in pure states).
+These capabilities are sufficient for handling the mathematical formulas of quantum mechanics in many-body systems.
 
-Labelled Dirac notation is not only ubiquitous in the description of many-body systems, but also plays an important role in program logics.
-Just as in classical program logic, variable names are used to construct logical formulas instead of describing them as global memory functions, in quantum program logic, variable names are used to label the subsystems on which quantum gates act, rather than lifting them to the global system.
-Indeed, the latter would cause exponential length of the formula w.r.t the number of variables, as discussed in \cite{QSL_Le_2022}.
+Labelled Dirac notation is not only pervasive in the description of many-body systems but also plays a crucial role in quantum program logic.
+Just as classical program logic uses variable names to construct logical formulas, avoiding the need for global memory functions, quantum program logic similarly uses variable names to label the subsystems on which quantum gates act, rather than lifting them to the global system.
+Actually, lifting operations would lead to an exponential increase in formula length relative to the number of variables, as discussed in \cite{QSL_Le_2022}.
 
-In the following,we use the motivating example as the main role in our demonstrations.
-Instead of GHZ states, it is a similar but simpler example on Bell states:
+In the following, we will use this motivating example as the primary focus in our demonstrations. Instead of GHZ states, we use a simpler example involving Bell states:
 \begin{example}
@@
 \end{example}
-Following the explanation on labels, we can consider the global system $(q,r)$ and transform the equation above into the plain Dirac notation:
+As explained earlier regarding labels, we can consider the global system $(q,r)$ and transform the equation above into the plain Dirac notation:
 \begin{align}
@@
 \end{align}
-The next sections introduce the formal language, labelled Dirac notation and the systematical way to reduce labels, showing how an automated system is built and used to solve equalities likewise.
+The following sections introduce the formal language and labelled Dirac notation, and present a systematic approach for reducing labels.
+We will also demonstrate how an automated system can be built and used to solve similar equalities.
 

### 2025-02-01 02language_design.tex
writing improved

@@
 \subsubsection{Denotational Semantics}
-The denotational semantics maps types to sets, and expressions and
-indices to values in the interpretation of types and indices
-respectively. As for other dependently typed systems, all
+Denotational semantics maps types to sets, and expressions and
+indices to values in the interpretation of types and indices,
+respectively. As in other dependently typed systems, all
 interpretations are parametrized by a valuation mapping \( v \), which
 assigns values to variables and indices. We let \( \sem{e}_v \) denote
-the interpretation of an expression $e$ w.r.t.\, a valuation \( v \);
-we use similar notations for types and indices. As usual, we say that
+the interpretation of an expression $e$ w.r.t.\, a valuation \( v \),
+and use similar notations for types and indices. As usual, we say that
 a valuation \( v \) is valid w.r.t.\, a context $\Gamma$ if for every
@@
 The entire axioms are separated into two sets $R$ and $E$.
-$R$ contains the axioms normalized by term rewriting. There are axioms requiring special algorithms, which are collected in the set $E$.
+$R$ contains the axioms normalized by term rewriting. Other axioms requiring special algorithms, which are collected in the set $E$.
 \begin{definition}[axiom set E]
@@
 
-We say an equation $e_1 = e_2$ is provable, written as $\Gamma \vdash e_1 = e_2 : T$, if $\Gamma \vdash e_1 : T$ and $\Gamma \vdash e_2 : T$ are provable, and $e_1 = e_2$ can be deduced in $\Gamma$ using the axioms and equational logic.
-An equation \( e_1 = e_2 \) is valid in context $\Gamma$, written \(
+We say an equation $e_1 = e_2$ is provable, denoted as $\Gamma \vdash e_1 = e_2 : T$, if $\Gamma \vdash e_1 : T$ and $\Gamma \vdash e_2 : T$ are provable, and $e_1 = e_2$ can be deduced in $\Gamma$ using the axioms and equational logic.
+An equation \( e_1 = e_2 \) is valid in context $\Gamma$, written as \(
 \Gamma \vDash e_1 = e_2 \), if \( \sem{e_1}_v = \sem{e_2}_v \) for
@@
 \subsection{Algorithm to Prove Equivalence}
-% \gb{I still think that the names first rewritings, second rewriting and de bruijn normalization are not good, in particular, de bruijn normalization is non-standard terminology that will make people familiar with the field cringe.}
 
-The equivalence of Dirac notations are proved by normalization, which transforms equivalent expressions under a set of axioms into the same syntax.
-We use an efficient algorithm to perform the normalization totally on $R \cup E$.
+The equivalence of Dirac notations is established through normalization, which transforms equivalent expressions into the same syntax under a set of axioms.
+We employ an efficient algorithm to perform the normalization fully on $R \cup E$.
 
@@
 
-Here steps 1 - 3 are term rewriting for $R$. 
-Term rewriting reduces a term repeatedly using a set of rules in the form of $l\ \reduce\ r$.
-The reduction proceeds by matching the subterms with the left-hand side, and replacing it with the right-hand side.
+Step 1 through 3 involve term rewriting for $R$. 
+Term rewriting is the process of repeatedly reducing a term using a set of rules in the form of $l\ \reduce\ r$.
+The reduction works by matching the subterms with the left-hand side of a rule and replacing it with the right-hand side.
 For example, the term $(x\times y) . \ket{t} + \ket{t}$ is matched by the rule $a.K + K\ \reduce\ (a + 1).K$, and is rewritten into $(x\times y + 1) . \ket{t}$.
-Step 1 and 3 uses the same set of rewriting rules in~\Cref{sec: rewriting rules}.
-Step 2 expands variables to their abstract element-wise representation, e.g., $K \ \reduce\ \sum_{i} (\bra{i} \cdot K). \ket{i}$. It improves the decidability with sum.
+Step 1 and 3 use the same set of rewriting rules in~\Cref{sec: rewriting rules}.
+Step 2 expands variables to their abstract element-wise representation, e.g., $K \ \reduce\ \sum_{i} (\bra{i} \cdot K). \ket{i}$, which improves the decidability when dealing with sum.
 
-Steps 4 - 6 are special algorithms to further normalize the axiom set $E$.
-Here the main difficulty is the coexistence of AC-equivalence and SUM-SWAP, so that naive sorting will not convert equivalent terms to the same form.
+Steps 4 through 6 are specialized algorithms designed to further normalize the axiom set $E$.
+The main challenge here is the coexistence of AC-equivalence and SUM-SWAP, which means that naive sorting cannot alwasy convert equivalent terms into the same form.
 For step 4 and 5, the key observation is that in a successive sum expression \( \sum_{i \in s_1} \cdots \sum_{j \in s_n} A \), the names and order of the bound variables \( i, \dots, j \) can be freely permuted.
 Therefore we first ignore bound variables and normalize AC-equivalence by sorting.
-Then the order of summation can be decided accordingly.
-The last step converts functions to de Bruijn indices~\cite{deBruijn1972lambda} to decide $\alpha$-equivalence.
-For detailed descriptions, see~\Cref{sec: decide}.
+Afterward, the order of summation can be decided accordingly.
+The final step uses de Bruijn indices~\cite{deBruijn1972lambda} to resolve $\alpha$-equivalence.
+For further details, refer to~\Cref{sec: decide}.
 
@@
 
-As a comparison, the previous work conducts normalization paritally on $R$, and equivalence by $E$ is proved through checking all possible permutations.
-Our algorithm totally normalize the term, as is illustrated below.
+In contrast, previous work performs normalization only partially on $R$, and proves equivalence by checking all possible permutations according to $E$.
+Our algorithm fully normalize the term, as illustrated below, and is more efficient as a result.
 \[
@@
 \qquad
-\text{(total)}\qquad e_1 \mathop{\twoheadrightarrow}^{R \cup E} e \equiv e \mathop{\twoheadleftarrow}^{R \cup E} e_2
+\text{(full)}\qquad e_1 \mathop{\twoheadrightarrow}^{R \cup E} e \equiv e \mathop{\twoheadleftarrow}^{R \cup E} e_2
 \]

### 2025-02-01 03labelled.tex
writing improved

@@
 
-In this section, we extend the language by allowing quantum variables to indicate the quantum system of vectors and operators it act. 
-As discussed, this allows us to express and reason about the states and operations locally, without referring to the whole system. 
-We further demonstrate how to transform the equivalence problem into that for the plain Dirac notation studied above.
+In this section, we extend the language by allowing quantum variables to indicate the quantum system on which vectors and operators act. 
+As discussed, this enables us to express and reason about the states and operations locally, without referring to the entire system. 
+We further demonstrate how to transform the equivalence problem into one involving the plain Dirac notation studied earlier.
 
 \subsection{Syntax, Typing and Semantics}
-We first introduce the notation of quantum registers for structured variable combinations.
+We begin by introducing the notation of quantum registers for structured variable combinations.
 This is necessary because, unlike assignments for classical variables, unitary transformations on composite systems--a quantum version of assignments--cannot generally be decomposed into separate unitary transformations on individual subsystems.
 Let $\cR$ be the set of quantum variables.
-%The first new notion is quantum registers, and we assume they are constructed from a set $\cR$.
 \begin{definition}[Quantum Register] Register $R$ is inductively generated by
@@
 \end{definition}
-For simplicity, we only allow pairing the registers, which corresponds to the structure of tensor product, i.e., $\cH_{(R_1,R_2)}$ is isomorphic to $\cH_{R_1}\otimes \cH_{R_2}$, which allows us to view the tensor product space as the space of paired registers.
+For simplicity, we restrict register formation to pairings, which corresponds to the structure of tensor product.
+In this context, $\cH_{(R_1,R_2)}$ is isomorphic to $\cH_{R_1}\otimes \cH_{R_2}$, which allows us to view the tensor product space as the space of paired registers.
 
@@
 This requires an additional check on the valid registers--they should not include repeated quantum variables--which is often handled in programming languages via linear types.
-As such, we define the \emph{order-free} variable set of a register as all quantum variables appearing in the register.
+As such, we define the \emph{order-free} variable set of a register as all quantum variables appearing in the register; we let $\var(R)$ denote the variable set of register $R$.
 We use the variable set to establish side conditions of typing valid registers and employ it as the type parameter for annotating labelled Dirac terms.
 
-% \begin{definition}
-% The variable set of a register is defined inductively by:
-% \begin{itemize}
-%     \item $\var(R) = \{r\}$, if $R\equiv r$; or
-%     \item $\var(R) = \var(R_1) \cup\var(R_2)$, if $R\equiv (R_1,R_2)$.
-% \end{itemize}
-% \end{definition}
-
-% \textbf{Remark: } Set operations: $\cup$ for union; $\cap$ for intersection; $\setminus$ for difference. So,
-% % \begin{itemize}
-% $ S_1 \cap S_2 \equiv S_1 \cup S_2 \setminus (S_1 \setminus S_2) \setminus (S_2 \setminus S_1) $.
-% \end{itemize}
-
-% The labelled Dirac notation is an extension on the existing type and syntax.
 Now, we are ready to extend the type syntax and term syntax as follows:
@@
 \end{gather*}
-The first rule says that a paired register is of the product type and its components should be disjoint.
-To lift a plain Dirac notation into the labelled one (line 2), we enforce that the term and register are of the same indices, reflecting the fact the state should be consistant to the corresponding subsystems.
-The third line gives the typing of labelled tensoring with a check whether the component subsystems are disjoint with each other.
-Since $\cR$ is given, let $\sigma_r : \Index$ denote type of $r$, i.e.,  $\Gamma\vdash r : \reg(\sigma_r)$, for simplicity.
+The first rule states that a paired register is of the product type and its components must be disjoint.
+To lift a plain Dirac notation into the labelled version (line 2), we enforce that the term and register share the same indices, reflecting the fact the state should be consistant to the corresponding subsystems.
+The third line provides the typing of labelled tensor product, with a check to ensure that the component subsystems are disjoint from each other.
 
@@
 
-\paragraph*{Semantics.} The labelled Dirac notation handles lifting and ordering for us, and its semantics faithfully capture these details.
-The key points are 1.
-cylindrical extension that lift a ket or bra or operator to larger domain and codomain; 2.
-general composition further employs cylindrical extension that obeys the principle of ``localizing objects as much as possible''~\cite{Zhou2023}.
+\paragraph*{Semantics.} 
+The labelled Dirac notation handles lifting and ordering for us, and its semantics accurately capture these details.
+The key points are: 1.
+cylindrical extension, which lift a ket or bra or operator to larger domain and codomain; 2.
+general composition, which further employs cylindrical extension that obeys the principle of ``localizing objects as much as possible''~\cite{Zhou2023}.
+Since $\cR$ is given, we let $\sigma_r : \Index$ denote type of $r$, i.e.,  $\Gamma\vdash r : \reg(\sigma_r)$, for simplicity.
 
@@
 \end{definition}
-Formally, we equip $\cR$ with a default order, say for example the alphabetical order of names.
+Formally, we equip $\cR$ with a default order, such as the alphabetical order of names.
 For any valid register $R$, there exists the operator $\Swap_R$ that sorts $R$; for example, $\Swap_{(q,p)} \triangleq \sum_{i\in \bU(\sigma_p)}\sum_{j\in \bU(\sigma_q)}|i\>\<j|\otimes |j\>\<i|$.
@@
 $$\mbox{$\sem{\cD(s,s')}_v \triangleq \mathcal{L}(\bigotimes_{j}\cH_{\sem{\sigma_{r'_j}}_v}, \bigotimes_{i}\cH_{\sem{\sigma_{r_i}}_v})$}$$ where $\mathcal{L}$ denotes the set of linear maps, 
-the sets $s = \{r_1,\cdots,r_n\}$ and $s' = \{r'_1,\cdots,r'_m\}$ ($r_i$ and $r'_j$ are sorted).
+$s = \{r_1,\cdots,r_n\}$ and $s' = \{r'_1,\cdots,r'_m\}$ ($r_i$ and $r'_j$ are sorted).
 % we define $\sigma_s\triangleq \sigma_1\times\cdots\times\sigma_n$ if $s$ is non-empty. \label{context: sigma s}
@@
 \paragraph*{2. Rewriting to the normal form.}
-We add three types of rules for dealing with operators on labeled terms, 1) recursively applying them to subterms, 2) push big operators out and 3) eliminate generalized composition and bra-ket pairs.
+We add three types of rules for dealing with operators on labelled terms, 1) recursively applying them to subterms, 2) pushing big operators out and 3) eliminating generalized composition and bra-ket pairs.
 %Other symbols on labelled Dirac notation are also 
@@
   \label{thm: normalization}
-  Assuming $\Gamma\vdash D_1 : \cD(s,s')$, $\Gamma\vdash D_2 : \cD(s,s')$, 
-  and no variables of $\DType(\cdot,\cdot)$ in $D_1,D_2$.
+  Assume $\Gamma\vdash D_1 : \cD(s,s')$, $\Gamma\vdash D_2 : \cD(s,s')$
+  and no variables of $\DType(\cdot,\cdot)$ appear in $D_1,D_2$.
   Let $e_1 = e_2$ be obtained by above normalization procedure on $D_1 = D_2$. Then $\Gamma\vDash D_1 = D_2$ if and only if $\Gamma\vDash e_1 = e_2$.

### 2025-02-01 05tool_eval.tex
writing improved

@@
 \section{Implementation and Case Study}
-The improvement and extensions above concludes in our implementation called D-Hammer, a solver written in \CC.
-It has a parser built by ANTLR4, and scalar reasonings are powered by Mathematica Engine.
-The user can use commands to make definitions and assumptions in the maintained context, conduct the normalization and equivalence checking, and obtain the rewriting trace output.
-It can be used from the command line interactively, or can be integrated into other \CC\ projects as a library.
-This implementation is tested on the benchmark of the previous work and new examples for labelled Dirac notation, demonstrating significant improvement in expressivity and efficiency. 
-The tool and example encodings will be open-source and publicly available.
+The improvement and extensions discussed above culminate in our implementation, called \textbf{D-Hammer}, a solver written in \CC. 
+It features a parser built using ANTLR4, and scalar reasoning is powered by the Mathematica Engine.
+Users can use commands to make definitions and assumptions in the maintained context, conduct the normalization and equivalence checking, and obtain the rewriting trace output.
+D-Hammer can be run interactively from the command line or integrated into other \CC\ projects as a library.
+This tool has been tested on benchmarks from previous work and new examples for labelled Dirac notation, showing significant improvements in expressivity and efficiency. 
+The tool, along with example encodings, will be made open-source and publicly available.
 
@@
 \begin{itemize}
-    \item \texttt{antlr4}: the third-party library for building the parser.
-    \item \texttt{WSTP interface}: the wrapper to link with Mathematica Engine.
-    \item \texttt{ualg}: the framework module for universal algebra, defining basic concepts like terms and substitutions.
-    \item \texttt{dhammer}: the main module containing symbols definitions, type checking, rewriting rules, normalization algorithm and the prover.
-    \item \texttt{example}: an example benchmark for evaluation.
-    \item \texttt{toplevel}: the command line application.
+    \item \texttt{antlr4}: A third-party library for building the parser.
+    \item \texttt{WSTP interface}: A wrapper to link with Mathematica Engine.
+    \item \texttt{ualg}: The framework module for universal algebra, defining basic concepts like terms and substitutions.
+    \item \texttt{dhammer}: The main module containing symbols definitions, type checking, rewriting rules, normalization algorithm and the prover.
+    \item \texttt{example}: An example benchmark for evaluation.
+    \item \texttt{toplevel}: The command line application.
 \end{itemize}
@@
 
-The internal data structure for terms is a pointer-based syntax tree following the function application style:
+The internal data structure for terms follows a pointer-based syntax tree, using the function application style:
 \[
@@
 \]
-The syntax tree can be an identifier, or an application with an identifier as the function head, and several syntax trees as arguments.
-There are several Dirac notation terms and their corresponding syntax trees.
+The syntax tree can either be an identifier, or an application with an identifier as the function head, and several syntax trees as arguments.
+Below are several examples of Dirac notation terms and their corresponding syntax trees.
 \footnotesize{
@@
 % The syntax tree structure is also compatible with the datatype of Mathematica. This improves the interoperability between D-Hammer and the Mathematica system, enabling them to work interleavingly.
-To improve usability, D-Hammer also supports many special notations for terms,  and most Dirac notation terms will be encoded in the natural way.
+To improve usability, D-Hammer also supports many special notations for terms, and most Dirac notation terms is encoded in the natural, intuitive way.
 Here are some examples for the parsing syntax.
@@
 
-In the left hand side of \texttt{\textcolor{NavyBlue}{CheckEq}} command, \texttt{M\_r1} denotes the labelled notation $M_{r_1}$, and \texttt{(phi T)\_(r1, r2)} denotes the entangled state $\ket{\Phi}_{(r_1, r_2)}$.
+In the left-hand side of \texttt{\textcolor{NavyBlue}{CheckEq}} command, \texttt{M\_r1} denotes the labelled notation $M_{r_1}$, and \texttt{(phi T)\_(r1, r2)} denotes the entangled state $\ket{\Phi}_{(r_1, r_2)}$.
 They are connected by a white space, which is parsed into the composition of Dirac notation, and will be reduced into the operator-ket multiplication after typing.
@@
 \subsubsection{CoqQ}
-As to expressibility, D-Hammer encodes 158 examples for the CoqQ part, 4 less than DiracDec.
-This is because the language for DiracDec has the support for projectors $\texttt{fst}$ and $\texttt{snd}$ on basis pairs, satisfying $\texttt{fst} (s, t) = s$ and $\texttt{snd} (s, t) = t$.
-We found this feature is almost not used and removed the support.
-For decidability, we improved the rewriting rules about sum for D-Hammer, and proved several more examples failed by DiracDec.
+In terms of expressibility, D-Hammer encodes 158 examples for the CoqQ part, which is 4 fewer than DiracDec.
+This discrepancy arises because DiracDec supports projectors $\texttt{fst}$ and $\texttt{snd}$ on basis pairs, where $\texttt{fst} (s, t) = s$ and $\texttt{snd} (s, t) = t$.
+However, we found that this feature is rarely used and removed the support in D-Hammer.
+Regarding decidability, we improved the rewriting rules for sums in D-Hammer, and proved several more examples failed by DiracDec.
 \Cref{fig: CoqQ plot} shows a direct comparison of their time efficiency.
-We can observe that for small examples, D-Hammer is slower because of the marginal overhead. 
-When the running time goes up, D-Hammer provides 2 to 40 times of acceleration in average.
-The relationship between efficiency improvement and example characteristics further supports the necessity of our algorithm. 
-We find that examples with great efficiency improvement, e.g. COQQ-129 and COQQ-148 shown in~\Cref{fig: CoqQ plot}, tend to use deeply nested summations, where our algorithm has an effect.
+For small examples, D-Hammer is slower due to the marginal overhead. 
+However, as running time increases, D-Hammer accelerates by an average factor of 2 to 40 times.
+The relationship between efficiency improvement and example characteristics further highlights the necessity of our algorithm. 
+Notably, examples with great efficiency improvement, e.g. COQQ-129 and COQQ-148 shown in~\Cref{fig: CoqQ plot}, tend to use deeply nested sums, where our algorithm has a more pronounced effect.
 
@@
 \subsubsection{Labelled Dirac Notation}
-We present a new set of examples for labelled Dirac notation (LDN).
-These examples, illustrated in~\Cref{fig: labelled examples}, include six representative cases drawn from a variety of sources, such as well-established theorems, research paper results, quantum circuit equivalence, and quantum program verification.
-D-Hammer normalizes them and successfully checks their equivalence using the algorithm in~\Cref{sec: labelled}.
+We present a new set of examples for labelled Dirac notation (LDN), as illustrated in~\Cref{fig: labelled examples}.
+These examples include six representative cases drawn from various sources, such as well-established theorems, research paper results and quantum circuit equivalence.
+D-Hammer successfully normalizes these examples and checks their equivalence using the algorithm outlined in~\Cref{sec: labelled}.
 Among the examples, LDN-12 shows the flexibility in combining labelled Dirac notations.
 LDN-14 shows how to calculate controlled-not gate in different ways.
-%In LDN-18, $q, q_1$ and $q_2$ represent variables for quantum programs, and the equation shows the change of the state after applying $X$ gate on $q_2$. 
-% Experiments show that our algorithm is relatively inefficient for labelled Dirac notations based on qubits.
 
-It is a noteworthy result that D-Hammer solved LDN-10, a highly complex and lengthy example.
-It is a theorem on quantum separation logic from~\cite{DBLP:conf/lics/ZhouBHYY21}, and is challenging to prove even for experts.
-Notice that 7 registers are involved here, and it would be impractical to organize and referring to the subsystems without using labels.
+A particularly noteworthy result is D-Hammer's solution to LDN-10, a highly complex and lengthy example.
+It is a theorem on quantum separation logic from~\cite{DBLP:conf/lics/ZhouBHYY21}, and proving it is challenging even for experts.
+Notably, it involves 7 registers, making it practically impossible to organize and referring to the subsystems without using labels.
 

### 2025-02-01 05tool_eval.tex
.

@@
     \label{fig: CoqQ plot}
+    \vspace{-0.4cm}
 \end{figure}
@@
 D-Hammer successfully normalizes these examples and checks their equivalence using the algorithm outlined in~\Cref{sec: labelled}.
-Among the examples, LDN-12 shows the flexibility in combining labelled Dirac notations.
+Among the examples, LDN-16 is a generalization of Example \ref{example1} and LDN-4 for Example \ref{ex: motivating}. LDN-12 shows the flexibility in combining labelled Dirac notations.
 LDN-14 shows how to calculate controlled-not gate in different ways.
@@
         LDN-14 & circuit & 7.37 & \scriptsize{\( \textsf{CNOT}_{rq}\ket{\textrm{GHZ}}_{pqr} = (\textsf{CNOT}\ket{00})_{rq}\ket{0}_p + (\textsf{CNOT}\ket{11})_{rq}\ket{1}_p \)} \\
-        LDN-16 & circuit & 0.08 & \scriptsize{\( M_{prq}\ket{\textrm{GHZ}}_{prq} = N_{rq} \ket{\textrm{GHZ}}_{pqr}, M \triangleq \sum_{ij}\ket{ij}\bra{ij}\otimes U_{ij} \cdots \)} \\
+        LDN-16 & theorem & 0.08 & \scriptsize{\( M_{prq}\ket{\textrm{GHZ}}_{prq} = N_{rq} \ket{\textrm{GHZ}}_{pqr}, M \triangleq \sum_{ij}\ket{ij}\bra{ij}\otimes U_{ij} \cdots \)} \\
         \hline
@@
     \caption{Part of examples for labelled Dirac notations. See~\Cref{sec: examples for labelled} for the full list.}
+    \vspace{-0.2cm}
     \label{fig: labelled examples}

### 2025-02-01 03labelled.tex
.

@@
 \paragraph*{}
-By routinely showing that Lemma \ref{lem: axiom sound} holds for the extra rules mentioned above and in Appendix \ref{sec: rewriting rules} (regard rules as equational theory), we further claim:
+By routinely showing that Theorem \ref{lem: axiom sound} holds for the extra rules mentioned above and in Appendix \ref{sec: rewriting rules} (regard rules as equational theory), we further claim:
 \begin{theorem}[Soundness and Completeness of Normalization]

### 2025-04-05 03labelled.tex
fix typos
.

@@
         $\sem{K_R}_v = \sem{\Swap_R}_v \cdot \sem{K}_v$; \\
-        $\sem{B_R}_v = \sem{B}_v\cdot \sem{\Swap_R}_v$;\quad
-        $\sem{O_{R_1,R_2}}_v = \sem{\Swap_{R_1}}_v\cdot \sem{O}_v\cdot \sem{\Swap_{R_2}}_v$;
-  \item $\sem{D_1\otimes \cdots \otimes D_n}_v = \sem{\Swap_{s_1,\cdot,s_n}}_v\cdot (\sem{D_1}_v\otimes \cdots \otimes \sem{D_1}_v)\cdot \sem{\Swap_{s'_1,\cdot,s'_n}}_v$;
+        $\sem{B_R}_v = \sem{B}_v\cdot \sem{\Swap_R}_v^\dagger$;\quad
+        $\sem{O_{R_1,R_2}}_v = \sem{\Swap_{R_1}}_v\cdot \sem{O}_v\cdot \sem{\Swap_{R_2}}_v^\dagger$;
+  \item $\sem{D_1\otimes \cdots \otimes D_n}_v = \sem{\Swap_{s_1,\cdot,s_n}}_v\cdot (\sem{D_1}_v\otimes \cdots \otimes \sem{D_1}_v)\cdot \sem{\Swap_{s'_1,\cdot,s'_n}}_v^\dagger$;
   \item $\sem{D_1\cdot D_2}_v = \sem{\cl(D_1, s_2\backslash s'_1)}_v\cdot \sem{\cl(D_2, s'_1\backslash s_2)}_v$.
@@
 
-\paragraph*{2. Rewriting to the normal form.}
+\paragraph*{2. Rewriting to normal form.}
 We add three types of rules for dealing with operators on labelled terms, 1) recursively applying them to subterms, 2) pushing big operators out and 3) eliminating generalized composition and bra-ket pairs.

### 2025-04-05 05tool_eval.tex
.

@@
 \section{Implementation and Case Study}
-The improvement and extensions discussed above culminate in our implementation, called \textbf{D-Hammer}, a solver written in \CC. 
-It features a parser built using ANTLR4, and scalar reasoning is powered by the Mathematica Engine.
-Users can use commands to make definitions and assumptions in the maintained context, conduct the normalization and equivalence checking, and obtain the rewriting trace output.
-D-Hammer can be run interactively from the command line or integrated into other \CC\ projects as a library.
-This tool has been tested on benchmarks from previous work and new examples for labelled Dirac notation, showing significant improvements in expressivity and efficiency. 
-The tool, along with example encodings, will be made open-source and publicly available.
+We present \textbf{D-Hammer}, an open-source and publicly
+available\footnote{{\textcolor{red}{URL}}} implementaion of our
+approach.  \textbf{D-Hammer} is an equational prover for Labelled
+Dirac notation written in \CC.  It features a parser built using
+ANTLR4, and scalar reasoning is powered by the Mathematica
+Engine. Users can use commands to make definitions and assumptions in
+the maintained context, conduct the normalization and equivalence
+checking, and obtain the rewriting trace output.  D-Hammer can be run
+interactively from the command line or integrated into other
+\CC\ projects as a library.
 
@@
 
+The results show significant performance improvements. The enhanced
+efficiency primarily results from algorithmic improvements rather than
+the shift to a C++ implementation. This is supported by the fact that
+efficiency gains are more significant for larger examples, as can be
+observed in \Cref{fig: CoqQ plot}.
+
 \subsubsection{Textbook (QCQI)}
-The 18 examples of QCQI textbook equations from DiracDec are all encoded and solved. The running time are short and close because these examples are simple and direct.
+As a warm-up, we consider 18 examples from Nielsen and Chuang's
+classic texbook~\cite{nielsen2010quantum}. All examples can be encoded
+in DiracDec and D-Hammer and are solved very very efficiently.
 
@@
 \subsubsection{CoqQ}
-In terms of expressibility, D-Hammer encodes 158 examples for the CoqQ part, which is 4 fewer than DiracDec.
-This discrepancy arises because DiracDec supports projectors $\texttt{fst}$ and $\texttt{snd}$ on basis pairs, where $\texttt{fst} (s, t) = s$ and $\texttt{snd} (s, t) = t$.
-However, we found that this feature is rarely used and removed the support in D-Hammer.
-Regarding decidability, we improved the rewriting rules for sums in D-Hammer, and proved several more examples failed by DiracDec.
-\Cref{fig: CoqQ plot} shows a direct comparison of their time efficiency.
-For small examples, D-Hammer is slower due to the marginal overhead. 
-However, as running time increases, D-Hammer accelerates by an average factor of 2 to 40 times.
-The relationship between efficiency improvement and example characteristics further highlights the necessity of our algorithm. 
-Notably, examples with great efficiency improvement, e.g. COQQ-129 and COQQ-148 shown in~\Cref{fig: CoqQ plot}, tend to use deeply nested sums, where our algorithm has a more pronounced effect.
+As a more substantial example, we consider the examples from
+CoqQ~\cite{Zhou2023}, an extensive formalization of quantum
+information theory and quantum programming languages in the
+Coq proof assistant.
+
+CoqQ has been used as the main benchmark for evaluating DiracDec.
+Specifically, \cite{diracdec} isolates 162 statements in CoqQ that
+are in scope of the DiracDec language.
+
+We have ported 158 out of 162 examples to D-Hammer. The remaining 4
+examples uses projectors $\texttt{fst}$ and $\texttt{snd}$ on basis
+pairs, where $\texttt{fst} (s, t) = s$ and $\texttt{snd} (s, t) =
+t$. However, we found that this feature is rarely used and removed the
+support in D-Hammer. Note that the omission of the projection rules
+has a limited influence on the performance evaluation, because the
+efficiency of the 158 examples is not affected by projection rules.
+
+D-Hammer verifies the whole 158 examples in less than 10 seconds,
+whereas DiracDec verifies 156/162 examples in more than 45 seconds.
+\Cref{fig: CoqQ plot} shows a direct comparison of the efficiency of
+the two tools. We observe that D-Hammer is slower than DiracDec on
+small examples, due to marginal overhead, but becomes faster by an
+average factor of 2 to 40 times as the running time of examples
+increases. This non-linear growth suggests that efficiency gains
+result from algorithmic improvements rather than the shift to a C++
+implementation. Furthermore, examples with great improvements,
+e.g. COQQ-129 and COQQ-148 shown in~\Cref{fig: CoqQ plot}, tend to use
+deeply nested sums, for which our algorithm is more efficient.
 

### 2025-04-05 06related_conclusion.tex
.

@@
 \section{Related Work}
-Categorical quantum mechanics (CQM)~\cite{DBLP:conf/lics/AbramskyC04}
-is a category-theoretic foundation for quantum mechanics. CQM is the
-basis of the ZX calculus~\cite{DBLP:conf/icalp/CoeckeD08}, a popular
-graphical calculus for quantum states~\cite{vandewetering2020zx}. A
-key aspect of the ZX calculus and related graphical calculi is that
-they can be equipped with powerful rewriting systems. These systems
-support a form of graph rewriting, i.e.\, they exploit the graph
-representation of quantum states. There exists several completeness
-results for ZX-like rewriting systems, as well as some practical
-tools, such as PyZX~\cite{kissinger2019pyzx}. Similarly, there exist
-completeness results and tools for the sum-over-paths representation
-of quantum states~\cite{amy2018towards,amy2023complete}.
-
-Automata-based approaches has also been successfully applied for reasoning about quantum circuit~\cite{AutoQ_pldi_2023}.
-Instead of representing a set of  $n$-qubit states in exponential size, \cite{AutoQ_pldi_2023} encode it as (finite) tree automata with much less (often $\mathit{poly}(n)$) transitions.
-They further established transformers to implement the semantics of quantum gates over this representation, providing better scalability compared to other approaches.
-It is developed as a practical tool \textsc{autoq} with the capabilities for symbolic reasoning~\cite{AutoQ2023}.
-More recently, \cite{AutoQ_popl2025} further explore level-synchronized tree automata to encoding state, a promising model for parameterized verification.
 
@@
 reasoning about complex, many-body, quantum states. However, there are
-several other important differences. First, our language leverages the
-power of higher-order dependent type theory. In particular, we
-leverage higher-order functions to provide a compact and expressive
+several other important differences. First, our language leverages
+higher-order functions to provide a compact and expressive
 representation of big operators. Second, our language adopts AC
@@
 a consequence, its behavior, and in particular, its typing rules are
-constrained by the lack of typing in Mathematica. In contrast, D-Hammer
-is designed as a separate tool; as a consequence, it benefits from an
-improved representation of terms (e.g.\, AC operators with indefinite
-arity), a more expressive type system (e.g.\, dependent types), and a
-more efficient rewriting engine. D-Hammer still relies on Mathematica
-to reason about functions that are not natively supported by rewriting
-rules, but these interactions are constrained and do not have negative
-effects on the overall efficiency of the system. This is reflected in
-our experimental comparison of D-Hammer and DiracDec, which shows how
-the former outperforms the latter.
+constrained by the lack of typing in Mathematica. In contrast,
+D-Hammer is designed as a separate tool; as a consequence, it benefits
+from an improved representation of terms (e.g.\, AC operators with
+indefinite arity), a more expressive type system (e.g.\, dependent
+types), and a more efficient rewriting engine. D-Hammer still relies
+on Mathematica to reason about functions that are not natively
+supported by rewriting rules, but these interactions are constrained
+and do not have negative effects on the overall efficiency of the
+system. This is reflected in our experimental comparison of D-Hammer
+and DiracDec, which shows how the former outperforms the latter.
+
+\paragraph*{Comparison with ZX Calculus}
+The ZX calculus~\cite{DBLP:conf/icalp/CoeckeD08,vandewetering2020zx}
+is a popular graphical calculus for reasoning about quantum states.
+The foundations of the ZX calculus are grounded in categorical quantum
+mechanics (CQM)~\cite{DBLP:conf/lics/AbramskyC04}, which provides an
+category-theoretic foundation for quantum mechanics.  A key aspect of
+the ZX calculus and related graphical calculi is that they can be
+equipped with powerful rewriting systems. These systems support a form
+of graph rewriting, i.e.\, they exploit the graph representation of
+quantum states. Two notable properties of these systems are:
+\begin{itemize}
+\item completeness:
+\item efficiency:
+\end{itemize}
+Both Labelled Dirac Notation (LDN) and the ZX calculus share the use
+of rewriting techniques. However, it is difficult to compare the ZX
+calculus and LDN. First, they have different scopes: the ZX calculus
+mainly focuses on quantum circuits, whereas LDN aims to provide
+general support for quantum physics. This difference in scope is
+reflected in the underlying languages: for instance, many works in ZX
+calculus consider concrete sets of quantum gates, whereas LDN
+emphasizes the use of big operators, and does not attempt to provide
+specialized support for a particular set of quantum gates. As a
+result, many of the examples handled by D-Hammer cannot be formalized
+in existing ZX-calculus tools. On the contrary, D-Hammer is not
+competitive on the examples that can be verified by ZX-calculus.
+
+\gb{give a few examples in terms of size and time}
+
+\paragraph*{Comparison with path-over sums approach}
+Similarly, there exist
+completeness results and tools for the sum-over-paths representation
+of quantum states~\cite{amy2018towards,amy2023complete}.
+
+\paragraph*{Comparison with equivalence checking}
+
+Automata-based approaches has also been successfully applied for
+reasoning about quantum circuits~\cite{AutoQ_pldi_2023}. Instead of
+representing a set of $n$-qubit states in exponential size,
+\cite{AutoQ_pldi_2023} encode it as (finite) tree automata with much
+less (often $\mathit{poly}(n)$) transitions.  They further established
+transformers to implement the semantics of quantum gates over this
+representation, providing better scalability compared to other
+approaches. It is developed as a practical tool \textsc{autoq} with
+the capabilities for symbolic reasoning~\cite{AutoQ2023}. More
+recently, \cite{AutoQ_popl2025} further explore level-synchronized
+tree automata to encoding state, a promising model for parameterized
+verification.
+
+\paragraph*{Canonical forms in multi-body quantum physics}
+
+\paragraph*{Comparison with egraphs}
+Our algorithm is based on term rewriting. However, it is a challenge
+to device well-behaved and efficient term rewriting for (labelled)
+Dirac notation. An alternative to term rewriting would be to use
+equality saturation~\cite{DBLP:journals/pacmpl/WillseyNWFTP21}, a
+powerful equational reasoning technique that does not require
+existence of normal forms. Equality saturation may be particularly
+useful when considering further extensions of labelled Dirac notation.
 

### 2025-04-05 06related_conclusion.tex
.

@@
 category-theoretic foundation for quantum mechanics.  A key aspect of
-the ZX calculus and related graphical calculi is that they can be
-equipped with powerful rewriting systems. These systems support a form
-of graph rewriting, i.e.\, they exploit the graph representation of
-quantum states. Two notable properties of these systems are:
+the ZX calculus and related graphical calculi is that they support a
+form of graph rewriting. Two notable properties of these rewriting
+systems are:
 \begin{itemize}

### 2025-04-06 02language_design.tex
.

@@
 
-\begin{lemma}[Soundness of equational theory]\label{lem: axiom sound}
+\begin{theorem}[Soundness of equational theory]\label{lem: axiom sound}
 If \( \Gamma \vdash e_1 = e_2 :T \) then \( \Gamma \vDash e_1 = e_2\).
-\end{lemma}
+\end{theorem}
 % \begin{proof}
@@
 Step 1 and 3 use the same set of rewriting rules in~\Cref{sec: rewriting rules}.
-Step 2 expands variables to their abstract element-wise representation, e.g., $K \ \reduce\ \sum_{i} (\bra{i} \cdot K). \ket{i}$, which improves the decidability when dealing with sum.
+Step 2 expands variables to their abstract element-wise representation, e.g., $K \ \reduce\ \sum_{i} (\bra{i} \cdot K). \ket{i}$, which is useful when reasoning about sums.
 
@@
 Therefore we first ignore bound variables and normalize AC-equivalence by sorting.
-Afterward, the order of summation can be decided accordingly.
+Afterward, the order of summation can be established accordingly.
 The final step uses de Bruijn indices~\cite{deBruijn1972lambda} to resolve $\alpha$-equivalence.

### 2025-04-06 06related_conclusion.tex
.

@@
 \paragraph*{Comparison with ZX Calculus}
-The ZX calculus~\cite{DBLP:conf/icalp/CoeckeD08,vandewetering2020zx}
-is a popular graphical calculus for reasoning about quantum states.
-The foundations of the ZX calculus are grounded in categorical quantum
-mechanics (CQM)~\cite{DBLP:conf/lics/AbramskyC04}, which provides an
-category-theoretic foundation for quantum mechanics.  A key aspect of
-the ZX calculus and related graphical calculi is that they support a
-form of graph rewriting. Two notable properties of these rewriting
-systems are:
-\begin{itemize}
-\item completeness:
-\item efficiency:
-\end{itemize}
-Both Labelled Dirac Notation (LDN) and the ZX calculus share the use
-of rewriting techniques. However, it is difficult to compare the ZX
+The ZX calculus~\cite{DBLP:conf/icalp/CoeckeD08,coecke2011interacting}
+is a graphical calculus for quantum states. A main appeal of the ZX
+calculus is that its foundations are grounded in categorical quantum
+mechanics~\cite{DBLP:conf/lics/AbramskyC04}, a powerful framework for
+modeling and reasoning about quantum physics. Another main appeal of
+the ZX calculus is that ZX diagrams have an operational interpretation
+using graph rewriting. There is a large body of work that studies
+properties of these rewriting systems, and in particular completeness
+(a set of rules is complete if it can prove all valid identities) and
+minimality (a complete set of rules is minimal if removing a rule
+leads to incompleteness). Two main proof techniques for completeness
+are via termination, or via interpretation. In the first case, one
+shows that a rewriting system has unique normal forms and that two
+expressions are semantically equivalent iff they have the same normal
+form---there exists many relaxations of this result---whereas in the
+second case one shows completeness by exhibiting a well-behaved
+translation to another system for which completeness holds. Both proof
+techniques have been used to prove completeness for multiple settings,
+including the Clifford fragment~\cite{backens2014zx}, the
+Toffoli+Hadamard fragment~\cite{hadzi15zx}, the Clifford+T
+fragment~\cite{jeandel2018complete} and the (qubit) universal
+fragment~\cite{HNW18,JeandelPV18beyond}. Subsequent works generalize
+completeness to qudits~\cite{Poor2023} for a calculus know as ZXW
+which encompasses the ZX and ZW calculi~\cite{zwcalculus}. We refer
+the reader to~\cite{vandewetering2020zx} for a (relatively) recent
+historical and technical account of completeness results.
+
+Because they are both grounded in rewriting techniques, there are many
+similarities between prior work on the ZX calculus and our own work on
+labelled Dirac notation. We highlight some key differences. First,
+there exist a number of differences between our language and the
+concrete fragments of the ZX-calculus for which completeness has been
+studied. For instance, our language features big operators, and
+operators for dual spaces. A second difference is that we do not claim
+completeness: our main theorem shows a reduction from Labelled Dirac
+notation to Dirac notation, for which completeness is only partially
+known.
+
+it is difficult to compare the ZX
 calculus and LDN. First, they have different scopes: the ZX calculus

### 2025-04-06 06related_conclusion.tex
.

@@
 completeness to qudits~\cite{Poor2023} for a calculus know as ZXW
-which encompasses the ZX and ZW calculi~\cite{zwcalculus}. We refer
-the reader to~\cite{vandewetering2020zx} for a (relatively) recent
-historical and technical account of completeness results.
+which encompasses the ZX and ZW calculi~\cite{zwcalculus}, quantum
+circuits~\cite{DBLP:conf/lics/ClementHMPV23} or optical quantum
+circuits~\cite{clement_et_al:2022}. We refer the interested reader
+to~\cite{vandewetering2020zx} for a historical and technical account
+of completeness results up to 2020.
 
@@
 notation to Dirac notation, for which completeness is only partially
-known.
+known. From a practical point of view, many of the examples handled by
+D-Hammer cannot be formalized in existing ZX-calculus tools. On the
+contrary, D-Hammer is not competitive on the examples that can be
+verified by ZX-calculus. \gb{give a few examples in terms of size and time}
 
-it is difficult to compare the ZX
-calculus and LDN. First, they have different scopes: the ZX calculus
-mainly focuses on quantum circuits, whereas LDN aims to provide
-general support for quantum physics. This difference in scope is
-reflected in the underlying languages: for instance, many works in ZX
-calculus consider concrete sets of quantum gates, whereas LDN
-emphasizes the use of big operators, and does not attempt to provide
-specialized support for a particular set of quantum gates. As a
-result, many of the examples handled by D-Hammer cannot be formalized
-in existing ZX-calculus tools. On the contrary, D-Hammer is not
-competitive on the examples that can be verified by ZX-calculus.
+\paragraph*{Comparison with quantum circuit equivalence checking tools}
+A main application of the ZX-calculus is to check equivalence of
+quantum circuits; for instance, PyZX~\cite{kissinger2020pyzx} is a
+popular Python library for automating equational reasoning in the
+ZX-calculus.
 
-\gb{give a few examples in terms of size and time}
-
-\paragraph*{Comparison with path-over sums approach}
-Similarly, there exist
-completeness results and tools for the sum-over-paths representation
-of quantum states~\cite{amy2018towards,amy2023complete}.
-
-\paragraph*{Comparison with equivalence checking}
-
-Automata-based approaches has also been successfully applied for
-reasoning about quantum circuits~\cite{AutoQ_pldi_2023}. Instead of
-representing a set of $n$-qubit states in exponential size,
-\cite{AutoQ_pldi_2023} encode it as (finite) tree automata with much
-less (often $\mathit{poly}(n)$) transitions.  They further established
-transformers to implement the semantics of quantum gates over this
-representation, providing better scalability compared to other
-approaches. It is developed as a practical tool \textsc{autoq} with
-the capabilities for symbolic reasoning~\cite{AutoQ2023}. More
-recently, \cite{AutoQ_popl2025} further explore level-synchronized
-tree automata to encoding state, a promising model for parameterized
-verification.
+There are many other tools for equivalence checking of quantum
+circuits. Amy~\cite{amy2018towards} uses the path-sums formalism
+formalism to check circuits with thousands of T gates. A series of
+works~\cite{AutoQ_pldi_2023,AutoQ2023,AutoQ_popl2025} explores
+automata-based approaches to verify quantum circuits at scale. Their
+main idea is to encode circuits as a compact, finite tree automata,
+often with $\mathit{poly}(n)$ transitions (instead of $2^n$
+transitions), where $n$ is the number of qubits of the state. By
+providing the semantics of quantum gates over this representation,
+they achieve practical verification. Their tool \textsc{autoq} is able
+to verify circuits with over 100,00 gates, and was recently extended
+to support parametrized verification.
 
 \paragraph*{Canonical forms in multi-body quantum physics}
+Canonical forms play a fundamental role in quantum physics. For
+instance, \cite{perezgarcia2007,RevModPhys.93.045003,Acuaviva_2023}
+discuss canonical forms of Matrix Product States (MPS) and Tensor
+Networks respectively. An exciting direction for further work is to
+further develop automated deduction techniques for quantum physics.
 

### 2025-04-07 02language_design.tex
merge

@@
   remain close to standard Dirac notation;
-\item higher-order, indexed (a.k.a.\, weak dependent) types. It allows
+\item higher-order, indexed (a.k.a.\, weakly dependent) types. It allows
   to formally encode defined symbols like transpose or trace, which
@@
     \item \textbf{scalar}: \( 0 \), \( 1 \), \( e_1 \times \cdots \times e_n \) and \( e^* \) are symbols for scalars.
-    \( \delta_{e_1, e_2} \) compares where two basis are the same and evaluates to $1$ or $0$ accordingly.
+    \( \delta_{e_1, e_2} \) compares whether two basis are the same and evaluates to $1$ or $0$ accordingly.
     \item \textbf{Dirac constant}: zero ket, zero bra, zero operator and identity operator.

### 2025-04-07 03labelled.tex
.
update the proof

@@
-\newcommand {\cD } {{\mathcal{D}}}
-\newcommand {\cl } {{\mathit{cl}}}
-
 % Labelled Dirac Notation
@@
 \end{equation}
-and thus it is equivalent to decide the equivalence of additions of subterms of:
+and thus it is equivalent to prove the equivalence of additions of subterms of:
 \begin{equation}
@@
 \paragraph*{}
-By routinely showing that Theorem \ref{lem: axiom sound} holds for the extra rules mentioned above and in Appendix \ref{sec: rewriting rules} (regard rules as equational theory), we further claim:
+We claim that Theorem \ref{lem: axiom sound} holds for the extra rules mentioned above and in Appendix \ref{sec: rewriting rules} (regard rules as equational theory), as well as the relative completeness of the normalization with respect to plain Dirac notation:
 \begin{theorem}[Soundness and Completeness of Normalization]
@@
 \end{theorem}
+\begin{hproof}
+  \lz{
+  Checking the soundness of every rule is routine. 
+  For relative completeness, it is sufficient to show that the normalization steps can succeed for every labelled Dirac equality. To this end, we give 
+  concrete steps to rewrite every labelled Dirac notation into forms of Eqn.
+(1) by induction on the structure of the notation.
+Further noticing that reordering of labelled basis preserves the type and semantics and thus yield Eqn.
+(4).
+The detailed proof can be found in Appendix \ref{sec: rewriting rules}.
+  }
+
+  \lz{Note that since we take a special algorithm (procedure) to rewrite equations, the relative completeness is weaker than showing the confluence and termination of all these rewriting rules.}
+\end{hproof}
+
 % \begin{proof}
@@
 
-% In this stage, we only need to check the scalars to decide the equivalence of labelled Dirac notation.
+% In this stage, we only need to check the scalars to prove the equivalence of labelled Dirac notation.
 

### 2025-04-07 05tool_eval.tex
.

@@
 We present \textbf{D-Hammer}, an open-source and publicly
-available\footnote{{\textcolor{red}{URL}}} implementaion of our
+available\footnote{https://github.com/LucianoXu/D-Hammer} implementaion of our
 approach.  \textbf{D-Hammer} is an equational prover for Labelled
@@
 
-\section{Evaluation}
+\section{Evaluation}\label{sec:eval}
 We evaluate D-Hammer on several example sets, and make a comparison with the previous tool DiracDec~\cite{diracdec}.
@@
 efficiency primarily results from algorithmic improvements rather than
-the shift to a C++ implementation. This is supported by the fact that
+the shift to a \CC implementation. This is supported by the fact that
 efficiency gains are more significant for larger examples, as can be
@@
 \end{figure}
+
+\subsubsection*{Equivalence checking of circuits}
+\gb{TODO}

### 2025-04-08 01intro_preliminary.tex
.

@@
 \subsubsection*{Contributions}
-This paper presents D-Hammer, an automated tool for reasoning about labelled Dirac notation.
-D-Hammer uses a rich, dependently typed language to formalize labelled Dirac notation and supports common idioms for describing quantum systems, including big operators of the form \( \sum_{i \in I} a_i \ket{\phi_i}_S \) to represent indexed superpositions of states.
-The semantics of typable expressions are given in terms of Hilbert spaces, tailored to interpret the tensor product as an AC symbol.
-We leverage this interpretation to define a rich equational theory for labelled Dirac notation and prove its soundness with respect to our denotational semantics.
-Finally, we define an efficient normalization procedure to prove equivalence between two expressions.
-We then evaluate our procedure with respect to examples from the literature.
-Our evaluation covers examples based on plain Dirac notation, as well as examples based on
-labelled Dirac notation.
-The main conclusions are that our approach outperforms DiracDec~\cite{diracdec} to reason about plain Dirac notation and is able of proving complex examples from the literature on labelled Dirac notation, including examples from prior work on quantum separation logic~\cite{DBLP:conf/lics/ZhouBHYY21}.
+This paper presents D-Hammer, an automated tool for reasoning about
+labelled Dirac notation. D-Hammer uses a rich, dependently typed
+language to formalize labelled Dirac notation and supports common
+idioms for describing quantum systems, including big operators of the
+form \( \sum_{i \in I} a_i \ket{\phi_i}_S \) to represent indexed
+superpositions of states. The semantics of typable expressions are
+given in terms of Hilbert spaces, tailored to interpret the tensor
+product as an AC symbol. We leverage this interpretation to define a
+rich equational theory for labelled Dirac notation and prove its
+soundness with respect to our denotational semantics. Finally, we
+define an efficient normalization procedure to prove equivalence
+between two expressions. We then evaluate our procedure with respect
+to examples from the literature. Our evaluation covers examples in
+plain and labelled Dirac notation. The main conclusions are that our
+approach outperforms DiracDec~\cite{diracdec} to reason about plain
+Dirac notation and is able of proving complex examples from the
+literature on labelled Dirac notation, including examples from prior
+work on quantum separation logic~\cite{DBLP:conf/lics/ZhouBHYY21}.
+For completeness, we also evaluate D-Hammer on examples from the
+literature on equivalence checking of (parametrized) quantum
+circuits.\gb{want to say our tool is not competitive?}
 
@@
 
-%% Recently, Xu et al.~\cite{diracdec} proposed a theory for deciding the equivalence of Dirac notation, alongside a prototype implementation in Mathematica called DiracDec.
-They demonstrated that the equivalence of basic Dirac notation is decidable.
+%% Recently, Xu et al.~\cite{diracdec} proposed a theory for proving the equivalence of Dirac notation, alongside a prototype implementation in Mathematica called DiracDec.
+They demonstrated that the equivalence of basic Dirac notation is ??.
 Their algorithm, based on a pure term rewriting system, has been proven to be confluent and terminating.
@@
 %% One challenge is the efficiency of the algorithm when dealing with equivalences beyond the scope of term rewriting.
-DiracDec decides the entire equational theory by rewriting modulo \( E \), where \( E \) represents a set of axioms that cannot be decided by normalization alone, such as the associativity and commutativity (AC) of certain function symbols.
-DiracDec uses a direct but inefficient algorithm to decide these axioms, which searches through all possible permutations and exhibits factorial complexity.
+DiracDec proves the entire equational theory by rewriting modulo \( E \), where \( E \) represents a set of axioms that cannot be proved by normalization alone, such as the associativity and commutativity (AC) of certain function symbols.
+DiracDec uses a direct but inefficient algorithm to validate these axioms, which searches through all possible permutations and exhibits factorial complexity.
 This inefficiency becomes particularly evident in "computational examples" containing many AC symbols, which are time-consuming to process.
@@
 %% \begin{itemize}
-%%     \item An efficient algorithm for deciding the equational theories in \( E \), based on equivalence checking through normalization, with the normal form for \( E \) being obtained via sorting.
-%%     \item Support for constant register labels, and reducing the equivalence decision of labelled Dirac notation to the unlabelled case.
+%%     \item An efficient algorithm for proving the equational theories in \( E \), based on equivalence checking through normalization, with the normal form for \( E \) being obtained via sorting.
+%%     \item Support for constant register labels, and reducing the equivalence  labelled Dirac notation to the unlabelled case.
 %%     \item A more user-friendly \CC\ solver, D-Hammer, featuring an abstract language and typing system. We also support the definition of symbols (e.g., transpose and trace) using function syntax.
@@
 %% We evaluated D-Hammer against the DiracDec benchmark and new examples involving labelled Dirac notation.
-The results show significant improvements in both decidability and efficiency compared to DiracDec.
+The results show significant improvements in both coverage and efficiency compared to DiracDec.
 
-%% % D-Hammer successfully decides all the examples that are expressable in its language, including those failed by DiracDec because of complexity or insufficient decision power.
+%% % D-Hammer successfully proves all the examples that are expressable in its language, including those failed by DiracDec because of complexity or insufficient proving power.
 
@@
 operation $\cnot$ to the global system. When $\cnot$ acts on $r,q$, it is
-straightforward, as $r,q$ consistent with the chosen order. We only need
+straightforward, as $r$ and $q$ are consistent with the chosen order. We only need
 to tensor it with an identity operator $I$ on $p$, i.e., $(I\otimes

### 2025-04-08 02language_design.tex
update the proof
.

@@
 \end{theorem}
+The proof of soundess is standard: we prove that all axioms are sound,
+and that all proof rules are sound.
+
 % \begin{proof}
@@
 Next, we formalize the motivating example~\Cref{ex: motivating} in
-plain Dirac notation.
+Dirac notation.
 \begin{example}[Motivating Example Formalization]
@@
 
-\subsection{Algorithm to Prove Equivalence}
+\subsection{Normalization}
 
@@
     \annotate[yshift=0em]{below}{swapsum}{\scriptsize{sum swapping}}
-    \caption{A nomralization outline for the left-hand side of~\Cref{eq: motivating plain}.
+    \caption{A normalization outline for the left-hand side of~\Cref{eq: motivating plain}.
 Matched subterms are marked with colors.

### 2025-04-08 03labelled.tex
.

--- 2025-04-05 03labelled.tex
@@
 \paragraph*{Typing rules.}
-There are various rules for computing types and checking vadility of registers and labelled terms. Here we display some of the rules and see Appendix for all the rest.
+There are various rules for computing types and checking vadility of
+registers and labelled terms. Here we display some of the rules and
+refer the reader to Appendix for the full set of rules.\gb{which appendix? add section}
 \begin{gather*}
@@
 
-\subsection{Normalization}
-The normalization procedure of labelled Dirac notation allows us to eliminate all labels and thus translate the equality of labelled terms to the equality of plain Dirac terms.
-This is achieved by the following three steps:
+\subsection{Elimination of labels}
+It is possible to eliminate labels from labelled Dirac expressions and
+thus to transform any equation in Labelled Dirac notation into an
+equation in plain Dirac notation. This is achieved by the following
+three steps:
 
@@
 \paragraph*{}
-By routinely showing that Theorem \ref{lem: axiom sound} holds for the extra rules mentioned above and in Appendix \ref{sec: rewriting rules} (regard rules as equational theory), we further claim:
-\begin{theorem}[Soundness and Completeness of Normalization]
+
+The procedure to eliminate labels is sound and complete, in the sense
+that expressions in Labelled Dirac notation are equivalent iff their
+translations to plain Dirac notation are equivalent.
+\begin{theorem}[Label Elimination]
   \label{thm: normalization}
-  Assume $\Gamma\vdash D_1 : \cD(s,s')$, $\Gamma\vdash D_2 : \cD(s,s')$
-  and no variables of $\DType(\cdot,\cdot)$ appear in $D_1,D_2$.
-  Let $e_1 = e_2$ be obtained by above normalization procedure on $D_1 = D_2$. Then $\Gamma\vDash D_1 = D_2$ if and only if $\Gamma\vDash e_1 = e_2$.
-  %All rewriting rules for labelled Dirac notations listed here and in Appendix are sound, i.e., for any interpretation $\sem{\cdot}$, $D\reduce D'$ implies $\sem{D} = \sem{D'}$.
+  Assume $\Gamma\vdash D_1 : \cD(s,s')$, $\Gamma\vdash D_2 :
+  \cD(s,s')$ and no variables of $\DType(\cdot,\cdot)$ appear in
+  $D_1,D_2$.  Let $e_1 = e_2$ be obtained by above normalization
+  procedure on $D_1 = D_2$.  Then $e!=e_2$ is an equation in plain
+  Dirac notation and $\Gamma\vDash D_1 = D_2$ if and only if
+  $\Gamma\vDash e_1 = e_2$.
 \end{theorem}
+The idea of the proof is as follows: first, we define a set of proof
+rules to rewrite every labelled Dirac notation into the form of
+Eqn. (1). We prove that each rule is sound w.r.t.\, semantics, and
+that every labelled Dirac expression can be rewritten to an expression
+of the form of Eqn. (1); the proof is by induction on the structure of
+the expression. Next, we show that reordering of labelled basis
+preserves the type and semantics and thus yield expressions of the
+form of Eqn. (4), as desired. Details appear in
+Appendix~\ref{app:label_elim}.\gb{check proof is really there}
+%% \begin{hproof}
+%%   \lz{
+%%   Checking the soundness of every rule is routine. 
+%%   For relative completeness, it is sufficient to show that the normalization steps can succeed for every labelled Dirac equality. To this end, we give 
+%%   concrete steps to  by induction on the structure of the notation.
+Further noticing that reordering of labelled basis preserves the type and semantics and thus yield Eqn.
+(4).
+The detailed proof can be found in Appendix \ref{sec: rewriting rules}.
+%%   }
+
+%%   \lz{Note that since we take a special algorithm (procedure) to rewrite equations, the relative completeness is weaker than showing the confluence and termination of all these rewriting rules.}
+%% \end{hproof}
+
+%% We claim that Theorem \ref{lem: axiom sound} holds for the extra rules mentioned above and in Appendix \ref{sec: rewriting rules} (regard rules as equational theory), as well as the relative completeness of the normalization with respect to plain Dirac notation:
+
 % \begin{proof}

### 2025-04-08 05tool_eval.tex
.

@@
 \section{Implementation and Case Study}
-We present \textbf{D-Hammer}, an open-source and publicly
-available\footnote{https://github.com/LucianoXu/D-Hammer} implementaion of our
-approach.  \textbf{D-Hammer} is an equational prover for Labelled
-Dirac notation written in \CC.  It features a parser built using
-ANTLR4, and scalar reasoning is powered by the Mathematica
-Engine. Users can use commands to make definitions and assumptions in
-the maintained context, conduct the normalization and equivalence
-checking, and obtain the rewriting trace output.  D-Hammer can be run
-interactively from the command line or integrated into other
-\CC\ projects as a library.
+We present \textbf{D-Hammer}, an open-source
+% and publicly available\footnote{https://github.com/LucianoXu/D-Hammer}
+implementation of our approach\footnote{The toll will be made publicly
+available in the final version of the paper}.  \textbf{D-Hammer} is an
+equational prover for Labelled Dirac notation written in \CC.  It
+features a parser built using ANTLR4, and scalar reasoning is powered
+by the Mathematica Engine. Users can use commands to make definitions
+and assumptions in the maintained context, conduct the normalization
+and equivalence checking, and obtain the rewriting trace output.
+D-Hammer can be run interactively from the command line or integrated
+into other \CC\ projects as a library.
 

### 2025-04-08 06related_conclusion.tex
.

@@
 
-\paragraph*{Comparison with DiracDec~\cite{diracdec}}
+\paragraph*{Comparison with DiracDec}
 Xu \emph{et al}~\cite{diracdec} define a language and an associate and
@@
 mechanics~\cite{DBLP:conf/lics/AbramskyC04}, a powerful framework for
-modeling and reasoning about quantum physics. Another main appeal of
-the ZX calculus is that ZX diagrams have an operational interpretation
-using graph rewriting. There is a large body of work that studies
-properties of these rewriting systems, and in particular completeness
-(a set of rules is complete if it can prove all valid identities) and
+modeling quantum physics. Another main appeal of the ZX calculus is
+that it has a natural operational interpretation based on graph
+rewriting. There is a large body of work that defines rewriting
+systems for fragments/extensions/variants of the ZX calculus, and
+studies their theoretical properties, in particular completeness (a
+set of rules is complete if it can prove all valid identities) and
 minimality (a complete set of rules is minimal if removing a rule
@@
 fragment~\cite{HNW18,JeandelPV18beyond}. Subsequent works generalize
-completeness to qudits~\cite{Poor2023} for a calculus know as ZXW
-which encompasses the ZX and ZW calculi~\cite{zwcalculus}, quantum
+completeness to qudits~\cite{Poor2023}, quantum
 circuits~\cite{DBLP:conf/lics/ClementHMPV23} or optical quantum
@@
 
-Because they are both grounded in rewriting techniques, there are many
-similarities between prior work on the ZX calculus and our own work on
-labelled Dirac notation. We highlight some key differences. First,
-there exist a number of differences between our language and the
-concrete fragments of the ZX-calculus for which completeness has been
-studied. For instance, our language features big operators, and
-operators for dual spaces. A second difference is that we do not claim
-completeness: our main theorem shows a reduction from Labelled Dirac
-notation to Dirac notation, for which completeness is only partially
-known. From a practical point of view, many of the examples handled by
-D-Hammer cannot be formalized in existing ZX-calculus tools. On the
-contrary, D-Hammer is not competitive on the examples that can be
-verified by ZX-calculus. \gb{give a few examples in terms of size and time}
+There are many similarities between the ZX calculus and our formalism
+for Labelled Dirac notation. However, there are also some differences
+between our formalism and the ones used in ZX calculus. In particular,
+our language supports big operators, tensors, and various operations
+on Hilbert spaces. These features are typically not considered in
+prior work on the ZX calculus, and as a consequence many examples
+handled by D-Hammer lack an immediate translation into ZX-calculus.
+This additional generality comes with a price. On the one hand, our
+theoretical results are weaker: we do not claim completeness or
+minimality. Similarly, practical implementations of the ZX-calculus,
+such as PyZX~\cite{kissinger2020pyzx}, outperform D-Hammer on examples
+that can be handled by both tools, as shown in \Cref{sec:eval}.
 
-\paragraph*{Comparison with quantum circuit equivalence checking tools}
-A main application of the ZX-calculus is to check equivalence of
-quantum circuits; for instance, PyZX~\cite{kissinger2020pyzx} is a
-popular Python library for automating equational reasoning in the
-ZX-calculus.
-
-There are many other tools for equivalence checking of quantum
-circuits. Amy~\cite{amy2018towards} uses the path-sums formalism
-formalism to check circuits with thousands of T gates. A series of
-works~\cite{AutoQ_pldi_2023,AutoQ2023,AutoQ_popl2025} explores
-automata-based approaches to verify quantum circuits at scale. Their
-main idea is to encode circuits as a compact, finite tree automata,
-often with $\mathit{poly}(n)$ transitions (instead of $2^n$
-transitions), where $n$ is the number of qubits of the state. By
-providing the semantics of quantum gates over this representation,
-they achieve practical verification. Their tool \textsc{autoq} is able
-to verify circuits with over 100,00 gates, and was recently extended
-to support parametrized verification.
+\paragraph*{Comparison with other tools}
+Beyond ZX calculus, there exists many other tools for simplifying and
+proving equivalence of quantum circuits; we refer the reader to recent
+surveys~\cite{QPV-survey2021,CharetonLVVBX23} for detailed accounts.
+Notable works include~\cite{amy2018towards}, which uses the path-sums
+formalism to check circuits with 1,000 of T gates,
+and~\cite{AutoQ_pldi_2023,AutoQ2023,AutoQ_popl2025}, which uses
+automata-based approaches to verify quantum circuits at scale---their
+tool \textsc{autoq} is able to verify circuits with over 100,00 gates,
+and was recently extended to support parametrized verification.
 

### 2025-04-09 01intro_preliminary.tex
.

@@
 literature on equivalence checking of (parametrized) quantum
-circuits.\gb{want to say our tool is not competitive?}
+circuits.
 

### 2025-04-09 03labelled.tex
.

@@
 registers and labelled terms. Here we display some of the rules and
-refer the reader to Appendix for the full set of rules.\gb{which appendix? add section}
+refer the reader to~\Cref{sec: full typing rules} for the full set of rules.
 \begin{gather*}
@@
   $D_1,D_2$.  Let $e_1 = e_2$ be obtained by above normalization
-  procedure on $D_1 = D_2$.  Then $e!=e_2$ is an equation in plain
+  procedure on $D_1 = D_2$.  Then $e_1=e_2$ is an equation in plain
   Dirac notation and $\Gamma\vDash D_1 = D_2$ if and only if
@@
 form of Eqn. (4), as desired. Details appear in
-Appendix~\ref{app:label_elim}.\gb{check proof is really there}
+Appendix~\ref{app:label_elim}
 %% \begin{hproof}

### 2025-04-09 05tool_eval.tex
.

@@
 We evaluate D-Hammer on several example sets, and make a comparison with the previous tool DiracDec~\cite{diracdec}.
-The experiments are carried out using a MacBook Pro with M3 Max chip. Results are summarized as follows.
+The experiments are carried out using a MacBook Pro with M3 Max chip. Results are summarized as follows, which indicates
+significant performance improvements.
 
@@
 
-The results show significant performance improvements. The enhanced
-efficiency primarily results from algorithmic improvements rather than
-the shift to a \CC implementation. This is supported by the fact that
-efficiency gains are more significant for larger examples, as can be
-observed in \Cref{fig: CoqQ plot}.
-
 \subsubsection{Textbook (QCQI)}
@@
 classic texbook~\cite{nielsen2010quantum}. All examples can be encoded
-in DiracDec and D-Hammer and are solved very very efficiently.
+in DiracDec and D-Hammer and are solved very efficiently.
 

### 2025-04-09 reviews-rebuttal.md
.

CAV 2025 Paper #83 Reviews and Comments
===========================================================================
Paper #83 D-Hammer: Efficient Equational Reasoning for Labelled Dirac
Notation

Review #83A
===========================================================================
* Updated: Mar 18, 2025

Overall merit
-------------
4. Accept

Reviewer expertise
------------------
3. Knowledgeable

Paper summary
-------------
Summary:
The paper introduces D-Hammer, a tool that enables equivalence checking of Dirac notation, including, for the first time, labeled Dirac notation. Their approach offers a complete normalization algorithm that avoids checking permutations of axioms that cannot be handled by simple term rewriting. This claim is supported by experimental evidence. A tool implementation is applied to examples of both plain and labeled notation and the results compare favorably with DiracDec (Xu et al., POPL’25).

From my perspective, the most interesting difference with DiacDec lies in how the D-Hammer language handles AC symbols. The main challenge, as already noted in Xu et al.'s paper, is that not all axioms in Dirac notation can be normalized through term rewriting alone. Both approaches address this by separating axioms into two sets: those that can be normalized by term rewriting (R) and those that cannot (E), with the latter primarily containing AC symbols.

In Xu et al.'s work, normalization is applied to R on both sides of the equality, after which equivalence is checked within E. This paper, on the other hand, takes a seemingly better approach. Instead of introducing auxiliary symbols with variable arities to handle E, the authors design a language that inherently supports AC symbols with indefinite arities. This allows them to apply specialized algorithms to normalize E directly, effectively normalizing both R and E together.

Comments for authors
--------------------
Strengths of the paper:
- The paper identifies the limitations of existing equivalence-checking tools designed for plain Dirac notation and presents a well-defined solution. This problem is effectively motivated in the introduction. Furthermore, since the most relevant expressions and equalities in quantum mechanics are often formulated in labeled Dirac notation, developing automated tools capable of handling this notation represents an essential advancement for the research community.
- The complete normalization approach.
- Another key strength of the proposed tool is its ability to overcome other practical limitations of DiracDec. Specifically, one of the biggest strengths of D-Hammer lies in its relationship with Mathematica. Since DiracDec is implemented within Mathematica, its behavior and typing rules are inherently constrained by Mathematica’s lack of a strong type system. In contrast, D-Hammer is programmed in C++, and is designed to benefit from a more structured term representation (e.g., AC operators with indefinite arity), a more expressive type system (e.g., dependent types), and a more efficient rewriting engine. While D-Hammer still interacts with Mathematica for reasoning about functions not natively supported by its rewriting rules, these interactions are carefully managed and do not negatively impact overall efficiency. This advantage is reflected in the experimental results, where D-Hammer consistently outperforms DiracDec in equivalence-checking tasks.

Weak points:
- It remains unclear whether the better results stem from a fundamental improvement in the algorithm (the complete normalization) or simply from a more efficient implementation—since the tool more often avoids using Mathematica as a backend. 
- Regarding the experimental evaluation, the results strongly suggest that D-Hammer outperforms DiracDec. However, it would have been more compelling to see a comparison using all 1031 lemmas from (Xu et al. POPL’25) to better assess how D-Hammer performs in a broader range of cases. For the evaluation of D-Hammer’s performance on labeled notation, solving the LDN-10 instance is impressive, but only 18 examples seem to fall short of asserting the tool’s performance.

Questions:
- I could not sufficiently check some of the key claims. For instance, Theorem 1 is stated without formal proof, whereas an equivalent theorem in Xu et al. (POPL’25) is given nearly a full page of justification in Section 7.2. This omission might be understandable if the authors use the same algorithm. However, since they claim their approach differs, it seems crucial that the paper at least discusses confluence and termination in more detail (maybe because they use special algorithms for set E, is it unnecessary?). Please clarify.
- A similar issue arises with Theorem 2, which asserts the soundness and completeness of the normalization procedure. Proving completeness is generally challenging; in fact, Xu et al. (POPL’25) explicitly state in Section 7.2 that they were unable to prove completeness and instead conjectured it. Furthermore, establishing completeness typically requires demonstrating confluence and termination, yet the paper does not address these concepts. Additionally, can the authors explain how they ensure that none of the expressions in Appendix D introduce circular dependencies?
- Did the authors consider using SMT solvers with non-linear arithmetic and trigonometric expressions instead of Mathematica?

Opinion:
I really enjoyed reading the paper. It is well-written and well-motivated, and the numerical results seem quite strong. I genuinely believe the work is valuable and could be even more useful than the plain Dirac notation case, particularly for applications in quantum communication and many-body physics. 
What would also be interesting is extending these methods to second quantization rules, where the algebra becomes even more tedious and complex. A development in that direction would represent a fundamental advancement.
As for the fit to CAV, I believe that the community automated reasoning methods can have an impact in the field of physics. For instance, the community of tensor networks is enormous and applies techniques very similar to those we use in decision diagrams. ZX calculus and model counting are also rapidly gaining popularity for these applications. We have but touched the surface of the possibilities here. Therefore, the paper fits at CAV in my opinion.

Post-rebuttal:
The rebuttal convinces me and resolves my concerns.

Minor comments
Small examples might help when introducing the syntax. For instance, how is the set product used for index sets?
p5 weak dependent —> weakly dependent
p9: alwasy
p17: quantum circuit —> quantum circuits
p36: step that discard  (what is the difference between 1 and 2 here?)
p37: Besids,
For future work, perhaps: The text primarily focuses on labeled Dirac notation due to its relevance in many-body physics. The second quantization formalism is also of significant interest. Since working with second quantization rules can be highly complex and prone to errors, especially due to the anti-commutation relations of fermionic operators, it would be interesting to explore whether your tool could be extended to incorporate second quantization. Could D-Hammer be extended to support this formalism?

Review #83B
===========================================================================

Overall merit
-------------
4. Accept

Reviewer expertise
------------------
3. Knowledgeable

Paper summary
-------------
The paper introduces D-Hammer, a tool for automated equational proof for labelled Dirac notation. It uses a dependently typed language to formalize labelled Dirac notation and defines an equational theory which is proved to be sound wrt a denotational semantics. The authors also propose a normalization procedure to prove equivalence of expressions. The tool is implemented in C++ and shown to have good performance.

Comments for authors
--------------------
The paper is well motivated. Since labelled Dirac notation is widely used to represent many-body quantum systems, equational reasoning of it is undoubtedly important. So far the reasoning has been done manually. The paper rigorously defines the syntax and semantics of a language for labelled Dirac notation as well as a tool for automated equational reasoning. D-Hammer is the counterpart of DiracDec for plain Dirac notation, but is more efficient as it is implemented in C++ while the latter is implemented in Mathematica. The basic idea of D-Hammer is similar to that of DiracDec, but there are notable differences because of the use of dependent types and the new implementation.

Theorem 2 states the soundness and completeness of normalization. As to the equational theory, only soundness is stated (Theorem 1). It would be interesting to discuss the completeness as well.

There are a lot of axioms and rewriting rules used by D-Hammer. How many of them are valid for plain Dirac notation (already used by DiracDec), and how many are newly added in this work?

Minor comments:
-	P.3, “as r,q consistent” -> “as r and 1 are consistent”
-	P.6, “compares where” -> “compares whether”
-	In Section 3.1, the syntax for types and terms are abstract. It would be better to give some concrete examples.

Review #83C
===========================================================================

Overall merit
-------------
2. Weak reject

Reviewer expertise
------------------
2. Some familiarity

Paper summary
-------------
This paper introduces D-Hammer, a tool for automating equational proofs in labelled Dirac notation, a formalism widely used in quantum computing and quantum programming. D-Hammer features a higher-order, dependently-typed language, an efficient normalization algorithm, and an optimized C++ implementation. The tool is evaluated on representative examples from both plain and labelled Dirac notation, demonstrating significant performance improvements over DiracDec (Xu et al., POPL’25) in the plain Dirac setting.

Comments for authors
--------------------
### Pros

+ The first tool to support automated equational proofs for labelled Dirac notation.

+ Achieves significantly faster equivalence verification compared to DiracDec.

### Cons

- The positioning of this work within the broader literature on equivalence reasoning in quantum computing is unclear. Please see my detailed questions below.

- It is not evident why existing equality saturation techniques cannot be applied to Dirac notation, provided that equivalence-preserving rewriting rules are well-defined.

### Detailed Comments

Thank you for the submission! While I am not an expert in quantum computing, I found this paper easy to follow, and the descriptions of the syntax and semantics of labelled Dirac notation are quite clear. However, I have two main technical concerns regarding this work, which I outline below:

- **Equivalence Checking in Quantum Computing**:  
  The central focus of this paper is equational reasoning for labelled Dirac notation, which seems closely related to a broader class of problems—*equivalence checking in quantum computing*. A well-established approach in this area is **ZX-calculus**, where any quantum circuit can be represented as a **ZX-diagram**. In this formalism, quantum gates such as **S, T, H, and CNOT** are represented as nodes in a graph, and a set of rewriting rules enables transformations and simplifications of these diagrams. Two circuits are considered equivalent if their ZX-diagrams can be reduced to the same (isomorphic) form.

- The approach proposed in this paper appears to share strong similarities with ZX-calculus. While this work does not use a graph-based representation, it explicitly expands tensor product computations. Additionally, the equivalence-preserving rewriting rules introduced in **Section 3.4** seem analogous to the rewriting rules used in ZX-diagrams. This raises the question: **What is the relationship between the techniques presented in this paper and ZX-calculus?**

- Given that ZX-calculus is also widely used for verifying quantum circuit equivalence, it would be valuable to compare D-Hammer against ZX-calculus-based methods. A direct comparison with ZX-calculus would make the evaluation more compelling. If the authors could clarify this connection and possibly include a comparison, I would reconsider my overall score.

  A relevant reference on ZX-calculus:  
  van de Wetering, John. *"ZX-calculus for the working quantum computer scientist."* arXiv preprint arXiv:2012.13966 (2020).

- **Technical Novelty and Equality Saturation**:  
  Another concern is the level of technical novelty in this work. The techniques presented here seem to apply equality saturation to a new domain—labelled Dirac notation. Given this, I wonder whether existing equality saturation techniques (e.g., egg [1]) could be directly utilized for reasoning in labelled Dirac notation. If not, it would be insightful for the authors to highlight the domain-specific challenges that prevent a direct application of egg or similar techniques.

  [1] Willsey, Max, Chandrakana Nandi, Yisu Remy Wang, Oliver Flatt, Zachary Tatlock, and Pavel Panchekha. *"Egg: Fast and extensible equality saturation."* Proceedings of the ACM on Programming Languages 5, no. POPL (2021): 1-29.

- **Figure 2 – Evaluation Clarifications**:  
  When comparing D-Hammer to DiracDec, the paper notes that D-Hammer omits certain rarely used features in CoqQ benchmarks, resulting in fewer verified cases. However, I am concerned that these omitted features might be the primary sources of performance overhead in DiracDec. If so, the performance improvement of D-Hammer may be due in part to excluding these computationally expensive cases rather than an inherent efficiency advantage. Could the authors provide more clarification on this point?

Response by Yingte Xu <yingte.xu@mpi-sp.org> (Author) (670 words)
---------------------------------------------------------------------------
We sincerely thank all reviewers for their comments, questions and innovative suggestions.

## Benchmark Evaluation and Clarification (A, C)
We will revise the paper to clarify the comparison with DiracDec. In particular, we will emphasize that:
- the enhanced efficiency primarily results from algorithmic improvements rather than the shift to a C++ implementation. This is supported by the fact that efficiency gains are more significant for larger examples, as can be observed in Figure 2.
- the omission of the projection rules has a limited influence on the performance evaluation. Only 4 examples involving projections are omitted, and the remaining cases are not affected by projection rules, so removing the projection feature does not influence too much.

Our paper presents a representative selection of labelled Dirac notation examples that can be handled by our tool. We will add further examples (~50) from [14,20,21] and possible applications in physics that can be handled by our tool—and they will be distributed with the tool.

## Comparison with ZX Calculus (C)
Labelled Dirac Notation (LDN) and ZX calculus have different scopes: the ZX calculus mainly focuses on quantum circuits, whereas LDN aims to provide general support for quantum physics. This difference in scope is reflected in the underlying languages: for instance,  many works in ZX calculus consider concrete sets of quantum gates, whereas LDN emphasizes the use of big operators. At a technical level, both the ZX calculus and LDN share the use of rewriting techniques. However, it is difficult to compare ZX and LDN: from a theoretical perspective, they pose different challenges. From a practical perspective, they target different examples, as observed in (Xu et al, POPL 2025). We will expand the related work section to include a longer comparison with ZX examples

## Comparison with Equality Saturation (C)
Our algorithm is based on term rewriting, and does not use equality saturation. We believe that equality saturation could be used to prove Dirac notation equations, but are not convinced that it would yield better results. Equality saturation has the advantage when it is hard to rewrite terms to a normal form. This may occur when we consider further extensions of labelled Dirac notation.

## Proof of Theorem 1 and 2 (A)
We provide the detailed proof of Theorem 1 and 2 in the **attached file** (rebuttal.pdf).

For Theorem 1 and soundness proof in Theorem 2, it suffices to prove that each rule is sound with respect to the axiomatic definition. The proof is routine.
The completeness proof for normalization in Theorem 2 is a relative completeness w.r.t. to the completeness of plain Dirac notation, and thus it is weaker and also easier to prove the confluence and termination of the whole system. The proof involves checking that step (1) - (3) can succeed rewriting every labelled Dirac formula to equation (4).

## Use SMT Solvers for Scalars (A)
Deciding the equivalence of scalars is not our main focus so we didn’t spend time exploring other tools. Also, SMT solvers have limited support for functions like sqrt(x) and exp(x) compared to Mathematica. But it is possible to try out with SMT solvers in the future.

## Rewriting Rules and Termination (A, B)
D-Hammer shares a common foundation for plain Dirac notation with DiracDec. There are 172 rules in DiracDec in total. We merged analogous rules by typing, removed 13 rules related to projections, and obtained 147 rules in D-Hammer. Part of the new procedure in D-Hammer are expressed as 8 rewriting rules. But it’s hard to quantify the difference as numbers of rewriting rules, because we use abstract rules and other algorithms.
The abstract and heterogeneous procedure also pose difficulty in proving confluence and termination. But we conjecture that termination still holds following the result from DiracDec, and no circular rewritings will happen.

## Applications in Physics (A)
We agree that exploring applications in physics is promising. Indeed, D-Hammer already possesses the capability to represent certain expressions in second quantization formalism. For instance, the creation operator in the occupation number representation is expressible as Sum[i, sqrt(i+1) |i+1><i|], enabling simplifications consistent with theoretical expectations.

Comment @A1 by Shepherd
---------------------------------------------------------------------------
 We thank the authors for their clear rebuttal, which proposes several improvements. The reviewers would like to see this materialize in the final version as follows: 
* An extended comparison with ZX calculus
* An empirical comparison with ZX calculus, for instance, for the circuit equivalence type queries found in Table 3.

### 2025-04-10 05tool_eval.tex
add equiavlence checking section

@@
 \subsubsection*{Equivalence checking of circuits}
-\gb{TODO}
+Quantum circuits are concrete computation models for qubit systems, and they can be described as unitary operators in Dirac notation. 
+D-Hammer can check the equivalence of small scaled quantum circuits through their Dirac notation representations. Nevertheless, this is not the intended application. 
+Its performance is not comparable with specialized algorithms and tools based on ZX-calculus or decision diagrams, which utilize the specific information of circuit classes to achieve better optimizations.
+An evaluation of some examples is given in~\Cref{fig: circuit eval}, and~\Cref{fig: QC-5} shows one of them. D-Hammer will always time out on larger quantum circuit examples.
+
+\begin{figure}[h]
+    \center
+    % \scriptsize
+    \setlength{\extrarowheight}{2pt}
+    \begin{tabular}{c @{\hspace{2em}} c @{\hspace{2em}} c @{\hspace{2em}} c}
+        \hline
+        example & D-Hammer & ZX-Calculus & Decision Diagrams \\
+        \hline
+        QC-1 & 0.029 & 0.0 & 0.0 \\
+        QC-2 & 0.16 & 0.00039 & 0.0038 \\
+        QC-3 & 0.29 & 5.7e-5 & 0.0057 \\
+        QC-4 & 0.013 & 4.3e-5 & 0.0017 \\
+        QC-5 & 15 & 6.4e-5 & 0.0044 \\
+        QC-6 & timeout & 0.00014 & 0.016 \\
+        \hline
+    \end{tabular}
+    \caption{Time consumptions (in seconds) for quantum circuit equivalence checking using different tools.}
+    \vspace{-0.2cm}
+    \label{fig: circuit eval}
+\end{figure}
+
+\vspace{-2em}
+
+\begin{figure}[h]
+    \begin{align*}
+        & \textbf{(circuit)}\ &&
+        \begin{quantikz}[scale=0.8]
+            \qw & \gate{R_x(\frac{\pi}{2})} & \gate{R_x(\frac{\pi}{2})} & \qw \\
+            \qw & \gate{T} & \qw & \qw \\
+        \end{quantikz}
+        =
+        \begin{quantikz}[scale=0.8]
+            \qw & \gate{H} & \gate{Z} & \gate{H} & \qw \\
+            \qw & \gate{T} & \gate{H} & \gate{H} & \qw \\
+        \end{quantikz} \\
+        & \textbf{(notation)}\ &&
+        \begin{aligned}
+            & (\textsf{Rx}(\pi/2) \otimes \ONEO) \cdot (\ONEO \otimes \textsf{T}) \cdot (\textsf{Rx}(\pi/2) \otimes \ONEO) \\
+            = & (\ONEO \otimes \textsf{H}) \cdot (\ONEO \otimes \textsf{H}) \cdot (\textsf{H} \otimes \ONEO) \cdot (\textsf{Z} \otimes \ONEO) \cdot (\textsf{H} \otimes \ONEO) \cdot (\ONEO \otimes \textsf{T})
+        \end{aligned}
+    \end{align*}
+    \caption{The quantum circuit and Dirac notation encoding for example QC-5.}
+    \label{fig: QC-5}
+\end{figure}

### 2025-04-10 05tool_eval.tex
add ref

@@
 D-Hammer can check the equivalence of small scaled quantum circuits through their Dirac notation representations. Nevertheless, this is not the intended application. 
-Its performance is not comparable with specialized algorithms and tools based on ZX-calculus or decision diagrams, which utilize the specific information of circuit classes to achieve better optimizations.
+Its performance is not comparable with specialized algorithms and tools based on ZX-calculus~\cite{9868772} or decision diagrams\cite{10.1145/3394885.3431590}, which utilize the specific information of circuit classes to achieve better optimizations.
 An evaluation of some examples is given in~\Cref{fig: circuit eval}, and~\Cref{fig: QC-5} shows one of them. D-Hammer will always time out on larger quantum circuit examples.
@@
         \hline
-        example & D-Hammer & ZX-Calculus & Decision Diagrams \\
+        example & D-Hammer & ZX-Calculus~\cite{9868772} & Decision Diagrams (simulation)~\cite{10.1145/3394885.3431590} \\
         \hline

### 2025-04-10 05tool_eval.tex
.

@@
 
-\subsubsection*{Equivalence checking of circuits}
-Quantum circuits are concrete computation models for qubit systems, and they can be described as unitary operators in Dirac notation. 
-D-Hammer can check the equivalence of small scaled quantum circuits through their Dirac notation representations. Nevertheless, this is not the intended application. 
-Its performance is not comparable with specialized algorithms and tools based on ZX-calculus~\cite{9868772} or decision diagrams\cite{10.1145/3394885.3431590}, which utilize the specific information of circuit classes to achieve better optimizations.
-An evaluation of some examples is given in~\Cref{fig: circuit eval}, and~\Cref{fig: QC-5} shows one of them. D-Hammer will always time out on larger quantum circuit examples.
+\subsubsection*{Quantum circuits}
+Quantum circuits is a prominent model of quantum computation. This
+model is adopted by numerous tools, which can evaluate, optimize, or
+prove equivalence of quantum circuits. These tools are based on a
+variety of approaches, based on ZX-calculus~\cite{9868772}, or
+decision diagrams\cite{10.1145/3394885.3431590}, or other formalisms
+discussed in related work. In general, these tools are aggressively
+optimized to achieve scalability.
+
+Quantum circuits can also be described as unitary operators in Dirac
+notation. Therefore, D-Hammer can check the equivalence of quantum
+circuits through their Dirac notation representations. Although it is
+not the intended application of D-Hammer, we include an evaluation of
+D-Hammer on some simple examples. An evaluation of some examples is
+given in~\Cref{fig: circuit eval}, and~\Cref{fig: QC-5} shows one of
+them. As expected, D-Hammer does not perform well in comparison with
+specialized tools: it is always clearly outperformed on small quantum
+circuit examples, and it always times out on larger quantum circuit
+examples. It would be of interest to make D-Hammer more competitive on
+quantum circuits in future work.
 

### 2025-04-10 05tool_eval.tex
.

@@
 variety of approaches, based on ZX-calculus~\cite{9868772}, or
-decision diagrams\cite{10.1145/3394885.3431590}, or other formalisms
-discussed in related work. In general, these tools are aggressively
+decision diagrams~\cite{10.1145/3394885.3431590}, or other formalisms
+discussed in \Cref{sec:relwork}. In general, these tools are aggressively
 optimized to achieve scalability.
@@
 them. As expected, D-Hammer does not perform well in comparison with
-specialized tools: it is always clearly outperformed on small quantum
-circuit examples, and it always times out on larger quantum circuit
-examples. It would be of interest to make D-Hammer more competitive on
-quantum circuits in future work.
+specialized tools: it is always outperformed, often by several order
+of magnitude, on small quantum circuit examples, and it always times
+out on larger quantum circuit examples. In the future it would be of
+interest to make D-Hammer more competitive on quantum circuits by
+adopting some of the aggressive optimizations for other works.
 

### 2025-04-11 06related_conclusion.tex
.
fix tool referenece

@@
-\section{Related Work}
+\section{Related Work}\label{sec:relwork}
 
@@
 theoretical results are weaker: we do not claim completeness or
-minimality. Similarly, practical implementations of the ZX-calculus,
-such as PyZX~\cite{kissinger2020pyzx}, outperform D-Hammer on examples
+minimality. Similarly, practical implementations of the ZX-calculus~\cite{kissinger2020pyzx,9868772} outperform D-Hammer on examples
 that can be handled by both tools, as shown in \Cref{sec:eval}.

### 2025-05-13 01intro_preliminary.tex
Update

@@
 In labelled Dirac notation, bras and kets are tagged with labels to identify the subsystems they operate on.
-For example, the labelled tensor expression \( \ket{\psi}_{S'} \otimes \ket{\phi}_{S} \) indicates that \( \ket{\psi}_{S} \) and \( \ket{\phi}_{S'} \) describe two quantum states over some subsystem $S$ and $S'$, respectively.
+For example, the labelled tensor \( \ket{\psi}_{S'} \otimes \ket{\phi}_{S} \) indicates that \( \ket{\psi}_{S} \) and \( \ket{\phi}_{S'} \) describe two quantum states over subsystem $S$ and $S'$, respectively.
 By considering the relationship between $S$ and $S'$, one can obtain identities for free, e.g.\,

### 2025-05-13 05tool_eval.tex
Update

@@
 We present \textbf{D-Hammer}, an open-source
-% and publicly available\footnote{https://github.com/LucianoXu/D-Hammer}
-implementation of our approach\footnote{The toll will be made publicly
-available in the final version of the paper}.  \textbf{D-Hammer} is an
+and publicly available\footnote{\texttt{https://github.com/LucianoXu/D-Hammer}}
+implementation of our approach.  \textbf{D-Hammer} is an
 equational prover for Labelled Dirac notation written in \CC.  It
@@
 \end{figure}
+
+\vspace{-2em}

### 2025-05-13 06related_conclusion.tex
Update

@@
 similar pattern. However, there are significant differences in terms
-of scope, expressiveness, and efficiency. We review these differences
-below.
+of scope, expressiveness, and efficiency. 
+% We review these differences below.
 
@@
 
+\subsection*{Acknowledgements}
+\small
+We thank the anonymous reviewers for their constructive feedback that helped improve this paper.
+We also extend our gratitude to Jam Kabeer Ali Khan and Ivan Ariel Renison for their work of testing of D-Hammer.
+
+\normalsize
+
+%% Disclosure of Interests
+\subsubsection*{Disclosure of Interests}
+\small
+This research was supported by the National Key R\&D Program of China under Grant No. 2023YFA1009403.
+\normalsize
+
 %% Based on the first equational reasoning tool for Dirac notation called DiracDec, this work improves and extends the theory for practical applications, and provides the solver D-Hammer.

### 2025-05-14 02language_design.tex
CAV'25 Camera Ready

@@
 Assumptions, on the other hand, define the types of variables.
-We say an expression \( t \) has type \( X \) in context \( \Gamma \) if the typing judgment \( \Gamma \vdash t : X \) can be proven through the rules in~\Cref{sec: full typing rules}.
-These are two instances:
+We say an expression \( t \) has type \( X \) in context \( \Gamma \) if the typing judgment \( \Gamma \vdash t : X \) can be proven through the rules in Appendix A. These are two instances:
 \begin{gather*}
@@
 result.  The complete interpretation of terms and types is provided
-in~\Cref{sec: full denotational sem}.
+in Appendix B.
 
@@
 The proof system for equivalence is based on equational logic, together with axioms that describe the properties of Dirac notation.
-A full list of these axioms can be found in~\Cref{sec: full axioms}.
+A full list of these axioms can be found in Appendix C.
 The axioms cover fundamental aspects of linear spaces, as well as other structures like the tensor and inner products.
@@
 For example, the term $(x\times y) . \ket{t} + \ket{t}$ is matched by the rule $a.K + K\ \reduce\ (a + 1).K$, and is rewritten into $(x\times y + 1) . \ket{t}$.
-Step 1 and 3 use the same set of rewriting rules in~\Cref{sec: rewriting rules}.
+Step 1 and 3 use the same set of rewriting rules in Appendix D.
 Step 2 expands variables to their abstract element-wise representation, e.g., $K \ \reduce\ \sum_{i} (\bra{i} \cdot K). \ket{i}$, which is useful when reasoning about sums.
@@
 The final step uses de Bruijn indices~\cite{deBruijn1972lambda} to resolve $\alpha$-equivalence.
-For further details, refer to~\Cref{sec: decide}.
+For further details, refer to Appendix E.
 

### 2025-05-14 03labelled.tex
Change Appendix D.1 title
CAV'25 Camera Ready

@@
 registers and labelled terms. Here we display some of the rules and
-refer the reader to~\Cref{sec: full typing rules} for the full set of rules.
+refer the reader to Appendix A for the full set of rules.
 \begin{gather*}
@@
 We can further define $\Swap_{s_1,s_2,\cdots}$ for merging disjoint sets orderly.
-See Appendix \ref{sec: full denotational sem} for details.
+See Appendix B for details.
 
@@
 \end{align*}
-where $|i_R\>$ and $\<i_{R'}|$ are constructed by tensoring the basis according to the structure of $R$ (see Appendix \ref{sec: full denotational sem}). The rules for $e_R$ (ket and bra) are similar.
+where $|i_R\>$ and $\<i_{R'}|$ are constructed by tensoring the basis according to the structure of $R$ (see Appendix B). The rules for $e_R$ (ket and bra) are similar.
 %This step reduces all labelled terms $e_R$ or $e_{R;R}$.
@@
 preserves the type and semantics and thus yield expressions of the
-form of Eqn. (4), as desired. Details appear in
-Appendix~\ref{app:label_elim}
+form of Eqn. (4), as desired. Details appear in Appendix D.1.
 %% \begin{hproof}

### 2025-05-14 05tool_eval.tex
CAV'25 Camera Ready

@@
     \item \texttt{\textcolor{NavyBlue}{Check} term.} Type checking the \texttt{term} and output the result.
-    \item \texttt{\textcolor{NavyBlue}{Normalize} term.} Normalize the \texttt{term} using the algorithm introduced in~\Cref{sec: decide}.
+    \item \texttt{\textcolor{NavyBlue}{Normalize} term.} Normalize the \texttt{term} using the algorithm in Appendix E.
     \item \texttt{\textcolor{NavyBlue}{CheckEq} term \textcolor{NavyBlue}{with} term.} Check the equivalence of the two terms calculating and comparing their normal forms.
@@
     \end{tabular}
-    \caption{Part of examples for labelled Dirac notations. See~\Cref{sec: examples for labelled} for the full list.}
+    \caption{Part of examples for labelled Dirac notations. See Appendix F for the full list.}
     \vspace{-0.2cm}

### 2025-05-20 main.tex
.
update the proof
add equiavlence checking section
Update
CAV'25 Camera Ready
add badge

@@
 % \usepackage[margin=0.8in]{geometry}
+
+% \usepackage{showframe} % print page margins, remove when positioning is satisfying
+\usepackage{graphicx} % necessary for inserting .pdfs
+\usepackage{hyperref} % "available" badge is a link to the actual DOI
+\usepackage[firstpage]{draftwatermark} % free badge placement
+
+\SetWatermarkAngle{280}
+
+\SetWatermarkText{\raisebox{9cm}{
+\hspace{4.8cm}
+\href{https://doi.org/10.5281/zenodo.15206045}{\includegraphics{1-available}}
+}}
 
@@
 
+\usepackage{quantikz}
+
 \newenvironment{ruletable}[1]
@@
 
+\usepackage[misc]{ifsym}
 %
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% %
-% %\titlerunning{Abbreviated paper title}
-% % If the paper title is too long for the running head, you can set
-% % an abbreviated paper title here
-% %
-% \author{First Author\inst{1}\orcidID{0000-1111-2222-3333} \and
-% Second Author\inst{2,3}\orcidID{1111-2222-3333-4444} \and
-% Third Author\inst{3}\orcidID{2222--3333-4444-5555}}
-% %
-% \authorrunning{F. Author et al.}
-% % First names are abbreviated in the running head.
-% % If there are more than two authors, 'et al.' is used.
-% %
-% \institute{Princeton University, Princeton NJ 08544, USA \and
-% Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
-% \email{lncs@springer.com}\\
-% \url{http://www.springer.com/gp/computer-science/lncs} \and
-% ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
-% \email{\{abc,lncs\}@uni-heidelberg.de}}
+%
+%\titlerunning{Abbreviated paper title}
+% If the paper title is too long for the running head, you can set
+% an abbreviated paper title here
+%
+\author{Yingte Xu\inst{1}\orcidID{0000-0001-9071-7862} \and
+Li Zhou\inst{2}\orcidID{0000-0002-9868-8477} \and
+\\
+Gilles Barthe\inst{1,3}\orcidID{0000-0002-3853-1777} \Letter}
+%
+\authorrunning{Yingte Xu, Li Zhou, and Gilles Barthe}
+% First names are abbreviated in the running head.
+% If there are more than two authors, 'et al.' is used.
+%
+\institute{MPI-SP, Germany
+\email{\{yingte.xu, gilles.barthe\}@mpi-sp.org}
+\and
+Key Laboratory of System Software (Chinese Academy of Sciences) and State Key Laboratory of Computer Science, Institute of Software, Chinese Academy of Sciences, China.
+\email{zhouli@ios.ac.cn}
+\and
+IMDEA Software Institute, Spain}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \renewcommand{\>}{\rangle}
+\newcommand {\cD } {{\mathcal{D}}}
+\newcommand {\cl } {{\mathit{cl}}}
+
+\newenvironment{hproof}{%
+  \renewcommand{\proofname}{Proof Sketch}\proof}{\endproof}
 
@@
 
-\input{04efficient_alg.tex}
+%\input{04efficient_alg.tex}
 
@@
 
-\appendix
-\include{appendix}
+% \appendix
+% \include{appendix}
 
